### 2014-07-25 TOMS_Paper/assembly.tex
Add basic stuff and empty paper skeleton

% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subfig}
\usepackage{tabulary}
\usepackage{lineno}

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}

\usepackage{listings}
\lstset{language=C}

\usepackage[cmex10]{amsmath}
\usepackage{url}

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
%\acmVolume{9}
%\acmNumber{4}
%\acmArticle{39}
%\acmYear{2010}
%\acmMonth{3}

% Document starts
\begin{document}

% Page heads
\markboth{F. Luporini et al.}{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}

% Title portion
\title{Expression Re-writing and Code Specialization for Finite Element Integration}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H.J. Kelly
\affil{Imperial College London}}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

\begin{abstract}
The numerical solution of partial differential equations using the
finite element method is one of the key applications of high performance
computing. Local assembly is its characteristic operation. This entails
the execution of a problem-specific kernel to numerically evaluate an
integral for each element in the discretized problem domain. Since the
domain size can be huge, executing efficient kernels is fundamental.
Their optimization is, however, a challenging issue. Even though affine
loop nests are generally present, the short trip counts and the
complexity of mathematical expressions make it hard to determine a
single or unique sequence of successful transformations. Therefore, we
present the design and systematic evaluation of COFFEE, a
domain-specific compiler for local assembly kernels. COFFEE manipulates
abstract syntax trees generated from a high-level domain-specific
language for PDEs by introducing domain-aware composable optimizations
aimed at improving instruction-level parallelism, especially SIMD
vectorization, and register locality. It then generates C code including
vector intrinsics. Experiments using a range of finite-element forms of
increasing complexity show that significant performance improvement is
achieved.
%
%The code is generated from a high level language for PDEs. An
%invocation of the kernel typically does tensor-like calculators over
%relatively small matrices; e.g., triply nested loops operating on
%about a dozen iterations. Because of the DSL context, COFFEE has a lot
%of flexibility on how it organizes/orders the calculations to tailor
%execution to a target architecture.
%
\end{abstract}

\category{G.1.8}{Numerical Analysis}{Partial Differential Equations -
  Finite element methods}

\category{G.4}{Mathematical Software}{Parallel and vector
  implementations}

\terms{Design, Performance}

\keywords{Finite element integration, local assembly, compilers,
  optimizations, SIMD vectorization}

\acmformat{Fabio Luporini, Ana Lucia Varbanescu, Florian Rathgeber,
  Gheorghe-Teodor Bercea, J. Ramanujam, David A. Ham, and Paul
  H. J. Kelly, 2014. COFFEE: an Optimizing Compilerfor Finite Element
  Local Assembly.}

% At a minimum you need to supply the author names, year and a title.
% IMPORTANT: Full first names whenever they are known, surname last,
% followed by a period.  In the case of two authors, 'and' is placed
% between them.  In the case of three or more authors, the serial
% comma is used, that is, all author names except the last one but
% including the penultimate author's name are followed by a comma, and
% then 'and' is placed before the final author's name.  If only first
% and middle initials are known, then each initial is followed by a
% period and they are separated by a space.  The remaining information
% (journal title, volume, article number, date, etc.) is
% 'auto-generated'.

\begin{bottomstuff}

This research is partly funded by the MAPDES project, by the
Department of Computing at Imperial College London, by EPSRC through
grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants
NE/K008951/1 and NE/K006789/1, by the U.S.  National Science
Foundation through grants 0811457 and 0926687, by the U.S. Army
through contract W911NF-10-1-000, and by a HiPEAC collaboration
grant. The authors would like to thank Dr. Carlo Bertolli,
Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
suggestions and their contribution to the Firedrake project.

Author's addresses: Fabio Luporini $\&$ Paul H. J. Kelly, Department of Computing,
Imperial College London; David A. Ham, Department of Computing and
Department of Mathematics, Imperial College London; 
\end{bottomstuff}

\maketitle

%Computational cost is a critical limitation in scientific computing,
%especially for finite element simulations. To provide one particular
%example we are particularly concerned about, it has been well
%established that mesh resolution Consequently, our aggressive
%optimization of local assembly, which may even take up to 60% of the
%overall FEM's execution time, directly impacts the performance of
%large-scale scientific simulations running on supercomputers.

\section{Introduction}

\section{Preliminaries}
\label{sec:background}
%Here we discuss about the computational characteristics of
%computational science kernels. Briefly cite op2 kernels. Emphasis on
%Finite Element Assembly. Generalization and formalization of a Finite
%Element Assembly kernel using quadrature representation.

%\NoCaptionOfAlgo
%\RestoreCaptionOfAlgo

\section{Automated Expression Re-writing}
\label{sec:code-transf}

%\begin{algorithm}[t]
%\SetAlgorithmName{LISTING}{}
%\footnotesize
%\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
%~~// K, det = Compute Jacobian (coords) \\
%~~\\
%~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
%~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
%~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
%~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
%~~~~~~\KwSty{$\#$pragma vector aligned}\\
%~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
%~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
%~~$\rbrace$\\
%$\rbrace$
%\caption{Local assembly code for the Helmholtz problem in
%  Listing~\ref{code:helmholtz} after application of padding, data
%  alignment, and \emph{licm}, for an AVX architecture. In this
%  example, sub-expressions invariant to \texttt{j} are identical to
%  those invariant to \texttt{k}, so they can be precomputed once in
%  the $r$ loop.}
%\label{code:helmholtz-licm}
%\end{algorithm}

\section{Code Specialization}

%\begin{figure}[b]
%\begin{center}
%\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
%\caption{High-level view of Firedrake. COFFEE is at the core,
%  receiving ASTs from a modified version of the FEniCS Form compiler
%  and producing optimized C code kernels.}
%\label{fig:coffee-pipeline}
%\end{center}
%\end{figure}

\section{Performance Evaluation}
\label{sec:perf-results}

\subsection{Experimental Setup}

%Experiments were run on a single core of two Intel architectures, a
%Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB
%L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode,
%32KB L1 cache and 512KB L2 cache). We have chosen these two
%architectures because of the differences in the number of logical
%registers and SIMD lanes, which can impact the effectiveness of the
%optimization strategy. The \texttt{icc 13.1} compiler was used. On the
%Sandy Bridge, the compilation flags used were \texttt{-O2} and
%\texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization
%level \texttt{-O3} was used. Other optimization levels performed, in
%general, slightly worse.

%\newcommand{\licmapresultsnorms}{
%\begin{tabulary}{1.0\textwidth}{cccccc|cccc}
%\cline{3-10}
%& & \multicolumn{4}{c}{\texttt{Sandy Bridge}} & \multicolumn{4}{c}{\texttt{Xeon Phi}} \\
%\cline{1-10}
%\texttt{problem} & \texttt{shape} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
%\texttt{Helmholtz} & \texttt{triangle} & 1.32 & 1.88 & 2.87 & 4.13 & 1,50 & 2,41 & 1,30 & 1,96 \\
%\texttt{Helmholtz} & \texttt{tetrahedron} & 1.35 & 3.32 & 2.66 & 3.27 & 1,41 & 1,50 & 2,79 & 2,81 \\
%\texttt{Helmholtz} & \texttt{prism} & 2.63 & 2.74 & 2.43 & 2.75 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
%%\cline{1-18}
%\texttt{Diffusion} & \texttt{triangle} & 1.38 & 1.99 & 3.07 & 4.28 & 1,08 & 1,88 & 1,20 & 1,97\\
%\texttt{Diffusion} & \texttt{tetrahedron} & 1.41 & 3.70 & 3.18 & 3.82 & 1,05 & 1,51 & 2,76 & 3,00\\
%\texttt{Diffusion} & \texttt{prism} & 2.55 & 3.13 & 2.73 & 2.69 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
%%\cline{1-18}
%\texttt{Burgers} & \texttt{triangle} & 1.56 & 2.28 & 2.61 & 2.77 & 2,84 & 2,26 & 3,96 & 4,27 \\
%\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.10 & 1.60 & 1.78 & 1,48 & 3,83 & 1,55 & 1,29 \\
%\texttt{Burgers} & \texttt{prism} & 2.19 & 2.32 & 1.64 & 1.42 & 2,18 & 2,82 & 1,24 & 1,25 \\
%\cline{1-10}
%\end{tabulary}
%}
%
%\begin{table*}[t]
%\tbl{Performance improvement due to generalized loop-invariant code
%  motion, data alignment, and padding, for different element shapes
%  (triangle, tetrahedron, prism) and polynomial orders ($p \in [1,
%    4]$), over the original non-optimized code, for the Helmholtz,
%  Diffusion and Burgers problems.}{ \scriptsize \licmapresultsnorms }
%\label{table:perf-results-licmap}
%\end{table*}

\section{Related Work}
\label{sec:related-work}

\section{Conclusions}
\label{sec:conclusions}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{biblio}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

\medskip

\end{document}

### 2014-07-25 notes/concept.tex
Add basic stuff and empty paper skeleton

\documentclass[a4paper]{article}
\usepackage{authblk}
\bibliographystyle{plainnat}
\usepackage[left=1.4in, right=1.4in]{geometry}
\setlength{\parskip}{0.5cm}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{soul}

\title{A few notes}
\date{}

\begin{document}
\lstset{language=C, breaklines=true}

\maketitle

\section{Prototyping a language and a compiler for unstructured mesh computations}
This section tries to delineate the research path done so far and the potential future work. In no ways these things will be useful for the next (if any) COFFEE paper, but hopefully they will clarify my (and your) mind about the status of the research. As of October 2013, I have worked on two projects 
\begin{itemize}
\item Sparse tiling for irregular computations: ST
\item A compiler for optimizing the sequential code of finite element integration (local assembly): COFFEE
\end{itemize}
Despite the original motivations for the two projects arise from different areas - finite element for COFFEE, finite volume/molecular dynamics and, more in general, irregular computations for ST - I think much ground is in common. For example, the programming (at least at a certain level of abstraction) and execution models of the motivating applications are identical: OP2 establishes a way of defining \emph{loop chains}, in which, as you all know, the computation is structured as a sequence of synchronized calls to kernels applied all over the mesh. Given that, ST and COFFEE can be abstracted into a single framework. Such a framework would formalize the way we optimize unstructured mesh applications.

What do I mean by framework? This is all about being aware of the domain in which we operate, in order to be able to introduce powerful optimizations. 
\begin{itemize}
\item Unstructured mesh computations are, in a certain sense, special: they can be \textit{always} represented as a sequence of loop nests, each loop nest being characterized by the presence of a non-affine outer loop (over mesh elements). We all know that compilers tend to fail at optimizing non-affine loops, for obvious reasons. \ul{We want to be in a position that allows us to optimize individual, non-affine loops}. 
\item Such a sequence of loop nests is known, at compile/code-generation time, and described by a loop chain abstraction, which also tells how the non-affine accesses are distributed (by a sequence of \textit{opparloop} calls in OP2). \ul{We want to exploit the knowledge of the loop chain, e.g. for inspector/executor strategies}.
\item In specific domains, kernels have a particular structure: for example, an affine (perfect or non-perfect) loop nest, in which mathematical expressions have certain properties, loops are fully parallel, etc. \ul{We want to exploit domain-knowledge}.
\end{itemize}
\textit{By themselves, these observations do not introduce anything new: however, what is important here is the formalization of a framework that allows exploiting these three different potential sources of optimization in a systematic way.}

It is worth highlighting the differences with respect to the standard OP2-like programming model:
\begin{itemize}
\item We will have two domain-specific languages: one is obviously an OP2-like language to write loop chains; the other, let's call it $X$ for the moment, will be used to drive the optimization process. The optimization process may involve transforming the kernel code, as well as specializing the wrapper code.
\item $X$ is used to specify multi-level optimizations:
	\begin{enumerate}
	\item Across the loop chain (e.g. ST, kernel fusion, kernel fission)
	\item Within a single loop of the loop chain (e.g. Inter-kernel vectorization; )
	\item Within a kernel (e.g. COFFEE's optimizations, i.e. domain-aware optimizations~\footnote{note that domain-aware is different than domain-specific.})
	\end{enumerate}
	And all of these optimizations are composable. This is summarized in Figure~\ref{fig:framework}, along with some examples of $X$ constructs expressed as \ul{pragma notation}. 
\item The output of our framework would be the input to the plan function, which would be responsible for scheduling the parallelism exposed in the input (e.g. standard execution of parallel loops with a synchronization between the execution of two loops, with kernels executed sequentially; parallel (threaded) execution of a kernel, if $X$ exposed intra-kernel parallelism; lazyness; various techniques for communication-computation overlap; etc.). 
\end{itemize}

\section{Optimizing finite element integration through expression rewriting and code specialization}
This is basically a \textbf{generalization} of the COFFEE paper. We claim an optimizer for local assembly routines needs to go through two neatly separated steps: expression rewriting and code specialization.
\begin{itemize}
\item \textbf{Expression rewriting.} \textit{Intuitively}, at this level we do transformations that minimize the amount of redundant computation. This is an interesting problem that involves several steps (here in random order):
	\begin{enumerate}
	\item factorization of common terms (a*c + a*b $\rightarrow$ a*(b+c))
	\item loop-invariant code motion (note that factorization at point 1 can expose more invariant terms)
	\item expansion of terms (note that this can expose more factorization opportunities - e.g. (b+c)*a + b*d + c*e$\rightarrow$ b*(a + c) + c*(a + e))
	\item eliminate arithmetic operations involving zero-valued terms (this requires re-writing loop bounds, splitting expressions, etc)
	\end{enumerate}  
Of these, only point 2 was presented in the COFFEE paper. The idea here is to formalize a set of rewriting rules (similar in spirit to SPIRAL ones? maybe not...) that deterministically bring a given mathematical expression from its original form to a new form in which we have to do less floating point operations.
\item \textbf{Code specialization.}
In this phase, we specialize and optimize code for the underlying architecture. We have several possibilities. Some of them were in the COFFEE paper already.
	\begin{enumerate}
	\item Padding, data alignment, trip count adjustment. This is for improving simd-ization.
	\item Loop fission based on expression splitting (i.e. on commutativity of some operations)
	\item Outer-product vectorization
	\end{enumerate}
The outer-product vectorization is just an example of something that can be generalized as: we know the memory access pattern of the expression, can we write specialized code for it? In general, we may think to ask a compiler ''Hey, it's really important to optimize the execution of this expression in this loop nest, can you check whether you know how to generate specialized code for this notably important memory access pattern?''. So the outer-product vectorization becomes a possible optimized implementation of a given expression in a certain iteration space. Another possibility is to transform the computation, for instance, in a sequence to BLAS calls. So, we have more opportunities here.
On top of these things, we have to consider that many more code variants should be tried because: 
\begin{itemize}
\item we may want to try a different loop interchange (partly neglected in the COFFEE paper)
\item we may want to manually unroll-and-jam loops (neglected in the COFFEE paper)
\item we may want to try different ways of re-writing the expression! Especially the elimination of arithmetic operations involving zero-valued terms poses several questions since it leads to break the original iteration space into multiple loop nests, in which padding/data alignment/trip count adjustment would not be safe anymore. For example, compare Figures 1 and 2. In the transformed code in Figure 2, the first two \texttt{jk} loop nests in which $A$ is evaluated are not decorated with \texttt{pragma alignment/simd} because adjusting loop bounds would not be safe and accesses would not be aligned to the vector length. This is a pity because alignment/padding, as shown in the COFFEE paper, have a great impact. To make use of them, the only possibility is to store partial evaluations from different loops in temporary arrays, and then store them back into $A$ at the end of the computation (see Figure 3). So what is the overhead of this approach? What is it best to do? Should we lose the benefit of padding/alignment, or should we use temporary arrays as in Figure 3, or should we just forget about the elimination of zeros (\textit{in a given problem})? 
\end{itemize}

\begin{figure}
\begin{center}
\includegraphics[scale=1.0]{pictures/coffee-scheme}
\label{fig:newcoffee}\caption{Overview of COFFEE. Numbers 1-2-3-4 indicate the flow.}
\end{center}
\end{figure}

\begin{figure}
\label{original-code}\caption{Original code}
\scriptsize
\lstinputlisting{listings/original.code}
\end{figure}

\begin{figure}
\label{zeros-code}\caption{Code after expression re-writing. Operations involving zero-valued entries in the arrays are removed by suitably adjusting the iteration space.}
\scriptsize
\lstinputlisting{listings/zeros.code}
\end{figure}

\begin{figure}
\label{zeros-tmps-code}\caption{Code after expression re-writing plus using temporaries to ensure aligned accesses.}
\scriptsize
\lstinputlisting{listings/zeros-tmps.code}
\end{figure}

Therefore, a possible thing to do for specializing code in a certain problem is
\begin{itemize}
\item Autotuning, based on domain-knowledge to keep the exploration space small enough
\end{itemize}
which would replace the cost model introduced in the COFFEE paper (here, we have more code variants). A scheme of the new COFFEE is in Figure~\ref{fig:newcoffee}
\end{itemize}

%We will have to prove  in significant applications showing also the speedup in the overall computation.

%\begin{thebibliography}{99}
%\setlength{\parskip}{0pt}
%
%\bibitem{FFC-Compiler} Robert~C. Kirby and Anders Logg. 
%\newblock A compiler for variational forms.
%\newblock {\em ACM Trans. Math. Softw.}, 32(3):417--444, September 2006.
%
%\end{thebibliography}

\end{document}

### 2014-07-28 TOMS_Paper/assembly.tex
Start writing a bit of preliminaries

@@
 \usepackage{listings}
-\lstset{language=C}
+\lstset{language=C, breaklines=true}
 
@@
 % Page heads
-\markboth{F. Luporini et al.}{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
+\markboth{F. Luporini et al.}{Optimizing Finite Element Integration through Automated Expression Re-writing and Code Specialization}
 
 % Title portion
-\title{Expression Re-writing and Code Specialization for Finite Element Integration}
+\title{Optimizing Finite Element Integration through Automated Expression Re-writing and Code Specialization}
 \author{Fabio Luporini
@@
 \begin{abstract}
-The numerical solution of partial differential equations using the
-finite element method is one of the key applications of high performance
-computing. Local assembly is its characteristic operation. This entails
-the execution of a problem-specific kernel to numerically evaluate an
-integral for each element in the discretized problem domain. Since the
-domain size can be huge, executing efficient kernels is fundamental.
-Their optimization is, however, a challenging issue. Even though affine
-loop nests are generally present, the short trip counts and the
-complexity of mathematical expressions make it hard to determine a
-single or unique sequence of successful transformations. Therefore, we
-present the design and systematic evaluation of COFFEE, a
-domain-specific compiler for local assembly kernels. COFFEE manipulates
-abstract syntax trees generated from a high-level domain-specific
-language for PDEs by introducing domain-aware composable optimizations
-aimed at improving instruction-level parallelism, especially SIMD
-vectorization, and register locality. It then generates C code including
-vector intrinsics. Experiments using a range of finite-element forms of
-increasing complexity show that significant performance improvement is
-achieved.
+Abstract goes here
 %
@@
 
-\acmformat{Fabio Luporini, Ana Lucia Varbanescu, Florian Rathgeber,
-  Gheorghe-Teodor Bercea, J. Ramanujam, David A. Ham, and Paul
-  H. J. Kelly, 2014. COFFEE: an Optimizing Compilerfor Finite Element
-  Local Assembly.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2014. 
+  Optimizing Finite Element Integration through Automated Expression Re-writing and Code Specialization.}
 
@@
 
+%This paper can be thought of as a continuation of the work presented in~\cite{Luporini}: in particular, we answer the remaining open questions, generalize and formalize an approach to minimize redundant computation using a similar philosophy, we show ~\cite{quadrature1}.
+
 \section{Preliminaries}
 \label{sec:background}
+
+\subsection{Quadrature for Finite Element Local Assembly}
 %Here we discuss about the computational characteristics of
@@
 
-%\NoCaptionOfAlgo
-%\RestoreCaptionOfAlgo
+\subsection{Code Generation for Quadrature Representation}
+Rapid implementation of high performance, robust, and portable code evaluating element matrices using quadrature can be achieved through automated code generation.
+This has been succesfully proved in the context of the popular FEniCS project~\cite{Fenics}.
+The FEniCS Form Compiler (FFC) accepts as input the variational form of a partial differential equation and generates C++ code implementing local assembly routines.
+The variational form is expressed at high-level by means of the domain-specific Unified Form Language (UFL).
+Local assembly code must be high performance: as the complexity of a form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen functions, the resulting kernels evaluating element matrices become more computationally expensive, which impacts significantly the run-time of the overall computation.
 
-\section{Automated Expression Re-writing}
-\label{sec:code-transf}
+Achieving high performance is non-trivial due to the complexity of the mathematical expressions involved in the numerical integration and because of the small sizes of loops and accessed arrays.
+In~\cite{quadrature1},~\cite{Kirby-FEM-opt} and~\cite{Francis}, it is shown how automated code generation allows the introduction of powerful optimizations, which a user cannot be expected to write ``by hand'', as well as the exploration of non-standard integration techniques, based, for instance, on symbolic manipulation.
+In~\cite{Luporini} we made one step forward by showing that different problems require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations, which are not supported by available compilers, are essential to maximize instruction-level parallelism and register locality.
+
+\subsection{Summary of Low-level Optimizations and Results for Quadrature Representation}
+To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, we summarize in this section the results of our previous work on automated code transformation for quadrature representation.
+
+...TODO...
+
+% Far notare che low-level si, ma sono guidate da high-level knowledge! Questo in risposta alla domanda: ma allora perche' non sfruttare la domain knowledge che hai, che senso ha separare...
+
+% Ci sono due goal: minimizzare flops e specializzare
+
+% Vantaggio di automatizzare rispetto che scrivere a mano
+% which would be practically unfeasible to achieve by hand-written code
+
+\section{Expression Re-writing}
+\label{sec:expr-rewriting}
+
+\begin{figure}
+\scriptsize
+\lstinputlisting{listings/simple.code}
+\label{original-code}\caption{Original code}
+\end{figure}
+
+\subsection{Re-writing Rules}
+
+\subsection{Avoiding Operations on Zeros}
+
+\section{Code Specialization}
+\label{sec:code-spec}
+
+\subsection{Standard Compiler Transformations}
+
+\subsection{Precomputation of Invariant Terms}
+
+\subsection{Exposing Linear Alegbra Operations}
+
+\subsection{Model-driven Autotuning}
 
@@
 %\end{algorithm}
-
-\section{Code Specialization}
 
@@
 
-\section{Related Work}
-\label{sec:related-work}
+\subsection{Results for Forms of Increasing Complexity}
 
@@
 \bibliography{biblio}
-                             % Sample .bib file with references that match those in
-                             % the 'Specifications Document (V1.5)' as well containing
-                             % 'legacy' bibs and bibs with 'alternate codings'.
-                             % Gerry Murray - March 2012
 

### 2014-07-29 TOMS_Paper/assembly.tex
Went on writing the ER stuff and ideas

@@
 
-\subsection{Summary of Low-level Optimizations and Results for Quadrature Representation}
-To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, we summarize in this section the results of our previous work on automated code transformation for quadrature representation.
+\subsection{Summary of Low-level Optimizations for Quadrature Representation}
+\label{sec:summary-opts}
+To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.
+
+...TODO...
+
+The work in~\cite{Luporini} resulted in the development of COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for the optimization of local assembly kernels relying on quadrature representation.
+
+...TODO...
+
+Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.
 
@@
 
+\section{A Compiler for Optimizing Quadrature-based Integration}
+In order to generate high performance code, mathematical expressions evaluating the element tensor must be optimized with regards to several interrelated aspects: 1) minimization of floating point operations, 2) instruction-level parallelism, and 3) data locality.
+In this paper, we tackle these three points building on our previous work (\cite{Luporini}).
+
+...We propose a novel structure of how we think a platform-independent, domain-specific optimizing compiler should look like....
+
+...The Expression Re-writer is a software module implemented in COFFEE dealing with the first point...
+
+...The Code Specializer (henceforth CS) is...
+A key point is that the ER has to perform transformations that do not break the code specializer. Having indirections is really dangerous...
+
 \section{Expression Re-writing}
 \label{sec:expr-rewriting}
+As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of a mathematical expression.
+The Expression Re-writer (henceforth ER) that we have designed and implemented in COFFEE enhances this technique making two steps forward.
 
+Firstly, exploiting arithmetic operations properties like associativity, distributivity, and commutativity, it manipulates the original expression to expose more opportunities to the code hoister.
+There are many ways an expression can be re-written into, and the search space can become considerably big.
+Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression.
+In Section~\ref{sec:rewriting-rules}, we formalize the set of re-writing rules that COFFEE follows to transform an expression.
+
+Secondly, the ER re-structures the loop nest so as to eliminate arithmetic operations over array columns that are statically known to be zero-valued.
+Zero columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements.
+A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, by using indirection arrays in the generated code, it breaks many of the optimizations that can be applied at the Code Specializer level, including SIMD vectorization.
+In Section~\ref{sec:zeros}, we show a novel approach to avoiding computation on zeros based on symbolic execution.
+
+%In COFFEE, the application of the re-writing rules preceds the elimination of operations on zeros.
+
+%It is worth noting how writing these transformations ``by hand'' would be practically unfeasible, especially when the expressions are particularly complex.
+%
+%
+%In this section, we describe our efforts in producing a more effective tool capable of explore how to leverage automated code generation to expose more opportunities to the code hoister. 
+%
+%Automated code generation, however, Mathematical expressions can be transformed to decrease the number of arithmetic operations required to evaluate the element tensor. A first 
+%
+%Minimizing the number of floating point operations for evaluating the element tensor of a generic form
+%
+
+\subsection{Re-writing Rules}
+\label{sec:rewriting-rules}
 \begin{figure}
@@
 
-\subsection{Re-writing Rules}
+Consider the simplified yet representative example of the element matrix computation in Figure~\ref{original-code}.
+...TODO: describe.
+We will use this example throughout the rest of the paper for illustrative purpose.
 
-\subsection{Avoiding Operations on Zeros}
+A first glimpse of the code easily suggests that the sub-expression ... . A closer look also highlights that ... .
+Also, ...can precompute once.
+And doing that in a loop, at the expense of extra memory, to trigger autovect.
+These observations lie at the heart of the generalized loop-invariant code motion technique presented in~\cite{Luporini}.
+
+\subsection{Avoiding Iteration on Zero-blocks with Symbolic Execution}
+\label{sec:zeros}
 

### 2014-07-30 TOMS_Paper/assembly.tex
Elaborate the Expression Rewriter

@@
 \usepackage{caption}
-\usepackage{subfig}
+\usepackage{subfigure}
+\usepackage{setspace}
 \usepackage{tabulary}
 \usepackage{lineno}
+\usepackage{xfrac}
 
@@
 % Page heads
-\markboth{F. Luporini et al.}{Optimizing Finite Element Integration through Automated Expression Re-writing and Code Specialization}
+\markboth{F. Luporini et al.}{Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization}
 
 % Title portion
-\title{Optimizing Finite Element Integration through Automated Expression Re-writing and Code Specialization}
+\title{Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization}
 \author{Fabio Luporini
@@
 \acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2014. 
-  Optimizing Finite Element Integration through Automated Expression Re-writing and Code Specialization.}
+  Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization.}
 
@@
 
-...The Expression Re-writer is a software module implemented in COFFEE dealing with the first point...
+...The Expression Rewriter is a software module implemented in COFFEE dealing with the first point...
 
@@
 
-\section{Expression Re-writing}
+\section{Expression Rewriting}
 \label{sec:expr-rewriting}
 As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of a mathematical expression.
-The Expression Re-writer (henceforth ER) that we have designed and implemented in COFFEE enhances this technique making two steps forward.
+The Expression Rewriter (henceforth ER) that we have designed and implemented in COFFEE enhances this technique by making two steps forward, which allow more redundant computation to be avoided.
 
 Firstly, exploiting arithmetic operations properties like associativity, distributivity, and commutativity, it manipulates the original expression to expose more opportunities to the code hoister.
-There are many ways an expression can be re-written into, and the search space can become considerably big.
+There are many possibilities of rewriting an expression, and the search space can quickly become too big.
 Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression.
-In Section~\ref{sec:rewriting-rules}, we formalize the set of re-writing rules that COFFEE follows to transform an expression.
+In Section~\ref{sec:rewriting-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.
 
@@
 Zero columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements.
-A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, by using indirection arrays in the generated code, it breaks many of the optimizations that can be applied at the Code Specializer level, including SIMD vectorization.
+A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with it is that by using indirection arrays in the generated code, it breaks many of the optimizations that can be applied at the Code Specializer level, including SIMD vectorization.
 In Section~\ref{sec:zeros}, we show a novel approach to avoiding computation on zeros based on symbolic execution.
@@
 
-\subsection{Re-writing Rules}
+\subsection{Objectives of the Expression Rewriter}
 \label{sec:rewriting-rules}
+
 \begin{figure}
@@
 \lstinputlisting{listings/simple.code}
-\label{original-code}\caption{Original code}
+\label{fig:original-code}\caption{Original code}
 \end{figure}
 
-Consider the simplified yet representative example of the element matrix computation in Figure~\ref{original-code}.
-...TODO: describe.
-We will use this example throughout the rest of the paper for illustrative purpose.
+Consider the simplified example of the element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a real Burgers problem.
+In practice, depending on the problem, the expression tree could be much more complex, with multiple levels of nesting.
+The example is however representative for a large class of problems, so we will use it throughout the rest of the paper for illustrative purpose.
 
-A first glimpse of the code easily suggests that the sub-expression ... . A closer look also highlights that ... .
-Also, ...can precompute once.
-And doing that in a loop, at the expense of extra memory, to trigger autovect.
-These observations lie at the heart of the generalized loop-invariant code motion technique presented in~\cite{Luporini}.
+A first glimpse of the code suggests that the sub-expression \texttt{F[i][j]*d + G[i][j]*e} is invariant with respect to the innermost loop, so it should be hoisted at the level of the outer loop \texttt{j}.
+This is a standard compiler transformation, which is supported by any available compilers.
+With a closer look we notice that the sub-expression \texttt{a*C[i][k] + b*E[i][k]} is also invariant, although, in this case, with respect to the outer loop \texttt{j}.
+In~\cite{Luporini}, we have showed that a \textit{generalized} loop-invariant code motion transformation - that is, given a non-trivial expression, the capability of analyzing all of its sub-expressions with respect to the enclosing loops to determine what code is hoistable- is not supported by available compilers.
+Moreover, the lack of cost models to ascertain both the optimal place where to hoist an expression and whether or not vectorizing it at the price of extra temporary memory is a fundamental limiting factor.
+We have addressed these problems by implementing a generalized loop-invariant code motion transformation in COFFEE.
+
+We now consider the case of forms that ``hide'' further opportunities for code hoisting.
+Such forms are by no means exotic: for example, the pattern described next is commonly found in elastic problems.
+By examining again the example in Figure~\ref{fig:original-code}, we notice that the basis function array \texttt{B}, iterating along the iteration space \texttt{[i,j]}, appears twice in the expression.
+If we expand the products in which \texttt{B} is accessed, we can apply product commutativity and then factorize the expression as in Figure~\ref{fig:factorized-code}.
+This has two effects: firstly, it reduces the total number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{(A[i][k]/c + T2[k]*f)} invariant with respect to loop \texttt{j}.
+Therefore, code hoisting can be performed.
+
+\begin{figure}
+\scriptsize
+\lstinputlisting{listings/factorized.code}
+\label{fig:factorized-code}\caption{Factorized code}
+\end{figure}
+
+The second observation we make concerns the register pressure induced by the expression.
+Once loop-invariant terms are lifted, we can think about data locality and, in particular, register allocation.
+Assume the local assembly kernel is executed on a state-of-the-art architecture having 16 logical registers, e.g.
+an Intel Haswell.
+Each value appearing in the expression is loaded and kept in a register as long as possible.
+For instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every $i$ iteration.
+This implies that at every iteration of the \texttt{jk} loop nest, 12$\%$ of the available registers are spent just to store constant values.
+In more complicated expressions, the percentage of registers destined to store loop-invariant terms can be even higher.
+Registers are, however, a precious resource, especially when evaluating intensive expressions.
+The smaller is the number of free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e.
+it can worsen data locality), or it may prevent the effective application of standard transformations like loop unrolling.
+The ER works around this problem by suitably expanding terms and introducing, where necessary, new temporary values.
+
+%An analogous analysis applies to processors with larger numbers of registers, since using loop unroll or loop unroll-and-jam to expose more instruction-level parallelism would increase the requirements on registers.
+
+\begin{figure}
+\scriptsize
+\lstinputlisting{listings/toexpand.code}
+\label{fig:toexpand-code}\caption{Expandable code}
+\end{figure}
+
+Consider the variant of the running (transformed) example shown in Figure~\ref{fig:toexpand-code}.
+Again, this is a representative example of what happens in real finite element forms.
+We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of \texttt{T1}, resulting in the code illustrated in Figure~\ref{fig:expanded-1-code}.
+Freeing the register destined to the constant \texttt{g} is more complicated: we cannot absorb it in the pre-computation of \texttt{T1} because the same array is accessed in the evaluation of \texttt{(T1[ j]*A[i][k])}.
+The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}.
+Generalizing, this is a problem of data dependencies; in order to solve it, we employ a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}.
+The dependency graph is initially empty; every time a new temporary is created due to loop-invariant code motion or expansion of terms is performed, it is updated by suitably adding vertices and edges.
+
+\begin{figure}[t]
+\tiny
+\centering     
+\subfigure[Expanded 1 code]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
+~~
+\subfigure[Expanded 2 code]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
+\caption{Expanded code.}\label{fig:expanded-code}
+\end{figure}
+
+\subsection{Rewrite Rules}
+In general, assembly expressions produced by automated code generation can be much more complex (more terms and operations involved) and nested.
+Our goal is to establish a portable, platform- and compiler-independent, and systematic way of reducing the strength of an expression.
+The technique should be simple; definitely it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems.
+Ideally, it should be naturally extendible to problems that will be supported in next releases of state-of-the-art frameworks like Firedrake and FEniCS: for instance, explicit support for outer-product finite elements will enable generation of kernels with much deeper loop nests, and the ER should transparently be able to deal with these structures as well.
+
+To address these issues, we have based the implementation of the ER in COFFEE on a set of formal rewrite rules.
+By applying these rules, it is possible to derive how an expression will be transformed, as well as what and where (i.e.
+at which level in the loop nest) temporaries will be introduced.
+When applying a rule, the ER needs to update the state of the loop nest, to reflect, for example, the use of a new temporary and the newly created data dependencies.
+We define, therefore, the state of a loop nest $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma$ maps invariant sub-expressions to identifiers of temporary arrays.
+We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma : Inv \rightarrow S$ such that
+\begin{gather*}
+\sigma[\sfrac{v}{x}] = 
+\begin{cases}
+\sigma(x) \text{~~~~~~~~if $x \in Inv$; $v$ is ignored}\\
+v \text{~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $\sigma$(x) = v}\\
+\end{cases}
+\end{gather*}
+That is, intuitively, if the invariant expression \texttt{x} has already been hoisted, then return the temporary identifiers that hosts its value; otherwise, hoist the expression.
+There is a special case when $v = \perp$, used to delete entries in $\sigma$.
+Specifically:
+\begin{gather*}
+\sigma[\sfrac{\perp}{x}] = 
+\begin{cases}
+\sigma(x) \text{~~~~~~~~if $x \in Inv$; $\sigma = \sigma \setminus (x, \sigma(x))$}\\
+t \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $t \notin Inv$}\\
+\end{cases}
+\end{gather*}
+In other words, the previously hoisted expression \texttt{x} is removed (if any) and the temporary identifier that was hosting its value is returned.
+This is useful to express updates of invariant expressions.
+Rewrite rules for the ER are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
+Conceptually, the ER visits the expression tree from the root, which is the outermost operation, and applies the transformations dictated by the rewrite rules.
+As an example, one can try instantiating the rules in the code of Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized code in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} is obtained, respectively.
+
+\begin{figure}
+\small
+\centering
+\begin{spacing}{1.5}
+\begin{align*}
+[a_i \cdot b_j]_{(\sigma, G)} &\rightarrow [a_i \cdot b_j]_{(\sigma, G)} ~~&~~&\\
+[(a_i + b_j)\cdot \alpha]_{(\sigma, G)} &\rightarrow [(a_i \cdot \alpha + b_j \cdot \alpha)]_{(\sigma, G)} ~~&~~ &\\
+[a_i \cdot b_j + a_i \cdot c_j]_{(\sigma, G)} &\rightarrow [(a_i \cdot (b_j + c_j)]_{(\sigma, G)} ~~&~~ &\\
+[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
+[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
+~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
+[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
+~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
+\end{align*}
+\end{spacing}
+\caption{Rewrite rules.}\label{fig:rewrite-rules}
+\end{figure}
+
+% blablabla molto complesse, nestings etc expose etc allora rewrite rules
+
+% we will show the effects of this stuff over plain LICM in section
+
+% hoisting constants release registers!!!!
 
@@
 
+% Mass DGEMM utile explicit method etc
+
 \section{Conclusions}

### 2014-07-31 TOMS_Paper/assembly.tex
Start code specializer section

@@
 \begin{figure}
-\scriptsize
+\tiny
 \lstinputlisting{listings/simple.code}
-\label{fig:original-code}\caption{Original code}
+\caption{Original code}\label{fig:original-code}
 \end{figure}
@@
 \begin{figure}
-\scriptsize
+\tiny
 \lstinputlisting{listings/factorized.code}
-\label{fig:factorized-code}\caption{Factorized code}
+\caption{Factorized code}\label{fig:factorized-code}
 \end{figure}
@@
 \begin{figure}
-\scriptsize
+\tiny
 \lstinputlisting{listings/toexpand.code}
-\label{fig:toexpand-code}\caption{Expandable code}
+\caption{Expandable code}\label{fig:toexpand-code}
 \end{figure}
@@
 
+%TODO: each level in sigma is then taken and wrapped in a loop
+
 % blablabla molto complesse, nestings etc expose etc allora rewrite rules
@@
 
-\subsection{Avoiding Iteration on Zero-blocks with Symbolic Execution}
+\subsection{Avoiding Iteration on Zero-blocks by Symbolic Execution}
 \label{sec:zeros}
+The second task of the ER consists of skipping arithmetic operations over blocks of zero-valued entries in basis function and derivatives of basis functions arrays.
+Zero columns in such arrays arise, for example, when taking derivatives on a reference element and when employing mixed elements.
+In~\cite{quadrature1}, a code transformation to eliminate floating point operations on zero columns was introduced; the technique was implemented in FEniCS.
+We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}.
+We propose a strategy in which we avoid the use of indirection arrays in the generated code (e.g.
+\texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), which otherwise would break the optimizations applicable at the Code Specializer level, especially SIMD vectorization.
+
+\begin{figure}[t]
+\tiny
+\centering     
+\subfigure[Factorized, showing zeros, code]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subfigure[Factorized, skipping zeros, code]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
+\caption{Without and with skipping zeros}\label{fig:skip-code}
+\end{figure}
+
+Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the transformed Burgers excerpt in Figure~\ref{fig:factorized-code}.
+The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh.
+The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points.
+We notice the presence of 4 zero-columns.
+Any multiplications or additions along these columns could (should) be skipped in order to minimize irrelevant floating point operations.
+One solution, as anticipated, is not to generate the zero columns, i.e.
+to generate a dense 6$\times$2 array, to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g.
+$ind = \lbrace 3, 5\rbrace$) to access the right entries in the element tensor $A$.
+As explained, this prevents, among the various optimizations, effective code vectorizability, because memory loads and store would eventually reference non-contiguous locations.
+%on the other hand, reshrinki tutto su un array solo...
+
+Our approach is based on domain knowledge and symbolic execution.
+We discern the origin of zero-columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements.
+In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to preserve the correctness of the element matrix evaluation while iterating along the space of test and trial functions.
+The two key observations are that 1) the number of zero-columns caused by mixed elements is much larger then that due to derivatives, and 2) such columns are contiguous in memory.
+Based on this, we decide to skip any computation involving zero-columns induced by mixed elements, whereas we still iterate over any other zero-columns.
+Consequently, the only modifications to the code that are eventually needed are adjusting loop bounds and introducing offsets when accessing the element matrix, as shown in Figure~\ref{fig:withzeros-skipped-code}.
+In general, multiple iteration spaces over test and trial functions (i.e.
+\texttt{jk} loops), characterized by their own loop bounds, may be needed, increasing loop overhead and decreasing data locality for the element matrix.
+
+The implementation is based on symbolic execution.
+The ER expects information from the higher layer about where the zero-blocks are located, in each tabulated basis function.
+This could come in the form of code annotation if the input to COFFEE were provided as a C string (as shown in Figure~\ref{fig:withzeros-skipped-code}) or by suitably decorating basis functions nodes in the abstract syntax tree representing the local assembly routine.
+Then, each statement in the local assembly loop nest is symbolically executed.
+For example, consider the statement \texttt{T2[r] = ((d*D[i][k])+(e*E[i][k]));} in Figure~\ref{fig:withzeros-skipped-code}.
+Array \texttt{D} has non-zeros in positions $NZ_D = [3,5]$; we also assume \texttt{E} has non-zeros in positions $NZ_E = [0,2]$.
+Multiplications by scalar do not affect the propagation of zero columns.
+On the other hand, when summing the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we record that the target identifier \texttt{T2} will have non-zero columns in positions $NZ_D \cup NZ_E = [0-5]$.
+Exploiting the $NZ$ information associated with each identifier, the assembly expression is split into several sets of sub-expressions, each set characterized by accesses to the same range of non-zero columns and, therefore, iterating over a suitably-sized iteration space.
+In our example, there is just one set, which contains two sub-expressions, enclosed in a doubly nested loop of size 3$\times$3.
 
@@
 \label{sec:code-spec}
+The Code Specializer is the second software component of COFFEE.
+It provides a range of code transformations tailored for optimizing instruction-level parallelism and register locality.
+As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying.
+In this paper, we enrich this set of transformations as detailed in the following sections.
 
-\subsection{Standard Compiler Transformations}
+\subsection{Vector-expansion of Invariant Terms}
+The Expression Rewriter's loop-invariant code motion hoists invariant sub-expressions in the body of the outermost loop along which they iterate.
+If such a sub-expression computes a value depending on the outermost loop only, then the only way of vectorizing it - if we exclude superword level parallelism~\cite{SLP}, which is in general not applicable to our kernels - is by vector-expansion of terms.
 
-\subsection{Precomputation of Invariant Terms}
+\begin{figure}[t]
+\tiny
+\centering     
+\subfigure[To vector expand code]{\label{fig:tovectexpand-code}\lstinputlisting{listings/tovectexpand.code}}
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subfigure[Vector-expanded code]{\label{fig:vectexpanded-code}\lstinputlisting{listings/vectexpanded.code}}
+\caption{Before and after vector expansion}\label{fig:skip-code}
+\end{figure}
+
+An example is given in Figure~\ref{fig:tovectexpand-code}: \texttt{F1, F2, F3, T1} are all scalar quantities, and their computation will not be auto-vectorized (as explained in Section~\ref{sec:perf-results}, we experimented with the Intel and GNU compilers).
+The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I \cdot T^2)$, is too superficial, in general.
+There are notable cases, for instance problems based on hyperlasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.
+
+We have implemented a procedure that vector-expands scalar expressions, while lifting them outside of the loop nest as shown in Figure~\ref{fig:vectexpanded-code}.
+The dependency graph is queried to assess the legality of the transformation.
+At the price of extra memory, vector-expanded terms can now be vectorized by the compiler.
+
+% Expose vectorizability of the outermost loop 
+% Render the loop nest perfect
 
@@
 
+% can i just add a rewrite rule to expose these stuff?
+
 \subsection{Model-driven Autotuning}
-
-%\begin{algorithm}[t]
-%\SetAlgorithmName{LISTING}{}
-%\footnotesize
-%\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
-%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
-%~~// K, det = Compute Jacobian (coords) \\
-%~~\\
-%~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
-%~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
-%~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
-%~~\\
-%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
-%~~~~double LI$\_$0[4] ALIGN;\\
-%~~~~double LI$\_$1[4] ALIGN;\\
-%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
-%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
-%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
-%~~~~$\rbrace$\\
-%~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
-%~~~~~~\KwSty{$\#$pragma vector aligned}\\
-%~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
-%~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
-%~~$\rbrace$\\
-%$\rbrace$
-%\caption{Local assembly code for the Helmholtz problem in
-%  Listing~\ref{code:helmholtz} after application of padding, data
-%  alignment, and \emph{licm}, for an AVX architecture. In this
-%  example, sub-expressions invariant to \texttt{j} are identical to
-%  those invariant to \texttt{k}, so they can be precomputed once in
-%  the $r$ loop.}
-%\label{code:helmholtz-licm}
-%\end{algorithm}
-
-%\begin{figure}[b]
-%\begin{center}
-%\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
-%\caption{High-level view of Firedrake. COFFEE is at the core,
-%  receiving ASTs from a modified version of the FEniCS Form compiler
-%  and producing optimized C code kernels.}
-%\label{fig:coffee-pipeline}
-%\end{center}
-%\end{figure}
+%\subsection{Standard Compiler Transformations}
 

### 2014-08-01 TOMS_Paper/assembly.tex
Went on writing stuff for Code Specializer

@@
 
+% ASSEMBLY EXPRESSION == THE EXPRE// COMPUTATION OF ELEMENT MATRIX
+
 \subsection{Code Generation for Quadrature Representation}
@@
 ...We propose a novel structure of how we think a platform-independent, domain-specific optimizing compiler should look like....
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.6]{pictures/coffee-scheme}
+\caption{Outline of COFFEE}\label{fig:coffee-outline}
+\end{center}
+\end{figure}
+
+The high-level structure of COFFEE is shown in Figure~\ref{fig:coffee-outline}.
 
@@
 
+...TODO...: dire come si riaggiustano gli indici quando si vuole usare AVX ma la parte dense non e' multiplo del vector length - se non e' paddata la tail, allora si decrementa lo starting index di un pochino, tanto sono zeri...
+
 \section{Code Specialization}
@@
 \subsection{Vector-expansion of Invariant Terms}
+\label{sec:vect-exp}
 The Expression Rewriter's loop-invariant code motion hoists invariant sub-expressions in the body of the outermost loop along which they iterate.
-If such a sub-expression computes a value depending on the outermost loop only, then the only way of vectorizing it - if we exclude superword level parallelism~\cite{SLP}, which is in general not applicable to our kernels - is by vector-expansion of terms.
+If such a sub-expression computes a value depending on the outermost loop only, then the only way of vectorizing it -- if we exclude superword level parallelism~\cite{SLP}, which is in general not applicable to our kernels -- is by vector-expansion of terms.
 
@@
 \subsection{Exposing Linear Alegbra Operations}
+In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrices multiply routines (henceforte DGEMM), for instance MKL or ATLAS BLAS.
+We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM calls.
+The study was conducted only in the case of a specific Helmholtz equation.
+There are cases, however, in which tabulated basis functions sizes grow up to a point for which certain BLAS implementations allow to achieve a better performance.
+For example, this may be the case of relatively-high polynomial orders or when the form is characterized by the presence of a number of pre-multiplying functions.
+To explore these extreme yet meaningful scenarios, we have developed an algorithm to reduce any generic element matrix evaluation to a sequence of DGEMM calls.
 
-% can i just add a rewrite rule to expose these stuff?
+The algorithm is shown in Figure~\ref{fig:algo-dgemm}.
+By using the rewrite rules in Figure~\ref{fig:rewrite-rules}, it can be proved that the Expression Rewriter can always reduce an assembly expression to a summation, over each quadrature point, of outer products along the test and trial functions.
+Each outer product is then isolated, i.e.
+the assembly expression is split into chunks, each chunk representing an outer product.
+Terms in the bodies of the various enclosing loops (e.g.
+coefficients evaluation at quadrature point, temporaries introduced by the Expression Rewriter) are vector-expanded and hoisted outside of the loop nest, as explained in Section~\ref{sec:vect-exp}.
+This implies that the loop nest is now perfect; that is, there is no intervening code among the various loops.
+The element matrix evaluation became a sequence of perfect loop nests, each loop nest computing a dense matrix-matrix multiply between temporaries hoisted by the Expression Rewriter or tabulated basis functions.
+Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g.
+two dimensional arrays are flatten as one dimensional arrays).
+The translation of each loop nest into a DGEMM call is the last, straightforward step.
 
-\subsection{Model-driven Autotuning}
+% 
+\subsection{Model-driven Dynamic Autotuning}
+We have demonstrated in~\cite{Luporini} that there is no optimial set of code transformations.
+What sequence of transformations applying to a given problem depends on a broad range of factors, including mathematical structure of the input form, polynomial order of employed function spaces, presence of pre-multiplying functions, and, of course, the characteristics of the underlying architecture.
+COFFEE allows the user to abstract from finding the optimal combination of transformations for a given problem by resorting to its model-driven, dynamic autotuner.
+This is based on the idea that spending order of seconds to determine the fastest kernel implementation in a set of possible variants is a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain even up to trillions of elements~\cite{Rossinelli2013}.
+
+In~\cite{Luporini}, we proposed a simple cost model that COFFEE could use to choose, given a problem, the transformations that were likely to maximize the performance.
+With this paper, however, we have added a significant number of options to the set of possible optimizations, which makes the selection problem far more challenging.
+The sole Expression Rewriter, for instance, could apply rewrite rules up to a number of pre-established depths, leading to the generation of many possible code variants.
+
+We believe autotuning is an effective approach to tackling this problem.
+For each problem, not only does it allow to determine the best combination of transformations out of the set presented so far, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying multiple unroll factors and loop permutations.
+By leveraging the cost model defined in our previous study, domain-awarness, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over purely cost-model-driven implementations.
+
+%1) dual approach to cost-model - more reliable? Easier...when zero-avoidance and low order, what is it best to do?
+%2) self-building
+%3) deeper exploration, to explore aggressive standard low-level transformations
+%4) dynamic and heuristic-driven: 
+%     - depending on kernel size, more or less unroll factors are used
+%     - domain knowledge: no interchange test-trial functions loops
+%     - no unroll innermost loop
 %\subsection{Standard Compiler Transformations}

### 2014-08-04 TOMS_Paper/assembly.tex
Write the autotuner section

@@
 Rapid implementation of high performance, robust, and portable code evaluating element matrices using quadrature can be achieved through automated code generation.
-This has been succesfully proved in the context of the popular FEniCS project~\cite{Fenics}.
+This has been successfully proved in the context of the popular FEniCS project~\cite{Fenics}.
 The FEniCS Form Compiler (FFC) accepts as input the variational form of a partial differential equation and generates C++ code implementing local assembly routines.
@@
 
-\begin{figure}
+\begin{figure}[t]
 \tiny
-\lstinputlisting{listings/simple.code}
-\caption{Original code}\label{fig:original-code}
+\centering     
+\subfigure[Original code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
+~~~~~~~~~~~~~
+\subfigure[Factorized code]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
+\caption{Original and factorized code.}
 \end{figure}
@@
 Therefore, code hoisting can be performed.
-
-\begin{figure}
-\tiny
-\lstinputlisting{listings/factorized.code}
-\caption{Factorized code}\label{fig:factorized-code}
-\end{figure}
 
@@
 
-%TODO: each level in sigma is then taken and wrapped in a loop
+... TODO... : each level in sigma is then taken and wrapped in a loop
 
+...TODO...: to what extent applying rewrite rules, is problem specific...
 % blablabla molto complesse, nestings etc expose etc allora rewrite rules
@@
 The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I \cdot T^2)$, is too superficial, in general.
-There are notable cases, for instance problems based on hyperlasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.
+There are notable cases, for instance problems based on hyperelasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.
 
@@
 \subsection{Exposing Linear Alegbra Operations}
-In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrices multiply routines (henceforte DGEMM), for instance MKL or ATLAS BLAS.
+In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrices multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS.
 We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM calls.
@@
 \subsection{Model-driven Dynamic Autotuning}
-We have demonstrated in~\cite{Luporini} that there is no optimial set of code transformations.
-What sequence of transformations applying to a given problem depends on a broad range of factors, including mathematical structure of the input form, polynomial order of employed function spaces, presence of pre-multiplying functions, and, of course, the characteristics of the underlying architecture.
-COFFEE allows the user to abstract from finding the optimal combination of transformations for a given problem by resorting to its model-driven, dynamic autotuner.
-This is based on the idea that spending order of seconds to determine the fastest kernel implementation in a set of possible variants is a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain even up to trillions of elements~\cite{Rossinelli2013}.
+We have demonstrated in~\cite{Luporini} that an optimal set of code transformations does not exist.
+What sequence of transformations applying to a problem to optimize its performance depends on a broad range of factors, including mathematical structure of the input form, polynomial order of employed function spaces, presence of pre-multiplying functions, and, of course, the characteristics of the underlying architecture.
 
 In~\cite{Luporini}, we proposed a simple cost model that COFFEE could use to choose, given a problem, the transformations that were likely to maximize the performance.
-With this paper, however, we have added a significant number of options to the set of possible optimizations, which makes the selection problem far more challenging.
-The sole Expression Rewriter, for instance, could apply rewrite rules up to a number of pre-established depths, leading to the generation of many possible code variants.
+By having added a significant number of options to the set of possible optimizations, the selection problem is now far more challenging.
+The sole Expression Rewriter, for instance, could apply rewrite rules up to a set of pre-established depths, leading to the generation of many possible code variants.
 
-We believe autotuning is an effective approach to tackling this problem.
+We tackle this problem by compiler autotuning.
 For each problem, not only does it allow to determine the best combination of transformations out of the set presented so far, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying multiple unroll factors and loop permutations.
-By leveraging the cost model defined in our previous study, domain-awarness, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over purely cost-model-driven implementations.
+By leveraging the cost model defined in our previous study, domain-awareness, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementation.
+In particular, our autotuner usually requires order of seconds to determine the fastest kernel implementation, a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain up to trillions of elements (e.g.
+~\cite{Rossinelli2013}).
 
-%1) dual approach to cost-model - more reliable? Easier...when zero-avoidance and low order, what is it best to do?
-%2) self-building
-%3) deeper exploration, to explore aggressive standard low-level transformations
-%4) dynamic and heuristic-driven: 
-%     - depending on kernel size, more or less unroll factors are used
-%     - domain knowledge: no interchange test-trial functions loops
-%     - no unroll innermost loop
-%\subsection{Standard Compiler Transformations}
+The structure of the autotuner is outlined in Figure~\ref{fig:autotuner}.
+COFFEE analyzes the input problem and decides what variants it is worth testing, as described later.
+It then provides the autotuner with all of the possible variants, in the form of abstract syntax trees.
+The autotuner is a template-based code generator.
+By inspecting an abstract syntax tree, it determines how to generate ``wrapping'' code that, repeatedly in a \emph{while} loop executed a pre-established amount of time (order of milliseconds), 1) initializes kernel's input variables with fictitious values and 2) calls the kernel.
+At the exit of the \emph{while} loop, the times the kernel was invoked is recorded.
+Eventually, the variant executing the largest number of iterations is designated as the fastest implementation.
+Suitable compiler directives are used to prevent inlining of all function calls: this avoids the situation in which some variants are inlined and some are not, which would fake the autotuner's output.
+
+The autotuning process is dynamic: depending on the complexity of the input problem, more or less variants are tried.
+General heuristics, which can be considered a revisited version of those presented in~\cite{nek5000}, are applied
+\begin{itemize}
+\item Loop permutations that are likely to worsen the performance are excluded from the search space.
+According to the cost model, and for the same reasons explained in~\cite{Luporini}, we enable only variants in which the loop over quadrature points is either the outermost or the innermost.
+This is due to the fact that versions of the code in which such loop lies between the test and trial functions loops are typically lower performing.
+\item The unroll factors must divide the loop bounds evenly to avoid the introduction of reminder (scalar) loops.
+\item The innermost loop is never explicitly unrolled. This is because we expect auto-vectorization along this loop, so memory accesses should be kept unit-stride. 
+\end{itemize}
+Also, the following heuristics, which capture properties of our computational domain, are used
+\begin{itemize}
+\item When both test and trial functions derive from the same function space, the lengths of their corresponding loops are identical.
+In this case, since for the employed storage layout the memory accesses are symmetrical along these two loops, their interchange is pruned from the search space.
+\item In general, if relatively high polynomial orders are employed, the loop nest is larger.
+In these cases, in order to avoid testing too many unroll factors, we impose a bound $X$ on the loop nest's overall unroll factor, which we found empirically.
+For example, the sum of the unroll factors of all outer loops cannot exceed $X$.
+\item On the other hand, if the polynomial order is low, i.e. for smaller loop nests, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
+\item We select two levels of expression rewriting.
+In the ``base'' level, generalized loop-invariant code motion only is applied.
+This means that only a subset of the rewrite rules exposed in~\ref{fig:rewrite-rules} are applicable.
+In the ``aggressive'' level, all of the rewrite rules are applied.
+\item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only only three split factors, namely 1, 2, 4.
+Also, if the input problem uses mixed function spaces, the iteration space is already split by the Expression Rewriter to avoid computation over zero-columns; in these cases, we do not further apply expression splitting.
+\item Based on the cost model, the padding and data alignment optimization is always applied. 
+\end{itemize}
+
+%COFFEE allows the user to abstract from finding the optimal optimization strategy for a given problem by resorting to its model-driven, dynamic autotuner.
 
@@
 
+In Figure~\ref{fig:helmholtz}, ...
+
+%\begin{figure}[t]
+%\tiny
+%\centering     
+%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p1_formhelmholtz_speedup}}
+%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p2_formhelmholtz_speedup}}
+%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p3_formhelmholtz_speedup}}
+%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p4_formhelmholtz_speedup}}
+%\caption{Helmholtz results.}\label{fig:helmholtz}
+%\end{figure}
+
 \subsection{Results for Forms of Increasing Complexity}

### 2014-08-04 TOMS_Paper/assembly.tex
Add rough preliminaries

@@
 % Page heads
-\markboth{F. Luporini et al.}{Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization}
+\markboth{F. Luporini et al.}{Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization}
 
 % Title portion
-\title{Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization}
+\title{Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization}
 \author{Fabio Luporini
@@
 \affil{Imperial College London}}
-% NOTE! Affiliations placed here should be for the institution where the
-%       BULK of the research was done. If the author has gone to a new
-%       institution, before publication, the (above) affiliation should NOT be changed.
-%       The authors 'current' address may be given in the "Author's addresses:" block (below).
-%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
-%       currently affiliated with NASA.
 
@@
 \acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2014. 
-  Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization.}
+  Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization.}
 
@@
 
+%is the computation of contributions of a specific cell in the discretized domain to the linear system which yields the PDE solution.
+The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector [Olgaard and Wells 2010; AMCG 2010], whose sizes depend on the order of the method.
+This operation is applied to all cells in the discretized domain.
+In this work we focus on local matrices, or “element matrices”, which are more costly to compute than element vectors.
+
 \subsection{Quadrature for Finite Element Local Assembly}
-%Here we discuss about the computational characteristics of
-%computational science kernels. Briefly cite op2 kernels. Emphasis on
-%Finite Element Assembly. Generalization and formalization of a Finite
-%Element Assembly kernel using quadrature representation.
+We summarize the basic concepts sustaining the finite element method following the notation adopted in~\cite{quadrature1} and~\cite{Francis}.
+We consider the weak formulation of a linear variational problem
+\begin{equation}
+\begin{split}
+Find\ u\ \in U\ such\ that \\
+a(u, v) = L(v), \forall v \in V
+\end{split}
+\end{equation}
+where $a$ and $L$ are called bilinear and linear form, respectively.
+The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces.
+For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$.
+The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$.
+From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
+\begin{equation}
+A\textbf{u} = b
+\end{equation}
+in which $A$ and $b$ discretize $a$ and $L$ respectively:
+\begin{equation}
+\centering
+\begin{split}
+A_{ij} = a(\phi_i(x), \phi_j(x)) \\
+b_i = L(\phi_i(x))
+\end{split}
+\end{equation}
+The matrix $A$ and the vector $b$ are computed in the so called assembly phase.
+Then, in a subsequent phase, the linear system is solved, usually by means of an iterative method, and $\textbf{u}$ is eventually evaluated.
 
-% ASSEMBLY EXPRESSION == THE EXPRE// COMPUTATION OF ELEMENT MATRIX
+We focus on the assembly phase, which is often characterized as a two-step procedure: \textit{local} and \textit{global} assembly.
+Local assembly is the subject of the paper: this is about computing the contributions that an element in the discretized domain provide to the approximated solution of the equation.
+Global assembly, on the other hand, is the process of suitably ``inserting'' such contributions in $A$ and $b$.
+
+Without loss of generality, we illustrate local assembly in a real example, the evaluation of the local assembly (or, equivalently, element) matrix for a Laplacian operator.
+Consider the weighted Laplace equation
+\begin{equation}
+- \nabla \cdot (w \nabla u) = 0
+\end{equation}
+in which $u$ is unknown, while $w$ is prescribed. The bilinear form associated with the weak variational form of the equation is:
+\begin{equation}
+a(v, u) = \int_\Omega w \nabla v \cdot \nabla u\ \mathrm{d}x
+\end{equation}
+The domain $\Omega$ of the equation is partitioned into a set of cells (elements) $T$ such that $\bigcup T = \Omega$ and $\bigcap T = \emptyset$.
+By defining $\lbrace \phi_i^K \rbrace$ as the set of local basis functions spanning $U$ on the element $K$, we can express the local element matrix as
+\begin{equation}
+\label{stiffness}
+A_{ij}^K = \int_K w \nabla \phi_i^K \cdot \nabla \phi_j^K\ \mathrm{d}x
+\end{equation}
+The local element vector $L$ can be determined in an analogous way. 
+%From the computational perspective, its evaluation is however less expensive than that of $A$.
+
+Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$.
+For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced.
+This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated.
+The numerical integration routine based on quadrature representation over an element $K$ can be expressed as follows
+\begin{equation}
+\label{quadrature}
+\scriptsize
+A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
+\end{equation}
+where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.
+
+...TODO...: say that "assembly expression" =THE EXPRE// COMPUTATION OF ELEMENT MATRIX
 
@@
 
-%Experiments were run on a single core of two Intel architectures, a
-%Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB
-%L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode,
-%32KB L1 cache and 512KB L2 cache). We have chosen these two
-%architectures because of the differences in the number of logical
-%registers and SIMD lanes, which can impact the effectiveness of the
-%optimization strategy. The \texttt{icc 13.1} compiler was used. On the
-%Sandy Bridge, the compilation flags used were \texttt{-O2} and
-%\texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization
-%level \texttt{-O3} was used. Other optimization levels performed, in
-%general, slightly worse.
+Experiments were run on a single core of an Intel architecture, a Sandy Bridge I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache.
+The Intel \texttt{icc 14.2} compiler was used.
+The compilation flags used were \texttt{-O3, -xHost, -xAVX, -ip}.
 
-%\newcommand{\licmapresultsnorms}{
-%\begin{tabulary}{1.0\textwidth}{cccccc|cccc}
-%\cline{3-10}
-%& & \multicolumn{4}{c}{\texttt{Sandy Bridge}} & \multicolumn{4}{c}{\texttt{Xeon Phi}} \\
-%\cline{1-10}
-%\texttt{problem} & \texttt{shape} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
-%\texttt{Helmholtz} & \texttt{triangle} & 1.32 & 1.88 & 2.87 & 4.13 & 1,50 & 2,41 & 1,30 & 1,96 \\
-%\texttt{Helmholtz} & \texttt{tetrahedron} & 1.35 & 3.32 & 2.66 & 3.27 & 1,41 & 1,50 & 2,79 & 2,81 \\
-%\texttt{Helmholtz} & \texttt{prism} & 2.63 & 2.74 & 2.43 & 2.75 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
-%%\cline{1-18}
-%\texttt{Diffusion} & \texttt{triangle} & 1.38 & 1.99 & 3.07 & 4.28 & 1,08 & 1,88 & 1,20 & 1,97\\
-%\texttt{Diffusion} & \texttt{tetrahedron} & 1.41 & 3.70 & 3.18 & 3.82 & 1,05 & 1,51 & 2,76 & 3,00\\
-%\texttt{Diffusion} & \texttt{prism} & 2.55 & 3.13 & 2.73 & 2.69 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
-%%\cline{1-18}
-%\texttt{Burgers} & \texttt{triangle} & 1.56 & 2.28 & 2.61 & 2.77 & 2,84 & 2,26 & 3,96 & 4,27 \\
-%\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.10 & 1.60 & 1.78 & 1,48 & 3,83 & 1,55 & 1,29 \\
-%\texttt{Burgers} & \texttt{prism} & 2.19 & 2.32 & 1.64 & 1.42 & 2,18 & 2,82 & 1,24 & 1,25 \\
-%\cline{1-10}
-%\end{tabulary}
-%}
-%
-%\begin{table*}[t]
-%\tbl{Performance improvement due to generalized loop-invariant code
-%  motion, data alignment, and padding, for different element shapes
-%  (triangle, tetrahedron, prism) and polynomial orders ($p \in [1,
-%    4]$), over the original non-optimized code, for the Helmholtz,
-%  Diffusion and Burgers problems.}{ \scriptsize \licmapresultsnorms }
-%\label{table:perf-results-licmap}
-%\end{table*}
+...TODO... Problems chosen...
+
+...TODO... We don't compare to tensor contraction by transitivity on olegaards work
+
+\subsection{Results for Forms of Increasing Complexity}
 
@@
 
-\subsection{Results for Forms of Increasing Complexity}
-
 % Mass DGEMM utile explicit method etc

### 2014-08-04 TOMS_Paper/assembly.tex
Add introduction

@@
 
+% Quadrature. Tensor. Quadrature <, >, = Tensor
+%code evaluating element matrices using quadrature
+
+The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
+This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become incredibly popular over the last years.
+In these frameworks, the weak variational form of a given problem is expressed at high-level by means of a domain-specific language.
+Such a mathematical specification is suitably manipulated and then passed as input to a form compiler, whose goal is to generate a representation of local assembly operations.
+These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution.
+Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more expensive, covering a significant fraction of overall computation run-time.
+
+Achieving high performance implementations is, however, non-trivial.
+The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process.
+Also, traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions.
+This has led to development of a number of higher-level approaches to optimize local assembly kernels.
+In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''.
+\cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of the problem.
+In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
+The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu}.
+
+Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework.
+While clearly separating the mathematical domain, which remains captured in the higher-level form compiler, from the optimization process, COFFEE also aims to be platform-agnostic.
+The code transformations occur on an intermediate representation of the assembly code, which is ultimately translated into platform-specific code.
+Domain knowledge is exploited in two ways: for simplifying the implementation of a broad range of code transformations, and, obviously, to make them extremely effective.
+Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input.
+For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process.
+Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only.
+We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.
+
+In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines.
+We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization.
+The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become sub-steps of code specialization; on the other hand, generalized loop-invariant code motion is a step of the expression rewriting process.
+More importantly, we complement and generalize our previous work with the following contributions.
+
+Expression rewriting is based on a formal set of rewrite rules.
+Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of arithmetic operators to expose ``hidden'' loop-invariant sub-expressions.
+Secondly, we show how to make use of domain knowledge to avoid computation over zero-valued regions in vector-valued basis functions arrays, while preserving code vectorizability.
+These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable by using FEniCS' built-in optimizations, presented in~\cite{quadrature1}.
+
+At code specialization time, transformations are applied to maximize the exploitation of the underlying platform's resources, e.g.
+SIMD lanes.
+On top of the work in~\cite{Luporini}, we provide a number of contributions.
+Firstly, we show the benefit of vector-expansion to achieve SIMD vectorization of otherwise scalar code.
+This is particularly useful in complex forms, like those based on hyperelasticity.
+Secondly, we answer an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix multiplies.
+BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods.
+However, we will show that in corner, yet important cases, especially in forms characterized by pre-multiplying functions and relatively high-order function spaces, a BLAS-based execution strategy can be a successful.
+Finally, we introduce a model-driven, dynamic autotuner that automatically and transparently compose the set of code transformations that are likely to maximize the performance of a given problem.
+The main challenge with the autotuner is to maintain, for any possible problem, the search space reasonably small, although comprising the most effective code variants, so that the overhead, which impacts the run-time, is negligible.
+
+Expression rewriting and code specialization have been implemented in COFFEE and are fully operating.
+Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation in a number of forms of increasing complexity, including problems based on hyperelasticity operators.
+We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions.
+To clearly distinguish the improvement achieved by this work, we will compare four sets of code variants, for each problem instance: 1) unoptimized code, i.e.
+a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e.
+the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper.
+Notable performance improvements of 4) over 1), 2) and 3) are reported and detailed.
+
+%The contributions of this paper can be summarized as follows:
+%\begin{enumerate}
+%\item 
+%\item extensive and unprecedented performance evaluation in a wide range of forms of increasing complexity
+%\end{enumerate}
+
+%Extensive performance evaluation...
+
 %This paper can be thought of as a continuation of the work presented in~\cite{Luporini}: in particular, we answer the remaining open questions, generalize and formalize an approach to minimize redundant computation using a similar philosophy, we show ~\cite{quadrature1}.
@@
 
-\subsection{Code Generation for Quadrature Representation}
-Rapid implementation of high performance, robust, and portable code evaluating element matrices using quadrature can be achieved through automated code generation.
-This has been successfully proved in the context of the popular FEniCS project~\cite{Fenics}.
-The FEniCS Form Compiler (FFC) accepts as input the variational form of a partial differential equation and generates C++ code implementing local assembly routines.
-The variational form is expressed at high-level by means of the domain-specific Unified Form Language (UFL).
-Local assembly code must be high performance: as the complexity of a form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen functions, the resulting kernels evaluating element matrices become more computationally expensive, which impacts significantly the run-time of the overall computation.
-
-Achieving high performance is non-trivial due to the complexity of the mathematical expressions involved in the numerical integration and because of the small sizes of loops and accessed arrays.
-In~\cite{quadrature1},~\cite{Kirby-FEM-opt} and~\cite{Francis}, it is shown how automated code generation allows the introduction of powerful optimizations, which a user cannot be expected to write ``by hand'', as well as the exploration of non-standard integration techniques, based, for instance, on symbolic manipulation.
-In~\cite{Luporini} we made one step forward by showing that different problems require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations, which are not supported by available compilers, are essential to maximize instruction-level parallelism and register locality.
-
-\subsection{Summary of Low-level Optimizations for Quadrature Representation}
-\label{sec:summary-opts}
-To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.
-
-...TODO...
-
-The work in~\cite{Luporini} resulted in the development of COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for the optimization of local assembly kernels relying on quadrature representation.
-
-...TODO...
-
-Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.
-
-...TODO...
-
-% Far notare che low-level si, ma sono guidate da high-level knowledge! Questo in risposta alla domanda: ma allora perche' non sfruttare la domain knowledge che hai, che senso ha separare...
-
-% Ci sono due goal: minimizzare flops e specializzare
-
-% Vantaggio di automatizzare rispetto che scrivere a mano
-% which would be practically unfeasible to achieve by hand-written code
-
-\section{A Compiler for Optimizing Quadrature-based Integration}
+\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
 In order to generate high performance code, mathematical expressions evaluating the element tensor must be optimized with regards to several interrelated aspects: 1) minimization of floating point operations, 2) instruction-level parallelism, and 3) data locality.
@@
 A key point is that the ER has to perform transformations that do not break the code specializer. Having indirections is really dangerous...
+
+\subsection{Summary of Optimizations for Automated Quadrature Representation}
+\label{sec:summary-opts}
+...TODO... : clearly state, once more, we operate in the regime of automated code generation
+
+To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.
+
+...TODO...
+
+The work in~\cite{Luporini} resulted in the development of COFFEE, a compiler for the optimization of local assembly kernels relying on quadrature representation.
+
+...TODO...
+
+Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.
+
+...TODO...
+
+% Far notare che low-level si, ma sono guidate da high-level knowledge! Questo in risposta alla domanda: ma allora perche' non sfruttare la domain knowledge che hai, che senso ha separare...
+
+% Ci sono due goal: minimizzare flops e specializzare
+
+% Vantaggio di automatizzare rispetto che scrivere a mano
+% which would be practically unfeasible to achieve by hand-written code
 
@@
 ...TODO...: to what extent applying rewrite rules, is problem specific...
+
+...TODO...: things get interesting if the kernel operates on a batch of elements...can save a LOT of flops
 % blablabla molto complesse, nestings etc expose etc allora rewrite rules

### 2014-08-05 TOMS_Paper/assembly.tex
Add COFFEE section

@@
 
-% Quadrature. Tensor. Quadrature <, >, = Tensor
-%code evaluating element matrices using quadrature
-
 The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
@@
 
-In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines.
+In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation.
 We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization.
@@
 
-...TODO...: say that "assembly expression" =THE EXPRE// COMPUTATION OF ELEMENT MATRIX
+In the next sections, we will often refer to the (local) element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} of the variational problem.
 
 \subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
-In order to generate high performance code, mathematical expressions evaluating the element tensor must be optimized with regards to several interrelated aspects: 1) minimization of floating point operations, 2) instruction-level parallelism, and 3) data locality.
+\label{sec:summary-opts}
+If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality.
 In this paper, we tackle these three points building on our previous work (\cite{Luporini}).
-
-...We propose a novel structure of how we think a platform-independent, domain-specific optimizing compiler should look like....
 
@@
 
-The high-level structure of COFFEE is shown in Figure~\ref{fig:coffee-outline}.
+COFFEE is a mature, platform-independent compiler capable of optimizing local assembly code.
+Its high-level structure can be outlined as in Figure~\ref{fig:coffee-outline}.
+The input is either suitably-annoted C code, explicitly provided by the user, or a decorated abstract syntax tree (henceforth AST) representation of an assembly kernel, automatically generated by the higher level in the stack.
+We have integrated COFFEE with the Firedrake framework; here, the input to COFFEE is originated by the FEniCS Form Compiler in the form of an AST.
+Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE.
+This is used to introduce more effective, specialized optimizations, as well as to simplify implementation.
 
-...The Expression Rewriter is a software module implemented in COFFEE dealing with the first point...
+Two conceptually distinct software modules can be individuated: the Expression Rewriter and the Code Specializer.
+The former targets arithmetic intensity.
+Its role is to transform the assembly expression and, therefore, the enclosing loop nest, so as to minimize the number of floating point operations that are performed to evaluate the element matrix (or vector).
+The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality.
+The Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer.
+As described later, we have invested some effort to ensure that the code transformations applied in the expression rewriting stage do not break any code specialization opportunity.
+This for example would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.
 
-...The Code Specializer (henceforth CS) is...
-A key point is that the ER has to perform transformations that do not break the code specializer. Having indirections is really dangerous...
+To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work.
+Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to~\cite{Luporini} (as well as to FEniCS' built-in optimizations).
 
-\subsection{Summary of Optimizations for Automated Quadrature Representation}
-\label{sec:summary-opts}
-...TODO... : clearly state, once more, we operate in the regime of automated code generation
+In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation.
+We have also shown how different subset of optimizations are needed to achieve close-to-peak performance in different problems.
+In particular, we can summarize our study as follows
 
-To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.
+\begin{itemize}
+\item \emph{Generalized Loop-invariant Code Motion}.
+Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest.
+As briefly described in Section~\ref{sec:expr-rewriting}, we work around this limitation, while also achieving vectorization of invariant code.
+\item \emph{Padding and data alignment}.
+The small size of the loop nest require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code.
+Data alignment can be enforced as a consequence of padding.
+\item \emph{Vector-register Tiling}.
+Blocking at the level of vector registers, which we perform exploiting the specific memory access pattern of the assembly expressions (i.e.
+a domain-aware transformation), improves data locality beyond traditional unroll-and-jam optimizations.
+This is especially true for relatively high polynomial order (i.e.
+greater than 2) or when pre-multiplying functions are present.
+\item \emph{Expression Splitting}.
+In certain assembly expressions the register pressure is significantly high: when the number of basis functions arrays (or, equivalently, temporaries introduced by loop-invariant code motion) and constants is large, spilling to L1 cache is a consequence for architectures with a relatively low number of logical registers (e.g.
+16/32).
+We exploit sum's associativity to ``split'' the assembly expression into multiple sub-expressions, which are computed individually.
+\end{itemize}
 
-...TODO...
+In Figure~\ref{fig:coffee-outline}, we show where these transformations are logically applied in COFFEE; also, the contributions of this work are highlighted.
+In the next two sections, we describe the new functionalities of, respectively, the Expression Rewriter and the Code Specializer.
 
-The work in~\cite{Luporini} resulted in the development of COFFEE, a compiler for the optimization of local assembly kernels relying on quadrature representation.
-
-...TODO...
-
-Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.
-
-...TODO...
+%Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.
 
 % Far notare che low-level si, ma sono guidate da high-level knowledge! Questo in risposta alla domanda: ma allora perche' non sfruttare la domain knowledge che hai, che senso ha separare...
-
-% Ci sono due goal: minimizzare flops e specializzare
 
@@
 \label{sec:expr-rewriting}
-As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of a mathematical expression.
+As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of an assembly expression.
 The Expression Rewriter (henceforth ER) that we have designed and implemented in COFFEE enhances this technique by making two steps forward, which allow more redundant computation to be avoided.
@@
 An example is given in Figure~\ref{fig:tovectexpand-code}: \texttt{F1, F2, F3, T1} are all scalar quantities, and their computation will not be auto-vectorized (as explained in Section~\ref{sec:perf-results}, we experimented with the Intel and GNU compilers).
-The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I \cdot T^2)$, is too superficial, in general.
+The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I T^2)$, is too superficial, in general.
 There are notable cases, for instance problems based on hyperelasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.
@@
 
+% Quadrature. Tensor. Quadrature <, >, = Tensor
+
 In Figure~\ref{fig:helmholtz}, ...

### 2014-08-05 TOMS_Paper/assembly.tex
Rewrite Rewrite rules. General clean up

@@
 We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of \texttt{T1}, resulting in the code illustrated in Figure~\ref{fig:expanded-1-code}.
-Freeing the register destined to the constant \texttt{g} is more complicated: we cannot absorb it in the pre-computation of \texttt{T1} because the same array is accessed in the evaluation of \texttt{(T1[ j]*A[i][k])}.
+Freeing the register destined to the constant \texttt{g} is more complicated: we cannot absorb it in the pre-computation of \texttt{T1} because the same array is accessed in the evaluation of \texttt{(T1[j]*A[i][k])}.
 The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}.
@@
 We define, therefore, the state of a loop nest $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma$ maps invariant sub-expressions to identifiers of temporary arrays.
+The notation $\sigma_i$ refers to invariants hoisted at the level of loop $i$.
 We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma : Inv \rightarrow S$ such that
@@
 This is useful to express updates of invariant expressions.
+
+A generic (sub-)expression is represented with roman letter \texttt{a, b, ...}; constant terms are considered a special case, so greek letters \texttt{$\alpha$, $\beta$} are used instead.
+The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression.
+We will refer to $i_0$ as the outermost enclosing loop.
+The notation $a_i$, therefore, indicates that the expression $a$ assume distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.
+
 Rewrite rules for the ER are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
@@
 \centering
-\begin{spacing}{1.5}
+\begin{spacing}{1.35}
 \begin{align*}
+~~~~~~&Rule &~~&~~~~~~~~~~~~~~~~~~~~~~~~Precondition \\
 [a_i \cdot b_j]_{(\sigma, G)} &\rightarrow [a_i \cdot b_j]_{(\sigma, G)} ~~&~~&\\
@@
 [a_i \cdot b_j + a_i \cdot c_j]_{(\sigma, G)} &\rightarrow [(a_i \cdot (b_j + c_j)]_{(\sigma, G)} ~~&~~ &\\
-[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
-[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
+[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
+[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
 ~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
-[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
+[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
 ~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
@@
 
-%\begin{figure}[t]
-%\tiny
-%\centering     
-%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p1_formhelmholtz_speedup}}
-%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p2_formhelmholtz_speedup}}
-%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p3_formhelmholtz_speedup}}
-%\subfigure[]{\includesvg[scale=0.25]{perf-results/FiredrakeFormsCoffee_bar_q1_p4_formhelmholtz_speedup}}
-%\caption{Helmholtz results.}\label{fig:helmholtz}
-%\end{figure}
+\begin{figure}[t]
+\centering     
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p1_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p2_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p3_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p4_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p1_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p2_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p3_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p4_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p1_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p2_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p3_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p4_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p1_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p2_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p3_formhelmholtz_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p4_formhelmholtz_speedup}}
+\caption{Helmholtz results.}\label{fig:helmholtz}
+\end{figure}
+
+In Figure~\ref{fig:elasticity}, ...
+
+\begin{figure}[t]
+\centering     
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p1_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p2_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p3_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p4_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p1_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p2_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p3_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p4_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p1_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p2_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p3_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p4_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p1_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p2_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p3_formelasticity_speedup}}
+\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p4_formelasticity_speedup}}
+\caption{Elasticity results.}\label{fig:elasticity}
+\end{figure}
 
@@
 \label{sec:conclusions}
+
+We have presented ...
 

### 2014-08-05 TOMS_Paper/perf-results/generated_grid_plots.py
Working on the plot generator

@@ -1,0 +1,74 @@
+"""
+Generate a grid of plots given raw data generated by pybench 
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1 import Grid
+from mpltools import style
+
+import os
+import sys
+from collections import defaultdict
+from ast import literal_eval
+
+filename = "FiredrakeFormsCoffee.dat"
+
+style.use('ggplot')
+
+if len(sys.argv) is not 2:
+    print "Usage: directory-where-to-find-FiredrakeFormsCoffee.dat"
+    sys.exit(0)
+
+with open(os.path.join(sys.argv[1], filename), "r") as f:
+    timings = literal_eval(f.read())['timings']
+
+# First, adjust times in a dictionary as I need
+_timings = defaultdict(dict)
+for info, vals in sorted(timings.items()):
+    q, p, name, opt = info
+    _timings[(q, p)][opt] = vals
+    
+fig = plt.figure()
+grid = Grid(fig, rect=111, nrows_ncols=(4, 4), axes_pad=0.25, label_mode='L')
+for ax, plot_times in zip(grid, sorted(_timings.items())):
+    info, vals = plot_times
+    q, p = info
+    base = vals[(0, False, False, False)]
+    ffco = vals[(0, False, False, True)]
+    licm = vals[(1, True, False, False)]
+    auto = vals[(0, False, True, False)]
+
+    # Plot parameters
+    N = 3  # We show speedup of ffco, licm, and auto over base
+    ind = np.arange(N)  # the x locations for the groups
+    width = 0.15  # the width of the bars
+    fontsize = 15
+
+    # Compute speed ups
+    nf0 = (base['nf 0']/ffco['nf 0'], base['nf 0']/licm['nf 0'], base['nf 0']/auto['nf 0'])
+    nf1 = (base['nf 1']/ffco['nf 1'], base['nf 1']/licm['nf 1'], base['nf 1']/auto['nf 1'])
+    nf2 = (base['nf 2']/ffco['nf 2'], base['nf 2']/licm['nf 2'], base['nf 2']/auto['nf 2'])
+    nf3 = (base['nf 3']/ffco['nf 3'], base['nf 3']/licm['nf 3'], base['nf 3']/auto['nf 3'])
+
+    # Draw the bars
[…]

### 2014-08-05 TOMS_Paper/assembly.tex
Now use newer pictures

@@
 \begin{figure}[t]  
-\centering     
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p1_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p2_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p3_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q1_p4_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p1_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p2_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p3_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q2_p4_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p1_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p2_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p3_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q3_p4_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p1_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p2_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p3_formhelmholtz_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/helmholtz/FiredrakeFormsCoffee_bar_q4_p4_formhelmholtz_speedup}}
+\includegraphics[scale=0.7]{perf-results/helmholtz}
 \caption{Helmholtz results.}\label{fig:helmholtz}
@@
 \begin{figure}[t]  
-\centering     
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p1_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p2_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p3_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q1_p4_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p1_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p2_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p3_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q2_p4_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p1_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p2_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p3_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q3_p4_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p1_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p2_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p3_formelasticity_speedup}}
-\subfigure{\includegraphics[scale=0.17]{perf-results/elasticity/FiredrakeFormsCoffee_bar_q4_p4_formelasticity_speedup}}
+\includegraphics[scale=0.7]{perf-results/elasticity}
 \caption{Elasticity results.}\label{fig:elasticity}
 \end{figure}
-
 % Mass DGEMM utile explicit method etc

### 2014-08-06 TOMS_Paper/assembly.tex
Rewrite lots of stuff...

@@
 \subfigure[Original code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
-~~~~~~~~~~~~~
+~~~~~~~~~~~~~~~~
+\subfigure[Invariant code]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
+~\\
+~\\
 \subfigure[Factorized code]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
-\caption{Original and factorized code.}
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subfigure[Scalar-expanded code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
+\caption{Original, invariant, factorized, and then scalar-expanded code.}
 \end{figure}
 
-Consider the simplified example of the element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a real Burgers problem.
-In practice, depending on the problem, the expression tree could be much more complex, with multiple levels of nesting.
-The example is however representative for a large class of problems, so we will use it throughout the rest of the paper for illustrative purpose.
+Consider the element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem.
+The assembly expression, produced by the FEniCS Form Compiler, has been deliberately simplified, and code details have been omitted for brevity and readability.
+In practice, assembly expressions can be much more complex, for example depending on the differential operators employed in the variational form; however, this example is representative enough for highlighting patterns that are common in a large class of problems.
 
-A first glimpse of the code suggests that the sub-expression \texttt{F[i][j]*d + G[i][j]*e} is invariant with respect to the innermost loop, so it should be hoisted at the level of the outer loop \texttt{j}.
-This is a standard compiler transformation, which is supported by any available compilers.
-With a closer look we notice that the sub-expression \texttt{a*C[i][k] + b*E[i][k]} is also invariant, although, in this case, with respect to the outer loop \texttt{j}.
-In~\cite{Luporini}, we have showed that a \textit{generalized} loop-invariant code motion transformation - that is, given a non-trivial expression, the capability of analyzing all of its sub-expressions with respect to the enclosing loops to determine what code is hoistable- is not supported by available compilers.
+A first glimpse of the code suggests that the sub-expression \texttt{a*f0*A[i][j]+b*f1*B[i][j]} is invariant with respect to the innermost (trial functions) loop \texttt{k}, so it can be hoisted at the level of the outer loop \texttt{j} to avoid redundant computation.
+This is indeed a standard compiler transformation, supported by any available compilers, so, in principle, there should be no need to transform the source code explicitly.
+With a closer look we notice that the sub-expression \texttt{d*D[i][k]+e*E[i][k]} is also invariant, although, this time, with respect to the outer (test functions) loop \texttt{j}.
+In~\cite{Luporini}, we have showed that available compilers limit the search for code motion opportunities to the innermost loop of a given loop nest.
 Moreover, the lack of cost models to ascertain both the optimal place where to hoist an expression and whether or not vectorizing it at the price of extra temporary memory is a fundamental limiting factor.
-We have addressed these problems by implementing a generalized loop-invariant code motion transformation in COFFEE.
+The \textit{generalized loop-invariant code motion} technique that we implemented in COFFEE, leading to the code in Figure~\ref{fig:invariant-code}, addressed these limitations, which are critical in the case of non-trivial assembly expressions, resulting in run-time improvements up to 3$\times$.
 
-We now consider the case of forms that ``hide'' further opportunities for code hoisting.
-Such forms are by no means exotic: for example, the pattern described next is commonly found in elastic problems.
-By examining again the example in Figure~\ref{fig:original-code}, we notice that the basis function array \texttt{B}, iterating along the iteration space \texttt{[i,j]}, appears twice in the expression.
-If we expand the products in which \texttt{B} is accessed, we can apply product commutativity and then factorize the expression as in Figure~\ref{fig:factorized-code}.
-This has two effects: firstly, it reduces the total number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{(A[i][k]/c + T2[k]*f)} invariant with respect to loop \texttt{j}.
-Therefore, code hoisting can be performed.
+In this paper, we target particularly complex variational forms for which a more aggressive transformation is required.
+We generalize the problem of finding and hoisting invariant sub-expressions by formulating the following critical question: \textit{how an assembly expression should be rewritten so as to minimize both floating point operations and data movement}?
+To minimize floating point operations, in terms of instructions executed, it is important to reduce the absolute number of arithmetic operations to evaluate the element matrix as well as to achieve full vectorization of the assembly loop nest.
 
-The second observation we make concerns the register pressure induced by the expression.
-Once loop-invariant terms are lifted, we can think about data locality and, in particular, register allocation.
-Assume the local assembly kernel is executed on a state-of-the-art architecture having 16 logical registers, e.g.
+We start considering the case of assembly expressions that, after generalized loop-invariant code motion has been applied, still  ``hide'' opportunities for code hoisting.
+By examining again the code in Figure~\ref{fig:invariant-code}, we notice that the basis function array \texttt{A} iterating along the \texttt{[i,j]} loops appears twice in the expression.
+By expanding the products in which \texttt{A} is accessed and applying sum commutativity, we can factorize the expression.
+This has two effects: firstly, it reduces the number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{A[i][k]/c+T2[k]*f} invariant with respect to loop \texttt{j}.
+Consequently, hoisting can be performed, resulting in the code in Figure~\ref{fig:factorized-code}.
+In general, exposing factorization opportunities requires traversing the whole expression tree, and then expanding and moving terms.
+It also needs heuristics to select a factorization strategy: there may be different opportunities of reorganizing sub-expressions, and, in our case, the best is the one that maximizes the invariant code eventually disclosed.
+We will discuss this aspect formally in Section~\ref{fig:rewrite-rules}.
+
+As a second observation, we note that integration-dependent expressions are inherently executed as scalar code.
+For example, the \texttt{f0*a} and \texttt{f1*b} products in Figure~\ref{fig:factorized-code} depend on the loop along quadrature points; these operations are performed in a non-vectorized way at every \texttt{i} iteration.
+This is not an issue in our running example, where scalar computation represents a small percentage of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}).
+In such forms, the amount of computation independent of both test and trial functions loops is so large that it has a significant impact on the run-time, despite being executed only $O(I)$ times (with $I$ number of quadrature points).
+We have therefore implemented an algorithm to scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.
+
+The third point concerns the register pressure induced by the assembly expression.
+Once the code has been optimized for arithmetic intensity, it is important to think about how the transformations impacted register allocation.
+Assume the local assembly kernel is executed on a state-of-the-art CPU architecture having 16 logical registers, e.g.
 an Intel Haswell.
 Each value appearing in the expression is loaded and kept in a register as long as possible.
-For instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every $i$ iteration.
-This implies that at every iteration of the \texttt{jk} loop nest, 12$\%$ of the available registers are spent just to store constant values.
-In more complicated expressions, the percentage of registers destined to store loop-invariant terms can be even higher.
-Registers are, however, a precious resource, especially when evaluating intensive expressions.
-The smaller is the number of free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e.
-it can worsen data locality), or it may prevent the effective application of standard transformations like loop unrolling.
-The ER works around this problem by suitably expanding terms and introducing, where necessary, new temporary values.
+In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every $i$ iteration.
+This implies that at every iteration of the \texttt{jk} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops.
+In more complicated expressions, the percentage of registers destined to store such constant terms can be even higher.
+Registers are, however, a precious resource, especially when evaluating compute-intensive expressions.
+The smaller is the number of available free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e.
+it can worsen data locality), or it may prevent the effective application of standard transformations, e.g.
+loop unrolling.
+We aim at relieving this problem by suitably expanding terms and introducing, where necessary, additional temporary values.
+We illustrate this in the following example.
 
 %An analogous analysis applies to processors with larger numbers of registers, since using loop unroll or loop unroll-and-jam to expose more instruction-level parallelism would increase the requirements on registers.
-
-\begin{figure}
-\tiny
-\lstinputlisting{listings/toexpand.code}
-\caption{Expandable code}\label{fig:toexpand-code}
-\end{figure}
-
-Consider the variant of the running (transformed) example shown in Figure~\ref{fig:toexpand-code}.
-Again, this is a representative example of what happens in real finite element forms.
-We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of \texttt{T1}, resulting in the code illustrated in Figure~\ref{fig:expanded-1-code}.
-Freeing the register destined to the constant \texttt{g} is more complicated: we cannot absorb it in the pre-computation of \texttt{T1} because the same array is accessed in the evaluation of \texttt{(T1[j]*A[i][k])}.
-The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}.
-Generalizing, this is a problem of data dependencies; in order to solve it, we employ a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}.
-The dependency graph is initially empty; every time a new temporary is created due to loop-invariant code motion or expansion of terms is performed, it is updated by suitably adding vertices and edges.
 
@@
 \centering
+\subfigure[Expandable code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
+~\\
 \subfigure[Expanded 1 code]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
@@
 \subfigure[Expanded 2 code]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
-\caption{Expanded code.}\label{fig:expanded-code}
+\caption{Expanding code.}\label{fig:expanded-code}
 \end{figure}
 
+Consider a variant of the Burgers local assembly kernel, shown in Figure~\ref{fig:toexpand-code}.
+This is still a representative, simplified example.
+We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of the invariant sub-expression stored in \texttt{T1}, resulting in code as in Figure~\ref{fig:expanded-1-code}.
+Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since the same array is accessed in \texttt{(T1[j]*A[i][k])}.
+The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}.
+Generalizing, this is a problem of data dependencies: to solve it, we use a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}.
+The dependency graph is initially empty, and it is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms.
+The dependency graph is then queried to understand when expansion can be performed without resorting to new temporary values.
+This aspect is formalized in the next section.
+
 \subsection{Rewrite Rules}
-In general, assembly expressions produced by automated code generation can be much more complex (more terms and operations involved) and nested.
-Our goal is to establish a portable, platform- and compiler-independent, and systematic way of reducing the strength of an expression.
-The technique should be simple; definitely it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems.
-Ideally, it should be naturally extendible to problems that will be supported in next releases of state-of-the-art frameworks like Firedrake and FEniCS: for instance, explicit support for outer-product finite elements will enable generation of kernels with much deeper loop nests, and the ER should transparently be able to deal with these structures as well.
+\label{sec:rewrite-rules}
+In general, assembly expressions produced by automated code generation can be much more complex than those we have used as examples, with dozens of terms involved (basis function arrays, derivatives, coefficients, ...) and hundreds of (nested) arithmetic operations.
+Our goal is to establish a portable, unified, platform-independent, and systematic way of reducing the computational strength of an expression exploiting the intuitions described in the previous section.
+This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems.
+In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform scalar-expansion to achieve full vectorization of the assembly code.
 
-To address these issues, we have based the implementation of the ER in COFFEE on a set of formal rewrite rules.
-By applying these rules, it is possible to derive how an expression will be transformed, as well as what and where (i.e.
-at which level in the loop nest) temporaries will be introduced.
-When applying a rule, the ER needs to update the state of the loop nest, to reflect, for example, the use of a new temporary and the newly created data dependencies.
-We define, therefore, the state of a loop nest $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma$ maps invariant sub-expressions to identifiers of temporary arrays.
+We centre such an algorithm around a set of rewrite rules.
+These rules drive the transformation of an expression, prescribe where invariant sub-expressions will be moved (i.e.
+at what level in the loop nest), and track the propagation of data dependencies.
+When applying a rule, the state of the loop nest must be updated to reflect, for example, the use of a new temporary and new data dependencies.
+We define the state of a loop nest as $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma : Inv \rightarrow S$ maps invariant sub-expressions to identifiers of temporary arrays.
 The notation $\sigma_i$ refers to invariants hoisted at the level of loop $i$.
-We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma : Inv \rightarrow S$ such that
+We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma$ such that
 \begin{gather*}
@@
 \end{gather*}
-That is, intuitively, if the invariant expression \texttt{x} has already been hoisted, then return the temporary identifiers that hosts its value; otherwise, hoist the expression.
-There is a special case when $v = \perp$, used to delete entries in $\sigma$.
-Specifically:
+That is, if the invariant expression \texttt{x} has already been hoisted, $[]$ returns the temporary identifier hosting its value; otherwise, \texttt{x} is hoisted and a new temporary $v$ is created.
+There is a special case when $v = \perp$, used for \textit{conditional deletion} of entries in $\sigma$.
+Specifically
 \begin{gather*}
@@
 \sigma(x) \text{~~~~~~~~if $x \in Inv$; $\sigma = \sigma \setminus (x, \sigma(x))$}\\
-t \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $t \notin Inv$}\\
+v \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $v \notin S$}\\
 \end{cases}
 \end{gather*}
-In other words, the previously hoisted expression \texttt{x} is removed (if any) and the temporary identifier that was hosting its value is returned.
-This is useful to express updates of invariant expressions.
+In other words, the invariant sub-expression \texttt{x} is removed and the temporary identifier that was hosting its value is returned if \texttt{x} had been previously hoisted; otherwise, a fresh identifier $v$ is returned.
+This is useful to express updates of hoisted invariant sub-expressions when expanding terms.
 
-A generic (sub-)expression is represented with roman letter \texttt{a, b, ...}; constant terms are considered a special case, so greek letters \texttt{$\alpha$, $\beta$} are used instead.
+In the following, a generic (sub-)expression is represented with roman letter \texttt{a, b, ...}; constant terms are considered a special case, so greek letters \texttt{$\alpha$, $\beta$} are used instead.
 The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression.
@@
 
-Rewrite rules for the ER are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
-Conceptually, the ER visits the expression tree from the root, which is the outermost operation, and applies the transformations dictated by the rewrite rules.
-As an example, one can try instantiating the rules in the code of Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized code in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} is obtained, respectively.
+Rewrite rules for expression rewriting are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
+The Expression Rewriter applies the rules while performing a depth-first traversal of the assembly expression tree.
+Given an arithmetic operation between two sub-expressions (i.e.
+a node in the expression tree), we first need to find an applicable rule.
+There cannot be ambiguities: only one rule can be matched.
+If the preconditions of the rule are satisfied, the corresponding transformation is performed; otherwise, no rewriting is performed, and the traversal proceeds.
+As examples, it is possible to instantiate the rules in the codes shown in Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized codes in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} are obtained, respectively.
 
@@
 
-... TODO... : each level in sigma is then taken and wrapped in a loop
-
-...TODO...: to what extent applying rewrite rules, is problem specific...
-
-...TODO...: things get interesting if the kernel operates on a batch of elements...can save a LOT of flops
-% blablabla molto complesse, nestings etc expose etc allora rewrite rules
-
-% we will show the effects of this stuff over plain LICM in section
-
-% hoisting constants release registers!!!!
+To what extent should rewrite rules be applied is a question that cannot be answered in general.
+In some problems, a full rewrite of the expression may be the best option; in other cases, on the other hand, an aggressive expansion of terms, for example, may lead to high register pressure in the loops computing invariant terms, worsening the performance.
+In Section~\ref{sec:autotune} we explain how to leverage the Code Specializer to select a suitable rewriting strategy for the problem at hand.
 
@@
 
-\subsection{Vector-expansion of Invariant Terms}
-\label{sec:vect-exp}
-The Expression Rewriter's loop-invariant code motion hoists invariant sub-expressions in the body of the outermost loop along which they iterate.
-If such a sub-expression computes a value depending on the outermost loop only, then the only way of vectorizing it -- if we exclude superword level parallelism~\cite{SLP}, which is in general not applicable to our kernels -- is by vector-expansion of terms.
-
-\begin{figure}[t]
-\tiny
-\centering     
-\subfigure[To vector expand code]{\label{fig:tovectexpand-code}\lstinputlisting{listings/tovectexpand.code}}
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subfigure[Vector-expanded code]{\label{fig:vectexpanded-code}\lstinputlisting{listings/vectexpanded.code}}
-\caption{Before and after vector expansion}\label{fig:skip-code}
-\end{figure}
-
-An example is given in Figure~\ref{fig:tovectexpand-code}: \texttt{F1, F2, F3, T1} are all scalar quantities, and their computation will not be auto-vectorized (as explained in Section~\ref{sec:perf-results}, we experimented with the Intel and GNU compilers).
-The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I T^2)$, is too superficial, in general.
-There are notable cases, for instance problems based on hyperelasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.
-
-We have implemented a procedure that vector-expands scalar expressions, while lifting them outside of the loop nest as shown in Figure~\ref{fig:vectexpanded-code}.
-The dependency graph is queried to assess the legality of the transformation.
-At the price of extra memory, vector-expanded terms can now be vectorized by the compiler.
-
-% Expose vectorizability of the outermost loop 
-% Render the loop nest perfect
-
 \subsection{Exposing Linear Alegbra Operations}
@@
 Terms in the bodies of the various enclosing loops (e.g.
-coefficients evaluation at quadrature point, temporaries introduced by the Expression Rewriter) are vector-expanded and hoisted outside of the loop nest, as explained in Section~\ref{sec:vect-exp}.
+coefficients evaluation at quadrature point, temporaries introduced by rewriting the assembly expression) are vector-expanded and hoisted outside of the loop nest, as explained in Section~\ref{sec:vect-exp}.
 This implies that the loop nest is now perfect; that is, there is no intervening code among the various loops.
-The element matrix evaluation became a sequence of perfect loop nests, each loop nest computing a dense matrix-matrix multiply between temporaries hoisted by the Expression Rewriter or tabulated basis functions.
+The element matrix evaluation has now become a sequence of perfect loop nests, each loop nest computing a dense matrix-matrix multiply between temporaries hoisted by the Expression Rewriter or tabulated basis functions.
 Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g.
@@
 \subsection{Model-driven Dynamic Autotuning}
+\label{sec:autotune}
 We have demonstrated in~\cite{Luporini} that an optimal set of code transformations does not exist.

### 2014-08-07 TOMS_Paper/assembly.tex
Rewrite zero-columns avoidance

@@
 \label{sec:zeros}
-The second task of the ER consists of skipping arithmetic operations over blocks of zero-valued entries in basis function and derivatives of basis functions arrays.
-Zero columns in such arrays arise, for example, when taking derivatives on a reference element and when employing mixed elements.
-In~\cite{quadrature1}, a code transformation to eliminate floating point operations on zero columns was introduced; the technique was implemented in FEniCS.
+We aim at skipping arithmetic operations over blocks of zero-valued entries in basis functions (and their derivatives) arrays.
+Zero-valued columns arise, for example, when taking derivatives on a reference element and when employing mixed elements.
+In~\cite{quadrature1}, a technique to avoid operations on zero-valued columns based on the use of indirection arrays was described (e.g.
+\texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}) and implemented in the context of FEniCS.
 We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}.
-We propose a strategy in which we avoid the use of indirection arrays in the generated code (e.g.
-\texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), which otherwise would break the optimizations applicable at the Code Specializer level, especially SIMD vectorization.
+Essentially, our strategy avoids indirection arrays in the generated code, which otherwise would break the optimizations applicable at the Code Specializer level, including SIMD vectorization.
 
@@
 
-Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the transformed Burgers excerpt in Figure~\ref{fig:factorized-code}.
+Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}.
 The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh.
 The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points.
-We notice the presence of 4 zero-columns.
-Any multiplications or additions along these columns could (should) be skipped in order to minimize irrelevant floating point operations.
-One solution, as anticipated, is not to generate the zero columns, i.e.
-to generate a dense 6$\times$2 array, to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g.
-$ind = \lbrace 3, 5\rbrace$) to access the right entries in the element tensor $A$.
-As explained, this prevents, among the various optimizations, effective code vectorizability, because memory loads and store would eventually reference non-contiguous locations.
+There are four zero-valued columns.
+Any multiplications or additions along these columns could (should) be skipped to avoid irrelevant floating point operations.
+The solution adopted in~\cite{quadrature1} is not to generate the zero-valued columns (i.e.
+to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g.
+$ind = \lbrace 3, 5\rbrace$) to update the right entries in the element tensor $A$.
+This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations.
 %on the other hand, reshrinki tutto su un array solo...
 
-Our approach is based on domain knowledge and symbolic execution.
-We discern the origin of zero-columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements.
-In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to preserve the correctness of the element matrix evaluation while iterating along the space of test and trial functions.
-The two key observations are that 1) the number of zero-columns caused by mixed elements is much larger then that due to derivatives, and 2) such columns are contiguous in memory.
-Based on this, we decide to skip any computation involving zero-columns induced by mixed elements, whereas we still iterate over any other zero-columns.
-Consequently, the only modifications to the code that are eventually needed are adjusting loop bounds and introducing offsets when accessing the element matrix, as shown in Figure~\ref{fig:withzeros-skipped-code}.
-In general, multiple iteration spaces over test and trial functions (i.e.
-\texttt{jk} loops), characterized by their own loop bounds, may be needed, increasing loop overhead and decreasing data locality for the element matrix.
+Our approach is based on using domain knowledge and symbolic execution.
+We discern the origin of zero-valued columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements.
+In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to correctly evaluate the local element matrix while iterating along the space of test and trial functions.
+The two key observations are that 1) the number of zero-valued columns caused by using vector function spaces is, often, much larger then that due to derivatives, and 2) such columns are contiguous in memory.
+Based on this, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.
 
-The implementation is based on symbolic execution.
-The ER expects information from the higher layer about where the zero-blocks are located, in each tabulated basis function.
-This could come in the form of code annotation if the input to COFFEE were provided as a C string (as shown in Figure~\ref{fig:withzeros-skipped-code}) or by suitably decorating basis functions nodes in the abstract syntax tree representing the local assembly routine.
-Then, each statement in the local assembly loop nest is symbolically executed.
-For example, consider the statement \texttt{T2[r] = ((d*D[i][k])+(e*E[i][k]));} in Figure~\ref{fig:withzeros-skipped-code}.
-Array \texttt{D} has non-zeros in positions $NZ_D = [3,5]$; we also assume \texttt{E} has non-zeros in positions $NZ_E = [0,2]$.
-Multiplications by scalar do not affect the propagation of zero columns.
-On the other hand, when summing the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we record that the target identifier \texttt{T2} will have non-zero columns in positions $NZ_D \cup NZ_E = [0-5]$.
-Exploiting the $NZ$ information associated with each identifier, the assembly expression is split into several sets of sub-expressions, each set characterized by accesses to the same range of non-zero columns and, therefore, iterating over a suitably-sized iteration space.
-In our example, there is just one set, which contains two sub-expressions, enclosed in a doubly nested loop of size 3$\times$3.
+%Our example is then transformed as in Figure~\ref{fig:withzeros-skipped-code}: loop bounds are adjusted and suitable offsets are introduce to access the element matrix and basis function arrays.
+In general, the element matrix evaluation may also have to be split over multiple iteration spaces (test and trial functions loops), each iteration space characterized by its own loop bounds; this has the side effect of both increasing loop overhead and decreasing data locality.
 
-...TODO...: dire come si riaggiustano gli indici quando si vuole usare AVX ma la parte dense non e' multiplo del vector length - se non e' paddata la tail, allora si decrementa lo starting index di un pochino, tanto sono zeri...
+We achieve our goal by symbolic execution.
+The Expression Rewriter expects some indication in the input exposing the location of the zero-valued columns due to mixed (vector) function spaces, in each tabulated basis function.
+This indication comes either in the form of code annotation, if the input to COFFEE were provided as pure C (as shown in Figure~\ref{fig:withzeros-code}), or by suitably decorating basis functions nodes, if the input were an abstract syntax tree.
+Then, the rewrite rules are applied and each statement is executed symbolically.
+For example, consider the assignment \texttt{T2[r] = d*D[i][k]+e*E[i][k]} in Figure~\ref{fig:withzeros-code}.
+Array \texttt{D} has non-zero-valued columns in positions $NZ_D = [3,5]$; we also assume array \texttt{E} has non-zero-valued columns in positions $NZ_E = [0,2]$.
+Multiplications by scalar values do not affect the propagation of non-zero-valued columns.
+On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in positions $NZ_D \cup NZ_E = [0,5]$.
+Eventually, exploiting the $NZ$ information computed and associated with each identifier, we split the original assembly expression into multiple sets of sub-expressions, each set characterized by the same range of non-zero-valued columns.
+In our example, assuming that $NZ_{T1} = [3,5]$ and $NZ_A = [3,5]$, there are two of such sets, this leads to the generation of two distinct iteration spaces, one for each set, as in Figure~\ref{fig:withzeros-skipped-code}.
 
@@
 \label{sec:code-spec}
-The Code Specializer is the second software component of COFFEE.
-It provides a range of code transformations tailored for optimizing instruction-level parallelism and register locality.
+Code specialization provides a range of code transformations tailored to optimizing for instruction-level parallelism and register locality.
 As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying.
-In this paper, we enrich this set of transformations as detailed in the following sections.
+In this paper, we enrich this set of transformations as explained in the following sections.
+
+\subsection{Padding and Data Alignment Revisited}
+Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided.
+We recall effective SIMD vectorization can be achieved only if the innermost loop size is a multiple of the vector register length.
+In the case of an AVX instruction set, for example, we want the size of innermost loops to be multiples of 4 (vector registers can fit up to four double-precision floats).
+Moreover, efficient loads and stores can be issued only if all of the involved arrays' base addresses are multiple of the vector register length.
+
+Consider again the code in Figure~\ref{fig:withzeros-skipped-code}.
+The arrays in the loop nest $[j1,k1]$ can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region $M[0:3][6:8]$ will be discarded.
+Then, by explicitly aligning arrays and using suitable pragmas (e.g.
+\texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained.
+
+There are some complications in the cose of loops $[j0,k0]$.
+Here, increasing the loop bound to 4, despite being still safe, has no effect: for instance, the starting addresses of load instructions are \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned.
+One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$, which would mean losing the effect of expression rewriting.
+Another possibility is to attain to non-aligned accesses.
+COFFEE can generate code for both situations, so we leave the autotuner, described in Section~\ref{sec:autotune}, in charge of determining the optimal transformation.
+
+Note that in some cases it is not even safe to increase loop bounds, since extra iterations would access non-zero entries in the local element matrix $M$.
+In such a situation, there may be no possibility of restoring data alignment at all.
 

### 2014-08-07 TOMS_Paper/assembly.tex
Rewrite BLAS and add poisson

@@
 \subsection{Exposing Linear Alegbra Operations}
-In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrices multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS.
-We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM calls.
-The study was conducted only in the case of a specific Helmholtz equation.
-There are cases, however, in which tabulated basis functions sizes grow up to a point for which certain BLAS implementations allow to achieve a better performance.
-For example, this may be the case of relatively-high polynomial orders or when the form is characterized by the presence of a number of pre-multiplying functions.
-To explore these extreme yet meaningful scenarios, we have developed an algorithm to reduce any generic element matrix evaluation to a sequence of DGEMM calls.
+In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrix-matrix multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS.
+We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM routines, which are usually tuned for large arrays.
+The study was conducted only in the case of a specific Helmholtz problem.
+There are scenarios, however, in which tabulated basis functions sizes grow up to a point for which turning to BLAS may actually lead to better performance.
+For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of tabulated basis functions.
+To explore these scenarios, we have developed an algorithm that reduces any assembly expression evaluation to a sequence of DGEMM calls.
 
-The algorithm is shown in Figure~\ref{fig:algo-dgemm}.
-By using the rewrite rules in Figure~\ref{fig:rewrite-rules}, it can be proved that the Expression Rewriter can always reduce an assembly expression to a summation, over each quadrature point, of outer products along the test and trial functions.
+Here, we informally provide the main steps of the algorithm.
+By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions.
 Each outer product is then isolated, i.e.
-the assembly expression is split into chunks, each chunk representing an outer product.
-Terms in the bodies of the various enclosing loops (e.g.
-coefficients evaluation at quadrature point, temporaries introduced by rewriting the assembly expression) are vector-expanded and hoisted outside of the loop nest, as explained in Section~\ref{sec:vect-exp}.
-This implies that the loop nest is now perfect; that is, there is no intervening code among the various loops.
-The element matrix evaluation has now become a sequence of perfect loop nests, each loop nest computing a dense matrix-matrix multiply between temporaries hoisted by the Expression Rewriter or tabulated basis functions.
+the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions.
+Statements in the bodies of the surrounding loops (e.g.
+coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:rewriting-rules}.
+This renders perfect the loop nest; that is, there is no intervening code among the various loops.
+The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
+\begin{equation*}
+A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
+\end{equation*}
+where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time.
 Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g.
 two dimensional arrays are flatten as one dimensional arrays).
-The translation of each loop nest into a DGEMM call is the last, straightforward step.
+The translation into a sequence of DGEMM calls is the last, straightforward step.
 

### 2014-08-07 TOMS_Paper/perf-results/generated_grid_plots.py
Add hyperelasticity and update plotter script

@@ -28,47 +28,88 @@ for info, vals in sorted(timings.items()):
     q, p, name, opt = info
     _timings[(q, p)][opt] = vals
     
-fig = plt.figure()
-grid = Grid(fig, rect=111, nrows_ncols=(4, 4), axes_pad=0.25, label_mode='L')
-for ax, plot_times in zip(grid, sorted(_timings.items())):
-    info, vals = plot_times
-    q, p = info
-    base = vals[(0, False, False, False)]
-    ffco = vals[(0, False, False, True)]
-    licm = vals[(1, True, False, False)]
-    auto = vals[(0, False, True, False)]
+if sys.argv[1] in ['helmholtz', 'poisson', 'elasticity']:
+    fig = plt.figure()
+    grid = Grid(fig, rect=111, nrows_ncols=(4, 4), axes_pad=0.25, label_mode='L')
+    for ax, plot_times in zip(grid, sorted(_timings.items())):
+        info, vals = plot_times
+        q, p = info
+        base = vals[(0, False, False, False)]
+        ffco = vals[(0, False, False, True)]
+        licm = vals[(1, True, False, False)]
+        auto = vals[(0, False, True, False)]
 
-    # Plot parameters
-    N = 3  # We show speedup of ffco, licm, and auto over base
-    ind = np.arange(N)  # the x locations for the groups
-    width = 0.15  # the width of the bars
-    fontsize = 15
+        # Plot parameters
+        N = 3  # We show speedup of ffco, licm, and auto over base
+        ind = np.arange(N)  # the x locations for the groups
+        width = 0.15  # the width of the bars
+        fontsize = 15
 
-    # Compute speed ups
-    nf0 = (base['nf 0']/ffco['nf 0'], base['nf 0']/licm['nf 0'], base['nf 0']/auto['nf 0'])
-    nf1 = (base['nf 1']/ffco['nf 1'], base['nf 1']/licm['nf 1'], base['nf 1']/auto['nf 1'])
-    nf2 = (base['nf 2']/ffco['nf 2'], base['nf 2']/licm['nf 2'], base['nf 2']/auto['nf 2'])
-    nf3 = (base['nf 3']/ffco['nf 3'], base['nf 3']/licm['nf 3'], base['nf 3']/auto['nf 3'])
+        # Compute speed ups
[…]

### 2014-08-07 TOMS_Paper/assembly.tex
Add first description of perf setup

@@
 \label{sec:autotune}
-We have demonstrated in~\cite{Luporini} that an optimal set of code transformations does not exist.
-What sequence of transformations applying to a problem to optimize its performance depends on a broad range of factors, including mathematical structure of the input form, polynomial order of employed function spaces, presence of pre-multiplying functions, and, of course, the characteristics of the underlying architecture.
+We have demonstrated in~\cite{Luporini} that determining the sequence of transformations that  maximizes the performance of a problem requires investigating a broad range of factors, including mathematical structure of the input form, polynomial order of employed function spaces, presence of pre-multiplying functions, and, of course, the characteristics of the underlying architecture.
 
-In~\cite{Luporini}, we proposed a simple cost model that COFFEE could use to choose, given a problem, the transformations that were likely to maximize the performance.
-By having added a significant number of options to the set of possible optimizations, the selection problem is now far more challenging.
-The sole Expression Rewriter, for instance, could apply rewrite rules up to a set of pre-established depths, leading to the generation of many possible code variants.
+In~\cite{Luporini}, we proposed a simple cost model to select, for a given a problem, the optimal combination of transformations.
+In this paper, we have added a significant number of options to the set of possible optimizations, so the selection problem is now far more challenging.
+The sole Expression Rewriter, for instance, could generate many possible code variants by applying rewrite rules to different extents.
+We found difficult extending the cost model to effectively cover such a large search space.
 
-We tackle this problem by compiler autotuning.
-For each problem, not only does it allow to determine the best combination of transformations out of the set presented so far, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying multiple unroll factors and loop permutations.
+We tackle the optimization selection problem by compiler autotuning.
+Not only does it allow to determine the best combination of transformations out of the set presented so far, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying different unroll factors and loop permutations.
 By leveraging the cost model defined in our previous study, domain-awareness, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementation.
@@
 
-The structure of the autotuner is outlined in Figure~\ref{fig:autotuner}.
 COFFEE analyzes the input problem and decides what variants it is worth testing, as described later.
-It then provides the autotuner with all of the possible variants, in the form of abstract syntax trees.
+Each variant is obtained by requesting specific transformations to the Expression Rewriter and the Code Specializer.
+The possible variants are then provided to the autotuner, in the form of abstract syntax trees.
 The autotuner is a template-based code generator.
-By inspecting an abstract syntax tree, it determines how to generate ``wrapping'' code that, repeatedly in a \emph{while} loop executed a pre-established amount of time (order of milliseconds), 1) initializes kernel's input variables with fictitious values and 2) calls the kernel.
+By inspecting an abstract syntax tree, it determines how to generate ``wrapping'' code that 1) initializes kernel's input variables with fictitious values and 2) calls the kernel.
+These two points are executed repeatedly in a \emph{while} loop for a pre-established amount of time (order of milliseconds).
 At the exit of the \emph{while} loop, the times the kernel was invoked is recorded.
-Eventually, the variant executing the largest number of iterations is designated as the fastest implementation.
+Eventually, the variant executed the largest number of iterations is designated as the fastest implementation.
 Suitable compiler directives are used to prevent inlining of all function calls: this avoids the situation in which some variants are inlined and some are not, which would fake the autotuner's output.
@@
 \end{itemize}
-Also, the following heuristics, which capture properties of our computational domain, are used
+The autotuner is also domain-aware: the following heuristics, which capture properties of the computational domain, are exploited
 \begin{itemize}
-\item When both test and trial functions derive from the same function space, the lengths of their corresponding loops are identical.
-In this case, since for the employed storage layout the memory accesses are symmetrical along these two loops, their interchange is pruned from the search space.
-\item In general, if relatively high polynomial orders are employed, the loop nest is larger.
-In these cases, in order to avoid testing too many unroll factors, we impose a bound $X$ on the loop nest's overall unroll factor, which we found empirically.
-For example, the sum of the unroll factors of all outer loops cannot exceed $X$.
-\item On the other hand, if the polynomial order is low, i.e. for smaller loop nests, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
+\item The lengths of test and trial functions loops are identical in some cases, for example when they originate from the same function space.
+In such cases, since for the employed storage layout the memory access pattern is symmetrical along these two loops, we prune their interchange from the search space.
+\item The larger is the polynomial order of the method, the larger is the assembly loop nest.
+In these cases, we impose a bound on the loop nest's overall unroll factor (which we found empirically) to avoid uselessly testing too many unroll factors.
+\item On the other hand, if the polynomial order is low, i.e. when the loop nest is small, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
 \item We select two levels of expression rewriting.
-In the ``base'' level, generalized loop-invariant code motion only is applied.
-This means that only a subset of the rewrite rules exposed in~\ref{fig:rewrite-rules} are applicable.
+In the ``base'' level, only generalized loop-invariant code motion, as described in~\cite{Luporini}, is applied.
+This means that only a subset of the rewrite rules exposed in~\ref{fig:rewrite-rules} will be considered.
 In the ``aggressive'' level, all of the rewrite rules are applied.
-\item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only only three split factors, namely 1, 2, 4.
+Many other trade-offs, which we do not explore, would be feasibile, however.
+\item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only three split factors, namely 1, 2, 4.
 Also, if the input problem uses mixed function spaces, the iteration space is already split by the Expression Rewriter to avoid computation over zero-columns; in these cases, we do not further apply expression splitting.
@@
 
-...TODO... Problems chosen...
+We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods.
+In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms.
+The Unified Form Language code for these forms is available at~\cite{Code-location}.
 
-...TODO... We don't compare to tensor contraction by transitivity on olegaards work
+We evaluate the effect of three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied.
+In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning).
+This way, we also highlight the advances achieved over our previous work.
 
-\subsection{Results for Forms of Increasing Complexity}
+The values that we report include the cost of local assembly as well as the cost of matrix insertion.
+However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the noise due to any operations that are not part of the element matrix evaluation itself.
+It has been discussed in many places already (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant.
+All codes were executed in the context of the Firedrake framework.
 
-% Quadrature. Tensor. Quadrature <, >, = Tensor
+We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superflous to repeat the same analysis.
+Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode.
+Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.
 
+We vary several aspects of each form, which follows the approach and the notation of~\cite{quadrature1} and~\cite{Francis}
+\begin{itemize}
+\item The polynomial order of basis functions, $q \in \lbrace1, 2, 3, 4\rbrace$
+\item The polynomial order of coefficient (also called pre-multiplying) functions, $p \in \lbrace1, 2, 3, 4\rbrace$
+\item The number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
+\end{itemize}
+On the other hand, other aspects are fixed 
+\begin{itemize}
+\item The space of both basis and coefficient functions is Lagrange
+\item The mesh is three-dimensional, made of tetrahedrons, for a total of 4374 cells
+\end{itemize}
+
+Each figure reported in the following is a grid of plots.
+Each grid has two axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
+The top-left plot will show values for $q=1, p=1$; the plot on its right will be $q=1, p=2$ and so on.
+The diagonal of the grid shows plots for which both basis and coefficient functions originate from the same function space.
+
+A plot reports speed-ups over non-optimized FEniCS-Form-Compiler-generated code.
+There are three groups of bars, each group referring to a particular version of the code (\texttt{ffc, fix, auto}).
+There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.
+Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.
+
+\subsection{Performance of Forms}
 In Figure~\ref{fig:helmholtz}, ...
@@
 \end{figure}
+
+In Figure~\ref{fig:poisson}, ...
+
+\begin{figure}[t]  
+\includegraphics[scale=0.7]{perf-results/poisson}
+\caption{Poisson results.}\label{fig:poisson}
+\end{figure}
+
+In Figure~\ref{fig:hyperelasticity}, ...
+
+\begin{figure}[t]  
+\includegraphics[scale=0.7]{perf-results/hyperelasticity}
+\caption{Hyperelasticity results.}\label{fig:hyperelasticity}
+\end{figure}
+
+\subsection{Details on the Autotuning Process}
+
 % Mass DGEMM utile explicit method etc

### 2014-08-08 TOMS_Paper/assembly.tex
Start writing the perf eval section

@@
 
-We evaluate the effect of three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied.
-In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning).
-This way, we also highlight the advances achieved over our previous work.
+We evaluate the \textit{speed ups} achieved by three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied.
+In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning as described in Section~\ref{sec:autotune}).
+The \texttt{auto} values do not include the autotuner cost, which is commented aside in Section~\ref{sec:auto-analysis}.
+
+%This way, we also highlight the advances achieved over our previous work.
 
@@
 
-Each figure reported in the following is a grid of plots.
+Each figure reported in the following corresponds to one specific problem and must be read as a grid of plots.
 Each grid has two axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
-The top-left plot will show values for $q=1, p=1$; the plot on its right will be $q=1, p=2$ and so on.
+The top-left plot will show values for $[q=1, p=1]$; the plot on its right will be $[q=1, p=2]$ and so on.
 The diagonal of the grid shows plots for which both basis and coefficient functions originate from the same function space.
+Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.
 
@@
 There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.
-Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.
 
 \subsection{Performance of Forms}
-In Figure~\ref{fig:helmholtz}, ...
+The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly.
+We limit ourselves to analyzing the cost of computing element matrices, although all of the techniques presented in this paper are immediately extendible to the evaluation of local vectors.
+As anticipated, in the following we comment speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the non-optimized, FEniCS-Form-Compiler-generated code.
+
+We first comment on results of general applicability.
+We note there is a trend of COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase.
+This is because most of the transformations applied aim at optimizing for arithmetic intensity and SIMD vectorization, which obviously have a strong impact when arrays and iteration spaces are large.
+The corner cases of this phenomenon are indeed $[q=1, p=1]$ and $[q=4, p=4]$.
+We also observe how \texttt{auto}, in almost all scenarios, outperforms all of the other variants.
+In particular, it is not a surprise that \texttt{auto} is faster than \texttt{fix}, since \texttt{fix} is one of the autotuner's tested variants, as explained in Section~\ref{sec:autotune}.
+This proves the quality of the work presented in this paper, which shows significant advances over~\cite{Luporini}.
+The reasons for which \texttt{auto} exceeds both original code and \texttt{ffc} are discussed for each specific problem next.
+Also, details on the ``optimal'' code variant determined by autotuning are given in Section~\ref{sec:auto-analysis}.
 
@@
 
-In Figure~\ref{fig:elasticity}, ...
+The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.
+We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$, as a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization.
+The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$.
+This is due to the fact that if the loop over quadrature points is relatively small, then close-to-peak performance is obtainable through basic expression rewriting and code specialization; in this circumstance, generalized loop-invariant code motion and padding plus data alignment.
+The trend changes dramatically as $nf$ and $q$ increase: a more ample spectrum of transformations must be considered to find the optimal local assembly implementation.
+We will comment on this in details in the next section.
 
@@
 
-In Figure~\ref{fig:poisson}, ...
+Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem.
+This form uses vector-valued spaces for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance.
+The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$.
+The use of inderection arrays allows to phisically eliminate zero-valued columns at code generation time; as a consequence, different tabulated basis functions are merged into a single array.
+Therefore, despite the execution being purely scalar because of indirection arrays, the reduction in arithmetic intensity and register pressure imply improvement in performance.
+Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase.
+
+For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceeds 1$\times$.
+On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.
+
+It is worth noting that speed ups of \texttt{auto} over \texttt{fix} decrease at $q=4$, particularly for low values of $p$.
+As we will discuss in Section~\ref{sec:auto-analysis}, this is because at $q=4$ the vector-register tiling transformation (in combination with loop unroll-and-jam) leads to the highest performance.
+In principle, vector-register tiling can be used in combination to the zero-avoidance technique; however, due to mere technical limitations, this is currently not supported in COFFEE.
+Once solved, we expect much higher speed ups in the $q=4$ regime as well.
 
@@
 
-In Figure~\ref{fig:hyperelasticity}, ...
+In Figure~\ref{fig:poisson}, ...
 
@@
 
+In Figure~\ref{fig:hyperelasticity}, ...
+
 \subsection{Details on the Autotuning Process}
+\label{sec:auto-analysis}
 
-% Mass DGEMM utile explicit method etc
+% selected variants
+% cost of autotuner
+% cached autotuning
 
@@
 We have presented ...
+...TODO...: straightforward extendibility to right-hand side assembly
+% Mass DGEMM utile explicit method etc
 

### 2014-08-08 TOMS_Paper/assembly.tex
Write a first draft of perf eval section

@@
 
+\paragraph{Helmholtz}
 \begin{figure}[t]  
@@
 \end{figure}
-
 The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.
@@
 
+\paragraph{Elasticity}
 \begin{figure}[t]  
@@
 \end{figure}
-
 Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem.
@@
 
+% TODO: mention aggressive expression rewriting somewhere
+
+\paragraph{Poisson}
 \begin{figure}[t]  
@@
 
-In Figure~\ref{fig:poisson}, ...
+In Figure~\ref{fig:poisson} we report speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the original code for the Poisson form.
+We note that, as a general trend, \texttt{ffc} exhibits drops in performance as $nf$ increases, notably when $nf=3$, for any values of $q$ and $p$.
+This is a consequence of the inherent complexity of the generated code.
+The way \texttt{ffc} performs loop-invariant code motion leads to the pre-computation of integration-dependent terms at the level of the integration loop, which are characterized by higher arithmetic intensity and redundant computation as $nf$ increases.
+Moreover, the absence of vectorization is another limiting factor.
 
+The \texttt{auto} variant generally shows the best performance.
+Significant improvements over \texttt{fix} are also achieved, notably as $q$, $p$ and $nf$ increase.
+As clarified in the next section, this is always due to a more aggressive expression rewriting in combination with the technique to avoid computation over the zero-valued columns induced by vector-valued function spaces.
+
+\paragraph{Hyperelasticity}
 \begin{figure}[t]  
@@
 
-In Figure~\ref{fig:hyperelasticity}, ...
+Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 3$ could not be executed because of FEniCS-Form-Compiler's technical limitations.
+
+For \texttt{auto}, massive speed ups for $q \geq 2$ must mainly be ascribed to aggressive and successful expression writing.
+Hyperelasticity problems are really compute-intensive, with thousands of operations being performed, so reductions in redundant and useless computation are crucial.
+Complex forms like hyperelasticity would benefit from further ``specialized'' optimizations: for example, it is a known technical limitation of COFFEE that, in some circumstances, less temporaries could (should) be generated and that hoisted code could (should) be suitably distributed over different loops to minimize register pressure (e.g.
+COFFEE could carefully apply loop fission for obtaining significantly better register usage).
+We expect to obtain considerably faster code once such optimizations will be incorporated.
+
+In the regime $q \geq 2$ and $nf=1$, peformance improvements are less pronounced moving from $p=1$ to $p=2$, although still significant; in particular, we notice a drop at $p=2$, followed by a raise up to $p=4$.
+It is worth observing that this effect is common to all set of optimizations.
+The hypothesis is that this is due to the way coefficient functions are evaluated at quadrature points (identical in all configurations), which cannot be easily vectorized unless a change in storage layout and loops order is implemented in the code (abstract syntax tree) generator on top of COFFEE.
 
@@
 
-% selected variants
+% selected variants (tables)
 % cost of autotuner
 % cached autotuning
+% show the difference of ``aggressive'' vs ``base'' expression rewriting
 

### 2014-08-11 TOMS_Paper/assembly.tex
Add autotuner details subsection

@@
 \usepackage{tabulary}
+\newcommand{\specialcell}[2][c]{%
+  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
 \usepackage{lineno}
@@
 Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression.
-In Section~\ref{sec:rewriting-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.
+In Section~\ref{sec:rewrite-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.
 
@@
 \subsection{Objectives of the Expression Rewriter}
-\label{sec:rewriting-rules}
+\label{sec:expr-rewriter-obj}
 
@@
 \subsection{Padding and Data Alignment Revisited}
+\label{sec:padding}
 Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided.
@@
 Statements in the bodies of the surrounding loops (e.g.
-coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:rewriting-rules}.
+coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:expr-rewriter-obj}.
 This renders perfect the loop nest; that is, there is no intervening code among the various loops.
@@
 %This way, we also highlight the advances achieved over our previous work. 
+%...TODO...: dire che i tempi sono la media di 3 runs
 
@@
 
-We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superflous to repeat the same analysis.
+We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superfluous to repeat the same analysis.
 Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode.
@@
 \subsection{Performance of Forms}
+\label{sec:perf-results-forms}
 The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly.
@@
 
+We first comment on the overhead of the autotuning process.
+In the context of Firedrake, the framework in which COFFEE is integrated, the autotuner is executed at run-time, once the local assembly kernels are provided by the FEniCS Form Compiler.
+Autotuning, therefore, introduces an overhead in the application execution time.
+For a given problem instance, such overhead originates from four sources: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution.
+Of these four points, we note that: the cost of 4) is relatively small, because each variant's execution time is bound by an empirically-found value (e.g.
+some milliseconds).
+The cost of all four points is constrained by our heuristics to prune the search space, as described in Section~\ref{sec:autotune}.
+Moreover, for a given form and a given discretization, the autotuner needs to be executed only once, since its output is saved and reused for later assemblies.
+This implies that if the assembly occurs in a time stepping loop, or the same form is executed on a different mesh, or known quantities of the input problem are changed, then the assembly cost is rapidly amortized.
+Premised that, the most important thing remains that when working with real unstructured meshes - which are likely to be composed of millions of elements, leading to long-lasting assembly phases - the autotuner overhead practically becomes completely negligible.
+In our experience, and in particular in the four examined problems, the autotuning process lasted less than a minute in the majority of cases.
+Most often, it took less than thirty seconds.
+Rarely it needed more than a minute, specifically in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$.
+This analysis certifies that the autotuner overhead is definitely sustainable in real-world applications.
+
+In this paper and in~\cite{Luporini}, we have repeatedly claimed that different forms (and different discretizations) require distinct sets of transformations to reach close-to-peak performance.
+To demonstrate this, we now report details about the output of the autotuning process.
+We show that for the four examined forms - more specifically, for the 224 problem instances $[form, q, p, nf]$ illustrated in the previous figures - a plethora of optimization strategies have been selected by the autotuner.
+We also complement and strengthen our claim by showing the autotuner's output for two additional forms whose performance results, for brevity, were not shown in Section~\ref{sec:perf-results-forms}: a Mass and a Mixed Poisson problems.
+
+Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner.
+To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$.
+Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible.
+If the value of the column is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one.
+Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}.
+The \texttt{split} column refers to the expression splitting transformation (~\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}).
+The \texttt{unroll} column indicates the application of explicit unrolling.
+Other columns are of obvious meaning.
+Values in the various columns illustrate the number of problem instances out of the total (column \texttt{mumber of variants}) in which an optimization was activated.
+Obviously, more transformations are typically used in combination in a same problem.
+
+%In the table, we implicitly assume that padding and data alignment (Section~\ref{sec:padding}), as well as precomputation of integration-dependent terms (Section~\ref{sec:expr-rewriter-obj}) have always been applied.
+
+\newcommand{\autotuneroutput}{
+\tiny
+\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
+\cline{3-10}
+\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
+\cline{1-10}
+problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & BLAS \\
+\cline{1-10}
+\texttt{helmholtz} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{mass} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{elasticity} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{poisson} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{mixed poisson} & 40 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{hyperelasticity} & 32 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\cline{1-10}
+\end{tabulary}
+}
+
+\begin{table*}[t]
+\tbl{Autotuner output.}{ \scriptsize \autotuneroutput }
+\label{table:autotuner-output}
+\end{table*}
+
 % selected variants (tables)
-% cost of autotuner
-% cached autotuning
 % show the difference of ``aggressive'' vs ``base'' expression rewriting

### 2014-08-11 TOMS_Paper/assembly.tex
Add some entries in autotuner table, still incomplete
Minor tweaks

@@
 These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution.
-Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more expensive, covering a significant fraction of overall computation run-time.
+Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more computationally expensive, covering a significant fraction of the overall computation run-time.
 
@@
 In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
-The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu}.
+The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.
 
@@
 \cline{1-10}
-\texttt{helmholtz} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
-\texttt{mass} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
-\texttt{elasticity} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
-\texttt{poisson} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
-\texttt{mixed poisson} & 40 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
-\texttt{hyperelasticity} & 32 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 8 & 17 & 15 & 0 \\
+\texttt{mass} & 64 & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{elasticity} & 64 & aggressive & 39 & 0 & all & 0 & 11 & 0 & 2 \\
+\texttt{poisson} & 64 & aggressive & 50 & 0 & all & X & X & 1 & 0 \\
+\texttt{mixed poisson} & X & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{hyperelasticity} & 32 & aggressive & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
 \cline{1-10}

### 2014-08-11 TOMS_Paper/assembly.tex
Rewrite some introduction

@@
 
-Achieving high performance implementations is, however, non-trivial.
+Producing high performance implementations is, however, non-trivial.
 The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process.
-Also, traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions.
-This has led to development of a number of higher-level approaches to optimize local assembly kernels.
+Traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions.
+Polyhedral-model-based source-to-source compilers, for instance~\cite{PLUTO}, mainly apply aggressive loop optimizations, such as tiling, but these are not particularly helpful in our context.
+This lack of suitable optimizing tools has led to the development of a number of higher-level approaches to maximize the performance of local assembly kernels.
 In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''.
-\cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of the problem.
-In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
-The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.
+\cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration.
+In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct set of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
+The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.
 
-Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework.
-While clearly separating the mathematical domain, which remains captured in the higher-level form compiler, from the optimization process, COFFEE also aims to be platform-agnostic.
-The code transformations occur on an intermediate representation of the assembly code, which is ultimately translated into platform-specific code.
-Domain knowledge is exploited in two ways: for simplifying the implementation of a broad range of code transformations, and, obviously, to make them extremely effective.
+Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework~\cite{Luporini}.
+Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic.
+The code transformations occur on an intermediate representation of the assembly operation, which is ultimately translated into platform-specific code.
+Domain knowledge is exploited in two ways: for simplifying the implementation of code transformations and to make them extremely effective.
 Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input.
@@
 
-In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation.
-We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization.
-The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become sub-steps of code specialization; on the other hand, generalized loop-invariant code motion is a step of the expression rewriting process.
-More importantly, we complement and generalize our previous work with the following contributions.
+In this paper, we build on our previous work~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation.
+We argue that for complex, realistic forms, peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to ensure effective register utilization and SIMD vectorization.
+The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become steps of code specialization, whereas generalized loop-invariant code motion is performed during the expression rewriting stage.
+We complement and generalize our previous work with the following contributions.
 
-Expression rewriting is based on a formal set of rewrite rules.
-Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of arithmetic operators to expose ``hidden'' loop-invariant sub-expressions.
-Secondly, we show how to make use of domain knowledge to avoid computation over zero-valued regions in vector-valued basis functions arrays, while preserving code vectorizability.
-These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable by using FEniCS' built-in optimizations, presented in~\cite{quadrature1}.
+Expression rewriting aims at reducing the computational intensity of local assembly kernels by rescheduling arithmetic operations based on a set of rewrite rules.
+Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage.
+While rewriting an assembly expression, domain knowledge is used in several flavours, for example to avoid redundant computation.
+Secondly, we propose an algorithm that, relying on symbolic execution, allows to restructure the assembly expression so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability.
+These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.
 
@@
 On top of the work in~\cite{Luporini}, we provide a number of contributions.
-Firstly, we show the benefit of vector-expansion to achieve SIMD vectorization of otherwise scalar code.
-This is particularly useful in complex forms, like those based on hyperelasticity.
-Secondly, we answer an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix multiplies.
+Firstly, we revisit the padding and data alignment transformation to account for the new code structures produced by expression rewriting.
+This is of fundamental importance to achieve effective SIMD vectorization.
+Secondly, we address an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix-matrix multiply routines.
 BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods.
-However, we will show that in corner, yet important cases, especially in forms characterized by pre-multiplying functions and relatively high-order function spaces, a BLAS-based execution strategy can be a successful.
-Finally, we introduce a model-driven, dynamic autotuner that automatically and transparently compose the set of code transformations that are likely to maximize the performance of a given problem.
-The main challenge with the autotuner is to maintain, for any possible problem, the search space reasonably small, although comprising the most effective code variants, so that the overhead, which impacts the run-time, is negligible.
+However, we will show that in corner, yet meaningful cases, especially in forms characterized by the presence of pre-multiplying functions and relatively high-order function spaces, a BLAS-based implementation can be successful.
+Finally, we introduce a model-driven, dynamic autotuner that transparently evaluates multiple sets of code transformations to determine the best optimization strategy for a given problem.
+The main challenge here is to build, for a generic form, a reasonably small search space that comprises most of the effective code variants.
 
-Expression rewriting and code specialization have been implemented in COFFEE and are fully operating.
-Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation in a number of forms of increasing complexity, including problems based on hyperelasticity operators.
+Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework.
+Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on hyperelasticity.
 We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions.
-To clearly distinguish the improvement achieved by this work, we will compare four sets of code variants, for each problem instance: 1) unoptimized code, i.e.
+To clearly distinguish the improvement achieved by this work, we will compare, for each test case, four sets of code variants: 1) unoptimized code, i.e.
 a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e.
 the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper.
-Notable performance improvements of 4) over 1), 2) and 3) are reported and detailed.
+Notable performance improvements of 4) over 1), 2) and 3) are reported and discussed.
 

### 2014-08-11 TOMS_Paper/assembly.tex
Some other minor fixes

@@
 
-Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework~\cite{Luporini}.
+Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Finit Element local assEmbly.}, fully integrated with the Firedrake framework~\cite{Luporini}.
 Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic.
@@
 Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework.
-Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on hyperelasticity.
+Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex hyperelasticity models.
 We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions.
@@
 
-\subsection{Avoiding Iteration on Zero-blocks by Symbolic Execution}
+\subsection{Avoiding Iteration on Zero-valued Blocks by Symbolic Execution}
 \label{sec:zeros}
@@
 Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}.
-The \texttt{split} column refers to the expression splitting transformation (~\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}).
+The \texttt{split} column refers to the expression splitting transformation (\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}).
 The \texttt{unroll} column indicates the application of explicit unrolling.

### 2014-08-12 TOMS_Paper/assembly.tex
Several fixes. Add some captions

@@
 \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration.
-In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct set of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
+In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct sets of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
 The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.
@@
 Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage.
-While rewriting an assembly expression, domain knowledge is used in several flavours, for example to avoid redundant computation.
-Secondly, we propose an algorithm that, relying on symbolic execution, allows to restructure the assembly expression so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability.
+While rewriting an assembly expression, domain knowledge is used in several ways, for example to avoid redundant computation.
+Secondly, we propose an algorithm that, relying on symbolic execution, restructures the code so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability.
 These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.
@@
 
-Without loss of generality, we illustrate local assembly in a real example, the evaluation of the local assembly (or, equivalently, element) matrix for a Laplacian operator.
-Consider the weighted Laplace equation
+Without loss of generality, we illustrate local assembly in a concrete example; that is, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
 \begin{equation}
@@
 
-In the next sections, we will often refer to the (local) element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} of the variational problem.
+In the next sections, we will often refer to the local element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} deriving from the weak variational problem.
 
@@
 If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality.
-In this paper, we tackle these three points building on our previous work (\cite{Luporini}).
+In this paper, we tackle these three points building on our previous work (\cite{Luporini}), in which we presented COFFEE, a mature, platform-independent compiler capable of optimizing local assembly code.
 
@@
 \includegraphics[scale=0.6]{pictures/coffee-scheme}
-\caption{Outline of COFFEE}\label{fig:coffee-outline}
+\caption{Outline of COFFEE.
+The input is either a string of C code or an abstract syntax tree representation.
+The optimization process goes through two main steps: \textit{expression rewriting} and \textit{code specialization}.
+The autotuner, incorporated with the Code Specializer, use heuristics to select the set of code variants that, for a given variational form, are likely to reach high performance.}\label{fig:coffee-outline}
 \end{center}
@@
 
-COFFEE is a mature, platform-independent compiler capable of optimizing local assembly code.
-Its high-level structure can be outlined as in Figure~\ref{fig:coffee-outline}.
-The input is either suitably-annoted C code, explicitly provided by the user, or a decorated abstract syntax tree (henceforth AST) representation of an assembly kernel, automatically generated by the higher level in the stack.
-We have integrated COFFEE with the Firedrake framework; here, the input to COFFEE is originated by the FEniCS Form Compiler in the form of an AST.
+The high-level structure of COFFEE is outlined in Figure~\ref{fig:coffee-outline}.
+The input is either C code, explicitly provided by the user as a string, or an abstract syntax tree (henceforth AST) representation automatically generated at the higher level of abstraction.
+We have integrated COFFEE with the Firedrake framework; here, the input is originated by the FEniCS Form Compiler~\cite{FFC-Compiler} in the form of an AST.
 Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE.
-This is used to introduce more effective, specialized optimizations, as well as to simplify implementation.
+This is used to introduce more effective, specialized optimizations, as well as to simplify the implementation.
 
@@
 The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality.
-The Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer.
-As described later, we have invested some effort to ensure that the code transformations applied in the expression rewriting stage do not break any code specialization opportunity.
-This for example would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.
+As shown in the Figure, the Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer.
+As described later, we have invested some effort to ensure that the code transformations applied at expression rewriting time would not break any code specialization opportunity.
+This, for example, would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.
 
 To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work.
-Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to~\cite{Luporini} (as well as to FEniCS' built-in optimizations).
+Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to both~\cite{Luporini} and~\cite{quadrature1}.
 
 In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation.
-We have also shown how different subset of optimizations are needed to achieve close-to-peak performance in different problems.
+We have also shown how different subsets of optimizations are needed to achieve close-to-peak performance in different variational forms.
 In particular, we can summarize our study as follows
@@
 Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest.
-As briefly described in Section~\ref{sec:expr-rewriting}, we work around this limitation, while also achieving vectorization of invariant code.
+As reiterated in Section~\ref{sec:expr-rewriting}, we work around this limitation by performing source-level code motion, while also achieving vectorization of invariant code.
 \item \emph{Padding and data alignment}.
-The small size of the loop nest require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code.
+The small size of the loop nest (integration, test, and trial functions loops) require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code.
 Data alignment can be enforced as a consequence of padding.
@@
 
-In Figure~\ref{fig:coffee-outline}, we show where these transformations are logically applied in COFFEE; also, the contributions of this work are highlighted.
-In the next two sections, we describe the new functionalities of, respectively, the Expression Rewriter and the Code Specializer.
+In Figure~\ref{fig:coffee-outline}, we show where these transformations are applied in the COFFEE pipeline; also, the novel contributions of this work are highlighted in red.
+In the next two sections, we describe the new functionalities of Expression Rewriter and Code Specializer, respectively.
 
@@
 \label{sec:expr-rewriting}
-As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of an assembly expression.
-The Expression Rewriter (henceforth ER) that we have designed and implemented in COFFEE enhances this technique by making two steps forward, which allow more redundant computation to be avoided.
+Loop-invariant code motion is the key to reduce the computational intensity of an assembly expression. The Expression Rewriter still relies on this technique, although it makes several steps forward.
 
-Firstly, exploiting arithmetic operations properties like associativity, distributivity, and commutativity, it manipulates the original expression to expose more opportunities to the code hoister.
-There are many possibilities of rewriting an expression, and the search space can quickly become too big.
-Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression.
-In Section~\ref{sec:rewrite-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.
+Firstly, exploiting operators properties like associativity, distributivity, and commutativity, it manipulates assembly expressions to expose more code hoisting and SIMD vectorization opportunities, as well as to balance the register pressure in the various levels of the loop nest.
+There are many possibilities of rewriting an expression, so the search space can be quite large.
+We address this problem by defining a simple yet systematic way of rewriting an expression based on a set of formal rewrite rules, which are meticulously applied by COFFEE.
 
-Secondly, the ER re-structures the loop nest so as to eliminate arithmetic operations over array columns that are statically known to be zero-valued.
-Zero columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements.
-A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with it is that by using indirection arrays in the generated code, it breaks many of the optimizations that can be applied at the Code Specializer level, including SIMD vectorization.
-In Section~\ref{sec:zeros}, we show a novel approach to avoiding computation on zeros based on symbolic execution.
+Secondly, the Expression Rewriter re-structures iteration spaces so as to avoid arithmetic operations over tabulated basis functions columns that are statically known to be zero-valued.
+Zero-valued columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements.
+A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with that is the use of indirection arrays in the generated code, which break many of the optimizations applicable at code specialization time, including SIMD vectorization.
+In Section~\ref{sec:zeros}, we present a novel approach based on symbolic execution.
 
@@
 
-\subsection{Objectives of the Expression Rewriter}
+\subsection{Motivations and Objectives}
 \label{sec:expr-rewriter-obj}
@@
 \centering     
-\subfigure[Original code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
+\subfigure[Original (simplified) code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
 ~~~~~~~~~~~~~~~~
-\subfigure[Invariant code]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
+\subfigure[The code after that generalized loop-invariant code motion has been applied]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
 ~\\
 ~\\
-\subfigure[Factorized code]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
+\subfigure[Once terms are factorized, further code hoisting opportunities arise.
+Here, \texttt{T3} stores the precomputation of an invariant sub-expression exposed by factorization]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subfigure[Scalar-expanded code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
-\caption{Original, invariant, factorized, and then scalar-expanded code.}
+\subfigure[Hoisting and scalar-expansion of integration-dependent terms allow to achieve full SIMD vectorization of the assembly code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
+\caption{Applying a sequence of transformations to an excerpt from a Burgers' local assembly kernel, suitably simplified to improve readability.}
 \end{figure}
 
-Consider the element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem.
+Consider the local element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem.
 The assembly expression, produced by the FEniCS Form Compiler, has been deliberately simplified, and code details have been omitted for brevity and readability.
@@
 By examining again the code in Figure~\ref{fig:invariant-code}, we notice that the basis function array \texttt{A} iterating along the \texttt{[i,j]} loops appears twice in the expression.
-By expanding the products in which \texttt{A} is accessed and applying sum commutativity, we can factorize the expression.
+By expanding the products in which \texttt{A} is accessed and by applying sum commutativity, terms can be factorized.
 This has two effects: firstly, it reduces the number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{A[i][k]/c+T2[k]*f} invariant with respect to loop \texttt{j}.
@@
 For example, the \texttt{f0*a} and \texttt{f1*b} products in Figure~\ref{fig:factorized-code} depend on the loop along quadrature points; these operations are performed in a non-vectorized way at every \texttt{i} iteration.
-This is not an issue in our running example, where scalar computation represents a small percentage of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}).
+This is not a big issue in our running example, in which the scalar computation represents a small fraction of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}).
 In such forms, the amount of computation independent of both test and trial functions loops is so large that it has a significant impact on the run-time, despite being executed only $O(I)$ times (with $I$ number of quadrature points).
-We have therefore implemented an algorithm to scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.
+We have therefore implemented an algorithm to move and scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.
 
-The third point concerns the register pressure induced by the assembly expression.
+The third motivation for expression rewriting concerns the register pressure in the innermost loop.
 Once the code has been optimized for arithmetic intensity, it is important to think about how the transformations impacted register allocation.
@@
 Each value appearing in the expression is loaded and kept in a register as long as possible.
-In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every $i$ iteration.
-This implies that at every iteration of the \texttt{jk} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops.
+In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every \texttt{i} iteration.
+This implies that at every iteration of the \texttt{[j,k]} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops.
 In more complicated expressions, the percentage of registers destined to store such constant terms can be even higher.
@@
 \centering
-\subfigure[Expandable code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
+\subfigure[Original (simplified) code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
 ~\\
-\subfigure[Expanded 1 code]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
+\subfigure[The code after expansion of \texttt{det*W[i]}]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
 ~~
-\subfigure[Expanded 2 code]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
-\caption{Expanding code.}\label{fig:expanded-code}
+\subfigure[The code after expansion of \texttt{g}. Note the need to introduce a new temporary array]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
+\caption{Expansion of terms to improve register pressure in a local assembly kernel}\label{fig:expanded-code}
 \end{figure}
@@
 Consider a variant of the Burgers local assembly kernel, shown in Figure~\ref{fig:toexpand-code}.
-This is still a representative, simplified example.
+This is again a representative, simplified example.
 We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of the invariant sub-expression stored in \texttt{T1}, resulting in code as in Figure~\ref{fig:expanded-1-code}.
-Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since the same array is accessed in \texttt{(T1[j]*A[i][k])}.
+Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since \texttt{T1} is also accessed in the \texttt{T1[j]*A[i][k]} sub-expression.
 The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}.
 Generalizing, this is a problem of data dependencies: to solve it, we use a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}.
-The dependency graph is initially empty, and it is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms.
+The dependency graph is initially empty, and is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms.
 The dependency graph is then queried to understand when expansion can be performed without resorting to new temporary values.
@@
 Our goal is to establish a portable, unified, platform-independent, and systematic way of reducing the computational strength of an expression exploiting the intuitions described in the previous section.
-This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems.
-In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform scalar-expansion to achieve full vectorization of the assembly code.
+This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems, such as COFFEE.
+In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform 4) scalar-expansion of integration-dependent terms to achieve full vectorization of the assembly code.
 
@@
 
-In the following, a generic (sub-)expression is represented with roman letter \texttt{a, b, ...}; constant terms are considered a special case, so greek letters \texttt{$\alpha$, $\beta$} are used instead.
+In the following, a generic (sub-)expression is represented with Roman letters \texttt{a, b, ...}; constant terms are considered a special case, so Greek letters \texttt{$\alpha$, $\beta$, ...} are used instead.
 The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression.
 We will refer to $i_0$ as the outermost enclosing loop.
-The notation $a_i$, therefore, indicates that the expression $a$ assume distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.
+The notation $a_i$, therefore, indicates that the expression $a$ assumes distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.
 
-Rewrite rules for expression rewriting are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
-The Expression Rewriter applies the rules while performing a depth-first traversal of the assembly expression tree.
-Given an arithmetic operation between two sub-expressions (i.e.
-a node in the expression tree), we first need to find an applicable rule.
-There cannot be ambiguities: only one rule can be matched.
-If the preconditions of the rule are satisfied, the corresponding transformation is performed; otherwise, no rewriting is performed, and the traversal proceeds.
-As examples, it is possible to instantiate the rules in the codes shown in Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized codes in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} are obtained, respectively.
-
-\begin{figure}
+\begin{figure}[t]
 \small
@@
 \end{spacing}
-\caption{Rewrite rules.}\label{fig:rewrite-rules}
+\caption{Rewrite rules driving the rewriting process of an assembly expression.}\label{fig:rewrite-rules}
 \end{figure}
+
+Rewrite rules for expression rewriting are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
+The Expression Rewriter applies the rules while performing a depth-first traversal of the assembly expression tree.
+Given an arithmetic operation between two sub-expressions (i.e.
+a node in the expression tree), we first need to find an applicable rule.
+There cannot be ambiguities: only one rule can be matched.
+If the preconditions of the rule are satisfied, the corresponding transformation is performed; otherwise, no rewriting is performed, and the traversal proceeds.
+As examples, it is possible to instantiate the rules in the codes shown in Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized codes in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} are obtained, respectively.
 
@@
 In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms.
-The Unified Form Language code for these forms is available at~\cite{Code-location}.
+The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{Code-location}.
 

### 2014-08-12 TOMS_Paper/assembly.tex
Some other tweaks + captions

@@
 \centering     
-\subfigure[Factorized, showing zeros, code]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
+\subfigure[Original (simplified) code.
+Note the annotation over the definition of the tabulated basis function \texttt{D}, which is used to identify the presence of zero-valued columns]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
 ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subfigure[Factorized, skipping zeros, code]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
-\caption{Without and with skipping zeros}\label{fig:skip-code}
+\subfigure[The code after symbolic execution took place]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
+\caption{Simplified excerpt of local assembly code from a Burgers form using vector-valued basis functions, before and after symbolic execution is performed to rewrite the iteration space}\label{fig:skip-code}
 \end{figure}
 
-Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}.
+Consider Figure~\ref{fig:withzeros-code}, an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}.
 The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh.
@@
 The two key observations are that 1) the number of zero-valued columns caused by using vector function spaces is, often, much larger then that due to derivatives, and 2) such columns are contiguous in memory.
-Based on this, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.
+Based on this observation, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.
 
@@
 
-We achieve our goal by symbolic execution.
-The Expression Rewriter expects some indication in the input exposing the location of the zero-valued columns due to mixed (vector) function spaces, in each tabulated basis function.
+We achieve our goal by means of symbolic execution.
+The Expression Rewriter expects some indication about the location of the zero-valued columns induced by mixed (vector) function spaces, for each tabulated basis function.
 This indication comes either in the form of code annotation, if the input to COFFEE were provided as pure C (as shown in Figure~\ref{fig:withzeros-code}), or by suitably decorating basis functions nodes, if the input were an abstract syntax tree.
@@
 For example, consider the assignment \texttt{T2[r] = d*D[i][k]+e*E[i][k]} in Figure~\ref{fig:withzeros-code}.
-Array \texttt{D} has non-zero-valued columns in positions $NZ_D = [3,5]$; we also assume array \texttt{E} has non-zero-valued columns in positions $NZ_E = [0,2]$.
+Array \texttt{D} has non-zero-valued columns in the range $NZ_D=[3,5]$; we also assume array \texttt{E} has non-zero-valued columns in the range $NZ_E=[0,2]$.
 Multiplications by scalar values do not affect the propagation of non-zero-valued columns.
-On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in positions $NZ_D \cup NZ_E = [0,5]$.
+On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in the range $NZ_E \| NZ_D=[0,5]$.
 Eventually, exploiting the $NZ$ information computed and associated with each identifier, we split the original assembly expression into multiple sets of sub-expressions, each set characterized by the same range of non-zero-valued columns.
-In our example, assuming that $NZ_{T1} = [3,5]$ and $NZ_A = [3,5]$, there are two of such sets, this leads to the generation of two distinct iteration spaces, one for each set, as in Figure~\ref{fig:withzeros-skipped-code}.
+In our example, assuming that $NZ_{T1}=[3,5]$ and $NZ_A=[3,5]$, there are two of such sets, which leads to the generation of two distinct iteration spaces (one for each set), as in Figure~\ref{fig:withzeros-skipped-code}.
 
@@
 As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying.
-In this paper, we enrich this set of transformations as explained in the following sections.
+In this paper, we revisit and enrich this set of transformations as explained in the following sections.
 
@@
 Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided.
-We recall effective SIMD vectorization can be achieved only if the innermost loop size is a multiple of the vector register length.
-In the case of an AVX instruction set, for example, we want the size of innermost loops to be multiples of 4 (vector registers can fit up to four double-precision floats).
-Moreover, efficient loads and stores can be issued only if all of the involved arrays' base addresses are multiple of the vector register length.
+We recall effective SIMD (auto-)vectorization can be achieved only if the innermost loop size is a multiple of the vector register length, in which case the compiler needs not to introduce a remainder scalar loop.
+In the case of an AVX instruction set, for example, we want to round the size of the innermost loops to the closest multiple of 4 (AVX registers can fit up to four double-precision floats).
+This can be done provided that arrays are padded, if necessary.
+Moreover, loads and stores instructions are efficient only if their addresses are suitably aligned to cache boundaries; this is obtainable by enforcing the base address of each padded array to be a multiple of the vector length.
 
 Consider again the code in Figure~\ref{fig:withzeros-skipped-code}.
-The arrays in the loop nest $[j1,k1]$ can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region $M[0:3][6:8]$ will be discarded.
+The arrays in the loop nest \texttt{[j1,k1]} can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region \texttt{M[0:3][6:8]} will be discarded.
 Then, by explicitly aligning arrays and using suitable pragmas (e.g.
-\texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained.
+\texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained for this loop nest.
 
-There are some complications in the cose of loops $[j0,k0]$.
-Here, increasing the loop bound to 4, despite being still safe, has no effect: for instance, the starting addresses of load instructions are \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned.
-One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$, which would mean losing the effect of expression rewriting.
+There are some complications in the case of loops \texttt{[j0,k0]}.
+Here, increasing the loop bound to 4 is still safe, assuming that both \texttt{T1} and \texttt{A} are padded with zero-valued entries, but it has no effect: the starting addresses of the load instructions would be \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned.
+One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$.
+However, this would imply losing (partially in general, totally for this loop nest) the effect of the zero-avoidance transformation.
 Another possibility is to attain to non-aligned accesses.
@@
 
-Note that in some cases it is not even safe to increase loop bounds, since extra iterations would access non-zero entries in the local element matrix $M$.
-In such a situation, there may be no possibility of restoring data alignment at all.
+Note that in some circumstances the previous solution cannot be applied, since extra iterations might end up accessing non-zero entries in the local element matrix \texttt{M}.
+In such a situation, there is no possibility of recovering data alignment.
 
@@
 We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM routines, which are usually tuned for large arrays.
-The study was conducted only in the case of a specific Helmholtz problem.
+The study was conducted only in the case of a specific Helmholtz form.
 There are scenarios, however, in which tabulated basis functions sizes grow up to a point for which turning to BLAS may actually lead to better performance.
-For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of tabulated basis functions.
+For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of basis functions tabulated at quadrature points.
 To explore these scenarios, we have developed an algorithm that reduces any assembly expression evaluation to a sequence of DGEMM calls.
 
-Here, we informally provide the main steps of the algorithm.
+We informally provide the main steps of the algorithm.
 By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions.
@@
 coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:expr-rewriter-obj}.
-This renders perfect the loop nest; that is, there is no intervening code among the various loops.
+This renders the loop nest perfect; that is, there is no intervening code among the various loops.
 The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
@@
 
-In~\cite{Luporini}, we proposed a simple cost model to select, for a given a problem, the optimal combination of transformations.
+In~\cite{Luporini}, we proposed a simple cost model to drive the optimization process.
 In this paper, we have added a significant number of options to the set of possible optimizations, so the selection problem is now far more challenging.
-The sole Expression Rewriter, for instance, could generate many possible code variants by applying rewrite rules to different extents.
+The sole Expression Rewriter, for instance, can generate a wide variety of implementations by just applying rewrite rules to different extents.
 We found difficult extending the cost model to effectively cover such a large search space.
 
-We tackle the optimization selection problem by compiler autotuning.
-Not only does it allow to determine the best combination of transformations out of the set presented so far, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying different unroll factors and loop permutations.
-By leveraging the cost model defined in our previous study, domain-awareness, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementation.
+We tackle the optimizations selection problem by compiler autotuning.
+Not only does it allow to determine the best combination of transformations, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying different unroll factors and loop permutations.
+By leveraging the cost model defined in our previous study, domain knowledge, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementations in all of the forms we have evaluated.
 In particular, our autotuner usually requires order of seconds to determine the fastest kernel implementation, a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain up to trillions of elements (e.g.
@@
 
-COFFEE analyzes the input problem and decides what variants it is worth testing, as described later.
+COFFEE analyzes the input problem and decides what variants it is worth testing through autotuning, as described later.
 Each variant is obtained by requesting specific transformations to the Expression Rewriter and the Code Specializer.
@@
 \item On the other hand, if the polynomial order is low, i.e. when the loop nest is small, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
-\item We select two levels of expression rewriting.
+\item We specifically select two levels of expression rewriting.
 In the ``base'' level, only generalized loop-invariant code motion, as described in~\cite{Luporini}, is applied.
@@
 \item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only three split factors, namely 1, 2, 4.
-Also, if the input problem uses mixed function spaces, the iteration space is already split by the Expression Rewriter to avoid computation over zero-columns; in these cases, we do not further apply expression splitting.
+Also, if the input problem uses mixed function spaces, the iteration space is already split at expression rewriting time to avoid computation over zero-valued columns; in these cases, we do not further apply expression splitting.
 \item Based on the cost model, the padding and data alignment optimization is always applied. 
@@
 
+All of the previous points contribute to minimize the overhead of the autotuning process. We will discuss the actual cost of autotuning in Section~\ref{sec:auto-analysis}.
 %COFFEE allows the user to abstract from finding the optimal optimization strategy for a given problem by resorting to its model-driven, dynamic autotuner.
@@
 In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms.
-The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{Code-location}.
+The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{ufl-code}.
 
@@
 The values that we report include the cost of local assembly as well as the cost of matrix insertion.
-However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the noise due to any operations that are not part of the element matrix evaluation itself.
-It has been discussed in many places already (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant.
+However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the ``noise'' due to any operations that are not part of the element matrix evaluation itself.
+However, it has been reiterated over and over (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant.
 All codes were executed in the context of the Firedrake framework.
@@
 We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superfluous to repeat the same analysis.
-Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode.
+Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode due to the inherent structure of the code.
 Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.
@@
 \item The polynomial order of basis functions, $q \in \lbrace1, 2, 3, 4\rbrace$
-\item The polynomial order of coefficient (also called pre-multiplying) functions, $p \in \lbrace1, 2, 3, 4\rbrace$
+\item The polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
 \item The number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
@@
 
-Each figure reported in the following corresponds to one specific problem and must be read as a grid of plots.
-Each grid has two axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
-The top-left plot will show values for $[q=1, p=1]$; the plot on its right will be $[q=1, p=2]$ and so on.
-The diagonal of the grid shows plots for which both basis and coefficient functions originate from the same function space.
+Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will be deeply commented in the next section, must be read as ``plots, or grids, of plots''.
+Each grid (figure) has two logical axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
+The top-left plot in a grid shows speed ups for $[q=1, p=1]$; the plot on its right does the same for $[q=1, p=2]$, and so on.
+The diagonal of the grid shows plots for which basis and coefficient functions have same polynomial order, that is $q=p$.
 Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.
@@
 We first comment on results of general applicability.
-We note there is a trend of COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase.
+By looking at the various figures, we note there is a trend in COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase.
 This is because most of the transformations applied aim at optimizing for arithmetic intensity and SIMD vectorization, which obviously have a strong impact when arrays and iteration spaces are large.
@@
 The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.
-We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$, as a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization.
+We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$.
+This is a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization.
 The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$.
@@
 The trend changes dramatically as $nf$ and $q$ increase: a more ample spectrum of transformations must be considered to find the optimal local assembly implementation.
-We will comment on this in details in the next section.
+We will provide details about the selected transformations in the next section.
 
@@
 Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem.
-This form uses vector-valued spaces for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance.
+This form uses a vector-valued space for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance.
 The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$.
@@
 
-For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceeds 1$\times$.
+For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceed 1$\times$.
 On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.
@@
 Significant improvements over \texttt{fix} are also achieved, notably as $q$, $p$ and $nf$ increase.
-As clarified in the next section, this is always due to a more aggressive expression rewriting in combination with the technique to avoid computation over the zero-valued columns induced by vector-valued function spaces.
+As clarified in the next section, this is always due to a more aggressive expression rewriting in combination with the zero-avoidance technique.
 
@@
 
-Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 3$ could not be executed because of FEniCS-Form-Compiler's technical limitations.
+Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 2$ could not be executed because of FEniCS-Form-Compiler's technical limitations.
 
-For \texttt{auto}, massive speed ups for $q \geq 2$ must mainly be ascribed to aggressive and successful expression writing.
+For \texttt{auto}, massive speed ups for $q \geq 2$ are to be ascribed to aggressive and successful expression writing.
 Hyperelasticity problems are really compute-intensive, with thousands of operations being performed, so reductions in redundant and useless computation are crucial.
 Complex forms like hyperelasticity would benefit from further ``specialized'' optimizations: for example, it is a known technical limitation of COFFEE that, in some circumstances, less temporaries could (should) be generated and that hoisted code could (should) be suitably distributed over different loops to minimize register pressure (e.g.
-COFFEE could carefully apply loop fission for obtaining significantly better register usage).
+COFFEE could apply loop fission for obtaining significantly better register usage).
 We expect to obtain considerably faster code once such optimizations will be incorporated.
@@
 In the regime $q \geq 2$ and $nf=1$, peformance improvements are less pronounced moving from $p=1$ to $p=2$, although still significant; in particular, we notice a drop at $p=2$, followed by a raise up to $p=4$.
-It is worth observing that this effect is common to all set of optimizations.
+It is worth observing that this effect is common to all sets of optimizations.
 The hypothesis is that this is due to the way coefficient functions are evaluated at quadrature points (identical in all configurations), which cannot be easily vectorized unless a change in storage layout and loops order is implemented in the code (abstract syntax tree) generator on top of COFFEE.

### 2014-08-13 TOMS_Paper/assembly.tex
Update autotuner table + minor tweaks

@@
 
+\paragraph{Autotuning Overhead}
 We first comment on the overhead of the autotuning process.
@@
 Autotuning, therefore, introduces an overhead in the application execution time.
-For a given problem instance, such overhead originates from four sources: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution.
+Such overhead originates from four operations: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution.
 Of these four points, we note that: the cost of 4) is relatively small, because each variant's execution time is bound by an empirically-found value (e.g.
@@
 Most often, it took less than thirty seconds.
-Rarely it needed more than a minute, specifically in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$.
+Rarely it needed more than a minute, as for instance in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$.
 This analysis certifies that the autotuner overhead is definitely sustainable in real-world applications.
 
+\paragraph{Selected Optimizations}
 In this paper and in~\cite{Luporini}, we have repeatedly claimed that different forms (and different discretizations) require distinct sets of transformations to reach close-to-peak performance.
@@
 Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible.
-If the value of the column is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one.
+If the value is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one.
 Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}.
@@
 Other columns are of obvious meaning.
-Values in the various columns illustrate the number of problem instances out of the total (column \texttt{mumber of variants}) in which an optimization was activated.
+Values in the various cells illustrate the number of problem instances (out of the total reported in column \texttt{number of variants}) in which an optimization was activated.
 Obviously, more transformations are typically used in combination in a same problem.
-
-%In the table, we implicitly assume that padding and data alignment (Section~\ref{sec:padding}), as well as precomputation of integration-dependent terms (Section~\ref{sec:expr-rewriter-obj}) have always been applied.
 
@@
 \cline{1-10}
-problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & BLAS \\
+problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & M \\
 \cline{1-10}
-\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 8 & 17 & 15 & 0 \\
-\texttt{mass} & 64 & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 12 & 26 & 20 & 0 \\
+\texttt{mass} & 64 & base & 0 & 0 & all & 0 & 0 & 31 & 2 \\
 \texttt{elasticity} & 64 & aggressive & 39 & 0 & all & 0 & 11 & 0 & 2 \\
-\texttt{poisson} & 64 & aggressive & 50 & 0 & all & X & X & 1 & 0 \\
-\texttt{mixed poisson} & X & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
-\texttt{hyperelasticity} & 32 & aggressive & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
+\texttt{poisson} & 64 & aggressive & 52 & 0 & all & all & 0 & 0 & 0 \\
+\texttt{mixed poisson} & 192 & base & 0 & 0 & all & 32 & 68 & 36 & 0 \\
+\texttt{hyperelasticity} & 32 & aggressive & all & all & all & 2 & 0 & 0 & 0 \\
 \cline{1-10}
@@
 \begin{table*}[t]
-\tbl{Autotuner output.}{ \scriptsize \autotuneroutput }
+\tbl{Summary of the optimizations selected by the autotuner in a number of forms}{ \scriptsize \autotuneroutput }
 \label{table:autotuner-output}

### 2014-08-14 TOMS_Paper/assembly.tex
Add draft abstract

@@
 \begin{abstract}
-Abstract goes here
-%
-%The code is generated from a high level language for PDEs. An
-%invocation of the kernel typically does tensor-like calculators over
-%relatively small matrices; e.g., triply nested loops operating on
-%about a dozen iterations. Because of the DSL context, COFFEE has a lot
-%of flexibility on how it organizes/orders the calculations to tailor
-%execution to a target architecture.
-%
+The need for rapidly implementing high performance and portable finite element methods has led to approaches based on automated code generation, in which local assembly operations are derived from the mathematical specification of a weak variational form.
+Local assembly code must be efficient, since the computational cost can dramatically increase with the complexity of the input form, covering a significant fraction of the overall computation run-time.
+Achieving high performance, however, is non-trivial, due to the complexity of mathematical expressions involved in the numerical integration and the small size of the loop nests, which prevent many standard compiler optimizations.
+We present an optimization strategy centred on expression rewriting, to minimize the executed floating point operations, and code specialization, to fully exploit the capability of the underlying platform, e.g.
+SIMD vectorization.
+By also relying on model-driven, dynamic autotuning, we report notable performance improvements over state-of-the-art optimization systems for local assembly operations.
 \end{abstract}

### 2014-08-15 TOMS_Paper/assembly.tex
Minor fixes

@@
 
-We have presented ...
-...TODO...: straightforward extendibility to right-hand side assembly
-% Mass DGEMM utile explicit method etc
+We started discussing how complex is the problem of optimizing local assembly operations in the context of automated code generation, in which an infinite range of possibly extremely complicated variational forms need to be supported.
+Available compilers and third-party libraries cannot exploit the structure inherent assembly expressions, and most of the time fail at producing efficient code.
+We have demonstrated that when optimizing local assembly kernels three points must be addressed: minimization of floating point operations, instruction-level parallelism (especially SIMD vectorization) and data (register) locality.
+Our approach is based on a two-stage optimization procedure: expression rewriting and code specialization.
+The first stage produces code that suits optimizations in the second stage.
+Our motivation for this structured strategy was that we felt state-of-the-art optimization systems were not adequate enough for moderately and remarkably complex variational forms.
+
+We have implemented the Expression Rewriter and the Code Specializer in COFFEE, a platform-independent optimizing compiler tailored to local assembly operations, integrated with the popular Firedrake framework.
+This, along with the extensive and successful performance evaluation provided, prove the maturity of our work.
 

### 2015-07-23 TOMS_Paper/assembly.tex
Overhaul the paper, deleting almost everything (it mostly was shit)

--- 2014-08-13 TOMS_Paper/assembly.tex
@@
 % Page heads
-\markboth{F. Luporini et al.}{Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization}
+\markboth{F. Luporini et al.}{Optimal Finite Element Integration}
 
 % Title portion
-\title{Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization}
+\title{Optimal Finite Element Integration}
 \author{Fabio Luporini
@@
 \begin{abstract}
-Abstract goes here
-%
-%The code is generated from a high level language for PDEs. An
-%invocation of the kernel typically does tensor-like calculators over
-%relatively small matrices; e.g., triply nested loops operating on
-%about a dozen iterations. Because of the DSL context, COFFEE has a lot
-%of flexibility on how it organizes/orders the calculations to tailor
-%execution to a target architecture.
-%
+...
 \end{abstract}
@@
 
-\category{G.4}{Mathematical Software}{Parallel and vector
-  implementations}
+\category{G.4}{Mathematical Software}{Parallel and vector implementations}
 
@@
 
-\keywords{Finite element integration, local assembly, compilers,
-  optimizations, SIMD vectorization}
+\keywords{Finite element integration, local assembly, compilers, optimizations, SIMD vectorization}
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2014. 
-  Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. Optimal Finite Element Integration.}
 
@@
 
-%Computational cost is a critical limitation in scientific computing,
-%especially for finite element simulations. To provide one particular
-%example we are particularly concerned about, it has been well
-%established that mesh resolution Consequently, our aggressive
-%optimization of local assembly, which may even take up to 60% of the
-%overall FEM's execution time, directly impacts the performance of
-%large-scale scientific simulations running on supercomputers.
-
-\section{Introduction}
+\section{Introduction and Motivations}
 
 The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
-This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become incredibly popular over the last years.
-In these frameworks, the weak variational form of a given problem is expressed at high-level by means of a domain-specific language.
-Such a mathematical specification is suitably manipulated and then passed as input to a form compiler, whose goal is to generate a representation of local assembly operations.
+This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become increasingly popular over the last years.
+In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language.
+The mathematical specification is manipulated and then passed to a form compiler, which generates a representation of local assembly operations.
 These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution.
@@
 
-Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Finit Element local assEmbly.}, fully integrated with the Firedrake framework~\cite{Luporini}.
-Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic.
-The code transformations occur on an intermediate representation of the assembly operation, which is ultimately translated into platform-specific code.
-Domain knowledge is exploited in two ways: for simplifying the implementation of code transformations and to make them extremely effective.
-Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input.
-For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process.
-Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only.
-We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.
+Our research has resulted in ....
+... we build on our previous work~\cite{Luporini} ... and present a ... We argue that for complex, realistic forms, peak performance can be achieved only by  ... 
+... also low-level optimisation ...
 
-In this paper, we build on our previous work~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation.
-We argue that for complex, realistic forms, peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to ensure effective register utilization and SIMD vectorization.
-The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become steps of code specialization, whereas generalized loop-invariant code motion is performed during the expression rewriting stage.
-We complement and generalize our previous work with the following contributions.
-
-Expression rewriting aims at reducing the computational intensity of local assembly kernels by rescheduling arithmetic operations based on a set of rewrite rules.
-Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage.
-While rewriting an assembly expression, domain knowledge is used in several ways, for example to avoid redundant computation.
-Secondly, we propose an algorithm that, relying on symbolic execution, restructures the code so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability.
-These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.
-
-At code specialization time, transformations are applied to maximize the exploitation of the underlying platform's resources, e.g.
-SIMD lanes.
-On top of the work in~\cite{Luporini}, we provide a number of contributions.
-Firstly, we revisit the padding and data alignment transformation to account for the new code structures produced by expression rewriting.
-This is of fundamental importance to achieve effective SIMD vectorization.
-Secondly, we address an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix-matrix multiply routines.
-BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods.
-However, we will show that in corner, yet meaningful cases, especially in forms characterized by the presence of pre-multiplying functions and relatively high-order function spaces, a BLAS-based implementation can be successful.
-Finally, we introduce a model-driven, dynamic autotuner that transparently evaluates multiple sets of code transformations to determine the best optimization strategy for a given problem.
-The main challenge here is to build, for a generic form, a reasonably small search space that comprises most of the effective code variants.
-
-Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework.
-Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex hyperelasticity models.
+This is all implemented in COFFEE, which in turn is integrated with the Firedrake framework.
+We provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex (hyperelasticity) models.
 We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions.
-To clearly distinguish the improvement achieved by this work, we will compare, for each test case, four sets of code variants: 1) unoptimized code, i.e.
+To clearly distinguish the improvement achieved by this work, we will compare, for each test case, X sets of code variants: 1) unoptimized code, i.e.
 a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e.
-the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper.
-Notable performance improvements of 4) over 1), 2) and 3) are reported and discussed.
-
-%The contributions of this paper can be summarized as follows:
-%\begin{enumerate}
-%\item 
-%\item extensive and unprecedented performance evaluation in a wide range of forms of increasing complexity
-%\end{enumerate}
-
-%Extensive performance evaluation...
-
-%This paper can be thought of as a continuation of the work presented in~\cite{Luporini}: in particular, we answer the remaining open questions, generalize and formalize an approach to minimize redundant computation using a similar philosophy, we show ~\cite{quadrature1}.
+the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; ....
 
@@
 
-\subsection{Quadrature for Finite Element Local Assembly}
 We summarize the basic concepts sustaining the finite element method following the notation adopted in~\cite{quadrature1} and~\cite{Francis}.
@@
 
+\subsection{Quadrature Mode}
 Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$.
@@
 
-In the next sections, we will often refer to the local element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} deriving from the weak variational problem.
+\subsection{Tensor Mode}
+...
 
-\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
-\label{sec:summary-opts}
-If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality.
-In this paper, we tackle these three points building on our previous work (\cite{Luporini}), in which we presented COFFEE, a mature, platform-independent compiler capable of optimizing local assembly code.
+\section{Optimality of Multilinear Loop Nests}
+\label{sec:optimal-impl}
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.6]{pictures/coffee-scheme}
-\caption{Outline of COFFEE.
-The input is either a string of C code or an abstract syntax tree representation.
-The optimization process goes through two main steps: \textit{expression rewriting} and \textit{code specialization}.
-The autotuner, incorporated with the Code Specializer, use heuristics to select the set of code variants that, for a given variational form, are likely to reach high performance.}\label{fig:coffee-outline}
-\end{center}
-\end{figure}
+We need to define: 
+\begin{itemize}
+\item multilinear loop nests, static loop nests; (note: the loop over the elements could also be considered static, if mesh didn't move...)
+\item concept of reuse in a loop;
+\item prove that reuse can always be eliminated from the innermost loop of multilinear loop nests. 
+\item definition of optimality
+\item insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
+\end{itemize}
 
-The high-level structure of COFFEE is outlined in Figure~\ref{fig:coffee-outline}.
-The input is either C code, explicitly provided by the user as a string, or an abstract syntax tree (henceforth AST) representation automatically generated at the higher level of abstraction.
-We have integrated COFFEE with the Firedrake framework; here, the input is originated by the FEniCS Form Compiler~\cite{FFC-Compiler} in the form of an AST.
-Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE.
-This is used to introduce more effective, specialized optimizations, as well as to simplify the implementation.
+Need to show examples.
 
-Two conceptually distinct software modules can be individuated: the Expression Rewriter and the Code Specializer.
-The former targets arithmetic intensity.
-Its role is to transform the assembly expression and, therefore, the enclosing loop nest, so as to minimize the number of floating point operations that are performed to evaluate the element matrix (or vector).
-The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality.
-As shown in the Figure, the Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer.
-As described later, we have invested some effort to ensure that the code transformations applied at expression rewriting time would not break any code specialization opportunity.
-This, for example, would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.
-
-To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work.
-Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to both~\cite{Luporini} and~\cite{quadrature1}.
-
-In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation.
-We have also shown how different subsets of optimizations are needed to achieve close-to-peak performance in different variational forms.
-In particular, we can summarize our study as follows
+\section{Automated Code Generation of Optimal Multilinear Loop Nests}
 
 \begin{itemize}
-\item \emph{Generalized Loop-invariant Code Motion}.
-Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest.
-As reiterated in Section~\ref{sec:expr-rewriting}, we work around this limitation by performing source-level code motion, while also achieving vectorization of invariant code.
-\item \emph{Padding and data alignment}.
-The small size of the loop nest (integration, test, and trial functions loops) require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code.
-Data alignment can be enforced as a consequence of padding.
-\item \emph{Vector-register Tiling}.
-Blocking at the level of vector registers, which we perform exploiting the specific memory access pattern of the assembly expressions (i.e.
-a domain-aware transformation), improves data locality beyond traditional unroll-and-jam optimizations.
-This is especially true for relatively high polynomial order (i.e.
-greater than 2) or when pre-multiplying functions are present.
-\item \emph{Expression Splitting}.
-In certain assembly expressions the register pressure is significantly high: when the number of basis functions arrays (or, equivalently, temporaries introduced by loop-invariant code motion) and constants is large, spilling to L1 cache is a consequence for architectures with a relatively low number of logical registers (e.g.
-16/32).
-We exploit sum's associativity to ``split'' the assembly expression into multiple sub-expressions, which are computed individually.
+\item Deficiencies of previous approaches
+\item  Building block operations in coffee: 1) eliminating reuse, 2)  insights on the various tree algorithms employed
+\item cost model for deciding what to do
+\item constrained increase in the working set size
+\item heuristic optimisation of constant and quadrature dependent expressions
 \end{itemize}
 
-In Figure~\ref{fig:coffee-outline}, we show where these transformations are applied in the COFFEE pipeline; also, the novel contributions of this work are highlighted in red.
-In the next two sections, we describe the new functionalities of Expression Rewriter and Code Specializer, respectively.
-
-%Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.
-
-% Far notare che low-level si, ma sono guidate da high-level knowledge! Questo in risposta alla domanda: ma allora perche' non sfruttare la domain knowledge che hai, che senso ha separare...
-
-% Vantaggio di automatizzare rispetto che scrivere a mano
-% which would be practically unfeasible to achieve by hand-written code
-
-\section{Expression Rewriting}
-\label{sec:expr-rewriting}
-Loop-invariant code motion is the key to reduce the computational intensity of an assembly expression. The Expression Rewriter still relies on this technique, although it makes several steps forward.
-
-Firstly, exploiting operators properties like associativity, distributivity, and commutativity, it manipulates assembly expressions to expose more code hoisting and SIMD vectorization opportunities, as well as to balance the register pressure in the various levels of the loop nest.
-There are many possibilities of rewriting an expression, so the search space can be quite large.
-We address this problem by defining a simple yet systematic way of rewriting an expression based on a set of formal rewrite rules, which are meticulously applied by COFFEE.
-
-Secondly, the Expression Rewriter re-structures iteration spaces so as to avoid arithmetic operations over tabulated basis functions columns that are statically known to be zero-valued.
-Zero-valued columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements.
-A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with that is the use of indirection arrays in the generated code, which break many of the optimizations applicable at code specialization time, including SIMD vectorization.
-In Section~\ref{sec:zeros}, we present a novel approach based on symbolic execution.
-
-%In COFFEE, the application of the re-writing rules preceds the elimination of operations on zeros.
-
-%It is worth noting how writing these transformations ``by hand'' would be practically unfeasible, especially when the expressions are particularly complex.
-%
-%
-%In this section, we describe our efforts in producing a more effective tool capable of explore how to leverage automated code generation to expose more opportunities to the code hoister. 
-%
-%Automated code generation, however, Mathematical expressions can be transformed to decrease the number of arithmetic operations required to evaluate the element tensor. A first 
-%
-%Minimizing the number of floating point operations for evaluating the element tensor of a generic form
-%
-
-\subsection{Motivations and Objectives}
-\label{sec:expr-rewriter-obj}
-
-\begin{figure}[t]
-\tiny
-\centering     
-\subfigure[Original (simplified) code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
-~~~~~~~~~~~~~~~~
-\subfigure[The code after that generalized loop-invariant code motion has been applied]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
-~\\
-~\\
-\subfigure[Once terms are factorized, further code hoisting opportunities arise.
-Here, \texttt{T3} stores the precomputation of an invariant sub-expression exposed by factorization]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subfigure[Hoisting and scalar-expansion of integration-dependent terms allow to achieve full SIMD vectorization of the assembly code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
-\caption{Applying a sequence of transformations to an excerpt from a Burgers' local assembly kernel, suitably simplified to improve readability.}
-\end{figure}
-
-Consider the local element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem.
-The assembly expression, produced by the FEniCS Form Compiler, has been deliberately simplified, and code details have been omitted for brevity and readability.
-In practice, assembly expressions can be much more complex, for example depending on the differential operators employed in the variational form; however, this example is representative enough for highlighting patterns that are common in a large class of problems.
-
-A first glimpse of the code suggests that the sub-expression \texttt{a*f0*A[i][j]+b*f1*B[i][j]} is invariant with respect to the innermost (trial functions) loop \texttt{k}, so it can be hoisted at the level of the outer loop \texttt{j} to avoid redundant computation.
-This is indeed a standard compiler transformation, supported by any available compilers, so, in principle, there should be no need to transform the source code explicitly.
-With a closer look we notice that the sub-expression \texttt{d*D[i][k]+e*E[i][k]} is also invariant, although, this time, with respect to the outer (test functions) loop \texttt{j}.
-In~\cite{Luporini}, we have showed that available compilers limit the search for code motion opportunities to the innermost loop of a given loop nest.
-Moreover, the lack of cost models to ascertain both the optimal place where to hoist an expression and whether or not vectorizing it at the price of extra temporary memory is a fundamental limiting factor.
-The \textit{generalized loop-invariant code motion} technique that we implemented in COFFEE, leading to the code in Figure~\ref{fig:invariant-code}, addressed these limitations, which are critical in the case of non-trivial assembly expressions, resulting in run-time improvements up to 3$\times$.
-
-In this paper, we target particularly complex variational forms for which a more aggressive transformation is required.
-We generalize the problem of finding and hoisting invariant sub-expressions by formulating the following critical question: \textit{how an assembly expression should be rewritten so as to minimize both floating point operations and data movement}?
-To minimize floating point operations, in terms of instructions executed, it is important to reduce the absolute number of arithmetic operations to evaluate the element matrix as well as to achieve full vectorization of the assembly loop nest.
-
-We start considering the case of assembly expressions that, after generalized loop-invariant code motion has been applied, still  ``hide'' opportunities for code hoisting.
-By examining again the code in Figure~\ref{fig:invariant-code}, we notice that the basis function array \texttt{A} iterating along the \texttt{[i,j]} loops appears twice in the expression.
-By expanding the products in which \texttt{A} is accessed and by applying sum commutativity, terms can be factorized.
-This has two effects: firstly, it reduces the number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{A[i][k]/c+T2[k]*f} invariant with respect to loop \texttt{j}.
-Consequently, hoisting can be performed, resulting in the code in Figure~\ref{fig:factorized-code}.
-In general, exposing factorization opportunities requires traversing the whole expression tree, and then expanding and moving terms.
-It also needs heuristics to select a factorization strategy: there may be different opportunities of reorganizing sub-expressions, and, in our case, the best is the one that maximizes the invariant code eventually disclosed.
-We will discuss this aspect formally in Section~\ref{fig:rewrite-rules}.
-
-As a second observation, we note that integration-dependent expressions are inherently executed as scalar code.
-For example, the \texttt{f0*a} and \texttt{f1*b} products in Figure~\ref{fig:factorized-code} depend on the loop along quadrature points; these operations are performed in a non-vectorized way at every \texttt{i} iteration.
-This is not a big issue in our running example, in which the scalar computation represents a small fraction of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}).
-In such forms, the amount of computation independent of both test and trial functions loops is so large that it has a significant impact on the run-time, despite being executed only $O(I)$ times (with $I$ number of quadrature points).
-We have therefore implemented an algorithm to move and scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.
-
-The third motivation for expression rewriting concerns the register pressure in the innermost loop.
-Once the code has been optimized for arithmetic intensity, it is important to think about how the transformations impacted register allocation.
-Assume the local assembly kernel is executed on a state-of-the-art CPU architecture having 16 logical registers, e.g.
-an Intel Haswell.
-Each value appearing in the expression is loaded and kept in a register as long as possible.
-In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every \texttt{i} iteration.
-This implies that at every iteration of the \texttt{[j,k]} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops.
-In more complicated expressions, the percentage of registers destined to store such constant terms can be even higher.
-Registers are, however, a precious resource, especially when evaluating compute-intensive expressions.
-The smaller is the number of available free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e.
-it can worsen data locality), or it may prevent the effective application of standard transformations, e.g.
-loop unrolling.
-We aim at relieving this problem by suitably expanding terms and introducing, where necessary, additional temporary values.
-We illustrate this in the following example.
-
-%An analogous analysis applies to processors with larger numbers of registers, since using loop unroll or loop unroll-and-jam to expose more instruction-level parallelism would increase the requirements on registers.
-
-\begin{figure}[t]
-\tiny
-\centering
-\subfigure[Original (simplified) code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
-~\\
-\subfigure[The code after expansion of \texttt{det*W[i]}]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
-~~
-\subfigure[The code after expansion of \texttt{g}. Note the need to introduce a new temporary array]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
-\caption{Expansion of terms to improve register pressure in a local assembly kernel}\label{fig:expanded-code}
-\end{figure}
-
-Consider a variant of the Burgers local assembly kernel, shown in Figure~\ref{fig:toexpand-code}.
-This is again a representative, simplified example.
-We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of the invariant sub-expression stored in \texttt{T1}, resulting in code as in Figure~\ref{fig:expanded-1-code}.
-Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since \texttt{T1} is also accessed in the \texttt{T1[j]*A[i][k]} sub-expression.
-The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}.
-Generalizing, this is a problem of data dependencies: to solve it, we use a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}.
-The dependency graph is initially empty, and is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms.
-The dependency graph is then queried to understand when expansion can be performed without resorting to new temporary values.
-This aspect is formalized in the next section.
-
-\subsection{Rewrite Rules}
-\label{sec:rewrite-rules}
-In general, assembly expressions produced by automated code generation can be much more complex than those we have used as examples, with dozens of terms involved (basis function arrays, derivatives, coefficients, ...) and hundreds of (nested) arithmetic operations.
-Our goal is to establish a portable, unified, platform-independent, and systematic way of reducing the computational strength of an expression exploiting the intuitions described in the previous section.
-This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems, such as COFFEE.
-In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform 4) scalar-expansion of integration-dependent terms to achieve full vectorization of the assembly code.
-
-We centre such an algorithm around a set of rewrite rules.
-These rules drive the transformation of an expression, prescribe where invariant sub-expressions will be moved (i.e.
-at what level in the loop nest), and track the propagation of data dependencies.
-When applying a rule, the state of the loop nest must be updated to reflect, for example, the use of a new temporary and new data dependencies.
-We define the state of a loop nest as $L = (\sigma, G)$, where $G = (V, E)$ represents the dependency graph, while $\sigma : Inv \rightarrow S$ maps invariant sub-expressions to identifiers of temporary arrays.
-The notation $\sigma_i$ refers to invariants hoisted at the level of loop $i$.
-We also introduce the \textit{conditional hoister} operator $[]$ on $\sigma$ such that
-\begin{gather*}
-\sigma[\sfrac{v}{x}] = 
-\begin{cases}
-\sigma(x) \text{~~~~~~~~if $x \in Inv$; $v$ is ignored}\\
-v \text{~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $\sigma$(x) = v}\\
-\end{cases}
-\end{gather*}
-That is, if the invariant expression \texttt{x} has already been hoisted, $[]$ returns the temporary identifier hosting its value; otherwise, \texttt{x} is hoisted and a new temporary $v$ is created.
-There is a special case when $v = \perp$, used for \textit{conditional deletion} of entries in $\sigma$.
-Specifically
-\begin{gather*}
-\sigma[\sfrac{\perp}{x}] = 
-\begin{cases}
-\sigma(x) \text{~~~~~~~~if $x \in Inv$; $\sigma = \sigma \setminus (x, \sigma(x))$}\\
-v \text{~~~~~~~~~~~~~~if $x$ $\notin$ $Inv$; $v \notin S$}\\
-\end{cases}
-\end{gather*}
-In other words, the invariant sub-expression \texttt{x} is removed and the temporary identifier that was hosting its value is returned if \texttt{x} had been previously hoisted; otherwise, a fresh identifier $v$ is returned.
-This is useful to express updates of hoisted invariant sub-expressions when expanding terms.
-
-In the following, a generic (sub-)expression is represented with Roman letters \texttt{a, b, ...}; constant terms are considered a special case, so Greek letters \texttt{$\alpha$, $\beta$, ...} are used instead.
-The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression.
-We will refer to $i_0$ as the outermost enclosing loop.
-The notation $a_i$, therefore, indicates that the expression $a$ assumes distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.
-
-\begin{figure}[t]
-\small
-\centering
-\begin{spacing}{1.35}
-\begin{align*}
-~~~~~~&Rule &~~&~~~~~~~~~~~~~~~~~~~~~~~~Precondition \\
-[a_i \cdot b_j]_{(\sigma, G)} &\rightarrow [a_i \cdot b_j]_{(\sigma, G)} ~~&~~&\\
-[(a_i + b_j)\cdot \alpha]_{(\sigma, G)} &\rightarrow [(a_i \cdot \alpha + b_j \cdot \alpha)]_{(\sigma, G)} ~~&~~ &\\
-[a_i \cdot b_j + a_i \cdot c_j]_{(\sigma, G)} &\rightarrow [(a_i \cdot (b_j + c_j)]_{(\sigma, G)} ~~&~~ &\\
-[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
-[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
-~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
-[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
-~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
-\end{align*}
-\end{spacing}
-\caption{Rewrite rules driving the rewriting process of an assembly expression.}\label{fig:rewrite-rules}
-\end{figure}
-
-Rewrite rules for expression rewriting are provided in Figure~\ref{fig:rewrite-rules}; obvious rules are omitted for brevity.
-The Expression Rewriter applies the rules while performing a depth-first traversal of the assembly expression tree.
-Given an arithmetic operation between two sub-expressions (i.e.
-a node in the expression tree), we first need to find an applicable rule.
-There cannot be ambiguities: only one rule can be matched.
-If the preconditions of the rule are satisfied, the corresponding transformation is performed; otherwise, no rewriting is performed, and the traversal proceeds.
-As examples, it is possible to instantiate the rules in the codes shown in Figures~\ref{fig:original-code} and~\ref{fig:toexpand-code}; eventually, the optimized codes in Figures~\ref{fig:factorized-code} and~\ref{fig:expanded-2-code} are obtained, respectively.
-
-To what extent should rewrite rules be applied is a question that cannot be answered in general.
-In some problems, a full rewrite of the expression may be the best option; in other cases, on the other hand, an aggressive expansion of terms, for example, may lead to high register pressure in the loops computing invariant terms, worsening the performance.
-In Section~\ref{sec:autotune} we explain how to leverage the Code Specializer to select a suitable rewriting strategy for the problem at hand.
+\section{Low-level Optimization}
+\label{sec:code-spec}
+...
 
@@
 \label{sec:zeros}
-We aim at skipping arithmetic operations over blocks of zero-valued entries in basis functions (and their derivatives) arrays.
-Zero-valued columns arise, for example, when taking derivatives on a reference element and when employing mixed elements.
-In~\cite{quadrature1}, a technique to avoid operations on zero-valued columns based on the use of indirection arrays was described (e.g.
-\texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}) and implemented in the context of FEniCS.
+%TODO : say that the reasons to have zeros is to keep the design of the form compiler simple, which just "puts zeros" and "outline the assembly expression".
+
+Operations over blocks of zero-valued entries in tabulated (derivatives of) basis functions can be skipped.
+Zero-valued columns arise, for example, when taking derivatives on a reference element, or when employing vector function spaces.
+In~\cite{quadrature1}, a technique to avoid iteration over zero-valued columns, based on the use of indirection arrays (e.g.
+\texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), was described and tested in \texttt{FFC}.
 We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}.
-Essentially, our strategy avoids indirection arrays in the generated code, which otherwise would break the optimizations applicable at the Code Specializer level, including SIMD vectorization.
+Our strategy differs by avoiding indirection arrays in the generated code.
 
@@
 Consider Figure~\ref{fig:withzeros-code}, an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}.
-The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh.
 The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points.
@@
 The solution adopted in~\cite{quadrature1} is not to generate the zero-valued columns (i.e.
-to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g.
+to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2) and to use an indirection array (e.g.
 $ind = \lbrace 3, 5\rbrace$) to update the right entries in the element tensor $A$.
-This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations.
 %on the other hand, reshrinki tutto su un array solo...
@@
 
-\section{Code Specialization}
-\label{sec:code-spec}
-Code specialization provides a range of code transformations tailored to optimizing for instruction-level parallelism and register locality.
-As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying.
-In this paper, we revisit and enrich this set of transformations as explained in the following sections.
+%TODO : rephrase: " This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations. "
 
-\subsection{Padding and Data Alignment Revisited}
+\subsection{Padding and Data Alignment}
 \label{sec:padding}
@@
 
-\subsection{Exposing Linear Alegbra Operations}
-In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrix-matrix multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS.
-We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM routines, which are usually tuned for large arrays.
-The study was conducted only in the case of a specific Helmholtz form.
-There are scenarios, however, in which tabulated basis functions sizes grow up to a point for which turning to BLAS may actually lead to better performance.
-For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of basis functions tabulated at quadrature points.
-To explore these scenarios, we have developed an algorithm that reduces any assembly expression evaluation to a sequence of DGEMM calls.
-
-We informally provide the main steps of the algorithm.
-By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions.
-Each outer product is then isolated, i.e.
-the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions.
-Statements in the bodies of the surrounding loops (e.g.
-coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:expr-rewriter-obj}.
-This renders the loop nest perfect; that is, there is no intervening code among the various loops.
-The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
-\begin{equation*}
-A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
-\end{equation*}
-where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time.
-Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g.
-two dimensional arrays are flatten as one dimensional arrays).
-The translation into a sequence of DGEMM calls is the last, straightforward step.
-
-% 
-\subsection{Model-driven Dynamic Autotuning}
-\label{sec:autotune}
-We have demonstrated in~\cite{Luporini} that determining the sequence of transformations that  maximizes the performance of a problem requires investigating a broad range of factors, including mathematical structure of the input form, polynomial order of employed function spaces, presence of pre-multiplying functions, and, of course, the characteristics of the underlying architecture.
-
-In~\cite{Luporini}, we proposed a simple cost model to drive the optimization process.
-In this paper, we have added a significant number of options to the set of possible optimizations, so the selection problem is now far more challenging.
-The sole Expression Rewriter, for instance, can generate a wide variety of implementations by just applying rewrite rules to different extents.
-We found difficult extending the cost model to effectively cover such a large search space.
-
-We tackle the optimizations selection problem by compiler autotuning.
-Not only does it allow to determine the best combination of transformations, also it enables exploring parametric low-level optimizations, such as loop unroll, unroll-and-jam, and interchange, by trying different unroll factors and loop permutations.
-By leveraging the cost model defined in our previous study, domain knowledge, and a set of heuristics, we manage to keep the autotuner overhead at a minimum, whilst achieving significant speed ups over the purely cost-model-based implementations in all of the forms we have evaluated.
-In particular, our autotuner usually requires order of seconds to determine the fastest kernel implementation, a negligible overhead when it comes to iterate over real-life unstructured meshes, which can contain up to trillions of elements (e.g.
-~\cite{Rossinelli2013}).
-
-COFFEE analyzes the input problem and decides what variants it is worth testing through autotuning, as described later.
-Each variant is obtained by requesting specific transformations to the Expression Rewriter and the Code Specializer.
-The possible variants are then provided to the autotuner, in the form of abstract syntax trees.
-The autotuner is a template-based code generator.
-By inspecting an abstract syntax tree, it determines how to generate ``wrapping'' code that 1) initializes kernel's input variables with fictitious values and 2) calls the kernel.
-These two points are executed repeatedly in a \emph{while} loop for a pre-established amount of time (order of milliseconds).
-At the exit of the \emph{while} loop, the times the kernel was invoked is recorded.
-Eventually, the variant executed the largest number of iterations is designated as the fastest implementation.
-Suitable compiler directives are used to prevent inlining of all function calls: this avoids the situation in which some variants are inlined and some are not, which would fake the autotuner's output.
-
-The autotuning process is dynamic: depending on the complexity of the input problem, more or less variants are tried.
-General heuristics, which can be considered a revisited version of those presented in~\cite{nek5000}, are applied
-\begin{itemize}
-\item Loop permutations that are likely to worsen the performance are excluded from the search space.
-According to the cost model, and for the same reasons explained in~\cite{Luporini}, we enable only variants in which the loop over quadrature points is either the outermost or the innermost.
-This is due to the fact that versions of the code in which such loop lies between the test and trial functions loops are typically lower performing.
-\item The unroll factors must divide the loop bounds evenly to avoid the introduction of reminder (scalar) loops.
-\item The innermost loop is never explicitly unrolled. This is because we expect auto-vectorization along this loop, so memory accesses should be kept unit-stride. 
-\end{itemize}
-The autotuner is also domain-aware: the following heuristics, which capture properties of the computational domain, are exploited
-\begin{itemize}
-\item The lengths of test and trial functions loops are identical in some cases, for example when they originate from the same function space.
-In such cases, since for the employed storage layout the memory access pattern is symmetrical along these two loops, we prune their interchange from the search space.
-\item The larger is the polynomial order of the method, the larger is the assembly loop nest.
-In these cases, we impose a bound on the loop nest's overall unroll factor (which we found empirically) to avoid uselessly testing too many unroll factors.
-\item On the other hand, if the polynomial order is low, i.e. when the loop nest is small, we prune variants that we know will be low-performing, e.g. those resorting to BLAS.
-\item We specifically select two levels of expression rewriting.
-In the ``base'' level, only generalized loop-invariant code motion, as described in~\cite{Luporini}, is applied.
-This means that only a subset of the rewrite rules exposed in~\ref{fig:rewrite-rules} will be considered.
-In the ``aggressive'' level, all of the rewrite rules are applied.
-Many other trade-offs, which we do not explore, would be feasibile, however.
-\item For the expression splitting optimization described in~\cite{Luporini} and summarized in Section~\ref{sec:summary-opts}, we test only three split factors, namely 1, 2, 4.
-Also, if the input problem uses mixed function spaces, the iteration space is already split at expression rewriting time to avoid computation over zero-valued columns; in these cases, we do not further apply expression splitting.
-\item Based on the cost model, the padding and data alignment optimization is always applied. 
-\end{itemize}
-
-All of the previous points contribute to minimize the overhead of the autotuning process. We will discuss the actual cost of autotuning in Section~\ref{sec:auto-analysis}.
-%COFFEE allows the user to abstract from finding the optimal optimization strategy for a given problem by resorting to its model-driven, dynamic autotuner.
+\subsection{Vector-promotion of Integration Quantities}
+....
 
@@
 All codes were executed in the context of the Firedrake framework.
-
-We do not compare to the FEniCS Form Compiler's tensor contraction mode~\cite{Kirby-FEM-opt} because of three reasons: first, in~\cite{quadrature1} it has been demonstrated the superiority of quadrature as the complexity of a form increases, so it would be superfluous to repeat the same analysis.
-Second, our aim is to show the effect of low-level optimizations on the code, especially SIMD vectorization, which is not feasible in tensor mode due to the inherent structure of the code.
-Third, tensor mode code generation fails due to hardware limitations in many of the test cases that we show below.
 
@@
 \begin{figure}[t]  
-\includegraphics[scale=0.7]{perf-results/helmholtz}
-\caption{Helmholtz results.}\label{fig:helmholtz}
+...
+%\includegraphics[scale=0.7]{perf-results/helmholtz}
+%\caption{Helmholtz results.}\label{fig:helmholtz}
 \end{figure}
@@
 We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$.
-This is a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization.
+This is a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler optimizations, SIMD auto-vectorization.
 The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$.
@@
 \begin{figure}[t]  
-\includegraphics[scale=0.7]{perf-results/elasticity}
-\caption{Elasticity results.}\label{fig:elasticity}
+...
+%\includegraphics[scale=0.7]{perf-results/elasticity}
+%\caption{Elasticity results.}\label{fig:elasticity}
 \end{figure}
@@
 
-% TODO: mention aggressive expression rewriting somewhere
-
 \paragraph{Poisson}
 \begin{figure}[t]  
-\includegraphics[scale=0.7]{perf-results/poisson}
-\caption{Poisson results.}\label{fig:poisson}
+...
+%\includegraphics[scale=0.7]{perf-results/poisson}
+%\caption{Poisson results.}\label{fig:poisson}
 \end{figure}
@@
 \begin{figure}[t]  
-\includegraphics[scale=0.7]{perf-results/hyperelasticity}
-\caption{Hyperelasticity results.}\label{fig:hyperelasticity}
+...
+%\includegraphics[scale=0.7]{perf-results/hyperelasticity}
+%\caption{Hyperelasticity results.}\label{fig:hyperelasticity}
 \end{figure}
@@
 
-\subsection{Details on the Autotuning Process}
-\label{sec:auto-analysis}
-
-\paragraph{Autotuning Overhead}
-We first comment on the overhead of the autotuning process.
-In the context of Firedrake, the framework in which COFFEE is integrated, the autotuner is executed at run-time, once the local assembly kernels are provided by the FEniCS Form Compiler.
-Autotuning, therefore, introduces an overhead in the application execution time.
-Such overhead originates from four operations: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution.
-Of these four points, we note that: the cost of 4) is relatively small, because each variant's execution time is bound by an empirically-found value (e.g.
-some milliseconds).
-The cost of all four points is constrained by our heuristics to prune the search space, as described in Section~\ref{sec:autotune}.
-Moreover, for a given form and a given discretization, the autotuner needs to be executed only once, since its output is saved and reused for later assemblies.
-This implies that if the assembly occurs in a time stepping loop, or the same form is executed on a different mesh, or known quantities of the input problem are changed, then the assembly cost is rapidly amortized.
-Premised that, the most important thing remains that when working with real unstructured meshes - which are likely to be composed of millions of elements, leading to long-lasting assembly phases - the autotuner overhead practically becomes completely negligible.
-In our experience, and in particular in the four examined problems, the autotuning process lasted less than a minute in the majority of cases.
-Most often, it took less than thirty seconds.
-Rarely it needed more than a minute, as for instance in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$.
-This analysis certifies that the autotuner overhead is definitely sustainable in real-world applications.
-
-\paragraph{Selected Optimizations}
-In this paper and in~\cite{Luporini}, we have repeatedly claimed that different forms (and different discretizations) require distinct sets of transformations to reach close-to-peak performance.
-To demonstrate this, we now report details about the output of the autotuning process.
-We show that for the four examined forms - more specifically, for the 224 problem instances $[form, q, p, nf]$ illustrated in the previous figures - a plethora of optimization strategies have been selected by the autotuner.
-We also complement and strengthen our claim by showing the autotuner's output for two additional forms whose performance results, for brevity, were not shown in Section~\ref{sec:perf-results-forms}: a Mass and a Mixed Poisson problems.
-
-Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner.
-To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$.
-Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible.
-If the value is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one.
-Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}.
-The \texttt{split} column refers to the expression splitting transformation (\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}).
-The \texttt{unroll} column indicates the application of explicit unrolling.
-Other columns are of obvious meaning.
-Values in the various cells illustrate the number of problem instances (out of the total reported in column \texttt{number of variants}) in which an optimization was activated.
-Obviously, more transformations are typically used in combination in a same problem.
-
-\newcommand{\autotuneroutput}{
-\tiny
-\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
-\cline{3-10}
-\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
-\cline{1-10}
-problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & M \\
-\cline{1-10}
-\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 12 & 26 & 20 & 0 \\
-\texttt{mass} & 64 & base & 0 & 0 & all & 0 & 0 & 31 & 2 \\
-\texttt{elasticity} & 64 & aggressive & 39 & 0 & all & 0 & 11 & 0 & 2 \\
-\texttt{poisson} & 64 & aggressive & 52 & 0 & all & all & 0 & 0 & 0 \\
-\texttt{mixed poisson} & 192 & base & 0 & 0 & all & 32 & 68 & 36 & 0 \\
-\texttt{hyperelasticity} & 32 & aggressive & all & all & all & 2 & 0 & 0 & 0 \\
-\cline{1-10}
-\end{tabulary}
-}
-
-\begin{table*}[t]
-\tbl{Summary of the optimizations selected by the autotuner in a number of forms}{ \scriptsize \autotuneroutput }
-\label{table:autotuner-output}
-\end{table*}
-
-% selected variants (tables)
-% show the difference of ``aggressive'' vs ``base'' expression rewriting
-
 \section{Conclusions}
 \label{sec:conclusions}
-
-We have presented ...
-...TODO...: straightforward extendibility to right-hand side assembly
-% Mass DGEMM utile explicit method etc
+...
 

### 2015-07-23 TOMS_Paper/perf-results/generated_grid_plots.py

@@ -1,115 +1,0 @@
-"""
-Generate a grid of plots given raw data generated by pybench 
-"""
-import numpy as np
-import matplotlib.pyplot as plt
-from mpl_toolkits.axes_grid1 import Grid
-from mpltools import style
-
-import os
-import sys
-from collections import defaultdict
-from ast import literal_eval
-
-filename = "FiredrakeFormsCoffee.dat"
-
-style.use('ggplot')
-
-if len(sys.argv) is not 2:
-    print "Usage: directory-where-to-find-FiredrakeFormsCoffee.dat"
-    sys.exit(0)
-
-with open(os.path.join(sys.argv[1], filename), "r") as f:
-    timings = literal_eval(f.read())['timings']
-
-# First, adjust times in a dictionary as I need
-_timings = defaultdict(dict)
-for info, vals in sorted(timings.items()):
-    q, p, name, opt = info
-    _timings[(q, p)][opt] = vals
-    
-if sys.argv[1] in ['helmholtz', 'poisson', 'elasticity']:
-    fig = plt.figure()
-    grid = Grid(fig, rect=111, nrows_ncols=(4, 4), axes_pad=0.25, label_mode='L')
-    for ax, plot_times in zip(grid, sorted(_timings.items())):
-        info, vals = plot_times
-        q, p = info
-        base = vals[(0, False, False, False)]
-        ffco = vals[(0, False, False, True)]
-        licm = vals[(1, True, False, False)]
-        auto = vals[(0, False, True, False)]
-
-        # Plot parameters
-        N = 3  # We show speedup of ffco, licm, and auto over base
-        ind = np.arange(N)  # the x locations for the groups
-        width = 0.15  # the width of the bars
-        fontsize = 15
-
-        # Compute speed ups
-        nf0 = (base['nf 0']/ffco['nf 0'], base['nf 0']/licm['nf 0'], base['nf 0']/auto['nf 0'])
-        nf1 = (base['nf 1']/ffco['nf 1'], base['nf 1']/licm['nf 1'], base['nf 1']/auto['nf 1'])
-        nf2 = (base['nf 2']/ffco['nf 2'], base['nf 2']/licm['nf 2'], base['nf 2']/auto['nf 2'])
-        nf3 = (base['nf 3']/ffco['nf 3'], base['nf 3']/licm['nf 3'], base['nf 3']/auto['nf 3'])
-
[…]

### 2015-07-27 TOMS_Paper/assembly.tex
Add optimality of loop nests

@@
 \usepackage{tabulary}
+\usepackage{fancybox}
+
+\makeatletter
+\newenvironment{CenteredBox}{% 
+\begin{Sbox}}{% Save the content in a box
+\end{Sbox}\centerline{\parbox{\wd\@Sbox}{\TheSbox}}}% And output it centered
+\makeatother
+
+\let\proof\relax
+\let\endproof\relax
+\usepackage{amsthm}
 \newcommand{\specialcell}[2][c]{%
@@
 \usepackage{xfrac}
+
+% Theorems, definitions, lemmas, etc
+\newtheorem{Def}{Definition}
+\newtheorem{Prop}{Proposition}
 
@@
 \usepackage{url}
-
-% Package to generate and customize Algorithm as per ACM style
-\usepackage[ruled]{algorithm2e}
-\renewcommand{\algorithmcfname}{ALGORITHM}
-\SetAlFnt{\small}
-\SetAlCapFnt{\small}
-\SetAlCapNameFnt{\small}
-\SetAlCapHSkip{0pt}
-\IncMargin{-\parindent}
 
@@
 ... also low-level optimisation ...
+...also for the first time we provide a formal explanation of the math in terms of compiler theory
 
@@
 
+% Magari la summation sui coefficients la sputo dentro? che dici?
+
 \subsection{Tensor Mode}
@@
 
-\section{Optimality of Multilinear Loop Nests}
+\section{Optimality of Loop Nests}
 \label{sec:optimal-impl}
 
-We need to define: 
+In this section, we characterize our definition of optimality and we describe the assumptions under which it holds.
+
+In order to make the document self-contained, we start by reviewing basic compiler terminology.
+
+\begin{Def}[Perfect and imperfect loop nests]
+A loop nest is said to be \textit{perfect} when code appears only in the body of the innermost loop.
+Conversely, the presence of code between any pair of loops in a nest renders the loop nest \textit{imperfect}.
+\end{Def}
+
+A straightforward property of perfect nests is that hoisting of invariant expressions from the innermost loop to the preheader (i.e., the block that precedes the entry point of the nest) is always safe.
+We will later make extensive use of this property.
+
+% In the following, we should represent loops by their iteration variable.
+
+\begin{Def}[Linear loop]
+A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if all expressions appearing in the body of $L$ that use $i$ to access some memory locations are linear in $I$.
+\end{Def}
+
+In this work, we are particularly interested in the following class since it naturally arises from the math described in~\ref{sec:background}.
+
+\begin{Def}[Perfect multilinear loop nest]
+A \textit{perfect multilinear} loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $l \in [0, n-1]$.
+\end{Def}
+
+Note that nothing prevents a perfect multilinear loop nest to be enclosed by an outer, possibly imperfect, nest.
+Indeed, this is the loop structure that will be at the centre of our study.
+We will later elaborate on this.
+First, we need to formulate a definition of optimality for perfect multilinear loop nests.
+To this purpose, it is convenient to introduce the notion of sharing.
+
+\begin{Def}[Sharing]
+A loop $L_i$ presents \textit{sharing} if it contains at least two expressions depending on $i$ that are symbolically identical. 
+\end{Def}
+
+At this point, we can present a simple yet fundamental result.
+
+\begin{Prop}
+\label{prop:share-removal}
+Sharing in a perfect multilinear loop nest $N = L_{i_{0}, i_{1}, ..., i_{n-1}}$ can always be eliminated.
+\end{Prop}
+\begin{proof}
+The demonstration is by construction and exploits linearity.
+We want to transform $N$ into $N'$ such that sharing is absent at all levels of $N'$.
+Starting from the innermost loop $n-1$, the expressions are ''flattened'' by expanding all products involving any of the terms depending on $i_{n-1}$.
+Subsequently, such terms can be factorized.
+Due to linearity, in each factored product there is only one term depending on $i_{n-1}$.
+The factors that do not depend on $i_{n-1}$ are, by definition, loop-invariant, and as such can be hoisted at the level of $i_{n-2}$.
+This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is always safe.
+\end{proof}
+
+Based on this proposition, we define optimality as follows.
+
+\begin{Def}[Optimality of a multilinear loop nest]
+The synthesis of a multilinear loop nest is \textit{optimal} if the amount of operations performed in the innermost loop is minimal.
+\end{Def}
+
+In other words, this implies there is no other nest syntheses that could decrease the number of operations in the body of the innermost loop any further.
+Note that this definition of optimality does not take into account memory requirements.
+If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' clearly makes no sense.
+In the following, we assume to operate in a CPU-bound regime, in which the body of loop nests are characterized by arithmetic-intensive expressions.
+This suits the context of finite element integration.
+
+A second result follows.
+\begin{Prop}
+An optimal synthesis for a perfect multilinear loop nest $N$ can be found in ....
+\end{Prop}
+\begin{proof}
+By construction.
+We consider the body of the innermost loop $L_{i_{n-1}}$.
+Constant expressions can be hoisted outside of $N$ and be replaced with suitable temporaries.
+Loop-dependent terms are grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on loop $i$.
+These sets are sorted in descending order based on their cardinality.
+By establishing a one-to-one mapping between set indices and loop indices, we produce a new loop permutation.
+The loop permutation is semantically correct because of perfectness.
+In this new order, loops are placed such that as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$.
+At this point, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
+This renders $L_{i_{n-1}}$ no further improvable in the number of operations executed.
+In particular, the number of operations is equal to $\sharp S_i + (\sharp S_i - 1)$, in which the second term represents the cost of the summation.
+\end{proof}
+
+In practice, this result is too weak if we consider generic loop nests in which multilinearity applies to a subset of loops only and reductions are present.
+Consider the example in Figure~\ref{code:loopnest}.
+
+\begin{figure}[h]\begin{CenteredBox}
+\lstinputlisting[basicstyle=\small]{listings/loopnest.code}
+\end{CenteredBox}\caption{Loop nest example}\label{code:loopnest}\end{figure}
+
+The imperfect nest \texttt{N}=\texttt{eijk} comprises a reduction loop \texttt{i} and a perfect doubly nested loop \texttt{jk}, which we assume to be multilinear.
+The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication.
+We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of \texttt{N}, provided that data dependencies are preserved (note that \texttt{N} is imperfect).
+One could think of pre-evaluating the reduction, thus obtaining a decrease proportional to \texttt{M} in the number of operations executed.
+However, finding or exposing such sub-expressions is, in general, challenging.
+Further, even though we assumed we could do so, the following issues should be addressed
 \begin{itemize}
-\item multilinear loop nests, static loop nests; (note: the loop over the elements could also be considered static, if mesh didn't move...)
-\item concept of reuse in a loop;
-\item prove that reuse can always be eliminated from the innermost loop of multilinear loop nests. 
-\item definition of optimality
-\item insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
+\item contrary to what happens with hoisting in perfect multilinear loop nests, in this case the size of the temporaries would be proportional to the number of non-reduction loops (in the example, \texttt{N$\cdot$O} for \texttt{ijk} sub-expressions, \texttt{L$\cdot$N$\cdot$O} for \texttt{eijk} ones).
+This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime;
+\item the process by which multi-invariant sub-expressions are exposed could require transformations like expansion and factorization.
+Consequently, the save originating from the elimination of the reduction loop could be overwhelmed by a potentially larger increase in the number of operations in the body of \texttt{k} (for example, consider how expansion tends to increase operations).
 \end{itemize}
 
-Need to show examples.
+We then review our definition of optimality for generic loop nests as follows
 
-\section{Automated Code Generation of Optimal Multilinear Loop Nests}
+\begin{Def}[Optimality of a loop nest]
+The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the amount of operations performed in the innermost loop is minimal.
+\end{Def}
+
+Clearly, this definition remains vague until instantiated in a certain model.
+Once the model is fixed, we can argue about the meaning of \texttt{C} and try to produce an algorithm to reach optimality.
+This is tackled in the next section.
+
+\section{Synthesis of Optimal Loop Nests in Finite Element Integration}
+The model, or loop nest structure, we are interested in arises in the domain of finite element integration.
+As explained in Section~\ref{sec:background}, integration occurs locally at the element level.
+In a weak variational formulation of a problem, a bilinear and a liner forms need be evaluated, typically using a quadrature scheme.
+The loop nest in Figure~\ref{code:loopnest} is a simplified view of a typical bilinear form implementation.
+The \texttt{e} loop represents iteration over the elements of a mesh; the \texttt{i} loop derives from using numerical quadrature; the perfect loop nest \texttt{jk} implements the computation over the test and trial functions' degrees of freedom.
+
+Domain knowledge tells us that in our model, for real-world scenarios, the following holds: 1) $L >>> M, N, O$; that is, even with high order basis functions, the number of elements is typically order of magnitude larger than both the number of quadrature points and degrees of freedom; 2) the loop nest enclosed by the integration loop is always perfect and multilinear (descends from the straightforward implementation of Equation~\ref{quadrature} into a loop nest).
+Based upon these observations, we now discuss optimality.
+
+\subsection{Operation-minimal synthesis}
+
+\subsection{Memory constraints}
+
+%\item cost model for deciding what to do
+%\item constrained increase in the working set size
+%\item heuristic optimisation of constant and quadrature dependent expressions
+
+%insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
+
+%Need to show examples.
+
+\section{Automated Code Generation}
 
@@
 \item  Building block operations in coffee: 1) eliminating reuse, 2)  insights on the various tree algorithms employed
-\item cost model for deciding what to do
-\item constrained increase in the working set size
-\item heuristic optimisation of constant and quadrature dependent expressions
 \end{itemize}
@@
 
-\subsection{Avoiding Iteration on Zero-valued Blocks by Symbolic Execution}
+\subsection{Avoiding Iteration over Zero-valued Blocks by Symbolic Execution}
 \label{sec:zeros}
@@
 
-Operations over blocks of zero-valued entries in tabulated (derivatives of) basis functions can be skipped.
-Zero-valued columns arise, for example, when taking derivatives on a reference element, or when employing vector function spaces.
-In~\cite{quadrature1}, a technique to avoid iteration over zero-valued columns, based on the use of indirection arrays (e.g.
-\texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), was described and tested in \texttt{FFC}.
-We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}.
-Our strategy differs by avoiding indirection arrays in the generated code.
+% say that we just don't kill al zero-columns and use indirections to not break vectorisation
 
-\begin{figure}[t]
-\tiny
-\centering     
-\subfigure[Original (simplified) code.
-Note the annotation over the definition of the tabulated basis function \texttt{D}, which is used to identify the presence of zero-valued columns]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subfigure[The code after symbolic execution took place]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
-\caption{Simplified excerpt of local assembly code from a Burgers form using vector-valued basis functions, before and after symbolic execution is performed to rewrite the iteration space}\label{fig:skip-code}
-\end{figure}
-
-Consider Figure~\ref{fig:withzeros-code}, an enriched version of the Burgers excerpt in Figure~\ref{fig:invariant-code}.
-The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points.
-There are four zero-valued columns.
-Any multiplications or additions along these columns could (should) be skipped to avoid irrelevant floating point operations.
-The solution adopted in~\cite{quadrature1} is not to generate the zero-valued columns (i.e.
-to generate a dense 6$\times$2 array), to reduce the size of the iteration space over test and trial functions (from 6 to 2) and to use an indirection array (e.g.
-$ind = \lbrace 3, 5\rbrace$) to update the right entries in the element tensor $A$.
-%on the other hand, reshrinki tutto su un array solo...
-
-Our approach is based on using domain knowledge and symbolic execution.
-We discern the origin of zero-valued columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements.
-In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to correctly evaluate the local element matrix while iterating along the space of test and trial functions.
-The two key observations are that 1) the number of zero-valued columns caused by using vector function spaces is, often, much larger then that due to derivatives, and 2) such columns are contiguous in memory.
-Based on this observation, we aim at avoiding iteration only along the block of zero-valued columns induced by mixed (vector) elements.
-
-%Our example is then transformed as in Figure~\ref{fig:withzeros-skipped-code}: loop bounds are adjusted and suitable offsets are introduce to access the element matrix and basis function arrays.
-In general, the element matrix evaluation may also have to be split over multiple iteration spaces (test and trial functions loops), each iteration space characterized by its own loop bounds; this has the side effect of both increasing loop overhead and decreasing data locality.
-
-We achieve our goal by means of symbolic execution.
-The Expression Rewriter expects some indication about the location of the zero-valued columns induced by mixed (vector) function spaces, for each tabulated basis function.
-This indication comes either in the form of code annotation, if the input to COFFEE were provided as pure C (as shown in Figure~\ref{fig:withzeros-code}), or by suitably decorating basis functions nodes, if the input were an abstract syntax tree.
-Then, the rewrite rules are applied and each statement is executed symbolically.
-For example, consider the assignment \texttt{T2[r] = d*D[i][k]+e*E[i][k]} in Figure~\ref{fig:withzeros-code}.
-Array \texttt{D} has non-zero-valued columns in the range $NZ_D=[3,5]$; we also assume array \texttt{E} has non-zero-valued columns in the range $NZ_E=[0,2]$.
-Multiplications by scalar values do not affect the propagation of non-zero-valued columns.
-On the other hand, when symbolically executing the sum of the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we track that the target identifier \texttt{T2} will have non-zero-valued columns in the range $NZ_E \| NZ_D=[0,5]$.
-Eventually, exploiting the $NZ$ information computed and associated with each identifier, we split the original assembly expression into multiple sets of sub-expressions, each set characterized by the same range of non-zero-valued columns.
-In our example, assuming that $NZ_{T1}=[3,5]$ and $NZ_A=[3,5]$, there are two of such sets, which leads to the generation of two distinct iteration spaces (one for each set), as in Figure~\ref{fig:withzeros-skipped-code}.
-
-%TODO : rephrase: " This prevents, among the various optimizations, effective SIMD vectorization, because memory loads and store would eventually reference non-contiguous locations. "
+% generic engine to remove zeros. Can be rows (RT elements), columns (derivatives, vector elements), blocks (vector elements, "TC mode"). Based on symbolic execution.
+% rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
+% tables shrunk and merged
 

### 2015-07-28 TOMS_Paper/assembly.tex
Going on adding stuff to optimal fem synthesis

@@
 
-We summarize the basic concepts sustaining the finite element method following the notation adopted in~\cite{quadrature1} and~\cite{Francis}.
+We review finite element integration and possible implementations using notation and examples adopted in~\cite{quadrature1} and~\cite{Francis}.
+
 We consider the weak formulation of a linear variational problem
@@
 \end{equation}
-where $a$ and $L$ are called bilinear and linear form, respectively.
+where $a$ and $L$ are, respectively, a bilinear and a linear form.
 The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces.
@@
 \begin{equation}
-\label{quadrature}
+\label{eq:quadrature}
 \scriptsize
@@
 
-\subsection{Tensor Mode}
-...
+\subsection{Tensor Contraction Mode}
+Starting from Equation~\ref{eq:quadrature}, exploiting basic mathematical properties we can rewrite the expression as
+\begin{equation}
+\label{eq:tensor}
+\scriptsize
+A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
+\end{equation}
+A generalization of this transformation has been proposed in~\cite{Kirby:TC}.
+By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in a temporary.
+The evaluation of the local tensor can then be abstracted as
+\begin{equation}
+A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
+\end{equation}
+in which the pre-evaluated "reference tensor" $A_{i_1 i_2 \alpha}$ and the cell-dependent "geometry tensor" $G_{K}^\alpha$ are exposed.
+
+\subsection{Qualitative Comparison}
+Depending on the form being considered, the relative performance of the two modes, in terms of number of operations executed, can vary even quite dramatically.
+The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode increasingly more indicate for "complex" forms.
+On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains "sufficiently small".
+
+These two modes have been implemented in the Fenics Form Compiler.
+In this compiler, a simple heuristic is used to choose the most suitable mode for a given form.
+It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically~\cite{Fenics}.
+We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-syntesis}.
+For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of selecting the optimal mode at the monomials level -- no heuristics, no "global" choice of the mode for all monomials in the form -- without affecting the cost of code generation.
 
@@
 \begin{Def}[Linear loop]
-A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if all expressions appearing in the body of $L$ that use $i$ to access some memory locations are linear in $I$.
+A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if all expressions appearing in the body of $L$ that use $i$ to access some memory locations are linear over $I$.
 \end{Def}
@@
 \label{prop:share-removal}
-Sharing in a perfect multilinear loop nest $N = L_{i_{0}, i_{1}, ..., i_{n-1}}$ can always be eliminated.
+Sharing in a perfect multilinear loop nest $N = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ can always be eliminated.
 \end{Prop}
@@
 \begin{Def}[Optimality of a multilinear loop nest]
-The synthesis of a multilinear loop nest is \textit{optimal} if the amount of operations performed in the innermost loop is minimal.
+The synthesis of a multilinear loop nest is \textit{optimal} if the amount of operations performed in the innermost loop is minimum.
 \end{Def}
@@
 
-The imperfect nest \texttt{N}=\texttt{eijk} comprises a reduction loop \texttt{i} and a perfect doubly nested loop \texttt{jk}, which we assume to be multilinear.
+The imperfect nest $N=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
 The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication.
@@
 \item the process by which multi-invariant sub-expressions are exposed could require transformations like expansion and factorization.
-Consequently, the save originating from the elimination of the reduction loop could be overwhelmed by a potentially larger increase in the number of operations in the body of \texttt{k} (for example, consider how expansion tends to increase operations).
+Consequently, the save originating from the elimination of the reduction loop could be overwhelmed by a potentially larger increase in the number of operations in the body of $L_k$ (for example, consider how expansion tends to increase operations).
 \end{itemize}
@@
 \begin{Def}[Optimality of a loop nest]
-The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the amount of operations performed in the innermost loop is minimal.
+\label{def:optimality}
+The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the amount of operations performed in the innermost loop is minimum.
 \end{Def}
@@
 \section{Synthesis of Optimal Loop Nests in Finite Element Integration}
-The model, or loop nest structure, we are interested in arises in the domain of finite element integration.
-As explained in Section~\ref{sec:background}, integration occurs locally at the element level.
-In a weak variational formulation of a problem, a bilinear and a liner forms need be evaluated, typically using a quadrature scheme.
-The loop nest in Figure~\ref{code:loopnest} is a simplified view of a typical bilinear form implementation.
-The \texttt{e} loop represents iteration over the elements of a mesh; the \texttt{i} loop derives from using numerical quadrature; the perfect loop nest \texttt{jk} implements the computation over the test and trial functions' degrees of freedom.
+\label{sec:optimal-syntesis}
 
-Domain knowledge tells us that in our model, for real-world scenarios, the following holds: 1) $L >>> M, N, O$; that is, even with high order basis functions, the number of elements is typically order of magnitude larger than both the number of quadrature points and degrees of freedom; 2) the loop nest enclosed by the integration loop is always perfect and multilinear (descends from the straightforward implementation of Equation~\ref{quadrature} into a loop nest).
-Based upon these observations, we now discuss optimality.
+In this section, we instantiate our definition of loop optimality in the domain of finite element integration.
+This requires elaborating on the junction between two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem, which we reviewed in Section~\ref{sec:background}; and the (partly multilinear) loop nests implementing such forms.
 
-\subsection{Operation-minimal synthesis}
+Our point of departure is the example loop nest in Figure~\ref{code:loopnest}.
+This loop nest is a simplified view of a typical bilinear form implementation.
+The $L_e$ loop represents iteration over the elements of a mesh; the $L_i$ loop derives from using numerical quadrature; the perfect loop nest $[L_j, L_k]$ implements the computation over the test and trial functions' degrees of freedom.
+We deliberately omitted useless portions of code to not hinder readability (e.g.
+matrix insertion) and to avoid tying our discussing to a specific form (e.g.
+\texttt{F} is unspecified).
+
+Before expanding on optimality, we make a few observations about properties of our domain.
+1) We know that $L >>> M, N, O$; that is, even with high order basis functions, the number of elements is typically order of magnitude larger than both the number of quadrature points and degrees of freedom.
+2) The loop nest enclosed by the integration loop is always perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
+
+% maybe say somewhere the order of the loop nest eijk is just the most natural --- the other present worse data locality for example
+% change L into some other letters
+% say somewhere that N and O are the number of dofs for test and trial functions
 
 \subsection{Memory constraints}
+The fact that the iteration space of $L_e$ is so larger than that of other loops suggests that we should be cautious when hoisting outside of the loop nest $N$.
+Imagine an additional loop $L_t$, the time stepping loop, wraps $N$.
+Theoretically, one could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and hoist them out of $L_e$.
+Unless adopting complex engineering solutions (e.g.
+aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by a factor $L$.
+It is our opinion, therefore, that the decrease in the number of operations would be completely overwhelmed by the bigger memory pressure.
+
+A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of "some level of local memory" (e.g.
+the last level of private cache on a conventional CPU, the shared memory on a GPU).
+We will provide precise details about this in the following sections.
+For the moment, and just as one of many possible examples, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporaries of size equal to $N\cdot O$; depending on $N$, $O$ and the cardinality of such set (or, equivalently, "the rank and size of the geometry tensor"), it is not so unlikely to eventually break the local memory threshold.
+
+Based upon these considerations, we impose two memory constraints
+\begin{itemize}
+\item The size of a temporary due to code motion cannot be bigger than the size of the multilinear loop nest iteration space ($N\cdot O$ for the bilinear form in the example).
+A corollary is that hoisting of expressions involving geometry terms outside of $L_e$ is forbidden.
+\item The total size of the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
+\end{itemize}
+
+\subsection{Operation-minimum synthesis}
+Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the number of operations in the innermost loop is minimum.
+We now discuss how we can systematically achieve this.
+
+Eliminating sharing from the multilinear loop nest does not suffice to reach nest optimality.
+This is because we are considering the whole nest, so the presence of the integration loop $L_i$ makes us wonder whether the reduction it imposes can be pre-evaluated outside of $L_e$, as already suggested in~\ref{def:optimality}, thus reducing the operation count.
+
+To answer this question, we make use of a result in~\cite{Kirby:TC}, the foundation of tensor contraction mode.
+Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the domain of product (of derivatives) of functions from discrete spaces; such monomials can always be reduced to the product of two tensor, a reference tensor and a geometry tensor.
+An interpretation of this result at the level of loop nest is that if our input is, conceptually, as in Figure~\ref{code:loopnest}, then we can always dissect \texttt{F} into distinct sub-expressions (one per monomial), refactorize each sub-expression to split constant from $[L_i, L_j, L_k]$-dependent terms, hoist the latter one outside of $L_e$, pre-evaluate the reduction.
 
 %\item cost model for deciding what to do
-%\item constrained increase in the working set size
 %\item heuristic optimisation of constant and quadrature dependent expressions

### 2015-07-28 TOMS_Paper/assembly.tex
Going on adding stuff to optimal fem synthesis

@@
 The fact that the iteration space of $L_e$ is so larger than that of other loops suggests that we should be cautious when hoisting outside of the loop nest $N$.
-Imagine an additional loop $L_t$, the time stepping loop, wraps $N$.
+Imagine a time stepping loop $L_t$ wraps $N$.
 Theoretically, one could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and hoist them out of $L_e$.
@@
 
-Eliminating sharing from the multilinear loop nest does not suffice to reach nest optimality.
-This is because we are considering the whole nest, so the presence of the integration loop $L_i$ makes us wonder whether the reduction it imposes can be pre-evaluated outside of $L_e$, as already suggested in~\ref{def:optimality}, thus reducing the operation count.
+Eliminating sharing from the multilinear loops does not suffice to reach nest optimality.
+In fact, we wonder whether the reduction imposed by $L_i$ could be pre-evaluated outside of $N$, as already suggested in~\ref{def:optimality}, thus potentially reducing the operation count.
 
-To answer this question, we make use of a result in~\cite{Kirby:TC}, the foundation of tensor contraction mode.
-Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the domain of product (of derivatives) of functions from discrete spaces; such monomials can always be reduced to the product of two tensor, a reference tensor and a geometry tensor.
-An interpretation of this result at the level of loop nest is that if our input is, conceptually, as in Figure~\ref{code:loopnest}, then we can always dissect \texttt{F} into distinct sub-expressions (one per monomial), refactorize each sub-expression to split constant from $[L_i, L_j, L_k]$-dependent terms, hoist the latter one outside of $L_e$, pre-evaluate the reduction.
+To answer this question, we make use of a result -- the foundation of tensor contraction mode -- in~\cite{Kirby:TC}.
+Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to the product of two tensor, a reference tensor and a geometry tensor.
+This result can be interpreted at the loop nest level: if our input is, conceptually, as in Figure~\ref{code:loopnest}, then we can always dissect \texttt{F} into distinct sub-expressions (the monomials), factorize each sub-expression to split constant from $[L_i, L_j, L_k]$-dependent terms, hoist the latter ones outside of $N$, pre-evaluate the reduction.
+
+The intuition of the algorithm to synthesize an optimal loop nest is shown in Figure~\ref{code:intuition}.
+
+\begin{figure}[h]
+\begin{lstlisting}[basicstyle=\small, frame=single]
+M = dissect the input expression into monomials
+for each monomial in M:
+    predict the cost of 
+    calculate the cost of reduction pre-evaluation 
+\end{lstlisting}
+\caption{Intuition of the main algorithm}
+\label{code:intuition}
+\end{figure}
 

### 2015-07-29 TOMS_Paper/assembly.tex
Trying to formalize optimality

@@
 \usepackage{fancybox}
+
+% font in listings
+\usepackage{courier}
 
@@
 \usepackage{listings}
-\lstset{language=C, breaklines=true}
+\lstset{language=C, breaklines=true, mathescape}
 
@@
 \begin{Prop}
+\label{prop:optimal-mln}
 An optimal synthesis for a perfect multilinear loop nest $N$ can be found in ....
@@
 This renders $L_{i_{n-1}}$ no further improvable in the number of operations executed.
-In particular, the number of operations is equal to $\sharp S_i + (\sharp S_i - 1)$, in which the second term represents the cost of the summation.
+In particular, the number of operations is equal to $\sharp S_{i_{n-1}} + (\sharp S_{i_{n-1}} - 1)$, in which the second term represents the cost of the summation.
 \end{proof}
@@
 \begin{figure}[h]\begin{CenteredBox}
-\lstinputlisting[basicstyle=\small]{listings/loopnest.code}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
 \end{CenteredBox}\caption{Loop nest example}\label{code:loopnest}\end{figure}
@@
 \begin{itemize}
-\item contrary to what happens with hoisting in perfect multilinear loop nests, in this case the size of the temporaries would be proportional to the number of non-reduction loops (in the example, \texttt{N$\cdot$O} for \texttt{ijk} sub-expressions, \texttt{L$\cdot$N$\cdot$O} for \texttt{eijk} ones).
+\item as opposed to what happens with hoisting in perfect multilinear loop nests, in this case the size of the temporaries would be proportional to the number of non-reduction loops (in the example, \texttt{N$\cdot$O} for \texttt{ijk} sub-expressions, \texttt{L$\cdot$N$\cdot$O} for \texttt{eijk} ones).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime;
@@
 \label{def:optimality}
-The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the amount of operations performed in the innermost loop is minimum.
+The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the amount of operations performed in all innermost loops of the nest is minimum.
 \end{Def}
@@
 To answer this question, we make use of a result -- the foundation of tensor contraction mode -- in~\cite{Kirby:TC}.
-Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to the product of two tensor, a reference tensor and a geometry tensor.
-This result can be interpreted at the loop nest level: if our input is, conceptually, as in Figure~\ref{code:loopnest}, then we can always dissect \texttt{F} into distinct sub-expressions (the monomials), factorize each sub-expression to split constant from $[L_i, L_j, L_k]$-dependent terms, hoist the latter ones outside of $N$, pre-evaluate the reduction.
+Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}).
+We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
+Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $N$ and finally pre-evaluated.
+As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
+In the following, we simply refer to this special sort of code hoisting as "\textit{pre-evaluation}".
 
@@
 
+% local optimization of monomials
+% local sharing elimination
+% global sharing elimination
+% we exclude pathological cases in which different monomials lead to pre-evaluating the same tables
+
 \begin{figure}[h]
-\begin{lstlisting}[basicstyle=\small, frame=single]
+\begin{lstlisting}[basicstyle=\small\ttfamily, frame=single]
 M = dissect the input expression into monomials
 for each monomial in M:
-    predict the cost of 
-    calculate the cost of reduction pre-evaluation 
+  C = estimate $\mbox{\texttt{ops}}$ after pre-evaluation
+  NC = estimate $\mbox{\texttt{ops}}$ without pre-evaluation
+  if C < NC and memory constraints satisfied:
+    extract the monomial into a separate loop nest
+    apply pre-evaluation
+for each expression:
+  remove sharing    
 \end{lstlisting}
@@
 
+The point of departure consists of understanding the impact, as number of operations performed ($\mbox{\texttt{ops}}$, in the following), of pre-evaluation.
+This is studied "locally"; that is, for each monomial, in isolation.
+We will later discuss the optimality of this approach and how we can determine $\mbox{\texttt{ops}}$ exploiting simple mathematical properties.
+
+%now show that local reasoning is optimal in our context, while we postpone the discussion about the variation in \texttt{ops}. 
+%For the moment, just note that if we exclude pathological cases in which different monomials reduce to multiple identical tables, local reasoning satisfies our needs.
+
+If we estimate that, for a given monomial, pre-evaluation will decrease the workload, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}, and the evaluation eventually performed.
+The result is a set of $n$-dimensional tables (these can be seen as "slices" of the reference tensor at the math level), $n$ being the arity of the multilinear form.
+Identical tables are mapped to the same temporary.
+Eventually, sharing is removed in all resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
+
+The transformed, operation-minimum loop nest is as in Figure~\ref{code:loopnest-opt}.
+
+\begin{figure}[h]\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
+\end{CenteredBox}\caption{Optimized Loop nest example}\label{code:loopnest-opt}\end{figure}
+
+Before elaborating on how we determine $C$ and $NC$ of Figure~\ref{code:intuition}, we introduce the following proposition about the optimality of this approach.
+
+\begin{Prop}
+Consider an expression comprising a set of monomials $M$.
+Let $P$ be the set of all monomials for which it has been estimated that pre-evaluation decreases their operation count.
+Then, if pre-evaluation is applied to all of the monomials in $P$, the resulting loop nest, enclosing an expression which is the sum of all monomials in $Z = M \setminus P$, is optimal under memory constraints $C$ once sharing is removed.
+\end{Prop}
+\begin{proof}
+We start noting that $C$ ensures the only hoisting operations allowed are those inherent to both pre-evaluation and sharing.
+This implies that if we 1) optimally decide what should be pre-evaluated and 2) remove sharing, no better synthesis can be found.
+We then need to prove that following: A) ignoring the prediction cost and pre-evaluating $Z_P: Z_P \subseteq Z$ can only increase the operation count; B) ignoring the prediction code and not pre-evaluating $P_Z: P_Z \subseteq P$ can only increase the operation count.
+That is, local analysis of monomials leads to an optimal synthesis.
+
+We observe that the cost of the loop nest when pre-evaluating only all monomials $P$ is $...$.
+
+A) If there is no sharing between terms in $Z_P$ and $Z \setminus Z_P$, then the statement is trivially true.
+Conversely, if present, sharing can be removed.
+However, the operation count would still be larger than when pre-evaluating $Z_P$.
+In fact, the cost of the loop nest would be $Z_P^{pre} + P + \sharp L_i \cdot Z \setminus Z_P$, which is greater than $...$, since $Z_P^{pre} > \sharp L_i \cdot Z_P$.
+
+B) ...
+
+\end{proof}
+%This minimizes, according to our definition of optimality, the operation count of the multilinear loop nest.
+
+We now only need to tie our loose hand...
+
 %\item cost model for deciding what to do
@@
 
+% apply memory constraints: reduction to knapsack problem ... not implemented, we use greedy
+
+\subsection{Heuristic Optimization of Integration-dependent Expressions}
 %insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
 
-%Need to show examples.
-
 \section{Automated Code Generation}
+\label{sec:codegen}
 
@@
 ...
+
+% MAYBE JUST A SUBSECTION, "INSIGHTS OF LOW-LEVEL OPTS ...."
 

### 2015-07-29 TOMS_Paper/assembly.tex
Minor fixes

@@
 
-\subsection{Operation-minimum synthesis}
-Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the number of operations in the innermost loop is minimum.
+\subsection{Operation Minimization under Memory Constraints}
+Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the number of operations in all innermost loops is minimum.
 We now discuss how we can systematically achieve this.
 
-Eliminating sharing from the multilinear loops does not suffice to reach nest optimality.
+Eliminating sharing from the multilinear loops does not suffice for nest optimality.
 In fact, we wonder whether the reduction imposed by $L_i$ could be pre-evaluated outside of $N$, as already suggested in~\ref{def:optimality}, thus potentially reducing the operation count.
 
-To answer this question, we make use of a result -- the foundation of tensor contraction mode -- in~\cite{Kirby:TC}.
+To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}.
 Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}).
 We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $N$ and finally pre-evaluated.
+Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $N$, and finally pre-evaluated.
 As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
@@
 Identical tables are mapped to the same temporary.
-Eventually, sharing is removed in all resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
+Eventually, sharing is removed from all resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
 

### 2015-07-30 TOMS_Paper/assembly.tex
Trying to prove optimality, pfff

@@
 
-The imperfect nest $N=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
+The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
 The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication.
@@
 \label{def:optimality}
-The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the amount of operations performed in all innermost loops of the nest is minimum.
+The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
 \end{Def}
@@
 \subsection{Memory constraints}
-The fact that the iteration space of $L_e$ is so larger than that of other loops suggests that we should be cautious when hoisting outside of the loop nest $N$.
-Imagine a time stepping loop $L_t$ wraps $N$.
+\label{sec:mem-const}
+The fact that the iteration space of $L_e$ is so larger than that of other loops suggests that we should be cautious when hoisting outside of $LN$.
+Imagine a time stepping loop $L_t$ wraps $LN$.
 Theoretically, one could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and hoist them out of $L_e$.
@@
 We will provide precise details about this in the following sections.
-For the moment, and just as one of many possible examples, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporaries of size equal to $N\cdot O$; depending on $N$, $O$ and the cardinality of such set (or, equivalently, "the rank and size of the geometry tensor"), it is not so unlikely to eventually break the local memory threshold.
+For the moment, and just as one of many possible examples, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporaries of size equal to $N\cdot O$; depending on the discretization employed it is not so uncommon to break the local memory threshold.
 
@@
 We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $N$, and finally pre-evaluated.
+Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated.
 As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
@@
 
-The point of departure consists of understanding the impact, as number of operations performed ($\mbox{\texttt{ops}}$, in the following), of pre-evaluation.
+The point of departure consists of understanding the impact, as number of operations saved or introduced, of pre-evaluation.
 This is studied "locally"; that is, for each monomial, in isolation.
-We will later discuss the optimality of this approach and how we can determine $\mbox{\texttt{ops}}$ exploiting simple mathematical properties.
+We will later discuss the optimality of this approach and how we can precisely quantify such impact exploiting simple mathematical properties.
 
@@
 
-The transformed, operation-minimum loop nest is as in Figure~\ref{code:loopnest-opt}.
+The transformed, optimal loop nest is as in Figure~\ref{code:loopnest-opt}.
 
@@
 
-Before elaborating on how we determine $C$ and $NC$ of Figure~\ref{code:intuition}, we introduce the following proposition about the optimality of this approach.
+Before elaborating on how we determine the pre-evaluation gain (or loss), we demonstrate the optimality of the approach.
+%We introduce the following notation ...
 
@@
 Let $P$ be the set of all monomials for which it has been estimated that pre-evaluation decreases their operation count.
-Then, if pre-evaluation is applied to all of the monomials in $P$, the resulting loop nest, enclosing an expression which is the sum of all monomials in $Z = M \setminus P$, is optimal under memory constraints $C$ once sharing is removed.
+Then, if pre-evaluation is applied to all of the monomials in $P$, the resulting loop nest $LN$, enclosing an expression which is the sum of all monomials in $Z = M \setminus P$, is optimal under memory constraints $C$ once sharing is removed.
 \end{Prop}
 \begin{proof}
-We start noting that $C$ ensures the only hoisting operations allowed are those inherent to both pre-evaluation and sharing.
-This implies that if we 1) optimally decide what should be pre-evaluated and 2) remove sharing, no better synthesis can be found.
-We then need to prove that following: A) ignoring the prediction cost and pre-evaluating $Z_P: Z_P \subseteq Z$ can only increase the operation count; B) ignoring the prediction code and not pre-evaluating $P_Z: P_Z \subseteq P$ can only increase the operation count.
-That is, local analysis of monomials leads to an optimal synthesis.
+We distinguish two classes of loop nests rooted in $LN$: $[L_e, L_j, L_k]$ for the pre-evaluated expressions in $P$, and $[L_e, L_i, L_j, L_k]$ enclosing the remaining monomials in $Z$.
+Since they only differer for the presence of $L_i$, we relieve the notation by omitting all shared loops when discussing operation counts.
+In particular, we use $I$ to refer to the iteration space size of $L_i$.
+The operation count of what we are proving to be the optimal $LN$ synthesis is, therefore, $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation cost of monomials in $P$ and $Z$.
 
-We observe that the cost of the loop nest when pre-evaluating only all monomials $P$ is $...$.
+We start noting that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting.
+This allows us to search the optimum in the syntheses space by only 1) selecting candidates for pre-evaluation and 2) dealing with sharing.
+We need to prove that reasoning about the operation count on a monomial basis, rather than globally, suffices to reach our definition of optimality.
+We reduce this problem to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $LN_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $LN_{ops}$.
+That is, we show there is no better way of constructing $P$ and $Z$ other than local reasoning.
 
-A) If there is no sharing between terms in $Z_P$ and $Z \setminus Z_P$, then the statement is trivially true.
-Conversely, if present, sharing can be removed.
-However, the operation count would still be larger than when pre-evaluating $Z_P$.
-In fact, the cost of the loop nest would be $Z_P^{pre} + P + \sharp L_i \cdot Z \setminus Z_P$, which is greater than $...$, since $Z_P^{pre} > \sharp L_i \cdot Z_P$.
+A) We prove that $LN_{ops}' = LN_{ops_1}' + LN_{ops_2}' > LN_{ops}$.
+It is rather obvious that $LN_{ops_1}' \geq LN_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
+We note that if all symbols in $Z_P$ are also shared in $\overline{Z} = Z \setminus Z_P$,  then we have $LN_{ops_2}' = LN_{ops_2}$, so out statement is true.
+If, on the other hand, at least one symbol is not shared, we obtain $LN_{ops_2}' < LN_{ops_2}$ or, equivalently, $LN_{ops_2}' = LN_{ops_2} - I \cdot\delta$.
+What we have to show now is that even by exposing more pre-evaluated terms, $LN_{ops_1}' \geq LN_{ops_1} +  I \cdot\delta$ holds.
+This is the case if and only if pre-evaluated terms are not identical to others in $P$, which would otherwise expose sharing.
+However, if we exclude the pathological cases in which different sub-expressions reduce to the same values (due to symmetries in basis functions, for example), this must be true, otherwise the assumption about the optimality of the cost prediction would be contradicted.
 
-B) ...
+B) This holds due to a simple observation.
+In the absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$ and we are assuming that pre-evaluation was optimal for the monomials at hand.
+In the worst case, instead, all non pre-evaluated monomials share terms with those in $Z$.
+Anyway, the best we can do
+
+%: regardless of whether there is sharing between terms in $Z_P$ and $\overline{Z} = Z \setminus Z_P$, then the statement is trivially true, since we are assuming that $Z$ contains all monomials for which pre-evaluation is sub-optimal.
+If, on the other hand, $\delta \in Z_P \cap \overline{Z}$, one can reason as follows.
+We can first remove sharing; this makes  to reach local optimality and we compare the operation counts.
+the cost of the loop nest would be $Z_P^{pre} + P + \sharp L_i \cdot Z \setminus Z_P$, which is greater than $...$, since $Z_P^{pre} > \sharp L_i \cdot Z_P$.
 
 \end{proof}
-%This minimizes, according to our definition of optimality, the operation count of the multilinear loop nest.
 

### 2015-07-30 TOMS_Paper/assembly.tex
Before refactoring to global reasoning

@@
 
-The intuition of the algorithm to synthesize an optimal loop nest is shown in Figure~\ref{code:intuition}.
+The intuition of the main algorithm for optimal loop nest synthesis is shown in Figure~\ref{code:intuition}.
 
@@
 Before elaborating on how we determine the pre-evaluation gain (or loss), we demonstrate the optimality of the approach.
-%We introduce the following notation ...
 
@@
 
-B) This holds due to a simple observation.
-In the absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$ and we are assuming that pre-evaluation was optimal for the monomials at hand.
-In the worst case, instead, all non pre-evaluated monomials share terms with those in $Z$.
-Anyway, the best we can do
+B) In the absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$ and we are starting from the assumption that pre-evaluation is optimal for the monomials considered.
+As opposed to this extreme case, there could be some degree of sharing between monomials in $P_Z$ an $\overline{P}$.
+However, as clarified in the next section, this situation is not possible: the cost function "sees sharing" and it would prevent pre-evaluation by counting on sharing-removal for optimality.
+Finally, partial sharing.
+Note that by not pre-evaluating $P_Z$, the following: $LN_{ops_1} - LN_{ops_1}' < LN_{ops_2}' - LN_{ops_2}$ trivially holds (i.e., the loss in operation count is larger than the gain).
+At best, once sharing is remove,
+\end{proof}
 
-%: regardless of whether there is sharing between terms in $Z_P$ and $\overline{Z} = Z \setminus Z_P$, then the statement is trivially true, since we are assuming that $Z$ contains all monomials for which pre-evaluation is sub-optimal.
-If, on the other hand, $\delta \in Z_P \cap \overline{Z}$, one can reason as follows.
-We can first remove sharing; this makes  to reach local optimality and we compare the operation counts.
-the cost of the loop nest would be $Z_P^{pre} + P + \sharp L_i \cdot Z \setminus Z_P$, which is greater than $...$, since $Z_P^{pre} > \sharp L_i \cdot Z_P$.
-
-\end{proof}
+\subsection{Operation Counting}
 

### 2015-07-30 TOMS_Paper/assembly.tex
Came up with a first, sketchy proof

@@
 
-\section{Introduction and Motivations}
+\section{Introduction, Motivations, Related Work}
 
@@
 
-In this work, we are particularly interested in the following class since it naturally arises from the math described in~\ref{sec:background}.
+In this work, we are particularly interested in the following class since it naturally arises from the math described in Section~\ref{sec:background}.
 
 \begin{Def}[Perfect multilinear loop nest]
-A \textit{perfect multilinear} loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $l \in [0, n-1]$.
+A \textit{perfect multilinear} loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
 \end{Def}
 
-Note that nothing prevents a perfect multilinear loop nest to be enclosed by an outer, possibly imperfect, nest.
-Indeed, this is the loop structure that will be at the centre of our study.
+Note that nothing prevents a perfect multilinear loop nest from being enclosed by an outer, possibly imperfect, nest.
+Indeed, this loop structure will be at the centre of our study.
 We will later elaborate on this.
-First, we need to formulate a definition of optimality for perfect multilinear loop nests.
+First, we need to define optimality for perfect multilinear loop nests.
 To this purpose, it is convenient to introduce the notion of sharing.
@@
 \label{prop:share-removal}
-Sharing in a perfect multilinear loop nest $N = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ can always be eliminated.
+Sharing in a perfect multilinear loop nest $LN = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ can always be eliminated.
 \end{Prop}
@@
 The demonstration is by construction and exploits linearity.
-We want to transform $N$ into $N'$ such that sharing is absent at all levels of $N'$.
-Starting from the innermost loop $n-1$, the expressions are ''flattened'' by expanding all products involving any of the terms depending on $i_{n-1}$.
-Subsequently, such terms can be factorized.
-Due to linearity, in each factored product there is only one term depending on $i_{n-1}$.
-The factors that do not depend on $i_{n-1}$ are, by definition, loop-invariant, and as such can be hoisted at the level of $i_{n-2}$.
+We want to transform $LN$ into $LN'$ such that there is no sharing is any $L_i \in LN'$.
+Starting from the innermost loop $L_{i_{n-1}}$, the expressions are ''flattened'' by expanding all products involving terms depending on $L{i_{n-1}}$.
+Being on the same level of the expression tree, such terms can then be factorized.
+Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$.
+The other terms, independent of $L_{i_{n-1}}$, are, by definition, loop-invariant, and as such can be hoisted at the level of $L_{i_{n-2}}$.
 This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is always safe.
@@
 
-In other words, this implies there is no other nest syntheses that could decrease the number of operations in the body of the innermost loop any further.
+In other words, this means there is no other nest synthesis able to decrease the number of operations in the innermost loop any further.
 Note that this definition of optimality does not take into account memory requirements.
-If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' clearly makes no sense.
+If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense.
 In the following, we assume to operate in a CPU-bound regime, in which the body of loop nests are characterized by arithmetic-intensive expressions.
@@
 \label{prop:optimal-mln}
-An optimal synthesis for a perfect multilinear loop nest $N$ can be found in ....
+An optimal synthesis for a perfect multilinear loop nest $LN$ can be found in ....
 \end{Prop}
@@
 By construction.
-We consider the body of the innermost loop $L_{i_{n-1}}$.
-Constant expressions can be hoisted outside of $N$ and be replaced with suitable temporaries.
-Loop-dependent terms are grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on loop $i$.
+We start from the innermost loop $L_{i_{n-1}}$.
+Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$.
 These sets are sorted in descending order based on their cardinality.
@@
 
-In practice, this result is too weak if we consider generic loop nests in which multilinearity applies to a subset of loops only and reductions are present.
-Consider the example in Figure~\ref{code:loopnest}.
+We now consider generic loop nests in which multilinearity applies to a subset of loops only. Consider the example in Figure~\ref{code:loopnest}.
 
@@
 The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication.
-We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of \texttt{N}, provided that data dependencies are preserved (note that \texttt{N} is imperfect).
-One could think of pre-evaluating the reduction, thus obtaining a decrease proportional to \texttt{M} in the number of operations executed.
-However, finding or exposing such sub-expressions is, in general, challenging.
+We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $LN$ as long as data dependencies are preserved (note that $LN$ is imperfect).
+One could think of pre-evaluating the reduction, thus obtaining a decrease proportional to $M$ in the number of operations executed.
+However, finding or exposing reducible sub-expressions is, in general, challenging.
 Further, even though we assumed we could do so, the following issues should be addressed
 \begin{itemize}
-\item as opposed to what happens with hoisting in perfect multilinear loop nests, in this case the size of the temporaries would be proportional to the number of non-reduction loops (in the example, \texttt{N$\cdot$O} for \texttt{ijk} sub-expressions, \texttt{L$\cdot$N$\cdot$O} for \texttt{eijk} ones).
+\item as opposed to what happens with hoisting in perfect multilinear loop nests, in this case the size of the temporaries would be proportional to the number of non-reduction loops crossed (in the example, \texttt{N$\cdot$O} for \texttt{ijk} sub-expressions, \texttt{L$\cdot$N$\cdot$O} for \texttt{eijk} ones).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime;
 \item the process by which multi-invariant sub-expressions are exposed could require transformations like expansion and factorization.
-Consequently, the save originating from the elimination of the reduction loop could be overwhelmed by a potentially larger increase in the number of operations in the body of $L_k$ (for example, consider how expansion tends to increase operations).
+Consequently, the save originating from the elimination of the reduction loop could be overwhelmed by a potentially larger increase in the number of operations in the body of $L_k$ (expansion could increase the number of operations, e.g.
+$A(B+C) = AB + AC$).
 \end{itemize}
 
-We then review our definition of optimality for generic loop nests as follows
+We then refine our definition of optimality for generic loop nests as follows
 
@@
 \item The size of a temporary due to code motion cannot be bigger than the size of the multilinear loop nest iteration space ($N\cdot O$ for the bilinear form in the example).
-A corollary is that hoisting of expressions involving geometry terms outside of $L_e$ is forbidden.
+A corollary is that hoisting expressions involving geometry terms outside of $L_e$ is forbidden.
 \item The total size of the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
@@
 
-\subsection{Operation Minimization under Memory Constraints}
+\subsection{Minimizing the Operation Cost}
 Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the number of operations in all innermost loops is minimum.
@@
 \begin{lstlisting}[basicstyle=\small\ttfamily, frame=single]
-M = dissect the input expression into monomials
-for each monomial in M:
+dissect the input expression into monomials
+for each monomial M:
   C = estimate $\mbox{\texttt{ops}}$ after pre-evaluation
@@
   if C < NC and memory constraints satisfied:
-    extract the monomial into a separate loop nest
-    apply pre-evaluation
+    mark M as candidate for pre-evaluation
+for each monomial M:
+  if M does not share terms with M$'$, an unmarked monomial:
+    extract M into a separate loop nest
+    apply pre-evaluation to M
 for each expression:
@@
 This is studied "locally"; that is, for each monomial, in isolation.
-We will later discuss the optimality of this approach and how we can precisely quantify such impact exploiting simple mathematical properties.
-
-%now show that local reasoning is optimal in our context, while we postpone the discussion about the variation in \texttt{ops}. 
-%For the moment, just note that if we exclude pathological cases in which different monomials reduce to multiple identical tables, local reasoning satisfies our needs.
-
-If we estimate that, for a given monomial, pre-evaluation will decrease the workload, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}, and the evaluation eventually performed.
+If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}, and the evaluation eventually performed.
 The result is a set of $n$-dimensional tables (these can be seen as "slices" of the reference tensor at the math level), $n$ being the arity of the multilinear form.
 Identical tables are mapped to the same temporary.
-Eventually, sharing is removed from all resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
+Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
 
-The transformed, optimal loop nest is as in Figure~\ref{code:loopnest-opt}.
+The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.
 
@@
 
-Before elaborating on how we determine the pre-evaluation gain (or loss), we demonstrate the optimality of the approach.
+Before elaborating on the profitability of pre-evaluation, we need to discuss under which conditions this approach, based on a "local analysis" of monomials, is optimal.
 
 \begin{Prop}
+\label{prop:optimal-approach}
 Consider an expression comprising a set of monomials $M$.
-Let $P$ be the set of all monomials for which it has been estimated that pre-evaluation decreases their operation count.
-Then, if pre-evaluation is applied to all of the monomials in $P$, the resulting loop nest $LN$, enclosing an expression which is the sum of all monomials in $Z = M \setminus P$, is optimal under memory constraints $C$ once sharing is removed.
+Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and let $Z$ be such that $Z = M \setminus P$.
+Assume that:
+\begin{enumerate}
+\item the cost function employed is optimal; that is, it predicts correctly the profitability of pre-evaluation
+\item pre-evaluating distinct monomials does not produce identical tables
+\item monomials in $P$ do not share terms
+\end{enumerate} 
+Then, the loop nest $LN$ is optimal under memory constraints $C$, once sharing is removed.
 \end{Prop}
 \begin{proof}
-We distinguish two classes of loop nests rooted in $LN$: $[L_e, L_j, L_k]$ for the pre-evaluated expressions in $P$, and $[L_e, L_i, L_j, L_k]$ enclosing the remaining monomials in $Z$.
-Since they only differer for the presence of $L_i$, we relieve the notation by omitting all shared loops when discussing operation counts.
+We first comment on the assumptions.
+1) How to create an optimal cost function is discussed in Section~\ref{sec:op_count}.
+2) A pathological case due to symmetries in basis functions, which in practice rarely happens.
+3) This could occur in complex forms with several monomials; for simplicity, we ignore this situation (otherwise, a "global" analysis of the monomials would be required).
+
+We distinguish two classes of loop nests rooted in $LN$: $[L_e, L_j, L_k]$, for the pre-evaluated monomials in $P$, and $[L_e, L_i, L_j, L_k]$, enclosing the remaining monomials in $Z$.
+Since they only differ for the presence of $L_i$, we relieve notation by omitting all shared loops when discussing operation counts.
 In particular, we use $I$ to refer to the iteration space size of $L_i$.
-The operation count of what we are proving to be the optimal $LN$ synthesis is, therefore, $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation cost of monomials in $P$ and $Z$.
+The operation count of what we are proving to be the optimal $LN$ synthesis is, therefore, $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation cost of monomials in $P$ and $Z$, respectively.
 
 We start noting that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting.
-This allows us to search the optimum in the syntheses space by only 1) selecting candidates for pre-evaluation and 2) dealing with sharing.
-We need to prove that reasoning about the operation count on a monomial basis, rather than globally, suffices to reach our definition of optimality.
-We reduce this problem to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $LN_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $LN_{ops}$.
-That is, we show there is no better way of constructing $P$ and $Z$ other than local reasoning.
+This narrows the proof to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $LN_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $LN_{ops}$.
 
@@
 It is rather obvious that $LN_{ops_1}' \geq LN_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
-We note that if all symbols in $Z_P$ are also shared in $\overline{Z} = Z \setminus Z_P$,  then we have $LN_{ops_2}' = LN_{ops_2}$, so out statement is true.
-If, on the other hand, at least one symbol is not shared, we obtain $LN_{ops_2}' < LN_{ops_2}$ or, equivalently, $LN_{ops_2}' = LN_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by exposing more pre-evaluated terms, $LN_{ops_1}' \geq LN_{ops_1} +  I \cdot\delta$ holds.
-This is the case if and only if pre-evaluated terms are not identical to others in $P$, which would otherwise expose sharing.
-However, if we exclude the pathological cases in which different sub-expressions reduce to the same values (due to symmetries in basis functions, for example), this must be true, otherwise the assumption about the optimality of the cost prediction would be contradicted.
+We note that if monomials in $Z_P$ share symbols with $\overline{Z} = Z \setminus Z_P$, then we have $LN_{ops_2}' = LN_{ops_2}$, so our statement is true.
+If, on the other hand, at least one monomial does not share any symbols, we obtain $LN_{ops_2}' < LN_{ops_2}$ or, equivalently, $LN_{ops_2}' = LN_{ops_2} - I \cdot\delta$.
+What we have to show now is that even by exposing more pre-evaluations, $LN_{ops_1}' \geq LN_{ops_1} +  I \cdot\delta$ holds.
+This is indeed the case since we rely on assumption 2, which ensures the uniqueness of the pre-evaluated tables (i.e., the absence of sharing) and, therefore, the optimality of $LN$.
 
-B) In the absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$ and we are starting from the assumption that pre-evaluation is optimal for the monomials considered.
-As opposed to this extreme case, there could be some degree of sharing between monomials in $P_Z$ an $\overline{P}$.
-However, as clarified in the next section, this situation is not possible: the cost function "sees sharing" and it would prevent pre-evaluation by counting on sharing-removal for optimality.
-Finally, partial sharing.
-Note that by not pre-evaluating $P_Z$, the following: $LN_{ops_1} - LN_{ops_1}' < LN_{ops_2}' - LN_{ops_2}$ trivially holds (i.e., the loss in operation count is larger than the gain).
-At best, once sharing is remove,
+B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, with the cost function being optimal by assumption.
+Assumption 3 guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities.
+The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$.
+This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are "declassed" from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so aggregating $p$ in $Z$ does not increase the operation count.
 \end{proof}
 
-\subsection{Operation Counting}
-
-We now only need to tie our loose hand...
+\subsection{A-Priori Operation Counting}
+\label{sec:op_count}
+In this section we discuss the optimality of the cost function.
 
@@
 
-\subsection{Heuristic Optimization of Integration-dependent Expressions}
-%insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
-
-\section{Automated Code Generation}
+\section{Code Generation}
 \label{sec:codegen}
@@
 
-\section{Low-level Optimization}
+\subsection{Heuristic Optimization of Integration-dependent Expressions}
+%insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
+
+\subsection{Low-level Optimization}
 \label{sec:code-spec}
@@
 
-\subsection{Avoiding Iteration over Zero-valued Blocks by Symbolic Execution}
+\subsubsection{Avoiding Iteration over Zero-valued Blocks by Symbolic Execution}
 \label{sec:zeros}
@@
 
-\subsection{Padding and Data Alignment}
+\subsubsection{Padding and Data Alignment}
 \label{sec:padding}
-Padding and data alignment as described in~\cite{Luporini} must be refined for the case in which computation over zero-valued columns is avoided.
-We recall effective SIMD (auto-)vectorization can be achieved only if the innermost loop size is a multiple of the vector register length, in which case the compiler needs not to introduce a remainder scalar loop.
-In the case of an AVX instruction set, for example, we want to round the size of the innermost loops to the closest multiple of 4 (AVX registers can fit up to four double-precision floats).
-This can be done provided that arrays are padded, if necessary.
-Moreover, loads and stores instructions are efficient only if their addresses are suitably aligned to cache boundaries; this is obtainable by enforcing the base address of each padded array to be a multiple of the vector length.
+...
 
-Consider again the code in Figure~\ref{fig:withzeros-skipped-code}.
-The arrays in the loop nest \texttt{[j1,k1]} can be padded and the right bound of loop \texttt{k1} can be safely increased to 8: eventually, values computed in the region \texttt{M[0:3][6:8]} will be discarded.
-Then, by explicitly aligning arrays and using suitable pragmas (e.g.
-\texttt{$\#$pragma simd} for the Intel compiler), effective SIMD auto-vectorization can be obtained for this loop nest.
-
-There are some complications in the case of loops \texttt{[j0,k0]}.
-Here, increasing the loop bound to 4 is still safe, assuming that both \texttt{T1} and \texttt{A} are padded with zero-valued entries, but it has no effect: the starting addresses of the load instructions would be \texttt{T1[3]} and \texttt{A[i][3]}, which are not aligned.
-One solution is to start iterating from the closest index that would ensure data alignment: in this specific case, $k0=0$.
-However, this would imply losing (partially in general, totally for this loop nest) the effect of the zero-avoidance transformation.
-Another possibility is to attain to non-aligned accesses.
-COFFEE can generate code for both situations, so we leave the autotuner, described in Section~\ref{sec:autotune}, in charge of determining the optimal transformation.
-
-Note that in some circumstances the previous solution cannot be applied, since extra iterations might end up accessing non-zero entries in the local element matrix \texttt{M}.
-In such a situation, there is no possibility of recovering data alignment.
-
-\subsection{Vector-promotion of Integration Quantities}
-....
+\subsubsection{Vector-promotion of Integration Quantities}
+...
 
@@
 \label{sec:perf-results-forms}
-The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly.
+%The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly.
 We limit ourselves to analyzing the cost of computing element matrices, although all of the techniques presented in this paper are immediately extendible to the evaluation of local vectors.
 As anticipated, in the following we comment speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the non-optimized, FEniCS-Form-Compiler-generated code.
-
-We first comment on results of general applicability.
+%
+%We first comment on results of general applicability.
 By looking at the various figures, we note there is a trend in COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase.
@@
 \paragraph{Helmholtz}
-\begin{figure}[t]  
-...
-%\includegraphics[scale=0.7]{perf-results/helmholtz}
-%\caption{Helmholtz results.}\label{fig:helmholtz}
-\end{figure}
-The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.
+%\begin{figure}[t]  
+%...
+%%\includegraphics[scale=0.7]{perf-results/helmholtz}
+%%\caption{Helmholtz results.}\label{fig:helmholtz}
+%\end{figure}
+%The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.
 We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$.
@@
 \paragraph{Elasticity}
-\begin{figure}[t]  
-...
-%\includegraphics[scale=0.7]{perf-results/elasticity}
-%\caption{Elasticity results.}\label{fig:elasticity}
-\end{figure}
-Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem.
+%\begin{figure}[t]  
+%...
+%%\includegraphics[scale=0.7]{perf-results/elasticity}
+%%\caption{Elasticity results.}\label{fig:elasticity}
+%\end{figure}
+%Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem.
 This form uses a vector-valued space for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance.
@@
 Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase.
-
-For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceed 1$\times$.
+%
+%For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceed 1$\times$.
 On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.
-
-It is worth noting that speed ups of \texttt{auto} over \texttt{fix} decrease at $q=4$, particularly for low values of $p$.
+%
+%It is worth noting that speed ups of \texttt{auto} over \texttt{fix} decrease at $q=4$, particularly for low values of $p$.
 As we will discuss in Section~\ref{sec:auto-analysis}, this is because at $q=4$ the vector-register tiling transformation (in combination with loop unroll-and-jam) leads to the highest performance.
@@
 \paragraph{Poisson}
-\begin{figure}[t]  
-...
-%\includegraphics[scale=0.7]{perf-results/poisson}
-%\caption{Poisson results.}\label{fig:poisson}
-\end{figure}
-
-In Figure~\ref{fig:poisson} we report speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the original code for the Poisson form.
+%\begin{figure}[t]  
+%...
+%%\includegraphics[scale=0.7]{perf-results/poisson}
+%%\caption{Poisson results.}\label{fig:poisson}
+%\end{figure}
+%
+%In Figure~\ref{fig:poisson} we report speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the original code for the Poisson form.
 We note that, as a general trend, \texttt{ffc} exhibits drops in performance as $nf$ increases, notably when $nf=3$, for any values of $q$ and $p$.
@@
 Moreover, the absence of vectorization is another limiting factor.
-
-The \texttt{auto} variant generally shows the best performance.
+%
+%The \texttt{auto} variant generally shows the best performance.
 Significant improvements over \texttt{fix} are also achieved, notably as $q$, $p$ and $nf$ increase.
@@
 \paragraph{Hyperelasticity}
-\begin{figure}[t]  
-...
-%\includegraphics[scale=0.7]{perf-results/hyperelasticity}
-%\caption{Hyperelasticity results.}\label{fig:hyperelasticity}
-\end{figure}
-
-Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 2$ could not be executed because of FEniCS-Form-Compiler's technical limitations.
-
-For \texttt{auto}, massive speed ups for $q \geq 2$ are to be ascribed to aggressive and successful expression writing.
+%\begin{figure}[t]  
+%...
+%%\includegraphics[scale=0.7]{perf-results/hyperelasticity}
+%%\caption{Hyperelasticity results.}\label{fig:hyperelasticity}
+%\end{figure}
+%
+%Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 2$ could not be executed because of FEniCS-Form-Compiler's technical limitations. 
+%
+%For \texttt{auto}, massive speed ups for $q \geq 2$ are to be ascribed to aggressive and successful expression writing.
 Hyperelasticity problems are really compute-intensive, with thousands of operations being performed, so reductions in redundant and useless computation are crucial.
@@
 We expect to obtain considerably faster code once such optimizations will be incorporated.
-
-In the regime $q \geq 2$ and $nf=1$, peformance improvements are less pronounced moving from $p=1$ to $p=2$, although still significant; in particular, we notice a drop at $p=2$, followed by a raise up to $p=4$.
+%
+%In the regime $q \geq 2$ and $nf=1$, peformance improvements are less pronounced moving from $p=1$ to $p=2$, although still significant; in particular, we notice a drop at $p=2$, followed by a raise up to $p=4$.
 It is worth observing that this effect is common to all sets of optimizations.

### 2015-07-31 TOMS_Paper/assembly.tex
Counting function

@@
 
-\section{Introduction, Motivations, Related Work}
+\section{Introduction}
 
@@
 \begin{Def}[Perfect and imperfect loop nests]
-A loop nest is said to be \textit{perfect} when code appears only in the body of the innermost loop.
-Conversely, the presence of code between any pair of loops in a nest renders the loop nest \textit{imperfect}.
+A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said \textit{imperfect}. 
 \end{Def}
 
-A straightforward property of perfect nests is that hoisting of invariant expressions from the innermost loop to the preheader (i.e., the block that precedes the entry point of the nest) is always safe.
-We will later make extensive use of this property.
+A straightforward property of perfect nests is that hoisting invariant expressions from the innermost loop to the preheader (i.e., the block that precedes the entry point of the nest) is a safe transformation.
+We will make use of this property.
 
@@
 \begin{Def}[Linear loop]
-A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if all expressions appearing in the body of $L$ that use $i$ to access some memory locations are linear over $I$.
+A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if all expressions appearing in the body of $L$ that use $i$ to access some memory locations are linear functions over $I$.
 \end{Def}
@@
 
-Note that nothing prevents a perfect multilinear loop nest from being enclosed by an outer, possibly imperfect, nest.
-Indeed, this loop structure will be at the centre of our study.
-We will later elaborate on this.
+Note that nothing prevents a perfect multilinear loop nest from being rooted in a deeper, possibly imperfect loop nest.
+In fact, we will later focus on this particular structure.
 First, we need to define optimality for perfect multilinear loop nests.
@@
 
-At this point, we can present a simple yet fundamental result.
+Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $k$.
+
+\begin{figure}[h]\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}
+\end{CenteredBox}\caption{Multilinear loop nest with sharing}\label{code:multi_loopnest}\end{figure}
+
+We can now present a simple yet fundamental result.
 
@@
 Being on the same level of the expression tree, such terms can then be factorized.
-Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$.
+Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$, and such term is now unique in the expression.
 The other terms, independent of $L_{i_{n-1}}$, are, by definition, loop-invariant, and as such can be hoisted at the level of $L_{i_{n-2}}$.
@@
 
-In other words, this means there is no other nest synthesis able to decrease the number of operations in the innermost loop any further.
-Note that this definition of optimality does not take into account memory requirements.
+In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop.
+Note that this definition does not take into account memory requirements.
 If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense.
@@
 \label{prop:optimal-mln}
-An optimal synthesis for a perfect multilinear loop nest $LN$ can be found in ....
+An optimal synthesis for a perfect multilinear loop nest $LN$ can deterministically be found.
 \end{Prop}
@@
 By construction.
-We start from the innermost loop $L_{i_{n-1}}$.
 Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$.
@@
 In this new order, loops are placed such that as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$.
-At this point, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
-This renders $L_{i_{n-1}}$ no further improvable in the number of operations executed.
-In particular, the number of operations is equal to $\sharp S_{i_{n-1}} + (\sharp S_{i_{n-1}} - 1)$, in which the second term represents the cost of the summation.
+Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
+This renders $L_{i_{n-1}}$ optimal.
+In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of the summation.
 \end{proof}
@@
 We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $LN$ as long as data dependencies are preserved (note that $LN$ is imperfect).
-One could think of pre-evaluating the reduction, thus obtaining a decrease proportional to $M$ in the number of operations executed.
-However, finding or exposing reducible sub-expressions is, in general, challenging.
-Further, even though we assumed we could do so, the following issues should be addressed
+One could think of searching for sub-expressions independent of $L_e$, for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
+However, finding or exposing such reducible sub-expressions requires, in general, a full exploration of the expression tree transformation space.
+This is obviously challenging, and so is the problem of synthesizing an optimal $LN'$ starting from $LN$.
+Even though we assumed we could generate $LN'$ in non-exponential time, the following issues should be addressed
 \begin{itemize}
-\item as opposed to what happens with hoisting in perfect multilinear loop nests, in this case the size of the temporaries would be proportional to the number of non-reduction loops crossed (in the example, \texttt{N$\cdot$O} for \texttt{ijk} sub-expressions, \texttt{L$\cdot$N$\cdot$O} for \texttt{eijk} ones).
+\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime;
-\item the process by which multi-invariant sub-expressions are exposed could require transformations like expansion and factorization.
-Consequently, the save originating from the elimination of the reduction loop could be overwhelmed by a potentially larger increase in the number of operations in the body of $L_k$ (expansion could increase the number of operations, e.g.
+\item the transformations exposing multi-invariant sub-expressions could require expansion and factorization.
+In terms of number of operations, the save originating from the elimination of the reduction loop could be overwhelmed by the increase in arithmetic complexity of $L_k$ (e.g., expansion can increase the operation count, e.g.
 $A(B+C) = AB + AC$).
@@
 
-Clearly, this definition remains vague until instantiated in a certain model.
-Once the model is fixed, we can argue about the meaning of \texttt{C} and try to produce an algorithm to reach optimality.
-This is tackled in the next section.
+Note how the definition contemplates the possibility for a nest to have multiple innermost loops.
+In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or result of suitable transformations.
 
@@
 
-In this section, we instantiate our definition of loop optimality in the domain of finite element integration.
-This requires elaborating on the junction between two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem, which we reviewed in Section~\ref{sec:background}; and the (partly multilinear) loop nests implementing such forms.
+In this section, we instantiate Definition~\ref{def:optimality} in our domain of interest, finite element integration.
+This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem, which we reviewed in Section~\ref{sec:background}; and the (partly multilinear) loop nests implementing such forms.
 
 Our point of departure is the example loop nest in Figure~\ref{code:loopnest}.
-This loop nest is a simplified view of a typical bilinear form implementation.
-The $L_e$ loop represents iteration over the elements of a mesh; the $L_i$ loop derives from using numerical quadrature; the perfect loop nest $[L_j, L_k]$ implements the computation over the test and trial functions' degrees of freedom.
+This loop nest is actually a simplified view of a typical bilinear form implementation.
+$L_e$ represents iteration over the elements of a mesh; $L_i$ derives from using numerical quadrature; $[L_j, L_k]$ implement the computation over test and trial functions.
 We deliberately omitted useless portions of code to not hinder readability (e.g.
-matrix insertion) and to avoid tying our discussing to a specific form (e.g.
+matrix insertion) and to avoid tying our discussion to specific forms (e.g.
 \texttt{F} is unspecified).
 
-Before expanding on optimality, we make a few observations about properties of our domain.
-1) We know that $L >>> M, N, O$; that is, even with high order basis functions, the number of elements is typically order of magnitude larger than both the number of quadrature points and degrees of freedom.
-2) The loop nest enclosed by the integration loop is always perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
+From domain knowledge, we make the following observations.
+1) $L >>> M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
 
 % maybe say somewhere the order of the loop nest eijk is just the most natural --- the other present worse data locality for example
-% change L into some other letters
-% say somewhere that N and O are the number of dofs for test and trial functions
 
@@
 \label{sec:mem-const}
-The fact that the iteration space of $L_e$ is so larger than that of other loops suggests that we should be cautious when hoisting outside of $LN$.
+The fact that the iteration space of $L_e$ is so larger than that of other loops suggests we should be cautious when hoisting out of $LN$.
 Imagine a time stepping loop $L_t$ wraps $LN$.
-Theoretically, one could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and hoist them out of $L_e$.
+One could then think of identifying time-invariant sub-expressions that access both geometry and reference element terms and pre-evaluate them within $L_t$.
 Unless adopting complex engineering solutions (e.g.
-aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by a factor $L$.
-It is our opinion, therefore, that the decrease in the number of operations would be completely overwhelmed by the bigger memory pressure.
+aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by $O(L)$.
+It is our opinion, therefore, that the drop in number of operations would be overwhelmed, from the run-time viewpoint, by the larger memory pressure.
 
@@
 We will provide precise details about this in the following sections.
-For the moment, and just as one of many possible examples, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporaries of size equal to $N\cdot O$; depending on the discretization employed it is not so uncommon to break the local memory threshold.
+For the moment, and just as one of many possible examples, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.
 
-Based upon these considerations, we impose two memory constraints
-\begin{itemize}
-\item The size of a temporary due to code motion cannot be bigger than the size of the multilinear loop nest iteration space ($N\cdot O$ for the bilinear form in the example).
-A corollary is that hoisting expressions involving geometry terms outside of $L_e$ is forbidden.
+Based upon these considerations, we add two constraints to $C$ (see Definition~\ref{def:optimality})  
+\begin{enumerate}
+\item The size of a temporary due to code motion cannot be bigger than the size of the multilinear loop nest iteration space. 
 \item The total size of the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
-\end{itemize}
+\end{enumerate}
+A corollary of $C1$ is that hoisting expressions involving geometry terms outside of $L_e$ becomes forbidden.
 
@@
 
-Eliminating sharing from the multilinear loops does not suffice for nest optimality.
-In fact, we wonder whether the reduction imposed by $L_i$ could be pre-evaluated outside of $N$, as already suggested in~\ref{def:optimality}, thus potentially reducing the operation count.
+Eliminating sharing from the multilinear loops does not suffice.
+In fact, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, as already suggested in~\ref{def:optimality}, and the operation count be reduced.
 
@@
 We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated.
+Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries.
 As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
@@
 
-The intuition of the main algorithm for optimal loop nest synthesis is shown in Figure~\ref{code:intuition}.
+The challenge is to understand when, and for which monomials, pre-evaluation is profitable. We propose an algorithm and a discussion of its optimality.
+
+The intuition of the algorithm for optimal loop nest synthesis is shown in Figure~\ref{code:intuition}.
 
@@
 % global sharing elimination
-% we exclude pathological cases in which different monomials lead to pre-evaluating the same tables
 
@@
 This is studied "locally"; that is, for each monomial, in isolation.
-If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}, and the evaluation eventually performed.
+If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed.
 The result is a set of $n$-dimensional tables (these can be seen as "slices" of the reference tensor at the math level), $n$ being the arity of the multilinear form.
@@
 Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
-
 The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.
@@
 Consider an expression comprising a set of monomials $M$.
-Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and let $Z$ be such that $Z = M \setminus P$.
+Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and be $Z = M \setminus P$.
 Assume that:
 \begin{enumerate}
-\item the cost function employed is optimal; that is, it predicts correctly the profitability of pre-evaluation
+\item the cost function employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
 \item pre-evaluating distinct monomials does not produce identical tables
@@
 \label{sec:op_count}
-In this section we discuss the optimality of the cost function.
+The loose hand we have to tie regards the construction of the pre-evaluation cost function.
+We introduce the cost function $\theta : M \to \mathbb{N} \times \mathbb{N}$ that, given a monomial, returns two natural numbers representing the ideal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation.
+Since $\theta$ is integrated in a compiler, a requirement is that its execution is fast.
 
-%\item cost model for deciding what to do
-%\item heuristic optimisation of constant and quadrature dependent expressions
+We start with $\theta_{wo}$.
+Without performing any kind of transformations, we can easily predict the number of operations in the innermost loop once sharing is removed.
+This is because of our key property, linearity.
+By just counting the number of unique terms along the multilinear loop nest, and by taking into account $I$, the size of $L_i$, we can infer the cost.
+This is also explained in Proposition~\ref{prop:optimal-mln}.
+
+For $\theta_{w}$, things are more complicated.
+We first need to account for the presence of (derivatives of) coefficients.
+This is because we want to estimate the \textit{increase factor} $\iota$.
+The increase factor captures the increase in arithmetic complexity within the multilinear loop nest if pre-evaluation is performed.
+To contextualize, consider the example in Figure~\ref{code:increase_factor}.
+
+\begin{figure}[h]\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
+\end{CenteredBox}\caption{Optimized Loop nest example}\label{code:increase_factor}\end{figure}
+
+One can think of this as the (simplified) loop nest originating from a pre-multiplied mass matrix.
+The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} expresses the field $f$ over a function space from which $B$ was derived.
+In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the other (reference element specific) quantities.
+This requires expanding a product, which, in turn, increases the number of $L_k$-dependent operations by a factor 3 (the number of the coefficient local degrees of freedom).
+$\iota$, or increase factor, precisely captures this growth in the number of $L_k$-dependent operations.
+Note that in an expression is not innermost loop dependent, then it can be hoisted, and therefore not significant for the operation count.
+
+To determine $\iota$, we distinguish two cases: coefficients expressed over different function spaces and coefficients expressed over the same function space.
+This is essential because there could be more than one (derivative of) coefficient pre-multiplying the same monomial.
+In the former case, $\iota$ is equal to the cost of expansion.
+In the latter, after expansion we obtain identical summands, which eventually pre-evaluate to the same table.
+Consequently, some sharing would be eventually removed from the loop nest resulting from pre-evaluation (e.g., expression X at line Y in Figure~\ref{code:loopnest-opt}).
+We then need to calculate the number of combinations with repetitions of the symbols appearing in the monomial's coefficients to precisely determine $\iota$.
+
+If $\iota \geq I$ we can immediately say that pre-evaluation will not be profitable.
+This is indeed a necessary condition that, intuitively, tells us that if we add to the innermost loop more operations than we actually save from eliminating $L_i$, then for sure $\theta_{wo} < \theta_{w}$.
+This observation can speed up the compilation time by decreasing the analysis cost.
+
+If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$.
+In particular, we need to calculate the number of reference-element specific terms, $\rho$, by means of which we can set $\theta = \rho \cdot \iota$.
+Consider again Figure~\ref{code:increase_factor}.
+In the case of the mass matrix, the body of $L_k$ is characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
+In general, $\rho$ varies with the discretization and the differential operators employed.
+For example, in the case of the bi-dimensional Poisson equation, $\rho = 3$.
+We determine $\rho$ by projecting the
 

### 2015-07-31 TOMS_Paper/assembly.tex
Rewrite bits of operation counting

@@
 for each monomial M:
-  C = estimate $\mbox{\texttt{ops}}$ after pre-evaluation
-  NC = estimate $\mbox{\texttt{ops}}$ without pre-evaluation
-  if C < NC and memory constraints satisfied:
+  $\theta_{w}$ = estimate $\mbox{\texttt{ops}}$ after pre-evaluation
+  $\theta_{wo}$ = estimate $\mbox{\texttt{ops}}$ without pre-evaluation
+  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
     mark M as candidate for pre-evaluation
@@
 \label{sec:op_count}
-The loose hand we have to tie regards the construction of the pre-evaluation cost function.
-We introduce the cost function $\theta : M \to \mathbb{N} \times \mathbb{N}$ that, given a monomial, returns two natural numbers representing the ideal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation.
-Since $\theta$ is integrated in a compiler, a requirement is that its execution is fast.
+It remains to tie one loose hand: the construction of the pre-evaluation cost function.
+We introduce the cost function $\theta : M \to \mathbb{N} \times \mathbb{N}$ that, given a monomial, returns two natural numbers representing the optimal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation.
+Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
 
-We start with $\theta_{wo}$.
-Without performing any kind of transformations, we can easily predict the number of operations in the innermost loop once sharing is removed.
-This is because of our key property, linearity.
-By just counting the number of unique terms along the multilinear loop nest, and by taking into account $I$, the size of $L_i$, we can infer the cost.
-This is also explained in Proposition~\ref{prop:optimal-mln}.
+We can easily predict $\theta_{wo}$ thanks to our key property, linearity.
+As described in Proposition~\ref{prop:optimal-mln}, a visit of the loop nest suffices to obtain the cost of the optimal synthesis of a multilinear loop nest, namely $MLN_{ops}$.
+Then, assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = MLN_{ops} \cdot I$.
 
 For $\theta_{w}$, things are more complicated.
-We first need to account for the presence of (derivatives of) coefficients.
-This is because we want to estimate the \textit{increase factor} $\iota$.
-The increase factor captures the increase in arithmetic complexity within the multilinear loop nest if pre-evaluation is performed.
+We first need to account for the presence of (derivatives of) coefficients to estimate the \textit{increase factor} $\iota$.
+This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation.
 To contextualize, consider the example in Figure~\ref{code:increase_factor}.
@@
 
-One can think of this as the (simplified) loop nest originating from a pre-multiplied mass matrix.
-The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} expresses the field $f$ over a function space from which $B$ was derived.
-In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the other (reference element specific) quantities.
-This requires expanding a product, which, in turn, increases the number of $L_k$-dependent operations by a factor 3 (the number of the coefficient local degrees of freedom).
-$\iota$, or increase factor, precisely captures this growth in the number of $L_k$-dependent operations.
-Note that in an expression is not innermost loop dependent, then it can be hoisted, and therefore not significant for the operation count.
+One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix.
+The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$.
+In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from other (reference element specific) quantities.
+By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient).
+$\iota$ serves to capture this growth.
 
-To determine $\iota$, we distinguish two cases: coefficients expressed over different function spaces and coefficients expressed over the same function space.
-This is essential because there could be more than one (derivative of) coefficient pre-multiplying the same monomial.
-In the former case, $\iota$ is equal to the cost of expansion.
-In the latter, after expansion we obtain identical summands, which eventually pre-evaluate to the same table.
-Consequently, some sharing would be eventually removed from the loop nest resulting from pre-evaluation (e.g., expression X at line Y in Figure~\ref{code:loopnest-opt}).
-We then need to calculate the number of combinations with repetitions of the symbols appearing in the monomial's coefficients to precisely determine $\iota$.
+With just a single coefficient, $\iota$ directly descends from the cost of expansion.
+In general, however, the calculation is less straightforward.
+Consider, for example, the case in which multiple coefficients originate from the same function space.
+Expansion would then lead to identical terms (i.e., identical pre-evaluated tables).
+Therefore, for a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the arity of the multilinear loop nest and $n$ the set of terms involved in the coefficient expansion.
 
@@
 If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$.
-In particular, we need to calculate the number of reference-element specific terms, $\rho$, by means of which we can set $\theta = \rho \cdot \iota$.
+In particular, we need to calculate the number of reference-element specific terms, $\rho$, such that $\theta_{w} = \rho \cdot \iota$.
 Consider again Figure~\ref{code:increase_factor}.
-In the case of the mass matrix, the body of $L_k$ is characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
+In the case of the mass matrix, the body of $L_k$ is characterized by just the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
 In general, $\rho$ varies with the discretization and the differential operators employed.
 For example, in the case of the bi-dimensional Poisson equation, $\rho = 3$.
-We determine $\rho$ by projecting the
+There are several ways of determining $\rho$.
+The fastest would be to extract it from high-level analysis of the operators in the form; for convenience, in our implementation we instead project (by analysis of the expression tree) the output of expansion and factorization in the monomial.
 

### 2015-07-31 TOMS_Paper/assembly.tex
Start talking about coffee language

@@
 ...also for the first time we provide a formal explanation of the math in terms of compiler theory
+-  Deficiencies of previous approaches
 
@@
 \label{sec:codegen}
+The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake.
+In this section, we describe the structure of the code generation system and we comment on a set of low-level optimizations.
 
-\begin{itemize}
-\item Deficiencies of previous approaches
-\item  Building block operations in coffee: 1) eliminating reuse, 2)  insights on the various tree algorithms employed
-\end{itemize}
+\subsection{Automation through the COFFEE Language}
+As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric "building-block" operations.
+This has several advantages.
+Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could ideally be expressed using the existing operators, or with small effort building on what is already available.
+Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler.
+Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.
 
-\subsection{Heuristic Optimization of Integration-dependent Expressions}
-%insights that optimisation of loops outside of the multilinear nest can be tackled greedily and heuristically
+A non-exhaustive list of such operations includes expansion, factorization, re-association, generalized code motion.
+These are parametric operations.
+For example, one could ask to factorize only constant or only linear terms, while hoisting can be limited by imposing a constraint on the size of the temporaries.
+Each of these operations is implemented by suitable manipulation of the abstract syntax tree representing the integration routine.
+
+\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
+As a proof-of-concept of our generality claim,
 

### 2015-08-03 TOMS_Paper/assembly.tex
Write optimization of Li quantities

@@
 
-In this section, we characterize our definition of optimality and we describe the assumptions under which it holds.
-
-In order to make the document self-contained, we start by reviewing basic compiler terminology.
+In this section, we provide a characterization of optimality for loop nests of various nature. In order to make the document self-contained, we start with reviewing basic compiler terminology.
 
@@
 
-Note that nothing prevents a perfect multilinear loop nest from being rooted in a deeper, possibly imperfect loop nest.
-In fact, we will later focus on this particular structure.
-First, we need to define optimality for perfect multilinear loop nests.
-To this purpose, it is convenient to introduce the notion of sharing.
+Note that perfect multilinear loop nests could actually be rooted in deeper, possibly imperfect loop nests. In fact, we will focus on this particular structure.
+
+We introduce the fundamental notion of sharing.
 
@@
 
-We can now present a simple yet fundamental result.
+We now have the ingredients to formulate a simple yet fundamental result.
 
@@
 
-Based on this proposition, we define optimality as follows.
+This result will be essential to prove that optimality, characterized as follows, can structurally be reached in this class of loops.
 
@@
 If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense.
-In the following, we assume to operate in a CPU-bound regime, in which the body of loop nests are characterized by arithmetic-intensive expressions.
+In the following, we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated.
 This suits the context of finite element integration.
@@
 \label{prop:optimal-mln}
-An optimal synthesis for a perfect multilinear loop nest $LN$ can deterministically be found.
+A perfect multilinear loop nest $LN$ can always be reduced to optimal form.
 \end{Prop}
@@
 This renders $L_{i_{n-1}}$ optimal.
-In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of the summation.
+In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
 \end{proof}
 
-We now consider generic loop nests in which multilinearity applies to a subset of loops only. Consider the example in Figure~\ref{code:loopnest}.
+For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_optimal}.
+
+\begin{figure}[h]\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}
+\end{CenteredBox}\caption{Multilinear loop nest in optimal form}\label{code:multi_loopnest_optimal}\end{figure}
+
+We now consider a wider class of loop nests in which perfectness and multilinearity apply to a sub-nest only. Consider the example in Figure~\ref{code:loopnest}.
 
@@
 \section{Synthesis of Optimal Loop Nests in Finite Element Integration}
-\label{sec:optimal-syntesis}
+\label{sec:optimal-synthesis}
 
@@
 
-Eliminating sharing from the multilinear loops does not suffice.
-In fact, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, as already suggested in~\ref{def:optimality}, and the operation count be reduced.
+Eliminating sharing from the inner multilinear loops does not suffice.
+In fact, as suggested in Section~\ref{def:optimality}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
 
@@
 for each monomial M:
-  $\theta_{w}$ = estimate $\mbox{\texttt{ops}}$ after pre-evaluation
-  $\theta_{wo}$ = estimate $\mbox{\texttt{ops}}$ without pre-evaluation
+  $\theta_{w}$ = estimate operation count with pre-evaluation
+  $\theta_{wo}$ = estimate operation count without pre-evaluation
   if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
@@
 We can easily predict $\theta_{wo}$ thanks to our key property, linearity.
-As described in Proposition~\ref{prop:optimal-mln}, a visit of the loop nest suffices to obtain the cost of the optimal synthesis of a multilinear loop nest, namely $MLN_{ops}$.
+As explained in Proposition~\ref{prop:optimal-mln}, a simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $MLN_{ops}$.
 Then, assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = MLN_{ops} \cdot I$.
@@
 The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$.
-In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from other (reference element specific) quantities.
+In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities.
 By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient).
-$\iota$ serves to capture this growth.
+Intuitively, $\iota$ captures this growth in non-hoistable operation.
 
@@
 If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$.
-In particular, we need to calculate the number of reference-element specific terms, $\rho$, such that $\theta_{w} = \rho \cdot \iota$.
+In particular, we need to calculate the number terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$.
 Consider again Figure~\ref{code:increase_factor}.
-In the case of the mass matrix, the body of $L_k$ is characterized by just the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
+In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
 In general, $\rho$ varies with the discretization and the differential operators employed.
-For example, in the case of the bi-dimensional Poisson equation, $\rho = 3$.
+For example, in the case of the bi-dimensional Poisson equation, after a suitable factorization, we have $\rho = 3$.
 There are several ways of determining $\rho$.
@@
 The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake.
-In this section, we describe the structure of the code generation system and we comment on a set of low-level optimizations.
+In this section, we describe the structure of the code generation system and we also comment on a set of low-level optimizations.
 
@@
 \subsubsection{Heuristic Optimization of Integration-dependent Expressions}
-As a proof-of-concept of our generality claim,
+As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions.
+These are expressions that should logically be placed within $L_i$.
+They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point.
+For some complex monomials (e.g., see~\citeN{quadrature1} and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, outweigh that in the multilinear loop nest.
+In these cases, our definition of optimality becomes weaker, since the underlying assumption is that the bulk of the computation is carried out in innermost loops.
+
+Despite the fact that we are not characterizing optimality for this wider class of loops, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis}, particularly for removing sharing, to minimize the operational cost.
+This is straightforward in our code generation system by composing rewrite operators.
+Our strategy is as follows.
+
+COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions.
+This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to trivial code without performing any sort of transformation.
+Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$.
+We then synthesize an optimal loop nest as described in the previous sections.
+This may in turn expose a set of $L_i$-dependent statements evaluating some temporaries.
+For each of this statements, we try to remove sharing by greedily applying factorization and code motion.
+In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named "factorize", "expand", etc.).
 
@@
 ...
-
-% MAYBE JUST A SUBSECTION, "INSIGHTS OF LOW-LEVEL OPTS ...."
 

### 2015-08-03 TOMS_Paper/assembly.tex
Write some low level opts

@@
 \subsubsection{Heuristic Optimization of Integration-dependent Expressions}
+\label{sec:heuristic-int-dep}
 As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions.
@@
 \label{sec:code-spec}
-...
+We comment on a set of low-level optimizations.
+These are essential to 1) achieve machine peak performance (Sections~\ref{sec:opt-review} and~\ref{sec:vect-prom}) and 2) make COFFEE independent of the high-level form compiler (Section~\ref{sec:zeros}).
+As we will see, there are interplays among different transformations.
+For completeness, we present all of the transformations available in the compiler, although we will only use a subset of them for a fair performance evaluation.
 
-\subsubsection{Avoiding Iteration over Zero-valued Blocks by Symbolic Execution}
+\subsubsection{Review of Existing Optimizations}
+\label{sec:opt-review}
+We start with briefly reviewing the low-level optimizations presented in~\citeN{Luporini}.
+
+\paragraph{Padding and data alignment} All of the arrays involved in the evaluation of the local element matrix or vector are padded to a multiple of the vector register length.
+This is a simple yet powerful transformation that maximizes the effectiveness of vectorization.
+Padding, and then loop bounds rounding, enable data alignment and avoid the introduction of scalar remainder loops.
+
+\paragraph{Vector-register Tiling} Blocking (or tiling) at the level of vector registers improves data locality beyond traditional unroll-and-jam transformations.
+The blocking strategy consists of evaluating outer products by just using two vector register without ever spilling to cache.
+
+\paragraph{Expression Splitting} When the number of basis functions arrays (or, equivalently, temporaries introduced by code motion) and constants is large, the chances of spilling to cache are high in architectures with a relatively low number of logical registers (e.g.
+16/32).
+By exploiting sum’s associativity, an expression can be fissioned so that the resulting sub-expressions can be computed in a separate loop nest.
+
+\subsubsection{Vector-promotion of Integration-dependent Expressions}
+\label{sec:vect-prom}
+Integration-dependent expressions are inherently executed as scalar code because vectorization (unless employing special hand-written schemes) occurs along a single loop, typically the innermost.
+In our loop nest, $L_i$ is clearly an outer loop.
+In situations such as in Section~\ref{sec:heuristic-int-dep}, we may want to vectorize also along $L_i$.
+One way to achieve this is vector-promotion.
+This implies creating a clone loop of $L_i$ in the preheader of the main loop nest, in which vector temporaries, instead of scalar ones, are evaluated.
+
+\subsubsection{Handling Block-sparse Tables}
 \label{sec:zeros}
-%TODO : say that the reasons to have zeros is to keep the design of the form compiler simple, which just "puts zeros" and "outline the assembly expression".
+Consider a tabulated set of basis functions having quadrature points along the rows and functions along the columns.
+For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}.
+Unless using a smart form compiler (and one of our goals is precisely to relieve it from the burden of smart code generation), there are several circumstances in which the tables can be block-sparse.
+Zero-valued columns arise when taking derivatives on a reference element or when employing vector-valued elements.
+Zero-valued rows can result from using non-standard functions spaces, like Raviart-Thomas.
+Zero-valued blocks can appear in pre-evaluated temporaries.
+Our objective is a general system that avoids useless iteration over these zero-entries while preserving the effectiveness of the other low-level optimizations, especially vectorization.
 
-% say that we just don't kill al zero-columns and use indirections to not break vectorisation
+In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g.
+\texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed.
+Our approach, which will be compared to this pioneering work, aims to free the generated code from any indirection arrays.
+This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.
 
-% generic engine to remove zeros. Can be rows (RT elements), columns (derivatives, vector elements), blocks (vector elements, "TC mode"). Based on symbolic execution.
+The idea is that if the dimension along which vectorization is performed (typically the innermost) has a contiguous slice (rows, columns) of zero-valued entries, and if that slice is smaller than the vector length, that we do nothing (i.e., the loop nest is not transformed).
+Otherwise, we restructure the loop nest.
+This has several complicated implications, including loop fission and a memory offsetting usually dangerous for padding and data alignment, which are, however, beyond the scope of this paper.
+The implementation is based on symbolic execution: the loop nests are traversed and for each statement the location of zeros in each of the involved symbols is tracked.
+Arithmetic operators have a different impact on the tracking.
+For example, for each loop dimension, multiplication requires computing the set intersection of the zero-valued slices, whereas addition requires computing the set union.
+
+% say that the reasons to have zeros is to keep the design of the form compiler simple, which just "puts zeros" and "outline the assembly expression".
 % rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
 % tables shrunk and merged
-
-\subsubsection{Padding and Data Alignment}
-\label{sec:padding}
-...
-
-\subsubsection{Vector-promotion of Integration Quantities}
-...
 

### 2015-08-03 TOMS_Paper/assembly.tex
Rewrite perf eval sect intro

@@
 
-Experiments were run on a single core of an Intel architecture, a Sandy Bridge I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache.
-The Intel \texttt{icc 14.2} compiler was used.
-The compilation flags used were \texttt{-O3, -xHost, -xAVX, -ip}.
+Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache.
+The Intel Turbo Boost and Intel Speed Step technologies were disabled.
+The Intel \texttt{icc 15.2} compiler was used.
+The compilation flags used were \texttt{-O3, -xHost, -ip}.
 
-We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods.
-In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms.
-The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{ufl-code}.
+We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
+In particular, we study the mass matrix and the bilinear forms arising in a Helmholtz equation, in an elastic model, and in a hyperelastic model.
+The complete specification of these forms is made publicly available (\cite{ufl-code}).
 
-We evaluate the \textit{speed ups} achieved by three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied.
-In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning as described in Section~\ref{sec:autotune}).
-The \texttt{auto} values do not include the autotuner cost, which is commented aside in Section~\ref{sec:auto-analysis}.
-
-%This way, we also highlight the advances achieved over our previous work. 
-%...TODO...: dire che i tempi sono la media di 3 runs
-
-The values that we report include the cost of local assembly as well as the cost of matrix insertion.
-However, the unstructured mesh has been made small enough to fit the L3 cache, so as to minimize the ``noise'' due to any operations that are not part of the element matrix evaluation itself.
-However, it has been reiterated over and over (e.g.~\cite{quadrature1}) that as the complexity of a form increseas, the cost of local assembly becomes dominant.
-All codes were executed in the context of the Firedrake framework.
-
-We vary several aspects of each form, which follows the approach and the notation of~\cite{quadrature1} and~\cite{Francis}
+We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
 \begin{itemize}
-\item The polynomial order of basis functions, $q \in \lbrace1, 2, 3, 4\rbrace$
-\item The polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
-\item The number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
+\item the FEniCS Form Compiler's built-in optimization systems, namely
+\begin{itemize}
+\item optimized quadrature mode, \texttt{ffc-quad}
+\item tensor mode, \texttt{ffc-tens}
+\item automatic mode, \texttt{ffc-auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
 \end{itemize}
-On the other hand, other aspects are fixed 
-\begin{itemize}
-\item The space of both basis and coefficient functions is Lagrange
-\item The mesh is three-dimensional, made of tetrahedrons, for a total of 4374 cells
+\item a novel back-end for the FEniCS Form Compiler, UFLACS \texttt{uflacs}, whose primary goals are improved code generation time and runtime
+\item the sole generalized loop-invariant code motion in COFFEE plus padding, \texttt{coffee-O1}, as discussed in~\citeN{Luporini}
+\item the optimal multilinear loop nest synthesis plus padding and symbolic execution, \texttt{coffee-O2}, which is the focus of this paper
 \end{itemize}
 
-Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will be deeply commented in the next section, must be read as ``plots, or grids, of plots''.
-Each grid (figure) has two logical axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
-The top-left plot in a grid shows speed ups for $[q=1, p=1]$; the plot on its right does the same for $[q=1, p=2]$, and so on.
-The diagonal of the grid shows plots for which basis and coefficient functions have same polynomial order, that is $q=p$.
-Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.
+The values that we report are the average of three runs with "warm cache" (no code generation time, no compilation time).
+They include the cost of local assembly as well as the cost of matrix insertion.
+However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.
+For a fair comparison, small patches (publicly available) were written to be able to use the aforementioned transformation systems in Firedrake; that is, all simulations are run through Firedrake.
+UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS (\cite{Fenics}), which employs a data storage layout different than Firedrake.
+Our small patches fix this problem by producing code with a "correct" data storage layout.
 
-A plot reports speed-ups over non-optimized FEniCS-Form-Compiler-generated code.
-There are three groups of bars, each group referring to a particular version of the code (\texttt{ffc, fix, auto}).
+Following the methodology (and the notation) used in ~\cite{quadrature1} and~\cite{Francis}, we increasingly vary the complexity of each form. In particular:
+\begin{itemize}
+\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
+\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
+\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
+\end{itemize}
+Constants of our analysis are instead
+\begin{itemize}
+\item the space of test, trial, and coefficient functions, Lagrange
+\item the mesh, tetrahedral with a total of 4374 elements
+\end{itemize}
+
+Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will deeply be commented in Section~\ref{sec:perf-results-forms}, can be interpreted as ``plots (grids) of plots''.
+Each grid (each figure) has two logical "outer" axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
+The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
+Looking at the diagonal of the grid one can obtain the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$.
+A grid (figure) can therefore be read in many different ways, which allows us to make structured considerations on the performance achieved.
+In each plot there are three groups of bars, each group referring to a particular version of the code (\texttt{ffc-quad, ffc-tens, ...}).
 There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.
@@
 Also, details on the ``optimal'' code variant determined by autotuning are given in Section~\ref{sec:auto-analysis}.
+
+\paragraph{Mass}
+%\begin{figure}[t]  
+%...
+%%\includegraphics[scale=0.7]{perf-results/mass}
+%%\caption{Mass results.}\label{fig:mass}
+%\end{figure}
+%
 
@@
 
-\paragraph{Poisson}
-%\begin{figure}[t]  
-%...
-%%\includegraphics[scale=0.7]{perf-results/poisson}
-%%\caption{Poisson results.}\label{fig:poisson}
-%\end{figure}
-%
-%In Figure~\ref{fig:poisson} we report speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the original code for the Poisson form.
-We note that, as a general trend, \texttt{ffc} exhibits drops in performance as $nf$ increases, notably when $nf=3$, for any values of $q$ and $p$.
-This is a consequence of the inherent complexity of the generated code.
-The way \texttt{ffc} performs loop-invariant code motion leads to the pre-computation of integration-dependent terms at the level of the integration loop, which are characterized by higher arithmetic intensity and redundant computation as $nf$ increases.
-Moreover, the absence of vectorization is another limiting factor.
-%
-%The \texttt{auto} variant generally shows the best performance.
-Significant improvements over \texttt{fix} are also achieved, notably as $q$, $p$ and $nf$ increase.
-As clarified in the next section, this is always due to a more aggressive expression rewriting in combination with the zero-avoidance technique.
-
 \paragraph{Hyperelasticity}

### 2015-08-03 TOMS_Paper/assembly.tex
Rewrite the introduction

@@
 The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
-This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become increasingly popular over the last years.
+This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years.
 In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language.
-The mathematical specification is manipulated and then passed to a form compiler, which generates a representation of local assembly operations.
-These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution.
-Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more computationally expensive, covering a significant fraction of the overall computation run-time.
+The mathematical specification is manipulated by a form compiler, which then generates a representation of assembly operators.
+By applying these operators to an element in the discretized domain a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed.
+The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.
 
-Producing high performance implementations is, however, non-trivial.
-The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process.
-Traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions.
-Polyhedral-model-based source-to-source compilers, for instance~\cite{PLUTO}, mainly apply aggressive loop optimizations, such as tiling, but these are not particularly helpful in our context.
-This lack of suitable optimizing tools has led to the development of a number of higher-level approaches to maximize the performance of local assembly kernels.
-In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''.
-\cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration.
-In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct sets of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality.
-The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.
+As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges.
+This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
+General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or "flops", than necessary).
+Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context.
+The lack of suitable third-party tools has led to the development of a number of domain-specific transformation systems or synthesizers.
+In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations that a user should not be expected to write ``by hand''.
+In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count.
+In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed.
+There is also an on-going effort to produce a novel form compiler, and therefore a novel code generation system for assembly operators, called UFLACS (\cite{Uflacs}).
 
-Our research has resulted in ....
-... we build on our previous work~\cite{Luporini} ... and present a ... We argue that for complex, realistic forms, peak performance can be achieved only by  ... 
-... also low-level optimisation ...
-...also for the first time we provide a formal explanation of the math in terms of compiler theory
--  Deficiencies of previous approaches
+In spite of a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
+With this paper, we aim to make a first step in the direction of solving this problem.
 
-This is all implemented in COFFEE, which in turn is integrated with the Firedrake framework.
-We provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex (hyperelasticity) models.
-We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions.
-To clearly distinguish the improvement achieved by this work, we will compare, for each test case, X sets of code variants: 1) unoptimized code, i.e.
-a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e.
-the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; ....
+We summarize the contributions of this research as follows
+\begin{itemize}
+\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration.
+As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
+\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
+\item We implemented the theory in our compiler, COFFEE\footnote{...}, which is the default optimization system in the Firedrake framework.
+Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
+\item We provide an unprecedented (for extension) performance evaluation comparing several code transformation/synthesizers systems in a wide range of forms of increasing complexity, which is essential to demonstrate our optimality claim.
+\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we deal with (i.e., optimize) them
+\item We enrich the set of low-level transformations described in~\citeN{Luporini}
+\end{itemize}
 
@@
 
-These two modes have been implemented in the Fenics Form Compiler.
+These two modes have been implemented in the FEniCS Form Compiler.
 In this compiler, a simple heuristic is used to choose the most suitable mode for a given form.
 It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically~\cite{Fenics}.
-We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-syntesis}.
+We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}.
 For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of selecting the optimal mode at the monomials level -- no heuristics, no "global" choice of the mode for all monomials in the form -- without affecting the cost of code generation.

### 2015-08-04 TOMS_Paper/assembly.tex
Add helmholtz and mass plots

@@
 \label{eq:quadrature}
-\scriptsize
 A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
@@
 \label{eq:tensor}
-\scriptsize
 A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
@@
 
-Figures~\ref{fig:helmholtz},~\ref{fig:elasticity},~\ref{fig:poisson}, and~\ref{fig:hyperelasticity}, which will deeply be commented in Section~\ref{sec:perf-results-forms}, can be interpreted as ``plots (grids) of plots''.
+Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}, which will deeply be commented in Section~\ref{sec:perf-results-forms}, can be interpreted as ``plots (grids) of plots''.
 Each grid (each figure) has two logical "outer" axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
@@
 \paragraph{Mass}
-%\begin{figure}[t]  
-%...
-%%\includegraphics[scale=0.7]{perf-results/mass}
-%%\caption{Mass results.}\label{fig:mass}
-%\end{figure}
-%
+\begin{figure}[t] 
+ \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
+\caption{Mass results.}\label{fig:mass}
+\end{figure}
 
 \paragraph{Helmholtz}
-%\begin{figure}[t]  
-%...
-%%\includegraphics[scale=0.7]{perf-results/helmholtz}
-%%\caption{Helmholtz results.}\label{fig:helmholtz}
-%\end{figure}
+\begin{figure}[t] 
+ \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
+\caption{Helmholtz results.}\label{fig:helmholtz}
+\end{figure}
+
 %The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.

### 2015-08-04 TOMS_Paper/assembly.tex
Add perf eval general comments

@@
 We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
-In particular, we study the mass matrix and the bilinear forms arising in a Helmholtz equation, in an elastic model, and in a hyperelastic model.
+In particular, we study the mass matrix ("\texttt{Mass}") and the bilinear forms arising in a Helmholtz equation ("\texttt{Helmholtz}"), in an elastic model ("\texttt{Elasticity}"), and in a hyperelastic model ("\texttt{Hyperelasticity}").
 The complete specification of these forms is made publicly available (\cite{ufl-code}).
@@
 \begin{itemize}
-\item optimized quadrature mode, \texttt{ffc-quad}
-\item tensor mode, \texttt{ffc-tens}
-\item automatic mode, \texttt{ffc-auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
+\item optimized quadrature mode, \texttt{quad}
+\item tensor mode, \texttt{tens}
+\item automatic mode, \texttt{auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
 \end{itemize}
-\item a novel back-end for the FEniCS Form Compiler, UFLACS \texttt{uflacs}, whose primary goals are improved code generation time and runtime
-\item the sole generalized loop-invariant code motion in COFFEE plus padding, \texttt{coffee-O1}, as discussed in~\citeN{Luporini}
-\item the optimal multilinear loop nest synthesis plus padding and symbolic execution, \texttt{coffee-O2}, which is the focus of this paper
+\item a novel back-end for the FEniCS Form Compiler, UFLACS, referred to as \texttt{ufls}, whose primary goals are improved code generation time and runtime
+\item the sole generalized loop-invariant code motion in COFFEE plus padding, \texttt{cfO1}, as discussed in~\citeN{Luporini}
+\item the optimal multilinear loop nest synthesis plus padding and symbolic execution, \texttt{cfO2}, which is the focus of this paper
 \end{itemize}
@@
 
-Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}, which will deeply be commented in Section~\ref{sec:perf-results-forms}, can be interpreted as ``plots (grids) of plots''.
-Each grid (each figure) has two logical "outer" axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis.
+Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}, which will be commented in Section~\ref{sec:perf-results-forms}, are "nested" plots.
+We refer to the outer plot as the "grid", while the inner are the actual "plots".
+In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis.
 The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
-Looking at the diagonal of the grid one can obtain the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$.
-A grid (figure) can therefore be read in many different ways, which allows us to make structured considerations on the performance achieved.
-In each plot there are three groups of bars, each group referring to a particular version of the code (\texttt{ffc-quad, ffc-tens, ...}).
+The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$.
+A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved.
+In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}).
 There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.
@@
 \label{sec:perf-results-forms}
-%The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly.
-We limit ourselves to analyzing the cost of computing element matrices, although all of the techniques presented in this paper are immediately extendible to the evaluation of local vectors.
-As anticipated, in the following we comment speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the non-optimized, FEniCS-Form-Compiler-generated code.
-%
-%We first comment on results of general applicability.
-By looking at the various figures, we note there is a trend in COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase.
-This is because most of the transformations applied aim at optimizing for arithmetic intensity and SIMD vectorization, which obviously have a strong impact when arrays and iteration spaces are large.
-The corner cases of this phenomenon are indeed $[q=1, p=1]$ and $[q=4, p=4]$.
-We also observe how \texttt{auto}, in almost all scenarios, outperforms all of the other variants.
-In particular, it is not a surprise that \texttt{auto} is faster than \texttt{fix}, since \texttt{fix} is one of the autotuner's tested variants, as explained in Section~\ref{sec:autotune}.
-This proves the quality of the work presented in this paper, which shows significant advances over~\cite{Luporini}.
-The reasons for which \texttt{auto} exceeds both original code and \texttt{ffc} are discussed for each specific problem next.
-Also, details on the ``optimal'' code variant determined by autotuning are given in Section~\ref{sec:auto-analysis}.
+The first, general observation is that optimality, as we defined it, does not necessarily imply best achievable performance.
+There are two reasons for this.
+The first reason is that low-level optimizations can impact the actual runtime.
+Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$.
+With $p \in [2,3]$, \texttt{tens} tend to slightly outperform \texttt{cfO2} even though pre-evaluation is triggered.
+With $p=4$, \texttt{tens} is quite faster then \texttt{cfO2}; in particular, in this case \texttt{cfO2} considers pre-evaluation unprofitable.
+The motivation is the aggressive unrolling performed in \texttt{tens}, which allows to eliminate any operations involving zero-valued entries.
+In COFFEE we never unroll loops to maximize the chances of autovectorization, which results in considerable speed ups throughout the majority of test cases.
+The second reason is a direct consequence of our axiomatic modeling of optimality: if we are in a regime as in Section~\ref{sec:heuristic-int-dep}, then other code generation strategies for the integration loop might result in larger performance improvements.
+This is the primary reason we are studying \texttt{Hyperelasticity}, a form characterized by extensive tensor algebra.
+Nevertheless, as we will see, we found \texttt{cfO2} to be the best scheme also in this circumstance.
 
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{Mass results.}\label{fig:mass}
+\caption{The mass matrix performance evaluation.}\label{fig:mass}
 \end{figure}

### 2015-08-04 TOMS_Paper/assembly.tex
Elaborating on the perf results

@@
 The Intel \texttt{icc 15.2} compiler was used.
-The compilation flags used were \texttt{-O3, -xHost, -ip}.
+The compilation flags used were \texttt{-O3, -xHost, -ip}, which can trigger \texttt{AVX} autovectorization.
 
@@
 
-Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}, which will be commented in Section~\ref{sec:perf-results-forms}, are "nested" plots.
+The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are "nested" plots.
 We refer to the outer plot as the "grid", while the inner are the actual "plots".
@@
 There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.
+
+%TODO : can we just say padding is useless because in 3D ndofs is a multiple of our vector length?
 
@@
 With $p=4$, \texttt{tens} is quite faster then \texttt{cfO2}; in particular, in this case \texttt{cfO2} considers pre-evaluation unprofitable.
-The motivation is the aggressive unrolling performed in \texttt{tens}, which allows to eliminate any operations involving zero-valued entries.
+The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate any operations involving zero-valued entries and 2) can drastically decrease the working set by not having to store entire temporaries.
 In COFFEE we never unroll loops to maximize the chances of autovectorization, which results in considerable speed ups throughout the majority of test cases.
@@
 
+A second observation concerns the triggering of pre-evaluation in \texttt{cfO2}.
+The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} means that pre-evaluation was found unprofitable.
+The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends to not be performed.
+This is a consequence of the larger increase factor, $\iota$, which we defined in Section~\ref{sec:op_count}.
+
+Also note some other minor things.
+The lack of improvements in all code variants with $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly.
+The slow downs experienced with \texttt{quad} should be imputable to the presence of indirection arrays in the generated code; when the tables are dense, these eventually break vectorization for a marginal gain due to avoiding iteration over isolated zero-valued columns.
+A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of \texttt{Hyperelasticity})
+
 \paragraph{Mass}
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{The mass matrix performance evaluation.}\label{fig:mass}
+\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
 \end{figure}
+Apart from a very small number of points in our exploration space, from analysis of Figure~\ref{fig:mass} we can infer how \texttt{cfO2} represents the best code variant among the tested.
+This suggests that our definition of optimality is consistent in what is the simplest possible bilinear form we can test.
+As we increase the complexity, by adding more coefficients or by varying the polynomial orders, \texttt{cfO2} is always capable of generating optimal loop nests.
+The biggest exception is $[q=4, p=4, nf=1]$ in which the save on operations over zero-valued entries in tensor mode is considerably higher than the save in flops in \texttt{cfO1} and \texttt{cfO2}.
 
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
-\caption{Helmholtz results.}\label{fig:helmholtz}
+\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
 \end{figure}
-
-%The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}.
-We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$.
-This is a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler optimizations, SIMD auto-vectorization.
-The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$.
-This is due to the fact that if the loop over quadrature points is relatively small, then close-to-peak performance is obtainable through basic expression rewriting and code specialization; in this circumstance, generalized loop-invariant code motion and padding plus data alignment.
-The trend changes dramatically as $nf$ and $q$ increase: a more ample spectrum of transformations must be considered to find the optimal local assembly implementation.
-We will provide details about the selected transformations in the next section.
+Figure~\ref{fig:helmholtz} shows the performance improvements achieved by various code variants over the original code can be significant.
+It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in most of the cases.
+One can also appreciate the evidence that the relative order of the employed function spaces plays a key role in the code synthesis choice.
+Take, for example, the case $[q=3, p=1, nf=2]$.
+The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over all other variants.
+On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceed the threshold within which tensor mode is applied), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.
 
@@
 \paragraph{Hyperelasticity}
-%\begin{figure}[t]  
-%...
-%%\includegraphics[scale=0.7]{perf-results/hyperelasticity}
-%%\caption{Hyperelasticity results.}\label{fig:hyperelasticity}
-%\end{figure}
+\begin{figure}[t]
+ \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
+\caption{Performance evaluation for the bilinear form arising in a Hyperelastic model}\label{fig:helmholtz}
+\end{figure}
+
 %

### 2015-08-04 TOMS_Paper/assembly.tex
Update hyperelasticity + minor fixes

@@
 \begin{abstract}
-...
+We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms.
+Optimality is defined in terms of floating point operations required to compute a result.
+We establish a model under which optimality can systematically be reached and comment on the general applicability of this model.
+The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived.
+Extensive performance evaluation proves the robustness of the model, showing considerable performance improvements over a number of alternative code generation systems.
+The effect of low-level optimization is also discussed.
 \end{abstract}
@@
 \item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
-\item We implemented the theory in our compiler, COFFEE\footnote{...}, which is the default optimization system in the Firedrake framework.
+\item We implemented the theory in our compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
+The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework.
 Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
@@
 %\end{figure}
-%Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem.
-This form uses a vector-valued space for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance.
-The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$.
-The use of inderection arrays allows to phisically eliminate zero-valued columns at code generation time; as a consequence, different tabulated basis functions are merged into a single array.
-Therefore, despite the execution being purely scalar because of indirection arrays, the reduction in arithmetic intensity and register pressure imply improvement in performance.
-Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase.
-%
-%For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceed 1$\times$.
-On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.
-%
-%It is worth noting that speed ups of \texttt{auto} over \texttt{fix} decrease at $q=4$, particularly for low values of $p$.
-As we will discuss in Section~\ref{sec:auto-analysis}, this is because at $q=4$ the vector-register tiling transformation (in combination with loop unroll-and-jam) leads to the highest performance.
-In principle, vector-register tiling can be used in combination to the zero-avoidance technique; however, due to mere technical limitations, this is currently not supported in COFFEE.
-Once solved, we expect much higher speed ups in the $q=4$ regime as well.
 
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
-\caption{Performance evaluation for the bilinear form arising in a Hyperelastic model}\label{fig:helmholtz}
+\caption{Performance evaluation for the bilinear form arising in a Hyperelastic model}\label{fig:hyperelasticity}
 \end{figure}
-
-%
-%Speed ups for the hyperelasticity form are shown in Figure~\ref{fig:hyperelasticity}. Experiments for $nf \geq 2$ could not be executed because of FEniCS-Form-Compiler's technical limitations. 
-%
-%For \texttt{auto}, massive speed ups for $q \geq 2$ are to be ascribed to aggressive and successful expression writing.
-Hyperelasticity problems are really compute-intensive, with thousands of operations being performed, so reductions in redundant and useless computation are crucial.
-Complex forms like hyperelasticity would benefit from further ``specialized'' optimizations: for example, it is a known technical limitation of COFFEE that, in some circumstances, less temporaries could (should) be generated and that hoisted code could (should) be suitably distributed over different loops to minimize register pressure (e.g.
-COFFEE could apply loop fission for obtaining significantly better register usage).
-We expect to obtain considerably faster code once such optimizations will be incorporated.
-%
-%In the regime $q \geq 2$ and $nf=1$, peformance improvements are less pronounced moving from $p=1$ to $p=2$, although still significant; in particular, we notice a drop at $p=2$, followed by a raise up to $p=4$.
-It is worth observing that this effect is common to all sets of optimizations.
-The hypothesis is that this is due to the way coefficient functions are evaluated at quadrature points (identical in all configurations), which cannot be easily vectorized unless a change in storage layout and loops order is implemented in the code (abstract syntax tree) generator on top of COFFEE.
+In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}.
+This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction.
+Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques to this purpose.
+Results are displayed in Figure~\ref{fig:hyperelasticity}.
+\texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions.
+\texttt{ufls} performs generally well and exhibits significant speed ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.
 

### 2015-08-05 TOMS_Paper/assembly.tex
Several minor fixes

@@
 The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived.
-Extensive performance evaluation proves the robustness of the model, showing considerable performance improvements over a number of alternative code generation systems.
+Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems.
 The effect of low-level optimization is also discussed.
@@
 In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language.
-The mathematical specification is manipulated by a form compiler, which then generates a representation of assembly operators.
-By applying these operators to an element in the discretized domain a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed.
+The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators.
+By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed.
 The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.
@@
 Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context.
-The lack of suitable third-party tools has led to the development of a number of domain-specific transformation systems or synthesizers.
-In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations that a user should not be expected to write ``by hand''.
+The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers.
+In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''.
 In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count.
 In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed.
-There is also an on-going effort to produce a novel form compiler, and therefore a novel code generation system for assembly operators, called UFLACS (\cite{Uflacs}).
+It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.
 
-In spite of a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-With this paper, we aim to make a first step in the direction of solving this problem.
+However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
+With this paper, we aim to make a first step towards solving this problem.
 
@@
 \item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
-\item We implemented the theory in our compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
+\item We implement the theory in a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
 The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework.
 Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
-\item We provide an unprecedented (for extension) performance evaluation comparing several code transformation/synthesizers systems in a wide range of forms of increasing complexity, which is essential to demonstrate our optimality claim.
-\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we deal with (i.e., optimize) them
+\item We provide an unprecedented (for its extension) performance evaluation comparing several code generation or transformation systems in a range of forms of increasing complexity.
+This is essential to demonstrate our optimality claim.
+\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we handle them
 \item We enrich the set of low-level transformations described in~\citeN{Luporini}
@@
 \label{sec:background}
-
-%is the computation of contributions of a specific cell in the discretized domain to the linear system which yields the PDE solution.
-The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector [Olgaard and Wells 2010; AMCG 2010], whose sizes depend on the order of the method.
-This operation is applied to all cells in the discretized domain.
-In this work we focus on local matrices, or “element matrices”, which are more costly to compute than element vectors.
-
 We review finite element integration and possible implementations using notation and examples adopted in~\cite{quadrature1} and~\cite{Francis}.
@@
 \end{equation}
-The matrix $A$ and the vector $b$ are computed in the so called assembly phase.
-Then, in a subsequent phase, the linear system is solved, usually by means of an iterative method, and $\textbf{u}$ is eventually evaluated.
+The matrix $A$ and the vector $b$ are "assembled" and subsequently used to solve the linear system through (typically) an iterative method.
 
 We focus on the assembly phase, which is often characterized as a two-step procedure: \textit{local} and \textit{global} assembly.
-Local assembly is the subject of the paper: this is about computing the contributions that an element in the discretized domain provide to the approximated solution of the equation.
-Global assembly, on the other hand, is the process of suitably ``inserting'' such contributions in $A$ and $b$.
+Local assembly is the subject of this article.
+It consists of computing the contributions of a single element in the discretized domain to the equation solution.
+In global assembly, such local contributions are "coupled" by suitably inserting them into $A$ and $b$.
 
-Without loss of generality, we illustrate local assembly in a concrete example; that is, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
+We illustrate local assembly in a concrete example, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
 \begin{equation}
@@
 This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated.
-The numerical integration routine based on quadrature representation over an element $K$ can be expressed as follows
+The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
 \begin{equation}
@@
 \subsection{Tensor Contraction Mode}
-Starting from Equation~\ref{eq:quadrature}, exploiting basic mathematical properties we can rewrite the expression as
+Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
 \begin{equation}
@@
 A generalization of this transformation has been proposed in~\cite{Kirby:TC}.
-By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in a temporary.
+By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables.
 The evaluation of the local tensor can then be abstracted as
@@
 \subsection{Qualitative Comparison}
-Depending on the form being considered, the relative performance of the two modes, in terms of number of operations executed, can vary even quite dramatically.
-The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode increasingly more indicate for "complex" forms.
+Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically.
+The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for "complex" forms.
 On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains "sufficiently small".
+The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.
 
-These two modes have been implemented in the FEniCS Form Compiler.
-In this compiler, a simple heuristic is used to choose the most suitable mode for a given form.
-It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically~\cite{Fenics}.
+These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-Compiler}).
+In this compiler, a heuristic is used to choose the most suitable mode for a given form.
+It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}).
 We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}.
-For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of selecting the optimal mode at the monomials level -- no heuristics, no "global" choice of the mode for all monomials in the form -- without affecting the cost of code generation.
+For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of reasoning at the monomials level -- not at the form level -- without affecting the code generation cost.
 
@@
 \begin{Def}[Perfect and imperfect loop nests]
-A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said \textit{imperfect}. 
+A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said imperfect. 
 \end{Def}
 
-A straightforward property of perfect nests is that hoisting invariant expressions from the innermost loop to the preheader (i.e., the block that precedes the entry point of the nest) is a safe transformation.
+\begin{Def}[Independent basic block]
+A sequence of statements that present no data dependencies compose an independent basic block
+\end{Def}
+
+We focus on perfect nests whose innermost loop body is an independent basic block.
+A straightforward property of this class is that hoisting invariant expressions from the innermost to any of the outer loops or the preheader (i.e., the block that precedes the entry point of the nest) is always safe, as long as the obvious loop dependencies are preserved.
 We will make use of this property.
+The results of this section could also be generalized to larger classes of loop nests, in which basic block independence does not hold, although this would require a refinement that goes beyond the scope of this paper.
 
 % In the following, we should represent loops by their iteration variable.
+We find convenient abstracting mathematical properties at the level of loops
 
@@
 
-In this work, we are particularly interested in the following class since it naturally arises from the math described in Section~\ref{sec:background}.
+In particular, we are interested in the following class since it naturally arises from the math of finite element integration, reviewed in Section~\ref{sec:background}.
 
@@
 
-Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $k$.
+Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.
 
@@
 \label{prop:share-removal}
-Sharing in a perfect multilinear loop nest $LN = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ can always be eliminated.
+Assume $LN = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a perfect multilinear loop nest with the body of $L_{i_{n-1}}$ being an independent basic block.
+Then sharing can always be eliminated from $LN$.
 \end{Prop}
@@
 The demonstration is by construction and exploits linearity.
-We want to transform $LN$ into $LN'$ such that there is no sharing is any $L_i \in LN'$.
-Starting from the innermost loop $L_{i_{n-1}}$, the expressions are ''flattened'' by expanding all products involving terms depending on $L{i_{n-1}}$.
-Being on the same level of the expression tree, such terms can then be factorized.
+We want to transform $LN$ into $LN'$ such that there is no sharing in any $L_i \in LN'$.
+Starting from the innermost loop $L_{i_{n-1}}$, the expressions are ''flattened'' by expanding all products involving terms depending on $L_{i_{n-1}}$.
+Being now on the same level of the expression tree, such terms can be factorized.
 Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$, and such term is now unique in the expression.
 The other terms, independent of $L_{i_{n-1}}$, are, by definition, loop-invariant, and as such can be hoisted at the level of $L_{i_{n-2}}$.
-This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is always safe.
+This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is safe.
 \end{proof}
 
-This result will be essential to prove that optimality, characterized as follows, can structurally be reached in this class of loops.
+This result will be essential to prove that optimality, characterized as follows, can always be reached in this class of loops.
 
@@
 \label{prop:optimal-mln}
-A perfect multilinear loop nest $LN$ can always be reduced to optimal form.
+A perfect multilinear loop nest $LN$ with basic block independence can always be reduced to optimal form.
 \end{Prop}
@@
 These sets are sorted in descending order based on their cardinality.
-By establishing a one-to-one mapping between set indices and loop indices, we produce a new loop permutation.
-The loop permutation is semantically correct because of perfectness.
-In this new order, loops are placed such that as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$.
+We produce a new loop permutation that reflects the ${S_i}$ ordering; that is, loops are placed such that, as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$.
+Note the loop permutation is semantically correct because of perfectness.
 Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
@@
 The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication.
-We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $LN$ as long as data dependencies are preserved (note that $LN$ is imperfect).
+We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $L_k$ as long as data dependencies are preserved (note that $LN$ is imperfect).
 One could think of searching for sub-expressions independent of $L_e$, for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
-However, finding or exposing such reducible sub-expressions requires, in general, a full exploration of the expression tree transformation space.
+However, finding or exposing such sub-expressions requires, in general, a full exploration of the expression tree transformation space.
 This is obviously challenging, and so is the problem of synthesizing an optimal $LN'$ starting from $LN$.
-Even though we assumed we could generate $LN'$ in non-exponential time, the following issues should be addressed
+Even though we assumed we could generate $LN'$ through a suitable transformation algorithm, the following issues should be addressed
 \begin{itemize}
 \item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
-This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime;
-\item the transformations exposing multi-invariant sub-expressions could require expansion and factorization.
-In terms of number of operations, the save originating from the elimination of the reduction loop could be overwhelmed by the increase in arithmetic complexity of $L_k$ (e.g., expansion can increase the operation count, e.g.
+This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
+\item the transformations exposing invariant sub-expressions could require expansion and factorization.
+In terms of number of operations, the save originating from hoisting could be overwhelmed by the increase in arithmetic complexity within $L_k$ (expansion can increase the operation count, e.g.
 $A(B+C) = AB + AC$).
@@
 
-We then refine our definition of optimality for generic loop nests as follows
+We shall then refine our definition of optimality for generic loop nests as follows
 
@@
 Note how the definition contemplates the possibility for a nest to have multiple innermost loops.
-In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or result of suitable transformations.
+In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations.
 

### 2015-08-05 TOMS_Paper/assembly.tex
Add elasticity + minor fixes

@@
 We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $L_k$ as long as data dependencies are preserved (note that $LN$ is imperfect).
-One could think of searching for sub-expressions independent of $L_e$, for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
+One could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
 However, finding or exposing such sub-expressions requires, in general, a full exploration of the expression tree transformation space.
@@
 
-From domain knowledge, we make the following observations.
+We make the following observations.
 1) $L >>> M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
@@
 Imagine a time stepping loop $L_t$ wraps $LN$.
-One could then think of identifying time-invariant sub-expressions that access both geometry and reference element terms and pre-evaluate them within $L_t$.
+One could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and pre-evaluate them within $L_t$.
 Unless adopting complex engineering solutions (e.g.
-aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by $O(L)$.
-It is our opinion, therefore, that the drop in number of operations would be overwhelmed, from the run-time viewpoint, by the larger memory pressure.
+aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by a factor $L$.
+It is our opinion that the drop in number of operations would be overwhelmed, from the run-time viewpoint, by a much larger memory pressure.
 
@@
 We will provide precise details about this in the following sections.
-For the moment, and just as one of many possible examples, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.
+For the moment, and just as an example, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.
 
@@
 \end{enumerate}
-A corollary of $C1$ is that hoisting expressions involving geometry terms outside of $L_e$ becomes forbidden.
+A corollary of $C_1$ is that hoisting expressions involving geometry terms outside of $L_e$ is now forbidden. Consequently, the search space for optimality becomes smaller.
 
 \subsection{Minimizing the Operation Cost}
-Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the number of operations in all innermost loops is minimum.
+Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the operation cost in the innermost loops is minimum.
 We now discuss how we can systematically achieve this.
@@
 
-The challenge is to understand when, and for which monomials, pre-evaluation is profitable. We propose an algorithm and a discussion of its optimality.
-
-The intuition of the algorithm for optimal loop nest synthesis is shown in Figure~\ref{code:intuition}.
-
-% local optimization of monomials
-% local sharing elimination
-% global sharing elimination
+The challenge is to understand when, and for which monomials, pre-evaluation is profitable.
+We propose an algorithm and a discussion of its optimality.
+The intuition of the main algorithm for reducing a loop nest to optimal form is shown in Figure~\ref{code:intuition}.
 
@@
 We first comment on the assumptions.
-1) How to create an optimal cost function is discussed in Section~\ref{sec:op_count}.
-2) A pathological case due to symmetries in basis functions, which in practice rarely happens.
-3) This could occur in complex forms with several monomials; for simplicity, we ignore this situation (otherwise, a "global" analysis of the monomials would be required).
+1) The creation of an optimal cost function is discussed in Section~\ref{sec:op_count}.
+2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions.
+This is however a pathological case.
+3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
+For simplicity, we do not handle this case, which otherwise would require a "global" analysis of the monomials in the form.
+Further, even though, in practice, assumptions 2) and 3) did not hold, the resulting loop nest would still be nearly-optimal.
 
 We distinguish two classes of loop nests rooted in $LN$: $[L_e, L_j, L_k]$, for the pre-evaluated monomials in $P$, and $[L_e, L_i, L_j, L_k]$, enclosing the remaining monomials in $Z$.
-Since they only differ for the presence of $L_i$, we relieve notation by omitting all shared loops when discussing operation counts.
-In particular, we use $I$ to refer to the iteration space size of $L_i$.
-The operation count of what we are proving to be the optimal $LN$ synthesis is, therefore, $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation cost of monomials in $P$ and $Z$, respectively.
+Since they only differ for the presence of $L_i$, we relieve the notation by omitting the shared loops when discussing operation counts.
+The operation count of what we are proving to be the optimal $LN$ synthesis is $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation cost of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.
+This can also be inferred from Figure~\ref{code:loopnest-opt}.
 
-We start noting that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting.
+We note that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting.
 This narrows the proof to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $LN_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $LN_{ops}$.
@@
 It is rather obvious that $LN_{ops_1}' \geq LN_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
-We note that if monomials in $Z_P$ share symbols with $\overline{Z} = Z \setminus Z_P$, then we have $LN_{ops_2}' = LN_{ops_2}$, so our statement is true.
-If, on the other hand, at least one monomial does not share any symbols, we obtain $LN_{ops_2}' < LN_{ops_2}$ or, equivalently, $LN_{ops_2}' = LN_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by exposing more pre-evaluations, $LN_{ops_1}' \geq LN_{ops_1} +  I \cdot\delta$ holds.
-This is indeed the case since we rely on assumption 2, which ensures the uniqueness of the pre-evaluated tables (i.e., the absence of sharing) and, therefore, the optimality of $LN$.
+We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $LN_{ops_2}' = LN_{ops_2}$, so our statement is true.
+If, on the other hand, at least one monomial does not share any terms, we obtain $LN_{ops_2}' < LN_{ops_2}$ or, equivalently, $LN_{ops_2}' = LN_{ops_2} - I \cdot\delta$.
+What we have to show now is that even by exposing more pre-evaluations, $LN_{ops_1}' - LN_{ops_1} =  \gamma \geq I \cdot\delta$ holds.
+Because of assumption 2), the new pre-evaluations cannot expose further sharing.
+Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.
 
-B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, with the cost function being optimal by assumption.
-Assumption 3 guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities.
+B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1).
+Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities.
 The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$.
-This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are "declassed" from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so aggregating $p$ in $Z$ does not increase the operation count.
+This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are "de-classed" from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count.
 \end{proof}
@@
 By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient).
-Intuitively, $\iota$ captures this growth in non-hoistable operation.
+Intuitively, $\iota$ captures such a growth in non-hoistable operation.
 
@@
 Expansion would then lead to identical terms (i.e., identical pre-evaluated tables).
-Therefore, for a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the arity of the multilinear loop nest and $n$ the set of terms involved in the coefficient expansion.
+Therefore, for a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are considering sets, so duplicates would be counted as 1).
 
@@
 \begin{itemize}
-\item optimized quadrature mode, \texttt{quad}
-\item tensor mode, \texttt{tens}
+\item optimized quadrature mode, \texttt{quad} (work presented in~\citeN{quadrature1})
+\item tensor mode, \texttt{tens} (work presented in~\citeN{FFC-TC})
 \item automatic mode, \texttt{auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
@@
 However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.
-For a fair comparison, small patches (publicly available) were written to be able to use the aforementioned transformation systems in Firedrake; that is, all simulations are run through Firedrake.
-UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS (\cite{Fenics}), which employs a data storage layout different than Firedrake.
-Our small patches fix this problem by producing code with a "correct" data storage layout.
 
-Following the methodology (and the notation) used in ~\cite{quadrature1} and~\cite{Francis}, we increasingly vary the complexity of each form. In particular:
+For a fair comparison, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants.
+UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS, which employs a data storage layout different than Firedrake.
+Our patches fix this problem by producing code with a data storage layout as expected by Firedrake.
+
+Following the methodology (and the notation) used in ~\citeN{quadrature1} and~\citeN{Francis}, we increasingly vary the complexity of each form. In particular:
 \begin{itemize}
@@
 The slow downs experienced with \texttt{quad} should be imputable to the presence of indirection arrays in the generated code; when the tables are dense, these eventually break vectorization for a marginal gain due to avoiding iteration over isolated zero-valued columns.
-A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of \texttt{Hyperelasticity})
+A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of \texttt{Hyperelasticity}).
+Finally, note that the bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 
 \paragraph{Mass}
-\begin{figure}[t] 
+\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
@@
 \paragraph{Helmholtz}
-\begin{figure}[t] 
+\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
@@
 \paragraph{Elasticity}
-%\begin{figure}[t]  
-%...
-%%\includegraphics[scale=0.7]{perf-results/elasticity}
-%%\caption{Elasticity results.}\label{fig:elasticity}
-%\end{figure}
+\begin{figure}
+\includegraphics[scale=0.7]{perf-results/elasticity}
+\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
+\end{figure}
+...
 
 \paragraph{Hyperelasticity}
-\begin{figure}[t]
+\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
-\caption{Performance evaluation for the bilinear form arising in a Hyperelastic model}\label{fig:hyperelasticity}
+\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
 \end{figure}

### 2015-08-05 TOMS_Paper/assembly.tex
Add conclusions + minor fixes

@@
 
-\keywords{Finite element integration, local assembly, compilers, optimizations, SIMD vectorization}
+\keywords{Finite element integration, local assembly, compilers, performance optimization}
 
@@
 through contract W911NF-10-1-000, and by a HiPEAC collaboration
-grant. The authors would like to thank Dr. Carlo Bertolli,
+grant. The authors would like to thank Mr. Andrew T.T. McRae,
 Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
@@
 \label{sec:conclusions}
-...
+With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration.
+The results are extremely encouraging, suggesting our model applies to a variety of contexts.
+We have discussed the conditions under which our optimality definition becomes weaker.
+An open problem is understanding how to refine this definition to include outer loops.
+This will probably require studying and exploiting properties of different differential operators.
+A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the high order function spaces.
 

### 2015-08-06 TOMS_Paper/assembly.tex
Several minor fixes in perf eval section

@@
 \label{sec:perf-results-forms}
-The first, general observation is that optimality, as we defined it, does not necessarily imply best achievable performance.
+% 52 test cases per form, so 52*4 = 208 test cases in total. In 7, there is a non-marginal slow down in cfO2.
+
+The first observation is that our optimality model does not imply minimum execution time.
+In particular, in the 0.03$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime.
 There are two reasons for this.
-The first reason is that low-level optimizations can impact the actual runtime.
+The former is that low-level optimization can have a significant impact.
 Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$.
-With $p \in [2,3]$, \texttt{tens} tend to slightly outperform \texttt{cfO2} even though pre-evaluation is triggered.
-With $p=4$, \texttt{tens} is quite faster then \texttt{cfO2}; in particular, in this case \texttt{cfO2} considers pre-evaluation unprofitable.
-The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate any operations involving zero-valued entries and 2) can drastically decrease the working set by not having to store entire temporaries.
-In COFFEE we never unroll loops to maximize the chances of autovectorization, which results in considerable speed ups throughout the majority of test cases.
-The second reason is a direct consequence of our axiomatic modeling of optimality: if we are in a regime as in Section~\ref{sec:heuristic-int-dep}, then other code generation strategies for the integration loop might result in larger performance improvements.
-This is the primary reason we are studying \texttt{Hyperelasticity}, a form characterized by extensive tensor algebra.
-Nevertheless, as we will see, we found \texttt{cfO2} to be the best scheme also in this circumstance.
+With $p \in [2,3]$, \texttt{tens} tends to slightly outperform \texttt{cfO2}.
+With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}.
+The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries.
+In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases.
+The latter reason is simply inherent to our optimality model.
+As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop.
+The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.
 
-A second observation concerns the triggering of pre-evaluation in \texttt{cfO2}.
-The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} means that pre-evaluation was found unprofitable.
-The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends to not be performed.
-This is a consequence of the larger increase factor, $\iota$, which we defined in Section~\ref{sec:op_count}.
+A second observation concerns the "triggering" of pre-evaluation in \texttt{cfO2}.
+The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable.
+The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed.
+This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.
 
 Also note some other minor things.
-The lack of improvements in all code variants with $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly.
-The slow downs experienced with \texttt{quad} should be imputable to the presence of indirection arrays in the generated code; when the tables are dense, these eventually break vectorization for a marginal gain due to avoiding iteration over isolated zero-valued columns.
-A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of \texttt{Hyperelasticity}).
-Finally, note that the bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
+\begin{itemize}
+\item The lack of improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
+\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
+\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of the hyperelastic model).
+\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
+ \end{itemize}
 
@@
 \end{figure}
-Apart from a very small number of points in our exploration space, from analysis of Figure~\ref{fig:mass} we can infer how \texttt{cfO2} represents the best code variant among the tested.
-This suggests that our definition of optimality is consistent in what is the simplest possible bilinear form we can test.
-As we increase the complexity, by adding more coefficients or by varying the polynomial orders, \texttt{cfO2} is always capable of generating optimal loop nests.
-The biggest exception is $[q=4, p=4, nf=1]$ in which the save on operations over zero-valued entries in tensor mode is considerably higher than the save in flops in \texttt{cfO1} and \texttt{cfO2}.
+From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} represents the best variant among the ones tested, apart from the few points previously discussed.
+This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints.
+It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$.
 
@@
 \end{figure}
-Figure~\ref{fig:helmholtz} shows the performance improvements achieved by various code variants over the original code can be significant.
-It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in most of the cases.
-One can also appreciate the evidence that the relative order of the employed function spaces plays a key role in the code synthesis choice.
+Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant.
+We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice.
 Take, for example, the case $[q=3, p=1, nf=2]$.
-The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over all other variants.
-On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceed the threshold within which tensor mode is applied), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.
+The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over the competitors.
+On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.
 
@@
 This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction.
-Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques to this purpose.
+Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques.
 Results are displayed in Figure~\ref{fig:hyperelasticity}.
@@
 The results are extremely encouraging, suggesting our model applies to a variety of contexts.
-We have discussed the conditions under which our optimality definition becomes weaker.
-An open problem is understanding how to refine this definition to include outer loops.
-This will probably require studying and exploiting properties of different differential operators.
-A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the high order function spaces.
+We have discussed the conditions under which the optimality model becomes weaker.
+An open problem is understanding how to refine this model to include outer loops.
+This will probably require exploiting properties of differential operators.
+A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.
 

### 2015-08-06 TOMS_Paper/assembly.tex
Several minor fixes

@@
 This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
-General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or "flops", than necessary).
+General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary).
 Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context.
@@
 \end{equation}
-The matrix $A$ and the vector $b$ are "assembled" and subsequently used to solve the linear system through (typically) an iterative method.
+The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.
 
@@
 It consists of computing the contributions of a single element in the discretized domain to the equation solution.
-In global assembly, such local contributions are "coupled" by suitably inserting them into $A$ and $b$.
+In global assembly, such local contributions are ``coupled'' by suitably inserting them into $A$ and $b$.
 
@@
 \end{equation}
-in which the pre-evaluated "reference tensor" $A_{i_1 i_2 \alpha}$ and the cell-dependent "geometry tensor" $G_{K}^\alpha$ are exposed.
+in which the pre-evaluated ``reference tensor'' $A_{i_1 i_2 \alpha}$ and the cell-dependent ``geometry tensor'' $G_{K}^\alpha$ are exposed.
 
@@
 Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically.
-The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for "complex" forms.
-On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains "sufficiently small".
+The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for ``complex'' forms.
+On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''.
 The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.
 
-These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-Compiler}).
+These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-TC}).
 In this compiler, a heuristic is used to choose the most suitable mode for a given form.
@@
 
-A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of "some level of local memory" (e.g.
+A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
@@
 As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
-In the following, we simply refer to this special sort of code hoisting as "\textit{pre-evaluation}".
+In the following, we simply refer to this special sort of code hoisting as ``\textit{pre-evaluation}''.
 
@@
 The point of departure consists of understanding the impact, as number of operations saved or introduced, of pre-evaluation.
-This is studied "locally"; that is, for each monomial, in isolation.
+This is studied ``locally''; that is, for each monomial, in isolation.
 If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed.
-The result is a set of $n$-dimensional tables (these can be seen as "slices" of the reference tensor at the math level), $n$ being the arity of the multilinear form.
+The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.
 Identical tables are mapped to the same temporary.
@@
 
-Before elaborating on the profitability of pre-evaluation, we need to discuss under which conditions this approach, based on a "local analysis" of monomials, is optimal.
+Before elaborating on the profitability of pre-evaluation, we need to discuss under which conditions this approach, based on a ``local analysis'' of monomials, is optimal.
 
@@
 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
-For simplicity, we do not handle this case, which otherwise would require a "global" analysis of the monomials in the form.
+For simplicity, we do not handle this case, which otherwise would require a ``global'' analysis of the monomials in the form.
 Further, even though, in practice, assumptions 2) and 3) did not hold, the resulting loop nest would still be nearly-optimal.
@@
 The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$.
-This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are "de-classed" from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count.
+This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count.
 \end{proof}
@@
 \subsection{Automation through the COFFEE Language}
-As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric "building-block" operations.
+As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric ``building-block'' operations.
 This has several advantages.
@@
 For each of this statements, we try to remove sharing by greedily applying factorization and code motion.
-In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named "factorize", "expand", etc.).
+In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named ``factorize'', ``expand'', etc.).
 
@@
 
-% say that the reasons to have zeros is to keep the design of the form compiler simple, which just "puts zeros" and "outline the assembly expression".
+% say that the reasons to have zeros is to keep the design of the form compiler simple, which just ``puts zeros'' and ``outline the assembly expression''.
 % rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
@@
 We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
-In particular, we study the mass matrix ("\texttt{Mass}") and the bilinear forms arising in a Helmholtz equation ("\texttt{Helmholtz}"), in an elastic model ("\texttt{Elasticity}"), and in a hyperelastic model ("\texttt{Hyperelasticity}").
+In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}'').
 The complete specification of these forms is made publicly available (\cite{ufl-code}).
@@
 \begin{itemize}
-\item the FEniCS Form Compiler's built-in optimization systems, namely
-\begin{itemize}
-\item optimized quadrature mode, \texttt{quad} (work presented in~\citeN{quadrature1})
-\item tensor mode, \texttt{tens} (work presented in~\citeN{FFC-TC})
-\item automatic mode, \texttt{auto} (here, heuristics are used to infer the optimal mode for the form between quadrature and tensor)
-\end{itemize}
-\item a novel back-end for the FEniCS Form Compiler, UFLACS, referred to as \texttt{ufls}, whose primary goals are improved code generation time and runtime
-\item the sole generalized loop-invariant code motion in COFFEE plus padding, \texttt{cfO1}, as discussed in~\citeN{Luporini}
-\item the optimal multilinear loop nest synthesis plus padding and symbolic execution, \texttt{cfO2}, which is the focus of this paper
+\item FEniCS Form Compiler's built-in optimization system: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
+\item FEniCS Form Compiler's built-in optimization system: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
+\item FEniCS Form Compiler's built-in optimization system: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, discussed in~\citeN{Fenics}).
+Referred to as \texttt{auto}
+\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
+\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
+\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
 \end{itemize}
 
-The values that we report are the average of three runs with "warm cache" (no code generation time, no compilation time).
+The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time).
 They include the cost of local assembly as well as the cost of matrix insertion.
@@
 
-Following the methodology (and the notation) used in ~\citeN{quadrature1} and~\citeN{Francis}, we increasingly vary the complexity of each form. In particular:
+Since we are running sequential programs, with respect to the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = \infty$; that is, temporaries can occupy an arbitrary amount of memory.
+This is because a single process that has exclusive access to all of the bandwidth available in the system can count on pre-fetching to minimize cache misses.
+If, on the other hand, a parallel program were run, then a more sensible value for $T_H$ would be the size of the last level of private cache; this would avoid memory bank conflicts and, therefore, potential drops in runtime.
+
+Following the methodology adopted in ~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
 \begin{itemize}
@@
 
-The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are "nested" plots.
-We refer to the outer plot as the "grid", while the inner are the actual "plots".
+The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are ``nested'' plots.
+We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''.
 In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis.
@@
 
-A second observation concerns the "triggering" of pre-evaluation in \texttt{cfO2}.
+A second observation concerns the ``triggering'' of pre-evaluation in \texttt{cfO2}.
 The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable.
@@
 \end{figure}
-...
+The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
+We note how the fact that\texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
+This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present.
 

### 2015-08-07 TOMS_Paper/assembly.tex
Say we are using exact quadrature
Update threshold discussion

@@
 
-Since we are running sequential programs, with respect to the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = \infty$; that is, temporaries can occupy an arbitrary amount of memory.
-This is because a single process that has exclusive access to all of the bandwidth available in the system can count on pre-fetching to minimize cache misses.
-If, on the other hand, a parallel program were run, then a more sensible value for $T_H$ would be the size of the last level of private cache; this would avoid memory bank conflicts and, therefore, potential drops in runtime.
+With respect to the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, since we are running sequential programs we set $T_H = LLC_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L3 cache.
+If, on the other hand, a parallel program were run, then a more sensible value for $T_H$ would be the size of the last level of private cache (L2, on our Sandy Bridge machine); this would avoid cache misses and memory conflicts and, therefore, potential drops in runtime.
 
@@
 \item the mesh, tetrahedral with a total of 4374 elements
+\item exact numerical quadrature (the same scheme as in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule, is employed)
 \end{itemize}
@@
 
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
+\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
+\end{figure}
+
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
+\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.7]{perf-results/elasticity}
+\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
+\end{figure}
+
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
+\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
+\end{figure}
+
 \subsection{Performance of Forms}
@@
 Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$.
-With $p \in [2,3]$, \texttt{tens} tends to slightly outperform \texttt{cfO2}.
+With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}.
 With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}.
@@
 \paragraph{Mass}
-\begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
-\end{figure}
 From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} represents the best variant among the ones tested, apart from the few points previously discussed.
@@
 \paragraph{Helmholtz}
-\begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
-\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
-\end{figure}
 Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant.
@@
 \paragraph{Elasticity}
-\begin{figure}
-\includegraphics[scale=0.7]{perf-results/elasticity}
-\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
-\end{figure}
 The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
@@
 This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present.
+With $q=4$, pre-evaluation is triggered in \texttt{cfO2} when $nf0$.
+When $nf=1$, pre-evaluation is found profitable in terms of operation count, but it is not performed because of exceeding the working set threshold $T_H$ imposed by memory constraint $C_2$.
+We have tried running the same experiments with $T_H = \infty$ and noticed a significant drop in runtime for all values of $p$ (\texttt{cfO2} was generally slower than \texttt{cfO1}).
+This confirms the importance of memory constraints.
 
 \paragraph{Hyperelasticity}
-\begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
-\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
-\end{figure}
 In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}.

### 2015-08-07 TOMS_Paper/assembly.tex
Minor fixes

@@
 In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''.
-In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count.
+In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count.
 In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed.
@@
 In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}'').
-The complete specification of these forms is made publicly available (\cite{ufl-code}).
+The complete specification of these forms is made publicly available\footnote{\url{https://github.com/firedrakeproject/firedrake-bench/blob/experiments/forms/firedrake_forms.py}}.
 
@@
 \item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
-\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of the hyperelastic model).
+\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
 \item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
@@
 The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
-We note how the fact that\texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
+We note how the fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
 This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present.
 With $q=4$, pre-evaluation is triggered in \texttt{cfO2} when $nf0$.
-When $nf=1$, pre-evaluation is found profitable in terms of operation count, but it is not performed because of exceeding the working set threshold $T_H$ imposed by memory constraint $C_2$.
+When $nf=1$, pre-evaluation is found profitable in terms of operation count, although it is not performed because of exceeding the working set threshold $T_H$ imposed by memory constraint $C_2$.
 We have tried running the same experiments with $T_H = \infty$ and noticed a significant drop in runtime for all values of $p$ (\texttt{cfO2} was generally slower than \texttt{cfO1}).

### 2015-08-07 TOMS_Paper/assembly.tex
Readjust Th discussion

@@
 
-Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache.
+Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared).
 The Intel Turbo Boost and Intel Speed Step technologies were disabled.
@@
 
-With respect to the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, since we are running sequential programs we set $T_H = LLC_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L3 cache.
-If, on the other hand, a parallel program were run, then a more sensible value for $T_H$ would be the size of the last level of private cache (L2, on our Sandy Bridge machine); this would avoid cache misses and memory conflicts and, therefore, potential drops in runtime.
+For what concerns the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache).
+For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ and $T_H = \infty$.
+Results of these trials are discussed below.
 
@@
 The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
-We note how the fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
+The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
 This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present.
-With $q=4$, pre-evaluation is triggered in \texttt{cfO2} when $nf0$.
-When $nf=1$, pre-evaluation is found profitable in terms of operation count, although it is not performed because of exceeding the working set threshold $T_H$ imposed by memory constraint $C_2$.
-We have tried running the same experiments with $T_H = \infty$ and noticed a significant drop in runtime for all values of $p$ (\texttt{cfO2} was generally slower than \texttt{cfO1}).
-This confirms the importance of memory constraints.
+With $q=4$ and $nf \in [0, 1]$, pre-evaluation in \texttt{cfO2} is found profitable in terms of operation count, although it is not triggered because of exceeding the working set threshold $T_H$.
+Running the same experiments with $T_H = \infty$ resulted in significant runtime increases for all values of $p$.
+Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement for $nf = 0$, with \texttt{cfO2} becoming the fastest alternative.
+Our explanation for this is that by running a sequential program, a process has exclusive access to the memory bandwidth of the system, so pre-fetching minimizes cache misses without hurting concurrent processes (since there are none).
 

### 2015-08-10 TOMS_Paper/assembly.tex
Many small fixes

@@
 We establish a model under which optimality can systematically be reached and comment on the general applicability of this model.
-The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived.
+The model is integrated in a compiler that, given a representation of an integration routine, derives a loop nest in optimal form.
 Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems.
@@
 General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary).
-Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context.
-The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers.
+Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on loop optimization for cache locality, so they are not particularly helpful in our context.
+The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems.
 In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''.
@@
 In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed.
-It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.
+It is also worth mentioning an on-going effort to produce a novel form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.
 
@@
 As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
-\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
-\item We implement the theory in a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
-The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework.
-Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
-\item We provide an unprecedented (for its extension) performance evaluation comparing several code generation or transformation systems in a range of forms of increasing complexity.
-This is essential to demonstrate our optimality claim.
-\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we handle them
-\item We enrich the set of low-level transformations described in~\citeN{Luporini}
+\item We provide an incremental, constructive proof that optimal implementations can be obtain by analyzing the monomials appearing in a form
+\item We automate the translation of forms into optimal loop nests by a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
+The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, the default optimization system in the Firedrake framework.
+\item We provide an unprecedented (for its extension) performance evaluation comparing several code transformation systems in a range of forms of increasing complexity.
+This is essential to demonstrate that our optimality model holds in practice.
+\item We comment on the cases in which our characterization of optimality becomes weaker, and provide an intuition of how we handle such forms
+\item We enrich the set of low-level transformations described in~\citeN{Luporini}. These make COFFEE a full-fledged system capable of decoupling form compilers from low level code generation.
 \end{itemize}
@@
 \label{sec:background}
-We review finite element integration and possible implementations using notation and examples adopted in~\cite{quadrature1} and~\cite{Francis}.
+We review finite element integration using the same notation and examples adopted in~\citeN{quadrature1} and~\citeN{Francis}.
 
@@
 The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces.
-For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$.
+For simplicity, we assume $U = V$.
+Let $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$.
 The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$.
@@
 \subsection{Quadrature Mode}
-Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$.
+Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$.
 For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced.
@@
 We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}.
-For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of reasoning at the monomials level -- not at the form level -- without affecting the code generation cost.
+For the moment, we just recall that one of the goals of this research is to produce an intelligent system that goes beyond the dichotomy between quadrature and tensor modes.
+We will reason in terms of loop nests, code motion, and code pre-evaluation, searching the entire implementation space for an optimal synthesis.
 
@@
 
-In this section, we provide a characterization of optimality for loop nests of various nature. In order to make the document self-contained, we start with reviewing basic compiler terminology.
+In this section, we provide a characterization of loop nest optimality. In order to make the document self-contained, we start with reviewing basic compiler terminology.
 
@@
 % In the following, we should represent loops by their iteration variable.
-We find convenient abstracting mathematical properties at the level of loops
+It is convenient to map mathematical properties to the loop nest level
 
@@
 
-Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.
+Figure~\ref{code:multi_loopnest_a} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.
 
 \begin{figure}[h]\begin{CenteredBox}
-\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}
-\end{CenteredBox}\caption{Multilinear loop nest with sharing}\label{code:multi_loopnest}\end{figure}
+{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
+~~~~~~~~~~~~~~
+{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
+\end{CenteredBox}\caption{A simple multilinear loop nest}\end{figure}
 
@@
 
-For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_optimal}.
-
-\begin{figure}[h]\begin{CenteredBox}
-\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}
-\end{CenteredBox}\caption{Multilinear loop nest in optimal form}\label{code:multi_loopnest_optimal}\end{figure}
+For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.
 
@@
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
-\end{CenteredBox}\caption{Loop nest example}\label{code:loopnest}\end{figure}
+\end{CenteredBox}\caption{An imperfect loop nest, with one reduction loop ($L_i$), an inner perfect multilinear loop nest ($[L_j, L_k]$), and sharing potentially present in $F$}\label{code:loopnest}\end{figure}
 
 The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
-The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication.
-We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $L_k$ as long as data dependencies are preserved (note that $LN$ is imperfect).
-One could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
-However, finding or exposing such sub-expressions requires, in general, a full exploration of the expression tree transformation space.
-This is obviously challenging, and so is the problem of synthesizing an optimal $LN'$ starting from $LN$.
+The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication, possibly characterized by sharing.
+Here, one could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
+This is exemplified through Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored} by fixing $F$ in a simpler loop nest.
+
+\begin{figure}[h]\begin{CenteredBox}
+{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
+~~~~~~~~~~~~~~
+{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
+\end{CenteredBox}\caption{The effect of pre-evaluating reductions. Note the decrease in operation count grows with the loop nest depth.}\end{figure}
+
+It is then obvious that reaching an optimal form $LN'$ is, in the general case, a challenging issue, since a full exploration of the expression tree transformation space is necessary.
 Even though we assumed we could generate $LN'$ through a suitable transformation algorithm, the following issues should be addressed
@@
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
-\item the transformations exposing invariant sub-expressions could require expansion and factorization.
-In terms of number of operations, the save originating from hoisting could be overwhelmed by the increase in arithmetic complexity within $L_k$ (expansion can increase the operation count, e.g.
-$A(B+C) = AB + AC$).
+\item the transformations exposing sub-expressions amenable to pre-evaluation could increase the arithmetic complexity within $L_k$ (for example, expansion may increase the operation count).
+This could overwhelm any gain due to pre-evaluation.
 \end{itemize}
 
-We shall then refine our definition of optimality for generic loop nests as follows
+We shall then refine our definition of optimality for this wider class of loop nests as follows
 
@@
 Note how the definition contemplates the possibility for a nest to have multiple innermost loops.
-In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations.
+In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).
 
@@
 
-In this section, we instantiate Definition~\ref{def:optimality} in our domain of interest, finite element integration.
-This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem, which we reviewed in Section~\ref{sec:background}; and the (partly multilinear) loop nests implementing such forms.
+In this section, we instantiate Definition~\ref{def:optimality} to our domain of interest, finite element integration.
+This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.
 
@@
 $L_e$ represents iteration over the elements of a mesh; $L_i$ derives from using numerical quadrature; $[L_j, L_k]$ implement the computation over test and trial functions.
-We deliberately omitted useless portions of code to not hinder readability (e.g.
-matrix insertion) and to avoid tying our discussion to specific forms (e.g.
-\texttt{F} is unspecified).
+We deliberately omitted useless portions of code to not hinder readability and to avoid tying our discussion to specific forms.
 
@@
 \label{sec:mem-const}
-The fact that the iteration space of $L_e$ is so larger than that of other loops suggests we should be cautious when hoisting out of $LN$.
-Imagine a time stepping loop $L_t$ wraps $LN$.
-One could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and pre-evaluate them within $L_t$.
-Unless adopting complex engineering solutions (e.g.
-aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by a factor $L$.
-It is our opinion that the drop in number of operations would be overwhelmed, from the run-time viewpoint, by a much larger memory pressure.
+The fact that $L >>> M, N, O$ suggests we should be cautious about hoisting out of $LN$.
+Imagine $LN$ is enclosed in a time stepping loop.
+One could think of exposing and then hoisting time-invariant sub-expressions.
+There are several kinds of time-invariant sub-expressions: for example, those depending on the geometry (i.e., depending on $L_e$) and those which do not (e.g., when only involving reference element terms).
+In the former case, code motion increases the working set by a factor $L$ (unless adopting complex engineering solutions, like blocking non-affine loop nests, which are practically difficult to devise and maintain).
+The drop in number of operations could then be overwhelmed, from the runtime viewpoint, by a much larger memory pressure.
 
-A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
+A second, more general thought is that, for certain forms and discretizations, excessive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-We will provide precise details about this in the following sections.
-For the moment, and just as an example, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.
+For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $LN$, requires temporary arrays of size $N\cdot O$; with certain discretizations, as detailed in Section~\ref{sec:perf-results-forms}, this can break the local memory threshold.
 
-Based upon these considerations, we add two constraints to $C$ (see Definition~\ref{def:optimality})  
+Based upon these considerations, we add two elements to the set of memory constraints $C$ (see Definition~\ref{def:optimality})  
 \begin{enumerate}
-\item The size of a temporary due to code motion cannot be bigger than the size of the multilinear loop nest iteration space. 
-\item The total size of the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
+\item The size of a temporary due to code motion cannot be larger than that of the multilinear iteration space. 
+\item The total amount of memory occupied by the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
 \end{enumerate}
-A corollary of $C_1$ is that hoisting expressions involving geometry terms outside of $L_e$ is now forbidden. Consequently, the search space for optimality becomes smaller.
+A corollary of $C_1$ is that hoisting expressions involving geometry-dependent terms outside of $LN$, which we discussed at the beginning of this section, is now forbidden.
+Consequently, the search space for optimality becomes smaller.
 
 \subsection{Minimizing the Operation Cost}
-Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the operation cost in the innermost loops is minimum.
+Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the innermost loop operation costs is minimum.
 We now discuss how we can systematically achieve this.
@@
 Eliminating sharing from the inner multilinear loops does not suffice.
-In fact, as suggested in Section~\ref{def:optimality}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
+In fact, as suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
 
@@
 Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}).
-We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
+We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
 Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries.
@@
 \begin{figure}[h]
-\begin{lstlisting}[basicstyle=\small\ttfamily, frame=single]
+\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
 dissect the input expression into monomials
@@
 
-The point of departure consists of understanding the impact, as number of operations saved or introduced, of pre-evaluation.
-This is studied ``locally''; that is, for each monomial, in isolation.
+We study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, in isolation.
 If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed.
@@
 \begin{enumerate}
-\item the cost function employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
-\item pre-evaluating distinct monomials does not produce identical tables
+\item the cost function $\theta$ employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
+\item pre-evaluating different monomials does not result in identical tables
 \item monomials in $P$ do not share terms
@@
 We first comment on the assumptions.
-1) The creation of an optimal cost function is discussed in Section~\ref{sec:op_count}.
+1) We postpone the discussion of how to build $\theta$ to Section~\ref{sec:op_count}.
 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions.
-This is however a pathological case.
+This is however a pathological case that we can ignore without compromises.
 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
-For simplicity, we do not handle this case, which otherwise would require a ``global'' analysis of the monomials in the form.
-Further, even though, in practice, assumptions 2) and 3) did not hold, the resulting loop nest would still be nearly-optimal.
+We abstract from this case, which otherwise would require a ``global'' analysis of the monomials in the form that we believe would not justify the gain (if any) in operation count.
 
@@
 B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1).
-Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities.
+Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities.
 The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$.
-This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count.
+This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8).
+The rationale is that since we would have to pay anyway the presence of $z$ terms in the innermost loop, adding $p$ to $Z$ does not augment the operation count in our optimality model, so we can just avoid pre-evaluation.
+
+% assunzione 3 è quella che previene la somma di due mass matrix moltiplicate da F0 una e da F1 l'altra...ma non solo !
+chesso magari la trial function è derivata in un monomial e nell'altra no, ma la test function è sempre uguale...
 \end{proof}
@@
 We can easily predict $\theta_{wo}$ thanks to our key property, linearity.
-As explained in Proposition~\ref{prop:optimal-mln}, a simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $MLN_{ops}$.
-Then, assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = MLN_{ops} \cdot I$.
+This was explained in Proposition~\ref{prop:optimal-mln}.
+A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $LN_{ops}^{sf}$.
+Assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.
 

### 2015-08-11 TOMS_Paper/assembly.tex
Several fixes to the seecond half of the paper

@@
 \subsection{Qualitative Comparison}
+\label{sec:qualitative}
 Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically.
@@
 It remains to tie one loose hand: the construction of the pre-evaluation cost function.
-We introduce the cost function $\theta : M \to \mathbb{N} \times \mathbb{N}$ that, given a monomial, returns two natural numbers representing the optimal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation.
+We introduce the cost function $\theta$ such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the optimal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned.
 Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
@@
 A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $LN_{ops}^{sf}$.
-Assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.
+Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.
 
 For $\theta_{w}$, things are more complicated.
-We first need to account for the presence of (derivatives of) coefficients to estimate the \textit{increase factor} $\iota$.
+We first need to estimate the \textit{increase factor}, $\iota$, to account for the presence of (derivatives of) coefficients.
 This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation.
@@
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities.
-By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient).
-Intuitively, $\iota$ captures such a growth in non-hoistable operation.
+By expanding the product we observe an increase in the number of $[L_j, L_k]$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient).
+Intuitively, $\iota$ captures this growth in non-hoistable operation.
 
-With just a single coefficient, $\iota$ directly descends from the cost of expansion.
-In general, however, the calculation is less straightforward.
-Consider, for example, the case in which multiple coefficients originate from the same function space.
-Expansion would then lead to identical terms (i.e., identical pre-evaluated tables).
-Therefore, for a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are considering sets, so duplicates would be counted as 1).
+With a single coefficient, as we just saw, $\iota$ directly descends from the cost of expansion.
+In general, however, predicting $\iota$ is less straightforward.
+For example, consider the case in which a monomial has multiple coefficients expressed over the same function space.
+The expansion would now lead to identical sub-expressions that, once pre-evaluated, would be mapped to the same temporary.
+In other words, the resulting loop nest would be characterized by sharing.
+Therefore, the actual operation count (i.e., once sharing is removed) would be smaller than that one could infer from analysing the expansion ``in isolation''.
+For a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
 
@@
 In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
-In general, $\rho$ varies with the discretization and the differential operators employed.
+In general, $\rho$ varies with the discretization and the differential operators used in the form.
 For example, in the case of the bi-dimensional Poisson equation, after a suitable factorization, we have $\rho = 3$.
 There are several ways of determining $\rho$.
-The fastest would be to extract it from high-level analysis of the operators in the form; for convenience, in our implementation we instead project (by analysis of the expression tree) the output of expansion and factorization in the monomial.
+The fastest would be to extract it from high-level analysis of form and discretization; for convenience, in our implementation we have algorithms that, based on analysis of the expression tree, project the output of monomial expansion and factorization, which in turn gives us $\rho$.
 
@@
 \subsection{Automation through the COFFEE Language}
-As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric ``building-block'' operations.
+As opposed to what happens in the FEniCS Form Compiler with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general.
+All optimizations are instead expressed as composition of parametric ``building-block'' operations.
 This has several advantages.
-Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could ideally be expressed using the existing operators, or with small effort building on what is already available.
+Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
 Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler.
@@
 A non-exhaustive list of such operations includes expansion, factorization, re-association, generalized code motion.
-These are parametric operations.
-For example, one could ask to factorize only constant or only linear terms, while hoisting can be limited by imposing a constraint on the size of the temporaries.
-Each of these operations is implemented by suitable manipulation of the abstract syntax tree representing the integration routine.
+These ``rewrite operators'' can be seen as the COFFEE language.
+They define parametric transformations: for example, one could ask to factorize constant rather linear terms, while hoisting could be driven by loop dependency.
+Their implementation is based on manipulation of the abstract syntax tree representing the integration routine.
 
@@
 They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point.
-For some complex monomials (e.g., see~\citeN{quadrature1} and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, outweigh that in the multilinear loop nest.
+For some complex monomials and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, even outweigh that of the multilinear loop nest.
 In these cases, our definition of optimality becomes weaker, since the underlying assumption is that the bulk of the computation is carried out in innermost loops.
 
-Despite the fact that we are not characterizing optimality for this wider class of loops, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis}, particularly for removing sharing, to minimize the operational cost.
+Despite the fact that we are not characterizing optimality for this wider class of problems, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} to remove sharing, thus reducing the operation count.
 This is straightforward in our code generation system by composing rewrite operators.
@@
 COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions.
-This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to trivial code without performing any sort of transformation.
+This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to code without performing any sort of analysis.
 Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$.
 We then synthesize an optimal loop nest as described in the previous sections.
-This may in turn expose a set of $L_i$-dependent statements evaluating some temporaries.
-For each of this statements, we try to remove sharing by greedily applying factorization and code motion.
-In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named ``factorize'', ``expand'', etc.).
+This may in turn expose a set of $L_i$-dependent expressions.
+For each of this expressions, we try to remove sharing by greedily applying factorization and code motion.
+In the COFFEE language, this process is expressed by simply composing five rewrite operators.
 
@@
 \label{sec:opt-review}
-We start with briefly reviewing the low-level optimizations presented in~\citeN{Luporini}.
+We start with briefly reviewing the low level optimizations presented in~\citeN{Luporini}.
 
@@
 \paragraph{Vector-register Tiling} Blocking (or tiling) at the level of vector registers improves data locality beyond traditional unroll-and-jam transformations.
-The blocking strategy consists of evaluating outer products by just using two vector register without ever spilling to cache.
+This blocking strategy consists of evaluating outer products by using just two vector register and without ever spilling to cache.
 
-\paragraph{Expression Splitting} When the number of basis functions arrays (or, equivalently, temporaries introduced by code motion) and constants is large, the chances of spilling to cache are high in architectures with a relatively low number of logical registers (e.g.
+\paragraph{Expression Splitting} When the number of basis functions arrays (or, equivalently, temporaries introduced by code motion) and constants is large, the chances of spilling to cache are high in architectures with a few logical registers (e.g.
 16/32).
-By exploiting sum’s associativity, an expression can be fissioned so that the resulting sub-expressions can be computed in a separate loop nest.
+By exploiting sum’s associativity, an expression can be fissioned so that the resulting sub-expressions can be computed in separate loop nests.
+This reduces the register pressure.
 
@@
 Integration-dependent expressions are inherently executed as scalar code because vectorization (unless employing special hand-written schemes) occurs along a single loop, typically the innermost.
-In our loop nest, $L_i$ is clearly an outer loop.
-In situations such as in Section~\ref{sec:heuristic-int-dep}, we may want to vectorize also along $L_i$.
+For the same reasons discussed in Section~\ref{sec:heuristic-int-dep}, we also want to vectorize along $L_i$.
 One way to achieve this is vector-promotion.
-This implies creating a clone loop of $L_i$ in the preheader of the main loop nest, in which vector temporaries, instead of scalar ones, are evaluated.
+This requires creating a ``clone'' of $L_i$ in the preheader of the loop nest, in which vector temporaries are evaluated in what is now an innermost loop.
 
-\subsubsection{Handling Block-sparse Tables}
+\subsubsection{Handling Sparse Tables}
 \label{sec:zeros}
-Consider a tabulated set of basis functions having quadrature points along the rows and functions along the columns.
+Consider a set of tabulated basis functions with quadrature points along rows and functions along columns.
 For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}.
-Unless using a smart form compiler (and one of our goals is precisely to relieve it from the burden of smart code generation), there are several circumstances in which the tables can be block-sparse.
+Unless using a smart form compiler (which we want to avoid), there are circumstances in which the tables are sparse.
 Zero-valued columns arise when taking derivatives on a reference element or when employing vector-valued elements.
@@
 Zero-valued blocks can appear in pre-evaluated temporaries.
-Our objective is a general system that avoids useless iteration over these zero-entries while preserving the effectiveness of the other low-level optimizations, especially vectorization.
+Our objective is a transformation that avoids useless iteration over zeros while preserving the effectiveness of the other low-level optimizations, especially vectorization.
 
@@
 \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed.
-Our approach, which will be compared to this pioneering work, aims to free the generated code from any indirection arrays.
+Our approach, which will be compared to this pioneering work, aims to free the generated code from such indirection arrays.
 This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.
 
-The idea is that if the dimension along which vectorization is performed (typically the innermost) has a contiguous slice (rows, columns) of zero-valued entries, and if that slice is smaller than the vector length, that we do nothing (i.e., the loop nest is not transformed).
-Otherwise, we restructure the loop nest.
-This has several complicated implications, including loop fission and a memory offsetting usually dangerous for padding and data alignment, which are, however, beyond the scope of this paper.
-The implementation is based on symbolic execution: the loop nests are traversed and for each statement the location of zeros in each of the involved symbols is tracked.
-Arithmetic operators have a different impact on the tracking.
-For example, for each loop dimension, multiplication requires computing the set intersection of the zero-valued slices, whereas addition requires computing the set union.
+The idea is that if the dimension along which vectorization is performed (typically the innermost) has a contiguous slice of zeros, but that slice is smaller than the vector length, then we do nothing (i.e., the loop nest is not transformed).
+Otherwise, we restructure the iteration space.
+This has several non-trivial implications.
+The most notable one is memory offsetting (e.g., \texttt{A[i+m,j+n]}), which dramatically enters in conflict with padding and data alignment.
+We use heuristics to retain the positive effects of both and to ensure correctness.
+Details are, however, beyond the scope of this paper.
+
+The implementation is based on symbolic execution: the loop nests are traversed and for each statement encountered the location of zeros in each of the involved symbols is tracked.
+Arithmetic operators have a different impact on tracking.
+For example, multiplication requires computing the set intersection of the zero-valued slices (for each loop dimension), whereas addition requires computing the set union.
 
@@
 \begin{itemize}
-\item FEniCS Form Compiler's built-in optimization system: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
-\item FEniCS Form Compiler's built-in optimization system: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
-\item FEniCS Form Compiler's built-in optimization system: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, discussed in~\citeN{Fenics}).
+\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
+\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
+\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}).
 Referred to as \texttt{auto}
@@
 
-For a fair comparison, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants.
+For the fairest comparison possible, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants.
 UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS, which employs a data storage layout different than Firedrake.
@@
 
-Following the methodology adopted in ~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
+Following the methodology adopted in~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
 \begin{itemize}
@@
 \end{itemize}
-Constants of our analysis are instead
+Constants of our analysis, instead, are
 \begin{itemize}
-\item the space of test, trial, and coefficient functions, Lagrange
-\item the mesh, tetrahedral with a total of 4374 elements
+\item the space of test, trial, and coefficient functions: Lagrange
+\item the mesh: tetrahedral with a total of 4374 elements
 \item exact numerical quadrature (the same scheme as in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule, is employed)
@@
 
-The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are ``nested'' plots.
+The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) can be interpreted as ``nested'' plots.
 We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''.
 In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis.
-The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
+The top-left plot in a grid shows the speed up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
 The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$.
@@
 There are two reasons for this.
-The former is that low-level optimization can have a significant impact.
+The former is that low level optimization can have a significant impact.
 Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$.
@@
 With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}.
-The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries.
+The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once).
 In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases.

### 2015-08-12 TOMS_Paper/assembly.tex
Refine second half of paper, update elasticity plot

@@
 For what concerns the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache).
-For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ and $T_H = \infty$.
+For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ to assess our hypotheses.
 Results of these trials are discussed below.
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
+\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:mass}
 \end{figure}
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
-\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
+\caption{Performance evaluation for the bilinear form of a Helmholtz equation.
+The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
 \end{figure}
@@
 \includegraphics[scale=0.7]{perf-results/elasticity}
-\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
+\caption{Performance evaluation for the bilinear form arising in a elastic model.
+The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
 \end{figure}
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
-\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
+\caption{Performance evaluation for the bilinear form arising in a hyperelastic model.
+The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
 \end{figure}
@@
 \paragraph{Mass}
-From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} represents the best variant among the ones tested, apart from the few points previously discussed.
+From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} is the best variant among the ones tested, apart from a few points already discussed.
 This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints.
 It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$.
+The fact that \texttt{cfO2} is slower than \texttt{tens} at $[q=4, p=4, nf=1]$ is explained by the memory threshold that prevents pre-evaluation.
+By just setting $T_H = L3_{size}$, \texttt{cfO2} becomes 1.48$\times$ faster than \texttt{tens}.
+This is probably due to the hardware prefetching being way more effective than in all other forms: the number of allocated temporaries is relatively small, and so is the number of memory access traces that need be tracked.
+However, as we already explained, in a parallel context this gain might diminish, since the L3 cache is shared by the cores of the CPU.
 
@@
 Take, for example, the case $[q=3, p=1, nf=2]$.
-The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over the competitors.
+The adoption of pre-evaluation makes a tremendous speed up be achievable by \texttt{cfO2} over the competitors.
 On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.
+The effects of the working set threshold are in general positive.
+Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow down of 2.16$\times$ at $[q=2, p=2]$.
 
@@
 The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
-This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present.
-With $q=4$ and $nf \in [0, 1]$, pre-evaluation in \texttt{cfO2} is found profitable in terms of operation count, although it is not triggered because of exceeding the working set threshold $T_H$.
-Running the same experiments with $T_H = \infty$ resulted in significant runtime increases for all values of $p$.
-Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement for $nf = 0$, with \texttt{cfO2} becoming the fastest alternative.
-Our explanation for this is that by running a sequential program, a process has exclusive access to the memory bandwidth of the system, so pre-fetching minimizes cache misses without hurting concurrent processes (since there are none).
+This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$.
+Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$.
+In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to completely removing sharing and avoiding iteration over zero-valued blocks.
+The latter case occurs with $q=4$ and $nf \in [0, 1]$.
+Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow downs characterized the cases of $nf=1$ for all values of $p$.
+Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$.
+This suggest that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.
 
@@
 A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.
+Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.
 

### 2015-08-13 TOMS_Paper/assembly.tex
Add vectorization perf eval discussion + minor fixes

@@
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
-\end{CenteredBox}\caption{Optimized Loop nest example}\label{code:loopnest-opt}\end{figure}
+\end{CenteredBox}\caption{A possible loop nest for the local assembly of a form after applying pre-evaluation to some of the monomials (resulting in sub-expressions \texttt{G'}, \texttt{G''}, ...)}\label{code:loopnest-opt}\end{figure}
 
@@
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
-\end{CenteredBox}\caption{Optimized Loop nest example}\label{code:increase_factor}\end{figure}
+\end{CenteredBox}\caption{Simplified loop nest for the local assembly of a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}
 
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:mass}
+\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
 \end{figure}
@@
 \caption{Performance evaluation for the bilinear form of a Helmholtz equation.
-The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
+The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
 \end{figure}
@@
 \caption{Performance evaluation for the bilinear form arising in a elastic model.
-The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
+The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
 \end{figure}
@@
 \caption{Performance evaluation for the bilinear form arising in a hyperelastic model.
-The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
+The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
 \end{figure}
@@
 The first observation is that our optimality model does not imply minimum execution time.
-In particular, in the 0.03$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime.
+In particular, in the 0.06$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime.
 There are two reasons for this.
@@
 
-A second observation concerns the ``triggering'' of pre-evaluation in \texttt{cfO2}.
+The second aspect regards the ``triggering'' of pre-evaluation in \texttt{cfO2}.
 The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable.
@@
 
+The third point concerns the effect of the chosen discretizations on autovectorization.
+The discretizations employed ensure both multilinear loop nest and tabulated basis function sizes to be multiple of the machine vector length.
+Given the suitable choice of compilation options, this promotes autovectorization in the majority of code variants.
+The biggest exception is \texttt{quad}, in which, due to the presence of indirection arrays in the generated code, vectorization is rarely triggered.
+In \texttt{tens}, the loop nests are fully unrolled, so the standard loop vectorization is simply not possible; however, manual inspection of the compiled code suggests block vectorization (\cite{SLP-vect}) is often triggered.
+In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces tend to have the same ``structure'' (there are very few exceptions due to low level optimization), with loop vectorization being regularly applied, as far as we could evince.
+It is worth noting the fact that the size of loops and tabulated basis functions is a multiple of the vector length render padding and data alignment (in \texttt{cfO1} and \texttt{cfO2}), to the best of our understanding, basically irrelevant.
+
 Also note some other minor things.
 \begin{itemize}
-\item The lack of improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
-\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
+\item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
+\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
 \item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
@@
 Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$.
-This suggest that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.
+This suggests that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.
 

### 2015-09-03 TOMS_Paper/assembly.tex
Incorporate Paul's advises and rework optimality

@@
 % Page heads
-\markboth{F. Luporini et al.}{Optimal Finite Element Integration}
+\markboth{F. Luporini et al.}{On Optimality of Finite Element Integration}
 
 % Title portion
-\title{Optimal Finite Element Integration}
+\title{On Optimality of Finite Element Integration}
 \author{Fabio Luporini
@@
 \begin{abstract}
-We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms.
-Optimality is defined in terms of floating point operations required to compute a result.
-We establish a model under which optimality can systematically be reached and comment on the general applicability of this model.
-The model is integrated in a compiler that, given a representation of an integration routine, derives a loop nest in optimal form.
-Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems.
+We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms.
+Optimality is defined in terms of floating point operations required to execute a loop nest.
+The generation of optimal loop nests is driven by a model that exploits mathematical properties of the domain of interest.
+A theoretical analysis and extensive experimentation prove the effectiveness of our approach, showing systematic performance improvements over a number of alternative code generation systems.
 The effect of low-level optimization is also discussed.
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. Optimal Finite Element Integration.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. On Optimality of Finite Element Integration.}
 
@@
 By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed.
-The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.
+The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.
 
-As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges.
+As demonstrated by the considerable body of research on the topic, automating the generation of such high performance implementations poses several challenges.
 This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
-General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary).
-Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on loop optimization for cache locality, so they are not particularly helpful in our context.
+General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary).
+Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and loop optimization for cache locality, so they are not particularly helpful in our context.
 The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems.
@@
 However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-With this paper, we aim to make a first step towards solving this problem.
+In this paper, we formulate an approach to solve this problem.
+Summarizing, our contributions are as follows
 
-We summarize the contributions of this research as follows
 \begin{itemize}
@@
 As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
-\item We provide an incremental, constructive proof that optimal implementations can be obtain by analyzing the monomials appearing in a form
-\item We automate the translation of forms into optimal loop nests by a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
-The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, the default optimization system in the Firedrake framework.
-\item We provide an unprecedented (for its extension) performance evaluation comparing several code transformation systems in a range of forms of increasing complexity.
+\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into flop-optimal loop nests.
+\item We comment on the cases in which such model does not lead to optimal loop nests.
+We will see that this might occur with particular forms that make extensive use of tensor algebra.
+We show, however, that our model still is capable of producing quasi-optimal loop nests.
+\item We integrate the model with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
+The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
+\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research.
 This is essential to demonstrate that our optimality model holds in practice.
-\item We comment on the cases in which our characterization of optimality becomes weaker, and provide an intuition of how we handle such forms
-\item We enrich the set of low-level transformations described in~\citeN{Luporini}. These make COFFEE a full-fledged system capable of decoupling form compilers from low level code generation.
+\end{itemize}
+
+In addition, in order to place COFFEE on the same level of other code generation systems from the viewpoint of low-level optimization (which is essential for a fair performance comparison)
+
+\begin{itemize}
+\item We introduce an engine based on symbolic execution that allows skipping irrelevant floating point operations (e.g., those involving zero-valued quantities).
+We elaborate on the performance impact of this optimization, making a clear distinction between flop optimality and efficient code.
 \end{itemize}
@@
 The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$.
-From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
+From the solution of the following linear system it is possible to determine a set of coefficients to express $u$:
 \begin{equation}
@@
 For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced.
-This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated.
+This implies that a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated.
 The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
@@
 \label{sec:qualitative}
-Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically.
-The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for ``complex'' forms.
-On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''.
+Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary quite dramatically.
+The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms.
+On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''.
 The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.
@@
 \begin{Def}[Perfect and imperfect loop nests]
-A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said imperfect. 
+A perfect loop nest is a loop whose body either 1) comprises only a sequence of non-loop statements or 2) is itself a perfect loop nest.
+If this condition does not hold, a loop nest is said imperfect.
 \end{Def}
@@
 \begin{Def}[Independent basic block]
-A sequence of statements that present no data dependencies compose an independent basic block
+An independent basic block is a sequence of statements that present no data dependencies
 \end{Def}
@@
 % In the following, we should represent loops by their iteration variable.
-It is convenient to map mathematical properties to the loop nest level
+It is convenient to map mathematical properties to the loop nest level.
 
@@
 
-In particular, we are interested in the following class since it naturally arises from the math of finite element integration, reviewed in Section~\ref{sec:background}.
+In particular, we are interested in the following class since it naturally arises in finite element integration (as reviewed in Section~\ref{sec:background}).
 
@@
 
-\begin{Def}[Optimality of a multilinear loop nest]
-The synthesis of a multilinear loop nest is \textit{optimal} if the amount of operations performed in the innermost loop is minimum.
+\begin{Def}[Relaxed optimality of a multilinear loop nest]
+The synthesis of a multilinear loop nest is optimal if the number of operations performed in the innermost loop is minimum.
 \end{Def}
@@
 In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop.
-Note that this definition does not take into account memory requirements.
-If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense.
+We use the attribute ``relaxed'' to denote that the definition does not take into account outer loops.
+In the following, we assume ``relaxed'' is implicit whenever we discuss optimality or mention ``optimal form''.
+Also note that this definition does not take into account memory requirements.
+If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
 In the following, we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated.
@@
 
-\begin{Def}[Optimality of a loop nest]
+\begin{Def}[Relaxed optimality of a loop nest]
 \label{def:optimality}
-The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
+The synthesis of a loop nest is optimal if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
 \end{Def}
@@
 In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).
+Again, we defined``relaxed optimality'' to highlight the fact that we are focussing on inner loops only.
 
@@
 We make the following observations.
-1) $L >>> M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
+1) $L \gg M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
 
@@
 \label{sec:mem-const}
-The fact that $L >>> M, N, O$ suggests we should be cautious about hoisting out of $LN$.
+The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting out of $LN$.
 Imagine $LN$ is enclosed in a time stepping loop.
@@
 The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake.
-In this section, we describe the structure of the code generation system and we also comment on a set of low-level optimizations.
+In this section, we describe the structure of the code generation system.
 
@@
 
-We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
+We evaluate the speed-ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
 \begin{itemize}
@@
 In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis.
-The top-left plot in a grid shows the speed up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
+The top-left plot in a grid shows the speed-up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
 The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$.
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
+\caption{Performance evaluation for the mass matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
 \end{figure}
@@
 \caption{Performance evaluation for the bilinear form of a Helmholtz equation.
-The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
+The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
 \end{figure}
@@
 \caption{Performance evaluation for the bilinear form arising in a elastic model.
-The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
+The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
 \end{figure}
@@
 \caption{Performance evaluation for the bilinear form arising in a hyperelastic model.
-The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
+The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
 \end{figure}
@@
 The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once).
-In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases.
+In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed-ups throughout the majority of test cases.
 The latter reason is simply inherent to our optimality model.
@@
 \item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
-\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
+\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
 \item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
@@
 Take, for example, the case $[q=3, p=1, nf=2]$.
-The adoption of pre-evaluation makes a tremendous speed up be achievable by \texttt{cfO2} over the competitors.
+The adoption of pre-evaluation makes a tremendous speed-up be achievable by \texttt{cfO2} over the competitors.
 On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.
 The effects of the working set threshold are in general positive.
-Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow down of 2.16$\times$ at $[q=2, p=2]$.
+Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow-down of 2.16$\times$ at $[q=2, p=2]$.
 
@@
 Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$.
-In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to completely removing sharing and avoiding iteration over zero-valued blocks.
+In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to the elimination of sharing and the avoidance of iteration over zero-valued blocks.
 The latter case occurs with $q=4$ and $nf \in [0, 1]$.
-Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow downs characterized the cases of $nf=1$ for all values of $p$.
-Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$.
-This suggests that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.
+Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow-downs characterized the cases of $nf=1$ for all values of $p$.
+Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only 40$\%$, while the cost function predicts a save in operation count of 5$\times$.
+This suggests that our model could be refined to handle the cases in which the gain in operation count is so large that non-extreme variations in working set size should be considered negligible.
 
@@
 \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions.
-\texttt{ufls} performs generally well and exhibits significant speed ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.
+\texttt{ufls} performs generally well and exhibits significant speed-ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.
 

### 2015-09-04 TOMS_Paper/assembly.tex
Several updates

@@
 \newtheorem{Prop}{Proposition}
+\newtheorem*{Rem*}{Remark}
 
@@
 
-% In the following, we should represent loops by their iteration variable.
-It is convenient to map mathematical properties to the loop nest level.
+By mapping mathematical properties to the loop nest level, we propose the concept of \textit{linear loop}.
 
@@
 
-In particular, we are interested in the following class since it naturally arises in finite element integration (as reviewed in Section~\ref{sec:background}).
+In particular, we are interested in the following class since it naturally arises in finite element integration.
 
@@
 
-\begin{Def}[Relaxed optimality of a multilinear loop nest]
+\begin{Def}[Optimality of a multilinear loop nest]
+\label{def:mln-optimality}
 The synthesis of a multilinear loop nest is optimal if the number of operations performed in the innermost loop is minimum.
@@
 In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop.
-We use the attribute ``relaxed'' to denote that the definition does not take into account outer loops.
-In the following, we assume ``relaxed'' is implicit whenever we discuss optimality or mention ``optimal form''.
-Also note that this definition does not take into account memory requirements.
+We will see that the restriction to inner loops has the effect of simplifying the transformation model; the price to pay will be the generation of quasi-optimal code in presence of costly outer loops (this, in finite element integration, happens when forms are characterized by extensive use of tensor algebra).
+
+Note that Definition~\ref{def:mln-optimality} does not take into account memory requirements.
 If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
@@
 
-A second result follows.
+\begin{Rem*}
+In the following, when discussing ``optimality'' or ``loop nests in optimal form'', we will always tacitly refer to inner optimality (unless explicitly specified differently). 
+\end{Rem*}
+
+As anticipated, perfect multilinear loop nests can be brought to optimal form. The following result proves this claim.
+
 \begin{Prop}
@@
 For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.
-
-We now consider a wider class of loop nests in which perfectness and multilinearity apply to a sub-nest only. Consider the example in Figure~\ref{code:loopnest}.
-
+\\
 \begin{figure}[h]\begin{CenteredBox}
@@
 
+We now consider a wider class of loop nests in which perfectness and multilinearity apply to a sub-nest only.
+Consider the example in Figure~\ref{code:loopnest}.
 The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
@@
 {\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
-~~~~~~~~~~~~~~
+~~~~~~~~~~
 {\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
@@
 \begin{itemize}
-\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example of Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
-\item the transformations exposing sub-expressions amenable to pre-evaluation could increase the arithmetic complexity within $L_k$ (for example, expansion may increase the operation count).
-This could overwhelm any gain due to pre-evaluation.
+\item the transformations exposing sub-expressions amenable to pre-evaluation could increase the arithmetic complexity within $L_k$ (e.g., expansion may increase the operation count; further examples will be provided later).
+This could overwhelm the gain inherent in pre-evaluation.
 \end{itemize}
 
-We shall then refine our definition of optimality for this wider class of loop nests as follows
+Before attacking this problem in Section~\ref{sec:optimal-synthesis}, we shall conclude our construction refining Definition~\ref{def:mln-optimality} for generic loop nests
 
-\begin{Def}[Relaxed optimality of a loop nest]
+\begin{Def}[Inner optimality of a loop nest]
 \label{def:optimality}
@@
 In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).
-Again, we defined``relaxed optimality'' to highlight the fact that we are focussing on inner loops only.
 
-\section{Synthesis of Optimal Loop Nests in Finite Element Integration}
+\section{Optimal-form Loop Nests in Finite Element Integration}
 \label{sec:optimal-synthesis}
@@
 One could think of exposing and then hoisting time-invariant sub-expressions.
-There are several kinds of time-invariant sub-expressions: for example, those depending on the geometry (i.e., depending on $L_e$) and those which do not (e.g., when only involving reference element terms).
-In the former case, code motion increases the working set by a factor $L$ (unless adopting complex engineering solutions, like blocking non-affine loop nests, which are practically difficult to devise and maintain).
-The drop in number of operations could then be overwhelmed, from the runtime viewpoint, by a much larger memory pressure.
+For those sub-expressions that depend on the mesh geometry (i.e., they would depend on $L_e$), code motion would increase the working set by a factor $L$.
+The gain in number of operations executed could then be overwhelmed, from a runtime viewpoint, by a much larger memory pressure.
 
-A second, more general thought is that, for certain forms and discretizations, excessive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
+A second, more general fact is that, for certain forms and discretizations, excessive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $LN$, requires temporary arrays of size $N\cdot O$; with certain discretizations, as detailed in Section~\ref{sec:perf-results-forms}, this can break the local memory threshold.
+For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $LN$, requires temporary arrays of size $N\cdot O$.
+This can sometimes break the local memory threshold.
+We will clarify and experiment this in Section~\ref{sec:perf-results-forms}.
 
-Based upon these considerations, we add two elements to the set of memory constraints $C$ (see Definition~\ref{def:optimality})  
+Based on these considerations, we characterize the set of memory constraints $C$ (see Definition~\ref{def:optimality}) as follows
 \begin{enumerate}
@@
 
-\subsection{Minimizing the Operation Cost}
-Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the innermost loop operation costs is minimum.
+\subsection{Minimizing the Operation Count in Inner Loops}
+Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the innermost loop operation counts is minimum.
 We now discuss how we can systematically achieve this.
@@
 
-\begin{figure}[h]\begin{CenteredBox}
+\begin{figure}\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
@@
 
-Before elaborating on the profitability of pre-evaluation, we need to discuss under which conditions this approach, based on a ``local analysis'' of monomials, is optimal.
+Before elaborating on the profitability of pre-evaluation (i.e., how to determine the cost function $\theta$ in Figure~\ref{code:intuition}), we need to discuss under which conditions this approach, based on a ``local analysis'' of monomials, is able to reduce loop nests to optimal form.
 
@@
 \label{prop:optimal-approach}
-Consider an expression comprising a set of monomials $M$.
-Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and be $Z = M \setminus P$.
+Consider an integration routine originating from an arbitrary multilinear form.
+A loop nest $LN$, as previously characterized, evaluates an expression comprising a set of monomials, $M$.
+Let $P$ be the set of monomials for which pre-evaluation is applied because found profitable, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
 Assume that:
@@
 \end{enumerate} 
-Then, the loop nest $LN$ is optimal under memory constraints $C$, once sharing is removed.
+Then, the main algorithm in Figure~\ref{code:intuition} leads to inner optimality under the set of memory constraints $C$ (defined in Section~\ref{sec:mem-const}).
 \end{Prop}
@@
 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions.
-This is however a pathological case that we can ignore without compromises.
+This is however a pathological case that we can harmlessly ignore.
 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
@@
 Since they only differ for the presence of $L_i$, we relieve the notation by omitting the shared loops when discussing operation counts.
-The operation count of what we are proving to be the optimal $LN$ synthesis is $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation cost of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.
-This can also be inferred from Figure~\ref{code:loopnest-opt}.
+The operation count of what we are proving to be the optimal $LN$ synthesis (inferable from Figure~\ref{code:loopnest-opt}) is $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation count of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.
 
@@
 B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1).
-Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities.
+Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities.
 The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$.
 This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8).
-The rationale is that since we would have to pay anyway the presence of $z$ terms in the innermost loop, adding $p$ to $Z$ does not augment the operation count in our optimality model, so we can just avoid pre-evaluation.
+The rationale is that since we would have to pay anyway the presence of $z$ in the innermost loop, adding $p$ to $Z$ would not augment the operation count in our model, so we can safely avoid pre-evaluation.
 
@@
 \label{sec:op_count}
-It remains to tie one loose hand: the construction of the pre-evaluation cost function.
-We introduce the cost function $\theta$ such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the optimal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned.
+It remains to tie one loose hand: the construction of the pre-evaluation cost function $\theta$.
+We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the sharing-free operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned.
 Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
@@
 The expansion would now lead to identical sub-expressions that, once pre-evaluated, would be mapped to the same temporary.
-In other words, the resulting loop nest would be characterized by sharing.
+The resulting loop nest would be characterized by sharing, as a result.
 Therefore, the actual operation count (i.e., once sharing is removed) would be smaller than that one could infer from analysing the expansion ``in isolation''.
-For a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
+For a precise estimate of $\iota$, we then need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
 
@@
 If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$.
-In particular, we need to calculate the number terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$.
+In particular, we need to calculate the number of terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$.
 Consider again Figure~\ref{code:increase_factor}.
@@
 In general, $\rho$ varies with the discretization and the differential operators used in the form.
-For example, in the case of the bi-dimensional Poisson equation, after a suitable factorization, we have $\rho = 3$.
+For example, in the case of the bilinear form originating from a standard bi-dimensional Poisson equation, the reader could verify that after a suitable factorization we would have $\rho = 3$.
 There are several ways of determining $\rho$.
@@
 \label{sec:codegen}
-The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake.
-In this section, we describe the structure of the code generation system.
+The model described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimizer of finite element integration routines used in Firedrake.
+In this section, we describe the features of this code generation system.
 
@@
 For some complex monomials and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, even outweigh that of the multilinear loop nest.
-In these cases, our definition of optimality becomes weaker, since the underlying assumption is that the bulk of the computation is carried out in innermost loops.
+In these cases, our optimality model becomes weaker, since its underlying assumption is that the bulk of the computation is carried out in innermost loops.
 
-Despite the fact that we are not characterizing optimality for this wider class of problems, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} to remove sharing, thus reducing the operation count.
+Despite the fact that we are not characterizing optimality for this wider class of problems, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} to try to eliminate sharing, thus reducing the operation count.
 This is straightforward in our code generation system by composing rewrite operators.
@@
 This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to code without performing any sort of analysis.
-Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$.
+Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent in $L_i$.
 We then synthesize an optimal loop nest as described in the previous sections.
@@
 The results are extremely encouraging, suggesting our model applies to a variety of contexts.
-We have discussed the conditions under which the optimality model becomes weaker.
+We have discussed the conditions under which the model only leads to quasi-optimal loop nests.
 An open problem is understanding how to refine this model to include outer loops.
-This will probably require exploiting properties of differential operators.
+This will probably require exploiting mathematical properties of differential operators.
 A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.

### 2015-09-17 TOMS_Paper/assembly.tex
Some cleanup of the definitions

@@
 \begin{Def}[Perfect and imperfect loop nests]
-A perfect loop nest is a loop whose body either 1) comprises only a sequence of non-loop statements or 2) is itself a perfect loop nest.
-If this condition does not hold, a loop nest is said imperfect.
+A perfect loop nest is a loop whose body either 1) comprises only a sequence
+of non-loop statements or 2) is itself a perfect loop nest. If this
+condition does not hold, a loop nest is said to be imperfect. 
 \end{Def}
@@
 \begin{Def}[Independent basic block]
-An independent basic block is a sequence of statements that present no data dependencies
+An independent basic block is a sequence of statements such that no data
+dependencies exist between statements in the block.
 \end{Def}
 
-We focus on perfect nests whose innermost loop body is an independent basic block.
-A straightforward property of this class is that hoisting invariant expressions from the innermost to any of the outer loops or the preheader (i.e., the block that precedes the entry point of the nest) is always safe, as long as the obvious loop dependencies are preserved.
+We focus on perfect nests whose innermost loop body is an independent basic
+block. A straightforward property of this class is that hoisting invariant
+expressions from the innermost to any of the outer loops or the preheader
+(i.e., the block that precedes the entry point of the nest) is always safe,
+as long as any dependencies on loop indices are honored.
 We will make use of this property.
-The results of this section could also be generalized to larger classes of loop nests, in which basic block independence does not hold, although this would require a refinement that goes beyond the scope of this paper.
+The results of this section could also be generalized to larger classes of loop nests, in which basic block independence does not hold, although this would require refinements beyond the scope of this paper.
 
-By mapping mathematical properties to the loop nest level, we propose the concept of \textit{linear loop}.
+By mapping mathematical properties to the loop nest level, we introduce the
+concept of a \textit{linear loop}.
 
 \begin{Def}[Linear loop]
-A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if all expressions appearing in the body of $L$ that use $i$ to access some memory locations are linear functions over $I$.
+A loop $L$ defining the iteration space $I$ through the iteration variable
+$i$, or simply $L_i$, is \textit{linear} if 
+\begin{enumerate}
+\item $i$ appears in the body of $L$
+only as an array index, and
+\item whenever an array $a$ is indexed by $i$
+($a[i]$), all expressions in which this appears are affine in $a$.
+\end{enumerate}
 \end{Def}
@@
 \label{prop:share-removal}
-Assume $LN = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a perfect multilinear loop nest with the body of $L_{i_{n-1}}$ being an independent basic block.
-Then sharing can always be eliminated from $LN$.
+Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a perfect multilinear loop nest with the body of $L_{i_{n-1}}$ being an independent basic block.
+Then sharing can always be eliminated from $\Lambda$.
 \end{Prop}
 \begin{proof}
-The demonstration is by construction and exploits linearity.
-We want to transform $LN$ into $LN'$ such that there is no sharing in any $L_i \in LN'$.
-Starting from the innermost loop $L_{i_{n-1}}$, the expressions are ''flattened'' by expanding all products involving terms depending on $L_{i_{n-1}}$.
-Being now on the same level of the expression tree, such terms can be factorized.
-Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$, and such term is now unique in the expression.
-The other terms, independent of $L_{i_{n-1}}$, are, by definition, loop-invariant, and as such can be hoisted at the level of $L_{i_{n-2}}$.
-This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is safe.
+  The demonstration is by construction and exploits linearity. We want to
+  transform $\Lambda$ into $\Lambda'$ such that there is no sharing in any
+  $L_i \in \Lambda'$. Starting from the innermost loop $L_{i_{n-1}}$, the
+  expressions are ``flattened'' by expanding all products involving terms
+  depending on $L_{i_{n-1}}$. Being now on the same level of the expression
+  tree, such terms can be factorized. Due to linearity, each factored
+  product only has one term depending on $L_{i_{n-1}}$, and this term is now
+  unique in the expression. The other terms, independent of $L_{i_{n-1}}$,
+  are, by definition, loop-invariant, and as such can be hoisted at the
+  level of $L_{i_{n-2}}$. This procedure can be applied recursively up to
+  $L_{i_{0}}$: multilinearity allows factorization at each level;
+  perfectness ensures hoisting is safe.
 \end{proof}
@@
 \label{def:mln-optimality}
-The synthesis of a multilinear loop nest is optimal if the number of operations performed in the innermost loop is minimum.
+The synthesis of a multilinear loop nest is optimal if the number of
+operations performed in the loop nest is minimal.
 \end{Def}
 
-In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop.
-We will see that the restriction to inner loops has the effect of simplifying the transformation model; the price to pay will be the generation of quasi-optimal code in presence of costly outer loops (this, in finite element integration, happens when forms are characterized by extensive use of tensor algebra).
+%We will see that the restriction to inner loops has the effect of simplifying the transformation model; the price to pay will be the generation of quasi-optimal code in presence of costly outer loops (this, in finite element integration, happens when forms are characterized by extensive use of tensor algebra).
 
-Note that Definition~\ref{def:mln-optimality} does not take into account memory requirements.
-If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
-In the following, we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated.
-This suits the context of finite element integration.
+Note that Definition~\ref{def:mln-optimality} does not take into account
+memory requirements. If the loop nest were memory-bound -- the ratio of
+operations to bytes transferred from memory to the CPU being too low -- then
+speaking of optimality would clearly make no sense. In the following, we
+assume to operate in a CPU-bound regime, in which arithmetic-intensive
+expressions need be evaluated. In the context of finite element, this is
+often true for more complex multilinear forms and/or higher order elements.
 
-\begin{Rem*}
-In the following, when discussing ``optimality'' or ``loop nests in optimal form'', we will always tacitly refer to inner optimality (unless explicitly specified differently). 
-\end{Rem*}
+%\begin{Rem*}
+%In the following, when discussing ``optimality'' or ``loop nests in optimal form'', we will always tacitly refer to inner optimality (unless explicitly specified differently). 
+%\end{Rem*}
 
@@
 \label{prop:optimal-mln}
-A perfect multilinear loop nest $LN$ with basic block independence can always be reduced to optimal form.
+A perfect multilinear loop nest $\Lambda$ with basic block independence can always be reduced to optimal form.
 \end{Prop}
 \begin{proof}
-By construction.
-Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$.
-These sets are sorted in descending order based on their cardinality.
-We produce a new loop permutation that reflects the ${S_i}$ ordering; that is, loops are placed such that, as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$.
-Note the loop permutation is semantically correct because of perfectness.
+By construction. Loop-dependent terms are logically grouped into $n$
+disjoint sets $S_i$, each $S_i$ containing all terms depending on
+$L_i$. These sets are sorted in descending order based on their
+cardinality. We produce a new loop permutation that reflects the ${S_i}$
+ordering; that is, loops are placed such that each loop has fewer unique
+terms than the outer loops containing it. Since the loop nest is perfect,
+any such permutation is semantically correct.
 Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
 This renders $L_{i_{n-1}}$ optimal.
-In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
+In particular, the number of operations is equal to $\Lambda_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
 \end{proof}
@@
 Consider the example in Figure~\ref{code:loopnest}.
-The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
 The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication, possibly characterized by sharing.
@@
 
-It is then obvious that reaching an optimal form $LN'$ is, in the general case, a challenging issue, since a full exploration of the expression tree transformation space is necessary.
-Even though we assumed we could generate $LN'$ through a suitable transformation algorithm, the following issues should be addressed
+It is then obvious that reaching an optimal form $\Lambda'$ is, in the general case, a challenging issue, since a full exploration of the expression tree transformation space is necessary.
+Even though we assumed we could generate $\Lambda'$ through a suitable transformation algorithm, the following issues should be addressed
 \begin{itemize}
@@
 \label{sec:mem-const}
-The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting out of $LN$.
-Imagine $LN$ is enclosed in a time stepping loop.
+The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting out of $\Lambda$.
+Imagine $\Lambda$ is enclosed in a time stepping loop.
 One could think of exposing and then hoisting time-invariant sub-expressions.
@@
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $LN$, requires temporary arrays of size $N\cdot O$.
+For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $\Lambda$, requires temporary arrays of size $N\cdot O$.
 This can sometimes break the local memory threshold.
@@
 \end{enumerate}
-A corollary of $C_1$ is that hoisting expressions involving geometry-dependent terms outside of $LN$, which we discussed at the beginning of this section, is now forbidden.
+A corollary of $C_1$ is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which we discussed at the beginning of this section, is now forbidden.
 Consequently, the search space for optimality becomes smaller.
@@
 We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries.
+Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries.
 As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
@@
 Consider an integration routine originating from an arbitrary multilinear form.
-A loop nest $LN$, as previously characterized, evaluates an expression comprising a set of monomials, $M$.
+A loop nest $\Lambda$, as previously characterized, evaluates an expression comprising a set of monomials, $M$.
 Let $P$ be the set of monomials for which pre-evaluation is applied because found profitable, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
@@
 
-We distinguish two classes of loop nests rooted in $LN$: $[L_e, L_j, L_k]$, for the pre-evaluated monomials in $P$, and $[L_e, L_i, L_j, L_k]$, enclosing the remaining monomials in $Z$.
+We distinguish two classes of loop nests rooted in $\Lambda$: $[L_e, L_j, L_k]$, for the pre-evaluated monomials in $P$, and $[L_e, L_i, L_j, L_k]$, enclosing the remaining monomials in $Z$.
 Since they only differ for the presence of $L_i$, we relieve the notation by omitting the shared loops when discussing operation counts.
-The operation count of what we are proving to be the optimal $LN$ synthesis (inferable from Figure~\ref{code:loopnest-opt}) is $LN_{ops} = LN_{ops_1} + LN_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation count of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.
+The operation count of what we are proving to be the optimal $\Lambda$ synthesis (inferable from Figure~\ref{code:loopnest-opt}) is $\Lambda_{ops} = \Lambda_{ops_1} + \Lambda_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation count of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.
 
 We note that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting.
-This narrows the proof to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $LN_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $LN_{ops}$.
+This narrows the proof to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $\Lambda_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $\Lambda_{ops}$.
 
-A) We prove that $LN_{ops}' = LN_{ops_1}' + LN_{ops_2}' > LN_{ops}$.
-It is rather obvious that $LN_{ops_1}' \geq LN_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
-We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $LN_{ops_2}' = LN_{ops_2}$, so our statement is true.
-If, on the other hand, at least one monomial does not share any terms, we obtain $LN_{ops_2}' < LN_{ops_2}$ or, equivalently, $LN_{ops_2}' = LN_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by exposing more pre-evaluations, $LN_{ops_1}' - LN_{ops_1} =  \gamma \geq I \cdot\delta$ holds.
+A) We prove that $\Lambda_{ops}' = \Lambda_{ops_1}' + \Lambda_{ops_2}' > \Lambda_{ops}$.
+It is rather obvious that $\Lambda_{ops_1}' \geq \Lambda_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
+We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda_{ops_2}' = \Lambda_{ops_2}$, so our statement is true.
+If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda_{ops_2}' < \Lambda_{ops_2}$ or, equivalently, $\Lambda_{ops_2}' = \Lambda_{ops_2} - I \cdot\delta$.
+What we have to show now is that even by exposing more pre-evaluations, $\Lambda_{ops_1}' - \Lambda_{ops_1} =  \gamma \geq I \cdot\delta$ holds.
 Because of assumption 2), the new pre-evaluations cannot expose further sharing.
@@
 
-B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1).
+B) In absence of sharing, the statement is trivially true since we would have $\Lambda_{ops_2}' > \Lambda_{ops_2}$, being the cost function optimal due to assumption 1).
 Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities.
@@
 This was explained in Proposition~\ref{prop:optimal-mln}.
-A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $LN_{ops}^{sf}$.
-Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.
+A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{sf}$.
+Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = \Lambda_{ops}^{sf} \cdot I$.
 

### 2015-09-21 TOMS_Paper/assembly.tex
Rework section 2 entirely...

@@
 
-\section{Optimality of Loop Nests}
+\section{Problem Characterization}
 \label{sec:optimal-impl}
 
-In this section, we provide a characterization of loop nest optimality. In order to make the document self-contained, we start with reviewing basic compiler terminology.
+In this section, we characterize optimality, as well as the transformation space that needs be explored to achieve it, for finite element integration.
+
+\subsection{Loop Nests and Optimality}
+In order to make the document self-contained, we start with reviewing basic compiler terminology.
 
@@
 By mapping mathematical properties to the loop nest level, we introduce the
-concept of a \textit{linear loop}.
+concepts of \textit{linear loop} and, more generally, (perfect) multilinear loop nest.
 
 \begin{Def}[Linear loop]
-A loop $L$ defining the iteration space $I$ through the iteration variable
-$i$, or simply $L_i$, is \textit{linear} if 
+A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if 
 \begin{enumerate}
-\item $i$ appears in the body of $L$
-only as an array index, and
-\item whenever an array $a$ is indexed by $i$
-($a[i]$), all expressions in which this appears are affine in $a$.
+\item $i$ appears in the body of $L$ only as an array index, and
+\item whenever an array $a$ is indexed by $i$ ($a[i]$), all expressions in which this appears are affine in $a$.
 \end{enumerate}
@@
 
-In particular, we are interested in the following class since it naturally arises in finite element integration.
-
 \begin{Def}[Perfect multilinear loop nest]
-A \textit{perfect multilinear} loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
+A perfect multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
 \end{Def}
 
-Note that perfect multilinear loop nests could actually be rooted in deeper, possibly imperfect loop nests. In fact, we will focus on this particular structure.
+Since our focus is on finite element integration, for simplicity we restrict ourselves to the following, notable subclass of perfect multilinear loop nests.
 
-We introduce the fundamental notion of sharing.
+\begin{Def}[Tensor-product loop nest]
+A tensor-product loop nest is a perfect multilinear loop nest $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ in which all of the expressions appearing in the body of $L_{i_{n-1}}$ are (summations of) tensor products $t_{i_{0}, i_{1}, ..., i_{n-1}} = a_{i_{0}}b_{i_1}...z_{i_{n-1}}$.
+\end{Def}
+
+As shown later, tensor-product loop nests are important because they possess properties that simplify the synthesis of optimal code.
+They arise naturally when translating a multilinear form into code.
+Consider Equation~\ref{eq:quadrature} and the loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the tensor-product loop-nest $[L_j, L_k]$ over test and trial functions.
+The expression \texttt{F} implements the operator.
+
+% (the Laplacian operator in the case of Equation~\ref{eq:quadrature}, but our discussion applies in general).
+\begin{figure}[h]\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
+\end{CenteredBox}\caption{The typical loop nest implementing a bilinear form.}\label{code:loopnest}\end{figure}
+
+Our aim is to formulate a strategy to synthesize optimal implementations of such loop nests. In particular, we characterize optimality as follows.
+
+\begin{Def}[Optimality of a loop nest]
+\label{def:mln-optimality}
+The synthesis of a loop nest is optimal if the number of operations performed in the loop nest is minimal.
+\end{Def}
+
+Note that Definition~\ref{def:mln-optimality} does not take into account memory requirements.
+If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
+Henceforth we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated.
+In the context of finite element, this is often true for more complex multilinear forms and/or higher order elements.
+
+\subsection{Transformation Space}
+To synthesize optimal implementations we need:
+\begin{itemize}
+\item a characterization of the transformation space for the class of loop nests considered
+\item a cost model to select the optimal point in the transformation space
+\end{itemize}
+In this section, we construct the transformation space. We leave a few loose hands, which we progressively tie in Section~\ref{sec:optimal-synthesis}.
+
+We start with introducing the fundamental notion of sharing.
 
@@
 
-Figure~\ref{code:multi_loopnest_a} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.
+Figure~\ref{code:multi_loopnest_a} shows an example of a trivial tensor-product loop nest of arity $n=2$ with sharing along dimension $j$.
 
@@
 {\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
-\end{CenteredBox}\caption{A simple multilinear loop nest}\end{figure}
+\end{CenteredBox}\caption{A simple tensor-product loop nest}\end{figure}
 
-We now have the ingredients to formulate a simple yet fundamental result.
-
-\begin{Prop}
-\label{prop:share-removal}
-Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a perfect multilinear loop nest with the body of $L_{i_{n-1}}$ being an independent basic block.
-Then sharing can always be eliminated from $\Lambda$.
-\end{Prop}
-\begin{proof}
-  The demonstration is by construction and exploits linearity. We want to
-  transform $\Lambda$ into $\Lambda'$ such that there is no sharing in any
-  $L_i \in \Lambda'$. Starting from the innermost loop $L_{i_{n-1}}$, the
-  expressions are ``flattened'' by expanding all products involving terms
-  depending on $L_{i_{n-1}}$. Being now on the same level of the expression
-  tree, such terms can be factorized. Due to linearity, each factored
-  product only has one term depending on $L_{i_{n-1}}$, and this term is now
-  unique in the expression. The other terms, independent of $L_{i_{n-1}}$,
-  are, by definition, loop-invariant, and as such can be hoisted at the
-  level of $L_{i_{n-2}}$. This procedure can be applied recursively up to
-  $L_{i_{0}}$: multilinearity allows factorization at each level;
-  perfectness ensures hoisting is safe.
-\end{proof}
-
-This result will be essential to prove that optimality, characterized as follows, can always be reached in this class of loops.
-
-\begin{Def}[Optimality of a multilinear loop nest]
-\label{def:mln-optimality}
-The synthesis of a multilinear loop nest is optimal if the number of
-operations performed in the loop nest is minimal.
-\end{Def}
-
-%We will see that the restriction to inner loops has the effect of simplifying the transformation model; the price to pay will be the generation of quasi-optimal code in presence of costly outer loops (this, in finite element integration, happens when forms are characterized by extensive use of tensor algebra).
-
-Note that Definition~\ref{def:mln-optimality} does not take into account
-memory requirements. If the loop nest were memory-bound -- the ratio of
-operations to bytes transferred from memory to the CPU being too low -- then
-speaking of optimality would clearly make no sense. In the following, we
-assume to operate in a CPU-bound regime, in which arithmetic-intensive
-expressions need be evaluated. In the context of finite element, this is
-often true for more complex multilinear forms and/or higher order elements.
-
-%\begin{Rem*}
-%In the following, when discussing ``optimality'' or ``loop nests in optimal form'', we will always tacitly refer to inner optimality (unless explicitly specified differently). 
-%\end{Rem*}
-
-As anticipated, perfect multilinear loop nests can be brought to optimal form. The following result proves this claim.
+We prove that tensor-product loop nests can be reduced to optimal form (i.e., the number of operations in the resulting loop nest is minimal) by eliminating sharing through a suitable algorithm.
 
@@
 \label{prop:optimal-mln}
-A perfect multilinear loop nest $\Lambda$ with basic block independence can always be reduced to optimal form.
+Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a tensor-product loop nest with the body of $L_{i_{n-1}}$ being an independent basic block.
+Then we can determine an optimal $\Lambda'$ applying a sequence of steps aiming at eliminating sharing.
 \end{Prop}
 \begin{proof}
-By construction. Loop-dependent terms are logically grouped into $n$
-disjoint sets $S_i$, each $S_i$ containing all terms depending on
-$L_i$. These sets are sorted in descending order based on their
-cardinality. We produce a new loop permutation that reflects the ${S_i}$
-ordering; that is, loops are placed such that each loop has fewer unique
-terms than the outer loops containing it. Since the loop nest is perfect,
-any such permutation is semantically correct.
+The demonstration is by construction and exploits linearity (recall that tensor-product loop nests are multilinear by definition).
+We start with ``flattening'' the expressions appearing in the body of $L_{i_{n-1}}$; that is, products are recursively expanded.
+This exposes the space of all possible factorizations.
+The terms appearing in the expanded expressions are tensor products of the form $a_{i_{0}}b_{i_1}...z_{i_{n-1}}$ (a tensor product involves $n$ arrays , each array indexed by a linear loop, or simply symbols).
+
+We logically group the symbols into $n$ disjoint sets $S_i$, each $S_i$ containing the symbols indexed through $L_i$.
+These sets are sorted in ascending order based on their cardinality.
+Based on this, we create the factorization list: the ordered list of all factorizable symbols.
+We then progressively try to factorize these symbols.
+That is, we start factorizing along the loop that has the least number of unique symbols, and then we go on with the other loops.
+
+Due to linearity, each factored product only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
+The other symbols in the term, independent of $L_i$, are, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
+This procedure is always semantically correct, regardless of the symbols (loops) being considered: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
+
+Two observations related to the aforementioned code motion: 1) hoisting might require the creation of ``clone loops'' outside of $\Lambda$ to honor the potential dependency on loop indices; 2) hoisted expressions are replaced with temporaries, which need be added to a suitable set $S_i$ of factorizable symbols .
+
+It remains to clarify why the chosen factorization ordering leads to optimality.
+Naturally, as we factorize a symbol, we are potentially pruning other factorization opportunities.
+If that is the case, we might wonder whether there was a better factorization strategy.
+.
+Therefore, this is not possible (in the best case, the resulting factorization would be identical).
+
+%The key observation is that in a tensor-product loop nest, all terms depend on $\Lambda$ through $n$ distinct symbols.
+
+Same cardinality, try both
+\end{proof}
+
+%\begin{proof}
+%By construction. Loop-dependent terms are logically grouped into $n$
+%disjoint sets $S_i$, each $S_i$ containing all terms depending on
+%$L_i$. These sets are sorted in descending order based on their
+%cardinality. We produce a new loop permutation that reflects the ${S_i}$
+%ordering; that is, loops are placed such that each loop has fewer unique
+%terms than the outer loops containing it. Since the loop nest is perfect,
+%any such permutation is semantically correct.
 Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
@@
 In particular, the number of operations is equal to $\Lambda_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
-\end{proof}
+%\end{proof}
 
-For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.
-\\
-\begin{figure}[h]\begin{CenteredBox}
-\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
-\end{CenteredBox}\caption{An imperfect loop nest, with one reduction loop ($L_i$), an inner perfect multilinear loop nest ($[L_j, L_k]$), and sharing potentially present in $F$}\label{code:loopnest}\end{figure}
+For example, by applying this procedure to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.
 
-We now consider a wider class of loop nests in which perfectness and multilinearity apply to a sub-nest only.
-Consider the example in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear.
-The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication, possibly characterized by sharing.
-Here, one could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count.
-This is exemplified through Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored} by fixing $F$ in a simpler loop nest.
+However, we observe that, for the larger class of finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
+Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
+We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$?
+The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} (input) and~\ref{code:loopnest_nored} (output); Figure~\ref{code:loopnest_red} can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.
 
@@
 {\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
-\end{CenteredBox}\caption{The effect of pre-evaluating reductions. Note the decrease in operation count grows with the loop nest depth.}\end{figure}
+\end{CenteredBox}\caption{Exposition (through factorization) and pre-evaluation of a reduction.}\end{figure}
 
-It is then obvious that reaching an optimal form $\Lambda'$ is, in the general case, a challenging issue, since a full exploration of the expression tree transformation space is necessary.
-Even though we assumed we could generate $\Lambda'$ through a suitable transformation algorithm, the following issues should be addressed
+Pre-evaluation opportunities could have to be exposed, for instance through an exploration of the expression tree transformation space, which can be challenging.
+Further, the following issues arise when considering pre-evaluation opportunities:
 \begin{itemize}
-\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example of Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item as opposed to what happens with hoisting in tensor-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
-\item the transformations exposing sub-expressions amenable to pre-evaluation could increase the arithmetic complexity within $L_k$ (e.g., expansion may increase the operation count; further examples will be provided later).
+\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later).
 This could overwhelm the gain inherent in pre-evaluation.
@@
 
-Before attacking this problem in Section~\ref{sec:optimal-synthesis}, we shall conclude our construction refining Definition~\ref{def:mln-optimality} for generic loop nests
+To summarize, so far we have highlighted four problems:
+\begin{enumerate}
+\item The need for an algorithm to expose pre-evaluation opportunities
+\item Potential explosion in working set size
+\item Potential increase in operation count due to manipulation of the expression tree
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities
+\end{enumerate}
 
-\begin{Def}[Inner optimality of a loop nest]
+We tackle these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2, it is useful to conclude this section refining our optimality statement as follows.
+
+\begin{Def}[Optimality of a loop nest with bounded working set]
 \label{def:optimality}
-The synthesis of a loop nest is optimal if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
+The synthesis of a loop nest is optimal if, under a set of memory constraints $C$, the number of operations performed in the loop nest is minimal.
 \end{Def}
 
-Note how the definition contemplates the possibility for a nest to have multiple innermost loops.
-In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).
-
-\section{Optimal-form Loop Nests in Finite Element Integration}
+\section{Optimal Integration Routines}
 \label{sec:optimal-synthesis}
 
-In this section, we instantiate Definition~\ref{def:optimality} to our domain of interest, finite element integration.
-This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.
+In this section, we will reason at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.
 
 Our point of departure is the example loop nest in Figure~\ref{code:loopnest}.
-This loop nest is actually a simplified view of a typical bilinear form implementation.
-$L_e$ represents iteration over the elements of a mesh; $L_i$ derives from using numerical quadrature; $[L_j, L_k]$ implement the computation over test and trial functions.
-We deliberately omitted useless portions of code to not hinder readability and to avoid tying our discussion to specific forms.
-
 We make the following observations.

### 2015-09-22 TOMS_Paper/assembly.tex
Reworked section 2 with outer-product loop nests

@@
 
-In this section, we characterize optimality, as well as the transformation space that needs be explored to achieve it, for finite element integration.
+In this section, we characterize optimality for finite element integration, as well as the transformation space that needs be explored to achieve it.
 
@@
 By mapping mathematical properties to the loop nest level, we introduce the
-concepts of \textit{linear loop} and, more generally, (perfect) multilinear loop nest.
+concepts of a \textit{linear loop} and, more generally, a (perfect) multilinear loop nest.
 
@@
 
-\begin{Def}[Tensor-product loop nest]
-A tensor-product loop nest is a perfect multilinear loop nest $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ in which all of the expressions appearing in the body of $L_{i_{n-1}}$ are (summations of) tensor products $t_{i_{0}, i_{1}, ..., i_{n-1}} = a_{i_{0}}b_{i_1}...z_{i_{n-1}}$.
+\begin{Def}[Outer-product loop nest]
+An outer-product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations of outer products.
 \end{Def}
 
-As shown later, tensor-product loop nests are important because they possess properties that simplify the synthesis of optimal code.
-They arise naturally when translating a multilinear form into code.
-Consider Equation~\ref{eq:quadrature} and the loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the tensor-product loop-nest $[L_j, L_k]$ over test and trial functions.
+As shown later, outer-product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code.
+They arise naturally when translating bilinear or linear forms into code.
+Consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the outer-product loop nest $[L_j, L_k]$ over test and trial functions.
 The expression \texttt{F} implements the operator.
@@
 
-Our aim is to formulate a strategy to synthesize optimal implementations of such loop nests. In particular, we characterize optimality as follows.
+Our aim is a strategy to synthesize optimal loop nests of this form. In particular, we characterize optimality as follows.
 
@@
 \label{def:mln-optimality}
-The synthesis of a loop nest is optimal if the number of operations performed in the loop nest is minimal.
+Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations that it performs to evaluate the result is minimal.
 \end{Def}
@@
 \begin{Def}[Sharing]
-A loop $L_i$ presents \textit{sharing} if it contains at least two expressions depending on $i$ that are symbolically identical. 
+A loop $L_i$ presents sharing if it contains at least one statement $S$ in which two (sub-)expressions depending on $i$ are symbolically identical.
 \end{Def}
 
-Figure~\ref{code:multi_loopnest_a} shows an example of a trivial tensor-product loop nest of arity $n=2$ with sharing along dimension $j$.
+\begin{Def}[Sharing factor]
+The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$).
+It provides an intuition of how many symbols can ideally be factored along $L_i$
+\end{Def}
+
+Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer-product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
 
@@
 {\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
-\end{CenteredBox}\caption{A simple tensor-product loop nest}\end{figure}
+\end{CenteredBox}\caption{A simple outer-product loop nest. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.}\end{figure}
 
-We prove that tensor-product loop nests can be reduced to optimal form (i.e., the number of operations in the resulting loop nest is minimal) by eliminating sharing through a suitable algorithm.
+We prove that outer-product loop nests can be reduced to optimal form.
+The proof focuses on outer-product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, which is beyond the scope of this paper, a more general treatise would be required to avoid sub-optimal outer loops in pathological cases).
 
@@
 \label{prop:optimal-mln}
-Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a tensor-product loop nest with the body of $L_{i_{n-1}}$ being an independent basic block.
-Then we can determine an optimal $\Lambda'$ applying a sequence of steps aiming at eliminating sharing.
+Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}]$ is an outer-product loop nest with the body of $L_{i_{1}}$ being an independent basic block.
+Then an optimal synthesis $\Lambda'$ can be determined by eliminating sharing from both loops, starting from the loop with smallest sharing factor.
 \end{Prop}
 \begin{proof}
-The demonstration is by construction and exploits linearity (recall that tensor-product loop nests are multilinear by definition).
-We start with ``flattening'' the expressions appearing in the body of $L_{i_{n-1}}$; that is, products are recursively expanded.
-This exposes the space of all possible factorizations.
-The terms appearing in the expanded expressions are tensor products of the form $a_{i_{0}}b_{i_1}...z_{i_{n-1}}$ (a tensor product involves $n$ arrays , each array indexed by a linear loop, or simply symbols).
+The demonstration is by construction and exploits linearity (recall that outer-product loop nests are multilinear by definition).
+We start with ``flattening'' the expressions appearing in the body of $L_{i_{1}}$; this is because, depending on the input structure, products might have to be recursively expanded in order to expose the space of all possible factorizations.
+The expanded expressions are summations of outer products (otherwise, the assumption of $\Lambda$ being an outer-product loop nest would trivially be contradicted).
 
-We logically group the symbols into $n$ disjoint sets $S_i$, each $S_i$ containing the symbols indexed through $L_i$.
-These sets are sorted in ascending order based on their cardinality.
-Based on this, we create the factorization list: the ordered list of all factorizable symbols.
-We then progressively try to factorize these symbols.
-That is, we start factorizing along the loop that has the least number of unique symbols, and then we go on with the other loops.
+We traverse the expression tree and determine the sharing factor of the two linear loops.
+Starting from the loop with smallest sharing factor, $L_i$, we incrementally factorize its symbols.
+Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
+The other expression in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
+This transformation is semantically correct, regardless of $L_i$ being $L_{i_{0}}$ or $L_{i_{1}}$: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about code motion).
+We repeat the same process for the loop with higher sharing factor, $L_j$.
 
-Due to linearity, each factored product only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
-The other symbols in the term, independent of $L_i$, are, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
-This procedure is always semantically correct, regardless of the symbols (loops) being considered: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
-
-Two observations related to the aforementioned code motion: 1) hoisting might require the creation of ``clone loops'' outside of $\Lambda$ to honor the potential dependency on loop indices; 2) hoisted expressions are replaced with temporaries, which need be added to a suitable set $S_i$ of factorizable symbols .
-
-It remains to clarify why the chosen factorization ordering leads to optimality.
-Naturally, as we factorize a symbol, we are potentially pruning other factorization opportunities.
-If that is the case, we might wonder whether there was a better factorization strategy.
-.
-Therefore, this is not possible (in the best case, the resulting factorization would be identical).
-
-%The key observation is that in a tensor-product loop nest, all terms depend on $\Lambda$ through $n$ distinct symbols.
-
-Same cardinality, try both
+We now have to demonstrate that the chosen factorization ordering, based on sharing factors, leads to optimality. This reduces to prove that the two following facts are true
+\begin{enumerate}
+\item Taking $L_j$, the loop with higher sharing factor, as point of departure leads to a sub-optimal $\Lambda'$ (optimal only in the best case).
+By contradiction, let us assume this is false.
+There are two possibilities: A) both $L_i$ and $L_j$ present sharing, but this occurs over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$.
+Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ is actually optimal (best case scenario).
+In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had taken $L_i$ as point of departure.
+Therefore, since $L^{sf}_{j} > L^{sf}_{i}$, the innermost loop results more expensive.
+We still wonder about the cost of hoisting, since clearly we are hoisting less expressions in this scenario.
+However, a hoisted symbol always costs one operation (a sum), while a non-factored outer product always costs two operations (a product and a sum).
+That is, factorization is always convenient.
+The initial assumption is therefore wrong.
+\item Partial elimination of sharing leads to a sub-optimal $\Lambda'$ (optimal only in the best case).
+By contradiction, let us assume this is false.
+Because of linearity, we know that factorization eliminates sharing.
+We wonder what happens in the case of partial factorization; that is, at least one factorizable symbol, say along $L_i$, is left outside of the factored product (e.g., $a_i$ in $a_i(b_j + c_j + ...) + a_i e_j$).
+It is fairly simple to see that either there is a factorization opportunity along $L_j$, in which case no penalty needs be paid; or $\Lambda'$ is sub-optimal, since eventually the innermost loop will perform more operations.
+The initial assumption is therefore contradicted.
+Note that, if we were considering arities $n > 2$, this fact would generally be false.
+\end{enumerate}
 \end{proof}
-
-%\begin{proof}
-%By construction. Loop-dependent terms are logically grouped into $n$
-%disjoint sets $S_i$, each $S_i$ containing all terms depending on
-%$L_i$. These sets are sorted in descending order based on their
-%cardinality. We produce a new loop permutation that reflects the ${S_i}$
-%ordering; that is, loops are placed such that each loop has fewer unique
-%terms than the outer loops containing it. Since the loop nest is perfect,
-%any such permutation is semantically correct.
-Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}.
-This renders $L_{i_{n-1}}$ optimal.
-In particular, the number of operations is equal to $\Lambda_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
-%\end{proof}
 
@@
 
-However, we observe that, for the larger class of finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
+We now observe that, for the larger class of finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
 Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
@@
 \begin{itemize}
-\item as opposed to what happens with hoisting in tensor-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item as opposed to what happens with hoisting in outer-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
@@
 This was explained in Proposition~\ref{prop:optimal-mln}.
-A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{sf}$.
-Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = \Lambda_{ops}^{sf} \cdot I$.
+A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$.
+Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.
 

### 2015-09-22 TOMS_Paper/assembly.tex
Start rewriting section 3...

@@
 
-Pre-evaluation opportunities could have to be exposed, for instance through an exploration of the expression tree transformation space, which can be challenging.
+Pre-evaluation opportunities could be exposed through an exploration of the expression tree transformation space.
+For arbitrary loop nests, this can be challenging.
 Further, the following issues arise when considering pre-evaluation opportunities:
@@
 \item Potential increase in operation count due to manipulation of the expression tree
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could prevent pre-evaluation; pre-evaluation might expose further sharing)
 \end{enumerate}
 
-We tackle these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2, it is useful to conclude this section refining our optimality statement as follows.
+We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we conclude this section refining our optimality statement as follows.
 
@@
 \label{def:optimality}
-The synthesis of a loop nest is optimal if, under a set of memory constraints $C$, the number of operations performed in the loop nest is minimal.
+Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and a set of memory constraints is satisfied.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations that it performs to evaluate the result is minimal.
 \end{Def}
 
-\section{Optimal Integration Routines}
+\section{Transformation Algorithm}
 \label{sec:optimal-synthesis}
 
-In this section, we will reason at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.
-
-Our point of departure is the example loop nest in Figure~\ref{code:loopnest}.
-We make the following observations.
-1) $L \gg M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.
-
-% maybe say somewhere the order of the loop nest eijk is just the most natural --- the other present worse data locality for example
+In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form.
+A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation, the transformation introduced in the previous section (and later further expanded).
+The optimality of the transformation algorithm is discussed.
 
@@
 \label{sec:mem-const}
-The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting out of $\Lambda$.
+We provide intuitions about the need for memory constraints.
+Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}.
+Analogous considerations apply to forms of arbitrary arity.
+
+The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions.
 Imagine $\Lambda$ is enclosed in a time stepping loop.
-One could think of exposing and then hoisting time-invariant sub-expressions.
-For those sub-expressions that depend on the mesh geometry (i.e., they would depend on $L_e$), code motion would increase the working set by a factor $L$.
-The gain in number of operations executed could then be overwhelmed, from a runtime viewpoint, by a much larger memory pressure.
+One could think of exposing (through some transformations) and hoisting any time-invariant sub-expressions to minimize redundant computation at every time step.
+The working set size could then increase by a factor $L$.
+The gain in number of operations executed could therefore be overwhelmed, from a runtime viewpoint, by a much larger memory pressure.
 
-A second, more general fact is that, for certain forms and discretizations, excessive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
+A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $\Lambda$, requires temporary arrays of size $N\cdot O$.
-This can sometimes break the local memory threshold.
-We will clarify and experiment this in Section~\ref{sec:perf-results-forms}.
+For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
+This can sometimes break a ``local memory threshold''.
+We will report results of our experiments on the memory threshold in Section~\ref{sec:perf-results-forms}.
 
-Based on these considerations, we characterize the set of memory constraints $C$ (see Definition~\ref{def:optimality}) as follows
+Based on these considerations, we establish the following set of memory constraints
 \begin{enumerate}
-\item The size of a temporary due to code motion cannot be larger than that of the multilinear iteration space. 
-\item The total amount of memory occupied by the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
+\item The size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space. 
+\item The total amount of memory occupied by the temporaries created by code motion cannot exceed a threshold \texttt{$T_H$}
 \end{enumerate}
-A corollary of $C_1$ is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which we discussed at the beginning of this section, is now forbidden.
-Consequently, the search space for optimality becomes smaller.
+A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden.
+This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.
 
-\subsection{Minimizing the Operation Count in Inner Loops}
-Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the innermost loop operation counts is minimum.
-We now discuss how we can systematically achieve this.
+\subsection{Minimization of the Operation Count}
+We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.
 
-Eliminating sharing from the inner multilinear loops does not suffice.
-In fact, as suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
+We reinforce that eliminating sharing from the outer-product loop nest does not suffice.
+As suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
 
-To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}.
+To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}.
 Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}).
 We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries.
-As part of this pre-evaluation, the reduction induced by $L_i$ vanishes.
-In the following, we simply refer to this special sort of code hoisting as ``\textit{pre-evaluation}''.
+Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
+These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries.
+As part of this process, the reduction induced by $L_i$ is computed and, therefore, disappears from $\Lambda$.
+This is actually what we were referring to when we introduced the pre-evaluation transformation in Section~\ref{sec:optimal-impl}.
 
-The challenge is to understand when, and for which monomials, pre-evaluation is profitable.
-We propose an algorithm and a discussion of its optimality.
-The intuition of the main algorithm for reducing a loop nest to optimal form is shown in Figure~\ref{code:intuition}.
+There are two issues: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter prevents the former.
+We start with the second issue.
+For this, it is convenient to introduce the transformation algorithm.
+Figure~\ref{code:intuition} provides the pseudo-code of the algorithm.
+An explanation and a discussion on its optimality follow.
 
@@
 \end{lstlisting}
-\caption{Intuition of the main algorithm}
+\caption{Intuition of the transformation algorithm}
 \label{code:intuition}
@@
 
-\subsection{A-Priori Operation Counting}
+\subsection{Cost Function}
 \label{sec:op_count}

### 2015-09-23 TOMS_Paper/assembly.tex
Rewrite bits of section 2 (the proof mostly)

@@
 \begin{itemize}
-\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration.
+\item We characterize loop nest optimality and we instantiate this concept to finite element integration.
 As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
-\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into flop-optimal loop nests.
+\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into optimal loop nests.
 \item We comment on the cases in which such model does not lead to optimal loop nests.
@@
 
-Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer-product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
-
 \begin{figure}[h]\begin{CenteredBox}
@@
 
-We prove that outer-product loop nests can be reduced to optimal form.
-The proof focuses on outer-product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, which is beyond the scope of this paper, a more general treatise would be required to avoid sub-optimal outer loops in pathological cases).
+Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer-product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
+Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}.
+More in general, arbitrarily complex outer-product loop nests can systematically be reduced to optimal form.
+The constructive proof that we provide below focuses on outer-product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, as we elaborate later, a more general treatise would be required).
 
@@
 \begin{proof}
-The demonstration is by construction and exploits linearity (recall that outer-product loop nests are multilinear by definition).
-We start with ``flattening'' the expressions appearing in the body of $L_{i_{1}}$; this is because, depending on the input structure, products might have to be recursively expanded in order to expose the space of all possible factorizations.
-The expanded expressions are summations of outer products (otherwise, the assumption of $\Lambda$ being an outer-product loop nest would trivially be contradicted).
+The demonstration is by construction and exploits linearity.
+We start with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_{i_{1}}$; this exposes the space of all possible factorizations.
+The resulting expressions are summations of outer products (otherwise, the assumption of $\Lambda$ being an outer-product loop nest would trivially be contradicted).
 
 We traverse the expression tree and determine the sharing factor of the two linear loops.
-Starting from the loop with smallest sharing factor, $L_i$, we incrementally factorize its symbols.
+We take the symbols depending on the loop with smallest sharing factor, $L_i$, and we incrementally factorize them.
 Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
-The other expression in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
-This transformation is semantically correct, regardless of $L_i$ being $L_{i_{0}}$ or $L_{i_{1}}$: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about code motion).
-We repeat the same process for the loop with higher sharing factor, $L_j$.
+The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
+We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$.
+The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about generalized code motion).
 
-We now have to demonstrate that the chosen factorization ordering, based on sharing factors, leads to optimality. This reduces to prove that the two following facts are true
+We now have to demonstrate that the chosen factorization strategy, based on sharing factors, leads to optimality.
+This reduces to prove that the two following facts are true (we do it by contradiction).
 \begin{enumerate}
-\item Taking $L_j$, the loop with higher sharing factor, as point of departure leads to a sub-optimal $\Lambda'$ (optimal only in the best case).
-By contradiction, let us assume this is false.
-There are two possibilities: A) both $L_i$ and $L_j$ present sharing, but this occurs over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$.
-Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ is actually optimal (best case scenario).
-In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had taken $L_i$ as point of departure.
-Therefore, since $L^{sf}_{j} > L^{sf}_{i}$, the innermost loop results more expensive.
-We still wonder about the cost of hoisting, since clearly we are hoisting less expressions in this scenario.
-However, a hoisted symbol always costs one operation (a sum), while a non-factored outer product always costs two operations (a product and a sum).
-That is, factorization is always convenient.
-The initial assumption is therefore wrong.
-\item Partial elimination of sharing leads to a sub-optimal $\Lambda'$ (optimal only in the best case).
-By contradiction, let us assume this is false.
-Because of linearity, we know that factorization eliminates sharing.
-We wonder what happens in the case of partial factorization; that is, at least one factorizable symbol, say along $L_i$, is left outside of the factored product (e.g., $a_i$ in $a_i(b_j + c_j + ...) + a_i e_j$).
-It is fairly simple to see that either there is a factorization opportunity along $L_j$, in which case no penalty needs be paid; or $\Lambda'$ is sub-optimal, since eventually the innermost loop will perform more operations.
-The initial assumption is therefore contradicted.
-Note that, if we were considering arities $n > 2$, this fact would generally be false.
+\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case).
+If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$.
+Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario).
+In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$.
+However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product.
+Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
+\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case).
+Let us assume, for a moment, that this is false.
+In this scenario, at least one symbol is excluded from the factored product.
+Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$.
+It is fairly obvious to see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal.
+Consequently, the hypothesis is generally contradicted.
 \end{enumerate}
+
+Note that if we were considering outer-product loop nests of arity $n > 2$, unless relaxing Definition~\ref{def:mln-optimality}, we would need a much more complex construction to guarantee optimality.
 \end{proof}
-
-For example, by applying this procedure to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.
 
@@
 
-\subsection{Minimization of the Operation Count}
+\subsection{Operation Count Minimization}
 We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.

### 2015-09-23 TOMS_Paper/assembly.tex
Rewrite bits of section 3

@@
 
-\subsection{Operation Count Minimization}
+\subsection{Reaching Optimality}
 We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.
 
-We reinforce that eliminating sharing from the outer-product loop nest does not suffice.
+We reinforce the idea that eliminating sharing from the outer-product loop nest does not suffice.
 As suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
@@
 Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
-These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries.
-As part of this process, the reduction induced by $L_i$ is computed and, therefore, disappears from $\Lambda$.
-This is actually what we were referring to when we introduced the pre-evaluation transformation in Section~\ref{sec:optimal-impl}.
+These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$ and stored into temporaries.
+As part of this process, the reduction induced by $L_i$ is evaluated.
+Consequently, $L_i$ disappears from $\Lambda$.
+This transformation is actually what we were referring to when we introduced pre-evaluation in Section~\ref{sec:optimal-impl}.
 
-There are two issues: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter prevents the former.
-We start with the second issue.
-For this, it is convenient to introduce the transformation algorithm.
-Figure~\ref{code:intuition} provides the pseudo-code of the algorithm.
-An explanation and a discussion on its optimality follow.
+Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former.
+In this section, we tackle the second issue, while we deal with the second in Section~\ref{sec:op_count}.
 
-\begin{figure}[h]
+We start with introducing the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion about its optimality follow.
+
+%TODO add this somewhere: Further, we note that because of the memory constraints discussed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space.
+
+\begin{figure}
 \begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
@@
 
-We study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, in isolation.
-If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed.
+Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''.
+If we estimate that, for a given monomial $M$, pre-evaluation will decrease the operation count, then: the sub-expression implementing $M$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
 The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.
-Identical tables are mapped to the same temporary.
-Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}.
-The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.
+
+Because of symmetries in basis functions, pre-evaluation could produce identical tables.
+These are mapped to the same temporary.
+The transformed expression, therefore, can be characterized by sharing.
+This is why the last two lines of the algorithm, which goal is to eliminate sharing following a procedure as described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all expressions, regardless of whether pre-evaluation was used.
+The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the input is still the usual loop nest in Figure~\ref{code:loopnest}.
 
@@
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
-\end{CenteredBox}\caption{A possible loop nest for the local assembly of a form after applying pre-evaluation to some of the monomials (resulting in sub-expressions \texttt{G'}, \texttt{G''}, ...)}\label{code:loopnest-opt}\end{figure}
+\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}
 
-Before elaborating on the profitability of pre-evaluation (i.e., how to determine the cost function $\theta$ in Figure~\ref{code:intuition}), we need to discuss under which conditions this approach, based on a ``local analysis'' of monomials, is able to reduce loop nests to optimal form.
+We elaborate on the optimality of our approach.
+First of all, we note that, because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space.
+The search for the optimal is then performed by the algorithm in a suitably constructed space.
+With the following Proposition, we finally aim to formalize the optimality claim.
 
@@
 \label{prop:optimal-approach}
-Consider an integration routine originating from an arbitrary multilinear form.
-A loop nest $\Lambda$, as previously characterized, evaluates an expression comprising a set of monomials, $M$.
-Let $P$ be the set of monomials for which pre-evaluation is applied because found profitable, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
+Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it.
+The form comprises a set of monomials, $M$.
+Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
 Assume that:
 \begin{enumerate}
-\item the cost function $\theta$ employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
+\item the cost function $\theta$ is optimal; that is, it predicts correctly the profitability of pre-evaluation
 \item pre-evaluating different monomials does not result in identical tables
-\item monomials in $P$ do not share terms
+\item there is no sharing of terms among monomials in $P$
 \end{enumerate} 
-Then, the main algorithm in Figure~\ref{code:intuition} leads to inner optimality under the set of memory constraints $C$ (defined in Section~\ref{sec:mem-const}).
+Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
 \end{Prop}
@@
 We first comment on the assumptions.
-1) We postpone the discussion of how to build $\theta$ to Section~\ref{sec:op_count}.
-2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions.
-This is however a pathological case that we can harmlessly ignore.
+1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}.
+2) This is inherently a pathological case, from which we abstract.
 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
-We abstract from this case, which otherwise would require a ``global'' analysis of the monomials in the form that we believe would not justify the gain (if any) in operation count.
+We also abstract from this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.
 
-We distinguish two classes of loop nests rooted in $\Lambda$: $[L_e, L_j, L_k]$, for the pre-evaluated monomials in $P$, and $[L_e, L_i, L_j, L_k]$, enclosing the remaining monomials in $Z$.
-Since they only differ for the presence of $L_i$, we relieve the notation by omitting the shared loops when discussing operation counts.
-The operation count of what we are proving to be the optimal $\Lambda$ synthesis (inferable from Figure~\ref{code:loopnest-opt}) is $\Lambda_{ops} = \Lambda_{ops_1} + \Lambda_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation count of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.
+The operation count for $\Lambda'$ is $\Lambda'_{ops} = \Lambda'_{ops_1} + \Lambda'_{ops_2} = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively.
+$\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.
 
-We note that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting.
-This narrows the proof to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $\Lambda_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $\Lambda_{ops}$.
+We prove our claim by demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$.
 
-A) We prove that $\Lambda_{ops}' = \Lambda_{ops_1}' + \Lambda_{ops_2}' > \Lambda_{ops}$.
-It is rather obvious that $\Lambda_{ops_1}' \geq \Lambda_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
-We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda_{ops_2}' = \Lambda_{ops_2}$, so our statement is true.
-If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda_{ops_2}' < \Lambda_{ops_2}$ or, equivalently, $\Lambda_{ops_2}' = \Lambda_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by exposing more pre-evaluations, $\Lambda_{ops_1}' - \Lambda_{ops_1} =  \gamma \geq I \cdot\delta$ holds.
+A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
+We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is true.
+If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$.
+What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} =  \gamma \geq I \cdot\delta$ holds.
 Because of assumption 2), the new pre-evaluations cannot expose further sharing.
@@
 
-B) In absence of sharing, the statement is trivially true since we would have $\Lambda_{ops_2}' > \Lambda_{ops_2}$, being the cost function optimal due to assumption 1).
+B) In absence of sharing, the statement is trivially true since we would have $\Lambda''_{ops_2} > \Lambda'_{ops_2}$, being the cost function optimal due to assumption 1).
 Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities.

### 2015-09-23 TOMS_Paper/assembly.tex
Update the plots

@@
 \item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
-\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
-\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
+\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
+\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
 \end{itemize}
@@
 
-For what concerns the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache).
+For what concerns the second memory constraint introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache).
 For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ to assess our hypotheses.
@@
 \begin{itemize}
-\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
-\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
+\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test, trial, and coefficient (or ``pre-multiplying'') functions always have same degree
 \item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
@@
 
-The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) can be interpreted as ``nested'' plots.
-We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''.
-In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis.
-The top-left plot in a grid shows the speed-up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on.
-The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$.
-A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved.
-In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}).
-There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.
+The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are three-dimensional plots.
+The three axes are for $q$, $nf$, and code transformation system.
+The code transformation system varies within different subplots, such that one can easily evince the fastest alternative for each problem instance: the taller a bar, the better, since, as explained, we are reporting speed-ups.
 
@@
 \begin{figure}
-\includegraphics[scale=0.7]{perf-results/elasticity}
+\includegraphics[scale=0.77]{perf-results/elasticity}
 \caption{Performance evaluation for the bilinear form arising in a elastic model.
@@
 The third point concerns the effect of the chosen discretizations on autovectorization.
-The discretizations employed ensure both multilinear loop nest and tabulated basis function sizes to be multiple of the machine vector length.
+The discretizations employed ensure both multilinear loop nests and tabulated basis function sizes to be multiple of the machine vector length.
 Given the suitable choice of compilation options, this promotes autovectorization in the majority of code variants.
 The biggest exception is \texttt{quad}, in which, due to the presence of indirection arrays in the generated code, vectorization is rarely triggered.
-In \texttt{tens}, the loop nests are fully unrolled, so the standard loop vectorization is simply not possible; however, manual inspection of the compiled code suggests block vectorization (\cite{SLP-vect}) is often triggered.
-In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces tend to have the same ``structure'' (there are very few exceptions due to low level optimization), with loop vectorization being regularly applied, as far as we could evince.
-It is worth noting the fact that the size of loops and tabulated basis functions is a multiple of the vector length render padding and data alignment (in \texttt{cfO1} and \texttt{cfO2}), to the best of our understanding, basically irrelevant.
+In \texttt{tens}, the loop nests are fully unrolled, so the standard loop vectorization is not possible; however, manual inspection of the compiled code suggests block vectorization (\cite{SLP-vect}) is often triggered.
+In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces tend to have the same structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could understand.
 

### 2015-09-24 TOMS_Paper/assembly.tex
Several fixes, update plots, descriptions, etc

@@
 The Intel \texttt{icc 15.2} compiler was used.
-The compilation flags used were \texttt{-O3, -xHost, -ip}, which can trigger \texttt{AVX} autovectorization.
+The compilation flags used were \texttt{-O3, -xHost, -ip}.
 
-We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
+We analyze the runtime performance of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
 In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}'').
@@
 
-We evaluate the speed-ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
+We evaluate the speed-ups achieved by a wide variety of transformation systems over ``original code'' produced by the FEniCS Form Compiler (i.e., no optimizations applied).
+We analyze the following transformation systems
 \begin{itemize}
@@
 They include the cost of local assembly as well as the cost of matrix insertion.
-However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.
+However, the unstructured mesh used for the simulations (details below) was chosen small enough to fit the L3 cache of the CPU so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.
 
-For the fairest comparison possible, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants.
-UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS, which employs a data storage layout different than Firedrake.
-Our patches fix this problem by producing code with a data storage layout as expected by Firedrake.
+For a fair comparison, small patches (publicly available) were written to run \textit{all} simulations through Firedrake.
+This means the costs of matrix insertion and mesh iteration are identical in all variants.
+Our patches make UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for Firedrake, which employs a data storage layout different than that of FEniCS (e.g., array of pointers instead of pointer to pointers).
 
-For what concerns the second memory constraint introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache).
-For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ to assess our hypotheses.
-Results of these trials are discussed below.
+In Section~\ref{sec:mem-const}, we discussed the importance of memory constraints.
+We then define $T_H$ as the maximum amount of space that temporaries due to code motion can take.
+We set $T_H = L2_{size}$, that is, the size of the processor L2 cache (the last level of private cache).
+We recall that exceeding this threshold prevents the application of pre-evaluation.
+In our experiments, this happened in some circumstances.
+In such cases, experiments were repeated with $T_H = L3_{size}$ to verify the hypotheses made in Section~\ref{sec:mem-const}.
+We later elaborate on this.
 
-Following the methodology adopted in~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
+Following the methodology adopted in~\citeN{quadrature1}, we vary the following parameters:
 \begin{itemize}
-\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test, trial, and coefficient (or ``pre-multiplying'') functions always have same degree
+\item the polynomial order of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
 \item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
 \end{itemize}
-Constants of our analysis, instead, are
+While constants of our study are
 \begin{itemize}
@@
 \item the mesh: tetrahedral with a total of 4374 elements
-\item exact numerical quadrature (the same scheme as in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule, is employed)
+\item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
 \end{itemize}
 
-The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are three-dimensional plots.
-The three axes are for $q$, $nf$, and code transformation system.
-The code transformation system varies within different subplots, such that one can easily evince the fastest alternative for each problem instance: the taller a bar, the better, since, as explained, we are reporting speed-ups.
-
-%TODO : can we just say padding is useless because in 3D ndofs is a multiple of our vector length?
+We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
+The axes represent $q$, $nf$, and code transformation system.
+We show one subplot for each problem instance $[q, nf]$, with the code transformation system varying within each subplot.
+The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
 
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
-\caption{Performance evaluation for the mass matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
+\caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
 \end{figure}
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
-\caption{Performance evaluation for the bilinear form of a Helmholtz equation.
+\caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation.
 The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
@@
 \includegraphics[scale=0.77]{perf-results/elasticity}
-\caption{Performance evaluation for the bilinear form arising in a elastic model.
+\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model.
 The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
@@
  \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
-\caption{Performance evaluation for the bilinear form arising in a hyperelastic model.
+\caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model.
 The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
@@
 
-\subsection{Performance of Forms}
+\subsection{Performance Results}
 \label{sec:perf-results-forms}
-% 52 test cases per form, so 52*4 = 208 test cases in total. In 7, there is a non-marginal slow down in cfO2.
 
-The first observation is that our optimality model does not imply minimum execution time.
-In particular, in the 0.06$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime.
+The main observation is that our transformation strategy does not always guarantee minimum execution time.
+In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime.
+The most significant of such test cases is the elastic model with $[q=4, nf=0]$.
 There are two reasons for this.
-The former is that low level optimization can have a significant impact.
-Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$.
-With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}.
-With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}.
-The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once).
-In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed-ups throughout the majority of test cases.
-The latter reason is simply inherent to our optimality model.
-As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop.
-The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.
+Firstly, low level optimization can have a significant impact on actual performance.
+For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization.
+Secondly, memory constraints are critical, particularly the transformation strategy adpopted when exceeding $T_H$.
+We will later thoroughly elaborate on all these aspects.
 
-The second aspect regards the ``triggering'' of pre-evaluation in \texttt{cfO2}.
-The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable.
+A second observation regards the use of pre-evaluation in \texttt{cfO2}.
+The fact that the \texttt{cfO2} and \texttt{cfO1} bars match suggests pre-evaluation was found unprofitable and, therefore, rejected.
 The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed.
@@
 
-The third point concerns the effect of the chosen discretizations on autovectorization.
-The discretizations employed ensure both multilinear loop nests and tabulated basis function sizes to be multiple of the machine vector length.
-Given the suitable choice of compilation options, this promotes autovectorization in the majority of code variants.
-The biggest exception is \texttt{quad}, in which, due to the presence of indirection arrays in the generated code, vectorization is rarely triggered.
-In \texttt{tens}, the loop nests are fully unrolled, so the standard loop vectorization is not possible; however, manual inspection of the compiled code suggests block vectorization (\cite{SLP-vect}) is often triggered.
-In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces tend to have the same structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could understand.
+A third point concerns autovectorization.
+The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length.
+This, combined with the chosen compilation flags, promotes autovectorization in the majority of code variants.
+An exception is \texttt{quad} due to the presence of indirection arrays in the generated code.
+In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; manual inspection of the compiled code suggests, however, that block vectorization (\cite{SLP-vect}) is often triggered.
+In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have similar structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could evince from compiler reports and manual inspection of assembly code.
 
-Also note some other minor things.
+Other less relevant aspects are
 \begin{itemize}
-\item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
-\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
-\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
-\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
+\item The lack of (or marginal) improvements in all code variants when $q=1$ is due to the cost of matrix insertion, which outweighs that of local assembly.
+\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading autovectorization for a relatively small save in floating point operations can be counter-productive (see also Section~\ref{sec:zeros}).
+\item If a bar or a subplot is missing, it means that the form compiler failed at generating code because of either exceeding the memory limit of the system or being unable to handle the form.
 \end{itemize}
@@
 \paragraph{Mass}
-From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} is the best variant among the ones tested, apart from a few points already discussed.
-This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints.
-It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$.
-The fact that \texttt{cfO2} is slower than \texttt{tens} at $[q=4, p=4, nf=1]$ is explained by the memory threshold that prevents pre-evaluation.
-By just setting $T_H = L3_{size}$, \texttt{cfO2} becomes 1.48$\times$ faster than \texttt{tens}.
-This is probably due to the hardware prefetching being way more effective than in all other forms: the number of allocated temporaries is relatively small, and so is the number of memory access traces that need be tracked.
-However, as we already explained, in a parallel context this gain might diminish, since the L3 cache is shared by the cores of the CPU.
+We start with the simplest of the bilinear forms that we are investigating, the mass matrix.
+From analysis of Figure~\ref{fig:mass}, we see that \texttt{cfO2} shows the highest speed-ups.
+It is worth noting how \texttt{auto} does not select the fastest implementation (out of \texttt{quad} and \texttt{tens}) in some of the test cases: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} tends to be preferable.
+On the other hand, one can infer that \texttt{cfO2} constantly makes correct decisions about pre-evaluation.
 
 \paragraph{Helmholtz}
-Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant.
-We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice.
-Take, for example, the case $[q=3, p=1, nf=2]$.
-The adoption of pre-evaluation makes a tremendous speed-up be achievable by \texttt{cfO2} over the competitors.
-On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.
-The effects of the working set threshold are in general positive.
-Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow-down of 2.16$\times$ at $[q=2, p=2]$.
+The main message of Figure~\ref{fig:helmholtz} is that \texttt{cfO2} still outperforms the other code generation systems.
+For $nf=0$, the adoption of pre-evaluation results in increasingly notable speed-ups over \texttt{cfO1} as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice.
+With $nf=1$, \texttt{auto} still picks \texttt{tens}, although this time the choice is sub-optimal: this can indirectly be inferred from \texttt{cfO2}, which rejects pre-evaluation for any $q$.
+For $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
+The sporadic slow-downs or marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.
+
+An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$.
+We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $nf=2, q=2$.
+Presumably, the effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is a shared memory.
 

### 2015-09-24 TOMS_Paper/assembly.tex
Reworked basically the whole perf eval section

@@
 
-We evaluate the speed-ups achieved by a wide variety of transformation systems over ``original code'' produced by the FEniCS Form Compiler (i.e., no optimizations applied).
+We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied).
 We analyze the following transformation systems
@@
 
+\subsection{Performance Results}
+\label{sec:perf-results-forms}
+
 We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
 The axes represent $q$, $nf$, and code transformation system.
-We show one subplot for each problem instance $[q, nf]$, with the code transformation system varying within each subplot.
+We show one subplot for each problem instance $\langle form, nf, q\rangle$, with the code transformation system varying within each subplot.
 The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
+We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.
+
+The rest of the section is structured as follows: we provide insights about the main message of the experimentation; we comment on the impact of autovectorization; we explain in detail, individually for each form, the performance results obtained.
 
@@
 
-\subsection{Performance Results}
-\label{sec:perf-results-forms}
-
+\paragraph{High level view}
 The main observation is that our transformation strategy does not always guarantee minimum execution time.
@@
 
-A second observation regards the use of pre-evaluation in \texttt{cfO2}.
-The fact that the \texttt{cfO2} and \texttt{cfO1} bars match suggests pre-evaluation was found unprofitable and, therefore, rejected.
-The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed.
-This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.
-
-A third point concerns autovectorization.
+\paragraph{Autovectorization}
 The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length.
@@
 
-Other less relevant aspects are
+\paragraph{Mass}
+We start with the simplest of the bilinear forms investigated, the mass matrix.
+Results are in Figure~\ref{fig:mass}.
+We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly.
+As $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups.
+It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} would tend to be preferable.
+On the other hand, \texttt{cfO2} always makes the optimal decision about whether applying pre-evaluation or not.
+
+\paragraph{Helmholtz}
+As happened with the mass matrix problem, when $q=1$ matrix insertion still hides the cost of local assembly.
+For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors.
+In particular, with
 \begin{itemize}
-\item The lack of (or marginal) improvements in all code variants when $q=1$ is due to the cost of matrix insertion, which outweighs that of local assembly.
-\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading autovectorization for a relatively small save in floating point operations can be counter-productive (see also Section~\ref{sec:zeros}).
-\item If a bar or a subplot is missing, it means that the form compiler failed at generating code because of either exceeding the memory limit of the system or being unable to handle the form.
+\item $nf=0$, the adoption of pre-evaluation by \texttt{cfO2} results in increasingly notable speed-ups over \texttt{cfO1}, as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice.
+\item $nf=1$, \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$.
+This can indirectly be inferred from the large gap between \texttt{cfO1/cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it avoids pre-evaluation.
+\item $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
 \end{itemize}
 
-\paragraph{Mass}
-We start with the simplest of the bilinear forms that we are investigating, the mass matrix.
-From analysis of Figure~\ref{fig:mass}, we see that \texttt{cfO2} shows the highest speed-ups.
-It is worth noting how \texttt{auto} does not select the fastest implementation (out of \texttt{quad} and \texttt{tens}) in some of the test cases: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} tends to be preferable.
-On the other hand, one can infer that \texttt{cfO2} constantly makes correct decisions about pre-evaluation.
-
-\paragraph{Helmholtz}
-The main message of Figure~\ref{fig:helmholtz} is that \texttt{cfO2} still outperforms the other code generation systems.
-For $nf=0$, the adoption of pre-evaluation results in increasingly notable speed-ups over \texttt{cfO1} as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice.
-With $nf=1$, \texttt{auto} still picks \texttt{tens}, although this time the choice is sub-optimal: this can indirectly be inferred from \texttt{cfO2}, which rejects pre-evaluation for any $q$.
-For $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
-The sporadic slow-downs or marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.
+The sporadic slow-downs or only marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.
 
 An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$.
-We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $nf=2, q=2$.
-Presumably, the effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is a shared memory.
+We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle nf=2, q=2\rangle$.
+The effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is shared by the cores.
 
 \paragraph{Elasticity}
-The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
-The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times.
-This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$.
-Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$.
-In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to the elimination of sharing and the avoidance of iteration over zero-valued blocks.
-The latter case occurs with $q=4$ and $nf \in [0, 1]$.
-Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow-downs characterized the cases of $nf=1$ for all values of $p$.
-Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only 40$\%$, while the cost function predicts a save in operation count of 5$\times$.
-This suggests that our model could be refined to handle the cases in which the gain in operation count is so large that non-extreme variations in working set size should be considered negligible.
+The results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
+The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should.
+To clarify this, consider the test case $\langle nf=0, q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
+\texttt{cfO2} finds pre-evaluation profitable -- that is, actually capable of reducing the operation count -- although it does not apply it because otherwise $T_H$ would be exceeded.
+However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that of \texttt{tens}.
+Our explanation is that despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this problem) that pre-evaluation would anyway be the optimal choice.
+This suggests that our model could be refined to handle the cases in which there is a significant gap between potential cache misses and save in flops.
+
+We also note that:
+\begin{itemize}
+\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to systematic sharing elimination and the use of symbolic execution to avoid iteration over the zero-valued regions in the basis function tables
+\item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
+\item \texttt{ufls} generally shows better runtime behaviour \texttt{quad} and \texttt{tens}.
+This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
+\end{itemize}
 
 \paragraph{Hyperelasticity}
-In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}.
-This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction.
+In the experiments on the hyperelastic model, which results are shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} manifests the largest gains out of all problem instances considered in this paper.
+This is a positive aspect: it means that our transformation algorithm scales with form complexity.
+
+% This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction.
 Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques.
@@
 \label{sec:conclusions}
-With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration.
+With this research we have made an important step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration.
 The results are extremely encouraging, suggesting our model applies to a variety of contexts.

### 2015-09-25 TOMS_Paper/assembly.tex
Rewritten bits here and there

@@
 
-In this section, we characterize optimality for finite element integration, as well as the transformation space that needs be explored to achieve it.
+In this section, we characterize optimality for finite element integration and the loop transformation space that we need to explore to achieve it.
 
@@
 \begin{Def}[Sharing]
-A loop $L_i$ presents sharing if it contains at least one statement $S$ in which two (sub-)expressions depending on $i$ are symbolically identical.
+A loop $L_i$ presents sharing if it contains at least one statement in which two (sub-)expressions depending on $i$ are symbolically identical.
 \end{Def}
@@
 The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$).
-It provides an intuition of how many symbols can ideally be factored along $L_i$
+It provides an intuition of how many symbols can ideally be factored along $L_i$.
 \end{Def}
@@
 
-Pre-evaluation opportunities could be exposed through an exploration of the expression tree transformation space.
+Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space.
 For arbitrary loop nests, this can be challenging.
@@
 \item Potential increase in operation count due to manipulation of the expression tree
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could prevent pre-evaluation; pre-evaluation might expose further sharing)
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could inhibit pre-evaluation; pre-evaluation might expose further sharing)
 \end{enumerate}
@@
 In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form.
-A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation, the transformation introduced in the previous section (and later further expanded).
+A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation (for which a deeper explanation will be provided).
 The optimality of the transformation algorithm is discussed.
@@
 For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
-This can sometimes break a ``local memory threshold''.
-We will report results of our experiments on the memory threshold in Section~\ref{sec:perf-results-forms}.
+This can sometimes break such a ``local memory threshold''.
+In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.
 
-Based on these considerations, we establish the following set of memory constraints
+Based on these considerations, we establish the following set of memory constraints:
 \begin{enumerate}
-\item The size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space. 
-\item The total amount of memory occupied by the temporaries created by code motion cannot exceed a threshold \texttt{$T_H$}
+\item the size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space;
+\item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
 \end{enumerate}
@@
 Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former.
-In this section, we tackle the second issue, while we deal with the second in Section~\ref{sec:op_count}.
+In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}.
 
 We start with introducing the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion about its optimality follow.
-
-%TODO add this somewhere: Further, we note that because of the memory constraints discussed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space.
 
@@
 These are mapped to the same temporary.
-The transformed expression, therefore, can be characterized by sharing.
-This is why the last two lines of the algorithm, which goal is to eliminate sharing following a procedure as described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all expressions, regardless of whether pre-evaluation was used.
-The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the input is still the usual loop nest in Figure~\ref{code:loopnest}.
+The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing.
+This is why the last two lines of the algorithm, which goal is to eliminate sharing following the procedure described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all sub-expressions, regardless of whether pre-evaluation was used.
+The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.
 
@@
 
-We elaborate on the optimality of our approach.
-First of all, we note that, because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space.
-The search for the optimal is then performed by the algorithm in a suitably constructed space.
-With the following Proposition, we finally aim to formalize the optimality claim.
+With the following Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach.
+First, however, note that because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space, with the search for the optimal by the algorithm being performed in a suitably constructed space.
 
@@
 
-The operation count for $\Lambda'$ is $\Lambda'_{ops} = \Lambda'_{ops_1} + \Lambda'_{ops_2} = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively.
+The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2}) = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively.
 $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.
@@
 If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} =  \gamma \geq I \cdot\delta$ holds.
+What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds.
 Because of assumption 2), the new pre-evaluations cannot expose further sharing.
@@
 \item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
-\item \texttt{ufls} generally shows better runtime behaviour \texttt{quad} and \texttt{tens}.
+\item \texttt{ufls} generally shows better runtime behaviour than \texttt{quad} and \texttt{tens}.
 This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
@@
 \paragraph{Hyperelasticity}
-In the experiments on the hyperelastic model, which results are shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} manifests the largest gains out of all problem instances considered in this paper.
+In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper.
 This is a positive aspect: it means that our transformation algorithm scales with form complexity.
-
-% This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction.
-Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques.
-Results are displayed in Figure~\ref{fig:hyperelasticity}.
-\texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions.
-\texttt{ufls} performs generally well and exhibits significant speed-ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.
+The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests several points.
+Firstly, the baseline is highly inefficient: with forms as complex as in this hyperelastic model, a trivial translation of integration routines into code should always be avoided since even one of the best general-purpose compilers available (Intel's on an Intel platform at maximum optimization level) is not capable of exploiting the structure inherent in the mathematical expressions generated.
+Secondly, the code motion strategy really makes a considerable impact.
+The sharing elimination performed by \texttt{cfO2} in each level of the loop nest ensures a critical reduction in operation count, which results is better execution times.
+In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the outer-product loop nest.
+Clearly, the operation count increases with $q$, and so do the speed-ups.
 
@@
 \label{sec:conclusions}
-With this research we have made an important step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration.
-The results are extremely encouraging, suggesting our model applies to a variety of contexts.
-We have discussed the conditions under which the model only leads to quasi-optimal loop nests.
-An open problem is understanding how to refine this model to include outer loops.
-This will probably require exploiting mathematical properties of differential operators.
+With this research we have set the foundation of optimal finite element integration.
+We have developed theory and implemented an automated system capable of applying it.
+The automated system, COFFEE, is integrated in Firedrake, a real-world framework for writing finite element methods.
+We believe the results are extremely positive.
+An open problem is understanding how to optimally handle non-linear loop nests.
 A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.

### 2015-09-25 TOMS_Paper/assembly.tex
Revisiting the cost function section

@@
 \begin{enumerate}
-\item the cost function $\theta$ is optimal; that is, it predicts correctly the profitability of pre-evaluation
-\item pre-evaluating different monomials does not result in identical tables
-\item there is no sharing of terms among monomials in $P$
+\item the cost function $\theta$ is optimal.
+We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, it returns two values $\theta_{w}$ and $\theta_{wo}$, representing the operation counts with and without pre-evaluation;
+\item pre-evaluating different monomials does not result in identical tables;
+\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
 \end{enumerate} 
@@
 
-The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2}) = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively.
+The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2})$ where $\Lambda'_{ops_1}$ and $\Lambda'_{ops_2}$ represent the operation counts for $P$ and $Z$, respectively.
 $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.
@@
 A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
-We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is true.
-If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds.
-Because of assumption 2), the new pre-evaluations cannot expose further sharing.
-Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.
+We note that if all monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is trivially true.
+If, on the other hand, there is no (or only partial) sharing, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$.
+What we have to show now is that even by pre-evaluating more monomials, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds.
+The intuition is that this should naturally be true since $\theta$ is optimal (assumption 1) and it predicted pre-evaluation to be a losing strategy for $Z_P$.
+However, to guarantee this we need to rely on assumption 2): if the new pre-evaluated tables were identical to those resulting from $P$, then we would have more sharing opportunities, which would generally break the optimality claim.
+Therefore, assumptions 1) and 2) are necessary for A) to be true.
 
-B) In absence of sharing, the statement is trivially true since we would have $\Lambda''_{ops_2} > \Lambda'_{ops_2}$, being the cost function optimal due to assumption 1).
-Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities.
-The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$.
-This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8).
-The rationale is that since we would have to pay anyway the presence of $z$ in the innermost loop, adding $p$ to $Z$ would not augment the operation count in our model, so we can safely avoid pre-evaluation.
-
-% assunzione 3 è quella che previene la somma di due mass matrix moltiplicate da F0 una e da F1 l'altra...ma non solo !
-chesso magari la trial function è derivata in un monomial e nell'altra no, ma la test function è sempre uguale...
+B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$.
+We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$ (both obvious due to assumption 1), we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
+The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
+An extreme example is the summation of mass matrices, in which each monomial is characterized by same test and trial functions, but also by a different pre-multiplying coefficient (i.e., known) function.
+Assumption 3) enables us abstracting from such a scenario.
 \end{proof}
@@
 \label{sec:op_count}
-It remains to tie one loose hand: the construction of the pre-evaluation cost function $\theta$.
-We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the sharing-free operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned.
-Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
+It remains to tie one loose hand: the definition of the cost function $\theta = [\theta_{w}, \theta_{wo}]$.
+Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.
 
@@
 This was explained in Proposition~\ref{prop:optimal-mln}.
-A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$.
-Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.
+A simple expression tree traversal suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$.
+Assuming $I$ to be the size of $L_i$, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.
 
@@
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
-\end{CenteredBox}\caption{Simplified loop nest for the local assembly of a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}
+\end{CenteredBox}\caption{Simplified loop nest for a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}
 
-One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix.
-The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$.
+One can think of this as the (simplified) loop nest originating from the local assembly of a pre-multiplied mass matrix.
+The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions (array $B$).
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities.
-By expanding the product we observe an increase in the number of $[L_j, L_k]$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient).
-Intuitively, $\iota$ captures this growth in non-hoistable operation.
+By expanding the products, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$).
+Intuitively, $\iota$ captures the increase in non-hoistable operations due to the transformations that expose pre-evaluation opportunities.
 
-With a single coefficient, as we just saw, $\iota$ directly descends from the cost of expansion.
+With a single coefficient, as we just saw, $\iota$ directly descends from the cost of product expansion.
 In general, however, predicting $\iota$ is less straightforward.
-For example, consider the case in which a monomial has multiple coefficients expressed over the same function space.
-The expansion would now lead to identical sub-expressions that, once pre-evaluated, would be mapped to the same temporary.
-The resulting loop nest would be characterized by sharing, as a result.
-Therefore, the actual operation count (i.e., once sharing is removed) would be smaller than that one could infer from analysing the expansion ``in isolation''.
-For a precise estimate of $\iota$, we then need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
+For example, consider the case in which a monomial has multiple coefficients over the same function space.
+Product expansion would now lead to identical sub-expressions that, as such, would be mapped to the same temporary once pre-evaluated.
+The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that expected by a naive analysis.
+For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
 

### 2015-09-25 TOMS_Paper/assembly.tex
Fixing a few things

@@
 \label{sec:codegen}
-The model described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimizer of finite element integration routines used in Firedrake.
-In this section, we describe the features of this code generation system.
+The transformation algorithm described in Section~\ref{sec:optimal-synthesis} has been implemented in COFFEE, the optimizer of finite element integration routines used in Firedrake.
+In this section, we describe several aspects related to code generation.
 
 \subsection{Automation through the COFFEE Language}
-As opposed to what happens in the FEniCS Form Compiler with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general.
-All optimizations are instead expressed as composition of parametric ``building-block'' operations.
+As opposed to what happens in the FEniCS Form Compiler, in which quadrature and tensor modes are completely separated code modules, COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''.
 This has several advantages.
 Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
-Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler.
-Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.
+Secondly, generality: any sort of codes characterized by (linear, but also non-linear) loop nests with sharing can be optimized (obviously, without guarantees on optimality).
+COFFEE exploits domain properties, although is not specifically tied to finite element in any way.
+Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
 
-A non-exhaustive list of such operations includes expansion, factorization, re-association, generalized code motion.
-These ``rewrite operators'' can be seen as the COFFEE language.
-They define parametric transformations: for example, one could ask to factorize constant rather linear terms, while hoisting could be driven by loop dependency.
+A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion.
+The rewrite operators can be seen as the COFFEE language.
+They are flexible: for example, one could ask to factorize constants instead of linear terms.
 Their implementation is based on manipulation of the abstract syntax tree representing the integration routine.
 
-\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
+\subsubsection{Heuristic Optimization of Non-linear Loops}
 \label{sec:heuristic-int-dep}
-As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions.
-These are expressions that should logically be placed within $L_i$.
-They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point.
-For some complex monomials and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, even outweigh that of the multilinear loop nest.
-In these cases, our optimality model becomes weaker, since its underlying assumption is that the bulk of the computation is carried out in innermost loops.
+As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for non-linear loops.
+Non-linear loops (expressions) are the natural result of non-linear coefficients in the form.
+The non-linear expressions logically depend on $L_i$.
+In complex monomials and/or with coarser discretizations, the operation count can experience a significant increase due to non-linear terms.
 
-Despite the fact that we are not characterizing optimality for this wider class of problems, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} to try to eliminate sharing, thus reducing the operation count.
+Despite the fact that we are not characterizing optimality for this wider class of problems, we still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} for (potentially) partial sharing elimination, thus reducing the operation count.
 This is straightforward in our code generation system by composing rewrite operators.

### 2015-09-25 TOMS_Paper/assembly.tex
Changes from Friday afternoon meeting

@@
 
-\begin{Def}[Outer-product loop nest]
-An outer-product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations of outer products.
+\begin{Def}[Outer product loop nest]
+An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs in every summand.
 \end{Def}
 
-As shown later, outer-product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code.
+As shown later, outer product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code.
 They arise naturally when translating bilinear or linear forms into code.
 Consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the outer-product loop nest $[L_j, L_k]$ over test and trial functions.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the outer product loop nest $[L_j, L_k]$ over test and trial functions.
 The expression \texttt{F} implements the operator.
@@
 \begin{itemize}
-\item a characterization of the transformation space for the class of loop nests considered
+\item a characterization of the space of legal loop transformations for the class of loop nests considered
 \item a cost model to select the optimal point in the transformation space
 \end{itemize}
-In this section, we construct the transformation space. We leave a few loose hands, which we progressively tie in Section~\ref{sec:optimal-synthesis}.
+In this section, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.
 
@@
 The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$).
-It provides an intuition of how many symbols can ideally be factored along $L_i$.
+It provides an indication of the maximal number of symbols which can be factorized to the level of $L_i$.
 \end{Def}
@@
 {\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
-\end{CenteredBox}\caption{A simple outer-product loop nest. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.}\end{figure}
+\end{CenteredBox}\caption{A simple outer product loop nest. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.}\end{figure}
 
-Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer-product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
+Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
 Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}.
-More in general, arbitrarily complex outer-product loop nests can systematically be reduced to optimal form.
-The constructive proof that we provide below focuses on outer-product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, as we elaborate later, a more general treatise would be required).
+More in general, arbitrarily complex outer product loop nests can systematically be reduced to optimal form.
+The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, as we elaborate later, a more general treatise would be required).
 
@@
 \label{prop:optimal-mln}
-Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}]$ is an outer-product loop nest with the body of $L_{i_{1}}$ being an independent basic block.
+Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}]$ is an outer product loop nest with the body of $L_{i_{1}}$ being an independent basic block.
 Then an optimal synthesis $\Lambda'$ can be determined by eliminating sharing from both loops, starting from the loop with smallest sharing factor.
@@
 We start with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_{i_{1}}$; this exposes the space of all possible factorizations.
-The resulting expressions are summations of outer products (otherwise, the assumption of $\Lambda$ being an outer-product loop nest would trivially be contradicted).
+Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products.
 
@@
 
-Note that if we were considering outer-product loop nests of arity $n > 2$, unless relaxing Definition~\ref{def:mln-optimality}, we would need a much more complex construction to guarantee optimality.
+Note that if we were considering outer product loop nests of arity $n > 2$ we would need a much more complex construction to guarantee optimality. 
 \end{proof}
@@
 \begin{itemize}
-\item as opposed to what happens with hoisting in outer-product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item as opposed to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
@@
 
-We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we conclude this section refining our optimality statement as follows.
+We expand on these points in Section~\ref{sec:optimal-synthesis}. In order to address point 2, we conclude this section by providing a more refined definition of optimality.
 
@@
 \label{sec:mem-const}
-We provide intuitions about the need for memory constraints.
+We will first consider two examples which motivate the need for memory constraints.
 Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}.
@@
 
-A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g.
+A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
@@
 \begin{enumerate}
-\item the size of a temporary due to code motion cannot be larger than that of the outer-product loop nest iteration space;
+\item the size of a temporary due to code motion cannot be larger than that of the outer product loop nest iteration space;
 \item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
@@
 
-We reinforce the idea that eliminating sharing from the outer-product loop nest does not suffice.
-As suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
+We reinforce the idea that eliminating sharing from the outer product loop nest does not suffice.
+As presaged in Section~\ref{sec:optimal-impl}, we consider whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
 
@@
 1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}.
-2) This is inherently a pathological case, from which we abstract.
+2) This is inherently a pathological case, which we neglect.
 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
-We also abstract from this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.
+We also neglect this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.
 
 The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2})$ where $\Lambda'_{ops_1}$ and $\Lambda'_{ops_2}$ represent the operation counts for $P$ and $Z$, respectively.
-$\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.
+$L$, $M$, $N$, and $O$ are the trip counts of the corresponding loops.
 
@@
 The sharing elimination performed by \texttt{cfO2} in each level of the loop nest ensures a critical reduction in operation count, which results is better execution times.
-In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the outer-product loop nest.
+In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the outer product loop nest.
 Clearly, the operation count increases with $q$, and so do the speed-ups.

### 2015-09-28 TOMS_Paper/assembly.tex
Rewrite the optimality proof

@@
 \usepackage{listings}
-\lstset{language=C, breaklines=true, mathescape}
+\lstset{language=C, breaklines=true, mathescape,
+			 morekeywords={and, each}}
 
@@
 In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}.
-
-We start with introducing the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion about its optimality follow.
+For this, we start with explaining the transformation algorithm, which pseudo-code is provided in Figure~\ref{code:intuition}.
 
@@
 dissect the input expression into monomials
-for each monomial M:
-  $\theta_{w}$ = estimate operation count with pre-evaluation
-  $\theta_{wo}$ = estimate operation count without pre-evaluation
-  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
-    mark M as candidate for pre-evaluation
-for each monomial M:
-  if M does not share terms with M$'$, an unmarked monomial:
-    extract M into a separate loop nest
-    apply pre-evaluation to M
+for each monomial $m$:
+  if $\theta(m)$ > 0 and memory constraints satisfied:
+    mark $m$ as candidate for pre-evaluation
+for each monomial $m$:
+  if $m$ does not share terms with $m'$, an unmarked monomial:
+    extract $m$ into a separate loop nest
+    apply pre-evaluation to $m$
 for each expression:
@@
 \end{lstlisting}
-\caption{Intuition of the transformation algorithm}
+\caption{High level view of the transformation algorithm}
 \label{code:intuition}
@@
 Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''.
-If we estimate that, for a given monomial $M$, pre-evaluation will decrease the operation count, then: the sub-expression implementing $M$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
+We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form.
+$\theta(m)$ represents the change in operation count if pre-evaluation were applied to monomial $m$: a negative value indicates an increase, whereas a positive value indicates a gain.
+If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
 The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.
@@
 
-With the following Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach.
-First, however, note that because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space, with the search for the optimal by the algorithm being performed in a suitably constructed space.
+Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched.
+We have defined them as high level operations on top of traditional expression-restructuring transformations such as code motion, re-association, and factorization.
+Other classes of loop transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.
+
+Finally, with Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach.
 
@@
 \begin{enumerate}
-\item the cost function $\theta$ is optimal.
-We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, it returns two values $\theta_{w}$ and $\theta_{wo}$, representing the operation counts with and without pre-evaluation;
 \item pre-evaluating different monomials does not result in identical tables;
@@
 We first comment on the assumptions.
-1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}.
-2) This is inherently a pathological case, which we neglect.
-3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms.
-We also neglect this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.
+First, we notice that neither of them is a sufficient condition for non-optimality.
+In addition: 1) is inherently a pathological case, which we neglect; 2) might actually occur in forms with several monomials, although, in practice, this was never the case in none of the problems that we have investigated (which are more than those studied in this article).
+Therefore, we also neglect this scenario, which would otherwise require a ``global analysis'' of $M$.
 
@@
 
-A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$).
-We note that if all monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is trivially true.
-If, on the other hand, there is no (or only partial) sharing, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$.
-What we have to show now is that even by pre-evaluating more monomials, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds.
-The intuition is that this should naturally be true since $\theta$ is optimal (assumption 1) and it predicted pre-evaluation to be a losing strategy for $Z_P$.
-However, to guarantee this we need to rely on assumption 2): if the new pre-evaluated tables were identical to those resulting from $P$, then we would have more sharing opportunities, which would generally break the optimality claim.
-Therefore, assumptions 1) and 2) are necessary for A) to be true.
+A) By pre-evaluating the monomials in $Z_P$, we clearly have $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$, where $\delta > 0$ is a parameter to indicate the reduction in operation count exhibited by the non pre-evaluated monomials.
+To prove A), we then need to convince ourselves that $\Lambda''_{ops_1} -  I \cdot\delta \geq \Lambda'_{ops_1}$ holds.
+The only (corner) situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$.
+If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim.
+Assumption 1) prevents this situation, so the statement is true.
 
 B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$.
-We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$ (both obvious due to assumption 1), we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
+We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
 The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
-An extreme example is the summation of mass matrices, in which each monomial is characterized by same test and trial functions, but also by a different pre-multiplying coefficient (i.e., known) function.
-Assumption 3) enables us abstracting from such a scenario.
+An extreme example is a summation of mass matrices, in which each monomial is characterized by same test and trial functions, and by a different pre-multiplying coefficient (i.e., known) function.
+The previously discussed assumption 2) allows us to neglect this possibility, which then makes B) true.
 \end{proof}
@@
 \label{sec:op_count}
-It remains to tie one loose hand: the definition of the cost function $\theta = [\theta_{w}, \theta_{wo}]$.
+It remains to tie up one loose end: the definition of the cost function $\theta$.
 Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.
+
+We define $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$, where $\theta_{wo}$ and $\theta_{w}$ represent the operation counts for a monomial $m$ without and with pre-evaluation, respectively.
+In the following, we explain how to derive these two values.
 
@@
 This was explained in Proposition~\ref{prop:optimal-mln}.
-A simple expression tree traversal suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$.
+A simple expression tree traversal suffices to obtain the cost of a  sharing-free outer product loop nest, namely $\Lambda_{ops}^{ns}$.
 Assuming $I$ to be the size of $L_i$, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.
 
-For $\theta_{w}$, things are more complicated.
+Determining $\theta_{w}$ is more complicated.
 We first need to estimate the \textit{increase factor}, $\iota$, to account for the presence of (derivatives of) coefficients.
@@
 Product expansion would now lead to identical sub-expressions that, as such, would be mapped to the same temporary once pre-evaluated.
-The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that expected by a naive analysis.
+The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that found with a more naive analysis.
 For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).

### 2015-09-28 TOMS_Paper/assembly.tex
Trying to fix code gen section

@@
 \subsection{Automation through the COFFEE Language}
-As opposed to what happens in the FEniCS Form Compiler, in which quadrature and tensor modes are completely separated code modules, COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''.
+COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''.
 This has several advantages.
@@
 Secondly, generality: any sort of codes characterized by (linear, but also non-linear) loop nests with sharing can be optimized (obviously, without guarantees on optimality).
-COFFEE exploits domain properties, although is not specifically tied to finite element in any way.
+In other words, COFFEE only exploits domain properties, and it is not specifically tied to finite element in any way.
 Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
 
+The rewrite operators, which implementation is based on abstract syntax tree manipulation, compose the COFFEE language.
 A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion.
-The rewrite operators can be seen as the COFFEE language.
-They are flexible: for example, one could ask to factorize constants instead of linear terms.
-Their implementation is based on manipulation of the abstract syntax tree representing the integration routine.
+We have primarily used these operators to implement sharing elimination and pre-evaluation.
+Further, we have exploited them for heuristic optimization of the integration loop when this is non-linear (e.g., whenever the form has non-linear coefficients).
+In such a case, expansion, factorization, and code motion are applied greedily, selecting the term with highest sharing factor at each iteration, based on the observation that this should minimize sharing and promote hoisting.
 
-\subsubsection{Heuristic Optimization of Non-linear Loops}
-\label{sec:heuristic-int-dep}
-As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for non-linear loops.
-Non-linear loops (expressions) are the natural result of non-linear coefficients in the form.
-The non-linear expressions logically depend on $L_i$.
-In complex monomials and/or with coarser discretizations, the operation count can experience a significant increase due to non-linear terms.
+\subsection{Independence from Form Compilers}
+\label{sec:zeros}
+COFFEE expects generic abstract syntax trees, so clearly any form compiler can be adapted to produce this kind of input.
+This is what we have done with the FEniCS Form Compiler, suitably modified to produce abstract syntax trees rather than strings.
+However, form compilers that translate quadrature-based expressions into code could decide, for simplicity, to implement non-scalar function spaces in trivial yet sub-optimal ways.
+For example ...by producing a trivial loop nest that iterates over basis function tables filled with blocks of zeros
 
-Despite the fact that we are not characterizing optimality for this wider class of problems, we still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} for (potentially) partial sharing elimination, thus reducing the operation count.
-This is straightforward in our code generation system by composing rewrite operators.
-Our strategy is as follows.
+Another viewpoint is: COFFEE takes care of the optimization process, ...
 
-COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions.
-This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to code without performing any sort of analysis.
-Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent in $L_i$.
-We then synthesize an optimal loop nest as described in the previous sections.
-This may in turn expose a set of $L_i$-dependent expressions.
-For each of this expressions, we try to remove sharing by greedily applying factorization and code motion.
-In the COFFEE language, this process is expressed by simply composing five rewrite operators.
-
-\subsection{Low-level Optimization}
-\label{sec:code-spec}
-We comment on a set of low-level optimizations.
-These are essential to 1) achieve machine peak performance (Sections~\ref{sec:opt-review} and~\ref{sec:vect-prom}) and 2) make COFFEE independent of the high-level form compiler (Section~\ref{sec:zeros}).
-As we will see, there are interplays among different transformations.
-For completeness, we present all of the transformations available in the compiler, although we will only use a subset of them for a fair performance evaluation.
-
-\subsubsection{Review of Existing Optimizations}
-\label{sec:opt-review}
-We start with briefly reviewing the low level optimizations presented in~\citeN{Luporini}.
-
-\paragraph{Padding and data alignment} All of the arrays involved in the evaluation of the local element matrix or vector are padded to a multiple of the vector register length.
-This is a simple yet powerful transformation that maximizes the effectiveness of vectorization.
-Padding, and then loop bounds rounding, enable data alignment and avoid the introduction of scalar remainder loops.
-
-\paragraph{Vector-register Tiling} Blocking (or tiling) at the level of vector registers improves data locality beyond traditional unroll-and-jam transformations.
-This blocking strategy consists of evaluating outer products by using just two vector register and without ever spilling to cache.
-
-\paragraph{Expression Splitting} When the number of basis functions arrays (or, equivalently, temporaries introduced by code motion) and constants is large, the chances of spilling to cache are high in architectures with a few logical registers (e.g.
-16/32).
-By exploiting sum’s associativity, an expression can be fissioned so that the resulting sub-expressions can be computed in separate loop nests.
-This reduces the register pressure.
-
-\subsubsection{Vector-promotion of Integration-dependent Expressions}
-\label{sec:vect-prom}
-Integration-dependent expressions are inherently executed as scalar code because vectorization (unless employing special hand-written schemes) occurs along a single loop, typically the innermost.
-For the same reasons discussed in Section~\ref{sec:heuristic-int-dep}, we also want to vectorize along $L_i$.
-One way to achieve this is vector-promotion.
-This requires creating a ``clone'' of $L_i$ in the preheader of the loop nest, in which vector temporaries are evaluated in what is now an innermost loop.
-
-\subsubsection{Handling Sparse Tables}
-\label{sec:zeros}
 Consider a set of tabulated basis functions with quadrature points along rows and functions along columns.

### 2015-09-29 TOMS_Paper/assembly.tex
Trying to rewrite the part on zero block elimination

@@
 \label{sec:zeros}
-COFFEE expects generic abstract syntax trees, so clearly any form compiler can be adapted to produce this kind of input.
-This is what we have done with the FEniCS Form Compiler, suitably modified to produce abstract syntax trees rather than strings.
-However, form compilers that translate quadrature-based expressions into code could decide, for simplicity, to implement non-scalar function spaces in trivial yet sub-optimal ways.
-For example ...by producing a trivial loop nest that iterates over basis function tables filled with blocks of zeros
+One of the key pillars of COFFEE is independence from form compilers.
+Being capable of handling generic abstract syntax trees, COFFEE can be leveraged by any suitably adapted form compiler.
+For example, in Firedrake a modified version of the FEniCS Form Compiler producing abstract syntax trees instead of strings is used.
+COFFEE itself provides an interface for building such abstract syntax trees.
+In particular, COFFEE aims to decouple the form manipulation (i.e., the ``math level'') from code optimization (flops executed, low level optimization such as vectorization).
+Another viewpoint is: developers of form compilers should not worry about expression optimization in any way, since this is handled at a lower level.
 
-Another viewpoint is: COFFEE takes care of the optimization process, ...
+To achieve this goal, COFFEE needs a strategy to minimize the execution of potentially ``useless'' flops due to simplistic implementations of non scalar-valued function spaces.
+For example, the most trivial quadrature mode in the FEniCS Form Compiler implements vector-valued function spaces by populating tabulated basis functions with blocks of zero-valued columns.
+For the local element matrix/vector evaluation, these blocks are iterated over in a single, rectangular iteration space.
+This preserves the semantics of the computation while keeping the implementation complexity extremely low, although it also introduces unnecessary operations (e.g., zero-valued summands due to multiplications by zero).
+COFFEE achieves full independence from form compilers (and so relives the implementation burden) by providing a transformation to avoid the execution of such useless flops.
+A challenge for this transformation is restructuring the iteration spaces while preserving the effectiveness of low level optimizations, especially (auto-)vectorization.
 
@@
 For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}.
-Unless using a smart form compiler (which we want to avoid), there are circumstances in which the tables are sparse.
-Zero-valued columns arise when taking derivatives on a reference element or when employing vector-valued elements.
-Zero-valued rows can result from using non-standard functions spaces, like Raviart-Thomas.
-Zero-valued blocks can appear in pre-evaluated temporaries.
-Our objective is a transformation that avoids useless iteration over zeros while preserving the effectiveness of the other low-level optimizations, especially vectorization.
-
 In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g.
 \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed.
-Our approach, which will be compared to this pioneering work, aims to free the generated code from such indirection arrays.
+Our approach aims to avoid such indirection arrays in the generated code.
 This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.
@@
 For example, multiplication requires computing the set intersection of the zero-valued slices (for each loop dimension), whereas addition requires computing the set union.
-
-% say that the reasons to have zeros is to keep the design of the form compiler simple, which just ``puts zeros'' and ``outline the assembly expression''.
-% rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
-% tables shrunk and merged
 

### 2015-09-30 TOMS_Paper/assembly.tex
Rewrite the contribution list

@@
 \begin{itemize}
-\item We characterize loop nest optimality and we instantiate this concept to finite element integration.
-As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
-\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into optimal loop nests.
-\item We comment on the cases in which such model does not lead to optimal loop nests.
-We will see that this might occur with particular forms that make extensive use of tensor algebra.
-We show, however, that our model still is capable of producing quasi-optimal loop nests.
-\item We integrate the model with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
+\item We define optimality for finite element integration. We study the transformation space for loops and expressions within which the optimal is sought.
+\item We provide an algorithm to select optimal points in the transformation space.
+The algorithm uses a cost model to: 1) understand whether a transformation reduces or increases the operation count; 2) choose between different (non-composable) transformations.
+\item We state for which class of problems optimality can be reached, and for which class heuristics are instead needed to minimize the operation count.
+\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
 The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
@@
 
-In addition, in order to place COFFEE on the same level of other code generation systems from the viewpoint of low-level optimization (which is essential for a fair performance comparison)
+In addition, in order to place COFFEE on the same level of other code generation systems from the viewpoint of low level optimization (which is essential for a fair performance comparison)
 
@@
 \item We introduce an engine based on symbolic execution that allows skipping irrelevant floating point operations (e.g., those involving zero-valued quantities).
-We elaborate on the performance impact of this optimization, making a clear distinction between flop optimality and efficient code.
 \end{itemize}

### 2015-09-30 TOMS_Paper/assembly.tex
Fix a few things

@@
 A theoretical analysis and extensive experimentation prove the effectiveness of our approach, showing systematic performance improvements over a number of alternative code generation systems.
-The effect of low-level optimization is also discussed.
 \end{abstract}
@@
 \begin{itemize}
-\item We define optimality for finite element integration. We study the transformation space for loops and expressions within which the optimal is sought.
+\item We define optimality for finite element integration and we build the transformation space within which the optimal is sought.
 \item We provide an algorithm to select optimal points in the transformation space.
 The algorithm uses a cost model to: 1) understand whether a transformation reduces or increases the operation count; 2) choose between different (non-composable) transformations.
-\item We state for which class of problems optimality can be reached, and for which class heuristics are instead needed to minimize the operation count.
+\item We state for which class of problems optimality can be reached, and for which class heuristics are instead needed to minimize (reduce) the operation count.
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.

### 2015-10-15 TOMS_Paper/assembly.tex
Several fixes, from the abstract up to the second proof

@@
 We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms.
-Optimality is defined in terms of floating point operations required to execute a loop nest.
-The generation of optimal loop nests is driven by a model that exploits mathematical properties of the domain of interest.
-A theoretical analysis and extensive experimentation prove the effectiveness of our approach, showing systematic performance improvements over a number of alternative code generation systems.
+Optimality is defined in terms of floating point operations given a memory bound.
+We provide an approach to explore the transformation space of loops and expressions typical of integration routines and discuss the conditions for optimality.
+A theoretical analysis and extensive experimentation, which shows systematic performance improvements over a number of alternative code generation systems, validate the approach.
 \end{abstract}
@@
 \begin{Def}[Outer product loop nest]
-An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs in every summand.
+An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs once in every summand.
 \end{Def}
@@
 \begin{itemize}
-\item a characterization of the space of legal loop transformations for the class of loop nests considered
+\item a characterization of the space of legal transformations for the class of loop nests considered impacting the operation count
 \item a cost model to select the optimal point in the transformation space
@@
 \begin{figure}[h]\begin{CenteredBox}
-{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
+{\subfigcapskip = 7pt \subfigure[With sharing.
+Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
 ~~~~~~~~~~~~~~
 {\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
-\end{CenteredBox}\caption{A simple outer product loop nest. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.}\end{figure}
+\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\end{figure}
 
@@
 More in general, arbitrarily complex outer product loop nests can systematically be reduced to optimal form.
-The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$ (for $n < 2$ the proof is trivial; for $n > 2$, as we elaborate later, a more general treatise would be required).
+The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$.
+Note that for $n < 2$ the proof is trivial, whereas for $n > 2$ a more general treatise would be required; fortunately, this case is never of interest in finite element integration.
 
@@
 \end{enumerate} 
-
-Note that if we were considering outer product loop nests of arity $n > 2$ we would need a much more complex construction to guarantee optimality. 
 \end{proof}
 
-We now observe that, for the larger class of finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
+We observe, however, that for finite element integration loop nests the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
 Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
 We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$?
-The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} (input) and~\ref{code:loopnest_nored} (output); Figure~\ref{code:loopnest_red} can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.
+The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.
 
@@
 {\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
-\end{CenteredBox}\caption{Exposition (through factorization) and pre-evaluation of a reduction.}\end{figure}
+\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\end{figure}
 
 Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space.
-For arbitrary loop nests, this can be challenging.
-Further, the following issues arise when considering pre-evaluation opportunities:
+This would be challenging if we were to deal with arbitrary loop nests and expressions, while we will see that a simpler algorithm can be adopted for the class arising in finite element integration.
+In any case, however, the following issues should be considered:
 \begin{itemize}
-\item as opposed to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item as opposed to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
@@
 \item Potential increase in operation count due to manipulation of the expression tree
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities (sharing elimination could inhibit pre-evaluation; pre-evaluation might expose further sharing)
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
 \end{enumerate}
@@
 \label{sec:mem-const}
-We will first consider two examples which motivate the need for memory constraints.
+Two observations motivate the need for memory constraints.
 Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}.
@@
 
-A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g.
+The second observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N\cdot O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
+For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
 This can sometimes break such a ``local memory threshold''.
@@
 
-Based on these considerations, we establish the following set of memory constraints:
+Based on these considerations, we establish the following memory constraints:
 \begin{enumerate}
@@
 Consequently, $L_i$ disappears from $\Lambda$.
-This transformation is actually what we were referring to when we introduced pre-evaluation in Section~\ref{sec:optimal-impl}.
+This transformation is actually what we were referring to when we more informally introduced pre-evaluation in Section~\ref{sec:optimal-impl}.
 
@@
   if $\theta(m)$ > 0 and memory constraints satisfied:
-    mark $m$ as candidate for pre-evaluation
-for each monomial $m$:
-  if $m$ does not share terms with $m'$, an unmarked monomial:
-    extract $m$ into a separate loop nest
-    apply pre-evaluation to $m$
+    extract into a separate loop nest
+    apply pre-evaluation
 for each expression:
@@
 
-Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''.
+Our choice is to evaluate the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''.
 We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form.
@@
 
-Because of symmetries in basis functions, pre-evaluation could produce identical tables.
-These are mapped to the same temporary.
+Because of symmetries in basis functions, pre-evaluation could produce identical tables, which would be mapped to the same temporary.
 The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing.
@@
 Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched.
-We have defined them as high level operations on top of traditional expression-restructuring transformations such as code motion, re-association, and factorization.
-Other classes of loop transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.
+We consider them high level operators defined on top of expression-restructuring transformations such as code motion, re-association, and factorization.
+Other classes of loop or expression transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.
 
@@
 We first comment on the assumptions.
-First, we notice that neither of them is a sufficient condition for non-optimality.
-In addition: 1) is inherently a pathological case, which we neglect; 2) might actually occur in forms with several monomials, although, in practice, this was never the case in none of the problems that we have investigated (which are more than those studied in this article).
-Therefore, we also neglect this scenario, which would otherwise require a ``global analysis'' of $M$.
-
-The operation count for $\Lambda'$ is $\Lambda'_{ops} = L N O (\Lambda'_{ops_1} + I \Lambda'_{ops_2})$ where $\Lambda'_{ops_1}$ and $\Lambda'_{ops_2}$ represent the operation counts for $P$ and $Z$, respectively.
-$L$, $M$, $N$, and $O$ are the trip counts of the corresponding loops.
+1) Inherently a pathological case, so we simply neglect it.
+2) Might occur in forms with several monomials, although this was never the case in the problems that we have so far investigated (which are obviously more than those presented in Section~\ref{sec:perf-results}).
+These assumptions allow us to avoid a ``global analysis'' of $M$.
 
@@
 
-A) By pre-evaluating the monomials in $Z_P$, we clearly have $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$, where $\delta > 0$ is a parameter to indicate the reduction in operation count exhibited by the non pre-evaluated monomials.
-To prove A), we then need to convince ourselves that $\Lambda''_{ops_1} -  I \cdot\delta \geq \Lambda'_{ops_1}$ holds.
-The only (corner) situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$.
+A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$.
 If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim.
@@
 
-B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$.
-We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
-The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
-An extreme example is a summation of mass matrices, in which each monomial is characterized by same test and trial functions, and by a different pre-multiplying coefficient (i.e., known) function.
-The previously discussed assumption 2) allows us to neglect this possibility, which then makes B) true.
+B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term.
+We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
+That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
+Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions.
+The previously discussed assumption 2) allows us to neglect this possibility.
 \end{proof}
@@
 \label{sec:op_count}
-It remains to tie up one loose end: the definition of the cost function $\theta$.
+It remains to tie up one loose end: the construction of the cost function $\theta$.
 Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.
 
-We define $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$, where $\theta_{wo}$ and $\theta_{w}$ represent the operation counts for a monomial $m$ without and with pre-evaluation, respectively.
+According to what established in the previous section, we have $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$; we recall that $\theta_{wo}$ and $\theta_{w}$ represent the operation counts without and with pre-evaluation.
 In the following, we explain how to derive these two values.

### 2015-10-17 TOMS_Paper/assembly.tex
Started integrating Paul's advises

@@
 
-The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
-This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years.
-In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language.
+The need for rapid implementation of high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
+This has been proven successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects.
+In these frameworks, the weak variational form of a problem is expressed at high level by means of a domain-specific language.
 The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators.
-By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed.
-The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.
+By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation approximated solution, are computed.
+The code for assembly operators must be carefully optimized: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.
 
-As demonstrated by the considerable body of research on the topic, automating the generation of such high performance implementations poses several challenges.
+As demonstrated by the substantial body of research on the topic, automating the generation of such high performance implementations poses several challenges.
 This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
-General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary).
-Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and loop optimization for cache locality, so they are not particularly helpful in our context.
+General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}).
+Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops.
 The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems.
@@
 In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count.
-In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed.
-It is also worth mentioning an on-going effort to produce a novel form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.
+In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed.
+It is also worth mentioning an on-going effort to produce a new form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.
+The performance evaluation in Section~\ref{sec:perf-results} includes most of these optimization systems.
 
@@
 In this paper, we formulate an approach to solve this problem.
-Summarizing, our contributions are as follows
+Summarizing, our contributions are as follows:
 
 \begin{itemize}
-\item We define optimality for finite element integration and we build the transformation space within which the optimal is sought.
+\item We define optimality for finite element integration as number of flops given a memory bound; we build the transformation space within which the optimal is sought.
 \item We provide an algorithm to select optimal points in the transformation space.
@@
 
-In addition, in order to place COFFEE on the same level of other code generation systems from the viewpoint of low level optimization (which is essential for a fair performance comparison)
+In addition, in order to place COFFEE on the same level as other code generation systems from the viewpoint of low level optimization (which is essential for a fair performance comparison)
 
@@
 \begin{split}
-Find\ u\ \in U\ such\ that \\
+\text{Find}\ u \in U\ \text{such that} \\
 a(u, v) = L(v), \forall v \in V
@@
 \begin{equation}
-A\textbf{u} = b
+Au = b
 \end{equation}
@@
 Local assembly is the subject of this article.
-It consists of computing the contributions of a single element in the discretized domain to the equation solution.
+It consists of computing the contributions of a single element in the discretized domain to the equation approximated solution.
 In global assembly, such local contributions are ``coupled'' by suitably inserting them into $A$ and $b$.
@@
 A generalization of this transformation has been proposed in~\cite{Kirby:TC}.
-By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables.
+Because of only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables.
 The evaluation of the local tensor can then be abstracted as
@@
 We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}.
-For the moment, we just recall that one of the goals of this research is to produce an intelligent system that goes beyond the dichotomy between quadrature and tensor modes.
+For the moment, we just recall that one of the goals of this research is to produce a system that goes beyond the dichotomy between quadrature and tensor modes.
 We will reason in terms of loop nests, code motion, and code pre-evaluation, searching the entire implementation space for an optimal synthesis.
@@
 \subsection{Loop Nests and Optimality}
-In order to make the document self-contained, we start with reviewing basic compiler terminology.
+In order to make the article self-contained, we start with reviewing basic compiler terminology.
 
@@
 Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations that it performs to evaluate the result is minimal.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
 \end{Def}
@@
 Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}.
-More in general, arbitrarily complex outer product loop nests can systematically be reduced to optimal form.
+More generally, arbitrarily complex outer product loop nests can systematically be reduced to optimal form.
 The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$.
-Note that for $n < 2$ the proof is trivial, whereas for $n > 2$ a more general treatise would be required; fortunately, this case is never of interest in finite element integration.
+Note that for $n < 2$ the proof is trivial, whereas for $n > 2$ a more general treatment would be required; fortunately, this case is never of interest in finite element integration.
 
@@
 Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$.
-It is fairly obvious to see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal.
+We see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal.
 Consequently, the hypothesis is generally contradicted.
@@
 Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and a set of memory constraints is satisfied.
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations that it performs to evaluate the result is minimal.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
 \end{Def}

### 2015-10-17 TOMS_Paper/assembly.tex
Incorporated most of Paul's suggestions

@@
 \newtheorem{Prop}{Proposition}
+\newtheorem{Algo}{Algorithm}
+\newtheorem{Const}{Constraint}
 \newtheorem*{Rem*}{Remark}
@@
 \subsection{Tensor Contraction Mode}
+\label{sec:tc}
 Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
@@
 Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}.
-More generally, arbitrarily complex outer product loop nests can systematically be reduced to optimal form.
-The constructive proof that we provide below focuses on outer product loop nests of arity $n=2$.
-Note that for $n < 2$ the proof is trivial, whereas for $n > 2$ a more general treatment would be required; fortunately, this case is never of interest in finite element integration.
+More generally, arbitrarily complex outer product loop nests can systematically be reduced to optimal form using an algorithm that exploits the properties of linearity and perfectness.
+The algorithm and the proof that we provide below focus on outer product loop nests of arity $n=2$.
+For $n < 2$ the proof would be trivial, whereas for $n > 2$ a more general treatment would be required; fortunately, this case is never of interest in finite element integration.
 
-\begin{Prop}
-\label{prop:optimal-mln}
-Assume $\Lambda = [L_{i_{0}}, L_{i_{1}}]$ is an outer product loop nest with the body of $L_{i_{1}}$ being an independent basic block.
-Then an optimal synthesis $\Lambda'$ can be determined by eliminating sharing from both loops, starting from the loop with smallest sharing factor.
-\end{Prop}
-\begin{proof}
-The demonstration is by construction and exploits linearity.
-We start with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_{i_{1}}$; this exposes the space of all possible factorizations.
+\begin{Algo}[Optimal code motion for outer product loop nests]
+\label{algo:outer-product-optimal}
+\normalfont
+Consider an outer product loop nest $\Lambda = [L_i, L_j]$, with the body of $L_j$ being an independent basic block.
+The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_j$; this exposes the space of all possible factorizations.
 Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products.
-
 We traverse the expression tree and determine the sharing factor of the two linear loops.
-We take the symbols depending on the loop with smallest sharing factor, $L_i$, and we incrementally factorize them.
+We take the symbols depending on the loop with smallest sharing factor, say $L_i$, and we incrementally factorize them.
 Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
@@
 The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about generalized code motion).
+\end{Algo}
 
-We now have to demonstrate that the chosen factorization strategy, based on sharing factors, leads to optimality.
-This reduces to prove that the two following facts are true (we do it by contradiction).
+\begin{Prop}
+\label{prop:optimal-mln}
+Let $\Lambda = [L_i, L_j]$ be an outer product loop nest with the body of $L_j$ being an independent basic block.
+Then Algorithm~\ref{algo:outer-product-optimal} leads to an optimal synthesis $\Lambda'$.
+\end{Prop}
+\begin{proof}
+Assume, without loss of generality, that $L_i$ is the loop with smallest sharing factor. We reduce the proof to showing that the two following facts are true (we do it by contradiction).
 \begin{enumerate}
@@
 \begin{itemize}
-\item as opposed to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
-This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
-\item the transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later).
-This could overwhelm the gain inherent in pre-evaluation.
+\item In contrast to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
+\item The transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later).
+This could outweigh the gain due to pre-evaluation.
 \end{itemize}
@@
 Two observations motivate the need for memory constraints.
-Our point of departure is again the bilinear form implementation in Figure~\ref{code:loopnest}.
+Our starting point is again the bilinear form implementation in Figure~\ref{code:loopnest}.
 Analogous considerations apply to forms of arbitrary arity.
@@
 The working set size could then increase by a factor $L$.
-The gain in number of operations executed could therefore be overwhelmed, from a runtime viewpoint, by a much larger memory pressure.
+The gain in number of operations executed could therefore be outweighed, from a runtime viewpoint, by a much larger memory pressure.
 
@@
 
-To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}.
+To partly answer this question, we make use of a result -- the foundation of tensor contraction mode (Section~\ref{sec:tc}) -- from~\citeN{Kirby:TC}.
 Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}).
@@
 In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}.
-For this, we start with explaining the transformation algorithm, which pseudo-code is provided in Figure~\ref{code:intuition}.
+For this, we start with explaining the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.
 

### 2015-10-17 TOMS_Paper/assembly.tex
Started incorporating David's suggestions

@@
 
-In this section, we characterize optimality for finite element integration and the loop transformation space that we need to explore to achieve it.
+In this section, we characterize optimality for finite element integration and the transformation space that we need to explore to achieve it.
 
@@
 \begin{Def}[Linear loop]
-A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if 
+A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is linear if in its body
 \begin{enumerate}
-\item $i$ appears in the body of $L$ only as an array index, and
+\item $i$ appears only as an array index, and
 \item whenever an array $a$ is indexed by $i$ ($a[i]$), all expressions in which this appears are affine in $a$.
@@
 As shown later, outer product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code.
-They arise naturally when translating bilinear or linear forms into code.
-Consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the outer product loop nest $[L_j, L_k]$ over test and trial functions.
+We will see that they arise naturally when translating bilinear or linear forms into code.
+
+Two other classes of loops need be characterized.
+
+\begin{Def}[Reduction loop]
+A loop $L_i$ is said to be a reduction loop if in its body
+\begin{enumerate}
+\item $i$ appears only as an array index, and
+\item for each augmented assignment statement $S$ (e.g., increments), arrays indexed by $i$ appear only on the right hand side of $S$.
+\end{enumerate}
+\end{Def}
+
+\begin{Def}[Free space loop]
+A loop $L_i$ is said to be a free space loop if its iterations can be executed in any arbitrary order; that is, no dependencies exist across different iterations. 
+\end{Def}
+
+Altogether, these definitions lead us to formalize the class of loop nests for which we seek optimality.
+
+\begin{Def}[Finite element integration loop nest]
+A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free space loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
+\end{Def}
+
+To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free space loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions).
 The expression \texttt{F} implements the operator.
 
-% (the Laplacian operator in the case of Equation~\ref{eq:quadrature}, but our discussion applies in general).
 \begin{figure}[h]\begin{CenteredBox}
@@
 
-Our aim is a strategy to synthesize optimal loop nests of this form. In particular, we characterize optimality as follows.
+We emphasize that our aim is a strategy to synthesize optimal loop nests of this specific form. In particular, we characterize optimality as follows.
 

### 2015-10-18 TOMS_Paper/assembly.tex
Add section on transf space completeness

@@
 \item $i$ appears only as an array index, and
-\item for each augmented assignment statement $S$ (e.g., increments), arrays indexed by $i$ appear only on the right hand side of $S$.
+\item for each augmented assignment statement $S$ (e.g., an increment), arrays indexed by $i$ appear only on the right hand side of $S$.
 \end{enumerate}
@@
 
-\begin{Def}[Free space loop]
-A loop $L_i$ is said to be a free space loop if its iterations can be executed in any arbitrary order; that is, no dependencies exist across different iterations. 
+\begin{Def}[Free order loop]
+A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, no loop-carried dependence exists across two different iterations. 
 \end{Def}
 
-Altogether, these definitions lead us to formalize the class of loop nests for which we seek optimality.
+Altogether, these definitions allow us to formalize the class of loop nests for which we seek optimality.
 
 \begin{Def}[Finite element integration loop nest]
-A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free space loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
+\label{def:fem-loopnest}
+A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
 \end{Def}
@@
 To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free space loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions).
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions).
 The expression \texttt{F} implements the operator.
@@
 
-\subsection{Transformation Space}
-To synthesize optimal implementations we need:
+To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
 \begin{itemize}
-\item a characterization of the space of legal transformations for the class of loop nests considered impacting the operation count
+\item a characterization of the space of legal transformations impacting the operation count
 \item a cost model to select the optimal point in the transformation space
 \end{itemize}
-In this section, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.
+In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space.
+We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.
 
+\subsection{Transformation Space - Sharing Elimination}
+\label{sec:sharing-elimination}
 We start with introducing the fundamental notion of sharing.
@@
 {\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
-\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\end{figure}
+\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}
 
@@
 We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$.
-The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about generalized code motion).
+The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
 \end{Algo}
@@
 
-We observe, however, that for finite element integration loop nests the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
+\subsection{Transformation Space - Introducing Pre-evaluation}
+\label{sec:pre-evaluation}
+For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
 Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
@@
 {\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
-\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\end{figure}
+\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\label{code:loopnest_rednored}\end{figure}
 
@@
 
-We expand on these points in Section~\ref{sec:optimal-synthesis}. In order to address point 2, we conclude this section by providing a more refined definition of optimality.
+We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we provide a more refined definition of optimality.
 
@@
 \end{Def}
+
+\subsection{Completeness of the Transformation Space}
+It is helpful to consider sharing elimination and pre-evaluation as high level operators defined on top of basic transformations such as code motion, expansion, and factorization.
+These two operators allow us to analyze all possible ways in which an expression can be factorized; that is, by grouping:
+\begin{itemize}
+\item Test or trial functions (sharing elimination in the outer product loop nest)
+\item Integration-dependent terms (sharing elimination in the reduction loop)
+\item Together test functions, trial functions, and integration-dependent terms (pre-evaluation)
+\end{itemize}
+Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.
+
+Sharing elimination and pre-evaluation also aim to remove \textit{temporal redundancy}, which arises whenever the same operation with the same operands is performed more than once.
+The problem of removing temporal redundancy reduces to minimize the execution of invariant sub-expressions.
+This requires two transformations:
+\begin{itemize}
+ \item Generalized code motion (occurs in both sharing elimination and pre-evaluation).
+Generalized, as explained in~\citeN{Luporini}, means that invariant sub-expressions are sought with respect to all loops composing a loop nest (and not just the innermost, as in traditional loop-invariant code motion).
+Examples were provided in Figures~\ref{code:multi_loopnest} and~\ref{code:loopnest_rednored}.
+\item Evaluation of reduction loops (occurs in pre-evaluation). An example was provided in Figure~\ref{code:loopnest_rednored}.
+ \end{itemize}
+
+Sharing elimination and pre-evaluation (and, as we will clarify, their composition) allow us to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.
 
@@
 
-Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched.
-We consider them high level operators defined on top of expression-restructuring transformations such as code motion, re-association, and factorization.
-Other classes of loop or expression transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.
-
-Finally, with Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach.
+Finally, with Proposition~\ref{prop:optimal-approach}, we show the optimality of this approach.
 

### 2015-10-18 TOMS_Paper/assembly.tex
Changed quite a lot of stuff

@@
 
-\subsection{Transformation Space - Introducing Pre-evaluation}
+\subsection{Transformation Space - Pre-evaluation}
 \label{sec:pre-evaluation}
@@
 Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space.
-This would be challenging if we were to deal with arbitrary loop nests and expressions, while we will see that a simpler algorithm can be adopted for the class arising in finite element integration.
-In any case, however, the following issues should be considered:
+This would be challenging if we were to deal with arbitrary loop nests and expressions.
+However, we can make use of a result -- the foundation of tensor contraction mode -- to build a relatively simple algorithm.
+As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors.
+This result can be turned into a loop and expression transformation algorithm, for which we provide a succinct description below.
+
+\begin{Algo}[Pre-evaluation]
+\label{algo:pre-evaluation}
+\normalfont
+Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$ as in Figure~\ref{code:loopnest}.
+We dissect \texttt{F} into distinct sub-expressions (the monomials).
+Each sub-expression is factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
+This transformation is feasible, as a consequence of the results in~\citeN{Kirby:TC}.
+These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries.
+As part of this process, the reduction induced by $L_i$ is evaluated.
+Consequently, $L_i$ disappears from $\Lambda$.
+\end{Algo}
+
+The pre-evaluation of a monomial introduces, however, some critical issues:
 \begin{itemize}
@@
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
-\item The transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later).
+\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase, in general, the arithmetic complexity (e.g., expansion may increase the operation count).
 This could outweigh the gain due to pre-evaluation.
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
 \end{itemize}
 
-To summarize, so far we have highlighted four problems:
-\begin{enumerate}
-\item The need for an algorithm to expose pre-evaluation opportunities
-\item Potential explosion in working set size
-\item Potential increase in operation count due to manipulation of the expression tree
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
-\end{enumerate}
-
-We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we provide a more refined definition of optimality.
+We address these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we conclude providing a more refined definition of optimality.
 
@@
 
-Sharing elimination and pre-evaluation (and, as we will clarify, their composition) allow us to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.
+Sharing elimination and pre-evaluation allow us to construct an algorithm to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.
 
@@
 
-Based on these considerations, we establish the following memory constraints:
-\begin{enumerate}
-\item the size of a temporary due to code motion cannot be larger than that of the outer product loop nest iteration space;
-\item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
-\end{enumerate}
-A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden.
-This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.
+Based on these considerations, we establish the following memory constraints.
+
+\begin{Const}
+\label{const:Le}
+The size of a temporary due to code motion must not be proportional to the size of $L_e$.
+\end{Const}
+
+\begin{Const}
+\label{const:TH}
+The total amount of memory occupied by the temporaries due to code motion must not exceed a certain threshold, \texttt{$T_H$}.
+\end{Const}
+
+Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on ...
+data objects.
+Consequently, generalized code motion as performed by sharing elimination is not allowed to hoist expressions outside of $L_e$.
 
 \subsection{Reaching Optimality}
-We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.
+We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
 
@@
 As presaged in Section~\ref{sec:optimal-impl}, we consider whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
-
-To partly answer this question, we make use of a result -- the foundation of tensor contraction mode (Section~\ref{sec:tc}) -- from~\citeN{Kirby:TC}.
-Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}).
-We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
-These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$ and stored into temporaries.
-As part of this process, the reduction induced by $L_i$ is evaluated.
-Consequently, $L_i$ disappears from $\Lambda$.
-This transformation is actually what we were referring to when we more informally introduced pre-evaluation in Section~\ref{sec:optimal-impl}.
 
@@
 
-Because of symmetries in basis functions, pre-evaluation could produce identical tables, which would be mapped to the same temporary.
+Because of reuse of basis functions, pre-evaluation may produce identical tables, which will be mapped to the same temporary.
 The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing.
-This is why the last two lines of the algorithm, which goal is to eliminate sharing following the procedure described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all sub-expressions, regardless of whether pre-evaluation was used.
+
 The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.

### 2015-10-18 TOMS_Paper/assembly.tex
Trying to rewrite algorithm...

@@
 \lstset{language=C, breaklines=true, mathescape,
-			 morekeywords={and, each}}
+			 morekeywords={each}}
 
@@
 
-\section{Problem Characterization}
+\section{Transformation Space}
 \label{sec:optimal-impl}
@@
 
-\subsection{Transformation Space - Sharing Elimination}
+\subsection{Sharing Elimination}
 \label{sec:sharing-elimination}
@@
 
-\subsection{Transformation Space - Pre-evaluation}
+\subsection{Pre-evaluation}
 \label{sec:pre-evaluation}
@@
 
-\subsection{Completeness of the Transformation Space}
+\subsection{Completeness}
 It is helpful to consider sharing elimination and pre-evaluation as high level operators defined on top of basic transformations such as code motion, expansion, and factorization.
@@
 
-Sharing elimination and pre-evaluation allow us to construct an algorithm to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.
+Sharing elimination and pre-evaluation allow us to construct an algorithm to minimize spatial and temporal redundancy in loop nests and expressions as in Definition~\ref{def:fem-loopnest}.
 
-\section{Transformation Algorithm}
+\section{Framework for Optimal Synthesis}
 \label{sec:optimal-synthesis}
@@
 In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form.
-A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation (for which a deeper explanation will be provided).
-The optimality of the transformation algorithm is discussed.
+A fundamental aspect is the definition of a suitable cost function to assess the profitability of applying pre-evaluation to a monomial.
 
@@
 
-\subsection{Reaching Optimality}
+\subsection{Transformation Algorithm}
 We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
 
-We reinforce the idea that eliminating sharing from the outer product loop nest does not suffice.
-As presaged in Section~\ref{sec:optimal-impl}, we consider whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.
+Two main issues should be addressed: 
+\begin{itemize}
+\item Understanding for which monomials pre-evaluation is profitable.
+In order to expose pre-evaluation opportunities, Algorithm~\ref{algo:pre-evaluation} transforms the input expression by expanding products.
+This increases the operation count.
+We therefore say that pre-evaluation is profitable if the gain due to pre-evaluating the reduction loop is larger than such increase.
+\item Consider a form comprising two monomials $m1$ and $m2$.
+Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g.
+some basis functions).
+If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost.
+We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- provides the highest operation count reduction.
+\end{itemize}
 
-Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former.
-In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}.
-For this, we start with explaining the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.
+For the moment, let us assume that we own a cost function $\theta : M \rightarrow \mathbb{Z}$ that, given a monomial $m \in M$, returns the number of operations saved or introduced if pre-evaluation were applied to $m$.
+$\theta(m) < 0$ indicates that pre-evaluation is not profitable, because the operation count would actually increase by $|\theta(m)|$.
+Analogous considerations stand for $\theta(m) > 0$.
+We will show how to construct a proper $\theta$ in Section~\ref{sec:op_count}.
+Having defined $\theta$, we can now elaborate on the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.
 
@@
 \begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
-dissect the input expression into monomials
-for each monomial $m$:
-  if $\theta(m)$ > 0 and memory constraints satisfied:
-    extract into a separate loop nest
-    apply pre-evaluation
+M = $\lbrace$m | m is a monomial in the input expression$\rbrace$
+S = $\lbrace$m | $\theta(m)$ < 0$\rbrace$
+P = $\lbrace$m | $\theta(m)$ $\geq$ 0 and sharing(m, S) = $\emptyset \rbrace$
+maybe_P = M $\setminus$ (S $\cup$ P)
+B = $\lbrace$b | b is a bipartite graph from complete graph G = (maybe_P, E)$\rbrace$
+for each $b \in B$:
+  for each $(b_S, b_P) \in b$:
+
 for each expression:
@@
 
-Our choice is to evaluate the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''.
-We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form.
-$\theta(m)$ represents the change in operation count if pre-evaluation were applied to monomial $m$: a negative value indicates an increase, whereas a positive value indicates a gain.
 If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.

### 2015-10-18 TOMS_Paper/assembly.tex
Trying to rewrite algorithm...

@@
 S = $\lbrace$m | $\theta(m)$ < 0$\rbrace$
-P = $\lbrace$m | $\theta(m)$ $\geq$ 0 and sharing(m, S) = $\emptyset \rbrace$
-maybe_P = M $\setminus$ (S $\cup$ P)
+maybe_P = M $\setminus$ S
 B = $\lbrace$b | b is a bipartite graph from complete graph G = (maybe_P, E)$\rbrace$
@@
   for each $(b_S, b_P) \in b$:
-
+    if memory_required($b_P$) $\geq$ $T_H $:
+      continue
+    op_count[b] = \eta(S $\cup$ b_S) + $|\theta(b_P)|$
 for each expression:
@@
 
-If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
+%If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
 The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

### 2015-10-21 TOMS_Paper/assembly.tex
Trying to work around the issue with sharing elimination...

@@
 
-As shown later, outer product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code.
+As shown later, outer product loop nests are important because they have a structure that we can take advantage of to synthesize optimal code.
 We will see that they arise naturally when translating bilinear or linear forms into code.
@@
 \label{def:fem-loopnest}
-A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
+A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest whose body is an independent basic block.
 \end{Def}
@@
 \begin{Def}[Sharing factor]
-The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$).
-It provides an indication of the maximal number of symbols which can be factorized to the level of $L_i$.
+The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of expressions depending on $i$ (e.g., number of unique arrays indexed by $i$ if the expressions are just symbols).
+It provides an indication of the maximal number of expressions that can be factorized to the level of $L_i$.
 \end{Def}
@@
 
-Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
-Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}.
-More generally, arbitrarily complex outer product loop nests can systematically be reduced to optimal form using an algorithm that exploits the properties of linearity and perfectness.
-The algorithm and the proof that we provide below focus on outer product loop nests of arity $n=2$.
-For $n < 2$ the proof would be trivial, whereas for $n > 2$ a more general treatment would be required; fortunately, this case is never of interest in finite element integration.
+For example, Figure~\ref{code:multi_loopnest_a} shows a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
+The optimal synthesis for this loop nest is provided in Figure~\ref{code:multi_loopnest_b}.
 
-\begin{Algo}[Optimal code motion for outer product loop nests]
+Our objective is a transformation capable of eliminating sharing from finite element integration loop nests.
+The transformation should find a factorization that maximizes code motion (as in the example) and common sub-expressions.
+In general, this would be a large combinatorial problem, but thanks to the specific structure of the loop nests considered, simpler approaches can be found.
+In particular, we describe two algorithms: the former leads to ``inner loop optimality'' by only performing an additional visit of the expression tree, whereas the latter allows reaching actual optimality at the price of a more costly worst-case analysis.
+
+By inner loop optimality we mean optimality at the level of the outer product loop nest; that is, factorization opportunities are sought only for terms depending on the inner loops.
+The cost of the outer loops may or may not be optimal.
+It is interesting to study this problem because of three reasons: 1) in many cases, especially for higher order methods, the bulk of the computation is carried out in the outer product loop nest; 2) it is easily solvable, so it can be implemented with minimum effort; 3) the analysis has negligible cost.
+This algorithm (and its proof) exploits the key properties of linearity and perfectness, as well as the structure inherent in the expressions.
+
+\begin{Algo}[Sharing elimination for inner-loop optimality]
 \label{algo:outer-product-optimal}
 \normalfont
-Consider an outer product loop nest $\Lambda = [L_i, L_j]$, with the body of $L_j$ being an independent basic block.
-The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_j$; this exposes the space of all possible factorizations.
-Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products.
+Consider the finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$.
+The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_k$ such that $L_j-$ and $L_k-$dependent terms lie on the same level of the expression tree.
+This exposes the space of all possible factorizations for (derivatives of) test and trial functions.
+
+Since $[L_j, L_k]$ is an outer product loop nest, the resulting expressions are summations of outer products.
 We traverse the expression tree and determine the sharing factor of the two linear loops.
-We take the symbols depending on the loop with smallest sharing factor, say $L_i$, and we incrementally factorize them.
-Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
-The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
-We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$.
+We take the symbols depending on the loop with smallest sharing factor, say $L_j$, and we incrementally factorize them.
+Due to linearity, each factored product $P$ only has one symbol depending on $L_j$, and this symbol is now unique in the expression.
+The other term in $P$, independent of $L_j$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
+We repeat the same process for the symbols along the loop with higher sharing factor, $L_k$.
 The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
@@
 \label{prop:optimal-mln}
-Let $\Lambda = [L_i, L_j]$ be an outer product loop nest with the body of $L_j$ being an independent basic block.
-Then Algorithm~\ref{algo:outer-product-optimal} leads to an optimal synthesis $\Lambda'$.
+Let $\Lambda=[L_e, L_i, L_j, L_k]$ be a finite element integration loop nest. Then Algorithm~\ref{algo:outer-product-optimal} leads to an optimal synthesis of $\Lambda'=[L_j', L_k']$.
 \end{Prop}
 \begin{proof}
-Assume, without loss of generality, that $L_i$ is the loop with smallest sharing factor. We reduce the proof to showing that the two following facts are true (we do it by contradiction).
+Assume, without loss of generality, that $L_j$ is the loop with smallest sharing factor. We reduce the proof to showing that the two following facts are true (we do it by contradiction).
 \begin{enumerate}
-\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case).
-If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$.
-Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario).
-In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$.
-However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product.
-Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
-\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case).
-Let us assume, for a moment, that this is false.
-In this scenario, at least one symbol is excluded from the factored product.
-Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$.
-We see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal.
+\item Taking $L_k$, the loop with higher sharing factor, as point of departure leads to a sub-optimal $\Lambda'$.
+If we assumed this were false, there would be two possibilities to consider: A) both $L_j$ and $L_k$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_j$ and $L_k$.
+Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal.
+In case B), due to linearity, factorization produces a summation of $L^{sf}_{k}$ outer products, as opposed to $L^{sf}_{j}$ if we had started with $L_j$.
+However, factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product.
+Since $L^{sf}_{k} > L^{sf}_{j}$, it is clear that $\Lambda'$ cannot be optimal, which contradicts the hypothesis.
+\item Partial elimination of sharing can lead to a sub-optimal $\Lambda'$.
+Let us assume that this is false.
+In this scenario, at least one term is excluded from a factored outer product, for instance $a_j e_k$ in $a_j(b_k + c_k + ...) + a_j e_k + ...$.
+We see that either there is another factorization opportunity along $L_k$ ($e_k$ is shared with some other outer products), or $\Lambda'$ is sub-optimal.
 Consequently, the hypothesis is generally contradicted.
 \end{enumerate}
+\end{proof}
+
+\begin{Rem*}
+\normalfont
+The algorithm and the proof focus on outer product loop nests of arity $n=2$, although a similar (simpler) reasoning is possible for $n < 2$.
+If $n > 2$, a more general treatment is required; this is however a case that, in finite element integration, is never of interest.
+\end{Rem*}
+
+To reach a global optimum (i.e., such that the number of operations at each level of the loop nest is minimum), we need to consider all factorization opportunities.
+In practice, however, the structure inherent in our expressions often allows to significantly prune the search space.
+As we explain below, the key observation to make is that an expression can always be reduced to a summation of terms of the form $a_{ij} b_{ik} \sigma_{ei}$, in which $a_{ij}$ and $b_{ik}$ represent test and trial functions, while $\sigma_{ei}$ a sub-expression including geometry- ($L_e$) and integration- ($L_i$) dependent terms.
+The idea is that this form exposes all possible factorizations leading to code motion.
+
+\begin{Algo}[Sharing elimination]
+\label{algo:fem-optimal}
+\normalfont
+Consider the finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$.
+We start with restructuring the expressions within $L_k$ such that they are in the form $\sum_{a \in V, b \in U} a_{ij} b_{ik} \sigma_{ei}^{ab}$, where $V$ and $U$ represent sets of test and trial functions.
+
+The loose end we have to tie up is the transformation of $\sigma_{ei}^{ab}$ terms.
+If $L_i$ is a linear loop, we can straightforwardly apply Algorithm~\ref{algo:outer-product-optimal} (with arity $n=1$) so that integration-dependent terms are grouped and geometry-dependent sub-expressions are hoisted.
+If, on the other hand, $L_i$ is non linear, we try all possible factorizations of $L_i$-dependent terms to maximize motion of geometry-dependent sub-expressions (although, we believe, a more refined approach, which avoids searching the whole space, can be found).
+\end{Algo}
+
+\begin{Prop}
+\label{prop:optimal-mln}
+Let $\Lambda=[L_e, L_i, L_j, L_k]$ be a finite element integration loop nest. Then Algorithm~\ref{algo:fem-optimal} leads to an optimal synthesis of $\Lambda'=[L_e', L_i', L_j', L_k']$.
+\end{Prop}
+\begin{proof}
+... This should be interpreted as our initial ``guess solution'' ...
 \end{proof}

### 2015-10-24 TOMS_Paper/assembly.tex
Change again sharing elimination---ufff

--- 2015-10-18 TOMS_Paper/assembly.tex
@@
 \renewcommand{\ttdefault}{txtt}
+
+% line over text in non math mode
+\newcommand{\textoverline}[1]{$\overline{\mbox{#1}}$}
 
@@
 \begin{Def}[Outer product loop nest]
+\label{def:jk-loop}
 An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs once in every summand.
@@
 \begin{Def}[Reduction loop]
+\label{def:i-loop}
 A loop $L_i$ is said to be a reduction loop if in its body
@@
 \begin{Def}[Free order loop]
-A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, no loop-carried dependence exists across two different iterations. 
+\label{def:e-loop}
+A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, there are no loop-carried dependencies across different iterations. 
 \end{Def}
 
-Altogether, these definitions allow us to formalize the class of loop nests for which we seek optimality.
+We use these definitions to formalize the class of loop nests for which we seek optimality.
 
@@
 \end{itemize}
-In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space.
-We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.
+In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space.
 
@@
 
-\begin{Def}[Sharing factor]
-The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$).
-It provides an indication of the maximal number of symbols which can be factorized to the level of $L_i$.
-\end{Def}
-
 \begin{figure}[h]\begin{CenteredBox}
-{\subfigcapskip = 7pt \subfigure[With sharing.
-Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
+{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
 ~~~~~~~~~~~~~~
@@
 
-Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
-Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}.
-More generally, arbitrarily complex outer product loop nests can systematically be reduced to optimal form using an algorithm that exploits the properties of linearity and perfectness.
-The algorithm and the proof that we provide below focus on outer product loop nests of arity $n=2$.
-For $n < 2$ the proof would be trivial, whereas for $n > 2$ a more general treatment would be required; fortunately, this case is never of interest in finite element integration.
+For instance, Figure~\ref{code:multi_loopnest_a} shows a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
+The optimal synthesis for this loop nest is provided in Figure~\ref{code:multi_loopnest_b}.
+This example illustrates how factorization can be applied to expose code motion opportunities in multilinear loop nests.
 
-\begin{Algo}[Optimal code motion for outer product loop nests]
-\label{algo:outer-product-optimal}
+In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (potentially) common sub-expression elimination.
+In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
+However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned.
+In the following, we assume outer product loop nest of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$ (which is actually just a special case).
+
+The first observation is that finite element expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of the coefficients.
+
+Given the peculiar structure of our terms, the key idea is that, in the space $S$ of all possible factorizations, the optimum almost always lies in a tiny subspace (typically of size one).
+We refer to this space as the \textit{potential search space} ($PSS$).
+Our approach is therefore as follows: we calculate lower and upper bounds (respectively, $lb$ and $ub$) for families of factorizations and we compare them: if, as expected, $ub(PSS) < lb(\overline{PPS})$, we prune $\overline{PPS}$.
+Otherwise, we have to search in $\overline{PPS}$ too.
+
+A potentially optimal factorization is one that groups together test and trial functions.
+Such a factorization exposes sub-expressions independent of at least one loop and, therefore, hoistable.
+The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits hoisting of sub-expressions including test and trial functions.
+Algorithm~\ref{algo:pps} provides a succinct description of how $PPS$ is built.
+
+\begin{Algo}[Construction of the potential search space]
+\label{algo:pps}
 \normalfont
-Consider an outer product loop nest $\Lambda = [L_i, L_j]$, with the body of $L_j$ being an independent basic block.
-The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_j$; this exposes the space of all possible factorizations.
-Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products.
-We traverse the expression tree and determine the sharing factor of the two linear loops.
-We take the symbols depending on the loop with smallest sharing factor, say $L_i$, and we incrementally factorize them.
-Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression.
-The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
-We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$.
-The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
+Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$.
+The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
+
+The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
+We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms (it can easily be demonstrated, by contradiction, that one should start from the loop with the least number of unique terms in order to obtain the smallest possible operation count).
+Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression.
+The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
+Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
 \end{Algo}
 
-\begin{Prop}
-\label{prop:optimal-mln}
-Let $\Lambda = [L_i, L_j]$ be an outer product loop nest with the body of $L_j$ being an independent basic block.
-Then Algorithm~\ref{algo:outer-product-optimal} leads to an optimal synthesis $\Lambda'$.
-\end{Prop}
-\begin{proof}
-Assume, without loss of generality, that $L_i$ is the loop with smallest sharing factor. We reduce the proof to showing that the two following facts are true (we do it by contradiction).
-\begin{enumerate}
-\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case).
-If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$.
-Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario).
-In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$.
-However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product.
-Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
-\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case).
-Let us assume, for a moment, that this is false.
-In this scenario, at least one symbol is excluded from the factored product.
-Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$.
-We see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal.
-Consequently, the hypothesis is generally contradicted.
-\end{enumerate} 
-\end{proof}
+%A special case is when test and trial functions belong to the same function space: this induces common sub-expressions \textit{over different loops}, so a potentially winning strategy consists of hoisting both $\alpha_{eij}$ and $\beta_{eik}$ and exploiting the fact that $\alpha = \beta$.
+---- If we note that, for some $w$, $\alpha = \beta$, we add the following strategy to $PPS$: hoist $\alpha_{eij}^{w}$ (or, equivalently, $\beta_{eik}^{w}$); replace common sub-expressions; minimize cost of $\sigma_{ei}$, which we discuss later.
+
+%The loose end we have to tie up is the transformation of $\sigma_{ei}^{ab}$ terms.
+If $L_i$ is a linear loop, we can straightforwardly apply Algorithm~\ref{algo:outer-product-optimal} (with arity $n=1$) so that integration-dependent terms are grouped and geometry-dependent sub-expressions are hoisted.
+If, on the other hand, $L_i$ is non linear, we try all possible factorizations of $L_i$-dependent terms to maximize motion of geometry-dependent sub-expressions (although, we believe, a more refined approach, which avoids searching the whole space, can be found).
 
@@
 \label{sec:pre-evaluation}
-For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form.
+Sharing elimination aims to reduce the operation count by applying factorization, code motion and/or common sub-expression elimination.
+There is a fourth operator impacting the operation count: \textit{reduction pre-evaluation}.
+In this section, we discuss role and impact of this operator.
+We also show that there exists a single point ($\in \overline{PPS}$) in the whole transformation space for which reduction pre-evaluation is legal.
+
+We start with an example.
 Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
@@
 
-Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space.
+Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space.
 This would be challenging if we were to deal with arbitrary loop nests and expressions.
-However, we can make use of a result -- the foundation of tensor contraction mode -- to build a relatively simple algorithm.
+However, we can make use of a result -- the foundation of tensor contraction mode -- to solve the problem by means of a relatively simple algorithm.
 As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors.

### 2015-10-25 TOMS_Paper/assembly.tex
Change again sharing elimination---ufff

@@
 
-The first observation is that finite element expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of the coefficients.
+From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+This particular form has led us to the key observation that, in the space $S$ of all possible sharing elimination alternatives, the optimum $opt$ is almost always reached through a specific strategy.
+Our approach is therefore as follows: we define families of factorization strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the upper bound ($ub$) of $opt$;  if, as expected, $ub(opt) < lb(S \setminus opt)$, then we know that $opt$ performs optimal sharing elimination.
 
-Given the peculiar structure of our terms, the key idea is that, in the space $S$ of all possible factorizations, the optimum almost always lies in a tiny subspace (typically of size one).
-We refer to this space as the \textit{potential search space} ($PSS$).
-Our approach is therefore as follows: we calculate lower and upper bounds (respectively, $lb$ and $ub$) for families of factorizations and we compare them: if, as expected, $ub(PSS) < lb(\overline{PPS})$, we prune $\overline{PPS}$.
-Otherwise, we have to search in $\overline{PPS}$ too.
+The potentially optimal factorization groups together test and trial functions in order to expose sub-expressions independent of at least one loop and, therefore, hoistable.
+The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits sub-expression hoisting due to dependencies on loop indices.
+Algorithm~\ref{algo:opt} describes how $opt$ is derived; Algorithm~\ref{algo:se-pick} discuss how we determine the actual optimality of $opt$.
 
-A potentially optimal factorization is one that groups together test and trial functions.
-Such a factorization exposes sub-expressions independent of at least one loop and, therefore, hoistable.
-The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits hoisting of sub-expressions including test and trial functions.
-Algorithm~\ref{algo:pps} provides a succinct description of how $PPS$ is built.
-
-\begin{Algo}[Construction of the potential search space]
-\label{algo:pps}
+\begin{Algo}[Construction of the potential optimum]
+\label{algo:opt}
 \normalfont
@@
 The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
-We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms (it can easily be demonstrated, by contradiction, that one should start from the loop with the least number of unique terms in order to obtain the smallest possible operation count).
+We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms (it can easily be demonstrated, by contradiction, that one should start from the loop with the least number of unique terms in order to obtain the smallest operation count).
 Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression.
@@
 Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
+
+If there exists $w_1$ and $w_2$ such that $\sigma_{ei}^{w_1} \neq \sigma_{ei}^{w_2}$, then we also need decide how to schedule operations in $L_i$.
+For the potential optimum, we factorize $L_i$-dependent terms, which promotes hoisting of $L_e$-dependent sub-expressions.
 \end{Algo}
+
+\begin{Algo}[Optimality check of $opt$]
+\label{algo:se-pick}
+\normalfont
+aaa
+\end{Algo}
+%There are two main possibilities: 1) factorization of $L_i$-dependent terms (i.e., coefficients and their derivatives) followed by code motion of $L_e$-dependent terms; 2) factorization of both $L_i$- and $L_e$-dependent terms, followed by common sub-expression elimination.
 
@@
 There is a fourth operator impacting the operation count: \textit{reduction pre-evaluation}.
-In this section, we discuss role and impact of this operator.
-We also show that there exists a single point ($\in \overline{PPS}$) in the whole transformation space for which reduction pre-evaluation is legal.
+In this section, we discuss the role of this operator.
+We also show that there exists a single point ($\in \overline{PSS}$) in the whole transformation space for which reduction pre-evaluation is legal.
 
@@
 
-\section{Framework for Optimal Synthesis}
+\section{Optimal Synthesis}
 \label{sec:optimal-synthesis}

### 2015-10-25 TOMS_Paper/assembly.tex
Other fixes

@@
 \begin{itemize}
-\item We define optimality for finite element integration as number of flops given a memory bound; we build the transformation space within which the optimal is sought.
+\item We define optimality for finite element integration as number of flops given a memory bound; we build the space of legal transformations within which the optimal is sought.
 \item We provide an algorithm to select optimal points in the transformation space.
 The algorithm uses a cost model to: 1) understand whether a transformation reduces or increases the operation count; 2) choose between different (non-composable) transformations.
-\item We state for which class of problems optimality can be reached, and for which class heuristics are instead needed to minimize (reduce) the operation count.
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 The optimal synthesis for this loop nest is provided in Figure~\ref{code:multi_loopnest_b}.
-This example illustrates how factorization can be applied to expose code motion opportunities in multilinear loop nests.
+This example illustrates how factorization, besides reducing the number of multiplications required at a given point in the loop nest, can expose code motion opportunities.
 
@@
 However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned.
-In the following, we assume outer product loop nest of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$ (which is actually just a special case).
+In the following, we assume outer product loop nests of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$.
 
 From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-This particular form has led us to the key observation that, in the space $S$ of all possible sharing elimination alternatives, the optimum $opt$ is almost always reached through a specific strategy.
-Our approach is therefore as follows: we define families of factorization strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the upper bound ($ub$) of $opt$;  if, as expected, $ub(opt) < lb(S \setminus opt)$, then we know that $opt$ performs optimal sharing elimination.
+This particular form has led us to the key observation that, in the space $S$ including all possible ways of performing sharing elimination, the optimum ($opt$) is almost always reached through a specific sequence of transformations.
+Our approach is therefore as follows: we define families of factorization strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the upper bound ($ub$) of $opt$;  if, as expected, $ub(\lbrace opt \rbrace) < lb(S \setminus \lbrace opt \rbrace)$, then we know that $opt$ performs optimal sharing elimination.
 
-The potentially optimal factorization groups together test and trial functions in order to expose sub-expressions independent of at least one loop and, therefore, hoistable.
-The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits sub-expression hoisting due to dependencies on loop indices.
-Algorithm~\ref{algo:opt} describes how $opt$ is derived; Algorithm~\ref{algo:se-pick} discuss how we determine the actual optimality of $opt$.
+The idea behind $opt$ is that by grouping together test and trial functions we expose sub-expressions independent of at least one loop and, therefore, hoistable.
+The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits code motion due to dependencies on loop indices.
+Algorithm~\ref{algo:opt} describes how $opt$ is derived.
+We postpone to Section~\ref{sec:optimal-synthesis} the discussion on how we assess the actual optimality of $opt$.
 
@@
 
-\begin{Algo}[Optimality check of $opt$]
-\label{algo:se-pick}
-\normalfont
-aaa
-\end{Algo}
-%There are two main possibilities: 1) factorization of $L_i$-dependent terms (i.e., coefficients and their derivatives) followed by code motion of $L_e$-dependent terms; 2) factorization of both $L_i$- and $L_e$-dependent terms, followed by common sub-expression elimination.
-
-%A special case is when test and trial functions belong to the same function space: this induces common sub-expressions \textit{over different loops}, so a potentially winning strategy consists of hoisting both $\alpha_{eij}$ and $\beta_{eik}$ and exploiting the fact that $\alpha = \beta$.
----- If we note that, for some $w$, $\alpha = \beta$, we add the following strategy to $PPS$: hoist $\alpha_{eij}^{w}$ (or, equivalently, $\beta_{eik}^{w}$); replace common sub-expressions; minimize cost of $\sigma_{ei}$, which we discuss later.
-
-%The loose end we have to tie up is the transformation of $\sigma_{ei}^{ab}$ terms.
-If $L_i$ is a linear loop, we can straightforwardly apply Algorithm~\ref{algo:outer-product-optimal} (with arity $n=1$) so that integration-dependent terms are grouped and geometry-dependent sub-expressions are hoisted.
-If, on the other hand, $L_i$ is non linear, we try all possible factorizations of $L_i$-dependent terms to maximize motion of geometry-dependent sub-expressions (although, we believe, a more refined approach, which avoids searching the whole space, can be found).
-
 \subsection{Pre-evaluation}
 \label{sec:pre-evaluation}
-Sharing elimination aims to reduce the operation count by applying factorization, code motion and/or common sub-expression elimination.
+Sharing elimination aims to reduce the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination.
 There is a fourth operator impacting the operation count: \textit{reduction pre-evaluation}.
-In this section, we discuss the role of this operator.
-We also show that there exists a single point ($\in \overline{PSS}$) in the whole transformation space for which reduction pre-evaluation is legal.
+In this section, we discuss its role and show that there exists a single point in the whole transformation space for which the transformation is legal.
 
@@
 This would be challenging if we were to deal with arbitrary loop nests and expressions.
-However, we can make use of a result -- the foundation of tensor contraction mode -- to solve the problem by means of a relatively simple algorithm.
+However, we can make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
 As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors.
-This result can be turned into a loop and expression transformation algorithm, for which we provide a succinct description below.
+This result can be turned into a transformation algorithm for loops and expressions, for which we provide a succinct description below.
 
@@
 \normalfont
-Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$ as in Figure~\ref{code:loopnest}.
+Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$.
 We dissect \texttt{F} into distinct sub-expressions (the monomials).
@@
 
-The pre-evaluation of a monomial introduces, however, some critical issues:
+The pre-evaluation of a monomial introduces some critical issues:
 \begin{itemize}
@@
 \subsection{Completeness}
-It is helpful to consider sharing elimination and pre-evaluation as high level operators defined on top of basic transformations such as code motion, expansion, and factorization.
-These two operators allow us to analyze all possible ways in which an expression can be factorized; that is, by grouping:
-\begin{itemize}
-\item Test or trial functions (sharing elimination in the outer product loop nest)
-\item Integration-dependent terms (sharing elimination in the reduction loop)
-\item Together test functions, trial functions, and integration-dependent terms (pre-evaluation)
-\end{itemize}
-Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.
+We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization.
+Factorization addresses \textit{spatial redundancy}.
+The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.
+Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.
 
-Sharing elimination and pre-evaluation also aim to remove \textit{temporal redundancy}, which arises whenever the same operation with the same operands is performed more than once.
-The problem of removing temporal redundancy reduces to minimize the execution of invariant sub-expressions.
-This requires two transformations:
-\begin{itemize}
- \item Generalized code motion (occurs in both sharing elimination and pre-evaluation).
-Generalized, as explained in~\citeN{Luporini}, means that invariant sub-expressions are sought with respect to all loops composing a loop nest (and not just the innermost, as in traditional loop-invariant code motion).
-Examples were provided in Figures~\ref{code:multi_loopnest} and~\ref{code:loopnest_rednored}.
-\item Evaluation of reduction loops (occurs in pre-evaluation). An example was provided in Figure~\ref{code:loopnest_rednored}.
- \end{itemize}
-
-Sharing elimination and pre-evaluation allow us to construct an algorithm to minimize spatial and temporal redundancy in loop nests and expressions as in Definition~\ref{def:fem-loopnest}.
+Sharing elimination and reduction pre-evaluation tackle the problem of minimizing spatial and temporal redundancy in the loop nests and expressions of Definition~\ref{def:fem-loopnest}.
+The space of all possible factorizations is explored, and for each point in this space, code motion, common sub-expression elimination, and reduction pre-evaluation opportunities are analyzed.
 
@@
 In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form.
-A fundamental aspect is the definition of a suitable cost function to assess the profitability of applying pre-evaluation to a monomial.
+A fundamental aspect is the definition of suitable cost functions to ascertain: 1) the profitability of pre-evaluation; 2) the optimality of sharing elimination (recall that Algorithm~\ref{algo:opt} provides, optimistically, a potentially optimal implementation of the transformation).
 
@@
 \subsection{Transformation Algorithm}
-We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
+We discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
 
-Two main issues should be addressed: 
+At this point, three main issues are still to be addressed: 
 \begin{itemize}
-\item Understanding for which monomials pre-evaluation is profitable.
-In order to expose pre-evaluation opportunities, Algorithm~\ref{algo:pre-evaluation} transforms the input expression by expanding products.
-This increases the operation count.
-We therefore say that pre-evaluation is profitable if the gain due to pre-evaluating the reduction loop is larger than such increase.
+\item 
+\item Given a monomial, deciding whether to apply pre-evaluation (instead of sharing elimination).
+Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation can actually increase the operation count because of product expansion.
+In the following, by saying that pre-evaluation is profitable, we simply mean that the resulting operation count is smaller than that produced through sharing elimination.
 \item Consider a form comprising two monomials $m1$ and $m2$.
@@
 If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost.
-We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- provides the highest operation count reduction.
+We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction.
+In other words, we seek a strategy for finding the global optimum.
 \end{itemize}
@@
 
+% apply memory constraints: reduction to knapsack problem ... not implemented, we use greedy
+
 %If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
 The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.
+
+%\begin{Algo}[Optimality check of $opt$]
+%\label{algo:se-pick}
+%\normalfont
+%aaa
+%\end{Algo}
+%There are two main possibilities: 1) factorization of $L_i$-dependent terms (i.e., coefficients and their derivatives) followed by code motion of $L_e$-dependent terms; 2) factorization of both $L_i$- and $L_e$-dependent terms, followed by common sub-expression elimination.
+
+%A special case is when test and trial functions belong to the same function space: this induces common sub-expressions \textit{over different loops}, so a potentially winning strategy consists of hoisting both $\alpha_{eij}$ and $\beta_{eik}$ and exploiting the fact that $\alpha = \beta$.
+---- If we note that, for some $w$, $\alpha = \beta$, we add the following strategy to $PPS$: hoist $\alpha_{eij}^{w}$ (or, equivalently, $\beta_{eik}^{w}$); replace common sub-expressions; minimize cost of $\sigma_{ei}$, which we discuss later.
 
@@
 
-\subsection{Cost Function}
+\subsection{Cost Functions}
+It remains to tie up one loose end: the construction of cost functions to 1) estimate the impact of pre-evaluating monomials (Section~\ref{sec:op_count}); 2) check that the default sharing elimination strategy is optimal (Section~\ref{sec:check-se}).
+Since the cost functions are expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
+
+\subsubsection{Profitability of pre-evaluation}
 \label{sec:op_count}
-It remains to tie up one loose end: the construction of the cost function $\theta$.
-Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.
-
+We build the cost function $\theta$.
 According to what established in the previous section, we have $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$; we recall that $\theta_{wo}$ and $\theta_{w}$ represent the operation counts without and with pre-evaluation.
@@
 
-% apply memory constraints: reduction to knapsack problem ... not implemented, we use greedy
+\subsubsection{Optimality of sharing elimination}
+\label{sec:check-se}
+aaa
 
@@
 A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion.
-We have primarily used these operators to implement sharing elimination and pre-evaluation.
-Further, we have exploited them for heuristic optimization of the integration loop when this is non-linear (e.g., whenever the form has non-linear coefficients).
-In such a case, expansion, factorization, and code motion are applied greedily, selecting the term with highest sharing factor at each iteration, based on the observation that this should minimize sharing and promote hoisting.
+Sharing elimination and pre-evaluation are implemented by applying (composing) in specific sequences these operators.
 

### 2015-10-25 TOMS_Paper/assembly.tex
Other fixes

@@
 At this point, three main issues are still to be addressed: 
-\begin{itemize}
-\item 
+\begin{enumerate}
 \item Given a monomial, deciding whether to apply pre-evaluation (instead of sharing elimination).
@@
 In the following, by saying that pre-evaluation is profitable, we simply mean that the resulting operation count is smaller than that produced through sharing elimination.
+\item 
 \item Consider a form comprising two monomials $m1$ and $m2$.
@@
 In other words, we seek a strategy for finding the global optimum.
-\end{itemize}
+\end{enumerate}
 
-For the moment, let us assume that we own a cost function $\theta : M \rightarrow \mathbb{Z}$ that, given a monomial $m \in M$, returns the number of operations saved or introduced if pre-evaluation were applied to $m$.
-$\theta(m) < 0$ indicates that pre-evaluation is not profitable, because the operation count would actually increase by $|\theta(m)|$.
-Analogous considerations stand for $\theta(m) > 0$.
-We will show how to construct a proper $\theta$ in Section~\ref{sec:op_count}.
-Having defined $\theta$, we can now elaborate on the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.
+For the moment, let us assume that we own a cost function $\theta : M \rightarrow \mathbb{Z}$ that, given a monomial $m \in M$, returns the gain/loss of pre-evaluating $m$ in place of sharing elimination.
+$\theta(m) < 0$ indicates that pre-evaluation is not profitable, with the operation count increasing by $|\theta(m)|$.
+Similarly for $\theta(m) > 0$.
+We return to the issue of constructing $\theta$ in Section~\ref{sec:op_count}.
+Having defined $\theta$, we now show the algorithm driving the transformation process.
+The pseudo-code is provided in Figure~\ref{code:intuition}.
 
@@
 \label{sec:op_count}
+
+%TODO : say that avoiding useless pre-evaluation is extremely important !!!!!!!!!!! Performing sharing elimination is fast and easy, pre-evaluationi s not !
+
 We build the cost function $\theta$.

### 2015-10-26 TOMS_Paper/assembly.tex
Reworking section 4

--- 2015-10-25 TOMS_Paper/assembly.tex
@@
 \lstset{language=C, breaklines=true, mathescape,
-			 morekeywords={each}}
+			 morekeywords={each,min,not}}
 
@@
 From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-This particular form has led us to the key observation that, in the space $S$ including all possible ways of performing sharing elimination, the optimum ($opt$) is almost always reached through a specific sequence of transformations.
-Our approach is therefore as follows: we define families of factorization strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the upper bound ($ub$) of $opt$;  if, as expected, $ub(\lbrace opt \rbrace) < lb(S \setminus \lbrace opt \rbrace)$, then we know that $opt$ performs optimal sharing elimination.
+This particular form has led us to the key observation that, in the space $S$ of all possible sharing elimination strategies, the optimum ($opt$) is almost always reached through a specific sequence of transformations.
+Our approach is therefore as follows: we define families of sharing elimination strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the upper bound ($ub$) of $opt$;  if, as expected, $ub(\lbrace opt \rbrace) < lb(S \setminus \lbrace opt \rbrace)$, then we know that $opt$ is actually optimal.
 
-The idea behind $opt$ is that by grouping together test and trial functions we expose sub-expressions independent of at least one loop and, therefore, hoistable.
+The idea behind $opt$ is that by grouping together test and trial functions sub-expressions independent of at least one loop (i.e., hoistable) are exposed.
 The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits code motion due to dependencies on loop indices.
 Algorithm~\ref{algo:opt} describes how $opt$ is derived.
-We postpone to Section~\ref{sec:optimal-synthesis} the discussion on how we assess the actual optimality of $opt$.
+We postpone to Section~\ref{sec:check-se} the discussion on how we assess the actual optimality of $opt$.
 
@@
 \label{sec:pre-evaluation}
-Sharing elimination aims to reduce the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination.
-There is a fourth operator impacting the operation count: \textit{reduction pre-evaluation}.
-In this section, we discuss its role and show that there exists a single point in the whole transformation space for which the transformation is legal.
+Sharing elimination reduces the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination.
+A fourth operator impacting the operation count is \textit{reduction pre-evaluation}.
+In this section, we discuss its role and show that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.
 
@@
 
-We address these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we conclude providing a more refined definition of optimality.
-
-\begin{Def}[Optimality of a loop nest with bounded working set]
-\label{def:optimality}
-Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and a set of memory constraints is satisfied.
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
-\end{Def}
-
-\subsection{Completeness}
-We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization.
-Factorization addresses \textit{spatial redundancy}.
-The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.
-Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.
-
-Sharing elimination and reduction pre-evaluation tackle the problem of minimizing spatial and temporal redundancy in the loop nests and expressions of Definition~\ref{def:fem-loopnest}.
-The space of all possible factorizations is explored, and for each point in this space, code motion, common sub-expression elimination, and reduction pre-evaluation opportunities are analyzed.
-
-\section{Optimal Synthesis}
-\label{sec:optimal-synthesis}
-
-In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form.
-A fundamental aspect is the definition of suitable cost functions to ascertain: 1) the profitability of pre-evaluation; 2) the optimality of sharing elimination (recall that Algorithm~\ref{algo:opt} provides, optimistically, a potentially optimal implementation of the transformation).
+We address these points in Section~\ref{sec:optimal-synthesis}.
 
@@
 \label{sec:mem-const}
-Two observations motivate the need for memory constraints.
-Our starting point is again the bilinear form implementation in Figure~\ref{code:loopnest}.
-Analogous considerations apply to forms of arbitrary arity.
 
-The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions.
+In the previous section, we provided an insight into the potentially negative effects of code motion.
+In this section, we expand on this matter.
+The following two observations, in particular, lead to the definition of \textit{memory constraints}.
+
+\begin{itemize}
+\item The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions.
 Imagine $\Lambda$ is enclosed in a time stepping loop.
@@
 The gain in number of operations executed could therefore be outweighed, from a runtime viewpoint, by a much larger memory pressure.
-
-The second observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g.
+\item For certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
+For example, as explained in Section~\ref{sec:pre-evaluation}, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
 This can sometimes break such a ``local memory threshold''.
 In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.
+\end{itemize}
 
@@
 
+We conclude providing a more refined definition of optimality.
+
+\begin{Def}[Optimality of a loop nest with bounded working set]
+\label{def:optimality}
+Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and Constraints~\ref{const:Le} and~\ref{const:TH} are satisfied.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
+\end{Def}
+
+\subsection{Completeness}
+We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization.
+Factorization addresses \textit{spatial redundancy}.
+The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.
+Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.
+
+Sharing elimination and reduction pre-evaluation tackle the problem of minimizing spatial and temporal redundancy in the loop nests and expressions of Definition~\ref{def:fem-loopnest}.
+The space of all possible factorizations is explored, and for each point in this space, code motion (restricted by Constraint~\ref{const:Le}), common sub-expression elimination, and reduction pre-evaluation opportunities (restricted by Constraint~\ref{const:TH}) are analyzed.
+
+\section{Optimal Synthesis}
+\label{sec:optimal-synthesis}
+In this section, we build a transformation algorithm capable of reaching the goal set by Definition~\ref{def:optimality} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
+In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.
+
 \subsection{Transformation Algorithm}
-We discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
 
 At this point, three main issues are still to be addressed: 
-\begin{itemize}
-\item 
-\item Given a monomial, deciding whether to apply pre-evaluation (instead of sharing elimination).
-Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation can actually increase the operation count because of product expansion.
-In the following, by saying that pre-evaluation is profitable, we simply mean that the resulting operation count is smaller than that produced through sharing elimination.
-\item Consider a form comprising two monomials $m1$ and $m2$.
+\begin{enumerate}
+\item \textit{Ascertain the optimality of sharing elimination.} Recall that Algorithm~\ref{algo:opt} performs a potentially optimal sharing elimination strategy.
+\item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation, which is legal only under a specific factorization of the expression, could either increase or decrease the operation count with respect to sharing elimination.
+\item \textit{Finding the global optimum.} Consider a form comprising two monomials $m1$ and $m2$.
 Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g.
@@
 If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost.
-We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction.
-In other words, we seek a strategy for finding the global optimum.
-\end{itemize}
+We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
+\end{enumerate}
 
-For the moment, let us assume that we own a cost function $\theta : M \rightarrow \mathbb{Z}$ that, given a monomial $m \in M$, returns the number of operations saved or introduced if pre-evaluation were applied to $m$.
-$\theta(m) < 0$ indicates that pre-evaluation is not profitable, because the operation count would actually increase by $|\theta(m)|$.
-Analogous considerations stand for $\theta(m) > 0$.
-We will show how to construct a proper $\theta$ in Section~\ref{sec:op_count}.
-Having defined $\theta$, we can now elaborate on the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.
+Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss of pre-evaluation over sharing elimination.
+In particular, we have that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$.
+$\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying to sharing elimination and pre-evaluation, respectively.
+We return to the issue of constructing $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
+$\theta$ is used by the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.
 
-\begin{figure}
+%numbers=left
+\begin{figure}[h]
 \begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
+// Initialization
 M = $\lbrace$m | m is a monomial in the input expression$\rbrace$
 S = $\lbrace$m | $\theta(m)$ < 0$\rbrace$
-maybe_P = M $\setminus$ S
-B = $\lbrace$b | b is a bipartite graph from complete graph G = (maybe_P, E)$\rbrace$
-for each $b \in B$:
-  for each $(b_S, b_P) \in b$:
-    if memory_required($b_P$) $\geq$ $T_H $:
+P = M $\setminus$ S
+
+// Find optimum such that memory constraints are honoured
+B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
+for each b $\in$ B:
+  for each (b$_S$, b$_P$) $\in$ b:
+    if memory_required(b$_P$) $\geq$ $T_H$:
       continue
-    op_count[b] = \eta(S $\cup$ b_S) + $|\theta(b_P)|$
-for each expression:
-  remove sharing
+    op_count[b] = $\theta_{se}$(S $\cup$ b$_S$) + $\theta_{pre}$(b$_P$)
+(b$_S$, b$_P$) = min(op_count)
+
+// Apply the transformations
+pre_evaluate(P $\cup$ b$_P$)
+eliminate_sharing(S $\cup$ b$_S$ $\cup$ P $\cup$ b$_P$)
 \end{lstlisting}
@@
 \end{figure}
+
+The algorithm starts with splitting the monomials into two disjoint sets: the monomials that surely do not take advantage of pre-evaluation ($S$) and those that instead potentially benefit from it ($P$).
+Some subsets of monomials in $P$: 1) might share terms with $S$, so global sharing elimination could outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}.
+The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of pre-evaluated monomials and Constraint~\ref{const:TH} is honoured.
+Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of evaluating $\theta_{se}$ and $\theta_{pre}$ is negligible.
+This aspect is elaborated in Section~\ref{sec:op_count}.
+In addition, observe that because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary.
+Therefore, sharing elimination is transparently applied to all monomials (i.e., to all sub-expressions, even those produced by pre-evaluation).
+The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.
+
+\begin{figure}\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
+\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}
+
+In the following sections, we tie up the remaining loose ends: the construction of $\theta$ (Section~\ref{sec:op_count}); 2) the optimality of sharing elimination (Section~\ref{sec:check-se}).
 
@@
 
-Because of reuse of basis functions, pre-evaluation may produce identical tables, which will be mapped to the same temporary.
-The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing.
-
-The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.
-
-\begin{figure}\begin{CenteredBox}
-\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
-\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}
-
-Finally, with Proposition~\ref{prop:optimal-approach}, we show the optimality of this approach.
-
-\begin{Prop}
-\label{prop:optimal-approach}
-Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it.
-The form comprises a set of monomials, $M$.
-Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
-Assume that:
-\begin{enumerate}
-\item pre-evaluating different monomials does not result in identical tables;
-\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
-\end{enumerate} 
-Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
-\end{Prop}
-\begin{proof}
-We first comment on the assumptions.
-1) Inherently a pathological case, so we simply neglect it.
-2) Might occur in forms with several monomials, although this was never the case in the problems that we have so far investigated (which are obviously more than those presented in Section~\ref{sec:perf-results}).
-These assumptions allow us to avoid a ``global analysis'' of $M$.
-
-We prove our claim by demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$.
-
-A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$.
-If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim.
-Assumption 1) prevents this situation, so the statement is true.
-
-B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term.
-We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
-That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
-Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions.
-The previously discussed assumption 2) allows us to neglect this possibility.
-\end{proof}
-
-\subsection{Cost Functions}
-It remains to tie up one loose end: the construction of cost functions to 1) estimate the impact of pre-evaluating monomials (Section~\ref{sec:op_count}); 2) check that the default sharing elimination strategy is optimal (Section~\ref{sec:check-se}).
-Since the cost functions are expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.
-
-\subsubsection{Profitability of pre-evaluation}
+\subsection{The cost function $\theta$}
 \label{sec:op_count}
 We build the cost function $\theta$.
-According to what established in the previous section, we have $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$; we recall that $\theta_{wo}$ and $\theta_{w}$ represent the operation counts without and with pre-evaluation.
+We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and with pre-evaluation.
+Since $\theta$ is expected to be used by a compiler, requirements are simplicity and velocity.
 In the following, we explain how to derive these two values.
 
-We can easily predict $\theta_{wo}$ thanks to our key property, linearity.
-This was explained in Proposition~\ref{prop:optimal-mln}.
-A simple expression tree traversal suffices to obtain the cost of a  sharing-free outer product loop nest, namely $\Lambda_{ops}^{ns}$.
-Assuming $I$ to be the size of $L_i$, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.
+The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the transformations and count the number of operations.
+If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} is fast), the code generation time would significantly increase if we were to apply pre-evaluation (Algorithm~\ref{algo:pre-evaluation}, evaluation of $L_i$ through symbolic execution) multiple times.
+We then need an analytic way of determining $\theta_{pre}$.
 
-Determining $\theta_{w}$ is more complicated.
-We first need to estimate the \textit{increase factor}, $\iota$, to account for the presence of (derivatives of) coefficients.
+The first step consists of estimating the \textit{increase factor}, $\iota$.
 This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation.
 To contextualize, consider the example in Figure~\ref{code:increase_factor}.
-
-\begin{figure}[h]\begin{CenteredBox}
-\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
-\end{CenteredBox}\caption{Simplified loop nest for a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}
-
 One can think of this as the (simplified) loop nest originating from the local assembly of a pre-multiplied mass matrix.
@@
 By expanding the products, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$).
-Intuitively, $\iota$ captures the increase in non-hoistable operations due to the transformations that expose pre-evaluation opportunities.
+$\iota$ captures the increase in non-hoistable operations due to the transformations that expose pre-evaluation opportunities.
+
+\begin{figure}\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
+\end{CenteredBox}\caption{Simplified loop nest for a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}
 
@@
 For example, consider the case in which a monomial has multiple coefficients over the same function space.
-Product expansion would now lead to identical sub-expressions that, as such, would be mapped to the same temporary once pre-evaluated.
-The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that found with a more naive analysis.
+Product expansion would lead to identical sub-expressions, and $\iota$ must reflect this.
 For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
@@
 If $\iota \geq I$ we can immediately say that pre-evaluation will not be profitable.
-This is indeed a necessary condition that, intuitively, tells us that if we add to the innermost loop more operations than we actually save from eliminating $L_i$, then for sure $\theta_{wo} < \theta_{w}$.
-This observation can speed up the compilation time by decreasing the analysis cost.
+This is a necessary condition telling us that if we add to the innermost loop more operations than we actually save from eliminating $L_i$, then for sure $\theta_{se} < \theta_{pre}$.
+This observation can speed up the analysis cost.
 
-If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$.
-In particular, we need to calculate the number of terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$.
+If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{pre}$.
+In particular, we need to calculate the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
 Consider again Figure~\ref{code:increase_factor}.
@@
 
-\subsubsection{Optimality of sharing elimination}
+\subsection{Optimality of sharing elimination}
 \label{sec:check-se}
 aaa
+
+\subsection{Observations and Heuristics}
+\label{sec:obsheu}
+\begin{itemize}
+\item SPEED UP THETA: We can easily predict $\theta_{se}$ thanks to our key property, linearity.
+This was explained in Proposition~\ref{prop:optimal-mln}.
+A simple expression tree traversal suffices to obtain the cost of a  sharing-free outer product loop nest, namely $\Lambda_{ops}^{ns}$.
+Assuming $I$ to be the size of $L_i$, we have that $\theta_{se} = \Lambda_{ops}^{ns} \cdot I$.
+\item SPEED UP ALGO: most of the time, only a single bipartition...
+\item SPEED UP SHARING EL: stick to opt
+\end{itemize}
 

### 2015-10-28 TOMS_Paper/assembly.tex
Trying to fix sharing elimination, again...

@@
 From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-This particular form has led us to the key observation that, in the space $S$ of all possible sharing elimination strategies, the optimum ($opt$) is almost always reached through a specific sequence of transformations.
-Our approach is therefore as follows: we define families of sharing elimination strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the upper bound ($ub$) of $opt$;  if, as expected, $ub(\lbrace opt \rbrace) < lb(S \setminus \lbrace opt \rbrace)$, then we know that $opt$ is actually optimal.
+We make the following considerations:
 
-The idea behind $opt$ is that by grouping together test and trial functions sub-expressions independent of at least one loop (i.e., hoistable) are exposed.
-The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits code motion due to dependencies on loop indices.
-Algorithm~\ref{algo:opt} describes how $opt$ is derived.
-We postpone to Section~\ref{sec:check-se} the discussion on how we assess the actual optimality of $opt$.
+\begin{enumerate}
+\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces.
+By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$) and then any $b_ik \in \beta{eik}$) (or vice-versa), sub-expressions independent of one of the outer product loops are exposed.
+This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
+\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$).
+Hoisting and capturing these sub-expressions can make the operation count of outer loops lower than in point 1), but the gain (if any) is, in practice, always negligible (note: the operation count in the innermost loop does not vary).
+\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in pathological scenarios in which the loops degenerate, i.e.
+loops having an iteration space of size 1).
+\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have different implications. We distinguish two cases:
+\begin{itemize}
+\item Performed before points 1) or 2).
+This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) reduces to cases 1 or 2 (for example, in the case of simple forms with scalar coefficients).
+\item After points 1) or 2) are applied; for example, $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei} + ...))$ and $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei} + ...$.
+Here, geometry-dependent expressions can be split from integration-dependent ones through suitable factorization and code motion.
+This is generally fundamental to minimize the operation count in the reduction loop.
+\end{itemize}
+\end{enumerate}
+
+Based on these observations, in Algorithm~\ref{algo:opt} we describe a sequence of transformations that, in most cases, lead to optimal sharing elimination.
 
@@
 \end{Algo}
+
+%Our approach is therefore as follows: we define families of sharing elimination strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the actual cost of $opt$; if, as expected, $cost(opt) < lb(S \setminus \lbrace opt \rbrace)$, then we know $opt$ is actually optimal.
+
+%We postpone to Section~\ref{sec:check-se} the discussion on how we assess the actual optimality of $opt$.
+
+%In the best and simplest case -- that is, in forms where $\alpha_{eij}$ and $\beta{eik}$ are multiplied by a single, scalar coefficient -- there is just one $\sigma_{ei}$ identical to all terms in the expression.
+Its factorization
+
+%Code motion tends to break due to , no longer through factorization would not be hoistable due to dependencies on loop indices.
+For example, assuming $f_i \in \sigma_{ei}^{v}, \sigma_{ei}^{w}$ and $\upsilon_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$, we would have $f_i (\alpha_{eij}^{v} \beta_{eik}^{v} \upsilon_{ei}^{v} + \alpha_{eij}^{w} \beta_{eik}^{w} \upsilon_{ei}^{w})$
 
@@
 
-At this point, three main issues are still to be addressed: 
+At this point, two main issues are to be addressed: 
 \begin{enumerate}
-\item \textit{Ascertain the optimality of sharing elimination.} Recall that Algorithm~\ref{algo:opt} performs a potentially optimal sharing elimination strategy.
 \item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation, which is legal only under a specific factorization of the expression, could either increase or decrease the operation count with respect to sharing elimination.
@@
 The algorithm starts with splitting the monomials into two disjoint sets: the monomials that surely do not take advantage of pre-evaluation ($S$) and those that instead potentially benefit from it ($P$).
-Some subsets of monomials in $P$: 1) might share terms with $S$, so global sharing elimination could outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}.
-The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of pre-evaluated monomials and Constraint~\ref{const:TH} is honoured.
-Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of evaluating $\theta_{se}$ and $\theta_{pre}$ is negligible.
+Some subsets of monomials in $P$: 1) might share terms with $S$, in which case sharing elimination could globally outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}.
+The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of monomials for which pre-evaluation is globally optimal and Constraint~\ref{const:TH} is honoured.
+Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible.
 This aspect is elaborated in Section~\ref{sec:op_count}.
@@
 
-In the following sections, we tie up the remaining loose ends: the construction of $\theta$ (Section~\ref{sec:op_count}); 2) the optimality of sharing elimination (Section~\ref{sec:check-se}).
-
-% apply memory constraints: reduction to knapsack problem ... not implemented, we use greedy
-
-%If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution.
-The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.
-
-%\begin{Algo}[Optimality check of $opt$]
-%\label{algo:se-pick}
-%\normalfont
-%aaa
-%\end{Algo}
-%There are two main possibilities: 1) factorization of $L_i$-dependent terms (i.e., coefficients and their derivatives) followed by code motion of $L_e$-dependent terms; 2) factorization of both $L_i$- and $L_e$-dependent terms, followed by common sub-expression elimination.
-
-%A special case is when test and trial functions belong to the same function space: this induces common sub-expressions \textit{over different loops}, so a potentially winning strategy consists of hoisting both $\alpha_{eij}$ and $\beta_{eik}$ and exploiting the fact that $\alpha = \beta$.
----- If we note that, for some $w$, $\alpha = \beta$, we add the following strategy to $PPS$: hoist $\alpha_{eij}^{w}$ (or, equivalently, $\beta_{eik}^{w}$); replace common sub-expressions; minimize cost of $\sigma_{ei}$, which we discuss later.
+In the following section, we tie up the remaining loose end, the construction of the cost function $\theta$.
 
@@
 \label{sec:op_count}
-We build the cost function $\theta$.
 We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and with pre-evaluation.
@@
 The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the transformations and count the number of operations.
-If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} is fast), the code generation time would significantly increase if we were to apply pre-evaluation (Algorithm~\ref{algo:pre-evaluation}, evaluation of $L_i$ through symbolic execution) multiple times.
+If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} tends to have negligible cost), the code generation time would become unacceptable if we were to apply pre-evaluation (Algorithm~\ref{algo:pre-evaluation}, symbolic execution of $L_i$) to each possible bipartition.
 We then need an analytic way of determining $\theta_{pre}$.
@@
 To contextualize, consider the example in Figure~\ref{code:increase_factor}.
-One can think of this as the (simplified) loop nest originating from the local assembly of a pre-multiplied mass matrix.
+One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix.
 The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions (array $B$).
-In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities.
-By expanding the products, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$).
-$\iota$ captures the increase in non-hoistable operations due to the transformations that expose pre-evaluation opportunities.
+In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities.
+By product expansion, we can observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$).
+$\iota$ captures this increase in non-hoistable operations.
 
@@
 Product expansion would lead to identical sub-expressions, and $\iota$ must reflect this.
-For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
+To evaluate $\iota$, we then calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$).
 
@@
 
-\subsection{Optimality of sharing elimination}
-\label{sec:check-se}
-aaa
-
-\subsection{Observations and Heuristics}
+\subsection{Trading optimality for analysis cost}
 \label{sec:obsheu}

### 2015-10-28 TOMS_Paper/assembly.tex
Trying to fix sharing elimination, still...UFF

@@
 % Page heads
-\markboth{F. Luporini et al.}{On Optimality of Finite Element Integration}
+\markboth{F. Luporini et al.}{Optimality and Heuristics for Finite Element Integration}
 
 % Title portion
-\title{On Optimality of Finite Element Integration}
+\title{Optimality and Heuristics for Finite Element Integration}
 \author{Fabio Luporini
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. On Optimality of Finite Element Integration.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. Optimality and Heuristics for Finite Element Integration.}
 
@@
 
-From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
 We make the following considerations:
@@
 \item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces.
-By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$) and then any $b_ik \in \beta{eik}$) (or vice-versa), sub-expressions independent of one of the outer product loops are exposed.
+By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed.
 This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
 \item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$).
-Hoisting and capturing these sub-expressions can make the operation count of outer loops lower than in point 1), but the gain (if any) is, in practice, always negligible (note: the operation count in the innermost loop does not vary).
-\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in pathological scenarios in which the loops degenerate, i.e.
-loops having an iteration space of size 1).
-\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have different implications. We distinguish two cases:
-\begin{itemize}
-\item Performed before points 1) or 2).
-This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) reduces to cases 1 or 2 (for example, in the case of simple forms with scalar coefficients).
-\item After points 1) or 2) are applied; for example, $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei} + ...))$ and $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei} + ...$.
-Here, geometry-dependent expressions can be split from integration-dependent ones through suitable factorization and code motion.
-This is generally fundamental to minimize the operation count in the reduction loop.
-\end{itemize}
+Hoisting and capturing these sub-expressions may make the operation count of outer loops lower than in point 1), but the gain (if any) will be negligible (note: the operation count in the innermost loop does not vary).
+\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in a scenario in which loops degenerate, i.e.
+with iteration spaces of size 1).
+\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have multiple implications. We distinguish two cases:
+\begin{enumerate}
+\item The factorization occurs before points 1) or 2).
+This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) simply reduces to cases 1) or 2) (for example, in the case of forms with scalar coefficients).
+\item The factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$.
+Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization.
+If the loop trip count is extremely low, however, this strategy may be sub-optimal.
+For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$.
+It is true that factorizing $v_i$ and $w_i$ would expose $L_i$-independent sub-expressions, but this would actually result in increased operation count if, for instance, the size of $L_i$ were 1.
+\end{enumerate}
 \end{enumerate}
 
-Based on these observations, in Algorithm~\ref{algo:opt} we describe a sequence of transformations that, in most cases, lead to optimal sharing elimination.
+Based on these observations, in Algorithm~\ref{algo:opt} we explain the sequence of transformations for potentially optimal sharing elimination.
+The algorithm does not take into account the concerns raised in points 2) and 4b).
+To address them, the simplest solution consists of comparing the operation counts by generating the corresponding code variants.
+We will further comment on this in Section~\ref{sec:obsheu}.
 
@@
 Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
-
-If there exists $w_1$ and $w_2$ such that $\sigma_{ei}^{w_1} \neq \sigma_{ei}^{w_2}$, then we also need decide how to schedule operations in $L_i$.
-For the potential optimum, we factorize $L_i$-dependent terms, which promotes hoisting of $L_e$-dependent sub-expressions.
+Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted.
 \end{Algo}
-
-%Our approach is therefore as follows: we define families of sharing elimination strategies; we calculate a lower bound ($lb$) for their operation count; we calculate the actual cost of $opt$; if, as expected, $cost(opt) < lb(S \setminus \lbrace opt \rbrace)$, then we know $opt$ is actually optimal.
-
-%We postpone to Section~\ref{sec:check-se} the discussion on how we assess the actual optimality of $opt$.
-
-%In the best and simplest case -- that is, in forms where $\alpha_{eij}$ and $\beta{eik}$ are multiplied by a single, scalar coefficient -- there is just one $\sigma_{ei}$ identical to all terms in the expression.
-Its factorization
-
-%Code motion tends to break due to , no longer through factorization would not be hoistable due to dependencies on loop indices.
-For example, assuming $f_i \in \sigma_{ei}^{v}, \sigma_{ei}^{w}$ and $\upsilon_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$, we would have $f_i (\alpha_{eij}^{v} \beta_{eik}^{v} \upsilon_{ei}^{v} + \alpha_{eij}^{w} \beta_{eik}^{w} \upsilon_{ei}^{w})$
 
@@
 The pre-evaluation of a monomial introduces some critical issues:
-\begin{itemize}
+\begin{enumerate}
 \item In contrast to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
@@
 This could outweigh the gain due to pre-evaluation.
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
-\end{itemize}
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination inhibits pre-evaluation, whereas pre-evaluation generally expose further sharing elimination opportunities.
+\end{enumerate}
 
-We address these points in Section~\ref{sec:optimal-synthesis}.
+We expand on point 1) in the next section. We address points 2) and 3) in Section~\ref{sec:optimal-synthesis}.
 

### 2015-10-28 TOMS_Paper/assembly.tex
Trying to fix sharing elimination, still...UFF

@@
 
-In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (potentially) common sub-expression elimination.
+In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination.
 In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
@@
 \begin{enumerate}
-\item The factorization occurs before points 1) or 2).
+\item Factorization occurs in place of points 1) or 2).
 This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) simply reduces to cases 1) or 2) (for example, in the case of forms with scalar coefficients).
-\item The factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$.
+\item Factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$.
 Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization.

### 2015-10-29 TOMS_Paper/assembly.tex
Fixes to shar el. Also, trading opt for analysis cost

@@
 In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space.
+We consider outer product loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
 
@@
 
-In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination.
+In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination.
 In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
-However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned.
-In the following, we assume outer product loop nests of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$.
+However, we explain that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.
 
@@
 \item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$).
-Hoisting and capturing these sub-expressions may make the operation count of outer loops lower than in point 1), but the gain (if any) will be negligible (note: the operation count in the innermost loop does not vary).
-\item Since loop trip counts are never lower than 3 (P1 Lagrange elements on triangles), it is never the case that applying points 1) or 2) increases the operation count (it might have been the case, otherwise, in a scenario in which loops degenerate, i.e.
-with iteration spaces of size 1).
-\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) can have multiple implications. We distinguish two cases:
+Compared to point 1), by hoisting and capturing these sub-expressions the reduction loop operation count becomes smaller, although the gain tends to be negligible (the operation count in the innermost loop is not affected).
+\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, which is the case of P0 discontinuous elements), factorizing and hoisting as in points 1) or 2) systematically reduces the operation count.
+The reduction is proportional to the outer product loop nest extent.
+\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
 \begin{enumerate}
-\item Factorization occurs in place of points 1) or 2).
-This either (i) inhibits (decrease) code motion by producing non-factorizable, non-hoistable sub-expressions or (ii) simply reduces to cases 1) or 2) (for example, in the case of forms with scalar coefficients).
-\item Factorization occurs after points 1) or 2); for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$.
+\item Test/trial functions are yet to be factorized.
+This, in general, inhibits (decreases) code motion by producing non-factorizable, non-hoistable sub-expressions.
+In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2).
+\item Within $\sigma_{ei}$ itself; for example, after applying point 1) we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$.
 Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization.
-If the loop trip count is extremely low, however, this strategy may be sub-optimal.
-For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$.
-It is true that factorizing $v_i$ and $w_i$ would expose $L_i$-independent sub-expressions, but this would actually result in increased operation count if, for instance, the size of $L_i$ were 1.
+Similarly to point 3), however, with very low trip counts this strategy may be sub-optimal.
+For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient.
+It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this will actually increase the operation count when the size of $L_i$ is 1.
 \end{enumerate}
@@
 
-Based on these observations, in Algorithm~\ref{algo:opt} we explain the sequence of transformations for potentially optimal sharing elimination.
-The algorithm does not take into account the concerns raised in points 2) and 4b).
-To address them, the simplest solution consists of comparing the operation counts by generating the corresponding code variants.
-We will further comment on this in Section~\ref{sec:obsheu}.
+Algorithm~\ref{algo:opt} turns these observations into a sequence of steps leading to optimal sharing elimination.
 
-\begin{Algo}[Construction of the potential optimum]
+\begin{Algo}[Sharing elimination]
 \label{algo:opt}
@@
 The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
-
-The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
-We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms (it can easily be demonstrated, by contradiction, that one should start from the loop with the least number of unique terms in order to obtain the smallest operation count).
+The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
+We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms.
+The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$.
+It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n tc$ is smallest leads to the maximal operation count reduction.
+Assume then we start with $L_j$.
 Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression.
@@
 Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
-Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted.
+Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted, checking that this actually results in reduced operation count (see point 4b) above).
 \end{Algo}
@@
 For example, in the case of the bilinear form originating from a standard bi-dimensional Poisson equation, the reader could verify that after a suitable factorization we would have $\rho = 3$.
-There are several ways of determining $\rho$.
-The fastest would be to extract it from high-level analysis of form and discretization; for convenience, in our implementation we have algorithms that, based on analysis of the expression tree, project the output of monomial expansion and factorization, which in turn gives us $\rho$.
+To determine $\rho$, we simply apply Algorithm~\ref{algo:pre-evaluation} until the evaluation of the reduction (obviously, excluded).
 
 \subsection{Trading optimality for analysis cost}
+Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing the constraints on optimality.
 \label{sec:obsheu}
 \begin{itemize}
-\item SPEED UP THETA: We can easily predict $\theta_{se}$ thanks to our key property, linearity.
-This was explained in Proposition~\ref{prop:optimal-mln}.
-A simple expression tree traversal suffices to obtain the cost of a  sharing-free outer product loop nest, namely $\Lambda_{ops}^{ns}$.
-Assuming $I$ to be the size of $L_i$, we have that $\theta_{se} = \Lambda_{ops}^{ns} \cdot I$.
-\item SPEED UP ALGO: most of the time, only a single bipartition...
-\item SPEED UP SHARING EL: stick to opt
+\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts render certain factorizations sub-optimal.
+To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level.
+Since this situation is particularly rare, we can relieve the algorithm by avoiding these checks.
+\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible ways of scheduling monomials candidate to pre-evaluation.
+In extremely complex forms, this could increase the analysis cost.
+Despite we have never encountered such a situation, a possibility is to disable these checks and to apply the transformation to all candidates.
+In the worst case scenario, this leads to a local optimum.
+\item \textit{Construction of $\theta$.} Instead of determining $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by manipulating the expression tree, we can approximate these values exploiting the key property of outer product loop nests, linearity.
+The number of unique terms in a multilinear loop nest can be used to calculate an upper bound on the cost of the innermost loop, where the bulk of the computation naturally tends to be performed.
+For example, in Figure~\ref{code:multi_loopnest_a} the fact that only a single term depends on $L_j$, $b_j$, suggests that in the innermost loop no more than one multiplication will be performed once sharing is eliminated.
+Therefore, an approximation for $\theta_{se}$ is, for example, $\theta_{se} = n_{iterations} n_{unique}$, where $n_{iterations}$ is the total number of iterations, and $n_{unique}$ is the smallest number of unique terms along one the linear loops.
 \end{itemize}
+
+In our code generation system, described in the next section, we make use of all of these heuristics.
 

### 2015-10-29 TOMS_Paper/assembly.tex
A few small fixes

@@
 % Page heads
-\markboth{F. Luporini et al.}{Optimality and Heuristics for Finite Element Integration}
+\markboth{F. Luporini et al.}{On Optimality of Finite Element Integration}
 
 % Title portion
-\title{Optimality and Heuristics for Finite Element Integration}
+\title{On Optimality of Finite Element Integration}
 \author{Fabio Luporini
@@
 \affil{Imperial College London}
-Paul H.J. Kelly
+Paul H. J. Kelly
 \affil{Imperial College London}}
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. Optimality and Heuristics for Finite Element Integration.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2015. On Optimality of Finite Element Integration.}
 

### 2015-10-30 TOMS_Paper/assembly.tex
Improve cost function and trade opt sections

@@
 \begin{enumerate}
-\item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation, which is legal only under a specific factorization of the expression, could either increase or decrease the operation count with respect to sharing elimination.
+\item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
 \item \textit{Finding the global optimum.} Consider a form comprising two monomials $m1$ and $m2$.
@@
 
-Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss of pre-evaluation over sharing elimination.
-In particular, we have that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$.
-$\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying to sharing elimination and pre-evaluation, respectively.
-We return to the issue of constructing $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
-$\theta$ is used by the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.
+Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination.
+In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively.
+Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$.
+We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
+Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.
 
@@
 
-// Find optimum such that memory constraints are honoured
+// Find optimum such that memory constraints are honored
 B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
@@
 Some subsets of monomials in $P$: 1) might share terms with $S$, in which case sharing elimination could globally outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}.
-The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of monomials for which pre-evaluation is globally optimal and Constraint~\ref{const:TH} is honoured.
+The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of monomials for which pre-evaluation is globally optimal and Constraint~\ref{const:TH} is honored.
 Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible.
@@
 Therefore, sharing elimination is transparently applied to all monomials (i.e., to all sub-expressions, even those produced by pre-evaluation).
-The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.
+The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming as input the loop nest in Figure~\ref{code:loopnest}.
 
@@
 
-In the following section, we tie up the remaining loose end, the construction of the cost function $\theta$.
+In the following section, we tie up the remaining loose end: the construction of the cost function $\theta$.
 
@@
 \label{sec:op_count}
-We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and with pre-evaluation.
+We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and pre-evaluation.
 Since $\theta$ is expected to be used by a compiler, requirements are simplicity and velocity.
@@
 
-The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the transformations and count the number of operations.
-If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} tends to have negligible cost), the code generation time would become unacceptable if we were to apply pre-evaluation (Algorithm~\ref{algo:pre-evaluation}, symbolic execution of $L_i$) to each possible bipartition.
-We then need an analytic way of determining $\theta_{pre}$.
+The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations.
+If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} tends to have negligible cost), the overhead becomes unacceptable when applying pre-evaluation to all possible bipartitions (Algorithm~\ref{algo:pre-evaluation}, due to the symbolic evaluation of $L_i$).
+We then seek an analytic way of determining $\theta_{pre}$.
 
@@
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities.
-By product expansion, we can observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$).
-$\iota$ captures this increase in non-hoistable operations.
+By performing product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.
 
@@
 
-With a single coefficient, as we just saw, $\iota$ directly descends from the cost of product expansion.
-In general, however, predicting $\iota$ is less straightforward.
-For example, consider the case in which a monomial has multiple coefficients over the same function space.
-Product expansion would lead to identical sub-expressions, and $\iota$ must reflect this.
-To evaluate $\iota$, we then calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$).
+Despite determining $\iota$ is straightforward in simple scenarios (e.g., when the form has a single coefficient, as we just saw), in general we must account for repeated sub-expressions that, once pre-evaluated, would result in identical tables.
+This is the case, for example, of a monomial having more than one coefficient in the same function space.
+One can verify this by taking the example in Figure~\ref{code:increase_factor} and adding a second coefficient $g$ over the same basis functions $B$.
+We then use combinatorics to evaluate $\iota$.
+We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$).
 
-If $\iota \geq I$ we can immediately say that pre-evaluation will not be profitable.
-This is a necessary condition telling us that if we add to the innermost loop more operations than we actually save from eliminating $L_i$, then for sure $\theta_{se} < \theta_{pre}$.
+If $\iota \geq I$ (the extent of the reduction loop), we already know that pre-evaluation will not be profitable.
+The condition $\iota < I$ is necessary to obtain a reduction in operation count.
+Intuitively, if this were not true, it would mean that we would be introducing more operations than we would save from pre-evaluating $L_i$.
 This observation can speed up the analysis cost.
 
-If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{pre}$.
-In particular, we need to calculate the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
-Consider again Figure~\ref{code:increase_factor}.
-In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$.
-In general, $\rho$ varies with the discretization and the differential operators used in the form.
-For example, in the case of the bilinear form originating from a standard bi-dimensional Poisson equation, the reader could verify that after a suitable factorization we would have $\rho = 3$.
-To determine $\rho$, we simply apply Algorithm~\ref{algo:pre-evaluation} until the evaluation of the reduction (obviously, excluded).
+With $\iota$ in place, it remains to find the total number of terms, $\rho$, that will be affected by the increase factor.
+Then we have $\theta_{pre} = \rho \cdot \iota$.
+Consider again the integration for the mass matrix in Figure~\ref{code:increase_factor}.
+This form is characterized by the dot product of test and trial functions (\texttt{B[j]*B[k]}), so here, trivially, $\rho = 1$.
+In general, $\rho$ depends on both form and discretization.
+As another example, take a standard Poisson equation using P1 Lagrange elements in 2D: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms.
+We determine $\rho$ by executing parts of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes the reduction pre-evaluation.
 
 \subsection{Trading optimality for analysis cost}
-Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing the constraints on optimality.
+Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing optimality.
 \label{sec:obsheu}
 \begin{itemize}
-\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts render certain factorizations sub-optimal.
+\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts may render certain factorizations sub-optimal.
 To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level.
-Since this situation is particularly rare, we can relieve the algorithm by avoiding these checks.
-\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible ways of scheduling monomials candidate to pre-evaluation.
-In extremely complex forms, this could increase the analysis cost.
-Despite we have never encountered such a situation, a possibility is to disable these checks and to apply the transformation to all candidates.
-In the worst case scenario, this leads to a local optimum.
-\item \textit{Construction of $\theta$.} Instead of determining $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by manipulating the expression tree, we can approximate these values exploiting the key property of outer product loop nests, linearity.
-The number of unique terms in a multilinear loop nest can be used to calculate an upper bound on the cost of the innermost loop, where the bulk of the computation naturally tends to be performed.
-For example, in Figure~\ref{code:multi_loopnest_a} the fact that only a single term depends on $L_j$, $b_j$, suggests that in the innermost loop no more than one multiplication will be performed once sharing is eliminated.
-Therefore, an approximation for $\theta_{se}$ is, for example, $\theta_{se} = n_{iterations} n_{unique}$, where $n_{iterations}$ is the total number of iterations, and $n_{unique}$ is the smallest number of unique terms along one the linear loops.
+Since such increases are, in practice, rare and negligible, the algorithm can be relieved by avoiding these checks.
+\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints.
+Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials.
+This matter is yet to be properly investigated.
+However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
+\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity.
+The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed).
+For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated.
+An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
 \end{itemize}
 
-In our code generation system, described in the next section, we make use of all of these heuristics.
+These heuristics have been implemented in the code generation system described in Section~\ref{sec:codegen}.
 
@@
 \label{sec:codegen}
-The transformation algorithm described in Section~\ref{sec:optimal-synthesis} has been implemented in COFFEE, the optimizer of finite element integration routines used in Firedrake.
-In this section, we describe several aspects related to code generation.
+Sharing elimination, pre-evaluation, the transformation algorithm, and all other features presented in Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} have been implemented in COFFEE, the optimizer for finite element integration routines used in Firedrake.
+In this section, we discuss implementation and code generation.
 

### 2015-10-31 TOMS_Paper/assembly.tex
Intermediate step with potentially useful sentences

@@
 \end{itemize}
-In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space.
+In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such transformation space.
 We consider outer product loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
@@
 In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
-However, we explain that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.
+However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.
 
-From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+Note that this respects Definition~\ref{def:fem-loopnest}.
 We make the following considerations:
@@
 \begin{enumerate}
-\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces.
-By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed.
-This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
-\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$).
-Compared to point 1), by hoisting and capturing these sub-expressions the reduction loop operation count becomes smaller, although the gain tends to be negligible (the operation count in the innermost loop is not affected).
-\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, which is the case of P0 discontinuous elements), factorizing and hoisting as in points 1) or 2) systematically reduces the operation count.
-The reduction is proportional to the outer product loop nest extent.
-\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
+\item In the special case in which test and trial functions are in the same function space, there exists $w1$ and $w2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$.
+That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$).
+By hoisting $\alpha$ or $\beta$ at the level of the closest common loop ($L_i$), the operation count is reduced.
+\item Let us now consider the general case in which test and trial functions belong to different function spaces.
+By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).
+Similarly to point 1), such a factorization aims to minimize the operation count ``locally'' at the level of the outer product loop nest.
+This transformation is independent of the chosen function spaces, but it may not be optimal in case point 1) is applicable (a common sub-expression elimination opportunity is lost).
+Two observations follow:
 \begin{enumerate}
-\item Test/trial functions are yet to be factorized.
-This, in general, inhibits (decreases) code motion by producing non-factorizable, non-hoistable sub-expressions.
+\item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
+\item However, the difference between the two tends to be negligible, since the outer product loop nests eventually perform the same number of operations.
+\end{enumerate}
+\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, the case of P0 discontinuous elements), points 1) and 2) always result in reduced operation count.
+\item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
+\begin{enumerate}
+\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form can be reduced to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$.
+\item There are no common symbols among the terms. Factorizing any $f_i \in \sigma_{ei}$ would then
+
+\item The starting point of the factorization is the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. Since the symbols in $\sigma_{ei}$ are generally not common factors
+
+In this case, the production of non-factorizable, non-hoistable sub-expressions tends to drastically inhibit code motion.
 In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2).
-\item Within $\sigma_{ei}$ itself; for example, after applying point 1) we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$.
+In other words, this is a ``non-natural factorization'', which tries to group potentially non-common factors and does not take advantage of loop dependencies.
+\item Point 1) was applied -- for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$ -- and we factorize symbols within $\sigma_{ei}$.
 Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization.
@@
 For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient.
-It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this will actually increase the operation count when the size of $L_i$ is 1.
+It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this may increase the operation count when the size of $L_i$ is 1.
 \end{enumerate}
@@
 A fourth operator impacting the operation count is \textit{reduction pre-evaluation}.
-In this section, we discuss its role and show that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.
+In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.
 

### 2015-10-31 TOMS_Paper/assembly.tex
Again, trying to improve section 2. non si vede la fine

@@
 In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
-However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.
+However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space is actually very small.
 
 Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-Note that this respects Definition~\ref{def:fem-loopnest}.
-We make the following considerations:
+Note that this normal form respects Definition~\ref{def:fem-loopnest}.
+We use the normal form for the following considerations:
 
@@
 \item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
-\item However, the difference between the two tends to be negligible, since the outer product loop nests eventually perform the same number of operations.
+\item However, the difference between the two tends to be negligible, since the operation count within the outer product loop nest is identical.
 \end{enumerate}
-\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, the case of P0 discontinuous elements), points 1) and 2) always result in reduced operation count.
 \item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
 \begin{enumerate}
-\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form can be reduced to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$.
-\item There are no common symbols among the terms. Factorizing any $f_i \in \sigma_{ei}$ would then
-
-\item The starting point of the factorization is the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. Since the symbols in $\sigma_{ei}$ are generally not common factors
-
-In this case, the production of non-factorizable, non-hoistable sub-expressions tends to drastically inhibit code motion.
-In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2).
-In other words, this is a ``non-natural factorization'', which tries to group potentially non-common factors and does not take advantage of loop dependencies.
-\item Point 1) was applied -- for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$ -- and we factorize symbols within $\sigma_{ei}$.
-Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization.
-Similarly to point 3), however, with very low trip counts this strategy may be sub-optimal.
-For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient.
-It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this may increase the operation count when the size of $L_i$ is 1.
+\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$.
+\item There is no common symbol among terms.
+Then, factorizing any $f_i \in \sigma_{ei}$ \textit{across two terms} $w1$ \textit{and} $w2$ always increases the operation count with respect to points 1) or 2).
+This is because of two facts.
+Firstly, the number of operations at every iteration is larger.
+Secondly, sub-expressions that are not hoistable due to the inherent dependencies on loop indices are exposed.
+On the other hand, factorizing any $f_i \in \sigma_{ei}$ \textit{within a generic term} $w$ promotes hoisting of geometry-dependent sub-expressions at the level of $L_e$.
+\end{enumerate}
+\item Inherent in the previous points is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for maximizing code motion.
+The fact that expressions result from sequences of inner products and loop nests adhere to a specific structure allows us to define a rigid transformation process leading, in general, to operation count minimization.
+\item There only are a few corner circumstances in which point 4) decades. Briefly:
+\begin{enumerate}
+\item As already explained, the dichotomy between points 1) and 2).
+\item The case in which loops degenerate (i.e., their trip counts is 1), which makes code motion irrelevant and certain factorizations sub-optimal (e.g., the factorization $(a f_i + b g_i) (c f_i + d f_i)$ is optimal if and only if the size of $L_i$ is 1).
 \end{enumerate}
@@
 
-Algorithm~\ref{algo:opt} turns these observations into a sequence of steps leading to optimal sharing elimination.
+These considerations can easily be turned into a constructive proof showing that any loop nest produced by Algorithm~\ref{algo:opt} is optimal if and only if: (i) we restrict, as specified at the beginning of the section, to factorization, code motion and common sub-expression elimination (what we mean by this is also clarified in the next section); (ii) we neglect the pathological cases in point 5) above.
 
@@
 The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$.
-It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n tc$ is smallest leads to the maximal operation count reduction.
+It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction.
 Assume then we start with $L_j$.

### 2015-11-02 TOMS_Paper/assembly.tex
Guess what? Still on fuckin section 2

@@
 
-\subsection{Quadrature Mode}
+\subsection{Quadrature mode}
 Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$.
@@
 
-\subsection{Tensor Contraction Mode}
+\subsection{Tensor contraction mode}
 \label{sec:tc}
@@
 
-\subsection{Loop Nests and Optimality}
+\subsection{Loop nests and optimality}
 In order to make the article self-contained, we start with reviewing basic compiler terminology.
@@
 
-\subsection{Sharing Elimination}
+\subsection{Sharing elimination}
 \label{sec:sharing-elimination}
@@
 
-Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-Note that this normal form respects Definition~\ref{def:fem-loopnest}.
-We use the normal form for the following considerations:
+\subsubsection{Normal form}
+Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry depending on the differential operators employed.
+Note how the normal form respects Definition~\ref{def:fem-loopnest}.
+The following sections exploit the normal form to explain how we perform sharing elimination in finite element integration loop nests.
 
-\begin{enumerate}
-\item In the special case in which test and trial functions are in the same function space, there exists $w1$ and $w2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$.
+\subsubsection{Outer product loop nest}
+\label{sec:se-opln}
+In the special case in which test and trial functions are in the same function space, there exists $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$.
 That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$).
-By hoisting $\alpha$ or $\beta$ at the level of the closest common loop ($L_i$), the operation count is reduced.
-\item Let us now consider the general case in which test and trial functions belong to different function spaces.
-By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).
-Similarly to point 1), such a factorization aims to minimize the operation count ``locally'' at the level of the outer product loop nest.
-This transformation is independent of the chosen function spaces, but it may not be optimal in case point 1) is applicable (a common sub-expression elimination opportunity is lost).
-Two observations follow:
-\begin{enumerate}
-\item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
-\item However, the difference between the two tends to be negligible, since the operation count within the outer product loop nest is identical.
-\end{enumerate}
-\item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
-\begin{enumerate}
+Both $\alpha$ and $\beta$ can be hoisted at the level of the closest common loop ($L_i$) and the common sub-expressions be replaced with suitable temporaries.
+This has been detailed in~\citeN{Luporini}.
+
+In the general case, instead, test and trial functions belong to different function spaces.
+By factorizing, in sequence, any test function $a_{ij} \in \alpha_{eij}$ and then any trial function $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops can be hoisted (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).
+
+These transformations minimize the operation count \textit{within} the outer product loop nest.
+However, when both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count.
+The difference in operation count tends however to be negligible, since only the outer loops are affected.
+
+\subsubsection{Reduction and free order loop}
+\label{sec:se-rln}
+Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops.
+The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination.
+This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators.
+We distinguish two cases:
+\begin{itemize}
 \item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$.
-\item There is no common symbol among terms.
-Then, factorizing any $f_i \in \sigma_{ei}$ \textit{across two terms} $w1$ \textit{and} $w2$ always increases the operation count with respect to points 1) or 2).
-This is because of two facts.
-Firstly, the number of operations at every iteration is larger.
-Secondly, sub-expressions that are not hoistable due to the inherent dependencies on loop indices are exposed.
-On the other hand, factorizing any $f_i \in \sigma_{ei}$ \textit{within a generic term} $w$ promotes hoisting of geometry-dependent sub-expressions at the level of $L_e$.
-\end{enumerate}
-\item Inherent in the previous points is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for maximizing code motion.
-The fact that expressions result from sequences of inner products and loop nests adhere to a specific structure allows us to define a rigid transformation process leading, in general, to operation count minimization.
-\item There only are a few corner circumstances in which point 4) decades. Briefly:
-\begin{enumerate}
-\item As already explained, the dichotomy between points 1) and 2).
-\item The case in which loops degenerate (i.e., their trip counts is 1), which makes code motion irrelevant and certain factorizations sub-optimal (e.g., the factorization $(a f_i + b g_i) (c f_i + d f_i)$ is optimal if and only if the size of $L_i$ is 1).
-\end{enumerate}
-\end{enumerate}
+\item There is no common symbol among all terms.
+There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost.
+On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion.
+\end{itemize}
 
-These considerations can easily be turned into a constructive proof showing that any loop nest produced by Algorithm~\ref{algo:opt} is optimal if and only if: (i) we restrict, as specified at the beginning of the section, to factorization, code motion and common sub-expression elimination (what we mean by this is also clarified in the next section); (ii) we neglect the pathological cases in point 5) above.
+\subsubsection{Observations and underlying assumption}
+\label{sec:se-assumptions}
+What we are advocating is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for minimizing the operation count.
+The idea is that sub-expressions must be placed within the loops that they logically belong to (and not in a deeper level), and this can only be accomplished by systematic factorization and code motion.
+
+There only are two pathological circumstances in which this approach might not lead to optimality: 
+\begin{itemize}
+\item As already explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes two alternative transformations.
+\item The loops degenerate (i.e., their trip count is 1).
+This makes the factorizations described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} sub-optimal since code motion is now irrelevant.
+For example, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ and factorizing $L_i$-dependent symbols is optimal if and only if the size of $L_i$ is strictly greater than 1 (the sub-expression was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient).
+\end{itemize}
+
+\subsubsection{Algorithm}
+Algorithm~\ref{algo:opt} describes how we manipulate the input loop nest and expression for performing sharing elimination.
+Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} can be seen as a constructive proof of optimality under the (reasonable) assumptions in Section~\ref{sec:se-assumptions}.
 
@@
 The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
+
 We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms.
-The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$.
+The choice of what we should factorize first impacts the resulting operation count.
+It depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$.
 It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction.
-Assume then we start with $L_j$.
+
+Assume, without loss of generality, that we start with $L_j$.
 Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression.
@@
 Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
-Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted, checking that this actually results in reduced operation count (see point 4b) above).
+
+For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted.
+Every time a sub-expression is hoisted, we search and replace all common sub-expressions.
+This last step is probably practically useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
 \end{Algo}
@@
 \label{sec:pre-evaluation}
-Sharing elimination reduces the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination.
+Sharing elimination reduces the operation count by applying three operators: factorization, code motion and common sub-expression elimination.
 A fourth operator impacting the operation count is \textit{reduction pre-evaluation}.
@@
 This would be challenging if we were to deal with arbitrary loop nests and expressions.
-However, we can make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
+However, we make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
 As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors.
@@
 
-\subsection{Transformation Algorithm}
+\subsection{Transformation algorithm}
 
@@
 
-\subsection{Automation through the COFFEE Language}
+\subsection{Automation through the COFFEE language}
 COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''.
@@
 
-\subsection{Independence from Form Compilers}
+\subsection{Independence from form compilers}
 \label{sec:zeros}
@@
 
-\subsection{Experimental Setup}
+\subsection{Experimental setup}
 
@@
 
-\subsection{Performance Results}
+\subsection{Performance results}
 \label{sec:perf-results-forms}

### 2015-11-04 TOMS_Paper/assembly.tex
Drop outer products. Revisit normal form. Minor fixes

@@
 \begin{itemize}
-\item We define optimality for finite element integration as number of flops given a memory bound; we build the space of legal transformations within which the optimal is sought.
+\item We formalize optimality for finite element integration and we build the space of legal transformations within which the optimal is sought.
 \item We provide an algorithm to select optimal points in the transformation space.
@@
 The local element vector $L$ can be determined in an analogous way.
+
+\subsection{Monomials}
+\label{sec:monomials}
+In general, it has been shown (e.g., in~\citeN{Kirby:TC}) that local element tensors can be expressed as a sum of integrals over $K$, each integral being the product of derivatives of functions from sets of discrete spaces and, possibly, functions of some spatially varying coefficients.
+An integral of this form is called \textit{monomial}.
+
 %From the computational perspective, its evaluation is however less expensive than that of $A$.
@@
 It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}).
-We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}.
+We will later comment on the efficacy of this approach (Section~\ref{sec:perf-results}).
 For the moment, we just recall that one of the goals of this research is to produce a system that goes beyond the dichotomy between quadrature and tensor modes.
@@
 \subsection{Loop nests and optimality}
+\label{sec:lnopt}
 In order to make the article self-contained, we start with reviewing basic compiler terminology.
@@
 
-Since our focus is on finite element integration, for simplicity we restrict ourselves to the following, notable subclass of perfect multilinear loop nests.
-
-\begin{Def}[Outer product loop nest]
-\label{def:jk-loop}
-An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs once in every summand.
-\end{Def}
-
-As shown later, outer product loop nests are important because they have a structure that we can take advantage of to synthesize optimal code.
+As shown later, multilinear loop nests are important because they have a structure that we can take advantage of to synthesize optimal code.
 We will see that they arise naturally when translating bilinear or linear forms into code.
 
-Two other classes of loops need be characterized.
+Two other classes of loops need be defined.
 
@@
 
-We use these definitions to formalize the class of loop nests for which we seek optimality.
+We build on these definitions to formalize the class of loop nests for which we seek optimality.
 
@@
 \label{def:fem-loopnest}
-A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest whose body is an independent basic block.
+A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block.
 \end{Def}
@@
 To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions).
-The expression \texttt{F} implements the operator.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions).
+In the body of $L_k$ we have one statement evaluating the local tensor for element $e$; more statements may be present if an element is a facet (instead of a cell, as we are assuming here).
+An expression implements a set of monomials.
+The expression we are showing is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that any expression from any problem can be reduced to this form.
 
@@
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
-\end{CenteredBox}\caption{The typical loop nest implementing a bilinear form.}\label{code:loopnest}\end{figure}
+\end{CenteredBox}\caption{The loop nest implementing a generic bilinear form.}\label{code:loopnest}\end{figure}
 
-We emphasize that our aim is a strategy to synthesize optimal loop nests of this specific form. In particular, we characterize optimality as follows.
+We emphasize that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.
 
@@
 In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such transformation space.
-We consider outer product loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
+We consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
 
@@
 {\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
-\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}
+\end{CenteredBox}\caption{Reducing a simple multilinear loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}
 
-For instance, Figure~\ref{code:multi_loopnest_a} shows a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
+For instance, Figure~\ref{code:multi_loopnest_a} shows a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
 The optimal synthesis for this loop nest is provided in Figure~\ref{code:multi_loopnest_b}.
@@
 
-In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination.
+In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination.
 In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
-However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space is actually very small.
+However, we show that by restricting ourselves to the class of loops and expressions constructed in Section~\ref{sec:lnopt}, the search space is actually very small.
 
-\subsubsection{Normal form}
-Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry depending on the differential operators employed.
-Note how the normal form respects Definition~\ref{def:fem-loopnest}.
-The following sections exploit the normal form to explain how we perform sharing elimination in finite element integration loop nests.
-
-\subsubsection{Outer product loop nest}
+\subsubsection{Multilinear loop nest}
 \label{sec:se-opln}
@@
 In the general case, instead, test and trial functions belong to different function spaces.
-By factorizing, in sequence, any test function $a_{ij} \in \alpha_{eij}$ and then any trial function $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops can be hoisted (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).
+By factorizing, in sequence, any test function $a_{ij} \in \alpha_{eij}$ and then any trial function $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the linear loops can be hoisted (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).
 
-These transformations minimize the operation count \textit{within} the outer product loop nest.
+These transformations minimize the operation count \textit{within} the multilinear loop nest.
 However, when both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count.
@@
 \begin{enumerate}
-\item In contrast to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item In contrast to what happens with hoisting in multilinear loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
@@
 One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix.
-The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions (array $B$).
+The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the field $f$ over (tabulated) basis functions (array $B$).
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities.
@@
 This is the case, for example, of a monomial having more than one coefficient in the same function space.
-One can verify this by taking the example in Figure~\ref{code:increase_factor} and adding a second coefficient $g$ over the same basis functions $B$.
+One can verify this by taking the example in Figure~\ref{code:increase_factor} and adding a second coefficient $g$ over the same basis functions $b$.
 We then use combinatorics to evaluate $\iota$.
-We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$).
+We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, $b_{i0}$, $b_{i1}$, and $b_{i2}$, so $n=3$).
 
@@
 Consider again the integration for the mass matrix in Figure~\ref{code:increase_factor}.
-This form is characterized by the dot product of test and trial functions (\texttt{B[j]*B[k]}), so here, trivially, $\rho = 1$.
+This form is characterized by the dot product of test and trial functions (\texttt{$b_{ij}$*$b_{ik}$}), so here, trivially, $\rho = 1$.
 In general, $\rho$ depends on both form and discretization.
-As another example, take a standard Poisson equation using P1 Lagrange elements in 2D: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms.
+As another example, take a standard Poisson equation using P1 Lagrange elements over triangles: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms.
 We determine $\rho$ by executing parts of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes the reduction pre-evaluation.
@@
 The sharing elimination performed by \texttt{cfO2} in each level of the loop nest ensures a critical reduction in operation count, which results is better execution times.
-In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the outer product loop nest.
+In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the multilinear loop nest.
 Clearly, the operation count increases with $q$, and so do the speed-ups.

### 2015-11-05 TOMS_Paper/assembly.tex
Minor fixes

@@
 
-\begin{Def}[Perfect multilinear loop nest]
-A perfect multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
+\begin{Def}[Multilinear loop nest]
+A multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
 \end{Def}
@@
 The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions).
-In the body of $L_k$ we have one statement evaluating the local tensor for element $e$; more statements may be present if an element is a facet (instead of a cell, as we are assuming here).
+In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$.
 An expression implements a set of monomials.
-The expression we are showing is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that any expression from any problem can be reduced to this form.
+The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that expressions from any problem can be reduced to this form.
 
@@
 
-We emphasize that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.
+We recall that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.
 
@@
 These transformations minimize the operation count \textit{within} the multilinear loop nest.
-However, when both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count.
-The difference in operation count tends however to be negligible, since only the outer loops are affected.
+When both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count.
+The difference in operation count, however, tends to be negligible, since only the outer loops are affected.
 
-\subsubsection{Reduction and free order loop}
+\subsubsection{Reduction and free order loops}
 \label{sec:se-rln}

### 2015-11-05 TOMS_Paper/assembly.tex
Minor fixes

@@
 
-\subsubsection{Observations and underlying assumption}
+\subsubsection{Observations and underlying assumptions}
 \label{sec:se-assumptions}
-What we are advocating is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for minimizing the operation count.
-The idea is that sub-expressions must be placed within the loops that they logically belong to (and not in a deeper level), and this can only be accomplished by systematic factorization and code motion.
-
-There only are two pathological circumstances in which this approach might not lead to optimality: 
+We observe that there almost never are two ore more conflicting factorizations among which to choose for minimizing the operation count.
+The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained.
+There only are two pathological circumstances in which such a reduction may not be obtained:
 \begin{itemize}
-\item As already explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes two alternative transformations.
-\item The loops degenerate (i.e., their trip count is 1).
-This makes the factorizations described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} sub-optimal since code motion is now irrelevant.
-For example, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ and factorizing $L_i$-dependent symbols is optimal if and only if the size of $L_i$ is strictly greater than 1 (the sub-expression was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient).
+\item As explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes sub-expressions over different loops, which introduces a choice between two alternative transformations.
+\item The trip count of a loop is extremely small (typically it needs be 1 for the following to happen).
+This decreases or nullifies the impact of code motion -- exposed through factorization, as described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln}.
+For instance, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ within a loop $L_i$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient} and factorizing $L_i$-dependent symbols leads to optimality if and only if the size of $L_i$ is strictly greater than 1.
 \end{itemize}
@@
 \subsubsection{Algorithm}
-Algorithm~\ref{algo:opt} describes how we manipulate the input loop nest and expression for performing sharing elimination.
-Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} can be seen as a constructive proof of optimality under the (reasonable) assumptions in Section~\ref{sec:se-assumptions}.
+Algorithm~\ref{algo:opt} details the steps for sharing elimination assuming as input an abstract syntax tree representation of the loop nest.
+Neglecting the pathological scenarios described in the previous section, the algorithm performs all necessary transformations and checks to reach optimality.
+A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln},~\ref{sec:se-rln} and~\ref{sec:se-assumptions}.
 
@@
 We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms.
-The choice of what we should factorize first impacts the resulting operation count.
-It depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$.
-It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction.
+The choice of starting with $L_j$ or $L_k$ depends on two parameters: the number of unique terms, $n_{L}$, and the trip count, $tc_{L}$, of a loop $L$.
+It is fairly simple to see (and easily demonstrable by contradiction) that taking first $L$ such that $n_L \cdot tc_L$ is smallest leads to the largest operation count reduction.
 
 Assume, without loss of generality, that we start with $L_j$.
-Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression.
-The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
+Thanks to linearity, each factored product only has one symbol depending on $L_j$, and this symbol is now unique in the expression tree.
+The other sub-expression, independent of $L_j$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
 Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
+We repeat the same process with $L_k$.
 
-For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted.
+For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions can be hoisted.
 Every time a sub-expression is hoisted, we search and replace all common sub-expressions.
-This last step is probably practically useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
+This last step is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
 \end{Algo}
@@
 In this section, we expand on this matter.
-The following two observations, in particular, lead to the definition of \textit{memory constraints}.
+The two following observations, in particular, lead to the definition of \textit{memory constraints}.
 
@@
 \begin{itemize}
-\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts may render certain factorizations sub-optimal.
-To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level.
-Since such increases are, in practice, rare and negligible, the algorithm can be relieved by avoiding these checks.
 \item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints.

### 2015-11-06 TOMS_Paper/assembly.tex
With Francis minor fixes

@@
 
-As shown later, multilinear loop nests are important because they have a structure that we can take advantage of to synthesize optimal code.
-We will see that they arise naturally when translating bilinear or linear forms into code.
+We will show that multilinear loop nests, which arise naturally when translating bilinear or linear forms into code, are important because they have a structure that we can take advantage of to synthesize optimal code.
 
-Two other classes of loops need be defined.
+We define two other classes of loops.
 
@@
 Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal.
 \end{Def}
@@
 \end{itemize}
-In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such transformation space.
+In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such a transformation space.
 We consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
@@
 \begin{itemize}
-\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$.
+\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
 \item There is no common symbol among all terms.
@@
 \label{sec:se-assumptions}
-We observe that there almost never are two ore more conflicting factorizations among which to choose for minimizing the operation count.
+We observe that there are almost never two ore more conflicting factorizations among which to choose for minimizing the operation count.
 The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained.
@@
 Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and Constraints~\ref{const:Le} and~\ref{const:TH} are satisfied.
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (sums, products) that it performs to evaluate the result is minimal.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal.
 \end{Def}

### 2015-11-09 TOMS_Paper/assembly.tex
Hopefully, this clarifies sharing elimination for L_i

@@
 Optimality is defined in terms of floating point operations given a memory bound.
-We provide an approach to explore the transformation space of loops and expressions typical of integration routines and discuss the conditions for optimality.
+We provide an approach to explore the transformation space of a generic integration routine and discuss the conditions for optimality.
 A theoretical analysis and extensive experimentation, which shows systematic performance improvements over a number of alternative code generation systems, validate the approach.
@@
 In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$.
-An expression implements a set of monomials.
+An expression (right hand side of a statement) implements a set of monomials.
 The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that expressions from any problem can be reduced to this form.
+We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and matrix operators, so $\sigma_{ei}$ can be arbitrarily complex.
+We observe that translating a monomial in matrix notation into code naturally results in this normal form; this applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}).
+Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests characterized by expressions in normal form.
 
-\begin{figure}[h]\begin{CenteredBox}
+\begin{figure}\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
@@
 
-We recall that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.
+Our aim is a strategy for reducing finite element integration loop nests to \textit{optimal form}. In particular, we characterize optimality as follows.
 
@@
 \label{sec:se-rln}
-Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops.
-The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination.
-This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators.
-We distinguish two cases:
-\begin{itemize}
+The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients).
+We distinguish the following cases.
+\begin{enumerate}
 \item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
-\item There is no common symbol among all terms.
-There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost.
-On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion.
-\end{itemize}
-
-\subsubsection{Observations and underlying assumptions}
-\label{sec:se-assumptions}
-We observe that there are almost never two ore more conflicting factorizations among which to choose for minimizing the operation count.
-The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained.
-There only are two pathological circumstances in which such a reduction may not be obtained:
-\begin{itemize}
-\item As explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes sub-expressions over different loops, which introduces a choice between two alternative transformations.
-\item The trip count of a loop is extremely small (typically it needs be 1 for the following to happen).
-This decreases or nullifies the impact of code motion -- exposed through factorization, as described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln}.
-For instance, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ within a loop $L_i$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient} and factorizing $L_i$-dependent symbols leads to optimality if and only if the size of $L_i$ is strictly greater than 1.
-\end{itemize}
+This is the simplest scenario since factorizing $f_i$ has no side effects.
+\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$.
+Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
+\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions.
+This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$.
+We provide an example.
+Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}.
+We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1.
+We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required.
+As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
+\end{enumerate}
 
 \subsubsection{Algorithm}
-Algorithm~\ref{algo:opt} details the steps for sharing elimination assuming as input an abstract syntax tree representation of the loop nest.
-Neglecting the pathological scenarios described in the previous section, the algorithm performs all necessary transformations and checks to reach optimality.
-A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln},~\ref{sec:se-rln} and~\ref{sec:se-assumptions}.
+Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on an abstract syntax tree representation of the loop nest.
+For simplicity, the algorithm neglects the pathological scenario described in Section~\ref{sec:se-opln}, in which the resulting operation count could only be very close to optimality when both test and trial functions are in the same function space.
+A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} (in which we have explored the entire factorization space).
 
@@
 \normalfont
-Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$.
-The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
-The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
+Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
 
+The algorithm starts with ``flattening'' the sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ (by product expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
 We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms.
@@
 It is fairly simple to see (and easily demonstrable by contradiction) that taking first $L$ such that $n_L \cdot tc_L$ is smallest leads to the largest operation count reduction.
-
 Assume, without loss of generality, that we start with $L_j$.
@@
 
-For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions can be hoisted.
-Every time a sub-expression is hoisted, we search and replace all common sub-expressions.
-This last step is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
+For each of the expressions $\eta_{ei} = \sum_w \sigma_{ei}^{w}$ lifted at the level of $L_i$, we need understand whether applying a similar process -- that is, factorizing $L_i$-dependent symbols such that $L_e$-dependent sub-expressions can be hoisted -- actually results in reduced operation count (the problem described in Section~\ref{sec:se-rln}, point 3).
+Consider two sub-expressions $a_{ei}$ and $b_{ei}$ in $\eta_{ei}$, with $n$ and $m$ being the respective number of summands, and with $c$ being the number of summands in common.
+Given a product $a_{ei} b_{ei}$, the increase in operation count after product expansion is $n \cdot m - \binom c2$.
+Given a sum $a_{ei} + b_{ei}$, the operation count after factorization of common terms is $n + m - c$.
+We use these simple formulae, together with the trip count of $L_i$, to predict the impact -- and then decide on the application -- of sharing elimination at the level of $L_i$.
+
+In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions.
+This is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
+
+% matrix notation, fixed normal form, so no need for fancy algors to do CSE
 \end{Algo}
@@
 A fourth operator impacting the operation count is \textit{reduction pre-evaluation}.
-In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.
+In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) making applying this operator legal.
 
@@
 
-With $\iota$ in place, it remains to find the total number of terms, $\rho$, that will be affected by the increase factor.
-Then we have $\theta_{pre} = \rho \cdot \iota$.
-Consider again the integration for the mass matrix in Figure~\ref{code:increase_factor}.
-This form is characterized by the dot product of test and trial functions (\texttt{$b_{ij}$*$b_{ik}$}), so here, trivially, $\rho = 1$.
+It remains to find the total number of terms, $\rho$, that will be affected by the increase factor; that is, $\theta_{pre} = \rho \cdot \iota$.
+Consider again the mass matrix in Figure~\ref{code:increase_factor}.
+This monomial is characterized by the dot product of test and trial functions (\texttt{$b_{ij}$*$b_{ik}$}), so here, trivially, $\rho = 1$.
 In general, $\rho$ depends on both form and discretization.

### 2015-11-23 TOMS_Paper/assembly.tex
BIG CHANGE IN PROGRESS: towards opt + quasi-opt

@@
 However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-In this paper, we formulate an approach to solve this problem.
+In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (``quasi-optimality'') in all other cases.
 Summarizing, our contributions are as follows:
@@
 \item We formalize optimality for finite element integration and we build the space of legal transformations within which the optimal is sought.
-\item We provide an algorithm to select optimal points in the transformation space.
-The algorithm uses a cost model to: 1) understand whether a transformation reduces or increases the operation count; 2) choose between different (non-composable) transformations.
+\item We provide an algorithm to select points in the transformation space.
+The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
+\item We explain under what conditions our algorithm leads to optimality.
+In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) in general, quasi-optimality is at least achieved (i.e., operation count optimal for innermost loops, close-to-optimal for outer loops).
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 \label{sec:optimal-impl}
-
 In this section, we characterize optimality for finite element integration and the transformation space that we need to explore to achieve it.
@@
 An expression (right hand side of a statement) implements a set of monomials.
-The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and matrix operators, so $\sigma_{ei}$ can be arbitrarily complex.
-We observe that translating a monomial in matrix notation into code naturally results in this normal form; this applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}).
-Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests characterized by expressions in normal form.
+The expression shown is in \textit{canonical form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex.
+We observe that translating a monomial in matrix notation into code naturally leads to the canonical form.
+This applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}).
+Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests with expressions in canonical form.
 
@@
 
-Our aim is a strategy for reducing finite element integration loop nests to \textit{optimal form}. In particular, we characterize optimality as follows.
+For this class of loop nests and expressions, we characterize optimality as follows.
 
@@
 
-To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
+Now that we have established the concepts of optimality and finite element integration loop nest, we can elaborate on goals and results of this article.
+The main objective is a strategy to systematically achieve optimality -- as formalized in Definition~\ref{def:mln-optimality} -- in classes of finite element integration loop nests -- as formalized in Definition~\ref{def:fem-loopnest}.
+For instance, a class of loop nests may be characterized by the use of only certain differential or tensor operators, or by certain properties in the resulting expressions.
+What our approach manages to achieve is: (i) optimality whenever the $\sigma_{ei}$  expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
+With quasi-optimality we mean that the operation count is minimal in the innermost loop and close to the ideal one in all outer loops.
+
+Our approach needs:
 \begin{itemize}
 \item a characterization of the space of legal transformations impacting the operation count
-\item a cost model to select the optimal point in the transformation space
+\item an algorithm to select points in the transformation space
 \end{itemize}
-In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such a transformation space.
-We consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
+In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space and on understanding the conditions for optimality.
+We will show that whenever an expression (i.e., its monomials) is ``too unstructured'', optimality may not be reached (unless, obviously, using brute-force based approaches).
+We will consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
 
@@
 
-In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination.
-In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
-However, we show that by restricting ourselves to the class of loops and expressions constructed in Section~\ref{sec:lnopt}, the search space is actually very small.
+In this section, we study \textit{sharing elimination}, a transformation that aims to minimize the operation count by applying factorization, code motion and (if necessary) common sub-expression elimination to remove sharing.
+In generic problems, transformations of this sort usually require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
+However, we show that since we restrict ourselves to a specific class of loops and expressions (described in Section~\ref{sec:lnopt}), we can formulate a simple algorithm ensuring quasi-optimality in general and optimality when expressions have a particular structure (typical of many finite element problems).
 
@@
 \begin{enumerate}
-\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
+\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
 This is the simplest scenario since factorizing $f_i$ has no side effects.
@@
 \normalfont
-Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
+Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in the canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
 
@@
 
-% matrix notation, fixed normal form, so no need for fancy algors to do CSE
+% matrix notation, fixed canonical form, so no need for fancy algors to do CSE
 \end{Algo}
@@
 \label{sec:optimal-synthesis}
-In this section, we build a transformation algorithm capable of reaching the goal set by Definition~\ref{def:optimality} for the class of loop nests in Definition~\ref{def:fem-loopnest}.
+In this section, we build a transformation algorithm capable of reaching optimality (Definition~\ref{def:optimality}) in a class of finite element integration loop nests (Definition~\ref{def:fem-loopnest}) and quasi-optimality in any other finite element integration loop nest.
 In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.

### 2015-11-26 TOMS_Paper/assembly.tex
Several minor fixes here and there...

@@
 \item We explain under what conditions our algorithm leads to optimality.
-In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) in general, quasi-optimality is at least achieved (i.e., operation count optimal for innermost loops, close-to-optimal for outer loops).
+In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, at least, always achieved (i.e., the operation count is at least optimal in innermost loops).
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 
-Note that Definition~\ref{def:mln-optimality} does not take into account memory requirements.
+\begin{Def}[Quasi-optimality of a loop nest]
+\label{def:mln-quasi-optimality}
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is at least minimal in the innermost loop.
+\end{Def}
+
+Note that Definitions~\ref{def:mln-optimality} and~\ref{def:mln-quasi-optimality} do not take into account memory requirements.
 If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
@@
 In the context of finite element, this is often true for more complex multilinear forms and/or higher order elements.
+We also note that quasi-optimality approximates optimality very well whenever the inner loop trip counts are ``sufficiently large''.
 
-Now that we have established the concepts of optimality and finite element integration loop nest, we can elaborate on goals and results of this article.
-The main objective is a strategy to systematically achieve optimality -- as formalized in Definition~\ref{def:mln-optimality} -- in classes of finite element integration loop nests -- as formalized in Definition~\ref{def:fem-loopnest}.
-For instance, a class of loop nests may be characterized by the use of only certain differential or tensor operators, or by certain properties in the resulting expressions.
-What our approach manages to achieve is: (i) optimality whenever the $\sigma_{ei}$  expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
-With quasi-optimality we mean that the operation count is minimal in the innermost loop and close to the ideal one in all outer loops.
-
-Our approach needs:
+Achieving optimality in polynomial time is clearly not always possible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily complex.
+Multilinearity, on the other hand, ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions.
+In the following sections, we build an approach that achieves: (i) optimality whenever the $\sigma_{ei}$  expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
+In particular, we need:
 \begin{itemize}
-\item a characterization of the space of legal transformations impacting the operation count
+\item the space of legal transformations impacting the operation count
 \item an algorithm to select points in the transformation space
 \end{itemize}
-In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space and on understanding the conditions for optimality.
-We will show that whenever an expression (i.e., its monomials) is ``too unstructured'', optimality may not be reached (unless, obviously, using brute-force based approaches).
+In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space and we explain the conditions for optimality.
+We show that whenever an expression (i.e., its monomials) is ``too unstructured'', optimality may not be reached.
 We will consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
@@
 \label{sec:pre-evaluation}
-Sharing elimination reduces the operation count by applying three operators: factorization, code motion and common sub-expression elimination.
-A fourth operator impacting the operation count is \textit{reduction pre-evaluation}.
-In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) making applying this operator legal.
+Sharing elimination explores the transformation space and applies three operators: factorization, code motion and common sub-expression elimination.
+In this section, we discuss role and legality of a fourth operator: \textit{reduction pre-evaluation}.
+We will see that what makes this operator special is the fact that there exists a single point in a monomial transformation space (i.e., specific factorization and code motion) preserving the correctness of the transformation.
 
@@
 
-\subsection{Completeness}
+An analogous definition can be provided for quasi-optimality.
+
+\subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
 We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization.
@@
 
-Sharing elimination and reduction pre-evaluation tackle the problem of minimizing spatial and temporal redundancy in the loop nests and expressions of Definition~\ref{def:fem-loopnest}.
-The space of all possible factorizations is explored, and for each point in this space, code motion (restricted by Constraint~\ref{const:Le}), common sub-expression elimination, and reduction pre-evaluation opportunities (restricted by Constraint~\ref{const:TH}) are analyzed.
+Sharing elimination and pre-evaluation tackle the problem of minimizing spatial and temporal redundancy with a bounded working set in the loop nests and expressions defined in Section~\ref{sec:lnopt} \textit{using a very specific set of operators}.
+This must be kept in mind: obviously, the operation count might even be lower if we considered operators exploiting domain properties that we are here neglecting, such as redundancies in basis functions.
 
@@
 
-// Find optimum such that memory constraints are honored
+// Find potential optimum such that memory constraints are honored
 B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
@@
 One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix.
-The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the field $f$ over (tabulated) basis functions (array $B$).
+The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$).
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities.
@@
 This is the case, for example, of a monomial having more than one coefficient in the same function space.
-One can verify this by taking the example in Figure~\ref{code:increase_factor} and adding a second coefficient $g$ over the same basis functions $b$.
-We then use combinatorics to evaluate $\iota$.
+One can verify this by taking the example in Figure~\ref{code:increase_factor}, adding a second coefficient $g$ over the same basis functions $b$, and expanding the products.
+To evaluate $\iota$, we then use combinatorics.
 We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, $b_{i0}$, $b_{i1}$, and $b_{i2}$, so $n=3$).
@@
 If $\iota \geq I$ (the extent of the reduction loop), we already know that pre-evaluation will not be profitable.
-The condition $\iota < I$ is necessary to obtain a reduction in operation count.
-Intuitively, if this were not true, it would mean that we would be introducing more operations than we would save from pre-evaluating $L_i$.
-This observation can speed up the analysis cost.
+Intuitively, this means that we are introducing more operations than we are saving from pre-evaluating $L_i$.
 
-It remains to find the total number of terms, $\rho$, that will be affected by the increase factor; that is, $\theta_{pre} = \rho \cdot \iota$.
+If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
 Consider again the mass matrix in Figure~\ref{code:increase_factor}.
-This monomial is characterized by the dot product of test and trial functions (\texttt{$b_{ij}$*$b_{ik}$}), so here, trivially, $\rho = 1$.
-In general, $\rho$ depends on both form and discretization.
-As another example, take a standard Poisson equation using P1 Lagrange elements over triangles: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms.
-We determine $\rho$ by executing parts of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes the reduction pre-evaluation.
+This monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
+However, in general, $\rho$ depends on both form and discretization.
+If we instead take a standard Poisson equation using P1 Lagrange elements over triangles we have, for example, $\rho = 3$ (this can be verified by factorizing derivatives of basis functions as explained in Algorithm~\ref{algo:pre-evaluation}).
+To determine $\rho$ we execute the initial phase of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes reduction pre-evaluation.
 

### 2015-11-26 TOMS_Paper/assembly.tex
Start attacking, again, sharing elimination...

@@
 
-In this section, we study \textit{sharing elimination}, a transformation that aims to minimize the operation count by applying factorization, code motion and (if necessary) common sub-expression elimination to remove sharing.
-In generic problems, transformations of this sort usually require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies.
-However, we show that since we restrict ourselves to a specific class of loops and expressions (described in Section~\ref{sec:lnopt}), we can formulate a simple algorithm ensuring quasi-optimality in general and optimality when expressions have a particular structure (typical of many finite element problems).
+In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by applying factorization, code motion and (if necessary) common sub-expression elimination to remove sharing.
+If optimality is the target and expressions are generally unstructured, transformations of this sort require solving large combinatorial problems, for instance to evaluate the impact of all possible factorization strategies.
+In our domain, however, expressions may be characterized by some properties.
+These properties hold either in general -- as inherent in finite element integration -- or in forms making ``special use'' of most common differential and tensor operators.
+We exploit them to build an algorithm that results in optimal or quasi-optimal sharing elimination depending on the problem to which it is applied.
+In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss these properties, while in Section~\ref{sec:sec:se-algo} we present the algorithm.
 
 \subsubsection{Multilinear loop nest}
-\label{sec:se-opln}
+\label{sec:se-mln}
 In the special case in which test and trial functions are in the same function space, there exists $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$.
@@
 \subsubsection{Algorithm}
+\label{sec:sec:se-algo}
 Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on an abstract syntax tree representation of the loop nest.
-For simplicity, the algorithm neglects the pathological scenario described in Section~\ref{sec:se-opln}, in which the resulting operation count could only be very close to optimality when both test and trial functions are in the same function space.
-A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} (in which we have explored the entire factorization space).
+For simplicity, the algorithm neglects the pathological scenario described in Section~\ref{sec:se-mln}, in which the resulting operation count could only be very close to optimality when both test and trial functions are in the same function space.
+A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} (in which we have explored the entire factorization space).
 

### 2015-11-26 TOMS_Paper/assembly.tex
Guess what...Refactoring sharing elimination

@@
 
-\subsection{Qualitative Comparison}
+\subsection{Qualitative comparison}
 \label{sec:qualitative}
@@
 
-\subsection{Loop nests and optimality}
+\subsection{Loop nests, expressions and optimality}
 \label{sec:lnopt}
@@
 \begin{Def}[Sharing]
-A loop $L_i$ presents sharing if it contains at least one statement in which two (sub-)expressions depending on $i$ are symbolically identical.
+A statement within a loop nest $\Lambda$ presents sharing if at least one of the following conditions hold:
+\begin{enumerate}
+\item two expressions are symbolically identical (spatial sharing)
+\item a non-trivial expression (at least an addition or a product) depending on $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ is present (temporal sharing)
+\end{enumerate}
 \end{Def}
 
+To illustrate the definition, we show in Figure~\ref{code:multi_loopnest} how sharing evolves as factorization and code motion are applied to a trivial multilinear loop nest.
+In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by $b_j$.
+Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}).
+Finally, generalized code motion (\cite{Luporini}) leads to optimality (Figure~\ref{code:multi_loopnest_c}).
+
 \begin{figure}[h]\begin{CenteredBox}
-{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
-~~~~~~~~~~~~~~
-{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
+{\subfigcapskip = 7pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
+~~~~~
+{\subfigcapskip = 7pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
+~~~~~
+{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_c}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
 \end{CenteredBox}\caption{Reducing a simple multilinear loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}
 
-For instance, Figure~\ref{code:multi_loopnest_a} shows a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$.
-The optimal synthesis for this loop nest is provided in Figure~\ref{code:multi_loopnest_b}.
-This example illustrates how factorization, besides reducing the number of multiplications required at a given point in the loop nest, can expose code motion opportunities.
-
-In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by applying factorization, code motion and (if necessary) common sub-expression elimination to remove sharing.
-If optimality is the target and expressions are generally unstructured, transformations of this sort require solving large combinatorial problems, for instance to evaluate the impact of all possible factorization strategies.
-In our domain, however, expressions may be characterized by some properties.
+In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, code motion and (if necessary) common sub-expression elimination.
+If the objective is reaching optimality and expressions are generally unstructured, transformations of this sort require solving large combinatorial problems, for instance to evaluate the impact of all possible factorization strategies.
+In our domain, however, expressions may be characterized by certain properties.
 These properties hold either in general -- as inherent in finite element integration -- or in forms making ``special use'' of most common differential and tensor operators.
@@
 In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss these properties, while in Section~\ref{sec:sec:se-algo} we present the algorithm.
+We assume the usual finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in the canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
 
-\subsubsection{Multilinear loop nest}
+\subsubsection{Common sub-expressions}
 \label{sec:se-mln}
-In the special case in which test and trial functions are in the same function space, there exists $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$.
-That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$).
-Both $\alpha$ and $\beta$ can be hoisted at the level of the closest common loop ($L_i$) and the common sub-expressions be replaced with suitable temporaries.
-This has been detailed in~\citeN{Luporini}.
+A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE).
+This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that (i) are symbolically identical and (ii) reduce to the same value.
 
-In the general case, instead, test and trial functions belong to different function spaces.
-By factorizing, in sequence, any test function $a_{ij} \in \alpha_{eij}$ and then any trial function $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the linear loops can be hoisted (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).
+We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block.
+More interestingly, we observe that in the ``special case'' (SCSE) in which test and trial functions are in the same function space, there may exist $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$, in which case common sub-expressions will arise along different loops ($L_j$ and $L_k$).
+General-purpose compilers cannot detect this (very domain-specific) property.
+Our algorithm for sharing elimination will exploit this extra information.
 
-These transformations minimize the operation count \textit{within} the multilinear loop nest.
-When both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count.
-The difference in operation count, however, tends to be negligible, since only the outer loops are affected.
+Both CSE and factorization target spatial sharing.
+We will see that in the cases in which factorization is found to increase the operation count (due to the need for preliminary transformations), CCSE will instead be employed to remove spatial sharing.
 
-\subsubsection{Reduction and free order loops}
+%TODO : if also the Ri loop is linear, then optimality full stop.
+
+\subsubsection{Structured inner products}
 \label{sec:se-rln}
-The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients).
+%Expressions depending on the linear loops $L_j$ and $L_k$ result from inner products involving test and trial functions (possibly, derivatives or vector-valued).
+The problem of reaching optimality (at least) at the level of the multilinear loop nest reduces, therefore, to understand how to optimally schedule these inner products.
+...
+...
+%The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients).
 We distinguish the following cases.
-\begin{enumerate}
-\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
+%\begin{enumerate}
+%\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
 This is the simplest scenario since factorizing $f_i$ has no side effects.
-\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$.
+%\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$.
 Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
-\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions.
+%\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions.
 This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$.
@@
 As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
-\end{enumerate}
+%\end{enumerate}
 
@@
 Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on an abstract syntax tree representation of the loop nest.
-For simplicity, the algorithm neglects the pathological scenario described in Section~\ref{sec:se-mln}, in which the resulting operation count could only be very close to optimality when both test and trial functions are in the same function space.
-A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} (in which we have explored the entire factorization space).
 
@@
 \normalfont
-Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in the canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
-
-The algorithm starts with ``flattening'' the sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ (by product expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
+%The algorithm starts with ``flattening'' the sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ (by product expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
 We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms.
@@
 We repeat the same process with $L_k$.
-
-For each of the expressions $\eta_{ei} = \sum_w \sigma_{ei}^{w}$ lifted at the level of $L_i$, we need understand whether applying a similar process -- that is, factorizing $L_i$-dependent symbols such that $L_e$-dependent sub-expressions can be hoisted -- actually results in reduced operation count (the problem described in Section~\ref{sec:se-rln}, point 3).
+%
+%For each of the expressions $\eta_{ei} = \sum_w \sigma_{ei}^{w}$ lifted at the level of $L_i$, we need understand whether applying a similar process -- that is, factorizing $L_i$-dependent symbols such that $L_e$-dependent sub-expressions can be hoisted -- actually results in reduced operation count (the problem described in Section~\ref{sec:se-rln}, point 3).
 Consider two sub-expressions $a_{ei}$ and $b_{ei}$ in $\eta_{ei}$, with $n$ and $m$ being the respective number of summands, and with $c$ being the number of summands in common.
@@
 We use these simple formulae, together with the trip count of $L_i$, to predict the impact -- and then decide on the application -- of sharing elimination at the level of $L_i$.
-
-In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions.
+%
+%In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions.
 This is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.

### 2015-11-30 TOMS_Paper/assembly.tex
trying to write structured tensor operations

@@
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is at least minimal in the innermost loop.
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is at least minimal in all innermost loops.
 \end{Def}
@@
 
-Achieving optimality in polynomial time is clearly not always possible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily complex.
+Achieving optimality in polynomial time is not generally possible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
 Multilinearity, on the other hand, ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions.
-In the following sections, we build an approach that achieves: (i) optimality whenever the $\sigma_{ei}$  expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
-In particular, we need:
+In the following, we formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
+In particular, we build:
 \begin{itemize}
-\item the space of legal transformations impacting the operation count
-\item an algorithm to select points in the transformation space
+\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
+\item an algorithm to select points in the transformation space (Sectionm~\ref{sec:optimal-synthesis})
 \end{itemize}
-In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space and we explain the conditions for optimality.
-We show that whenever an expression (i.e., its monomials) is ``too unstructured'', optimality may not be reached.
-We will consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.
 
@@
 In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, code motion and (if necessary) common sub-expression elimination.
-If the objective is reaching optimality and expressions are generally unstructured, transformations of this sort require solving large combinatorial problems, for instance to evaluate the impact of all possible factorization strategies.
-In our domain, however, expressions may be characterized by certain properties.
-These properties hold either in general -- as inherent in finite element integration -- or in forms making ``special use'' of most common differential and tensor operators.
-We exploit them to build an algorithm that results in optimal or quasi-optimal sharing elimination depending on the problem to which it is applied.
-In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss these properties, while in Section~\ref{sec:sec:se-algo} we present the algorithm.
-We assume the usual finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in the canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
+If the objective were reaching optimality and expressions were systematically unstructured, transformations of this sort would require solving large combinatorial problems -- for instance to evaluate the impact of all possible factorization strategies.
+Our sharing elimination strategy, however, attempts to exploit the structure that often characterizes finite element integration expressions to achieve quasi-optimality (after coordination with other transformations, as discussed in the following sections); optimality is achieved if stronger properties hold.
+
+In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss structural (i.e., inherent in finite element integration) and potential (i.e., due to the particular use of differential and vector operators) properties of our expressions, while in Section~\ref{sec:sec:se-algo} we present the sharing elimination algorithm.
+We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$\footnote{A similar reasoning applies to forms and loops that are not bilinear}.
 
@@
 This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that (i) are symbolically identical and (ii) reduce to the same value.
-
 We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block.
@@
 Our algorithm for sharing elimination will exploit this extra information.
+Although both CSE and factorization target spatial sharing, the latter may actually increase the operation count if preliminary transformations (e.g., expansion) are required.
+In such a case, CCSE should be preferred.
 
-Both CSE and factorization target spatial sharing.
-We will see that in the cases in which factorization is found to increase the operation count (due to the need for preliminary transformations), CCSE will instead be employed to remove spatial sharing.
+\subsubsection{Structured tensor operations}
+\label{sec:se-rln}
+Finite element expressions can be seen as sums of nested operations between tensors.
+A tensor (i.e., its entries) always depends on a specific subset of loops, and the set of loops is, obviously, always well-defined.
+A recurring scenario when multiplying two vectors together is that the optimal operation scheduling is to be found in a space of size 2.
 
+Consider a generic matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$.
+Let us assume that $y_{\Psi \Delta}$ itself is just an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$.
+As an example, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions.
+An entry in $y_{\Psi \Delta}$, therefore, is of the form $(a_{\Psi} f_{\Delta} + b_{\Psi} g_{\Delta}) (c_{\Psi} f_{\Delta} + d_{\Psi} g_{\Delta})$.
+To illustrate the concept, consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$, in which we have  $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$.
+To schedule this expression within $L_i$ we basically have two choices:
+
+Many analogous examples can easily be found.
+
+the expression $\dot(J^{-T} \nabla v, J^{-T} \nabla u)$ resulting from a Laplacian operator.
+
+the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the coordinates field Jacobian matrix with the derivative of a coefficient}.
+We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1.
+We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required.
+As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics
+
+One can find this pattern in a wide range of variational forms.
+In fact, all of the forms used for experimentation (Section~\ref{sec:perf-results}) -- and many others that we have tried beyond these -- are characterized by this property.
+
+In the base case, clearly, there is only one inner product between test and trial functions.
+In general, there may be multiple inner products, whose vectors  The problem of reaching optimality reduces to understand how to schedule these inner products in the most effective way
+
+multilinearity -> avoids ambiguity
 %TODO : if also the Ri loop is linear, then optimality full stop.
+we can relax linearity by talking of structured inner products which avoids ambiguity in collection
 
-\subsubsection{Structured inner products}
-\label{sec:se-rln}
 %Expressions depending on the linear loops $L_j$ and $L_k$ result from inner products involving test and trial functions (possibly, derivatives or vector-valued).
@@
 \subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
+\label{sec:completeness}
 We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization.
@@
 
-\section{Optimal Synthesis}
+\section{Implementation Synthesis}
 \label{sec:optimal-synthesis}
@@
 
-\subsection{Trading optimality for analysis cost}
-Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing optimality.
+\subsection{Trading operation count for analysis cost}
+We can relax the constraints on (quasi-)optimality to relieve the analysis phase. In this section, we discuss ways of achieving this.
 \label{sec:obsheu}

### 2015-11-30 TOMS_Paper/assembly.tex
Written a first draft of structured tensor ops

@@
 \begin{enumerate}
-\item two expressions are symbolically identical (spatial sharing)
-\item a non-trivial expression (at least an addition or a product) depending on $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ is present (temporal sharing)
+\item there are at least two symbolically identical sub-expressions (spatial sharing)
+\item there is at least a non-trivial sub-expression (an addition or a product) depending on $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
 \end{enumerate}
@@
 If the objective were reaching optimality and expressions were systematically unstructured, transformations of this sort would require solving large combinatorial problems -- for instance to evaluate the impact of all possible factorization strategies.
-Our sharing elimination strategy, however, attempts to exploit the structure that often characterizes finite element integration expressions to achieve quasi-optimality (after coordination with other transformations, as discussed in the following sections); optimality is achieved if stronger properties hold.
+Our sharing elimination strategy, however, attempts to exploit the structure that often characterizes finite element integration expressions to achieve quasi-optimality (after coordination with other transformations, as discussed in the following sections); optimality is achieved if stronger conditions hold.
 
@@
 \label{sec:se-rln}
-Finite element expressions can be seen as sums of nested operations between tensors.
-A tensor (i.e., its entries) always depends on a specific subset of loops, and the set of loops is, obviously, always well-defined.
-A recurring scenario when multiplying two vectors together is that the optimal operation scheduling is to be found in a space of size 2.
+Finite element expressions can be seen as composition of operations between tensors.
+When multiplying two tensors it often happens that the optimal scheduling strategy is to be searched in a space of size 2.
 
-Consider a generic matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$.
-Let us assume that $y_{\Psi \Delta}$ itself is just an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$.
-As an example, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions.
-An entry in $y_{\Psi \Delta}$, therefore, is of the form $(a_{\Psi} f_{\Delta} + b_{\Psi} g_{\Delta}) (c_{\Psi} f_{\Delta} + d_{\Psi} g_{\Delta})$.
-To illustrate the concept, consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$, in which we have  $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$.
-To schedule this expression within $L_i$ we basically have two choices:
+To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$.
+Forms are typically rich of operations of this kind, or equivalent.
+Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$.
+For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$.
+Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$.
+To schedule this expression, we basically have two options:
+\begin{enumerate}
+\item We leave the expression unmodified in the body of $L_i$; we only assign it to a temporary in the case it is a common sub-expression
+\item We notice the presence of spatial sharing. We can re-factorize the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
+\end{enumerate}
+The reader can verify that option 2) reduces the operation count with respect to 1) if and only if the size of $L_i$ is strictly greater than $c=1$.
+One can find this dichotomy in a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with values of $c$ varying with the operational cost inherent in removing spatial sharing.
 
-Many analogous examples can easily be found.
+In general, given a product of two sub-expressions in a finite element integration loop nest, we observe that whenever there is no ambiguity in the elimination of spatial sharing, the problem of finding an optimal scheduling reduces to the comparison of two alternatives.
+We note that this condition simply holds if a loop is guaranteed to be linear; in the next section, we will show how take advantage of this property to ensure quasi-optimal sharing elimination.
+Further, if the $\sigma_{ei}$ expressions are sufficiently regular, then the same approach can be applied to all expressions logically belonging to $L_i$.
+The regularity of $\sigma_{ei}$ is not as uncommon as it may seem at first glance: for instance, this property holds for many of the tensors that need be evaluated in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).
 
-the expression $\dot(J^{-T} \nabla v, J^{-T} \nabla u)$ resulting from a Laplacian operator.
+%
+%can make the factorization space explode 
+%
+%On the other hand, the possibly unstructured nature of the 
+%
+%In absence of further coefficient functions or if their use is such that  So optimal 
+%Due to the possibly unstructured nature of the $\sigma_{ei}$ expressions, Clearly, in general, the problem is much more complex
 
-the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the coordinates field Jacobian matrix with the derivative of a coefficient}.
-We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1.
-We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required.
-As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics
+%the expression $\dot(J^{-T} \nabla v, J^{-T} \nabla u)$ resulting from a Laplacian operator.
 
-One can find this pattern in a wide range of variational forms.
-In fact, all of the forms used for experimentation (Section~\ref{sec:perf-results}) -- and many others that we have tried beyond these -- are characterized by this property.
-
-In the base case, clearly, there is only one inner product between test and trial functions.
-In general, there may be multiple inner products, whose vectors  The problem of reaching optimality reduces to understand how to schedule these inner products in the most effective way
-
-multilinearity -> avoids ambiguity
+%multilinearity -> avoids ambiguity
 %TODO : if also the Ri loop is linear, then optimality full stop.
-we can relax linearity by talking of structured inner products which avoids ambiguity in collection
+%we can relax linearity by talking of structured inner products which avoids ambiguity in collection
 
@@
 The problem of reaching optimality (at least) at the level of the multilinear loop nest reduces, therefore, to understand how to optimally schedule these inner products.
-...
-...
-%The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients).
-We distinguish the following cases.
-%\begin{enumerate}
-%\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$.
-This is the simplest scenario since factorizing $f_i$ has no side effects.
-%\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$.
-Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
-%\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions.
-This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$.
-We provide an example.
-Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}.
-We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1.
-We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required.
-As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
-%\end{enumerate}
 

### 2015-12-03 TOMS_Paper/assembly.tex
Some adjustments here and there

@@
 We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms.
-Optimality is defined in terms of floating point operations given a memory bound.
-We provide an approach to explore the transformation space of a generic integration routine and discuss the conditions for optimality.
-A theoretical analysis and extensive experimentation, which shows systematic performance improvements over a number of alternative code generation systems, validate the approach.
+Optimality and quasi-optimality are defined in terms of floating point operations given a memory bound.
+We provide an approach to explore the space of legal transformations and discuss under what conditions optimality or quasi-optimality hold.
+A theoretical analysis and extensive experimentation, which shows systematic performance improvements over several state-of-the-art code generation systems, validate the approach.
 \end{abstract}
@@
 \label{sec:optimal-impl}
-In this section, we characterize optimality for finite element integration and the transformation space that we need to explore to achieve it.
+In this section, we characterize optimality and quasi-optimality for finite element integration as well as the space of legal transformations that we need to explore to achieve it.
+How the exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.
 
@@
 
-We build on these definitions to formalize the class of loop nests for which we seek optimality.
-
-\begin{Def}[Finite element integration loop nest]
-\label{def:fem-loopnest}
-A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block.
-\end{Def}
+\begin{figure}\begin{CenteredBox}
+\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
+\end{CenteredBox}\caption{The loop nest implementing a generic bilinear form.}\label{code:loopnest}\end{figure}
 
@@
 In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$.
-An expression (right hand side of a statement) implements a set of monomials.
-The expression shown is in \textit{canonical form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+Expressions (right hand side of a statement) result from the translation of a form in high level matrix notation into code.
+In particular, $m$ is the number of ``terms'' (a monomial can be implemented by one or more terms), $\alpha_{eij}$ ($\beta_{eik}$) represents the product of a coefficient function (e.g., the inverse Jacobian matrix for the change of coordinates) with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
 We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex.
-We observe that translating a monomial in matrix notation into code naturally leads to the canonical form.
-This applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}).
-Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests with expressions in canonical form.
+We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact (e.g., products have not been expanded yet, distinct monomials can still be identified, etc.).
+This bring us to formalize the class of loop nests for which we seek optimality.
 
-\begin{figure}\begin{CenteredBox}
-\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
-\end{CenteredBox}\caption{The loop nest implementing a generic bilinear form.}\label{code:loopnest}\end{figure}
+%TODO is "not expanded yet == still contracted" ?
 
-For this class of loop nests and expressions, we characterize optimality as follows.
+\begin{Def}[Finite element integration loop nest]
+\label{def:fem-loopnest}
+A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which each statement has expressions in normal form.
+\end{Def}
+
+For a finite element integration loop nest, we characterize optimality and quasi-optimality as follows.
 
@@
 
-Achieving optimality in polynomial time is not generally possible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
-Multilinearity, on the other hand, ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions.
-In the following, we formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
-In particular, we build:
+Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
+On the other hand, multilinearity ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions, which is reflected in the innermost loop.
+In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
+To this purpose, we will construct:
 \begin{itemize}
 \item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
-\item an algorithm to select points in the transformation space (Sectionm~\ref{sec:optimal-synthesis})
+\item an algorithm to select points in the transformation space (Section~\ref{sec:optimal-synthesis})
 \end{itemize}
@@
 \item there are at least two symbolically identical sub-expressions (spatial sharing)
-\item there is at least a non-trivial sub-expression (an addition or a product) depending on $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
+\item there is at least one non-trivial sub-expression (an addition or a product) that is redundantly executed as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
 \end{enumerate}
@@
 
-In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, code motion and (if necessary) common sub-expression elimination.
-If the objective were reaching optimality and expressions were systematically unstructured, transformations of this sort would require solving large combinatorial problems -- for instance to evaluate the impact of all possible factorization strategies.
-Our sharing elimination strategy, however, attempts to exploit the structure that often characterizes finite element integration expressions to achieve quasi-optimality (after coordination with other transformations, as discussed in the following sections); optimality is achieved if stronger conditions hold.
+In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, generalized code motion and (if necessary) common sub-expression elimination.
+If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorization strategies.
+Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This require coordination with other transformations, as discussed in the following sections.
+For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved whenever stronger conditions hold.
+Relaxing the problem is essential to produce simple and computationally efficient algorithms -- which, as such, can be integrated with a compiler.
 
-In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss structural (i.e., inherent in finite element integration) and potential (i.e., due to the particular use of differential and vector operators) properties of our expressions, while in Section~\ref{sec:sec:se-algo} we present the sharing elimination algorithm.
-We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$\footnote{A similar reasoning applies to forms and loops that are not bilinear}.
+In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties of our expressions, while in Section~\ref{sec:sec:se-algo} we present the sharing elimination algorithm.
+We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
 
@@
 Although both CSE and factorization target spatial sharing, the latter may actually increase the operation count if preliminary transformations (e.g., expansion) are required.
-In such a case, CCSE should be preferred.
+In such a case, CCSE will be preferred.
 
@@
 
-% matrix notation, fixed canonical form, so no need for fancy algors to do CSE
+% matrix notation, fixed normal form, so no need for fancy algors to do CSE
 \end{Algo}
@@
 
-\section{Implementation Synthesis}
+\section{Selection and composition of transformations}
 \label{sec:optimal-synthesis}
-In this section, we build a transformation algorithm capable of reaching optimality (Definition~\ref{def:optimality}) in a class of finite element integration loop nests (Definition~\ref{def:fem-loopnest}) and quasi-optimality in any other finite element integration loop nest.
+In this section, we build a transformation algorithm that produces optimal or quasi-optimal finite element integration loop nests.
 In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.

### 2015-12-03 TOMS_Paper/assembly.tex
Written a first version of the algorithm...

@@
 Finite element expressions can be seen as composition of operations between tensors.
-When multiplying two tensors it often happens that the optimal scheduling strategy is to be searched in a space of size 2.
-
+We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2.
 To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$.
-Forms are typically rich of operations of this kind, or equivalent.
 Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$.
@@
 \begin{enumerate}
-\item We leave the expression unmodified in the body of $L_i$; we only assign it to a temporary in the case it is a common sub-expression
+\item We leave the expression unmodified in the body of $L_i$; we may want to assign it to a temporary in the case it is a common sub-expression
 \item We notice the presence of spatial sharing. We can re-factorize the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
@@
 
-%
-%can make the factorization space explode 
-%
-%On the other hand, the possibly unstructured nature of the 
-%
-%In absence of further coefficient functions or if their use is such that  So optimal 
-%Due to the possibly unstructured nature of the $\sigma_{ei}$ expressions, Clearly, in general, the problem is much more complex
-
-%the expression $\dot(J^{-T} \nabla v, J^{-T} \nabla u)$ resulting from a Laplacian operator.
-
-%multilinearity -> avoids ambiguity
-%TODO : if also the Ri loop is linear, then optimality full stop.
-%we can relax linearity by talking of structured inner products which avoids ambiguity in collection
-
-%Expressions depending on the linear loops $L_j$ and $L_k$ result from inner products involving test and trial functions (possibly, derivatives or vector-valued).
-The problem of reaching optimality (at least) at the level of the multilinear loop nest reduces, therefore, to understand how to optimally schedule these inner products.
+%It is beyond the scope of this paper, but one could think of formalizing the conditions for which sigma_ei is structured...
 
@@
 \label{sec:sec:se-algo}
-Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on an abstract syntax tree representation of the loop nest.
+Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on a tree representation of the loop nest.
+The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest.
 
@@
 \normalfont
+The first step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
+This is to ensure that the operands of a generic operation are unique.
+For example, the expression $(a + a + b + c + a)$, in which we have a single n-ary operator \texttt{+}, is transformed into $(3a + b + c)$.
+
+We then perform a second depth-first visit of the expression tree and we recursively label each node $n$ with a couple ${<}I_{jk},\ cost{>}$, where: $I_{jk}$ represents the intersection of the symbols depending on the multilinear loop nest (i.e., depending on test and trial functions) within $n$, whereas $cost$ denotes the operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided).
+A crucial observation is that $I_{jk}$ is, in practice, always very small (a few units), even if the form is complex and includes several monomials.
+In addition, its computation can be made extremely efficient by also tracking the union of the symbols rooted in $n$.
+
+On a final depth-first visit, we examine the label of each operator \texttt{+}.
+If $I_{jk} = \emptyset$ -- that is, there is no spatial sharing -- we just proceed to the parent node.
+Otherwise, we speculatively assume that structured tensor operations are performed in the sub-tree rooted in \texttt{+}, and compare the operational cost of the two alternatives described in Section~\ref{sec:se-rln} (see points (1) and (2)).
+Alternative (1) has an operation count equal to $cost$.
+Alternative (2) requires collecting the shared symbols and estimating the impact of generalized code motion on the new factorization.
+For this, we need decide the collection order; that is, whether to start with symbols in $I_j$ or $I_k$.
+It can easily be demonstrated by contradiction that starting with $I_o$ such that $|L_o| |I_o|$ is minimum ($o \in {\lbrace} j, k{\rbrace}, |.|$ represents a generic size operator) leads to the lowest operation count.
+If found to be the most favorable, alternative (2) is straightforwardly applied; alternative (1) is instead applied only after all operands rooted in \texttt{+} have been examined.
+
+%If alternative (1) results better, the node \texttt{+} is marked as no longer expandable, which makes the recursion more efficient. Conversely,
+
+%We then perform a second depth-first visit of the expression tree and we recursively label each node $n$ with a triple ${<}I_{jk},\ U_{jk},\ cost{>}$, where: $I_{jk}$ and $U_{jk}$ represent, respectively, the intersection and the union of the symbols depending on the multilinear loop nest (i.e., depending on test and trial functions) within $n$, whereas $cost$ denotes the operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided).
+A crucial observation is that $I_{jk}$ and $U_{jk}$ are, in practice, always very small (a few units), even if the form is complex and includes several monomials.
+In addition, we note that $U_{jk}$ is tracked for the sole purpose of making the calculation of the parent $I_{jk}$ sets even faster.
+
 %The algorithm starts with ``flattening'' the sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ (by product expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
@@
 This could outweigh the gain due to pre-evaluation.
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination inhibits pre-evaluation, whereas pre-evaluation generally expose further sharing elimination opportunities.
+\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination inhibits pre-evaluation, whereas pre-evaluation generally exposes further sharing elimination opportunities.
 \end{enumerate}
@@
 \end{itemize}
+
+%TODO forget about the special case alpha=beta (ie test and trial from same fs)
 

### 2015-12-03 TOMS_Paper/assembly.tex
minor fixes and todos

@@
 
-%If alternative (1) results better, the node \texttt{+} is marked as no longer expandable, which makes the recursion more efficient. Conversely,
+%TODO sharing elimination of Li dependent stuff
 
-%We then perform a second depth-first visit of the expression tree and we recursively label each node $n$ with a triple ${<}I_{jk},\ U_{jk},\ cost{>}$, where: $I_{jk}$ and $U_{jk}$ represent, respectively, the intersection and the union of the symbols depending on the multilinear loop nest (i.e., depending on test and trial functions) within $n$, whereas $cost$ denotes the operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided).
-A crucial observation is that $I_{jk}$ and $U_{jk}$ are, in practice, always very small (a few units), even if the form is complex and includes several monomials.
-In addition, we note that $U_{jk}$ is tracked for the sole purpose of making the calculation of the parent $I_{jk}$ sets even faster.
-
-%The algorithm starts with ``flattening'' the sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ (by product expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.
-We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms.
-The choice of starting with $L_j$ or $L_k$ depends on two parameters: the number of unique terms, $n_{L}$, and the trip count, $tc_{L}$, of a loop $L$.
-It is fairly simple to see (and easily demonstrable by contradiction) that taking first $L$ such that $n_L \cdot tc_L$ is smallest leads to the largest operation count reduction.
-Assume, without loss of generality, that we start with $L_j$.
-Thanks to linearity, each factored product only has one symbol depending on $L_j$, and this symbol is now unique in the expression tree.
-The other sub-expression, independent of $L_j$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided.
-Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
-We repeat the same process with $L_k$.
-%
-%For each of the expressions $\eta_{ei} = \sum_w \sigma_{ei}^{w}$ lifted at the level of $L_i$, we need understand whether applying a similar process -- that is, factorizing $L_i$-dependent symbols such that $L_e$-dependent sub-expressions can be hoisted -- actually results in reduced operation count (the problem described in Section~\ref{sec:se-rln}, point 3).
-Consider two sub-expressions $a_{ei}$ and $b_{ei}$ in $\eta_{ei}$, with $n$ and $m$ being the respective number of summands, and with $c$ being the number of summands in common.
-Given a product $a_{ei} b_{ei}$, the increase in operation count after product expansion is $n \cdot m - \binom c2$.
-Given a sum $a_{ei} + b_{ei}$, the operation count after factorization of common terms is $n + m - c$.
-We use these simple formulae, together with the trip count of $L_i$, to predict the impact -- and then decide on the application -- of sharing elimination at the level of $L_i$.
-%
-%In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions.
+%TODO CSE / CCSE /  SCSE In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions.
 This is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
-
-% matrix notation, fixed normal form, so no need for fancy algors to do CSE
 \end{Algo}
@@
 
-Sharing elimination and pre-evaluation tackle the problem of minimizing spatial and temporal redundancy with a bounded working set in the loop nests and expressions defined in Section~\ref{sec:lnopt} \textit{using a very specific set of operators}.
-This must be kept in mind: obviously, the operation count might even be lower if we considered operators exploiting domain properties that we are here neglecting, such as redundancies in basis functions.
+Sharing elimination and pre-evaluation tackle the problem of minimizing spatial and temporal redundancy given a set of memory constraints in finite element integration loop nests, using \textit{a very specific set of operators}.
+This needs be emphasized since, theoretically, one could find an even lower operation count by exploiting domain-specific properties, such as redundancies in basis functions.
 

### 2015-12-07 TOMS_Paper/assembly.tex
MAH

@@
 As demonstrated by the substantial body of research on the topic, automating the generation of such high performance implementations poses several challenges.
-This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
+This is a result of the complexity inherent in the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
 General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}).
@@
 \begin{itemize}
-\item We formalize optimality for finite element integration and we build the space of legal transformations within which the optimal is sought.
+\item We formalize finite element integration loop nests and we build the space of legal transformations impacting their operation count.
 \item We provide an algorithm to select points in the transformation space.
@@
 \item We explain under what conditions our algorithm leads to optimality.
-In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, at least, always achieved (i.e., the operation count is at least optimal in innermost loops).
+In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, in general, always achieved (i.e., the operation count is at least optimal in innermost loops).
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, generalized code motion and (if necessary) common sub-expression elimination.
-If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorization strategies.
+If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
 Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This require coordination with other transformations, as discussed in the following sections.
-For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved whenever stronger conditions hold.
-Relaxing the problem is essential to produce simple and computationally efficient algorithms -- which, as such, can be integrated with a compiler.
+For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved if stronger hypotheses hold.
+Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
 
-In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties of our expressions, while in Section~\ref{sec:sec:se-algo} we present the sharing elimination algorithm.
+Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions.
+Section~\ref{sec:sec:se-algo} presents the sharing elimination algorithm.
 We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
@@
 A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE).
-This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that (i) are symbolically identical and (ii) reduce to the same value.
+This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that are (i) symbolically identical and (ii) reduce to the same value.
 We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block.
@@
 \begin{enumerate}
-\item We leave the expression unmodified in the body of $L_i$; we may want to assign it to a temporary in the case it is a common sub-expression
-\item We notice the presence of spatial sharing. We can re-factorize the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
+\item We leave the expression unmodified in the body of $L_i$ (in general, one may consider applying generalized code motion or creating a temporary for CCSE -- but it is not the case of this simple example).
+\item We notice the presence of spatial sharing. We can recast the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
 \end{enumerate}
-The reader can verify that option 2) reduces the operation count with respect to 1) if and only if the size of $L_i$ is strictly greater than $c=1$.
-One can find this dichotomy in a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with values of $c$ varying with the operational cost inherent in removing spatial sharing.
+The reader can verify that option 2) improves the operation count over 1) if and only if the size of $L_i$ is strictly greater than $c=1$.
+One can appreciate this dichotomy in a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with varying values of $c$.
 
 In general, given a product of two sub-expressions in a finite element integration loop nest, we observe that whenever there is no ambiguity in the elimination of spatial sharing, the problem of finding an optimal scheduling reduces to the comparison of two alternatives.
-We note that this condition simply holds if a loop is guaranteed to be linear; in the next section, we will show how take advantage of this property to ensure quasi-optimal sharing elimination.
-Further, if the $\sigma_{ei}$ expressions are sufficiently regular, then the same approach can be applied to all expressions logically belonging to $L_i$.
-The regularity of $\sigma_{ei}$ is not as uncommon as it may seem at first glance: for instance, this property holds for many of the tensors that need be evaluated in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).
+We note that this condition simply holds if a loop is guaranteed to be linear; in the next section, we will show how to take advantage of this property to ensure quasi-optimal sharing elimination.
+Further, if the $\sigma_{ei}$ expressions are sufficiently regular, the same approach can be applied to all of the $L_i$-dependent expressions.
+Structured tensor operations within $\sigma_{ei}$ are not as uncommon as one may think in the first place: for instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).
 
@@
 The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest.
+It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration expressions.
 
@@
 \normalfont
-The first step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
-This is to ensure that the operands of a generic operation are unique.
-For example, the expression $(a + a + b + c + a)$, in which we have a single n-ary operator \texttt{+}, is transformed into $(3a + b + c)$.
+The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions, to ensure the uniqueness of operands.
+For example, the expression $(a + a + b + c + a)$ is transformed into $(3a + b + c)$.
 
-We then perform a second depth-first visit of the expression tree and we recursively label each node $n$ with a couple ${<}I_{jk},\ cost{>}$, where: $I_{jk}$ represents the intersection of the symbols depending on the multilinear loop nest (i.e., depending on test and trial functions) within $n$, whereas $cost$ denotes the operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided).
+We perform a second depth-first visit of the expression tree to collect information about multilinear ``symbols'' and ``operands''.
+For example, any sub-expression $\alpha_{eij}$ ($\beta_{eik}$) is regarded as an operand, while a symbol could be any test (trial) function (or derivatives of) $a_{ij} \in \alpha_{eij}$.
+In particular, for each symbol $s$, we store a 2-tuple ${<}n,\ \lbrace ops \rbrace{>}$, where $n$ is the number of occurrences of $s$ in the input expression, while $\lbrace ops \rbrace$ represents the set of operands in which $s$ appears.
+
+Let $\mathbb{O}$ be the set of all $\lbrace ops \rbrace$s.
+
+%ops\ =\ \lbrace \alpha_{eij}^{1}, ..., \alpha_{eij}^{N}\rbrace$
+
+%and we recursively label each node $n$ with a 2-tuple ${<}I_{jk},\ op_{gcm}{>}$, where: $I_{jk}$ represents the intersection of the operands depending on the multilinear loop nest within $n$, while $op_{gcm}$ denotes the gain in operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided).
 A crucial observation is that $I_{jk}$ is, in practice, always very small (a few units), even if the form is complex and includes several monomials.
 In addition, its computation can be made extremely efficient by also tracking the union of the symbols rooted in $n$.
-
-On a final depth-first visit, we examine the label of each operator \texttt{+}.
-If $I_{jk} = \emptyset$ -- that is, there is no spatial sharing -- we just proceed to the parent node.
-Otherwise, we speculatively assume that structured tensor operations are performed in the sub-tree rooted in \texttt{+}, and compare the operational cost of the two alternatives described in Section~\ref{sec:se-rln} (see points (1) and (2)).
-Alternative (1) has an operation count equal to $cost$.
-Alternative (2) requires collecting the shared symbols and estimating the impact of generalized code motion on the new factorization.
+%
+%On a subsequent depth-first visit, we examine the label of each operator \texttt{+}.
+If $I_{jk} = \emptyset$ -- that is, no spatial sharing -- we just proceed to the parent node.
+Otherwise, we speculatively assume that structured tensor operations are being performed in the sub-tree rooted in \texttt{+}, and compare the operational cost of the two alternatives described in Section~\ref{sec:se-rln} (see points (1) and (2)).
+Alternative (1) has an operation count equal to $op_{gcm}$.
+Alternative (2) requires collecting the shared symbols and estimating the impact of generalized code motion on the new factorization (similarly to Figure~\ref{code:multi_loopnest}).
 For this, we need decide the collection order; that is, whether to start with symbols in $I_j$ or $I_k$.
-It can easily be demonstrated by contradiction that starting with $I_o$ such that $|L_o| |I_o|$ is minimum ($o \in {\lbrace} j, k{\rbrace}, |.|$ represents a generic size operator) leads to the lowest operation count.
-If found to be the most favorable, alternative (2) is straightforwardly applied; alternative (1) is instead applied only after all operands rooted in \texttt{+} have been examined.
+It can easily be demonstrated by contradiction that starting with $I_o$ such that $|L_o| |I_o|$ is minimum ($o \in {\lbrace} j, k{\rbrace}$, with $|.|$ representing a generic ``\texttt{sizeof}'' operator) leads to the lowest operation count.
+Once all of the operands rooted in Alternative (2) is immediately applied if determined to be ; alternative (1) is instead applied only after all operands rooted in \texttt{+} have been examined.
 

### 2015-12-07 TOMS_Paper/assembly.tex
writing stuff lol

@@
 The reader can verify that option 2) improves the operation count over 1) if and only if the size of $L_i$ is strictly greater than $c=1$.
-One can appreciate this dichotomy in a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with varying values of $c$.
+One can appreciate the dichotomy between 1) and 2) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with different values of $c$.
 
-In general, given a product of two sub-expressions in a finite element integration loop nest, we observe that whenever there is no ambiguity in the elimination of spatial sharing, the problem of finding an optimal scheduling reduces to the comparison of two alternatives.
-We note that this condition simply holds if a loop is guaranteed to be linear; in the next section, we will show how to take advantage of this property to ensure quasi-optimal sharing elimination.
-Further, if the $\sigma_{ei}$ expressions are sufficiently regular, the same approach can be applied to all of the $L_i$-dependent expressions.
-Structured tensor operations within $\sigma_{ei}$ are not as uncommon as one may think in the first place: for instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).
-
-%It is beyond the scope of this paper, but one could think of formalizing the conditions for which sigma_ei is structured...
+We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.
+Of notable importance, in the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations.
+Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too.
+For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).
 
@@
 The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions, to ensure the uniqueness of operands.
-For example, the expression $(a + a + b + c + a)$ is transformed into $(3a + b + c)$.
+For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
 
-We perform a second depth-first visit of the expression tree to collect information about multilinear ``symbols'' and ``operands''.
-For example, any sub-expression $\alpha_{eij}$ ($\beta_{eik}$) is regarded as an operand, while a symbol could be any test (trial) function (or derivatives of) $a_{ij} \in \alpha_{eij}$.
-In particular, for each symbol $s$, we store a 2-tuple ${<}n,\ \lbrace ops \rbrace{>}$, where $n$ is the number of occurrences of $s$ in the input expression, while $\lbrace ops \rbrace$ represents the set of operands in which $s$ appears.
+We perform a second depth-first visit of the expression tree to track all occurrences of multilinear operands (i.e., any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression in normal form).
+We create disjoint partitions of multilinear operands such that no spatial sharing is present among them.
+For each partition, we have two alternatives: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding the expression to expose spatial sharing and enable a subsequent factorization.
+These two alternatives basically correspond to points 1) and 2) in Section~\ref{sec:se-rln}.
 
-Let $\mathbb{O}$ be the set of all $\lbrace ops \rbrace$s.
+Given a partition $P$, to decide between (i) and (ii) we adopt the two following criteria:
+\begin{enumerate}
+\item We compare the cardinality of $P$ to that of the set of multilinear symbols (e.g., test functions) in $P$.
+If the former is lower, we use (i); if the latter is lower, we use (ii); if they are identical, we check the next criterion, which attempts to optimize for the outer loops.
+\item We know how many times a multilinear symbol $s$ appears in the input expression.
+Let $n_{s}$ denote this quantity.
+Let $c_{op}$ be the cost of (i.e., number of operations in) $op$.
+Also, recall from Figure~\ref{code:loopnest} that $L_i$ and $L_k$ have size $M$ and $O$.
+Assume, for example, that $P$ depends on $[L_i, L_k]$.
+We observe that the increase in operation count at the level of the outer loops would be: $M O \sum_{w \in P} c_{op}^{w}$ if (i) were applied; $M n_{s}$ if (ii) were applied.
+Therefore, we take (i) or (ii) depending on which quantity is minimum.
+\end{enumerate}
 
-%ops\ =\ \lbrace \alpha_{eij}^{1}, ..., \alpha_{eij}^{N}\rbrace$
+On a last depth-first visit of the expression tree, we apply point (2) above to all of the $L_i$-dependent expressions. 
+\end{Algo}
 
-%and we recursively label each node $n$ with a 2-tuple ${<}I_{jk},\ op_{gcm}{>}$, where: $I_{jk}$ represents the intersection of the operands depending on the multilinear loop nest within $n$, while $op_{gcm}$ denotes the gain in operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided).
-A crucial observation is that $I_{jk}$ is, in practice, always very small (a few units), even if the form is complex and includes several monomials.
-In addition, its computation can be made extremely efficient by also tracking the union of the symbols rooted in $n$.
-%
-%On a subsequent depth-first visit, we examine the label of each operator \texttt{+}.
-If $I_{jk} = \emptyset$ -- that is, no spatial sharing -- we just proceed to the parent node.
-Otherwise, we speculatively assume that structured tensor operations are being performed in the sub-tree rooted in \texttt{+}, and compare the operational cost of the two alternatives described in Section~\ref{sec:se-rln} (see points (1) and (2)).
-Alternative (1) has an operation count equal to $op_{gcm}$.
-Alternative (2) requires collecting the shared symbols and estimating the impact of generalized code motion on the new factorization (similarly to Figure~\ref{code:multi_loopnest}).
-For this, we need decide the collection order; that is, whether to start with symbols in $I_j$ or $I_k$.
-It can easily be demonstrated by contradiction that starting with $I_o$ such that $|L_o| |I_o|$ is minimum ($o \in {\lbrace} j, k{\rbrace}$, with $|.|$ representing a generic ``\texttt{sizeof}'' operator) leads to the lowest operation count.
-Once all of the operands rooted in Alternative (2) is immediately applied if determined to be ; alternative (1) is instead applied only after all operands rooted in \texttt{+} have been examined.
-
-%TODO sharing elimination of Li dependent stuff
-
-%TODO CSE / CCSE /  SCSE In all this process, whenever a sub-expression is hoisted, we search and replace all common sub-expressions.
-This is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
-\end{Algo}
+Section~\ref{sec:proof} elucidates the claims of quasi-optimality and optimality.
 
@@
 To determine $\rho$ we execute the initial phase of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes reduction pre-evaluation.
+
+\subsection{Again on optimality}
+\label{sec:proof}
 

### 2015-12-07 TOMS_Paper/assembly.tex
Doesnt look that bad now

@@
 It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration expressions.
+Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.
 
@@
 If the former is lower, we use (i); if the latter is lower, we use (ii); if they are identical, we check the next criterion, which attempts to optimize for the outer loops.
-\item We know how many times a multilinear symbol $s$ appears in the input expression.
-Let $n_{s}$ denote this quantity.
-Let $c_{op}$ be the cost of (i.e., number of operations in) $op$.
-Also, recall from Figure~\ref{code:loopnest} that $L_i$ and $L_k$ have size $M$ and $O$.
-Assume, for example, that $P$ depends on $[L_i, L_k]$.
-We observe that the increase in operation count at the level of the outer loops would be: $M O \sum_{w \in P} c_{op}^{w}$ if (i) were applied; $M n_{s}$ if (ii) were applied.
-Therefore, we take (i) or (ii) depending on which quantity is minimum.
+\item We know how many times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
+Let $n_{o}$ and $n_{s}$ denote these quantities.
+Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols.
+Let $c_{o}$ be the cost of $o$ (i.e., the number of operations).
+Assume, for this example, that $P$ depends on $[L_i, L_k]$ (which we know, from Figure~\ref{code:loopnest}, to have extent $M$ and $O$, respectively).
+We observe that the increase in operation count at the level of the outer loops would be: $M (O \sum_{o \in X} c_{o} + \sum_{o \in X} (n_{o} - 1))$ if (i) were applied; $M \sum_{s \in Y} (n_{s} - 1)$ if (ii) were applied.
+We then take (i) or (ii) depending on which quantity is minimum.
 \end{enumerate}
@@
 On a last depth-first visit of the expression tree, we apply point (2) above to all of the $L_i$-dependent expressions.
+
+%TODO what partition do we take first ? we need a partial ordering here
 \end{Algo}
 
-Section~\ref{sec:proof} elucidates the claims of quasi-optimality and optimality.
-
-\subsection{Pre-evaluation}
+\subsection{Monomial pre-evaluation}
 \label{sec:pre-evaluation}
@@
 In this section, we discuss role and legality of a fourth operator: \textit{reduction pre-evaluation}.
-We will see that what makes this operator special is the fact that there exists a single point in a monomial transformation space (i.e., specific factorization and code motion) preserving the correctness of the transformation.
+We will see that what makes this operator special is the fact that there exists a single point in the transformation space of a monomial (i.e., specific factorization and code motion) preserving the correctness of the transformation.
 
@@
 Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
-We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$?
+We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$?
 The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.
@@
 
-Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on ...
-data objects.
+Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects.
 Consequently, generalized code motion as performed by sharing elimination is not allowed to hoist expressions outside of $L_e$.
@@
 \subsection{Transformation algorithm}
-
-At this point, two main issues are to be addressed: 
+At this point, two main issues still need be addressed: 
 \begin{enumerate}
@@
 \label{sec:proof}
+
+\begin{Prop}
+\label{prop:optimal-approach}
+Consider a multilinear form and the corresponding loop nest $\Lambda$ implementing it.
+The form comprises a set of monomials, $M$.
+Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
+Assume that:
+\begin{enumerate}
+\item pre-evaluating different monomials does not result in identical tables;
+\item ...
+\end{enumerate} 
+Then, the transformation algorithm produces a quasi-optimal $\Lambda'$ as established by Definition ?. In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
+\end{Prop}
+\begin{proof}
+....
+%We first comment on the assumptions.
+1) Inherently a pathological case, so we simply neglect it.
+2) Might occur in forms with several monomials, although this was never the case in the problems that we have so far investigated (which are obviously more than those presented in Section~\ref{sec:perf-results}).
+These assumptions allow us to avoid a ``global analysis'' of $M$.
+%
+%We prove our claim by demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$.
+%
+%A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$.
+If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim.
+Assumption 1) prevents this situation, so the statement is true.
+%
+%B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term.
+We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
+That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
+Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions.
+The previously discussed assumption 2) allows us to neglect this possibility.
+\end{proof}
 

### 2015-12-08 TOMS_Paper/assembly.tex
Start writing again the proof...

@@
 \subsubsection{Algorithm}
-\label{sec:sec:se-algo}
-Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on a tree representation of the loop nest.
+\label{sec:se-algo}
+Algorithm~\ref{algo:sharing-elimination} details the steps for performing sharing elimination on a tree representation of the loop nest.
 The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest.
@@
 \begin{Algo}[Sharing elimination]
-\label{algo:opt}
+\label{algo:sharing-elimination}
 \normalfont
@@
 
-We conclude providing a more refined definition of optimality.
-
-\begin{Def}[Optimality of a loop nest with bounded working set]
-\label{def:optimality}
-Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$) and Constraints~\ref{const:Le} and~\ref{const:TH} are satisfied.
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal.
-\end{Def}
-
-An analogous definition can be provided for quasi-optimality.
-
 \subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
@@
 \label{sec:optimal-synthesis}
-In this section, we build a transformation algorithm that produces optimal or quasi-optimal finite element integration loop nests.
+In this section, we build a transformation algorithm that, given a memory bound, produces optimal or quasi-optimal finite element integration loop nests.
 In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.
@@
 \subsection{Transformation algorithm}
-At this point, two main issues still need be addressed: 
+We address the two following issues: 
 \begin{enumerate}
-\item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
-\item \textit{Finding the global optimum.} Consider a form comprising two monomials $m1$ and $m2$.
+\item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
+\item \textit{Search for a global optimum.} Consider a form comprising two monomials $m1$ and $m2$.
 Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g.
@@
 The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations.
-If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} tends to have negligible cost), the overhead becomes unacceptable when applying pre-evaluation to all possible bipartitions (Algorithm~\ref{algo:pre-evaluation}, due to the symbolic evaluation of $L_i$).
+If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead becomes unacceptable when applying pre-evaluation to all possible bipartitions (Algorithm~\ref{algo:pre-evaluation}, due to the symbolic evaluation of $L_i$).
 We then seek an analytic way of determining $\theta_{pre}$.
@@
 
-\begin{figure}\begin{CenteredBox}
+\begin{figure}[b]\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
@@
 
-\subsection{Again on optimality}
+\subsection{Formalization}
 \label{sec:proof}
+The following proposition states that our approach guarantees quasi-optimality. The proof re-uses most of the concepts and explanations provided throughout the paper.
 
@@
 \label{prop:optimal-approach}
-Consider a multilinear form and the corresponding loop nest $\Lambda$ implementing it.
+Consider a multilinear form and the corresponding finite element integration loop nest $\Lambda$ implementing it.
 The form comprises a set of monomials, $M$.
-Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$).
-Assume that:
+Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$.
+Assume that pre-evaluating different monomials does not result in identical tables.
+Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal (as in Definition~\ref{def:mln-quasi-optimality}).
+\end{Prop}
+\begin{proof}
+Given a monomial $m$, $\theta(m)$ provides the cost/gain due to pre-evaluating $m$.
+$\theta(m)$ is \textit{not} an estimate\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}.
+The estimate is actually for the whole form, rather than for its monomials.}, so $P$ (and $Z$) are constructed correctly.
+
+Let $\Lambda_{ops}$ denote the operation count (additions, products) in $\Lambda$.
+Let $\Lambda'_I \subset \Lambda'$ be the set of all innermost loops produced by the transformation algorithm (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
+We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$.
+For this, we need to show that:
 \begin{enumerate}
-\item pre-evaluating different monomials does not result in identical tables;
+\item the coordination of pre-evaluation with sharing elimination is optimal. This boils down to show that
+\begin{enumerate}
+\item pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$
+\item not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$
+\end{enumerate}
+\item sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal (again, in the sense of Definition~\ref{def:mln-quasi-optimality})
+\end{enumerate}
+
+We discuss each of these points individually
+
+\begin{enumerate}
+\begin{enumerate}
+\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$.
+Consider two generic monomials $m_1, m_2 \in M$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
+In a scenario in which $m_{T_{1}} \cap m_{T_{2}}$, we would have that at least one table would be assigned to the same temporary, with the transformation algorithm skipping a potential global optimum.
+By assumption, we neglect this possibility.
+The assumption is sustained by (i) the pathological nature of the scenario, (ii) the complexity inherent in handling it, and (iii) the fact that such a potential global optimum would probably be very close to the retrieved solution.
 \item ...
 \end{enumerate}
-Then, the transformation algorithm produces a quasi-optimal $\Lambda'$ as established by Definition ?. In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
-\end{Prop}
-\begin{proof}
-....
-%We first comment on the assumptions.
-1) Inherently a pathological case, so we simply neglect it.
-2) Might occur in forms with several monomials, although this was never the case in the problems that we have so far investigated (which are obviously more than those presented in Section~\ref{sec:perf-results}).
-These assumptions allow us to avoid a ``global analysis'' of $M$.
-%
-%We prove our claim by demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$.
+\item ...
+\end{enumerate}
 %
@@
 
+%TODO : In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
+
 \subsection{Trading operation count for analysis cost}
@@
 However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
-\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity.
+\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:sharing-elimination} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity.
 The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed).

### 2015-12-09 TOMS_Paper/assembly.tex
Minor fixes

@@
 Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions.
-Section~\ref{sec:sec:se-algo} presents the sharing elimination algorithm.
+Section~\ref{sec:se-algo} presents the sharing elimination algorithm.
 We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
@@
 Assume that pre-evaluating different monomials does not result in identical tables.
-Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal (as in Definition~\ref{def:mln-quasi-optimality}).
+Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
 \end{Prop}
 \begin{proof}
-Given a monomial $m$, $\theta(m)$ provides the cost/gain due to pre-evaluating $m$.
-$\theta(m)$ is \textit{not} an estimate\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}.
-The estimate is actually for the whole form, rather than for its monomials.}, so $P$ (and $Z$) are constructed correctly.
+The function $\theta$ provides cost or gain of pre-evaluation.
+$\theta$ retrieves exact values, rather than estimates\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}.
+The estimate is actually for the whole form, rather than for its monomials.}, so $P$ and $Z$ are constructed correctly.
 
-Let $\Lambda_{ops}$ denote the operation count (additions, products) in $\Lambda$.
+Let $\Lambda_{ops}$ denote the operation count (number of additions and products) in $\Lambda$.
 Let $\Lambda'_I \subset \Lambda'$ be the set of all innermost loops produced by the transformation algorithm (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
-We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$.
-For this, we need to show that:
+We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$, which holds if and only if
 \begin{enumerate}
-\item the coordination of pre-evaluation with sharing elimination is optimal. This boils down to show that
+\item \textit{The coordination of pre-evaluation with sharing elimination is optimal}. This boils down to prove that
 \begin{enumerate}
-\item pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$
-\item not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$
+\item \textit{pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$}
+\item \textit{not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$}
 \end{enumerate}
-\item sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal (again, in the sense of Definition~\ref{def:mln-quasi-optimality})
-\end{enumerate}
-
-We discuss each of these points individually
-
-\begin{enumerate}
+~
 \begin{enumerate}
@@
 \end{enumerate}
-\item ...
+
+\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal.}\\
+...
 \end{enumerate}
+
 %

### 2015-12-09 TOMS_Paper/assembly.tex
ADd stuff to sharing elimination algo..

@@
 \label{sec:se-algo}
-Algorithm~\ref{algo:sharing-elimination} details the steps for performing sharing elimination on a tree representation of the loop nest.
-The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest.
-It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration expressions.
+Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest.
+The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest.
+It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration.
 Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.
@@
 \normalfont
-The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions, to ensure the uniqueness of operands.
+In the following, we assume $|.|$ to be a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop.
+The cost, in number of operations, of a sub-expression $o$ is represented by $c_o$.
+
+The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
+This ensures the uniqueness of operands.
 For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
 
-We perform a second depth-first visit of the expression tree to track all occurrences of multilinear operands (i.e., any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression in normal form).
-We create disjoint partitions of multilinear operands such that no spatial sharing is present among them.
-For each partition, we have two alternatives: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding the expression to expose spatial sharing and enable a subsequent factorization.
-These two alternatives basically correspond to points 1) and 2) in Section~\ref{sec:se-rln}.
+We perform a second depth-first visit of the expression tree to build a partitioning of multilinear operands; that is, any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression in normal form.
+We partition the multilinear operands into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$.
+The multilinear operands in a partition $P_i$ share the same set of symbols (e.g., all operands in which the first order derivatives of test functions appear), while there is no sharing across different partitions.
+We have to establish:
+\begin{enumerate}
+\item the order in which sharing elimination will be applied to the partitions in $\mathbb{P}$ (this has a relevant impact on the outcome of the transformation process).
+\item a scheduling strategy for each $P \in \mathbb{P}$.
+\end{enumerate}
 
-Given a partition $P$, to decide between (i) and (ii) we adopt the two following criteria:
+We start with point 1).
+We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{L_j}$ and $\mathbb{P}^{L_k}$, which logically separates partitions based on the loop they depend on.
+We then order the partitions lexicographically, with the position of partition $P^{L_l} \in \mathbb{P}^{L_l}$ given by ${<}|\mathbb{P}^{L_l}|,\ |L_l| \sum_{o \in P^{L_l}} c_o{>}$.
+
+%TODO need intuition here!
+
+We now tackle point 2).
+For each partition $P \in \mathbb{P}$, we have to decide between two options, as already explained in Section~\ref{sec:se-rln}.
+We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion.
+For this choice, we adopt the following criteria:
 \begin{enumerate}
-\item We compare the cardinality of $P$ to that of the set of multilinear symbols (e.g., test functions) in $P$.
-If the former is lower, we use (i); if the latter is lower, we use (ii); if they are identical, we check the next criterion, which attempts to optimize for the outer loops.
+\item Let $S_P$ be set of multilinear symbols in $P$ (e.g., test functions).
+If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops
 \item We know how many times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
@@
 Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols.
-Let $c_{o}$ be the cost of $o$ (i.e., the number of operations).
 Assume, for this example, that $P$ depends on $[L_i, L_k]$ (which we know, from Figure~\ref{code:loopnest}, to have extent $M$ and $O$, respectively).
@@
 
-On a last depth-first visit of the expression tree, we apply point (2) above to all of the $L_i$-dependent expressions.
+%TODO need intuition here!
 
-%TODO what partition do we take first ? we need a partial ordering here
+It remains to schedule $L_i$-dependent expressions; that is, operations that, logically, do not belong to the multilinear loop nest.
+We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above.
+In doing this, we make the assumption that an $L_i$-dependent expression is the result of a structured tensor operation.
+If this were not the case, the factorization is chosen greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
 \end{Algo}

### 2015-12-10 TOMS_Paper/assembly.tex
Changing algo...

@@
 \normalfont
-In the following, we assume $|.|$ to be a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop.
-The cost, in number of operations, of a sub-expression $o$ is represented by $c_o$.
+We introduce the following notation and terminology
+\begin{itemize}
+\item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
+\item $c_o$: the cost (i.e., number of operations) of an expression $o$
+\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression; a \textit{multilinear symbol} is any symbol appearing in a multilinear operand (e.g., test functions)
+\end{itemize}
 
@@
 
-We perform a second depth-first visit of the expression tree to build a partitioning of multilinear operands; that is, any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression in normal form.
-We partition the multilinear operands into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$.
-The multilinear operands in a partition $P_i$ share the same set of symbols (e.g., all operands in which the first order derivatives of test functions appear), while there is no sharing across different partitions.
+We perform a second depth-first visit of the expression tree to collect all multilinear operands and partition them into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$.
+The multilinear operands in a partition $P_i$ share a set of symbols (e.g., all operands in which first order derivatives of test functions appear), whereas there is no sharing across different partitions.
 We have to establish:
 \begin{enumerate}
-\item the order in which sharing elimination will be applied to the partitions in $\mathbb{P}$ (this has a relevant impact on the outcome of the transformation process).
-\item a scheduling strategy for each $P \in \mathbb{P}$.
+\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, $P_i$ is transformed before $P_j$, and
+\item a scheduling strategy for each $P_i$
 \end{enumerate}
@@
 We start with point 1).
-We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{L_j}$ and $\mathbb{P}^{L_k}$, which logically separates partitions based on the loop they depend on.
-We then order the partitions lexicographically, with the position of partition $P^{L_l} \in \mathbb{P}^{L_l}$ given by ${<}|\mathbb{P}^{L_l}|,\ |L_l| \sum_{o \in P^{L_l}} c_o{>}$.
-
-%TODO need intuition here!
+We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{j}$ and $\mathbb{P}^{k}$, which logically separates partitions based on loop dependency.
+We then order $\mathbb{P}$ lexicographically, with the position of $P^{l}$ given by ${<}|\mathbb{P}^{l}|,\ |L_l| \sum_{o \in P^{l}} c_o{>}$.
+Intuitively, this simply means that operands appearing more often in the expression will be factorized first, which is essential to minimize the operation count within a loop.
 
 We now tackle point 2).
-For each partition $P \in \mathbb{P}$, we have to decide between two options, as already explained in Section~\ref{sec:se-rln}.
+For each partition, we have to decide between two options, as already explained in Section~\ref{sec:se-rln}.
 We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion.
@@
 \begin{enumerate}
-\item Let $S_P$ be set of multilinear symbols in $P$ (e.g., test functions).
+\item Let $S_P$ be set of multilinear symbols in $P$.
 If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops
-\item We know how many times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
-Let $n_{o}$ and $n_{s}$ denote these quantities.
+\item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
 Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols.

### 2015-12-10 TOMS_Paper/assembly.tex
Add examples to sharing elimination. Minor fixes

@@
 The need for rapid implementation of high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
-This has been proven successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects.
+This has been proven successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-paper}) projects.
 In these frameworks, the weak variational form of a problem is expressed at high level by means of a domain-specific language.
@@
 \begin{figure}[h]\begin{CenteredBox}
-{\subfigcapskip = 7pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
+{\subfigcapskip = 11pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
 ~~~~~
-{\subfigcapskip = 7pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
+{\subfigcapskip = 11pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
 ~~~~~
@@
 
-Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions.
+Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that characterize our expressions.
 Section~\ref{sec:se-algo} presents the sharing elimination algorithm.
@@
 \item $c_o$: the cost (i.e., number of operations) of an expression $o$
-\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression; a \textit{multilinear symbol} is any symbol appearing in a multilinear operand (e.g., test functions)
+\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the expression; a \textit{multilinear symbol} is a symbol appearing in a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, trial functions, first order derivatives of test functions, etc.)
 \end{itemize}
@@
 The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
-This ensures the uniqueness of operands.
 For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
@@
 We perform a second depth-first visit of the expression tree to collect all multilinear operands and partition them into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$.
-The multilinear operands in a partition $P_i$ share a set of symbols (e.g., all operands in which first order derivatives of test functions appear), whereas there is no sharing across different partitions.
+The multilinear operands in a partition $P_i$ share a set of multilinear symbols, whereas there is no sharing across different partitions.
 We have to establish:
 \begin{enumerate}
-\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, $P_i$ is transformed before $P_j$, and
+\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, then $P_i$ is scheduled before $P_j$, and
 \item a scheduling strategy for each $P_i$
@@
 We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{j}$ and $\mathbb{P}^{k}$, which logically separates partitions based on loop dependency.
-We then order $\mathbb{P}$ lexicographically, with the position of $P^{l}$ given by ${<}|\mathbb{P}^{l}|,\ |L_l| \sum_{o \in P^{l}} c_o{>}$.
-Intuitively, this simply means that operands appearing more often in the expression will be factorized first, which is essential to minimize the operation count within a loop.
+We then order $\mathbb{P}$ lexicographically, with the position of $P \in \mathbb{P}^j$ given by ${<}|\mathbb{P}^{j}|,\ |L_j| \sum_{o \in P} c_o{>}$ (and analogously for $P \in \mathbb{P}^k$).
+Intuitively, this simply means that operands appearing more often in the expression are handled first.
+We will later elaborate on this.
 
 We now tackle point 2).
-For each partition, we have to decide between two options, as already explained in Section~\ref{sec:se-rln}.
-We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion.
-For this choice, we adopt the following criteria:
+For each partition, we have two options, as already explained in Section~\ref{sec:se-rln}.
+We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE; (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion.
+We adopt the following criteria:
 \begin{enumerate}
-\item Let $S_P$ be set of multilinear symbols in $P$.
-If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops
+\item Let $S_P$ be the set of multilinear symbols in $P$.
+If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops.
 \item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
-Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols.
-Assume, for this example, that $P$ depends on $[L_i, L_k]$ (which we know, from Figure~\ref{code:loopnest}, to have extent $M$ and $O$, respectively).
-We observe that the increase in operation count at the level of the outer loops would be: $M (O \sum_{o \in X} c_{o} + \sum_{o \in X} (n_{o} - 1))$ if (i) were applied; $M \sum_{s \in Y} (n_{s} - 1)$ if (ii) were applied.
-We then take (i) or (ii) depending on which quantity is minimum.
+Assume that $P$ depends on $[L_i, L_k]$.
+The increase in operation count at the level of the outer loops is: if (i) were applied, $|L_i| (|L_k| \sum_{o \in P} c_{o} + \sum_{o \in P} (n_{o} - 1))$; if (ii) were applied, $|L_i| \sum_{s \in P_S} (n_{s} - 1)$.
+We pick (i) or (ii) such that the increase is minimum.
+Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that they both minimize the cost of the innermost loop.
 \end{enumerate}
 
-%TODO need intuition here!
+It remains to schedule $L_i$-dependent expressions.
+We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above.
+In doing this, we speculatively assume that all expressions result from structured tensor operations.
+If this were not the case, factorization would be applied greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
+\end{Algo}
 
-It remains to schedule $L_i$-dependent expressions; that is, operations that, logically, do not belong to the multilinear loop nest.
-We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above.
-In doing this, we make the assumption that an $L_i$-dependent expression is the result of a structured tensor operation.
-If this were not the case, the factorization is chosen greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
-\end{Algo}
+\subsubsection{Examples}
+Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D.
+We observe that $P_0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
+Clearly, $\mathbb{P} = \lbrace P_0, P_1 \rbrace$, with $\mathbb{P}^k = \lbrace P_0 \rbrace$ and $\mathbb{P}^j = \lbrace P_1 \rbrace$.
+A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
+\begin{figure}\begin{CenteredBox}
+{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
+~~~~~~~~~~
+{\subfigcapskip = 7pt \subfigure[Optimal form]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
+\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}
 
@@
 Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
-We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$?
+We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $I$?
 The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.
@@
 ~~~~~~~~~~
-{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
+{\subfigcapskip = 5pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
 \end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\label{code:loopnest_rednored}\end{figure}
@@
 \begin{enumerate}
-\item In contrast to what happens with hoisting in multilinear loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $).
+\item In contrast to what happens with hoisting in multilinear loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $J K$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $E J K$ for those depending on $[L_e, L_i, L_j, L_k] $).
 This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
@@
 \begin{itemize}
-\item The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions.
+\item The fact that $E \gg I, J, K$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions.
 Imagine $\Lambda$ is enclosed in a time stepping loop.
 One could think of exposing (through some transformations) and hoisting any time-invariant sub-expressions to minimize redundant computation at every time step.
-The working set size could then increase by a factor $L$.
+The working set size could then increase by a factor $E$.
 The gain in number of operations executed could therefore be outweighed, from a runtime viewpoint, by a much larger memory pressure.
@@
 the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, as explained in Section~\ref{sec:pre-evaluation}, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N O$ for bilinear forms and of size $N$ (or $O$) for linear forms.
+For example, as explained in Section~\ref{sec:pre-evaluation}, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $J K$ for bilinear forms and of size $J$ (or $K$) for linear forms.
 This can sometimes break such a ``local memory threshold''.

### 2015-12-11 TOMS_Paper/assembly.tex
Several fixes all around the paper

@@
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is at least minimal in all innermost loops.
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal in all innermost loops.
 \end{Def}
@@
 Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
-On the other hand, multilinearity ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions, which is reflected in the innermost loop.
+On the other hand, multilinearity ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions.
 In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
@@
 Section~\ref{sec:se-algo} presents the sharing elimination algorithm.
-We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.
+We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form.
 
@@
 ~~~~~~~~~~
-{\subfigcapskip = 7pt \subfigure[Optimal form]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
+{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
 \end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}
@@
 We start with an example.
-Consider again the bilinear form implementation in Figure~\ref{code:loopnest}.
-We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $I$?
-The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.
+Consider again the loop nest and the expression in Figure~\ref{code:loopnest}.
+We pose the following question: are we able to identify sub-expressions for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $I$?
+The transformation we look for is exemplified in Figure~\ref{code:loopnest_rednored} with a simple loop nest.
+The reader can easily verify that a similar transformation is applicable to the example in Figure~\ref{code:poisson_a}.
 
@@
 This would be challenging if we were to deal with arbitrary loop nests and expressions.
-However, we make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
+We make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
 As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors.
@@
 \label{sec:mem-const}
-
-In the previous section, we provided an insight into the potentially negative effects of code motion.
-In this section, we expand on this matter.
-The two following observations, in particular, lead to the definition of \textit{memory constraints}.
+In the previous section, we provided an insight into the potentially negative effects of code motion. We now expand on this matter, starting with the following observations.
 
@@
 
-Based on these considerations, we establish the following memory constraints.
+Based on these considerations, we establish two \textit{memory constraints}.
 
@@
 In this section, we build a transformation algorithm that, given a memory bound, produces optimal or quasi-optimal finite element integration loop nests.
-In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.
 
@@
 %numbers=left
-\begin{figure}[h]
+\begin{figure}
 \begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
@@
 If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
-Consider again the mass matrix in Figure~\ref{code:increase_factor}.
-This monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
-However, in general, $\rho$ depends on both form and discretization.
-If we instead take a standard Poisson equation using P1 Lagrange elements over triangles we have, for example, $\rho = 3$ (this can be verified by factorizing derivatives of basis functions as explained in Algorithm~\ref{algo:pre-evaluation}).
-To determine $\rho$ we execute the initial phase of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes reduction pre-evaluation.
+Consider again the mass matrix operator in Figure~\ref{code:increase_factor}.
+The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
+If we instead take the Poisson example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after suitable factorization of basis functions.
+In general, therefore, $\rho$ depends on both form and discretization employed.
+To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and count the terms amenable to pre-evaluation.
 
@@
 
-\subsection{Trading operation count for analysis cost}
-We can relax the constraints on (quasi-)optimality to relieve the analysis phase. In this section, we discuss ways of achieving this.
-\label{sec:obsheu}
-\begin{itemize}
-\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints.
-Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials.
-This matter is yet to be properly investigated.
-However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
-\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:sharing-elimination} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity.
-The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed).
-For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated.
-An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
-\end{itemize}
-
-%TODO forget about the special case alpha=beta (ie test and trial from same fs)
-
-These heuristics have been implemented in the code generation system described in Section~\ref{sec:codegen}.
-
 \section{Code Generation}
 \label{sec:codegen}
-Sharing elimination, pre-evaluation, the transformation algorithm, and all other features presented in Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} have been implemented in COFFEE, the optimizer for finite element integration routines used in Firedrake.
-In this section, we discuss implementation and code generation.
+Sharing elimination and pre-evaluation, as well as the transformation algorithm, have been implemented in COFFEE, the compiler for finite element integration routines adopted in Firedrake.
+In this section, we discuss a few aspects concerning implementation and features of the compiler.
 
 \subsection{Automation through the COFFEE language}
-COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''.
+COFFEE implements sharing elimination and pre-evaluation by composing ``building-block'' operators, which we refer to as ``rewrite operators''.
 This has several advantages.
 Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
-Secondly, generality: any sort of codes characterized by (linear, but also non-linear) loop nests with sharing can be optimized (obviously, without guarantees on optimality).
-In other words, COFFEE only exploits domain properties, and it is not specifically tied to finite element in any way.
+Secondly, generality: COFFEE can be seen as a tiny, low level computer algebra system, not specifically tied to finite element integration.
 Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
 
-The rewrite operators, which implementation is based on abstract syntax tree manipulation, compose the COFFEE language.
+The rewrite operators, whose (Python) implementation is based on manipulation of abstract syntax trees (ASTs), compose the COFFEE language.
 A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion.
-Sharing elimination and pre-evaluation are implemented by applying (composing) in specific sequences these operators.
+Sharing elimination and pre-evaluation are implemented by composing, in special ways, these operators.
 
 \subsection{Independence from form compilers}
+COFFEE is independent from the high level form compiler.
+COFFEE handles generic ASTs, so any form compiler can use it as long as such a representation is properly emitted (with expressions in normal form, or sufficiently close to it).
+In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs (instead of strings) is used.
+COFFEE itself provides an interface for building an AST.
+In particular, COFFEE aims to decouple the mathematical manipulation of a form from code optimization.
+Another viewpoint is that a form compiler developers should not worry about performance of the generated code.
+
+\subsection{Handling block-sparse tables}
 \label{sec:zeros}
-One of the key pillars of COFFEE is independence from form compilers.
-Being capable of handling generic abstract syntax trees, COFFEE can be leveraged by any suitably adapted form compiler.
-For example, in Firedrake a modified version of the FEniCS Form Compiler producing abstract syntax trees instead of strings is used.
-COFFEE itself provides an interface for building such abstract syntax trees.
-In particular, COFFEE aims to decouple the form manipulation (i.e., the ``math level'') from code optimization (flops executed, low level optimization such as vectorization).
-Another viewpoint is: developers of form compilers should not worry about expression optimization in any way, since this is handled at a lower level.
+Basis function tables may be block-sparse (e.g., containing zero-valued columns) depending on certain implementation choices at the level of the form compiler.
+For example, the FEniCS Form Compiler handles vector-valued function spaces by populating tabulated basis functions with blocks of zero-valued columns -- this makes code generation easier, but performance sub-optimal due to the execution of ``useless'' flops (e.g., by executing operations like \texttt{a + 0}).
 
-To achieve this goal, COFFEE needs a strategy to minimize the execution of potentially ``useless'' flops due to simplistic implementations of non scalar-valued function spaces.
-For example, the most trivial quadrature mode in the FEniCS Form Compiler implements vector-valued function spaces by populating tabulated basis functions with blocks of zero-valued columns.
-For the local element matrix/vector evaluation, these blocks are iterated over in a single, rectangular iteration space.
-This preserves the semantics of the computation while keeping the implementation complexity extremely low, although it also introduces unnecessary operations (e.g., zero-valued summands due to multiplications by zero).
-COFFEE achieves full independence from form compilers (and so relives the implementation burden) by providing a transformation to avoid the execution of such useless flops.
-A challenge for this transformation is restructuring the iteration spaces while preserving the effectiveness of low level optimizations, especially (auto-)vectorization.
-
-Consider a set of tabulated basis functions with quadrature points along rows and functions along columns.
-For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}.
+If, from one hand, it is true that form compilers usually provide optimization options to work around this issue, we must also consider the fact that not always their outcome is ideal for low level efficiency.
+Consider, for example, a set of tabulated basis functions with quadrature points along rows and functions along columns.
+\texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}.
 In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g.
 \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed.
-Our approach aims to avoid such indirection arrays in the generated code.
-This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.
+This technique, however, promotes non-contiguous memory loads and stores, which nullify the benefits of vectorization.
 
-The idea is that if the dimension along which vectorization is performed (typically the innermost) has a contiguous slice of zeros, but that slice is smaller than the vector length, then we do nothing (i.e., the loop nest is not transformed).
-Otherwise, we restructure the iteration space.
-This has several non-trivial implications.
-The most notable one is memory offsetting (e.g., \texttt{A[i+m,j+n]}), which dramatically enters in conflict with padding and data alignment.
-We use heuristics to retain the positive effects of both and to ensure correctness.
-Details are, however, beyond the scope of this paper.
+COFFEE provides a mechanism to restructure loops that preserves the effectiveness of low level optimization, such as vectorization.
+By executing the code symbolically, loop bounds are adjusted and memory offsets are introduced to minimize ``useless'' flops.
+This is done performing several checks to preserve the semantics of the computation whilst trying not to affect low level optimization (e.g., padding and data alignment are affected by choice of the offsets -- this is, however, out of the scope of this article).
 
-The implementation is based on symbolic execution: the loop nests are traversed and for each statement encountered the location of zeros in each of the involved symbols is tracked.
-Arithmetic operators have a different impact on tracking.
-For example, multiplication requires computing the set intersection of the zero-valued slices (for each loop dimension), whereas addition requires computing the set union.
+As discussed in the next section, all code generation systems used for performance evaluation handle block-sparse tables.
 
@@
 
-We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
-The axes represent $q$, $nf$, and code transformation system.
-We show one subplot for each problem instance $\langle form, nf, q\rangle$, with the code transformation system varying within each subplot.
-The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
-We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.
-
-The rest of the section is structured as follows: we provide insights about the main message of the experimentation; we comment on the impact of autovectorization; we explain in detail, individually for each form, the performance results obtained.
-
 \begin{figure}
@@
 \end{figure}
+
+We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
+The axes represent $q$, $nf$, and code transformation system.
+We show one subplot for each problem instance $\langle form, nf, q\rangle$, with the code transformation system varying within each subplot.
+The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
+We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.
+
+The rest of the section is structured as follows: we provide insights about the main message of the experimentation; we comment on the impact of autovectorization; we explain in detail, individually for each form, the performance results obtained.
 

### 2015-12-11 TOMS_Paper/assembly.tex
Add stuff to proof plus minor fixes

@@
 
-Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that characterize our expressions.
+Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural and algebraic properties that characterize our expressions.
 Section~\ref{sec:se-algo} presents the sharing elimination algorithm.
@@
 \item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
-\item $c_o$: the cost (i.e., number of operations) of an expression $o$
+\item $c_o^l$: the cost of an expression $o$ within a loop $L_l$; that is, the number of operations in $o$ times $|L_l|$.
 \item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the expression; a \textit{multilinear symbol} is a symbol appearing in a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, trial functions, first order derivatives of test functions, etc.)
@@
 
-We perform a second depth-first visit of the expression tree to collect all multilinear operands and partition them into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$.
-The multilinear operands in a partition $P_i$ share a set of multilinear symbols, whereas there is no sharing across different partitions.
+We perform a second depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, with $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$.
+The multilinear operands within a partition share the same set of multilinear symbols, whereas there is no sharing across different partitions.
 We have to establish:
 \begin{enumerate}
-\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, then $P_i$ is scheduled before $P_j$, and
-\item a scheduling strategy for each $P_i$
+\item an ordering for $\mathbb{P}$, such that if $P^i \prec P^j$, then $P^i$ is handled before $P^j$, and
+\item a scheduling strategy for each $P^i$
 \end{enumerate}
@@
 We start with point 1).
-We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{j}$ and $\mathbb{P}^{k}$, which logically separates partitions based on loop dependency.
-We then order $\mathbb{P}$ lexicographically, with the position of $P \in \mathbb{P}^j$ given by ${<}|\mathbb{P}^{j}|,\ |L_j| \sum_{o \in P} c_o{>}$ (and analogously for $P \in \mathbb{P}^k$).
+We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}_{j}$ and $\mathbb{P}_{k}$, which logically separates partitions based on loop dependency.
+We then build a reverse lexicographic ordering for $\mathbb{P}$, with the position of $P \in \mathbb{P}_l$, $l \in \lbrace j, k \rbrace$, given by ${<}|\mathbb{P}_{l}|,\ \sum_{o \in P} c_o^l{>}$.
 Intuitively, this simply means that operands appearing more often in the expression are handled first.
@@
 \item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
-Assume that $P$ depends on $[L_i, L_k]$.
-The increase in operation count at the level of the outer loops is: if (i) were applied, $|L_i| (|L_k| \sum_{o \in P} c_{o} + \sum_{o \in P} (n_{o} - 1))$; if (ii) were applied, $|L_i| \sum_{s \in P_S} (n_{s} - 1)$.
+The increase in operation count at the level of the outer loops is: if (i) were applied, $\sum_{o \in P_l} (c_{o}^{l} + n_{o} - 1)$; if (ii) were applied, $\sum_{s \in S_P} (n_{s} - 1)$.
 We pick (i) or (ii) such that the increase is minimum.
-Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that they both minimize the cost of the innermost loop.
+Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that both will eventually minimize the innermost loop cost.
 \end{enumerate}
@@
 Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D.
-We observe that $P_0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
-Clearly, $\mathbb{P} = \lbrace P_0, P_1 \rbrace$, with $\mathbb{P}^k = \lbrace P_0 \rbrace$ and $\mathbb{P}^j = \lbrace P_1 \rbrace$.
+We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
 A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
@@
 We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
-Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.
+Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{algo:gamma}.
 
@@
 
-// Find potential optimum such that memory constraints are honored
+// Find global quasi-optimum honoring memory constraints
 B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
@@
 \caption{High level view of the transformation algorithm}
-\label{code:intuition}
+\label{algo:gamma}
 \end{figure}
@@
 
-\begin{figure}\begin{CenteredBox}
+\begin{figure}[b]\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
@@
 
-\begin{figure}[b]\begin{CenteredBox}
+\begin{figure}[t]\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
@@
 \label{sec:proof}
-The following proposition states that our approach guarantees quasi-optimality. The proof re-uses most of the concepts and explanations provided throughout the paper.
+The following proposition states that our approach guarantees quasi-optimality. The proof re-uses some of the concepts and explanations provided throughout the paper.
 
@@
 \label{prop:optimal-approach}
-Consider a multilinear form and the corresponding finite element integration loop nest $\Lambda$ implementing it.
-The form comprises a set of monomials, $M$.
+Consider a multilinear form comprising a set of monomials $M$, and let $\Lambda$ be the corresponding finite element integration loop nest.
+Let $\Gamma$ be the transformation algorithm.
 Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$.
 Assume that pre-evaluating different monomials does not result in identical tables.
-Then, given the memory Constraints~\ref{const:Le} and~\ref{const:TH}, the transformation algorithm produces a $\Lambda'$ which is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
+Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
 \end{Prop}
 \begin{proof}
-The function $\theta$ provides cost or gain of pre-evaluation.
-$\theta$ retrieves exact values, rather than estimates\footnote{An estimate is used, instead, in~\citeN{Kirby:TC}.
-The estimate is actually for the whole form, rather than for its monomials.}, so $P$ and $Z$ are constructed correctly.
+We first observe that the cost function $\theta$ predicts the exact gain/loss in monomial pre-evaluation, so $P$ is always correct by construction.
 
-Let $\Lambda_{ops}$ denote the operation count (number of additions and products) in $\Lambda$.
-Let $\Lambda'_I \subset \Lambda'$ be the set of all innermost loops produced by the transformation algorithm (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
-We need to show that there is no other synthesis $\Lambda''_{I}$ such that $\Lambda''_{I_{ops}} < \Lambda'_{I_{ops}}$, which holds if and only if
+Let $c_\Lambda$ denote the operation count in $\Lambda$.
+Let $\Lambda'_I \subset \Lambda'$ be the subset of all innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
+We need to show that there is no other synthesis $\Lambda''_{I}$ such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
+This holds if and only if
 \begin{enumerate}
@@
 \begin{enumerate}
-\item \textit{pre-evaluating any $Z_P: Z_P \subseteq Z$ would result $\Lambda''_{ops} > \Lambda'_{ops}$}
-\item \textit{not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops} > \Lambda'_{ops}$}
+\item \textit{pre-evaluating any $Z_P \subseteq Z$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
+\item \textit{not pre-evaluating any $P_Z \subseteq P$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
 \end{enumerate}
-~
+\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces (at least) quasi-optimal loop nests.}\\
+\end{enumerate}
+
+We discuss these points separately
+
+\begin{enumerate}
+\item 
 \begin{enumerate}
 \item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$.
-Consider two generic monomials $m_1, m_2 \in M$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
-In a scenario in which $m_{T_{1}} \cap m_{T_{2}}$, we would have that at least one table would be assigned to the same temporary, with the transformation algorithm skipping a potential global optimum.
-By assumption, we neglect this possibility.
-The assumption is sustained by (i) the pathological nature of the scenario, (ii) the complexity inherent in handling it, and (iii) the fact that such a potential global optimum would probably be very close to the retrieved solution.
+Consider two generic monomials $m_1, m_2 \in Z$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
+If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary.
+$\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis).
+By assumption, this case is neglected.
+The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
+\item Let $p \in P$ and $z \in Z$ two monomials that share some subsets of multilinear symbols.
+If, as by assumption $\theta(p)$ suggests, $p$ will be pre-evaluated, there may be a potential gain in sharing elimination going lost, which potentially leads to a locally optimal solution -- instead of a global one.
+We avoid this situation by construction.
+$\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}).
+Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
+\end{enumerate}
 \item ...
-\end{enumerate}
-
-\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces loop nests that are at least quasi-optimal.}\\
-...
 \end{enumerate}

### 2015-12-12 TOMS_Paper/assembly.tex
Several minor fixes

@@
 \usepackage{fancybox}
+\usepackage{enumitem}
 
@@
 \begin{figure}[h]\begin{CenteredBox}
-{\subfigcapskip = 11pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
+{\subfigcapskip = 13pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
 ~~~~~
-{\subfigcapskip = 11pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
+{\subfigcapskip = 13pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
 ~~~~~
-{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_c}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
+{\subfigcapskip = 4pt \subfigure[Optimal form]{\label{code:multi_loopnest_c}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
 \end{CenteredBox}\caption{Reducing a simple multilinear loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}
@@
 To schedule this expression, we basically have two options:
-\begin{enumerate}
-\item We leave the expression unmodified in the body of $L_i$ (in general, one may consider applying generalized code motion or creating a temporary for CCSE -- but it is not the case of this simple example).
-\item We notice the presence of spatial sharing. We can recast the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
+\begin{enumerate}[label=(\roman*)]
+\item No transformation takes place. In general, one should consider applying generalized code motion or creating a temporary for CCSE -- but it is not the case of this simple example.
+\item Spatial sharing is found, so the expression is recast as $f_i (a + c) + g_i (b + d)$. This promotes temporal sharing and, therefore, code motion. 
 \end{enumerate}
@@
 
-\begin{Algo}[Sharing elimination]
-\label{algo:sharing-elimination}
-\normalfont
-We introduce the following notation and terminology
+The algorithm makes use of the following notation and terminology:
 \begin{itemize}
@@
 
+\begin{Algo}[Sharing elimination]
+\label{algo:sharing-elimination}
+\normalfont
 The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
@@
 The multilinear operands within a partition share the same set of multilinear symbols, whereas there is no sharing across different partitions.
-We have to establish:
-\begin{enumerate}
-\item an ordering for $\mathbb{P}$, such that if $P^i \prec P^j$, then $P^i$ is handled before $P^j$, and
-\item a scheduling strategy for each $P^i$
-\end{enumerate}
 
-We start with point 1).
-We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}_{j}$ and $\mathbb{P}_{k}$, which logically separates partitions based on loop dependency.
-We then build a reverse lexicographic ordering for $\mathbb{P}$, with the position of $P \in \mathbb{P}_l$, $l \in \lbrace j, k \rbrace$, given by ${<}|\mathbb{P}_{l}|,\ \sum_{o \in P} c_o^l{>}$.
+We build a reverse lexicographic ordering for $\mathbb{P}$, with the position of $P$ given by ${<}|\mathbb{P}_{l}|,\ \sum_{o \in P} c_o^l{>}$, in which $\mathbb{P}_{l}$ represents the set of partitions depending on loop $L_l$.
 Intuitively, this simply means that operands appearing more often in the expression are handled first.
@@
 
-We now tackle point 2).
-For each partition, we have two options, as already explained in Section~\ref{sec:se-rln}.
-We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE; (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion.
-We adopt the following criteria:
+For each partition, we need choose the optimal strategy between the points (i) and (ii) described in Section~\ref{sec:se-rln}. We adopt the following criteria.
 \begin{enumerate}
 \item Let $S_P$ be the set of multilinear symbols in $P$.
-If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops.
+If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we go on to criterion 2), which attempts to optimize for the outer loops.
 \item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
 The increase in operation count at the level of the outer loops is: if (i) were applied, $\sum_{o \in P_l} (c_{o}^{l} + n_{o} - 1)$; if (ii) were applied, $\sum_{s \in S_P} (n_{s} - 1)$.
-We pick (i) or (ii) such that the increase is minimum.
+We then pick (i) or (ii) such that the increase is minimum.
 Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that both will eventually minimize the innermost loop cost.
@@
 
-It remains to schedule $L_i$-dependent expressions.
+Eventually, $L_i$-dependent expressions are scheduled.
 We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above.
 In doing this, we speculatively assume that all expressions result from structured tensor operations.
-If this were not the case, factorization would be applied greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
+If this were not the case, factorization would be applied greedily (i.e., some code motion opportunities will be exposed, but this may lead to a local optimum).
 \end{Algo}
@@
 We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
-A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
+A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
 \begin{figure}\begin{CenteredBox}
@@
 
-\begin{figure}[h]\begin{CenteredBox}
+\begin{figure}\begin{CenteredBox}
 {\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
 ~~~~~~~~~~
-{\subfigcapskip = 5pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
+{\subfigcapskip = 5pt \subfigure[After pre-evaluation]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
 \end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\label{code:loopnest_rednored}\end{figure}

### 2015-12-12 TOMS_Paper/assembly.tex
Several minor fixes

@@
 \end{enumerate}
-\item \textit{Sharing elimination (Algorithm~\ref{algo:sharing-elimination}) produces (at least) quasi-optimal loop nests.}\\
+\item \textit{Sharing elimination produces (at least) quasi-optimal loop nests.}\\
 \end{enumerate}
@@
 \item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$.
-Consider two generic monomials $m_1, m_2 \in Z$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
+Consider two monomials $m_1, m_2 \in Z$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
 If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary.
@@
 The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
-\item Let $p \in P$ and $z \in Z$ two monomials that share some subsets of multilinear symbols.
-If, as by assumption $\theta(p)$ suggests, $p$ will be pre-evaluated, there may be a potential gain in sharing elimination going lost, which potentially leads to a locally optimal solution -- instead of a global one.
-We avoid this situation by construction.
+\item Let $p \in P$ and $z \in Z$ be two monomials that share some generic multilinear symbols.
+If $p$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one.
+We prevent this situation by construction.
 $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}).
@@
 \end{enumerate}
-\item ...
+\item Consider Algorithm~\ref{algo:sharing-elimination}. Partitions are ordered to enforce a specific sequence of factorizations. 
 \end{enumerate}

### 2015-12-12 TOMS_Paper/assembly.tex
Still working on the algos...

@@
 \begin{itemize}
-\item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
-\item $c_o^l$: the cost of an expression $o$ within a loop $L_l$; that is, the number of operations in $o$ times $|L_l|$.
-\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the expression; a \textit{multilinear symbol} is a symbol appearing in a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, trial functions, first order derivatives of test functions, etc.)
+\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop)
+\item $c_o^l$: the cost of an expression $o$ in a loop $L_l$ (i.e., operations in $o$ times $|L_l|$).
+\item $n_o$: the number of occurrences of a sub-expression $o$ appears in the input expression.
+\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the expression in normal form;
+\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.)
 \end{itemize}
@@
 
-We build a reverse lexicographic ordering for $\mathbb{P}$, with the position of $P$ given by ${<}|\mathbb{P}_{l}|,\ \sum_{o \in P} c_o^l{>}$, in which $\mathbb{P}_{l}$ represents the set of partitions depending on loop $L_l$.
-Intuitively, this simply means that operands appearing more often in the expression are handled first.
-We will later elaborate on this.
+We build a reverse lexicographic ordering for $\mathbb{P}$, in which the value of $P$ is given by ${<} ...,\ \sum_{o \in P} c_o^l{>}$.  ...
 
-For each partition, we need choose the optimal strategy between the points (i) and (ii) described in Section~\ref{sec:se-rln}. We adopt the following criteria.
+%By handling more costly operands first, the operation count achieved by temporal sharing elimination will be minimized.
+
+For each partition, we then need choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}. We adopt the following criteria.
 \begin{enumerate}
@@
 If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we go on to criterion 2), which attempts to optimize for the outer loops.
-\item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression.
-The increase in operation count at the level of the outer loops is: if (i) were applied, $\sum_{o \in P_l} (c_{o}^{l} + n_{o} - 1)$; if (ii) were applied, $\sum_{s \in S_P} (n_{s} - 1)$.
+\item The increase in operation count at the level of the outer loops is: if (i) were applied, $\sum_{o \in P_l} (c_{o}^{l} + n_{o} - 1)$; if (ii) were applied, $\sum_{s \in S_P} (n_{s} - 1)$.
 We then pick (i) or (ii) such that the increase is minimum.
@@
 Let $\Gamma$ be the transformation algorithm.
-Let $P$ be the set of pre-evaluated monomials, and let $Z = M \setminus P$.
+Let $X$ be the set of pre-evaluated monomials, and let $Y = M \setminus X$.
 Assume that pre-evaluating different monomials does not result in identical tables.
@@
 \begin{proof}
-We first observe that the cost function $\theta$ predicts the exact gain/loss in monomial pre-evaluation, so $P$ is always correct by construction.
+We first observe that the cost function $\theta$ predicts the exact gain/loss in monomial pre-evaluation, so $X$ is always correct by construction.
 
@@
 \begin{enumerate}
-\item \textit{pre-evaluating any $Z_P \subseteq Z$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
-\item \textit{not pre-evaluating any $P_Z \subseteq P$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
+\item \textit{pre-evaluating any $m \in Y$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
+\item \textit{not pre-evaluating any $m \in X$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
 \end{enumerate}
@@
 \item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$.
-Consider two monomials $m_1, m_2 \in Z$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
+Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
 If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary.
@@
 The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
-\item Let $p \in P$ and $z \in Z$ be two monomials that share some generic multilinear symbols.
-If $p$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one.
+\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols.
+If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one.
 We prevent this situation by construction.
@@
 \end{enumerate}
-\item Consider Algorithm~\ref{algo:sharing-elimination}. Partitions are ordered to enforce a specific sequence of factorizations. 
+\item Consider Algorithm~\ref{algo:sharing-elimination}.
+Partitions are ordered to promote a specific factorization sequence; this is essential to minimize the operation count in outer loops, although it is non influential for quasi-optimality.
+Criterion (2) also does not affect innermost loops.
+Criterion (1) is instead a discriminating factor.
+Consider a generic partition $P \in \mathbb{P}$.
+multilinearity...
 \end{enumerate}
 
-%
-%A) The only corner situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$.
-If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim.
-Assumption 1) prevents this situation, so the statement is true.
-%
-%B) Let $\theta_{wo}(m)$ and $\theta_{w}(m)$ represent, respectively, the operation count for a monomial $m$ without and with pre-evaluation and let us consider the non-trivial case in which two monomials $p \in P_Z$ and $z \in Z$ share at least one term.
-We have to contemplate the fact that despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle.
-That is, there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation.
-Potential examples are forms in which multiple monomials differ only for the number and/or the space of coefficient functions.
-The previously discussed assumption 2) allows us to neglect this possibility.
 \end{proof}
@@
 %TODO : In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
+
+% ordering of mathbb{P} is fundamental, because you start with something, then you collect stuff, and what you've collected is gonna be hoisted, which means having made a choice in terms of minimizing the innermost loop cost --- multilinearity and normal form here are super super imporant !
+% |P| <, ==, > then |S_P| is important because think of like: z0*FEik .... F0*FEik .... F1*FEik ...
 

### 2015-12-14 TOMS_Paper/assembly.tex
minor changes...

@@
 \begin{itemize}
-\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop)
+\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
 \item $c_o^l$: the cost of an expression $o$ in a loop $L_l$ (i.e., operations in $o$ times $|L_l|$).
-\item $n_o$: the number of occurrences of a sub-expression $o$ appears in the input expression.
-\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the expression in normal form;
-\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.)
+\item $n_o$: the number of occurrences of a sub-expression $o$ in the input expression.
+\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
+\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
 \end{itemize}
@@
 \normalfont
-The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
+The algorithm has three main steps: initialization (step 1); scheduling of multilinear operands, to reach quasi-optimality (steps 2-4); scheduling of all other sub-expressions (step 5).
+\begin{enumerate}
+\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
 For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
-
-We perform a second depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, with $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$.
-The multilinear operands within a partition share the same set of multilinear symbols, whereas there is no sharing across different partitions.
-
-We build a reverse lexicographic ordering for $\mathbb{P}$, in which the value of $P$ is given by ${<} ...,\ \sum_{o \in P} c_o^l{>}$.  ...
-
-%By handling more costly operands first, the operation count achieved by temporal sharing elimination will be minimized.
-
-For each partition, we then need choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}. We adopt the following criteria.
-\begin{enumerate}
-\item Let $S_P$ be the set of multilinear symbols in $P$.
-If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we go on to criterion 2), which attempts to optimize for the outer loops.
-\item The increase in operation count at the level of the outer loops is: if (i) were applied, $\sum_{o \in P_l} (c_{o}^{l} + n_{o} - 1)$; if (ii) were applied, $\sum_{s \in S_P} (n_{s} - 1)$.
-We then pick (i) or (ii) such that the increase is minimum.
-Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that both will eventually minimize the innermost loop cost.
+\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$.
+$\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols, whereas there is no sharing across different partitions.
+Let $S_P$ be the set of multilinear symbols shared in $P$.
+\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}\sum_{o \in P} n_o,\ \sum_{s \in S_P} n_s{)}$.
+Intuitively, $c_P$ indicates the number of operations induced by $P$ at every innermost loop iteration once sharing elimination will be applied.
+Take $P$ such that $c_P$ is minimum.
+\item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}.
+\begin{enumerate}[label=(\Alph*)]
+\item If $|P| < |S_P|$, (i) is applied.
+If $|P| > |S_P|$, (ii) is applied.
+If $|P| = |S_P|$, go to 4B.
+Intuitively, this means that (i) and (ii) have the same impact on the innermost loop, but the increase in operation count in the outer loops, $i_l$, may be different.
+\item If (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$. If (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$. We pick (i) or (ii) such that $i_l$ is minimum. 
 \end{enumerate}
-
-Eventually, $L_i$-dependent expressions are scheduled.
-We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above.
-In doing this, we speculatively assume that all expressions result from structured tensor operations.
-If this were not the case, factorization would be applied greedily (i.e., some code motion opportunities will be exposed, but this may lead to a local optimum).
+\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
+\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination (as in point 4B above) assuming a structured tensor operation.
+The result may clearly be sub-optimal if the assumption does not hold in practice (i.e., only a subset of all possible code motion opportunities are exposed).
+\end{enumerate}
 \end{Algo}

### 2015-12-14 TOMS_Paper/assembly.tex
Adding stuff to algo and proof...

@@
 Let $S_P$ be the set of multilinear symbols shared in $P$.
-\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}\sum_{o \in P} n_o,\ \sum_{s \in S_P} n_s{)}$.
+\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}|P|,\ |S_P|{)}$.
 Intuitively, $c_P$ indicates the number of operations induced by $P$ at every innermost loop iteration once sharing elimination will be applied.
@@
 \item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}.
-\begin{enumerate}[label=(\Alph*)]
-\item If $|P| < |S_P|$, (i) is applied.
+If $|P| < |S_P|$, (i) is applied.
 If $|P| > |S_P|$, (ii) is applied.
-If $|P| = |S_P|$, go to 4B.
-Intuitively, this means that (i) and (ii) have the same impact on the innermost loop, but the increase in operation count in the outer loops, $i_l$, may be different.
-\item If (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$. If (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$. We pick (i) or (ii) such that $i_l$ is minimum. 
-\end{enumerate}
+Otherwise, $|P| = |S_P|$ means that (i) and (ii) have the same impact on the innermost loop.
+However, the increase in operation count in the outer loops, $i_l$, may still be different.
+In particular, if (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$, while if (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$.
+Note that $n_o$ and $n_s$ capture the number of additions required after collection of $o$ and $s$.
+We then select (i) or (ii) such that $i_l$ is minimum.
 \item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
-\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination (as in point 4B above) assuming a structured tensor operation.
-The result may clearly be sub-optimal if the assumption does not hold in practice (i.e., only a subset of all possible code motion opportunities are exposed).
+\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above).
+The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
 \end{enumerate}
@@
 Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D.
-We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
+We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$, with $c_{P^0} = c_{P^1} = {min(}2, 2{)} = 2$.
 A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
-\begin{figure}\begin{CenteredBox}
+\begin{figure}[h]\begin{CenteredBox}
 {\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
@@
 
-\begin{figure}\begin{CenteredBox}
+\begin{figure}[h]\begin{CenteredBox}
 {\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Partitions are ordered to promote a specific factorization sequence; this is essential to minimize the operation count in outer loops, although it is non influential for quasi-optimality.
+The operation count induced by a partition $P$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$.
+This descends directly from two facts: the assumption that the input expression is in normal form and multilinearity.
+Therefore, what we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality.
+We do it by contradiction.
+Let us assume that taking any $P$ such that $c_{P} > c_{P_{min}}$ would lead to a lower operation count.
+We first observe that a partition (i.e., its multilinear symbols) depends on one of the two multilinear loops; we then indicate with $P^l$ the fact that $P$ depends on $L_l$, $l \in \lbrace j, k\rbrace$.
+Assuming $P_{min}^l$, we have two possibilities:
+\begin{itemize}
+\item likewise, $P^{l}$; that is, both $P_{min}$ and $P$ depend on the same linear loop.
+This is the simplest case, since swapping the order in which the two partitions are handled would not have any side effects -- there is no clash between all possible different re-factorizations.
+The assumption is then contradicted because the operation count in the innermost loop is the same irrespective of starting with $P_{min}$ or $P$.
+\item $P^{l'}$ and $l' \neq l$.
+In this case, it is possible that operands (symbols) in $P^{l'}$ are factorized and code motion takes place such that $P_{min}$ ``loses'' some of its operands (symbols).
+Linearity and the assumption $c_{P} > c_{P_{min}}$ tell us that once $P$ is scheduled, we will have $c_{\Lambda''_I} \geq c_{\Lambda'_I} + 1$; that is, there is at least one extra operation (an addition) due to
+\end{itemize}
+
+\end{enumerate}
+
+%Partitions are ordered to promote a specific factorization sequence; this is essential to minimize the operation count in outer loops, although it is non influential for quasi-optimality.
 Criterion (2) also does not affect innermost loops.
@@
 multilinearity...
-\end{enumerate}
 
 \end{proof}
+
+%TODO : even simpler if we considered linera forms. If we had trilinear forms, we would be fucked
 

### 2015-12-15 TOMS_Paper/assembly.tex
minor fixes

@@
 \normalfont
-The algorithm has three main steps: initialization (step 1); scheduling of multilinear operands, to reach quasi-optimality (steps 2-4); scheduling of all other sub-expressions (step 5).
+The algorithm has three main steps: initialization (step 1); scheduling of multilinear operands, to reach quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
 \begin{enumerate}
@@
 \item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$.
-$\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols, whereas there is no sharing across different partitions.
-Let $S_P$ be the set of multilinear symbols shared in $P$.
+$\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
 \item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}|P|,\ |S_P|{)}$.
@@
 In particular, if (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$, while if (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$.
-Note that $n_o$ and $n_s$ capture the number of additions required after collection of $o$ and $s$.
+Note that $n_o$ and $n_s$ capture the number of additions required after factorization of $o$ and $s$.
 We then select (i) or (ii) such that $i_l$ is minimum.

### 2015-12-16 TOMS_Paper/assembly.tex
minor fixes

@@
 \normalfont
-The algorithm has three main steps: initialization (step 1); scheduling of multilinear operands, to reach quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
+The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
 \begin{enumerate}
@@
 Take $P$ such that $c_P$ is minimum.
+%; if multiple partitions have same cost, the preference is for the one with the largest occurrences in the input expression.
 \item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}.
@@
 The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
+%TODO : this needs to be applied to all lifted sub-expressions as well
 \end{enumerate}
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-The operation count induced by a partition $P$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$.
-This descends directly from two facts: the assumption that the input expression is in normal form and multilinearity.
-Therefore, what we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality.
+Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$.
+What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality.
 We do it by contradiction.
-Let us assume that taking any $P$ such that $c_{P} > c_{P_{min}}$ would lead to a lower operation count.
-We first observe that a partition (i.e., its multilinear symbols) depends on one of the two multilinear loops; we then indicate with $P^l$ the fact that $P$ depends on $L_l$, $l \in \lbrace j, k\rbrace$.
+Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$.
+In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved.
+We distinguish two cases.
+\begin{itemize}
+\item $|P| < |S_P|$.
+This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place.
+We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
+\item $|P| \geq |S_P|$.
+The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected.
+Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion.
+It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$.
+Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions).
+Conversely, if we scheduled $P_{min}$ first, the increase would be -- assuming, without loss of generality, that $|S_{P_{min}}| = n - 1$ -- equal to $2(n - 1) - 1$.
+\end{itemize}
+
+%In total, therefore, 
+%If none of the $n$ hoisted sub-expressions include operands in $P_{min}$, we have again $c_{\Lambda''_I} = c_{\Lambda'_I}$. If that is not the case,
+
+%We first observe that a partition (i.e., its multilinear symbols) depends on one of the two multilinear loops; we then indicate with $P^l$ the fact that $P$ depends on $L_l$, $l \in \lbrace j, k\rbrace$.
 Assuming $P_{min}^l$, we have two possibilities:
-\begin{itemize}
-\item likewise, $P^{l}$; that is, both $P_{min}$ and $P$ depend on the same linear loop.
+%\begin{itemize}
+%\item likewise, $P^{l}$; that is, both $P_{min}$ and $P$ depend on the same linear loop.
 This is the simplest case, since swapping the order in which the two partitions are handled would not have any side effects -- there is no clash between all possible different re-factorizations.
 The assumption is then contradicted because the operation count in the innermost loop is the same irrespective of starting with $P_{min}$ or $P$.
-\item $P^{l'}$ and $l' \neq l$.
+%\item $P^{l'}$ and $l' \neq l$.
 In this case, it is possible that operands (symbols) in $P^{l'}$ are factorized and code motion takes place such that $P_{min}$ ``loses'' some of its operands (symbols).
 Linearity and the assumption $c_{P} > c_{P_{min}}$ tell us that once $P$ is scheduled, we will have $c_{\Lambda''_I} \geq c_{\Lambda'_I} + 1$; that is, there is at least one extra operation (an addition) due to
-\end{itemize}
+%\end{itemize}
 

### 2015-12-16 TOMS_Paper/assembly.tex
minor fixes

@@
 \begin{enumerate}[label=(\roman*)]
-\item No transformation takes place. In general, one should consider applying generalized code motion or creating a temporary for CCSE -- but it is not the case of this simple example.
-\item Spatial sharing is found, so the expression is recast as $f_i (a + c) + g_i (b + d)$. This promotes temporal sharing and, therefore, code motion. 
+\item Applying generalized code motion (or simply creating a temporary for CCSE) -- not necessary in this example since there is just one loop.
+\item Searching for spatial sharing, expanding and then factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
 \end{enumerate}
-The reader can verify that option 2) improves the operation count over 1) if and only if the size of $L_i$ is strictly greater than $c=1$.
-One can appreciate the dichotomy between 1) and 2) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with different values of $c$.
-
+The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$.
+In general, however, the optimal option depends on both the expansion cost in (ii) and the loop size.
+One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}.
 We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.
+
 Of notable importance, in the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations.
@@
 For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
-\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$.
+\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
 $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
-\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}|P|,\ |S_P|{)}$.
-Intuitively, $c_P$ indicates the number of operations induced by $P$ at every innermost loop iteration once sharing elimination will be applied.
-Take $P$ such that $c_P$ is minimum.
-%; if multiple partitions have same cost, the preference is for the one with the largest occurrences in the input expression.
-\item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}.
-If $|P| < |S_P|$, (i) is applied.
-If $|P| > |S_P|$, (ii) is applied.
-Otherwise, $|P| = |S_P|$ means that (i) and (ii) have the same impact on the innermost loop.
-However, the increase in operation count in the outer loops, $i_l$, may still be different.
-In particular, if (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$, while if (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$.
-Note that $n_o$ and $n_s$ capture the number of additions required after factorization of $o$ and $s$.
-We then select (i) or (ii) such that $i_l$ is minimum.
-\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
+\item Calculate the operation count induced by applying strategy (i) or (ii) -- as described in Section~\ref{sec:se-rln} -- to $P \in \mathbb{P}$.
+In particular: $c_P^i = |P||\Lambda| + \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$ and $c_P^{ii} = |S_P||\Lambda| + \sum_{s \in S_P} (n_{s} - 1)$.
+Note that:
+\begin{itemize}
+\item $|P||\Lambda|$ ($|S_P||\Lambda|$) represents the operation count in the innermost loop
+\item $n_o$ ($n_s$) captures the number of additions once $o$ ($s$) is factorized
+\end{itemize}
+We define the cost of $P$ as $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$.
+\item For all $P \in \mathbb{P}$ such that $c_P = c_P^i$, for all multilinear operands $o \in P$, apply generalized code motion.
+Update $\mathbb{P}$ by replacing any hoisted multilinear operand with the corresponding temporary symbol.
+\item For all $P \in \mathbb{P}$ such that $c_P^{ii} - |S_P||\Lambda| > 0$ -- that is, 
 \item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above).
@@
 \end{Algo}
+
+%If there exist $P^i$, $P^j$, and $P^k$
+%\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
 
@@
 Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions).
-Conversely, if we scheduled $P_{min}$ first, the increase would be -- assuming, without loss of generality, that $|S_{P_{min}}| = n - 1$ -- equal to $2(n - 1) - 1$.
+Conversely, assuming $|S_{P_{min}}| = n - 1$, if we scheduled $P_{min}$ first, .
 \end{itemize}

### 2015-12-16 TOMS_Paper/assembly.tex
Change the algo quite a lot

@@
 
-In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, generalized code motion and (if necessary) common sub-expression elimination.
+In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion.
 If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
@@
 
-Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural and algebraic properties that characterize our expressions.
+Section~\ref{sec:se-rln} discusses structural and algebraic properties characterizing our expressions.
 Section~\ref{sec:se-algo} presents the sharing elimination algorithm.
+Examples are provided in Section~\ref{sec:se-examples}.
 We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form.
-
-\subsubsection{Common sub-expressions}
-\label{sec:se-mln}
-A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE).
-This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that are (i) symbolically identical and (ii) reduce to the same value.
-We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block.
-More interestingly, we observe that in the ``special case'' (SCSE) in which test and trial functions are in the same function space, there may exist $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$, in which case common sub-expressions will arise along different loops ($L_j$ and $L_k$).
-General-purpose compilers cannot detect this (very domain-specific) property.
-Our algorithm for sharing elimination will exploit this extra information.
-Although both CSE and factorization target spatial sharing, the latter may actually increase the operation count if preliminary transformations (e.g., expansion) are required.
-In such a case, CCSE will be preferred.
 
@@
 \begin{enumerate}[label=(\roman*)]
-\item Applying generalized code motion (or simply creating a temporary for CCSE) -- not necessary in this example since there is just one loop.
-\item Searching for spatial sharing, expanding and then factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
+\item Applying generalized code motion -- not needed in the example as we have one loop.
+\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
 \end{enumerate}
@@
 $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
-\item Calculate the operation count induced by applying strategy (i) or (ii) -- as described in Section~\ref{sec:se-rln} -- to $P \in \mathbb{P}$.
-In particular: $c_P^i = |P||\Lambda| + \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$ and $c_P^{ii} = |S_P||\Lambda| + \sum_{s \in S_P} (n_{s} - 1)$.
-Note that:
-\begin{itemize}
-\item $|P||\Lambda|$ ($|S_P||\Lambda|$) represents the operation count in the innermost loop
-\item $n_o$ ($n_s$) captures the number of additions once $o$ ($s$) is factorized
-\end{itemize}
-We define the cost of $P$ as $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$.
-\item For all $P \in \mathbb{P}$ such that $c_P = c_P^i$, for all multilinear operands $o \in P$, apply generalized code motion.
-Update $\mathbb{P}$ by replacing any hoisted multilinear operand with the corresponding temporary symbol.
-\item For all $P \in \mathbb{P}$ such that $c_P^{ii} - |S_P||\Lambda| > 0$ -- that is, 
-\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above).
-The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
-%TODO : this needs to be applied to all lifted sub-expressions as well
+\item Calculate the number of operations in the innermost loop required for the execution of $P \in \mathbb{P}$.
+There are two possibilities, depending on which strategy -- (i) or (ii), as described in Section~\ref{sec:se-rln} -- is employed.
+In particular, we have $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$, with $c_P^i = |P|$ and $c_P^{ii} = |S_P|$.
+\item For all multilinear operands $o \in P$ such that $c_P = c_P^i$, apply strategy (i). Update $\mathbb{P}$ by replacing hoisted multilinear operands with the corresponding temporary symbols.
+\item For all $P \in \mathbb{P}$, determine the profitability of strategy (ii).
+We introduce $g_P$, which measures the gain in operation count in the innermost loop if (ii) were applied.
+That is, $g_P = \sum_{Q \in \mathbb{Q}} c_{Q_{old}} - c_{Q_{new}}$, with $\mathbb{Q} = \mathbb{P} \setminus \lbrace P \rbrace$.
+Note that $c_{Q_{new}}$ can be evaluated without applying the actual transformation.
+Eventually, apply (ii) if $g_P > 0$.
+\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent or previously hoisted sub-expression, apply (i) or (ii).
+This pass speculatively assumes that structured tensor operations are in place.
+If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may have been exposed.
 \end{enumerate}
@@
 %\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
+%|\Lambda| + \sum_{s \in S_P} (n_{s} - 1)$. Note that:
+%\begin{itemize}
+%\item $|P||\Lambda|$ ($|S_P||\Lambda|$) represents the operation count in the innermost loop
+%\item $n_o$ ($n_s$) captures the number of additions once $o$ ($s$) is factorized
+%\end{itemize}
+%We define the cost of $P$ as $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$.
 
 \subsubsection{Examples}
+\label{sec:se-examples}
 Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D.
@@
 \label{sec:pre-evaluation}
-Sharing elimination explores the transformation space and applies three operators: factorization, code motion and common sub-expression elimination.
+Sharing elimination explores the transformation space and applies three operators: expansion, factorization, and code motion.
 In this section, we discuss role and legality of a fourth operator: \textit{reduction pre-evaluation}.

### 2015-12-16 TOMS_Paper/assembly.tex
Other small fixes

@@
 Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}).
-Finally, generalized code motion (\cite{Luporini}) leads to optimality (Figure~\ref{code:multi_loopnest_c}).
+Finally, generalized code motion \cite{Luporini} leads to optimality (Figure~\ref{code:multi_loopnest_c}).
 
@@
 If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
-Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This require coordination with other transformations, as discussed in the following sections.
-For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved if stronger hypotheses hold.
+Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
+For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold.
 Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
@@
 Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$.
-For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$.
+For example, suppose that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$.
 Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$.
@@
 The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$.
-In general, however, the optimal option depends on both the expansion cost in (ii) and the loop size.
+In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and the factorization gain.
 One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}.
@@
 
-Of notable importance, in the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations.
+Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations.
 Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too.
@@
 \item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
-\item $c_o^l$: the cost of an expression $o$ in a loop $L_l$ (i.e., operations in $o$ times $|L_l|$).
-\item $n_o$: the number of occurrences of a sub-expression $o$ in the input expression.
 \item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
@@
 \end{Algo}
-
-%If there exist $P^i$, $P^j$, and $P^k$
-%\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
-%|\Lambda| + \sum_{s \in S_P} (n_{s} - 1)$. Note that:
-%\begin{itemize}
-%\item $|P||\Lambda|$ ($|S_P||\Lambda|$) represents the operation count in the innermost loop
-%\item $n_o$ ($n_s$) captures the number of additions once $o$ ($s$) is factorized
-%\end{itemize}
-%We define the cost of $P$ as $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$.
 
@@
 
-\begin{figure}[b]\begin{CenteredBox}
+\begin{figure}[t]\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
@@
 Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$.
-What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality.
+
+%What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality.
 We do it by contradiction.
@@
 We distinguish two cases.
-\begin{itemize}
-\item $|P| < |S_P|$.
+%\begin{itemize}
+%\item $|P| < |S_P|$.
 This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place.
 We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
-\item $|P| \geq |S_P|$.
+%\item $|P| \geq |S_P|$.
 The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected.
@@
 Conversely, assuming $|S_{P_{min}}| = n - 1$, if we scheduled $P_{min}$ first, .
-\end{itemize}
+%\end{itemize}
 

### 2015-12-17 TOMS_Paper/assembly.tex
small fixes

@@
 \item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
-For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
+
+\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
 \item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
@@
 \item Calculate the number of operations in the innermost loop required for the execution of $P \in \mathbb{P}$.
-There are two possibilities, depending on which strategy -- (i) or (ii), as described in Section~\ref{sec:se-rln} -- is employed.
+There are two possibilities, depending on which strategy -- (i) or (ii), as described in Section~\ref{sec:se-rln} -- is used.
 In particular, we have $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$, with $c_P^i = |P|$ and $c_P^{ii} = |S_P|$.
@@
 We introduce $g_P$, which measures the gain in operation count in the innermost loop if (ii) were applied.
-That is, $g_P = \sum_{Q \in \mathbb{Q}} c_{Q_{old}} - c_{Q_{new}}$, with $\mathbb{Q} = \mathbb{P} \setminus \lbrace P \rbrace$.
-Note that $c_{Q_{new}}$ can be evaluated without applying the actual transformation.
-Eventually, apply (ii) if $g_P > 0$.
-\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent or previously hoisted sub-expression, apply (i) or (ii).
-This pass speculatively assumes that structured tensor operations are in place.
-If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may have been exposed.
+We have $g_P = \sum_{Q \in \mathbb{Q}} c_{Q_{old}} - c_{Q_{new}}$, with $\mathbb{Q} = \mathbb{P} \setminus \lbrace P \rbrace$.
+Apply (ii) if $g_P > 0$ and the expansion cost does not exceed the gain in operation count.
+
+\textit{Note: both $c_{Q_{new}}$ and the expansion cost are easily determined through tree visits.}
+
+\textit{Note: $\mathbb{P}$ is in practice very small, so this pass can be accomplished rapidly.}
+\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent or previously hoisted sub-expression, apply the most profitable between (i) and (ii).
+
+\textit{Note: this pass speculatively assumes that structured tensor operations are in place.
+If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may have been exposed.}
 \end{enumerate}
 \end{Algo}
+
+%TODO : dire che il costo dell'espansione è semplicemente dato da ciascun s in | S_P | times la 2 evalato al depth di quell s.
 

### 2015-12-21 TOMS_Paper/assembly.tex
Rework, again, the algorithm...including the sharing graph

@@
 The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$.
-In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and the factorization gain.
+In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and code motion.
 One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}.
@@
 \item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
+\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured tensor operations.
 \end{itemize}
@@
 \normalfont
-The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
+The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-4); scheduling of all other sub-expressions (step 5).
 \begin{enumerate}
@@
 \textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
-\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
+\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
 $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
-\item Calculate the number of operations in the innermost loop required for the execution of $P \in \mathbb{P}$.
-There are two possibilities, depending on which strategy -- (i) or (ii), as described in Section~\ref{sec:se-rln} -- is used.
-In particular, we have $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$, with $c_P^i = |P|$ and $c_P^{ii} = |S_P|$.
-\item For all multilinear operands $o \in P$ such that $c_P = c_P^i$, apply strategy (i). Update $\mathbb{P}$ by replacing hoisted multilinear operands with the corresponding temporary symbols.
-\item For all $P \in \mathbb{P}$, determine the profitability of strategy (ii).
-We introduce $g_P$, which measures the gain in operation count in the innermost loop if (ii) were applied.
-We have $g_P = \sum_{Q \in \mathbb{Q}} c_{Q_{old}} - c_{Q_{new}}$, with $\mathbb{Q} = \mathbb{P} \setminus \lbrace P \rbrace$.
-Apply (ii) if $g_P > 0$ and the expansion cost does not exceed the gain in operation count.
-
-\textit{Note: both $c_{Q_{new}}$ and the expansion cost are easily determined through tree visits.}
-
-\textit{Note: $\mathbb{P}$ is in practice very small, so this pass can be accomplished rapidly.}
-\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent or previously hoisted sub-expression, apply the most profitable between (i) and (ii).
+\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply strategy (i) to the multilinear operands.
+
+\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, strategies (i) or (ii) were applied.
+We will clarify that this directly descends from loop linearity.}
+
+\item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step 3.
+An edge $(S^i$, $S^j)$ indicates that if the sub-expressions including $S^i$ and $S_j$ were expanded, a product $S^i S^j$ would be exposed.
+Let $\mathbb{T}$ be the subset of vertices with a single incident edge, or ``terminals''.
+We apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
+\begin{enumerate}[label=(\Alph*)]
+\item $\lbrace T^0, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^0, ..., T^n \rbrace$.\\ \textit{Note: strategy (ii) involves exposing $S (...T^0 + ...
++ ...T^n)$ and, subsequently, hoisting $(...T^0 + ...
++ ...T^n)$}.
+\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) as in rule (A), but including one non-terminal symbol $S^i$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^i \rbrace$.
+\item $\mathbb{T} = \emptyset$: take $S$, apply strategy (ii), eventually hoisting any two non-terminal symbols.\\ \textit{Note: this represents the ``fallback'' approach in presence of cyclic dependency}
+\end{enumerate}
+\item Perform a depth-first visit of the expression tree and, for each yet unhandled or previously hoisted sub-expression, apply the most profitable between strategies (i) and (ii).
 
 \textit{Note: this pass speculatively assumes that structured tensor operations are in place.
-If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may have been exposed.}
+If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be exposed.}
 \end{enumerate}
@@
 %TODO : dire che il costo dell'espansione è semplicemente dato da ciascun s in | S_P | times la 2 evalato al depth di quell s.
+%\textit{Note: $\mathbb{S}$ is in practice very small, so this pass can be accomplished rapidly.}
 
@@
 \label{sec:se-examples}
-Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D.
-We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$, with $c_{P^0} = c_{P^1} = {min(}2, 2{)} = 2$.
-A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
+Consider again Figure~\ref{code:multi_loopnest_a}.
+We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$.
+For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying strategy (i) in step 3 has no effect.
+The sharing graph is $\mathbb{G} = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $\mathbb{T} = \lbrace c_i, d_i \rbrace$.
+Rewrite rule (4A) is hit, which leads to the code in Figure~\ref{code:multi_loopnest_c}.
+
+Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to a realistic example, the bilinear form arising from a Poisson equation in 2D.
+We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
+In addition, $|P^i| = |S_{P^i}| = 2$, so strategy (i) is applied to both partitions (step 3).
+We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{E} = \emptyset$.
+
 \begin{figure}[h]\begin{CenteredBox}
@@
 {\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
-\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}
+\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}
+
+A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}
 

### 2015-12-21 TOMS_Paper/assembly.tex
Rework, again, the algorithm...including the sharing graph

@@
 We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
-Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{algo:gamma}.
-
-%numbers=left
-\begin{figure}
-\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
-// Initialization
-M = $\lbrace$m | m is a monomial in the input expression$\rbrace$
-S = $\lbrace$m | $\theta(m)$ < 0$\rbrace$
-P = M $\setminus$ S
-
-// Find global quasi-optimum honoring memory constraints
-B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
-for each b $\in$ B:
-  for each (b$_S$, b$_P$) $\in$ b:
-    if memory_required(b$_P$) $\geq$ $T_H$:
-      continue
-    op_count[b] = $\theta_{se}$(S $\cup$ b$_S$) + $\theta_{pre}$(b$_P$)
-(b$_S$, b$_P$) = min(op_count)
-
-// Apply the transformations
-pre_evaluate(P $\cup$ b$_P$)
-eliminate_sharing(S $\cup$ b$_S$ $\cup$ P $\cup$ b$_P$)
-\end{lstlisting}
-\caption{High level view of the transformation algorithm}
+Having defined $\theta$, we can focus on the transformation algorithm, provided in Algorithm~\ref{algo:gamma}.
+
+\begin{Algo}[Transformation algorithm]
 \label{algo:gamma}
-\end{figure}
-
-The algorithm starts with splitting the monomials into two disjoint sets: the monomials that surely do not take advantage of pre-evaluation ($S$) and those that instead potentially benefit from it ($P$).
-Some subsets of monomials in $P$: 1) might share terms with $S$, in which case sharing elimination could globally outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}.
-The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of monomials for which pre-evaluation is globally optimal and Constraint~\ref{const:TH} is honored.
-Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible.
-This aspect is elaborated in Section~\ref{sec:op_count}.
-In addition, observe that because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary.
-Therefore, sharing elimination is transparently applied to all monomials (i.e., to all sub-expressions, even those produced by pre-evaluation).
+\normalfont
+The algorithm has three main phases: initialization (step 1); determination of the monomials preserving the memory constraints that should be pre-evaluated (steps 2-4); application of sharing elimination and pre-evaluation (step 5).
+\begin{enumerate}
+\item Perform a depth-first visit of the expression tree and determine the set of monomials $M$.
+Let $S$ be the subset of monomials $m$ such that $\theta(m) < 0$.
+The set of monomials that will \textit{potentially} be pre-evaluated is $P = M \setminus S$.
+\\ \textit{Note: there are two subtle yet fundamental reasons for not pre-evaluating $m_1 \in P$: 1) the presence of spatial sharing between $m_1$ and $m_2 \in S$, which impacts the search for the global optimum; 2) breaking memory constraints.}
+\item Build the set $B$ of all possible bipartitions of $P$.
+\item For each $b = (b_S, b_P) \in B$, if the memory required to store the pre-evaluated tables from the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}), discard $b$; otherwise, add an entry to the dictionary $d$ of potential operation counts, such that $d[b] = \theta_{se}(S \cup b_S) + \theta_{pre}(b_P)$.
+\\ \textit{Note: $\mathbb{B}$ is in practice very small, since even complex forms usually have only a few monomials.
+This pass can then be accomplished rapidly as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible.
+We elaborate on this aspect in Section~\ref{sec:op_count}.}
+\item Take $b$ such that $min(d[b])$.
+\item Apply pre-evaluation to all monomials in $P \cup b_P$.
+Apply sharing elimination to the whole expression.
+\\ \textit{Note: because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary.
+Sharing elimination is therefore transparently applied to the whole expression, including what results from pre-evaluation.}
+\end{enumerate}
+\end{Algo}
+
 The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming as input the loop nest in Figure~\ref{code:loopnest}.
 
-\begin{figure}[t]\begin{CenteredBox}
+\begin{figure}[h]\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
@@
 
-In the following section, we tie up the remaining loose end: the construction of the cost function $\theta$.
-
 \subsection{The cost function $\theta$}
 \label{sec:op_count}
+We tie up the remaining loose end: the construction of the cost function $\theta$.
+
 We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and pre-evaluation.
@@
 
-\begin{figure}[t]\begin{CenteredBox}
+\begin{figure}[h]\begin{CenteredBox}
 \lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
@@
 We prevent this situation by construction.
-$\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}).
+$\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim.
 Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.

### 2015-12-21 TOMS_Paper/assembly.tex
Minor stuff, plus additions to the proof

@@
 + ...T^n)$}.
-\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) as in rule (A), but including one non-terminal symbol $S^i$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^i \rbrace$.
-\item $\mathbb{T} = \emptyset$: take $S$, apply strategy (ii), eventually hoisting any two non-terminal symbols.\\ \textit{Note: this represents the ``fallback'' approach in presence of cyclic dependency}
+\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) as in rule (A), but including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
+\item $\mathbb{T} = \emptyset$: take $S$, apply strategy (ii), hoist any two non-terminal symbols.\\ \textit{Note: this rule represents the ``fallback'' in presence of cyclic dependency}
 \end{enumerate}
@@
 
-The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming as input the loop nest in Figure~\ref{code:loopnest}.
+The output of the transformation algorithm is provided in Figure~\ref{code:loopnest-opt}, assuming as input the loop nest in Figure~\ref{code:loopnest}.
 
@@
 Assume that pre-evaluating different monomials does not result in identical tables.
-Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}.
+Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and preserves Constraints~\ref{const:Le} and~\ref{const:TH}.
 \end{Prop}
 \begin{proof}
-We first observe that the cost function $\theta$ predicts the exact gain/loss in monomial pre-evaluation, so $X$ is always correct by construction.
+We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ is constructed correctly.
 
-Let $c_\Lambda$ denote the operation count in $\Lambda$.
-Let $\Lambda'_I \subset \Lambda'$ be the subset of all innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
-We need to show that there is no other synthesis $\Lambda''_{I}$ such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
+Let $c_\Lambda$ denote the operation count in $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
+We need to show that there is no other synthesis $\Lambda''_{I}$ preserving the memory constraints such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
 This holds if and only if
@@
 \end{enumerate}
-\item \textit{Sharing elimination produces (at least) quasi-optimal loop nests.}\\
+\item \textit{Sharing elimination produces quasi-optimal (or optimal) loop nests.}\\
 \end{enumerate}
@@
 \begin{enumerate}
-\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$.
-Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$.
-If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary.
-$\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis).
-By assumption, this case is neglected.
-The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
+\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$.
+Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$.
+If $T_{m_{1}} \cap T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary.
+$\Gamma$, therefore, may skip a potential global optimum, since $X$ and $Y$ are constructed reasoning -- through $\theta$ -- only on monomials in ``isolation''.
+This case is neglected by assumption.
+The assumption is reasonable because such a scenario is purely pathological and, with extreme probability, of insignificant importance.
 \item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols.
-If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one.
-We prevent this situation by construction.
-$\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim.
-Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
+If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum.
+This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum that also preserves the memory constraints.
+Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
+Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
 \end{enumerate}
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$.
-
-%What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality.
-We do it by contradiction.
-Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$.
-In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved.
-We distinguish two cases.
-%\begin{itemize}
-%\item $|P| < |S_P|$.
-This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place.
-We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
-%\item $|P| \geq |S_P|$.
-The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected.
-Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion.
-It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$.
-Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions).
-Conversely, assuming $|S_{P_{min}}| = n - 1$, if we scheduled $P_{min}$ first, .
-%\end{itemize}
-
-%In total, therefore, 
-%If none of the $n$ hoisted sub-expressions include operands in $P_{min}$, we have again $c_{\Lambda''_I} = c_{\Lambda'_I}$. If that is not the case,
-
-%We first observe that a partition (i.e., its multilinear symbols) depends on one of the two multilinear loops; we then indicate with $P^l$ the fact that $P$ depends on $L_l$, $l \in \lbrace j, k\rbrace$.
-Assuming $P_{min}^l$, we have two possibilities:
-%\begin{itemize}
-%\item likewise, $P^{l}$; that is, both $P_{min}$ and $P$ depend on the same linear loop.
-This is the simplest case, since swapping the order in which the two partitions are handled would not have any side effects -- there is no clash between all possible different re-factorizations.
-The assumption is then contradicted because the operation count in the innermost loop is the same irrespective of starting with $P_{min}$ or $P$.
-%\item $P^{l'}$ and $l' \neq l$.
-In this case, it is possible that operands (symbols) in $P^{l'}$ are factorized and code motion takes place such that $P_{min}$ ``loses'' some of its operands (symbols).
-Linearity and the assumption $c_{P} > c_{P_{min}}$ tell us that once $P$ is scheduled, we will have $c_{\Lambda''_I} \geq c_{\Lambda'_I} + 1$; that is, there is at least one extra operation (an addition) due to
-%\end{itemize}
+Loop linearity is a sufficient (although not necessary) condition for removing spatial sharing through factorization.
+This induces a lower bound on the innermost loop operation count induced by $P \in \mathbb{P}$, which cannot be less than $c_P = 2{min(}|P|,\ |S_P|{)} - 1$.
+This observation ensures that step (3), which hoists loop-invariant multilinear operands, never prunes optimal solutions from the search space.
+Step (4) determines an optimal ordering for the factorization of multilinear symbols and temporaries produced at step (3).
+In particular, we need to demonstrate that the rewrite rules based on properties of the sharing graph bring to a quasi-optimal synthesis:
+\begin{enumerate}[label=(\Alph*)]
+\item Rule (A) has no side-effects since $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$.
+\item By contradiction, we can prove that applying rule (B) is always a step forward towards the optimal solution, assuming (as stated in Algorithm~\ref{algo:sharing-elimination}) no hit is possible for rule (A).
+Assume then that (B) prunes the path towards quasi-optimality by choosing a sub-optimal factorization.
+This means that there exists ...
+\item Rule (C) ...
+\end{enumerate}
+Finally, note that step (5) is irrelevant for quasi-optimality.
 
@@
 
-%Partitions are ordered to promote a specific factorization sequence; this is essential to minimize the operation count in outer loops, although it is non influential for quasi-optimality.
-Criterion (2) also does not affect innermost loops.
-Criterion (1) is instead a discriminating factor.
-Consider a generic partition $P \in \mathbb{P}$.
-multilinearity...
-
 \end{proof}
 
-%TODO : even simpler if we considered linera forms. If we had trilinear forms, we would be fucked
-
 %TODO : In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
-
-% ordering of mathbb{P} is fundamental, because you start with something, then you collect stuff, and what you've collected is gonna be hoisted, which means having made a choice in terms of minimizing the innermost loop cost --- multilinearity and normal form here are super super imporant !
-% |P| <, ==, > then |S_P| is important because think of like: z0*FEik .... F0*FEik .... F1*FEik ...
 
@@
 Referred to as \texttt{auto}
-\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
+\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation and execution times). Referred to as \texttt{ufls} 
 \item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 

### 2015-12-22 TOMS_Paper/assembly.tex
Improve cost function description

@@
 \item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
-\item \textit{Search for a global optimum.} Consider a form comprising two monomials $m1$ and $m2$.
-Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g.
+\item \textit{Search for a global optimum.} Consider a form comprising two monomials $m_1$ and $m_2$.
+Assume that pre-evaluation is profitable for $m_1$ but not for $m_2$, and that $m_1$ and $m_2$ share at least one term (e.g.
 some basis functions).
-If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost.
+If pre-evaluation were applied to $m_1$, sharing between $m_1$ and $m_2$ would be lost.
 We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
@@
 We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
-Having defined $\theta$, we can focus on the transformation algorithm, provided in Algorithm~\ref{algo:gamma}.
+Having defined $\theta$, we can now focus on the transformation algorithm (Algorithm~\ref{algo:gamma}).
 
@@
 \normalfont
-The algorithm has three main phases: initialization (step 1); determination of the monomials preserving the memory constraints that should be pre-evaluated (steps 2-4); application of sharing elimination and pre-evaluation (step 5).
+The algorithm has three main phases: initialization (step 1); determination of the monomials preserving the memory constraints that should be pre-evaluated (steps 2-4); application of pre-evaluation and sharing elimination (step 5).
 \begin{enumerate}
@@
 The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations.
-If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead becomes unacceptable when applying pre-evaluation to all possible bipartitions (Algorithm~\ref{algo:pre-evaluation}, due to the symbolic evaluation of $L_i$).
+If, on one hand, this is tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}.
 We then seek an analytic way of determining $\theta_{pre}$.
@@
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities.
-By performing product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.
+By product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.
 
@@
 
-Despite determining $\iota$ is straightforward in simple scenarios (e.g., when the form has a single coefficient, as we just saw), in general we must account for repeated sub-expressions that, once pre-evaluated, would result in identical tables.
-This is the case, for example, of a monomial having more than one coefficient in the same function space.
-One can verify this by taking the example in Figure~\ref{code:increase_factor}, adding a second coefficient $g$ over the same basis functions $b$, and expanding the products.
+In general, however, determining $\iota$ is not so straightforward since redundant tabulations may result from common sub-expressions.
+Take the previous example.
+One can add one coefficient in the same function space as $f$, repeat the expansion, and observe that multiple sub-expressions (e.g., $b_{10}*b_{01}*...$ and $b_{01}*b_{10}*...$) will reduce to identical tables.
 To evaluate $\iota$, we then use combinatorics.
-We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, $b_{i0}$, $b_{i1}$, and $b_{i2}$, so $n=3$).
+We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product; (ii) $n$ is the number of unique basis functions involved in the expansion.
+In the original example, we have $n=3$ (for $b_{i0}$, $b_{i1}$, and $b_{i2}$) and $k=1$, which confirms what we intuitively found before, namely $\iota=3$.
+In the modified example there are two coefficients, so $k=2$, which means $\iota=6$.
 
@@
 The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
-If we instead take the Poisson example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after suitable factorization of basis functions.
+If we instead take the Poisson example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after a suitable factorization of basis functions.
 In general, therefore, $\rho$ depends on both form and discretization employed.

### 2015-12-22 TOMS_Paper/assembly.tex
Reworking the proof...

@@
 The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
-If we instead take the Poisson example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after a suitable factorization of basis functions.
+If we instead take the example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after a suitable factorization of basis functions.
 In general, therefore, $\rho$ depends on both form and discretization employed.
-To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and count the terms amenable to pre-evaluation.
+To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and simply count the terms amenable to pre-evaluation.
 
@@
 \label{sec:proof}
-The following proposition states that our approach guarantees quasi-optimality. The proof re-uses some of the concepts and explanations provided throughout the paper.
+The following proposition states that our approach guarantees quasi-optimality.
+The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.
 
@@
 Let $X$ be the set of pre-evaluated monomials, and let $Y = M \setminus X$.
-Assume that pre-evaluating different monomials does not result in identical tables.
+Assume that the pre-evaluation of different monomials does not result in identical tables.
 Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and preserves Constraints~\ref{const:Le} and~\ref{const:TH}.
@@
 \begin{proof}
-We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ is constructed correctly.
+We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ and $Y$ can actually be constructed correctly.
 
-Let $c_\Lambda$ denote the operation count in $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
-We need to show that there is no other synthesis $\Lambda''_{I}$ preserving the memory constraints such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
+Let $c_\Lambda$ denote the operation count for $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
+We need to show that there is no other synthesis $\Lambda''_{I}$ preserving Constraints~\ref{const:Le} and~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
 This holds if and only if
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Loop linearity is a sufficient (although not necessary) condition for removing spatial sharing through factorization.
-This induces a lower bound on the innermost loop operation count induced by $P \in \mathbb{P}$, which cannot be less than $c_P = 2{min(}|P|,\ |S_P|{)} - 1$.
-This observation ensures that step (3), which hoists loop-invariant multilinear operands, never prunes optimal solutions from the search space.
-Step (4) determines an optimal ordering for the factorization of multilinear symbols and temporaries produced at step (3).
-In particular, we need to demonstrate that the rewrite rules based on properties of the sharing graph bring to a quasi-optimal synthesis:
+Based on loop linearity and normal form of input expressions, we observe that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols, respectively leading to $|P|$ and $|S_P|$ multiplications.
+An upper bound for $P$ is, therefore, $ub_P = 2{min(}|P|,\ |S_P|{)} - 1$.
+Thus, when generalized code motion in step (3) is applied, we do know that we are not pruning any potentially optimal solution from the search space.
+In fact, we are just moving towards the lower bound, given by $lb_P = min(0, ub_P)$, with $0$ reached only if the symbols in $P$ will be grouped and hoisted at step (4).
+
+Step (4) factorizes multilinear symbols and temporaries produced at step (3) in optimal order; that is, an order that leads to a quasi-optimal synthesis.
+To demonstrate this, we consider the sharing graph rewrite rules.
 \begin{enumerate}[label=(\Alph*)]
@@
 \end{enumerate}
-Finally, note that step (5) is irrelevant for quasi-optimality.
+Finally, we note that step (5) is irrelevant for quasi-optimality.
 

### 2015-12-22 TOMS_Paper/assembly.tex
Improving the proof...

@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Based on loop linearity and normal form of input expressions, we observe that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols, respectively leading to $|P|$ and $|S_P|$ multiplications.
-An upper bound for $P$ is, therefore, $ub_P = 2{min(}|P|,\ |S_P|{)} - 1$.
-Thus, when generalized code motion in step (3) is applied, we do know that we are not pruning any potentially optimal solution from the search space.
-In fact, we are just moving towards the lower bound, given by $lb_P = min(0, ub_P)$, with $0$ reached only if the symbols in $P$ will be grouped and hoisted at step (4).
+Because of loop linearity and normal form of expressions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols.
+These two strategies lead, respectively, to performing $|P|$ and $|S_P|$ products at every iteration.
+Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space; in fact, it just moves towards quasi-optimality.
 
-Step (4) factorizes multilinear symbols and temporaries produced at step (3) in optimal order; that is, an order that leads to a quasi-optimal synthesis.
-To demonstrate this, we consider the sharing graph rewrite rules.
+Step (4) factorizes symbols following an ordering that brings to quasi-optimality.
+The order is dictated by the sharing graph and its rewrite rules.
+We first recall that rewrite rules are tried following the order in which they were presented in Algorithm~\ref{algo:sharing-elimination}; for example, we do not attempt rule (B) until we have handled all possible hits for rule (A).
 \begin{enumerate}[label=(\Alph*)]
-\item Rule (A) has no side-effects since $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$.
-\item By contradiction, we can prove that applying rule (B) is always a step forward towards the optimal solution, assuming (as stated in Algorithm~\ref{algo:sharing-elimination}) no hit is possible for rule (A).
-Assume then that (B) prunes the path towards quasi-optimality by choosing a sub-optimal factorization.
-This means that there exists ...
-\item Rule (C) ...
+\item Rule (A) is side effects free since $\lbrace T^0, ..., T^n \rbrace$ can be grouped only through the factorization of $S$.
+\item We prove that rule (B) makes the algorithm progress towards the optimal solution.
+We first observe that the choice of $S^{nt}$ is arbitrary.
+Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
+We then consider the alternative factorization, namely $S^{nt}(...S + ...
++ ...V)$.
+If $V$ were a terminal, we would simply be back to where we started, because rule (B) would be applicable by setting $V = T$ and $S = S^{nt}$.
+If, on the other hand, $V$ were a non terminal, we would have $S$ appearing in two different sub-expressions, which implies the presence of spatial sharing.
+\item In order to have $\mathbb{T} = \emptyset$, which would trigger rule (C), it must be the case that all non terminals have the same number of incoming edges.  
 \end{enumerate}

### 2015-12-28 TOMS_Paper/assembly.tex
Improving the proof...
Trying to finalise the proof...

@@
 \item Applying generalized code motion -- not needed in the example as we have one loop.
-\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
+\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
 \end{enumerate}
@@
 One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}.
-We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.
+We then regard as \textit{structured tensor operation} any tensor operation that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.
 
@@
 \normalfont
-The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-4); scheduling of all other sub-expressions (step 5).
+The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
 \begin{enumerate}
-\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions.
+\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. \\ 
+\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
 
-\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
 \item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
 $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
-\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply strategy (i) to the multilinear operands.
 
+\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply strategy (i) to the multilinear operands. \\
 \textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, strategies (i) or (ii) were applied.
@@
 \item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step 3.
-An edge $(S^i$, $S^j)$ indicates that if the sub-expressions including $S^i$ and $S_j$ were expanded, a product $S^i S^j$ would be exposed.
+An edge $(S^i$, $S^j)$ indicates that a product $S^i S^j$ would appear if the sub-expressions including $S^i$ and $S_j$ were expanded.
 Let $\mathbb{T}$ be the subset of vertices with a single incident edge, or ``terminals''.
-We apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
+
+\item If $\mathbb{S} = \emptyset$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^0, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^0, ..., T^n \rbrace$.\\ \textit{Note: strategy (ii) involves exposing $S (...T^0 + ...
-+ ...T^n)$ and, subsequently, hoisting $(...T^0 + ...
+\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: strategy (ii) recasts the expression as $S (...T^0 + ...
 + ...T^n)$}.
-\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) as in rule (A), but including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
-\item $\mathbb{T} = \emptyset$: take $S$, apply strategy (ii), hoist any two non-terminal symbols.\\ \textit{Note: this rule represents the ``fallback'' in presence of cyclic dependency}
+\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
+\item $\mathbb{T} = \emptyset$: take $S$ adjacent to $\lbrace S^1, ..., S^d \rbrace$ such that $d$ is maximum (i.e.
+$S$ is the vertex with highest degree in $\mathbb{G}$) and apply strategy (ii);  $\mathbb{S} = \mathbb{S} \setminus \lbrace S^1, ..., S^d \rbrace$.
 \end{enumerate}
-\item Perform a depth-first visit of the expression tree and, for each yet unhandled or previously hoisted sub-expression, apply the most profitable between strategies (i) and (ii).
+Go back to step (2).
 
+\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
 \textit{Note: this pass speculatively assumes that structured tensor operations are in place.
@@
 %TODO : dire che il costo dell'espansione è semplicemente dato da ciascun s in | S_P | times la 2 evalato al depth di quell s.
+%\item $\mathbb{T} = \emptyset$: take $S$ at will such that the number of adjacent non terminals is maximum, and apply strategy (ii). 
 %\textit{Note: $\mathbb{S}$ is in practice very small, so this pass can be accomplished rapidly.}
@@
 
-Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to a realistic example, the bilinear form arising from a Poisson equation in 2D.
+In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, the bilinear form arising from a Poisson equation in 2D.
 We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
@@
 Assume that the pre-evaluation of different monomials does not result in identical tables.
-Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and preserves Constraints~\ref{const:Le} and~\ref{const:TH}.
+Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and satisfies Constraints~\ref{const:Le} and~\ref{const:TH}.
 \end{Prop}
@@
 Let $c_\Lambda$ denote the operation count for $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
-We need to show that there is no other synthesis $\Lambda''_{I}$ preserving Constraints~\ref{const:Le} and~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
+We need to show that there is no other synthesis $\Lambda''_{I}$ satisfying Constraints~\ref{const:Le} and~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
 This holds if and only if
@@
 If $T_{m_{1}} \cap T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary.
-$\Gamma$, therefore, may skip a potential global optimum, since $X$ and $Y$ are constructed reasoning -- through $\theta$ -- only on monomials in ``isolation''.
-This case is neglected by assumption.
-The assumption is reasonable because such a scenario is purely pathological and, with extreme probability, of insignificant importance.
+$\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''.
+We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
 \item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols.
@@
 Because of loop linearity and normal form of expressions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols.
-These two strategies lead, respectively, to performing $|P|$ and $|S_P|$ products at every iteration.
-Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space; in fact, it just moves towards quasi-optimality.
+These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration.
+Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space; in fact, it just represents a move towards quasi-optimality.
 
-Step (4) factorizes symbols following an ordering that brings to quasi-optimality.
-The order is dictated by the sharing graph and its rewrite rules.
-We first recall that rewrite rules are tried following the order in which they were presented in Algorithm~\ref{algo:sharing-elimination}; for example, we do not attempt rule (B) until we have handled all possible hits for rule (A).
-\begin{enumerate}[label=(\Alph*)]
-\item Rule (A) is side effects free since $\lbrace T^0, ..., T^n \rbrace$ can be grouped only through the factorization of $S$.
-\item We prove that rule (B) makes the algorithm progress towards the optimal solution.
-We first observe that the choice of $S^{nt}$ is arbitrary.
-Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
-We then consider the alternative factorization, namely $S^{nt}(...S + ...
+Steps (4)-(5) factorize symbols following the ordering dictated by the sharing graph and its rewrite rules.
+Recall that rules are applied in sequence: first all rule (A) hits are handled, then all (B)s and finally all (C)s, with applications of rule (C) that may expose further possibilities for (A) and (B).
+We prove that the application of these rules, in this sequence, represents a step towards quasi-optimality.
+\begin{description}
+\item[Rule (A)] $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule is side effects free (i.e., no need to choose between different factorizations).
+\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
+Let us consider the alternative factorization $S^{nt}(...S + ...
 + ...V)$.
 If $V$ were a terminal, we would simply be back to where we started, because rule (B) would be applicable by setting $V = T$ and $S = S^{nt}$.
-If, on the other hand, $V$ were a non terminal, we would have $S$ appearing in two different sub-expressions, which implies the presence of spatial sharing.
-\item In order to have $\mathbb{T} = \emptyset$, which would trigger rule (C), it must be the case that all non terminals have the same number of incoming edges.  
-\end{enumerate}
-Finally, we note that step (5) is irrelevant for quasi-optimality.
+Otherwise ($V$ non terminal), we would have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
+\item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
+For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d) \in \mathbb{B}$ we consider two possibilities, namely $\mathbb{B^d}$ is 1) not complete or 2) complete.
+Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound for the operation count (as number of products performed at every loop iteration) is given by ${min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
+This is reachable by taking the vertex with highest degree, as the rule states (note that other choices may break factorization opportunities).
+In case 2), the lower bound is 1 because factorization can be applied iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).
+
+% At this point, recall . 
+%In order to have $\mathbb{T} = \emptyset$, which would trigger rule (C), it must be the case that ....
+\end{description}
+Finally, we note that step (6) is irrelevant for quasi-optimality.
 

### 2015-12-28 TOMS_Paper/assembly.tex
Minor fixes. Still need to ultimate the proof

@@
 These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration.
-Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space; in fact, it just represents a move towards quasi-optimality.
+Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space.
+%; in fact, it simply shifts towards the objective.
 
@@
 + ...V)$.
-If $V$ were a terminal, we would simply be back to where we started, because rule (B) would be applicable by setting $V = T$ and $S = S^{nt}$.
-Otherwise ($V$ non terminal), we would have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
+If $V$ is a terminal, we are back to where we started, because rule (B) will be applicable by setting $V = T$ and $S = S^{nt}$.
+Otherwise ($V$ non terminal), we have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
 \item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
-For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d) \in \mathbb{B}$ we consider two possibilities, namely $\mathbb{B^d}$ is 1) not complete or 2) complete.
-Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound for the operation count (as number of products performed at every loop iteration) is given by ${min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
-This is reachable by taking the vertex with highest degree, as the rule states (note that other choices may break factorization opportunities).
-In case 2), the lower bound is 1 because factorization can be applied iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).
-
-% At this point, recall . 
-%In order to have $\mathbb{T} = \emptyset$, which would trigger rule (C), it must be the case that ....
+For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete.
+Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
+This is reachable by factorizing the vertex with highest degree (as the rule states), as any other choice may break factorization opportunities.
+In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).
 \end{description}
-Finally, we note that step (6) is irrelevant for quasi-optimality.
-
+The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \emptyset$, before applying the rewrite rules. 
 \end{enumerate}
-
 \end{proof}
@@
 We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied).
-We analyze the following transformation systems
-\begin{itemize}
-\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
-\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
-\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}).
-Referred to as \texttt{auto}
-\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation and execution times). Referred to as \texttt{ufls} 
-\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
-\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
-\end{itemize}
+We analyze the following transformation systems:
+\begin{description}
+\item[quad] Optimized quadrature mode. Work presented in~\citeN{quadrature1}, implemented in  in the FEniCS Form Compiler. 
+\item[tens] Tensor contraction mode. Work presented in~\citeN{FFC-TC}, implemented in the FEniCS Form Compiler.
+\item[auto] Automatic choice between \texttt{tens} and \texttt{quad} driven by heuristic (detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}).
+Implemented in the FEniCS Form Compiler.
+\item[ufls] UFLACS, a novel back-end for the FEniCS Form Compiler whose primary goals are improved code generation and execution times.
+\item[cfO1] Generalized loop-invariant code motion. Work presented in~\citeN{Luporini}, implemented in COFFEE.
+\item[cfO2] Optimal loop nest synthesis with handling of block-sparse tables. Work presented in this article, implemented in COFFEE.
+\end{description}
 

### 2015-12-28 TOMS_Paper/assembly.tex
Minor fixes.

@@
 \item Only one terminal $T$ adjacent to $S$: apply strategy (ii) including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
-\item $\mathbb{T} = \emptyset$: take $S$ adjacent to $\lbrace S^1, ..., S^d \rbrace$ such that $d$ is maximum (i.e.
-$S$ is the vertex with highest degree in $\mathbb{G}$) and apply strategy (ii);  $\mathbb{S} = \mathbb{S} \setminus \lbrace S^1, ..., S^d \rbrace$.
+\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii).
 \end{enumerate}

### 2016-01-14 TOMS_Paper/assembly.tex
Several fixes, but very important

@@
 Optimality and quasi-optimality are defined in terms of floating point operations given a memory bound.
-We provide an approach to explore the space of legal transformations and discuss under what conditions optimality or quasi-optimality hold.
-A theoretical analysis and extensive experimentation, which shows systematic performance improvements over several state-of-the-art code generation systems, validate the approach.
+We provide an approach for exploring the space of legal transformations, which systematically reaches quasi-optimality.
+The conditions under which optimality is achieved are also discussed.
+A theoretical analysis and extensive experimentation, which shows significant performance improvements over several state-of-the-art code generation systems, validate the approach.
 \end{abstract}
@@
 However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (``quasi-optimality'') in all other cases.
+In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (we will define ``quasi-optimality'', which relates operation count with inner loops) in all other cases.
 Summarizing, our contributions are as follows:
@@
 The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
-\item We explain under what conditions our algorithm leads to optimality.
-In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, in general, always achieved (i.e., the operation count is at least optimal in innermost loops).
+\item We demonstrate that our algorithm systematically leads to quasi-optimality. We also explain under what conditions of the input problem optimality is achieved.
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 In particular, $m$ is the number of ``terms'' (a monomial can be implemented by one or more terms), $\alpha_{eij}$ ($\beta_{eik}$) represents the product of a coefficient function (e.g., the inverse Jacobian matrix for the change of coordinates) with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex.
+We do not pose any restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex.
 We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact (e.g., products have not been expanded yet, distinct monomials can still be identified, etc.).
@@
 \label{def:fem-loopnest}
-A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which each statement has expressions in normal form.
+A finite element integration loop nest is a loop nest in which appear, in order, an imperfect free order loop, an imperfect (perfect only in special cases), linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which all expressions are in normal form.
 \end{Def}
@@
 \item there are at least two symbolically identical sub-expressions (spatial sharing)
-\item there is at least one non-trivial sub-expression (an addition or a product) that is redundantly executed as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
+\item there is at least one non-trivial sub-expression (an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
 \end{enumerate}
@@
 Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}).
-Finally, generalized code motion \cite{Luporini} leads to optimality (Figure~\ref{code:multi_loopnest_c}).
+Finally, generalized code motion \cite{Luporini}, which seeks redundant sub-expressions analyzing all levels of the loop nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only considers innermost loops.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).
 
@@
 In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion.
-If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
+If the objective were reaching optimality and the expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
 Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
 For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold.
-Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
+Setting quasi-optimality, instead of optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
 
-Section~\ref{sec:se-rln} discusses structural and algebraic properties characterizing our expressions.
+%Section~\ref{sec:se-rln} discusses structural and algebraic properties characterizing our expressions.
 Section~\ref{sec:se-algo} presents the sharing elimination algorithm.
@@
 We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2.
-To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$.
-Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$.
-For example, suppose that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$.
-Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$.
-To schedule this expression, we basically have two options:
+To illustrate the concept, consider the tensor operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
+One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$.
+To schedule sub-expressions of this kind, we have two options:
 \begin{enumerate}[label=(\roman*)]
-\item Applying generalized code motion -- not needed in the example as we have one loop.
-\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
+\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
+\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
 \end{enumerate}
-The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$.
-In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and code motion.
-One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}.
-We then regard as \textit{structured tensor operation} any tensor operation that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.
-
-Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations.
-Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too.
-For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).
+The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$.
+In general, however, the optimal choice depends on multiple factors: the loop size, the increase in operation count due to expansion, and the gain in code motion.
+We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii).
+Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured; this observation is fundamental for understanding the sharing elimination algorithm, described in the next section\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form.
+For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}.
 
@@
 \label{sec:se-algo}
-Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest.
+Algorithm~\ref{algo:sharing-elimination} describes how we perform sharing elimination assuming as input a tree representation of the loop nest.
 The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest.
-It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration.
 Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.

### 2016-01-15 TOMS_Paper/assembly.tex
Several fixes here and there

@@
 However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (we will define ``quasi-optimality'', which relates operation count with inner loops) in all other cases.
+In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``quasi-optimality'', which relates operation count with inner loops).
 Summarizing, our contributions are as follows:
@@
 \label{sec:monomials}
-In general, it has been shown (e.g., in~\citeN{Kirby:TC}) that local element tensors can be expressed as a sum of integrals over $K$, each integral being the product of derivatives of functions from sets of discrete spaces and, possibly, functions of some spatially varying coefficients.
+It has been shown (e.g., in~\citeN{Kirby:TC}) that local element tensors can be expressed as a sum of integrals over $K$, each integral being the product of derivatives of functions from sets of discrete spaces and, possibly, functions of some spatially varying coefficients.
 An integral of this form is called \textit{monomial}.
@@
 \label{sec:optimal-impl}
-In this section, we characterize optimality and quasi-optimality for finite element integration as well as the space of legal transformations that we need to explore to achieve it.
+In this section, we characterize optimality and quasi-optimality for finite element integration as well as the space of legal transformations that needs be explored to achieve them.
 How the exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.
@@
 \label{def:fem-loopnest}
-A finite element integration loop nest is a loop nest in which appear, in order, an imperfect free order loop, an imperfect (perfect only in special cases), linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which all expressions are in normal form.
+A finite element integration loop nest is a loop nest in which appear, in order, an imperfect free order loop, an imperfect (perfect only in special cases), linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which expressions are in normal form.
 \end{Def}
@@
 \item there are at least two symbolically identical sub-expressions (spatial sharing)
-\item there is at least one non-trivial sub-expression (an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
+\item there is at least one non-trivial sub-expression (i.e., at least an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
 \end{enumerate}
@@
 If the objective were reaching optimality and the expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
-Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
+Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
 For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold.
@@
 We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii).
-Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured; this observation is fundamental for understanding the sharing elimination algorithm, described in the next section\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form.
+Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form.
 For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}.
+This observation is fundamental for understanding the sharing elimination algorithm.
 
@@
 \label{sec:se-algo}
-Algorithm~\ref{algo:sharing-elimination} describes how we perform sharing elimination assuming as input a tree representation of the loop nest.
+Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest.
 The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest.
@@
 \item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
-$\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
+$\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
 
@@
 \textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, strategies (i) or (ii) were applied.
-We will clarify that this directly descends from loop linearity.}
+We will clarify that this directly descends from loop linearity and the structure inherent in tensor operations.}
 
@@
 
-\item If $\mathbb{S} = \emptyset$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
+\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: strategy (ii) recasts the expression as $S (...T^0 + ...
+\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^0 + ...
 + ...T^n)$}.
@@
 
-In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, the bilinear form arising from a Poisson equation in 2D.
+In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, the bilinear form arising from a Poisson equation in two dimensions.
 We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
 In addition, $|P^i| = |S_{P^i}| = 2$, so strategy (i) is applied to both partitions (step 3).
-We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{E} = \emptyset$.
+We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{S} = \mathbb{T}$.
 
@@
 
-A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}
+A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
 
@@
 \label{sec:pre-evaluation}
-Sharing elimination explores the transformation space and applies three operators: expansion, factorization, and code motion.
-In this section, we discuss role and legality of a fourth operator: \textit{reduction pre-evaluation}.
-We will see that what makes this operator special is the fact that there exists a single point in the transformation space of a monomial (i.e., specific factorization and code motion) preserving the correctness of the transformation.
+Sharing elimination uses three operators: expansion, factorization, and code motion.
+In this section, we discuss role and legality of a fourth operator: reduction pre-evaluation.
+We will see that what makes this operator special is the fact that there exists a single point in the transformation space of a monomial (i.e., a specific factorization of test, trial, and coefficient functions) ensuring its correctness.
 
@@
 The transformation we look for is exemplified in Figure~\ref{code:loopnest_rednored} with a simple loop nest.
-The reader can easily verify that a similar transformation is applicable to the example in Figure~\ref{code:poisson_a}.
+The reader may verify that a similar transformation is applicable to the example in Figure~\ref{code:poisson_a}.
 
@@
 We make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
-As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors.
-This result can be turned into a transformation algorithm for loops and expressions, for which we provide a succinct description below.
+As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces.
+A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors.
+In our model, a reference tensor is simply represented by a particular sub-expression independent of $L_e$.
+This leads to the following algorithm.
 
@@
 Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$.
-We dissect \texttt{F} into distinct sub-expressions (the monomials).
-Each sub-expression is factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
-This transformation is feasible, as a consequence of the results in~\citeN{Kirby:TC}.
+We dissect the normal form input expression into distinct sub-expressions, each of them representing a specific monomial.
+Each sub-expression is then factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
+This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py}}, as a consequence of the results in~\citeN{Kirby:TC}.
 These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries.
-As part of this process, the reduction induced by $L_i$ is evaluated.
-Consequently, $L_i$ disappears from $\Lambda$.
+As part of this process, the reduction induced by $L_i$ is computed by means of symbolic execution.
+Finally, $L_i$ is removed from $\Lambda$.
 \end{Algo}
@@
 \begin{enumerate}
-\item In contrast to what happens with hoisting in multilinear loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $J K$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $E J K$ for those depending on $[L_e, L_i, L_j, L_k] $).
-This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
-\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase, in general, the arithmetic complexity (e.g., expansion may increase the operation count).
+\item Depending on the complexity of a monomial, a certain number $t$ of temporary variables is required if pre-evaluation is performed.
+Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms).
+For certain values of ${<}t, n, S{>}$, pre-evaluation might then shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
+\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count).
 This could outweigh the gain due to pre-evaluation.
-\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination inhibits pre-evaluation, whereas pre-evaluation generally exposes further sharing elimination opportunities.
+\item A strategy for coordinating sharing elimination and pre-evaluation is needed.
+We observe that sharing elimination inhibits pre-evaluation, whereas pre-evaluation could expose further sharing elimination opportunities.
 \end{enumerate}
 
-We expand on point 1) in the next section. We address points 2) and 3) in Section~\ref{sec:optimal-synthesis}.
+We expand on point 1) in the next section, while we address points 2) and 3) in Section~\ref{sec:optimal-synthesis}.
 
@@
 \label{sec:mem-const}
-In the previous section, we provided an insight into the potentially negative effects of code motion. We now expand on this matter, starting with the following observations.
+We have just observed that code motion, as induced by monomial pre-evaluation, may excessively increase the working set size.
+Interestingly, even more aggressive code motion strategies are theoretically conceivable.
+Imagine $\Lambda$ is enclosed in a time stepping loop.
+One could think of exposing (through some transformations) and hoisting time-invariant sub-expressions for minimizing redundant computation at each time step.
+The working set size would then increase by a factor $E$, and since $E \gg I, J, K$, the gain in operation count would probably be outweighed, from a runtime viewpoint, by a much larger memory pressure.
 
-\begin{itemize}
-\item The fact that $E \gg I, J, K$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions.
-Imagine $\Lambda$ is enclosed in a time stepping loop.
-One could think of exposing (through some transformations) and hoisting any time-invariant sub-expressions to minimize redundant computation at every time step.
-The working set size could then increase by a factor $E$.
-The gain in number of operations executed could therefore be outweighed, from a runtime viewpoint, by a much larger memory pressure.
-\item For certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g.
-the last level of private cache on a conventional CPU, the shared memory on a GPU).
-For example, as explained in Section~\ref{sec:pre-evaluation}, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $J K$ for bilinear forms and of size $J$ (or $K$) for linear forms.
-This can sometimes break such a ``local memory threshold''.
-In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.
-\end{itemize}
-
-Based on these considerations, we establish two \textit{memory constraints}.
+This observation that, for certain forms and discretizations, hoisting may bring the working set to exceed the size of some level of local memory (e.g.
+the last level of private cache on a conventional CPU, the shared memory on a GPU) led us to formulating the following \textit{memory constraints}.
 
@@
 
-Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects.
-Consequently, generalized code motion as performed by sharing elimination is not allowed to hoist expressions outside of $L_e$.
+Note that Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects.
+We will carefully analyze both theoretical and practical implications of these memory constraints.
 
@@
 Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination.
-In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively.
+In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation to, respectively.
 Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$.
 We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
-Having defined $\theta$, we can now focus on the transformation algorithm (Algorithm~\ref{algo:gamma}).
+Having defined $\theta$, we can now describe the transformation algorithm (Algorithm~\ref{algo:gamma}).
 
@@
 The set of monomials that will \textit{potentially} be pre-evaluated is $P = M \setminus S$.
-\\ \textit{Note: there are two subtle yet fundamental reasons for not pre-evaluating $m_1 \in P$: 1) the presence of spatial sharing between $m_1$ and $m_2 \in S$, which impacts the search for the global optimum; 2) breaking memory constraints.}
-\item Build the set $B$ of all possible bipartitions of $P$.
-\item For each $b = (b_S, b_P) \in B$, if the memory required to store the pre-evaluated tables from the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}), discard $b$; otherwise, add an entry to the dictionary $d$ of potential operation counts, such that $d[b] = \theta_{se}(S \cup b_S) + \theta_{pre}(b_P)$.
+\\ \textit{Note: there are two fundamental reasons for not pre-evaluating $m_1 \in P$ straight away: 1) the potential presence of spatial sharing between $m_1$ and $m_2 \in S$, which impacts the search for the global optimum; 2) the risk of breaking Constraint~\ref{const:TH}.}
+\item Build the set $B$ of all possible bipartitions of $P$. Let $D$ be the dictionary that will store the operation counts of different alternatives.
+\item Discard $b = (b_S, b_P) \in B$ if the memory required after applying pre-evaluation to the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}); otherwise, add $D[b] = \theta_{se}(S \cup b_S) + \theta_{pre}(b_P)$.
 \\ \textit{Note: $\mathbb{B}$ is in practice very small, since even complex forms usually have only a few monomials.
@@
 We elaborate on this aspect in Section~\ref{sec:op_count}.}
-\item Take $b$ such that $min(d[b])$.
-\item Apply pre-evaluation to all monomials in $P \cup b_P$.
-Apply sharing elimination to the whole expression.
-\\ \textit{Note: because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary.
-Sharing elimination is therefore transparently applied to the whole expression, including what results from pre-evaluation.}
+\item Take $b$ such that $min(D[b])$.
+\item Apply pre-evaluation to all monomials in $b_P$.
+Apply sharing elimination to all resulting expressions.
+\\ \textit{Note: because of reuse of basis functions, pre-evaluation may produce some identical tables, which will be mapped to the same temporary variable.
+Sharing elimination is therefore transparently applied to all expressions, including those resulting from pre-evaluation.}
 \end{enumerate}

### 2016-01-16 TOMS_Paper/assembly.tex
again fixes here and there

@@
 We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2.
-To illustrate the concept, consider the tensor operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
+To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
 One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$.
@@
 The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$.
-In general, however, the optimal choice depends on multiple factors: the loop size, the increase in operation count due to expansion, and the gain in code motion.
+In general, however, the optimal choice depends on multiple factors: the loop(s) size, the increase in operation count due to expansion, and the gain in code motion.
 We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii).
@@
 \label{sec:se-algo}
-Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest.
-The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest.
-Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.
-
-The algorithm makes use of the following notation and terminology:
+Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
 \begin{itemize}
@@
 \begin{enumerate}
-\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. \\ 
+\item Perform a depth-first visit of the expression tree to factorize identical terms within additions. \\ 
 \textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
@@
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^0 + ...
+\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^1 + ...
 + ...T^n)$}.
@@
 \textit{Note: this pass speculatively assumes that structured tensor operations are in place.
-If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be exposed.}
+If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be considered.}
 \end{enumerate}
@@
 
-In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, the bilinear form arising from a Poisson equation in two dimensions.
+In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions.
 We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
@@
 A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors.
-In our model, a reference tensor is simply represented by a particular sub-expression independent of $L_e$.
+In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after special transformation of the expression tree.
 This leads to the following algorithm.
@@
 Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$.
-We dissect the normal form input expression into distinct sub-expressions, each of them representing a specific monomial.
+We dissect the normal form input expression into distinct sub-expressions, each of them representing a monomial.
 Each sub-expression is then factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
@@
 Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms).
-For certain values of ${<}t, n, S{>}$, pre-evaluation might then shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
+For certain values of ${<}t, n, S{>}$, pre-evaluation will shift the loop nest from a CPU-bound to a memory-bound regime, which may be counter-productive for actual execution time.
 \item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count).
@@
 
-This observation that, for certain forms and discretizations, hoisting may bring the working set to exceed the size of some level of local memory (e.g.
-the last level of private cache on a conventional CPU, the shared memory on a GPU) led us to formulating the following \textit{memory constraints}.
+Since, for certain forms and discretizations, hoisting may bring the working set to exceed the size of some level of local memory (e.g.
+the last level of private cache on a conventional CPU, the shared memory on a GPU), we introduce the following \textit{memory constraints}.
 
@@
 Note that Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects.
-We will carefully analyze both theoretical and practical implications of these memory constraints.
+It has the effect of shrinking the transformation space.
+Constraint~\ref{const:TH} has both theoretical and practical implications, which will be carefully analyzed in the next sections.
 
@@
 Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.
-
-Sharing elimination and pre-evaluation tackle the problem of minimizing spatial and temporal redundancy given a set of memory constraints in finite element integration loop nests, using \textit{a very specific set of operators}.
-This needs be emphasized since, theoretically, one could find an even lower operation count by exploiting domain-specific properties, such as redundancies in basis functions.
+We remark that sharing elimination and pre-evaluation aim to minimize spatial and temporal redundancy relying on \textit{a specific set of operators}.
+Theoretically, lower operation counts may be found by exploiting other domain-specific properties, such as redundancies in basis functions.
+Constraint~\ref{const:Le} also limits the applicability of code motion, which further reduced the transformation space to be analyzed.
 
@@
 
-The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations.
-If, on one hand, this is tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}.
+The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ would consist of applying the actual transformations and simply count the number of operations.
+If, on one hand, this would be tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}.
 We then seek an analytic way of determining $\theta_{pre}$.
@@
 The first step consists of estimating the \textit{increase factor}, $\iota$.
-This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation.
+This number captures the increase in arithmetic complexity due to the transformations exposing pre-evaluation opportunities.
 To contextualize, consider the example in Figure~\ref{code:increase_factor}.
@@
 The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$).
-In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities.
+In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $[L_i, L_j, L_k]$-dependent quantities (see Algorithm~\ref{algo:pre-evaluation}).
 By product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.
@@
 In general, however, determining $\iota$ is not so straightforward since redundant tabulations may result from common sub-expressions.
-Take the previous example.
-One can add one coefficient in the same function space as $f$, repeat the expansion, and observe that multiple sub-expressions (e.g., $b_{10}*b_{01}*...$ and $b_{01}*b_{10}*...$) will reduce to identical tables.
+Consider the previous example.
+One may add one coefficient in the same function space as $f$, repeat the expansion, and observe that multiple sub-expressions (e.g., $b_{10}*b_{01}*...$ and $b_{01}*b_{10}*...$) will reduce to identical tables.
 To evaluate $\iota$, we then use combinatorics.
 We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product; (ii) $n$ is the number of unique basis functions involved in the expansion.
-In the original example, we have $n=3$ (for $b_{i0}$, $b_{i1}$, and $b_{i2}$) and $k=1$, which confirms what we intuitively found before, namely $\iota=3$.
-In the modified example there are two coefficients, so $k=2$, which means $\iota=6$.
+In the original example, we had $n=3$ (for $b_{i0}$, $b_{i1}$, and $b_{i2}$) and $k=1$, which confirms $\iota=3$.
+In the modified example, there are two coefficients, so $k=2$, which means $\iota=6$.
 
@@
 Intuitively, this means that we are introducing more operations than we are saving from pre-evaluating $L_i$.
+If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
+The mass matrix monomial in Figure~\ref{code:increase_factor} is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
+In the example in Figure~\ref{code:poisson}, instead, we have $\rho = 3$ after a suitable factorization of basis functions.
+In general, therefore, $\rho$ depends on both form and discretization.
+To determine this parameter, we look at the re-factorized expression (as established by Algorithm~\ref{algo:pre-evaluation}), and simply count the terms amenable to pre-evaluation.
 
-If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
-Consider again the mass matrix operator in Figure~\ref{code:increase_factor}.
-The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
-If we instead take the example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after a suitable factorization of basis functions.
-In general, therefore, $\rho$ depends on both form and discretization employed.
-To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and simply count the terms amenable to pre-evaluation.
-
-\subsection{Formalization}
+\section{Formalization}
 \label{sec:proof}
-The following proposition states that our approach guarantees quasi-optimality.
+We demonstrate that the orchestration of sharing elimination and pre-evaluation performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}).
 The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.
@@
 Let $\Gamma$ be the transformation algorithm.
-Let $X$ be the set of pre-evaluated monomials, and let $Y = M \setminus X$.
+Let $X$ be the set of monomials that $\Gamma$ establishes to pre-evaluate, and let $Y = M \setminus X$.
 Assume that the pre-evaluation of different monomials does not result in identical tables.
-Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and satisfies Constraints~\ref{const:Le} and~\ref{const:TH}.
+Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and satisfies Constraint~\ref{const:TH}.
 \end{Prop}
 \begin{proof}
-We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ and $Y$ can actually be constructed correctly.
+We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ and $Y$ can actually be constructed.
 
 Let $c_\Lambda$ denote the operation count for $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}).
-We need to show that there is no other synthesis $\Lambda''_{I}$ satisfying Constraints~\ref{const:Le} and~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
+We need to show that there is no other synthesis $\Lambda''_{I}$ satisfying Constraint~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$.
 This holds if and only if
@@
 Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$.
-If $T_{m_{1}} \cap T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary.
+If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary.
 $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''.
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Because of loop linearity and normal form of expressions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols.
+Because of normal form of expressions, loop linearity and structure of tensor operations, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols.
 These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration.
-Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space.
+Since (i) is applied if and only if $|P| < |S_P|$ and does not require neither expansion nor factorization, step (3) will not prune any optimal solution from the search space.
 %; in fact, it simply shifts towards the objective.
@@
 Steps (4)-(5) factorize symbols following the ordering dictated by the sharing graph and its rewrite rules.
-Recall that rules are applied in sequence: first all rule (A) hits are handled, then all (B)s and finally all (C)s, with applications of rule (C) that may expose further possibilities for (A) and (B).
+Recall that rules (A), (B), and (C) are applied in sequence, with applications of rule (C) that may expose further possibilities for (A) and (B).
 We prove that the application of these rules, in this sequence, represents a step towards quasi-optimality.
@@
 COFFEE handles generic ASTs, so any form compiler can use it as long as such a representation is properly emitted (with expressions in normal form, or sufficiently close to it).
-In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs (instead of strings) is used.
-COFFEE itself provides an interface for building an AST.
+In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs (instead of strings, as in FEniCS) is used.
+COFFEE itself provides an interface for building ASTs.
 In particular, COFFEE aims to decouple the mathematical manipulation of a form from code optimization.
-Another viewpoint is that a form compiler developers should not worry about performance of the generated code.
+Another viewpoint is that form compiler developers should not worry about performance of the generated code.
 

### 2016-01-20 TOMS_Paper/assembly.tex
Remove edges if too costly in sharing graph plus minor fixes

@@
 
-For a finite element integration loop nest, we characterize optimality and quasi-optimality as follows.
+We then characterize optimality for a finite element integration loop nest as follows.
 
@@
 Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) performed to evaluate the result is minimal.
+\end{Def}
+
+The concept of quasi-optimality, which relies on the particular class of \textit{f-decreasing} transformations, is also introduced.
+
+\begin{Def}[F-decreasing transformation]
+A sequence of transformations reducing the operation count is called f-decreasing (abbreviation for ``flop-decreasing'').
 \end{Def}
@@
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is minimal in all innermost loops.
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) performed to evaluate the result, achieved through a sequence of f-decreasing transformations, is minimal in all innermost loops.
 \end{Def}
 
-Note that Definitions~\ref{def:mln-optimality} and~\ref{def:mln-quasi-optimality} do not take into account memory requirements.
+The restriction to f-decreasing transformations aims to exclude those ``deceptive'' solutions that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing a ``global'' increase in operation count.
+
+We also observe that Definitions~\ref{def:mln-optimality} and~\ref{def:mln-quasi-optimality} do not take into account memory requirements.
 If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
@@
 In the context of finite element, this is often true for more complex multilinear forms and/or higher order elements.
-We also note that quasi-optimality approximates optimality very well whenever the inner loop trip counts are ``sufficiently large''.
 
@@
 
-\item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step 3.
+\item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step (3).
 An edge $(S^i$, $S^j)$ indicates that a product $S^i S^j$ would appear if the sub-expressions including $S^i$ and $S_j$ were expanded.
-Let $\mathbb{T}$ be the subset of vertices with a single incident edge, or ``terminals''.
+Let $\mathbb{T}$ be the subset of vertices with at most one incident edge, or ``terminals''.
+
+\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying strategy (ii) would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
+\textit{Note: this ensures the f-decreasing nature of the transformations at step (6)}.\\
+\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost can be derived.}
 
@@
 
-%TODO : dire che il costo dell'espansione è semplicemente dato da ciascun s in | S_P | times la 2 evalato al depth di quell s.
-%\item $\mathbb{T} = \emptyset$: take $S$ at will such that the number of adjacent non terminals is maximum, and apply strategy (ii). 
-%\textit{Note: $\mathbb{S}$ is in practice very small, so this pass can be accomplished rapidly.}
+\begin{Rem*}
+\normalfont
+It is true that Algorithm~\ref{algo:sharing-elimination} aims for quasi-optimality, but clearly this does not mean that only those sub-expressions depending on the multilinear loop nest are considered for optimization.
+We emphasize once more that the enforcement of f-decreasing transformations (step 5) and the re-scheduling within outer loops (step 7) attempt to minimize the operation count of the whole loop nest.
+We will further elaborate this aspect in Section~\ref{sec:proof}.
+\end{Rem*}
 
@@
 Let $\Gamma$ be the transformation algorithm.
-Let $X$ be the set of monomials that $\Gamma$ establishes to pre-evaluate, and let $Y = M \setminus X$.
+Let $X$ be the set of monomials that, according to $\Gamma$, need be pre-evaluated, and let $Y = M \setminus X$.
 Assume that the pre-evaluation of different monomials does not result in identical tables.
@@
 If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum.
-This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum that also preserves the memory constraints.
+This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum preserving Constraint~\ref{const:TH}.
 Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Because of normal form of expressions, loop linearity and structure of tensor operations, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols.
-These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration.
-Since (i) is applied if and only if $|P| < |S_P|$ and does not require neither expansion nor factorization, step (3) will not prune any optimal solution from the search space.
-%; in fact, it simply shifts towards the objective.
+Because of the normal form of expressions and the structure inherent in tensor operations involving test and trial functions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (strategy (i)) or factorization of multilinear symbols (achieved through strategy (ii)).
+If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
+Since strategy (i) is applied if and only if $|P| < |S_P|$ and requires neither expansion nor factorization, step (3) will not prune any optimal solution from the search space; in fact, it simply represents a step forward towards quasi-optimality.
 
-Steps (4)-(5) factorize symbols following the ordering dictated by the sharing graph and its rewrite rules.
-Recall that rules (A), (B), and (C) are applied in sequence, with applications of rule (C) that may expose further possibilities for (A) and (B).
-We prove that the application of these rules, in this sequence, represents a step towards quasi-optimality.
+Step (6) factorizes symbols following the ordering dictated by the sharing graph and its rewrite rules.
+Non f-decreasing factorizations are excluded by construction (step (5)).
+We now have to prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), always represents progress towards quasi-optimality.
 \begin{description}
@@
 \item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
-Let us consider the alternative factorization $S^{nt}(...S + ...
+Let us consider a generic alternative factorization $S^{nt}(...S + ...
 + ...V)$.
-If $V$ is a terminal, we are back to where we started, because rule (B) will be applicable by setting $V = T$ and $S = S^{nt}$.
-Otherwise ($V$ non terminal), we have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
+If $V$ were a terminal, we would be back to where we started, because rule (B) with $V = T$ and $S = S^{nt}$ would still be applicable.
+Otherwise ($V$ non terminal), we would have $S$ appearing in two different sub-expressions, which would imply sub-optimality as a consequence of spatial sharing.
 \item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
-For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete.
+For each disconnected component $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete.
 Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
-This is reachable by factorizing the vertex with highest degree (as the rule states), as any other choice may break factorization opportunities.
-In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).
+This is reachable by factorizing the vertex with highest degree, as any other choice may break factorization opportunities.
+In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (6).
 \end{description}
-The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \emptyset$, before applying the rewrite rules. 
+The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \mathbb{T}$. 
 \end{enumerate}
@@
 %TODO : In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
+%TODO : A list of preconditions that eventually result in optimality
 

### 2016-01-20 TOMS_Paper/assembly.tex
Went on adding stuff to formalization

@@
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^1 + ...
-+ ...T^n)$}.
-\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
-\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii).
+\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii).\\
+\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
+\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply strategy (ii).\\
+\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}. 
+\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii) such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
+\textit{Note: this recasts the expression as $S (...S^1 + ... + ...S^n)$}.
 \end{enumerate}
@@
 \label{sec:proof}
-We demonstrate that the orchestration of sharing elimination and pre-evaluation performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}).
+We demonstrate that the orchestration of sharing elimination and pre-evaluation as performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}).
 The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.
@@
 If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
-Since strategy (i) is applied if and only if $|P| < |S_P|$ and requires neither expansion nor factorization, step (3) will not prune any optimal solution from the search space; in fact, it simply represents a step forward towards quasi-optimality.
+Since strategy (i) is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) will not prune optimal solutions from the search space.
 
-Step (6) factorizes symbols following the ordering dictated by the sharing graph and its rewrite rules.
-Non f-decreasing factorizations are excluded by construction (step (5)).
-We now have to prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), always represents progress towards quasi-optimality.
+Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules.
+Non f-decreasing factorizations are excluded by construction at step (5).
+We then prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality.
 \begin{description}
-\item[Rule (A)] $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule is side effects free (i.e., no need to choose between different factorizations).
+\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
 \item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
-Let us consider a generic alternative factorization $S^{nt}(...S + ...
+Let us consider the alternative factorization $S^{nt}(...S + ...
 + ...V)$.
-If $V$ were a terminal, we would be back to where we started, because rule (B) with $V = T$ and $S = S^{nt}$ would still be applicable.
-Otherwise ($V$ non terminal), we would have $S$ appearing in two different sub-expressions, which would imply sub-optimality as a consequence of spatial sharing.
-\item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
+If $V$ were a terminal, it is easy to see that setting $V = T$ and $S = S^{nt}$ would just expose another possible hit for rule (B).
+Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ...
++ ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$.
+From this, we see that applying rule (C) prior to rule (B) may prune the optimal solution from the search space.
+\item[Rule (C)] We interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
 For each disconnected component $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete.
-Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
+Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), and thanks to loop linearity, the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
 This is reachable by factorizing the vertex with highest degree, as any other choice may break factorization opportunities.
-In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (6).
+In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$).
+This lower bound is achieved by repeating the sharing elimination algorithm (notice the jump back to step (2) at the end of step (6)).
 \end{description}
-The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \mathbb{T}$. 
+The algorithm is guaranteed to terminate because of the jump to step (6) as $\mathbb{S} = \mathbb{T}$. 
 \end{enumerate}
@@
 
-%TODO : In addition, if X holds, then $\Lambda'$ is optimal as established by Definition ?.
-%TODO : A list of preconditions that eventually result in optimality
+%\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii)
+%\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply strategy (ii).\\
+%\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii) such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
+
+Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if some preconditions are satisfied by the input variational form.
+We here discuss such preconditions, in increasing order of complexity.
+\begin{enumerate}
+\item There is a single monomial and only a specific coefficient, the coordinates field.
+This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
+\item There is a single monomial, but multiple coefficients are present.
+Optimality is achieved if only structured tensor operations are present (see Section~\ref{sec:se-rln}), since there is no ambiguity in factorization, which is exploited during step (7) of Algorithm~\ref{algo:sharing-elimination}.
+\item There are multiple monomials, but no coefficients other than the coordinates field are present, or there is no spatial sharing in coefficients across different monomials.
+This reduces to cases (1) and (2) above.
+\item There are multiple monomials and there is sharing of coefficients across different monomials. We can still in some cases ....
+\end{enumerate}
 

### 2016-01-22 TOMS_Paper/assembly.tex
Various fixes all around

@@
 
-\subsubsection{Structured tensor operations}
+\subsubsection{Identification and exploitation of structure}
 \label{sec:se-rln}
@@
 To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
-One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$.
-To schedule sub-expressions of this kind, we have two options:
+The tensor operation will reduce to $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$.
+To minimize the operation count for sub-expressions of this kind, we have two options:
 \begin{enumerate}[label=(\roman*)]
@@
 In general, however, the optimal choice depends on multiple factors: the loop(s) size, the increase in operation count due to expansion, and the gain in code motion.
-We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii).
-Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form.
+We regard as \textit{structured} any sub-expression resulting from tensor operations for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii).
+Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions involving test and trial functions are structured\footnote{It is also remarkable that structured sub-expressions often characterize the \textit{entire} variational form.
 For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}.
@@
 \item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
-\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured tensor operations.
+\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured sub-expressions.
 \end{itemize}
@@
 \item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
-\textit{Note: this pass speculatively assumes that structured tensor operations are in place.
-If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be considered.}
+\textit{Note: this pass speculatively assumes all sub-expressions to be structured.
+If the assumption does not hold, the result will generally be sub-optimal since only a subset of factorizations and code motion opportunities may be considered.}
 \end{enumerate}
@@
 \label{sec:optimal-synthesis}
-In this section, we build a transformation algorithm that, given a memory bound, produces optimal or quasi-optimal finite element integration loop nests.
+In this section, we build a transformation algorithm that, given a memory bound, systematically produces quasi-optimal finite element integration loop nests.
 
@@
 
-Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if some preconditions are satisfied by the input variational form.
+Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if stronger preconditions on the input variational form are satisfied.
 We here discuss such preconditions, in increasing order of complexity.
 \begin{enumerate}
-\item There is a single monomial and only a specific coefficient, the coordinates field.
+\item There is a single monomial and only a specific coefficient (e.g., the coordinates field).
 This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
 \item There is a single monomial, but multiple coefficients are present.
-Optimality is achieved if only structured tensor operations are present (see Section~\ref{sec:se-rln}), since there is no ambiguity in factorization, which is exploited during step (7) of Algorithm~\ref{algo:sharing-elimination}.
-\item There are multiple monomials, but no coefficients other than the coordinates field are present, or there is no spatial sharing in coefficients across different monomials.
-This reduces to cases (1) and (2) above.
-\item There are multiple monomials and there is sharing of coefficients across different monomials. We can still in some cases ....
+Optimality is achieved if and only if all sub-expressions depending on coefficients are structured (see Section~\ref{sec:se-rln}).
+This would avoid ambiguity in factorization, which in turn would guarantee the output of step (7) in Algorithm~\ref{algo:sharing-elimination} be optimality.
+\item There are multiple monomials, but either at most one coefficient (e.g., the coordinates field) or multiple coefficients not inducing sharing across different monomials are present.
+This reduces, respectively, to cases (1) and (2) above.
+\item There are multiple monomials, and coefficients are shared across monomials.
+Optimality is reached if and only if the coefficient-dependent sub-expressions produced by Algorithm~\ref{algo:sharing-elimination} -- that is, the by-product of factorizing test/trial functions from distinct monomials -- preserve structure.
 \end{enumerate}
@@
 
-As discussed in the next section, all code generation systems used for performance evaluation handle block-sparse tables.
+%As discussed in the next section, all code generation systems used for performance evaluation handle block-sparse tables.
 
@@
 The Intel \texttt{icc 15.2} compiler was used.
-The compilation flags used were \texttt{-O3, -xHost, -ip}.
+The compilation flags used were \texttt{-O3, -xHost}.
 
-We analyze the runtime performance of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
+We analyze the execution time of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
 In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}'').
@@
 The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time).
-They include the cost of local assembly as well as the cost of matrix insertion.
-However, the unstructured mesh used for the simulations (details below) was chosen small enough to fit the L3 cache of the CPU so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.
+They include the cost of both local assembly and matrix insertion, with the latter minimized through the choice of a mesh (details below) small enough to fit the L3 cache of the CPU.
 
-For a fair comparison, small patches (publicly available) were written to run \textit{all} simulations through Firedrake.
-This means the costs of matrix insertion and mesh iteration are identical in all variants.
-Our patches make UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for Firedrake, which employs a data storage layout different than that of FEniCS (e.g., array of pointers instead of pointer to pointers).
+For a fair comparison, small patches (publicly available) were written for compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake.
+By executing all simulations in Firedrake, we guarantee that matrix insertion and mesh iteration have a fixed cost independent of the transformation system employed.
+The patches, in particular, adjust the data storage layout (e.g., by generating array of pointers instead of pointer to pointers and by replacing flattened arrays with bi-dimensional ones, as Firedrake expects).
 
-In Section~\ref{sec:mem-const}, we discussed the importance of memory constraints.
-We then define $T_H$ as the maximum amount of space that temporaries due to code motion can take.
-We set $T_H = L2_{size}$, that is, the size of the processor L2 cache (the last level of private cache).
-We recall that exceeding this threshold prevents the application of pre-evaluation.
-In our experiments, this happened in some circumstances.
-In such cases, experiments were repeated with $T_H = L3_{size}$ to verify the hypotheses made in Section~\ref{sec:mem-const}.
-We later elaborate on this.
+For Constraint~\ref{const:TH}, discussed in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the size of the processor L2 cache (the last level of private cache).
+When the threshold had an impact on the transformation process, the experiments were repeated with $T_H = L3_{size}$.
+The results are documented later, individually for each problem.
 
@@
 
-The rest of the section is structured as follows: we provide insights about the main message of the experimentation; we comment on the impact of autovectorization; we explain in detail, individually for each form, the performance results obtained.
+The rest of the section is organized as follows: we first provide insights into the overall outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we explain, in detail and individually for each form, the performance results obtained.
 
@@
 For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization.
-Secondly, memory constraints are critical, particularly the transformation strategy adpopted when exceeding $T_H$.
+Secondly, memory constraints are critical, particularly the transformation strategy adopted when exceeding $T_H$.
 We will later thoroughly elaborate on all these aspects.
@@
 In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper.
-This is a positive aspect: it means that our transformation algorithm scales with form complexity.
-The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests several points.
-Firstly, the baseline is highly inefficient: with forms as complex as in this hyperelastic model, a trivial translation of integration routines into code should always be avoided since even one of the best general-purpose compilers available (Intel's on an Intel platform at maximum optimization level) is not capable of exploiting the structure inherent in the mathematical expressions generated.
-Secondly, the code motion strategy really makes a considerable impact.
-The sharing elimination performed by \texttt{cfO2} in each level of the loop nest ensures a critical reduction in operation count, which results is better execution times.
-In particular, at higher order, the main difference between \texttt{ufls} and \texttt{cfO2} is due to the application of this transformation to the multilinear loop nest.
-Clearly, the operation count increases with $q$, and so do the speed-ups.
+This is a positive aspect, since it indicates that our transformation algorithm scales well with form complexity.
+The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points.
+Firstly, the baseline is highly inefficient.
+With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) cannot exploit the structure inherent in the expressions.
+Secondly, the strategy for removing spatial and temporal sharing has a tremendous impact.
+Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.
 
@@
 \label{sec:conclusions}
-With this research we have set the foundation of optimal finite element integration.
-We have developed theory and implemented an automated system capable of applying it.
-The automated system, COFFEE, is integrated in Firedrake, a real-world framework for writing finite element methods.
-We believe the results are extremely positive.
-An open problem is understanding how to optimally handle non-linear loop nests.
-A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.
-Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.
+We have developed transformation algorithms and a theory of optimality for finite element integration.
+The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made available to an increasingly significant user base.
+Performance results show the effectiveness of our methodology.
+There are still several open problems.
+The main one is an elaborate classification of forms for which optimality can be reached.
+Secondly, we would like to extend our methodology to classes of loops arising in spectral methods, in which interaction with low level optimization becomes even more critical due to the larger working sets.
 

### 2016-01-23 TOMS_Paper/assembly.tex
intermediate commit ASD

@@
 \newtheorem{Const}{Constraint}
+\newtheorem{Strategy}{Strategy}
 \newtheorem*{Rem*}{Remark}
@@
 \begin{Def}[Linear loop]
+\label{def:linear-loop}
 A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is linear if in its body
@@
 
-\begin{Def}[Free order loop]
+\begin{Def}[Order-free loop]
 \label{def:e-loop}
-A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, there are no loop-carried dependencies across different iterations. 
+A loop $L_i$ is said to be an order-free loop if its iterations can be executed in any arbitrary order. 
 \end{Def}
@@
 
-To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
-The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions).
+Consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}.
+The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises an order-free loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions).
 In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$.
@@
 \label{def:fem-loopnest}
-A finite element integration loop nest is a loop nest in which appear, in order, an imperfect free order loop, an imperfect (perfect only in special cases), linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which expressions are in normal form.
+A finite element integration loop nest is a loop nest in which the following appear, in order: an imperfect order-free loop, an imperfect (perfect only in some special cases), linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which expressions are in normal form.
 \end{Def}
@@
 Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
-We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the number of operations (additions, products) performed to evaluate the result is minimal.
+We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the total number of operations (additions, products) performed to evaluate the result is minimal.
 \end{Def}
@@
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) performed to evaluate the result, achieved through a sequence of f-decreasing transformations, is minimal in all innermost loops.
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) in the innermost loops performed to evaluate the result is minimal and was achieved through a sequence of f-decreasing transformations.
 \end{Def}
@@
 Henceforth we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated.
-In the context of finite element, this is often true for more complex multilinear forms and/or higher order elements.
+In the context of finite elements, this is often true for more complex multilinear forms and/or higher order elements.
 
 Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
-On the other hand, multilinearity ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions.
+On the other hand, multilinearity suggest a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
 In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
@@
 A statement within a loop nest $\Lambda$ presents sharing if at least one of the following conditions hold:
-\begin{enumerate}
-\item there are at least two symbolically identical sub-expressions (spatial sharing)
-\item there is at least one non-trivial sub-expression (i.e., at least an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
-\end{enumerate}
+\begin{description}
+\item[Spatial sharing] there are at least two symbolically identical sub-expressions
+\item[Temporal sharing] there is at least one non-trivial sub-expression (e.g., an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_1}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$
+\end{description}
 \end{Def}
@@
 In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion.
-If the objective were reaching optimality and the expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
+If the objective were reaching optimality and the expressions lacked structure, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
 Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
@@
 Finite element expressions can be seen as composition of operations between tensors.
-We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2.
+We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy can be determined by comparing two possibilities.
 To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
-The tensor operation will reduce to $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$.
-To minimize the operation count for sub-expressions of this kind, we have two options:
-\begin{enumerate}[label=(\roman*)]
-\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
-\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
-\end{enumerate}
-The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$.
-In general, however, the optimal choice depends on multiple factors: the loop(s) size, the increase in operation count due to expansion, and the gain in code motion.
-We regard as \textit{structured} any sub-expression resulting from tensor operations for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii).
-Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions involving test and trial functions are structured\footnote{It is also remarkable that structured sub-expressions often characterize the \textit{entire} variational form.
+The tensor operation will reduce to $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$.
+To minimize the operation count for sub-expressions of this kind, we have two options.
+\begin{Strategy}
+\label{strategy:i}
+Applying generalized code motion, in presence of temporal sharing.
+\end{Strategy}
+\begin{Strategy}
+\label{strategy:ii}
+Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
+\end{Strategy}
+In the current example, we observe that applying Strategy~\ref{strategy:ii} would improve the operation count independently of the size of $L_i$, since the expression would be recast as $x^0_i (a + c) + x^1_i (b + d)$, thus exposing the two hoistable sub-expressions $(a + c)$ and $(b + d)$.
+On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$.
+In general, however, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion.
+Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
+These examples motivate the introduction of a particular class of expressions.
+\begin{Def}[Structured expression]
+\label{def:struct-expr}
+An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in how to eliminate the spatial sharing induced by the symbols depending on $L$.
+\end{Def}
+The fundamental observation is the fact that a multilinear loop nest $L$ makes expressions structured with respect to $L$.
+In finite element integration loop nests, for example, a factorization of test and trial functions always results in exposing temporal sharing (Strategy~\ref{strategy:ii}).
+This will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality.
+Finally, it is remarkable that structured sub-expressions often characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
+
+In the following, we regard as \textit{structured} any sub-expression resulting from tensor operations for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
+From the perspective of achieving quasi-optimality, it is fundamental to observe how multilinearity promotes structured expressions.
+: factorization of test and trial functions will always expose temporal sharing, \footnote{It is also remarkable that structured sub-expressions often characterize the \textit{entire} variational form.
 For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}.
-This observation is fundamental for understanding the sharing elimination algorithm.
+This is a consequence of multilinearity, which ensures that the factorization of any test or trial function will always expose some hoistable sub-expression (i.e., temporal sharing).
 
@@
 \item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
-\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured sub-expressions.
 \end{itemize}
@@
 
-\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply strategy (i) to the multilinear operands. \\
-\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, strategies (i) or (ii) were applied.
+\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i} to the multilinear operands. \\
+\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied.
 We will clarify that this directly descends from loop linearity and the structure inherent in tensor operations.}
@@
 
-\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying strategy (ii) would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
+\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying Strategy~\ref{strategy:ii} would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
 \textit{Note: this ensures the f-decreasing nature of the transformations at step (6)}.\\
@@
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii).\\
+\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply Strategy~\ref{strategy:ii}.\\
 \textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
-\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply strategy (ii).\\
+\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
 \textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}. 
-\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii) such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
+\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii} such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
 \textit{Note: this recasts the expression as $S (...S^1 + ... + ...S^n)$}.
@@
 We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$.
-For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying strategy (i) in step 3 has no effect.
+For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step 3 has no effect.
 The sharing graph is $\mathbb{G} = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $\mathbb{T} = \lbrace c_i, d_i \rbrace$.
@@
 We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
-In addition, $|P^i| = |S_{P^i}| = 2$, so strategy (i) is applied to both partitions (step 3).
+In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step 3).
 We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{S} = \mathbb{T}$.
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Because of the normal form of expressions and the structure inherent in tensor operations involving test and trial functions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (strategy (i)) or factorization of multilinear symbols (achieved through strategy (ii)).
+Because of the normal form of expressions and the structure inherent in tensor operations involving test and trial functions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (achieved through Strategy~\ref{strategy:ii}).
 If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
-Since strategy (i) is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) will not prune optimal solutions from the search space.
+Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) will not prune optimal solutions from the search space.
 

### 2016-01-23 TOMS_Paper/assembly.tex
Trying to improve again description of sharing elimination

@@
 Finite element expressions can be seen as composition of operations between tensors.
-We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy can be determined by comparing two possibilities.
-To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
-The tensor operation will reduce to $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$.
+Often, the optimal implementation strategy for these operations is to be determined out of two alternatives.
+To illustrate the concept, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
+The tensor operation will reduce to $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$.
 To minimize the operation count for sub-expressions of this kind, we have two options.
@@
 \label{strategy:i}
-Applying generalized code motion, in presence of temporal sharing.
+Eliminating temporal sharing through generalized code motion.
 \end{Strategy}
@@
 \end{Strategy}
-In the current example, we observe that applying Strategy~\ref{strategy:ii} would improve the operation count independently of the size of $L_i$, since the expression would be recast as $x^0_i (a + c) + x^1_i (b + d)$, thus exposing the two hoistable sub-expressions $(a + c)$ and $(b + d)$.
+In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count, since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed.
 On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$.
-In general, however, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion.
+In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion.
 Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
@@
 \label{def:struct-expr}
-An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in how to eliminate the spatial sharing induced by the symbols depending on $L$.
-\end{Def}
-The fundamental observation is the fact that a multilinear loop nest $L$ makes expressions structured with respect to $L$.
-In finite element integration loop nests, for example, a factorization of test and trial functions always results in exposing temporal sharing (Strategy~\ref{strategy:ii}).
-This will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality.
-Finally, it is remarkable that structured sub-expressions often characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
-
-In the following, we regard as \textit{structured} any sub-expression resulting from tensor operations for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
-From the perspective of achieving quasi-optimality, it is fundamental to observe how multilinearity promotes structured expressions.
-: factorization of test and trial functions will always expose temporal sharing, \footnote{It is also remarkable that structured sub-expressions often characterize the \textit{entire} variational form.
-For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}.
-This is a consequence of multilinearity, which ensures that the factorization of any test or trial function will always expose some hoistable sub-expression (i.e., temporal sharing).
+An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in spatial sharing elimination for the symbols depending on $L$.
+\end{Def}
+The fundamental observation is that if a loop nest $L$ is multilinear, then all of its expressions will be structured with respect to $L$.
+In finite element integration loop nests, for example, a factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}).
+This property will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality.
+Finally, we observe that it is remarkable that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
 
@@
 \textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied.
-We will clarify that this directly descends from loop linearity and the structure inherent in tensor operations.}
+This is a consequence of the expression being structured with respect to $[L_j, L_k]$.}
 
@@
 \textit{Note: this ensures the f-decreasing nature of the transformations at step (6)}.\\
-\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost can be derived.}
-
-\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
+\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several kinds of information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost will be derived.}
+
+\item If $\mathbb{S} = \mathbb{T}$, jump to step (7). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
 \begin{enumerate}[label=(\Alph*)]
@@
 
-\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
-\textit{Note: this pass speculatively assumes all sub-expressions to be structured.
-If the assumption does not hold, the result will generally be sub-optimal since only a subset of factorizations and code motion opportunities may be considered.}
+\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted expression, apply the most profitable between Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
+\\
+\textit{Note: this pass speculatively assumes that all expressions are structured.
+If the assumption does not hold, the operation count within the reduction loop will generally be sub-optimal because only a subset of factorizations and code motion opportunities may eventually be considered.}
 \end{enumerate}
@@
 \normalfont
-It is true that Algorithm~\ref{algo:sharing-elimination} aims for quasi-optimality, but clearly this does not mean that only those sub-expressions depending on the multilinear loop nest are considered for optimization.
-We emphasize once more that the enforcement of f-decreasing transformations (step 5) and the re-scheduling within outer loops (step 7) attempt to minimize the operation count of the whole loop nest.
+Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also aim at optimizing the loop nest ``globally''.
 We will further elaborate this aspect in Section~\ref{sec:proof}.

### 2016-01-23 TOMS_Paper/assembly.tex
Other improvements to sharing elimination description

@@
 \label{def:struct-expr}
-An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in spatial sharing elimination for the symbols depending on $L$.
-\end{Def}
-The fundamental observation is that if a loop nest $L$ is multilinear, then all of its expressions will be structured with respect to $L$.
-In finite element integration loop nests, for example, a factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}).
-This property will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality.
-Finally, we observe that it is remarkable that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
+An expression is \textit{structured} with respect to a set of loops $\Lambda$ if and only if there is no ambiguity in the elimination of spatial sharing for the symbols depending on $\Lambda$.
+\end{Def}
+This leads us to the following, fundamental observation.
+\begin{Prop}
+\label{prop:multi-struct}
+In a multilinear loop nest $\Lambda$, all of the expressions are structured with respect to $\Lambda$.
+\end{Prop}
+In finite element integration loop nests, for example, any factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}).
+Conversely, in a generic loop nest $\Lambda$, the instances of a symbol depending on $\Lambda$ may appear everywhere (e.g., twice in a summand and only once in all other summands; as argument of a higher level function; in the denominator of a division), thus posing the challenge of finding the optimal factorization that maximizes temporal sharing.
+Proposition~\ref{prop:multi-struct} played a key role in the development of Algorithm~\ref{algo:sharing-elimination}.
+Finally, we observe that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
 
@@
 \end{itemize}
+Examples will be provided in Section~\ref{sec:se-examples}.
 
@@
 \begin{enumerate}
-\item Perform a depth-first visit of the expression tree to factorize identical terms within additions. \\ 
+\item Perform a depth-first visit of the expression tree to group identical terms within the same sub-expression. \\ 
 \textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
@@
 
-\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i} to the multilinear operands. \\
+\item For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
 \textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied.
@@
 \normalfont
-Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also aim at optimizing the loop nest ``globally''.
+Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also attempt to optimize the loop nest ``globally''.
 We will further elaborate this aspect in Section~\ref{sec:proof}.

### 2016-01-27 TOMS_Paper/assembly.tex
Many improvementes, including structured expressions

@@
 
-We will show that multilinear loop nests, which arise naturally when translating bilinear or linear forms into code, are important because they have a structure that we can take advantage of to synthesize optimal code.
+We will show that multilinear loop nests, which arise naturally when translating bilinear or linear forms into code, are important because they have a structure that we can take advantage of to synthesize quasi-optimal code.
 
@@
 
-The restriction to f-decreasing transformations aims to exclude those ``deceptive'' solutions that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing a ``global'' increase in operation count.
+The restriction to f-decreasing transformations aims to exclude those ``deceptive'' solutions that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing, in fact, a ``global'' increase in operation count.
 
@@
 On the other hand, multilinearity suggest a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
-In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases.
+In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) optimality whenever the monomials are ``sufficiently structured'';  (ii) quasi-optimality in all other cases.
 To this purpose, we will construct:
@@
 To illustrate the definition, we show in Figure~\ref{code:multi_loopnest} how sharing evolves as factorization and code motion are applied to a trivial multilinear loop nest.
-In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by $b_j$.
+In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by the symbol $b_j$.
 Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}).
-Finally, generalized code motion \cite{Luporini}, which seeks redundant sub-expressions analyzing all levels of the loop nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only considers innermost loops.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).
+Finally, generalized code motion \cite{Luporini}, which seeks sub-expressions that are redundantly executed by at least one loop in the nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only checks invariance with respect to the innermost loop.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).
 
@@
 Often, the optimal implementation strategy for these operations is to be determined out of two alternatives.
-To illustrate the concept, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
-The tensor operation will reduce to $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$.
-To minimize the operation count for sub-expressions of this kind, we have two options.
+For instance, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
+The tensor operation will reduce to the scalar expression $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on $L_i$.
+To minimize the operation count for expressions of this kind, we have two options.
 \begin{Strategy}
@@
 \end{Strategy}
-In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count, since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed.
+In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed.
 On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$.
 In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion.
-Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
-These examples motivate the introduction of a particular class of expressions.
+A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
+With the objective of characterizing when these two strategies suffice to reach quasi-optimality, we introduce a particular class of expressions and its relationship with multilinear loop nests.
 \begin{Def}[Structured expression]
 \label{def:struct-expr}
-An expression is \textit{structured} with respect to a set of loops $\Lambda$ if and only if there is no ambiguity in the elimination of spatial sharing for the symbols depending on $\Lambda$.
-\end{Def}
-This leads us to the following, fundamental observation.
+We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if for eliminating the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
+\end{Def}
 \begin{Prop}
 \label{prop:multi-struct}
-In a multilinear loop nest $\Lambda$, all of the expressions are structured with respect to $\Lambda$.
+An expression along a multilinear loop nest is structured.
 \end{Prop}
-In finite element integration loop nests, for example, any factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}).
-Conversely, in a generic loop nest $\Lambda$, the instances of a symbol depending on $\Lambda$ may appear everywhere (e.g., twice in a summand and only once in all other summands; as argument of a higher level function; in the denominator of a division), thus posing the challenge of finding the optimal factorization that maximizes temporal sharing.
-Proposition~\ref{prop:multi-struct} played a key role in the development of Algorithm~\ref{algo:sharing-elimination}.
-Finally, we observe that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
+\begin{proof}
+This is an obvious consequence of multilinearity.
+\end{proof}
+If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing.
+If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} we can build the space of f-decreasing transformations for quasi-optimality by just composing Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
+The challenge for Algorithm~\ref{algo:sharing-elimination} will be to select the best f-decreasing transformations.
+
+Finally, we observe that the $\sigma_{ei}$ sub-expressions are sometimes ``weakly structured''; that is, Definition~\ref{def:struct-expr} applies to them by replacing spatial sharing ``elimination'' with ``minimization''.
+This happens, for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}, and intuitively explains why under some circumstances (elaborated in Section~\ref{sec:proof}) Algorithm~\ref{algo:sharing-elimination} is capable of reaching actual optimality.
+
+%.the application of Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} to all sub-expressions in which $s_{\Lambda}$ appears
+% the spatial sharing induced by $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- can be eliminated through factorization of all instances of $s_{\Lambda}$.
 
@@
 \textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied.
-This is a consequence of the expression being structured with respect to $[L_j, L_k]$.}
+This is a consequence of Proposition~\ref{prop:multi-struct}.}
 
@@
 \\
-\textit{Note: this pass speculatively assumes that all expressions are structured.
-If the assumption does not hold, the operation count within the reduction loop will generally be sub-optimal because only a subset of factorizations and code motion opportunities may eventually be considered.}
+\textit{Note: this pass speculatively assumes that expressions are (weakly) structured along the reduction loop.
+If the assumption does not hold, the operation count will generally be sub-optimal because only a subset of factorizations and code motion opportunities may eventually be considered.}
 \end{enumerate}
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Because of the normal form of expressions and the structure inherent in tensor operations involving test and trial functions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (achieved through Strategy~\ref{strategy:ii}).
+Proposition~\ref{prop:multi-struct} states that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (achieved through Strategy~\ref{strategy:ii}).
 If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
-Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) will not prune optimal solutions from the search space.
+Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimal solution from the search space.
 

### 2016-01-27 TOMS_Paper/assembly.tex
Many improvementes, including structured expressions

@@
 A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
-With the objective of characterizing when these two strategies suffice to reach quasi-optimality, we introduce a particular class of expressions and its relationship with multilinear loop nests.
+These examples motivate the introduction of a particular class of expressions, for which the two strategies have notable importance.
 \begin{Def}[Structured expression]
@@
 If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing.
-If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} we can build the space of f-decreasing transformations for quasi-optimality by just composing Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
-The challenge for Algorithm~\ref{algo:sharing-elimination} will be to select the best f-decreasing transformations.
+If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of f-decreasing transformations for quasi-optimality can be built by simply composing Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
+The challenge for Algorithm~\ref{algo:sharing-elimination} will then be to select the best f-decreasing transformations.
 

### 2016-02-01 TOMS_Paper/assembly.tex
minor tweaks here and there

@@
 \begin{Def}[Multilinear loop nest]
+\label{def:multi-linear-loop}
 A multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
@@
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) in the innermost loops performed to evaluate the result is minimal and was achieved through a sequence of f-decreasing transformations.
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) in the innermost loops performed to evaluate the result is minimal and was achieved through f-decreasing transformations.
 \end{Def}
@@
 Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
-On the other hand, multilinearity suggest a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
-In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) optimality whenever the monomials are ``sufficiently structured'';  (ii) quasi-optimality in all other cases.
+On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
+In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least quasi-optimality in all cases; (ii) optimality whenever the monomials are ``sufficiently structured''.
 To this purpose, we will construct:
@@
 \begin{description}
-\item[Spatial sharing] there are at least two symbolically identical sub-expressions
-\item[Temporal sharing] there is at least one non-trivial sub-expression (e.g., an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_1}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$
+\item[Spatial sharing] There are at least two symbolically identical sub-expressions
+\item[Temporal sharing] There is at least one non-trivial sub-expression (e.g., an addition or a product) that, as independent of $\lbrace L_{i_1}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$, is redundantly executed.
 \end{description}
@@
 A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
-These examples motivate the introduction of a particular class of expressions, for which the two strategies have notable importance.
+These examples motivate the introduction of a particular class of expressions, for which the two strategies assume notable importance.
 \begin{Def}[Structured expression]
 \label{def:struct-expr}
-We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if for eliminating the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
+We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if in order to eliminate the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
 \end{Def}
@@
 \begin{proof}
-This is an obvious consequence of multilinearity.
+This descends directly from Definition~\ref{def:linear-loop} and Definition~\ref{def:multi-linear-loop}, which essentially restrict the number of occurrences of a symbol $s_{\Lambda}$ in a summand to at most 1.
 \end{proof}
 If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing.
-If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of f-decreasing transformations for quasi-optimality can be built by simply composing Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
-The challenge for Algorithm~\ref{algo:sharing-elimination} will then be to select the best f-decreasing transformations.
-
-Finally, we observe that the $\sigma_{ei}$ sub-expressions are sometimes ``weakly structured''; that is, Definition~\ref{def:struct-expr} applies to them by replacing spatial sharing ``elimination'' with ``minimization''.
-This happens, for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}, and intuitively explains why under some circumstances (elaborated in Section~\ref{sec:proof}) Algorithm~\ref{algo:sharing-elimination} is capable of reaching actual optimality.
+If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of f-decreasing transformations is constructed by ``composition'' of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}, as illustrated in Algorithm~\ref{algo:sharing-elimination}.
+
+Finally, we observe that the $\sigma_{ei}$ sub-expressions can sometimes be considered ``weakly structured''.
+This happens when a relaxed version of Definition~\ref{def:struct-expr}, in which the factorization of $s_{\Lambda}$ only ``minimizes'' (rather than ``eliminates'') spatial sharing, applies (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
+Weak structure will be exploited by Algorithm~\ref{algo:sharing-elimination} in the attempt to achieve optimality.
 
@@
 Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
-
 The rewrite operators, whose (Python) implementation is based on manipulation of abstract syntax trees (ASTs), compose the COFFEE language.
 A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion.
-Sharing elimination and pre-evaluation are implemented by composing, in special ways, these operators.
 
 \subsection{Independence from form compilers}
-COFFEE is independent from the high level form compiler.
-COFFEE handles generic ASTs, so any form compiler can use it as long as such a representation is properly emitted (with expressions in normal form, or sufficiently close to it).
-In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs (instead of strings, as in FEniCS) is used.
-COFFEE itself provides an interface for building ASTs.
-In particular, COFFEE aims to decouple the mathematical manipulation of a form from code optimization.
-Another viewpoint is that form compiler developers should not worry about performance of the generated code.
+COFFEE aims to be independent of the high level form compiler.
+It provides an interface to build generic ASTs and only expects expressions to be in normal form (or sufficiently close to it).
+In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs, instead of strings, is used.
+Thus, COFFEE decouples the mathematical manipulation of a form from code optimization, or, in other words, relieves form compiler developers of caring about performance of generated code.
 
@@
 
-If, from one hand, it is true that form compilers usually provide optimization options to work around this issue, we must also consider the fact that not always their outcome is ideal for low level efficiency.
+Form compilers work around this issue in many different ways, which impacts low level efficiency.
 Consider, for example, a set of tabulated basis functions with quadrature points along rows and functions along columns.

### 2016-02-06 TOMS_Paper/assembly.tex
Integrate David's suggestions

@@
 \label{def:multi-linear-loop}
-A multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
+A multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are affine in each loop $L_i$ separately.
 \end{Def}
@@
 We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact (e.g., products have not been expanded yet, distinct monomials can still be identified, etc.).
-This bring us to formalize the class of loop nests for which we seek optimality.
+This bring us to formalize the class of loop nests that we aim to optimize.
 
@@
 \label{def:mln-optimality}
-Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a generic transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
+Let $\Lambda$ be a generic loop nest, and let $\Gamma$ be a transformation function $\Gamma : \Lambda \rightarrow \Lambda'$ such that $\Lambda'$ is semantically equivalent to $\Lambda$ (possibly, $\Lambda' = \Lambda$).
 We say that $\Lambda' = \Gamma (\Lambda)$ is an optimal synthesis of $\Lambda$ if the total number of operations (additions, products) performed to evaluate the result is minimal.
@@
 
-The concept of quasi-optimality, which relies on the particular class of \textit{f-decreasing} transformations, is also introduced.
-
-\begin{Def}[F-decreasing transformation]
-A sequence of transformations reducing the operation count is called f-decreasing (abbreviation for ``flop-decreasing'').
+The concept of quasi-optimality, which relies on the particular class of \textit{flop-decreasing} transformations, is also introduced.
+
+\begin{Def}[Flop-decreasing transformation]
+A sequence of transformations reducing the operation count is called flop-decreasing.
 \end{Def}
@@
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) in the innermost loops performed to evaluate the result is minimal and was achieved through f-decreasing transformations.
-\end{Def}
-
-The restriction to f-decreasing transformations aims to exclude those ``deceptive'' solutions that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing, in fact, a ``global'' increase in operation count.
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if:
+\begin{itemize}
+\item the number of operations (additions, products) in the innermost loops performed to evaluate the result is minimal, and
+\item $\Gamma$ is expressed as composition of flop-decreasing transformations.
+\end{itemize}
+\end{Def}
+
+The restriction to flop-decreasing transformations aims to exclude those ``deceptive'' solutions that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing, in fact, a ``global'' increase in operation count.
 
@@
 \item[Spatial sharing] There are at least two symbolically identical sub-expressions
-\item[Temporal sharing] There is at least one non-trivial sub-expression (e.g., an addition or a product) that, as independent of $\lbrace L_{i_1}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$, is redundantly executed.
+\item[Temporal sharing] There is at least one non-trivial sub-expression (e.g., an addition or a product) that is redundantly executed because it is independent of $\lbrace L_{i_1}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$.
 \end{description}
@@
 In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by the symbol $b_j$.
-Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}).
+Factorization eliminates spatial sharing and creates temporal sharing (Figure~\ref{code:multi_loopnest_b}).
 Finally, generalized code motion \cite{Luporini}, which seeks sub-expressions that are redundantly executed by at least one loop in the nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only checks invariance with respect to the innermost loop.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).
@@
 Often, the optimal implementation strategy for these operations is to be determined out of two alternatives.
-For instance, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector.
-The tensor operation will reduce to the scalar expression $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on $L_i$.
+For instance, consider $J^{-1} \nabla v \cdot J^{-1} \nabla v$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $v$ a generic two-dimensional vector.
+The tensor operation will reduce to the scalar expression $(a v^0_i + b v^1_i) (a v^0_i + b v^1_i) + ...$, in which $v^0_i$ and $v^1_i$ represent components of $v$ that depend on $L_i$.
 To minimize the operation count for expressions of this kind, we have two options.
@@
 \end{Strategy}
-In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed.
-On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$.
+In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $v^0_i v^0_i a a + v^0_i v^1_i (ab + ab) + v^1_i v^1_i c c + ...$ and some hoistable sub-expressions would be exposed.
+On the other hand, Strategy~\ref{strategy:i} would have no effect as $v$ only depends on a single loop, $L_i$.
 In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion.
@@
 If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing.
-If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of f-decreasing transformations is constructed by ``composition'' of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}, as illustrated in Algorithm~\ref{algo:sharing-elimination}.
+If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of flop-decreasing transformations is constructed by ``composition'' of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}, as illustrated in Algorithm~\ref{algo:sharing-elimination}.
 
@@
 \normalfont
-The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
+The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-6); scheduling of all other sub-expressions (step 7).
 \begin{enumerate}
@@
 \item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying Strategy~\ref{strategy:ii} would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
-\textit{Note: this ensures the f-decreasing nature of the transformations at step (6)}.\\
+\textit{Note: this ensures the flop-decreasing nature of the transformations at step (6)}.\\
 \textit{Note: this pass can be accomplished by visiting the expression tree and collecting several kinds of information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost will be derived.}
@@
 \normalfont
-Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also attempt to optimize the loop nest ``globally''.
+Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also attempt to optimize the loop nest ``globally''.
 We will further elaborate this aspect in Section~\ref{sec:proof}.
@@
 {\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
-\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}
+\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D.
+The operation counts are $E(f(z^0, z^1, ...) + IJK \cdot 18)$ (left) and $E(f(z^0, z^1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z^0, z^1, ...)$ representing the operation count for evaluating $z^0, z^1, ...$, and common sub-expressions being counted once.
+Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality, apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}
 
@@
 Sharing elimination uses three operators: expansion, factorization, and code motion.
-In this section, we discuss role and legality of a fourth operator: reduction pre-evaluation.
+In this section, we discuss the role and legality of a fourth operator: reduction pre-evaluation.
 We will see that what makes this operator special is the fact that there exists a single point in the transformation space of a monomial (i.e., a specific factorization of test, trial, and coefficient functions) ensuring its correctness.
@@
 
-Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space.
-This would be challenging if we were to deal with arbitrary loop nests and expressions.
-We make use of a result -- the foundation of tensor contraction mode -- to simplify our task.
-As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces.
+Pre-evaluation can be seen as the generalization of tensor contraction (Section~\ref{sec:tc}) to a wider class of sub-expressions.
+We know that multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces.
 A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors.
-In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after special transformation of the expression tree.
+In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after particular transformations of the expression tree.
 This leads to the following algorithm.
@@
 \label{sec:mem-const}
-We have just observed that code motion, as induced by monomial pre-evaluation, may excessively increase the working set size.
+We have just observed that the code motion induced by monomial pre-evaluation may excessively increase the working set size.
 Interestingly, even more aggressive code motion strategies are theoretically conceivable.
@@
 
-Since, for certain forms and discretizations, hoisting may bring the working set to exceed the size of some level of local memory (e.g.
+Since, for certain forms and discretizations, hoisting may cause the working set to exceed the size of some level of local memory (e.g.
 the last level of private cache on a conventional CPU, the shared memory on a GPU), we introduce the following \textit{memory constraints}.
@@
 
-Note that Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects.
+Constraint~\ref{const:Le} is a policy decision that the compiler should not silently consume memory on global data objects.
 It has the effect of shrinking the transformation space.
@@
 
-\subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
+\subsection{Limitation in the optimization space considered}
 \label{sec:completeness}
@@
 \item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
-\item \textit{Search for a global optimum.} Consider a form comprising two monomials $m_1$ and $m_2$.
+\item \textit{Optimizing over composite operations.} Consider a form comprising two monomials $m_1$ and $m_2$.
 Assume that pre-evaluation is profitable for $m_1$ but not for $m_2$, and that $m_1$ and $m_2$ share at least one term (e.g.
@@
 Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination.
-In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation to, respectively.
-Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$.
-We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}.
+In particular, we define $\theta(m) = \mathrm{\theta}^{pre}(m) - \mathrm{\theta}^{se}(m)$, where $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation to, respectively.
+Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) < 0$.
+We return to the issue of deriving $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ in Section~\ref{sec:op_count}.
 Having defined $\theta$, we can now describe the transformation algorithm (Algorithm~\ref{algo:gamma}).
@@
 \item Perform a depth-first visit of the expression tree and determine the set of monomials $M$.
-Let $S$ be the subset of monomials $m$ such that $\theta(m) < 0$.
+Let $S$ be the subset of monomials $m$ such that $\theta(m) > 0$.
 The set of monomials that will \textit{potentially} be pre-evaluated is $P = M \setminus S$.
@@
 \item Build the set $B$ of all possible bipartitions of $P$. Let $D$ be the dictionary that will store the operation counts of different alternatives.
-\item Discard $b = (b_S, b_P) \in B$ if the memory required after applying pre-evaluation to the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}); otherwise, add $D[b] = \theta_{se}(S \cup b_S) + \theta_{pre}(b_P)$.
+\item Discard $b = (b_S, b_P) \in B$ if the memory required after applying pre-evaluation to the monomials in $b_P$ exceeds $T_H$ (see Constraint~\ref{const:TH}); otherwise, add $D[b] = \mathrm{\theta}^{se}(S \cup b_S) + \mathrm{\theta}^{pre}(b_P)$.
 \\ \textit{Note: $\mathbb{B}$ is in practice very small, since even complex forms usually have only a few monomials.
-This pass can then be accomplished rapidly as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible.
+This pass can then be accomplished rapidly as long as the cost of calculating $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ is negligible.
 We elaborate on this aspect in Section~\ref{sec:op_count}.}
-\item Take $b$ such that $min(D[b])$.
+\item Take $b$ such that $min(b \in D)$.
 \item Apply pre-evaluation to all monomials in $b_P$.
 Apply sharing elimination to all resulting expressions.
-\\ \textit{Note: because of reuse of basis functions, pre-evaluation may produce some identical tables, which will be mapped to the same temporary variable.
+\\ \textit{Note: because of the reuse of basis functions, pre-evaluation may produce some identical tables, which will be mapped to the same temporary variable.
 Sharing elimination is therefore transparently applied to all expressions, including those resulting from pre-evaluation.}
@@
 
-We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and pre-evaluation.
+We recall that $\theta(m) = \mathrm{\theta}^{se}(m) - \mathrm{\theta}^{pre}(m)$, with $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ representing the operation counts after applying sharing elimination and pre-evaluation.
 Since $\theta$ is expected to be used by a compiler, requirements are simplicity and velocity.
@@
 
-The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ would consist of applying the actual transformations and simply count the number of operations.
-If, on one hand, this would be tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}.
-We then seek an analytic way of determining $\theta_{pre}$.
+The most trivial way of evaluating $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ would consist of applying the actual transformations and simply count the number of operations.
+This would be tolerable for $\mathrm{\theta}^{se}$, as Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost.
+However, the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}.
+We therefore seek an analytic way of determining $\mathrm{\theta}^{pre}$.
 
@@
 Intuitively, this means that we are introducing more operations than we are saving from pre-evaluating $L_i$.
-If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$.
+If $\iota < I$, we still need to find the number of terms $\rho$ such that $\mathrm{\theta}^{pre} = \rho \cdot \iota$.
 The mass matrix monomial in Figure~\ref{code:increase_factor} is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.
@@
 If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary.
-$\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''.
+$\Gamma$, therefore, may not be optimal, since $\theta$ only distinguishes monomials in ``isolation''.
 We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
 \item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols.
-If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum.
-This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum preserving Constraint~\ref{const:TH}.
+If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum.
+This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions for determining an optimum preserving Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}.
 Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
-Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
 \end{enumerate}
@@
 Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules.
-Non f-decreasing factorizations are excluded by construction at step (5).
+Non flop-decreasing factorizations are excluded by construction at step (5).
 We then prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality.

### 2016-02-06 TOMS_Paper/assembly.tex
Minor fixes in the code generation sect

@@
 Sharing elimination and pre-evaluation, as well as the transformation algorithm, have been implemented in COFFEE, the compiler for finite element integration routines adopted in Firedrake.
-In this section, we discuss a few aspects concerning implementation and features of the compiler.
-
-\subsection{Automation through the COFFEE language}
+In this section, we briefly discuss the aspects of the compiler that are relevant for this article.
+
+\subsection{Expressing transformations through the COFFEE language}
 COFFEE implements sharing elimination and pre-evaluation by composing ``building-block'' operators, which we refer to as ``rewrite operators''.
@@
 Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
-Secondly, generality: COFFEE can be seen as a tiny, low level computer algebra system, not specifically tied to finite element integration.
+Secondly, generality: COFFEE can be seen as a lightweight, low level computer algebra system, not necessarily tied to finite element integration.
 Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
@@
 In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs, instead of strings, is used.
-Thus, COFFEE decouples the mathematical manipulation of a form from code optimization, or, in other words, relieves form compiler developers of caring about performance of generated code.
+Thus, COFFEE decouples the mathematical manipulation of a form from code optimization; or, in other words, relieves form compiler developers of caring about performance of generated code.
 
@@
 \label{sec:zeros}
-Basis function tables may be block-sparse (e.g., containing zero-valued columns) depending on certain implementation choices at the level of the form compiler.
-For example, the FEniCS Form Compiler handles vector-valued function spaces by populating tabulated basis functions with blocks of zero-valued columns -- this makes code generation easier, but performance sub-optimal due to the execution of ``useless'' flops (e.g., by executing operations like \texttt{a + 0}).
-
-Form compilers work around this issue in many different ways, which impacts low level efficiency.
-Consider, for example, a set of tabulated basis functions with quadrature points along rows and functions along columns.
-\texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}.
+For several reasons, basis function tables may be block-sparse (e.g., containing zero-valued columns).
+For example, the FEniCS Form Compiler implements vector-valued functions by adding blocks of zero-valued columns to the corresponding tabulations; this extremely simplifies code generation (particularly, the construction of loop nests), but also affects the performance of the generated code due to the execution of ``useless'' flops (e.g., operations like \texttt{a + 0}).
 In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g.
 \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed.
-This technique, however, promotes non-contiguous memory loads and stores, which nullify the benefits of vectorization.
-
-COFFEE provides a mechanism to restructure loops that preserves the effectiveness of low level optimization, such as vectorization.
-By executing the code symbolically, loop bounds are adjusted and memory offsets are introduced to minimize ``useless'' flops.
-This is done performing several checks to preserve the semantics of the computation whilst trying not to affect low level optimization (e.g., padding and data alignment are affected by choice of the offsets -- this is, however, out of the scope of this article).
-
-%As discussed in the next section, all code generation systems used for performance evaluation handle block-sparse tables.
+This technique, however, promotes non-contiguous memory loads and stores, which nullify the potential benefits of vectorization.
+COFFEE, instead, handles block-sparse basis function tables by restructuring loops in a way that low level optimization (especially vectorization) is only marginally affected.
+This is based on symbolic execution of the code, which enables a series of checks on array indices and loop bounds for determining which zero-valued blocks can be skipped without affecting data alignment.
 

### 2016-02-10 TOMS_Paper/assembly.tex
Several changes, but before changing the rewrite rules proof

@@
 \usepackage{url}
+
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 \item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step (3).
-An edge $(S^i$, $S^j)$ indicates that a product $S^i S^j$ would appear if the sub-expressions including $S^i$ and $S_j$ were expanded.
+An edge $(S^i$, $S^j)$ indicates that a product $S^i S^j$ would appear if the sub-expressions including $S^i$ and $S^j$ were expanded.
 Let $\mathbb{T}$ be the subset of vertices with at most one incident edge, or ``terminals''.
@@
 \textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
-\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii} such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
-\textit{Note: this recasts the expression as $S (...S^1 + ... + ...S^n)$}.
-\end{enumerate}
+\item There exists a complete bipartite sub-graph $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$, with $\mathbb{S}_j$ and $\mathbb{S}_k$ being partitions of symbols depending on $L_j$ and $L_k$, respectively: take an arbitrary symbol in $\mathbb{S}_j^d \cup \mathbb{S}_k^d$ and apply Strategy~\ref{strategy:ii}.
+\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii}. \end{enumerate}
 Go back to step (2).
@@
 We elaborate on this aspect in Section~\ref{sec:op_count}.}
-\item Take $b$ such that $min(b \in D)$.
+\item Take $\argmin_b D[b]$.
 \item Apply pre-evaluation to all monomials in $b_P$.
@@
 Assume that the pre-evaluation of different monomials does not result in identical tables.
-Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and satisfies Constraint~\ref{const:TH}.
+Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality} and satisfies Constraint~\ref{const:TH}.
 \end{Prop}
@@
 We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
-\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols.
+\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials sharing some generic multilinear symbols.
 If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum.
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Proposition~\ref{prop:multi-struct} states that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (achieved through Strategy~\ref{strategy:ii}).
+Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (through Strategy~\ref{strategy:ii}).
 If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
-Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimal solution from the search space.
+Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimum from the search space.
 
@@
 Non flop-decreasing factorizations are excluded by construction at step (5).
-We then prove that applying in sequence rewrite rules (A), (B), and (C), with applications of rule (C) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality.
+We have to prove that applying in sequence rewrite rules (A), (B), (C), and (D), with applications of rules (C) and (D) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality.
 \begin{description}
 \item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
-\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
+\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
 Let us consider the alternative factorization $S^{nt}(...S + ...
 + ...V)$.
-If $V$ were a terminal, it is easy to see that setting $V = T$ and $S = S^{nt}$ would just expose another possible hit for rule (B).
+If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B).
 Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ...
 + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$.
-From this, we see that applying rule (C) prior to rule (B) may prune the optimal solution from the search space.
+From this, we see that applying rule (C) prior to rule (B) may prune the optimum from the search space.
 \item[Rule (C)] We interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
-For each disconnected component $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete.
-Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), and thanks to loop linearity, the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$.
-This is reachable by factorizing the vertex with highest degree, as any other choice may break factorization opportunities.
-In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$).
-This lower bound is achieved by repeating the sharing elimination algorithm (notice the jump back to step (2) at the end of step (6)).
+There can be multiple connected components, each of which we denote by $\mathbb{B}^d$.
+We identify with $lb^d$ the lower bound on the operation count, which we recall is an affine function of the number of products, induced by the symbols in $\mathbb{B}^d$.
+The factorization in any $\mathbb{B}^d$ is independent of that in other connected components.
+It therefore suffices to consider an arbitrary $\mathbb{B}^d$.
+We distinguish the two following possibilities.
+\begin{description}
+\item[$\mathbb{B^d}$ is complete] We simply have that $lb^d = 1$, which the algorithm reaches through iterative factorization (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$), by jumping back to step (2) at the end of step (6).
+\item[$\mathbb{B^d}$ is not complete] As a consequence of loop linearity and the absence of further hits for rules (A) and (B), $lb^d = {\min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. 
+By not choosing the vertex with highest degree (as the rule states),  ...
 \end{description}
-The algorithm is guaranteed to terminate because of the jump to step (6) as $\mathbb{S} = \mathbb{T}$. 
+\end{description}
+The algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
 \end{enumerate}

### 2016-02-13 TOMS_Paper/assembly.tex
before removing complete bipartite graphs

@@
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply Strategy~\ref{strategy:ii}.\\
+\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
 \textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
@@
 \textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
-\item There exists a complete bipartite sub-graph $\mathbb{B}^d = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$, with $\mathbb{S}_j$ and $\mathbb{S}_k$ being partitions of symbols depending on $L_j$ and $L_k$, respectively: take an arbitrary symbol in $\mathbb{S}_j^d \cup \mathbb{S}_k^d$ and apply Strategy~\ref{strategy:ii}.
-\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii}. \end{enumerate}
+\item There exists a complete bipartite sub-graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, with $\mathbb{S}_j$ and $\mathbb{S}_k$ being partitions of symbols depending on $L_j$ and $L_k$, respectively: apply Strategy~\ref{strategy:ii} to either the symbols in $\mathbb{S}_j$ or $\mathbb{S}_k$ (choice is arbitrary).
+\item $\mathbb{T} = \emptyset$, $S$ is in a cycle (i.e., a path starting and ending in $S$ with at least one edge and no repeated edges or vertices): take $S$ and apply Strategy~\ref{strategy:ii}.\\
+\textit{Note: the precondition basically discards factorizations that can be ``distributed within larger factorizations''. Further details will be provided in Section~\ref{sec:proof}}.
+\end{enumerate}
 Go back to step (2).
@@
 The operation counts are $E(f(z^0, z^1, ...) + IJK \cdot 18)$ (left) and $E(f(z^0, z^1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z^0, z^1, ...)$ representing the operation count for evaluating $z^0, z^1, ...$, and common sub-expressions being counted once.
-Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality, apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}
+Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}
 
@@
 Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination.
-In particular, we define $\theta(m) = \mathrm{\theta}^{pre}(m) - \mathrm{\theta}^{se}(m)$, where $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation to, respectively.
+In particular, we define $\theta(m) = \mathrm{\theta}^{pre}(m) - \mathrm{\theta}^{se}(m)$, where $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively.
 Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) < 0$.
@@
 \item Consider Algorithm~\ref{algo:sharing-elimination}.
-Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (through Strategy~\ref{strategy:ii}).
+Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (via Strategy~\ref{strategy:ii}).
 If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
@@
 Non flop-decreasing factorizations are excluded by construction at step (5).
-We have to prove that applying in sequence rewrite rules (A), (B), (C), and (D), with applications of rules (C) and (D) that may expose further opportunities for (A) and (B), represents a step forward towards quasi-optimality.
+We first discuss the individual impact of the rewrite rules; then, we explain why applying them in a certain order represents a step forward towards quasi-optimality.
+
 \begin{description}
@@
 + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$.
-From this, we see that applying rule (C) prior to rule (B) may prune the optimum from the search space.
-\item[Rule (C)] We interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively.
-There can be multiple connected components, each of which we denote by $\mathbb{B}^d$.
-We identify with $lb^d$ the lower bound on the operation count, which we recall is an affine function of the number of products, induced by the symbols in $\mathbb{B}^d$.
-The factorization in any $\mathbb{B}^d$ is independent of that in other connected components.
-It therefore suffices to consider an arbitrary $\mathbb{B}^d$.
-We distinguish the two following possibilities.
-\begin{description}
-\item[$\mathbb{B^d}$ is complete] We simply have that $lb^d = 1$, which the algorithm reaches through iterative factorization (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$), by jumping back to step (2) at the end of step (6).
-\item[$\mathbb{B^d}$ is not complete] As a consequence of loop linearity and the absence of further hits for rules (A) and (B), $lb^d = {\min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. 
-By not choosing the vertex with highest degree (as the rule states),  ...
+From this, we evince that applying rule (D) prior to rule (B) may prune the optimum from the search space.
+\item[Rule (C)] The fact that $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$ is complete implies that a sub-expression can be implemented by only one multiplication at every inner loop iteration.
+For example, the expression $S^0_j S^0_k + S^0_j S^1_k + S^1_j S^0_k + S^1_j S^1_k$ can first be recast (factorization) as $(S^0_j + S^1_j) (S^0_k + S^1_k)$ and finally (code motion) as $P_j P_k$.
+Rule (C) will apply Strategy~\ref{strategy:ii} ${\min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$ times, thus reaching the objective.
+\item[Rule (D)] The precondition implies that whatever factorization is chosen, no symbols will be eradicated from the expression.
+The rule then acts conservatively: the absence of a cycle through $S$ simply implies that $S$ will be hoisted through other applications of Strategy~\ref{strategy:ii}, so there is no need to perform a factorization.
 \end{description}
-\end{description}
-The algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
+
+Instances of (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid regardless of how the applications of Strategy~\ref{strategy:ii} are scheduled.
+(C), similarly, guarantees a single multiplication for a whole set of symbols.
+From this, we deduce that what Algorithm~\ref{algo:sharing-elimination} actually has to guarantee is a partial ordering: (A), (B), and (C) can be applied in arbitrary order, whereas (D), as explained above, must be considered only when none of the other rules can be hit.
+Rule (D) is clearly the most expensive from a computational point of view, because it must prune all choices that will not lead to the optimum.
+However, this rule becomes relevant only in extremely complex forms, in which extensive use of differential operators is made.
+
+Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
 \end{enumerate}
@@
 
-For a fair comparison, small patches (publicly available) were written for compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake.
-By executing all simulations in Firedrake, we guarantee that matrix insertion and mesh iteration have a fixed cost independent of the transformation system employed.
-The patches, in particular, adjust the data storage layout (e.g., by generating array of pointers instead of pointer to pointers and by replacing flattened arrays with bi-dimensional ones, as Firedrake expects).
+For a fair comparison, small patches (publicly available) were written for the compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake.
+By executing all simulations in Firedrake, we guarantee that both matrix insertion and mesh iteration have a fixed cost, independent of the transformation system employed.
+The patches adjust the data storage layout to what Firedrake expects (e.g., by generating an array of pointers instead of a pointer to pointers, by replacing flattened arrays with bi-dimensional ones).
 

### 2016-02-14 TOMS_Paper/assembly.tex
Before changing algo desc

@@
 \textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
-\item There exists a complete bipartite sub-graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, with $\mathbb{S}_j$ and $\mathbb{S}_k$ being partitions of symbols depending on $L_j$ and $L_k$, respectively: apply Strategy~\ref{strategy:ii} to either the symbols in $\mathbb{S}_j$ or $\mathbb{S}_k$ (choice is arbitrary).
-\item $\mathbb{T} = \emptyset$, $S$ is in a cycle (i.e., a path starting and ending in $S$ with at least one edge and no repeated edges or vertices): take $S$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: the precondition basically discards factorizations that can be ``distributed within larger factorizations''. Further details will be provided in Section~\ref{sec:proof}}.
+\item There is no cyclic path $[S, W, ..., S]$: take $S$ and apply Strategy~\ref{strategy:ii}, similarly to (B).\\
+\textit{Note: this recasts the expression as $S (...W + ...)$}.\\
+\textit{Note: the idea is that the product ${S}\ {W}$ cannot be ``hidden within larger factorizations''. Further details will be provided in Section~\ref{sec:proof}}.
+\item $[S, ..., V, ..., S]$ is a cycle, $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
+\textit{Note: this recasts the expression as $S (...S^1 + ...S^n)$}.
 \end{enumerate}
@@
 Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules.
-Non flop-decreasing factorizations are excluded by construction at step (5).
-We first discuss the individual impact of the rewrite rules; then, we explain why applying them in a certain order represents a step forward towards quasi-optimality.
-
+We here demonstrate why this is a step forwards towards the optimum.
+Note that, at this point, non flop-decreasing factorizations have already been pruned (step (5)).
+
+Instances of rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid regardless of how the applications of Strategy~\ref{strategy:ii} are scheduled.
+In particular:
 \begin{description}
@@
 + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$.
-From this, we evince that applying rule (D) prior to rule (B) may prune the optimum from the search space.
-\item[Rule (C)] The fact that $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$ is complete implies that a sub-expression can be implemented by only one multiplication at every inner loop iteration.
-For example, the expression $S^0_j S^0_k + S^0_j S^1_k + S^1_j S^0_k + S^1_j S^1_k$ can first be recast (factorization) as $(S^0_j + S^1_j) (S^0_k + S^1_k)$ and finally (code motion) as $P_j P_k$.
-Rule (C) will apply Strategy~\ref{strategy:ii} ${\min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$ times, thus reaching the objective.
-\item[Rule (D)] The precondition implies that whatever factorization is chosen, no symbols will be eradicated from the expression.
-The rule then acts conservatively: the absence of a cycle through $S$ simply implies that $S$ will be hoisted through other applications of Strategy~\ref{strategy:ii}, so there is no need to perform a factorization.
+From this, we evince that applying rules (C) and (D) prior to rule (B) may prune the optimum from the search space.
 \end{description}
 
-Instances of (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid regardless of how the applications of Strategy~\ref{strategy:ii} are scheduled.
-(C), similarly, guarantees a single multiplication for a whole set of symbols.
-From this, we deduce that what Algorithm~\ref{algo:sharing-elimination} actually has to guarantee is a partial ordering: (A), (B), and (C) can be applied in arbitrary order, whereas (D), as explained above, must be considered only when none of the other rules can be hit.
-Rule (D) is clearly the most expensive from a computational point of view, because it must prune all choices that will not lead to the optimum.
-However, this rule becomes relevant only in extremely complex forms, in which extensive use of differential operators is made.
+Rule (C) is the critical one.
+Let $\mathbb{A}$ be an acyclic sub-graph rooted in $S$, and let $W$ be the neighbor of $S$ in $\mathbb{A}$.
+Since $\mathbb{T} = \emptyset$, $S$ is also the root a second sub-graph $\mathbb{C}$.
+The non-trivial case in when $\mathbb{C}$ is cyclic.
+If any vertex in $\mathbb{C}$ were factorized prior to $S$, the product $S\ W$ could eventually appear in the transformed expression.
+Conversely, is Strategy~\ref{strategy:ii} were first applied to $S$, $W$ and one or more symbols in $\mathbb{C}$ would be grouped in the same sub-expression and hoisted.
 
@@
 \end{proof}
-
-%\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii)
-%\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply strategy (ii).\\
-%\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii) such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
 

### 2016-02-15 TOMS_Paper/assembly.tex
Before removing rewrite rules

@@
 \begin{itemize}
-\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
 \item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
@@
 \normalfont 
-The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-6); scheduling of all other sub-expressions (step 7).
+~
 \begin{enumerate}
-\item Perform a depth-first visit of the expression tree to group identical terms within the same sub-expression. \\ 
-\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
-
 \item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
 $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
-
-\item For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
+For all multilinear operands in $P \in \mathbb{P}$ such that $|P| < |S_P|$, apply Strategy~\ref{strategy:i}.
+\\
 \textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied.
@@
 
-\item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step (3).
-An edge $(S^i$, $S^j)$ indicates that a product $S^i S^j$ would appear if the sub-expressions including $S^i$ and $S^j$ were expanded.
-Let $\mathbb{T}$ be the subset of vertices with at most one incident edge, or ``terminals''.
-
-\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying Strategy~\ref{strategy:ii} would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
-\textit{Note: this ensures the flop-decreasing nature of the transformations at step (6)}.\\
-\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several kinds of information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost will be derived.}
-
-\item If $\mathbb{S} = \mathbb{T}$, jump to step (7). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
+\item Build the \textit{sharing graph} $G= (S, E)$.
+Each $s \in S$ represents a multilinear symbol or a temporary produced at step (2).
+An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.
+
+\item Let $T \subset S$ be the subset of vertices with at most one incident edges, or ``terminals''.
+Partition $T$ into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that the terminals in $T^i$ are either adjacent to the same vertex or disconnected.
+Apply Strategy~\ref{strategy:ii} factorizing the terminals in each $T^i$, provided that this leads to a reduction in operation count.\\
+\textit{Note: the last check ensures the flop-decreasing nature of the transformation; the cases in which the expansion cost exceeds the code motion gain are discarded.}
+
+\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression.
+Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.\\
+\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping the elements in $S^i$, which impact the expansion cost.}
+
+\item Restructure the expression based on the structure of $G$ and a set of rewrite rules $<$\textit{precondition}: \textit{action}$>$, to be applied in the order given below.
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
-\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
-\item There is no cyclic path $[S, W, ..., S]$: take $S$ and apply Strategy~\ref{strategy:ii}, similarly to (B).\\
-\textit{Note: this recasts the expression as $S (...W + ...)$}.\\
-\textit{Note: the idea is that the product ${S}\ {W}$ cannot be ``hidden within larger factorizations''. Further details will be provided in Section~\ref{sec:proof}}.
-\item $[S, ..., V, ..., S]$ is a cycle, $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: this recasts the expression as $S (...S^1 + ...S^n)$}.
-\end{enumerate}
-Go back to step (2).
+\item $\lbrace t^1, ..., t^n \rbrace \subset T$ adjacent to $s \in S$, $n > 1$: take $s$ and apply Strategy~\ref{strategy:ii}.\\
+\textit{Note: this recasts the expression as $s (...t^1 + ... + ...t^n)$}.
+\item Only one terminal $t$ adjacent to $s$: take an arbitrary non-terminal symbol $s^{nt}$ adjacent to $s$ and apply Strategy~\ref{strategy:ii}.\\
+\textit{Note: this recasts the expression as $s (...t + ...s^{nt})$}. 
+\item $T = \emptyset$, $C = [s, ..., v, ..., s]$ is a cycle, $v$ has highest degree in $C$, $\lbrace s^1, ..., s^n \rbrace$ adjacent to $v$: take $v$ and apply Strategy~\ref{strategy:ii}.\\
+\textit{Note: this recasts the expression as $v (...s^1 + ... + ...s^n)$}.
+\end{enumerate}
+Rule (C) may expose further hits for rules (A) and (B). Rules are therefore applied cyclically until no more hits are possible.\\
+\textit{Note: the last two steps only impact bilinear forms, since otherwise $E = \emptyset$.}
 
@@
 \normalfont
-Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also attempt to optimize the loop nest ``globally''.
+Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (3) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''.
 We will further elaborate this aspect in Section~\ref{sec:proof}.
@@
 We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$.
-For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step 3 has no effect.
-The sharing graph is $\mathbb{G} = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $\mathbb{T} = \lbrace c_i, d_i \rbrace$.
+For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step (1) has no effect.
+The sharing graph is $G = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $T = \lbrace c_i, d_i \rbrace$.
 Rewrite rule (4A) is hit, which leads to the code in Figure~\ref{code:multi_loopnest_c}.
@@
 We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
-In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step 3).
-We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{S} = \mathbb{T}$.
+In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
+We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule is applicable since $S = T$
 
@@
 If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration.
-Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimum from the search space.
-
-Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules.
+Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (1) cannot prune the optimum from the search space.
+
+After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, a set of rewrite rules are applied.
+We here discuss why applying, in rigorous order, rules (A), (B), and (C), leads to the optimum.
+
+Rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid no matter how Strategy~\ref{strategy:ii} is scheduled. In particular:
+\begin{description}
+\item[Rule (A)] The terminals $\lbrace t^1, ..., t^n \rbrace$ can only be grouped through the factorization of $s$ (by definition of terminal), so this rule has no side effects.
+\item[Rule (B)] Since $s^{nt}$ is not a terminal, there must exist at least one edge between $s^{nt}$ and a generic vertex $v$.
+Let us consider the alternative factorization $s^{nt}(...s + ...
++ ...v)$.
+If $v$ were a terminal, we observe that setting $v = t$ and $s = s^{nt}$ would simply expose another hit for rule (B).
+Otherwise ($v$ non terminal), we would have $s$ now appearing in two different sub-expressions ($(...s + ...
++ ...v)$ and $(...s\ t)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $s$.
+From this, we evince that applying (C) prior (B) may prune the optimum from the search space.
+\end{description}
+To reach the optimum, rule (C) discards all those potential factorizations that can be rather ``aggregated'' with \textit{essential} factorizations.
+A factorization is essential if a vertex is in a cycle.
+Being in a cycle is the fundamental condition to
+
+Having higher degree simply means grouping more symbols (i.e., avoiding more products);
+
+%Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules.
 We here demonstrate why this is a step forwards towards the optimum.
 Note that, at this point, non flop-decreasing factorizations have already been pruned (step (5)).
-
-Instances of rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid regardless of how the applications of Strategy~\ref{strategy:ii} are scheduled.
-In particular:
-\begin{description}
-\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
-\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
+%
+%Instances of 
+%\begin{description}
+%\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
+%\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
 Let us consider the alternative factorization $S^{nt}(...S + ...
@@
 From this, we evince that applying rules (C) and (D) prior to rule (B) may prune the optimum from the search space.
-\end{description}
-
-Rule (C) is the critical one.
+%\end{description}
+%
+%Rule (C) is the critical one.
 Let $\mathbb{A}$ be an acyclic sub-graph rooted in $S$, and let $W$ be the neighbor of $S$ in $\mathbb{A}$.
@@
 Conversely, is Strategy~\ref{strategy:ii} were first applied to $S$, $W$ and one or more symbols in $\mathbb{C}$ would be grouped in the same sub-expression and hoisted.
-
-Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
+% 
+%Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
 \end{enumerate}

### 2016-02-15 TOMS_Paper/assembly.tex
Rewrote algo talking about cycles etc

@@
 \usepackage{xfrac}
+\usepackage{amssymb}
 
@@
 
-\item Restructure the expression based on the structure of $G$ and a set of rewrite rules $<$\textit{precondition}: \textit{action}$>$, to be applied in the order given below.
+\item Use rewrite rules to restructure the expression.
+Let $V = \lbrace v^1, ..., v^n \rbrace$ be the vertices adjacent to $s \in S$; the preconditions for the three rewrite rules (A), (B), and (C) are:
 \begin{enumerate}[label=(\Alph*)]
-\item $\lbrace t^1, ..., t^n \rbrace \subset T$ adjacent to $s \in S$, $n > 1$: take $s$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: this recasts the expression as $s (...t^1 + ... + ...t^n)$}.
-\item Only one terminal $t$ adjacent to $s$: take an arbitrary non-terminal symbol $s^{nt}$ adjacent to $s$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: this recasts the expression as $s (...t + ...s^{nt})$}. 
-\item $T = \emptyset$, $C = [s, ..., v, ..., s]$ is a cycle, $v$ has highest degree in $C$, $\lbrace s^1, ..., s^n \rbrace$ adjacent to $v$: take $v$ and apply Strategy~\ref{strategy:ii}.\\
-\textit{Note: this recasts the expression as $v (...s^1 + ... + ...s^n)$}.
-\end{enumerate}
-Rule (C) may expose further hits for rules (A) and (B). Rules are therefore applied cyclically until no more hits are possible.\\
-\textit{Note: the last two steps only impact bilinear forms, since otherwise $E = \emptyset$.}
+\item $A \subsetneq V$ is a set of vertices not standing on cyclic paths, $A \neq \emptyset$
+\item $s$ is in a cycle $C$ and has highest degree among all vertices in $C$
+\item $\lbrace v^1, ..., v^{n-1} \rbrace$ are leaves of a tree, $p = v^n$ is the parent of $s$ (unless $s$ is the root)
+\end{enumerate}
+If a precondition is met, we take $s$ and apply Strategy~\ref{strategy:ii} (i.e., we factorize $s$ and perform code motion), and we remove all edges incident to $s$.
+The preconditions are checked cyclically (first all (A)s, then all (B)s, with (B) that may expose further hits for (A), and finally all (C)s) until no more hits are possible (i.e., until $S = T$).
+\\
+\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}
 
@@
 In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
-We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule is applicable since $S = T$
+We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule (step (5)) is applicable since $S = T$
 
@@
 After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, a set of rewrite rules are applied.
-We here discuss why applying, in rigorous order, rules (A), (B), and (C), leads to the optimum.
-
-Rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid no matter how Strategy~\ref{strategy:ii} is scheduled. In particular:
-\begin{description}
-\item[Rule (A)] The terminals $\lbrace t^1, ..., t^n \rbrace$ can only be grouped through the factorization of $s$ (by definition of terminal), so this rule has no side effects.
-\item[Rule (B)] Since $s^{nt}$ is not a terminal, there must exist at least one edge between $s^{nt}$ and a generic vertex $v$.
-Let us consider the alternative factorization $s^{nt}(...s + ...
-+ ...v)$.
-If $v$ were a terminal, we observe that setting $v = t$ and $s = s^{nt}$ would simply expose another hit for rule (B).
-Otherwise ($v$ non terminal), we would have $s$ now appearing in two different sub-expressions ($(...s + ...
-+ ...v)$ and $(...s\ t)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $s$.
-From this, we evince that applying (C) prior (B) may prune the optimum from the search space.
-\end{description}
-To reach the optimum, rule (C) discards all those potential factorizations that can be rather ``aggregated'' with \textit{essential} factorizations.
-A factorization is essential if a vertex is in a cycle.
-Being in a cycle is the fundamental condition to
-
-Having higher degree simply means grouping more symbols (i.e., avoiding more products);
-
-%Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules.
-We here demonstrate why this is a step forwards towards the optimum.
-Note that, at this point, non flop-decreasing factorizations have already been pruned (step (5)).
-%
-%Instances of 
-%\begin{description}
-%\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
-%\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$.
-Let us consider the alternative factorization $S^{nt}(...S + ...
-+ ...V)$.
-If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B).
-Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ...
-+ ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$.
-From this, we evince that applying rules (C) and (D) prior to rule (B) may prune the optimum from the search space.
-%\end{description}
-%
-%Rule (C) is the critical one.
-Let $\mathbb{A}$ be an acyclic sub-graph rooted in $S$, and let $W$ be the neighbor of $S$ in $\mathbb{A}$.
-Since $\mathbb{T} = \emptyset$, $S$ is also the root a second sub-graph $\mathbb{C}$.
-The non-trivial case in when $\mathbb{C}$ is cyclic.
-If any vertex in $\mathbb{C}$ were factorized prior to $S$, the product $S\ W$ could eventually appear in the transformed expression.
-Conversely, is Strategy~\ref{strategy:ii} were first applied to $S$, $W$ and one or more symbols in $\mathbb{C}$ would be grouped in the same sub-expression and hoisted.
-% 
-%Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
+We here discuss why applying in order rules (A), (B) and (C) leads to the optimum.
+
+Rule (A) is the critical one because it coordinates factorization at the junction between a cyclic and an acyclic components.
+It exploits the observation that a vertex outside of cyclic paths may be ``aggregated'' within one or more ``potentially unavoidable'' factorizations, thus saving on the final number of multiplications.
+This is quite evident for terminals (for which there is at most one way of being factorized), although not so obvious for vertices of arbitrary degree.
+Let $a$ be a vertex outside of cyclic paths, and let $V = \lbrace v^1, ..., v^n \rbrace$ be the set of vertices adjacent to $a$.
+Consider the non-trivial case in which at least one vertex $s \in V$ is within a cycle.
+If $a$ and $s$ were factorized separately, we could end up with the two sub-expressions $a(...v^1 + ...
+s + ...v^n)$ and $s(...)$.
+The spatial sharing induced by $s$ would make the synthesis sub-optimal.
+Conversely, if $s$ is factorized such that a single sub-expression $s(...a + ...)$ will be produced, the operation count is at worst unchanged.
+
+Rule (B) acts within cycles by iteratively picking the vertex with highest degree.
+Since the assumption is that there are no more hits for rule (A), the problem here is as simple as selecting the minimum number of vertices whose incident edges would cover the whole cycle.
+With a very similar concept, rule (C) schedules the remaining (disconnected) sub-trees.
 \end{enumerate}
@@
 
-The rest of the section is organized as follows: we first provide insights into the overall outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we explain, in detail and individually for each form, the performance results obtained.
+The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained.
 
 \paragraph{High level view}
-The main observation is that our transformation strategy does not always guarantee minimum execution time.
+Our transformation strategy does not always guarantee minimum execution time.
 In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime.
@@
 There are two reasons for this.
-Firstly, low level optimization can have a significant impact on actual performance.
+Firstly, low level optimization can have a significant impact on the actual performance.
 For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization.
-Secondly, memory constraints are critical, particularly the transformation strategy adopted when exceeding $T_H$.
-We will later thoroughly elaborate on all these aspects.
+Secondly, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.
 
 \paragraph{Autovectorization}
-The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length.
-This, combined with the chosen compilation flags, promotes autovectorization in the majority of code variants.
-An exception is \texttt{quad} due to the presence of indirection arrays in the generated code.
-In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; manual inspection of the compiled code suggests, however, that block vectorization (\cite{SLP-vect}) is often triggered.
-In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have similar structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could evince from compiler reports and manual inspection of assembly code.
+We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length.
+This ensured autovectorization in the majority of code variants\footnote{The compilation flag \texttt{xHost} helped the Intel compiler generating efficient vector code.
+We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}.
+The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code.
+In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization (\cite{SLP-vect}) is often triggered.
+In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.
 
@@
 We have developed transformation algorithms and a theory of optimality for finite element integration.
-The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made available to an increasingly significant user base.
-Performance results show the effectiveness of our methodology.
+The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made publicly available.
+Performance results suggest the effectiveness of our methodology.
 There are still several open problems.
-The main one is an elaborate classification of forms for which optimality can be reached.
-Secondly, we would like to extend our methodology to classes of loops arising in spectral methods, in which interaction with low level optimization becomes even more critical due to the larger working sets.
+Among them, a classification of forms for which optimality can be reached and a generalization of the methodology to other loop nests, for instance those arising in spectral methods.
 

### 2016-02-16 TOMS_Paper/assembly.tex
Swith to ILP formulation (TBF)

@@
 
-\item Use rewrite rules to restructure the expression.
-Let $V = \lbrace v^1, ..., v^n \rbrace$ be the vertices adjacent to $s \in S$; the preconditions for the three rewrite rules (A), (B), and (C) are:
-\begin{enumerate}[label=(\Alph*)]
-\item $A \subsetneq V$ is a set of vertices not standing on cyclic paths, $A \neq \emptyset$
-\item $s$ is in a cycle $C$ and has highest degree among all vertices in $C$
-\item $\lbrace v^1, ..., v^{n-1} \rbrace$ are leaves of a tree, $p = v^n$ is the parent of $s$ (unless $s$ is the root)
-\end{enumerate}
-If a precondition is met, we take $s$ and apply Strategy~\ref{strategy:ii} (i.e., we factorize $s$ and perform code motion), and we remove all edges incident to $s$.
-The preconditions are checked cyclically (first all (A)s, then all (B)s, with (B) that may expose further hits for (A), and finally all (C)s) until no more hits are possible (i.e., until $S = T$).
-\\
-\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}
+\item Map $G$ to an Integer Linear Programming (ILP) problem for determining how to optimally apply Strategy~\ref{strategy:ii}.
+The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}.
+Let $|S| = n$; the ILP model then is as follows:
+\begin{gather*}
+x_i \text{: a vertex in } S \text{ (1 if a symbol must be factorized, 0 otherwise)}\\
+y_{ij} \text{: a direct edge in } E \text{ (1 if the product } s^i s^j \text{ is factorized through } s^i \text{, 0 otherwise)}\\
+n_i \text{: the number of edges incident to } x_i \\
+\begin{align*}
+\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j=1}^{n} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
+&\sum_{i=1}^{n} y_{ij} = 1,\ j = 1, ..., n
+\end{align*}
+\phantom{\hspace{6cm}}
+\end{gather*}
 
@@
 The sharing graph is $G = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $T = \lbrace c_i, d_i \rbrace$.
-Rewrite rule (4A) is hit, which leads to the code in Figure~\ref{code:multi_loopnest_c}.
+The ILP formulation leads to the code in Figure~\ref{code:multi_loopnest_c}.
 
@@
 In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
-We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule (step (5)) is applicable since $S = T$
+We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, and since $S = T$ the ILP formulation becomes irrelevant because there are no factorization opportunities.
 
@@
 
-After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, a set of rewrite rules are applied.
-We here discuss why applying in order rules (A), (B) and (C) leads to the optimum.
-
-Rule (A) is the critical one because it coordinates factorization at the junction between a cyclic and an acyclic components.
-It exploits the observation that a vertex outside of cyclic paths may be ``aggregated'' within one or more ``potentially unavoidable'' factorizations, thus saving on the final number of multiplications.
-This is quite evident for terminals (for which there is at most one way of being factorized), although not so obvious for vertices of arbitrary degree.
-Let $a$ be a vertex outside of cyclic paths, and let $V = \lbrace v^1, ..., v^n \rbrace$ be the set of vertices adjacent to $a$.
-Consider the non-trivial case in which at least one vertex $s \in V$ is within a cycle.
-If $a$ and $s$ were factorized separately, we could end up with the two sub-expressions $a(...v^1 + ...
-s + ...v^n)$ and $s(...)$.
-The spatial sharing induced by $s$ would make the synthesis sub-optimal.
-Conversely, if $s$ is factorized such that a single sub-expression $s(...a + ...)$ will be produced, the operation count is at worst unchanged.
-
-Rule (B) acts within cycles by iteratively picking the vertex with highest degree.
-Since the assumption is that there are no more hits for rule (A), the problem here is as simple as selecting the minimum number of vertices whose incident edges would cover the whole cycle.
-With a very similar concept, rule (C) schedules the remaining (disconnected) sub-trees.
+After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, the ILP model is instantiated.
+At this point, the optimality of the algorithm reduces to prove the correctness of the model.
+The model aims ...
 \end{enumerate}

### 2016-02-17 TOMS_Paper/assembly.tex
fix ILP formulation
Added ILP formulation. Now going to change quasi- with locally optimal

@@
 \begin{gather*}
-x_i \text{: a vertex in } S \text{ (1 if a symbol must be factorized, 0 otherwise)}\\
-y_{ij} \text{: a direct edge in } E \text{ (1 if the product } s^i s^j \text{ is factorized through } s^i \text{, 0 otherwise)}\\
+x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
+y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
 n_i \text{: the number of edges incident to } x_i \\
 \begin{align*}
-\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j=1}^{n} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
-&\sum_{i=1}^{n} y_{ij} = 1,\ j = 1, ..., n
+\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
+& y_{ij} + y_{ji} = 1,\ (i, j) \in E
 \end{align*}
@@
 In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
-We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, and since $S = T$ the ILP formulation becomes irrelevant because there are no factorization opportunities.
+We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$.
+Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.
 
@@
 
-A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
+For reasons of space, further examples, including the hyperelastic model evaluated in Section~\ref{sec:perf-results} and other non-trivial ILP instances, are made available online\footnote{Sharing elimination examples: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
 
@@
 After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, the ILP model is instantiated.
-At this point, the optimality of the algorithm reduces to prove the correctness of the model.
-The model aims ...
+At this point, the optimality of the algorithm reduces to prove the correctness of the model, which is relatively straightforward because of its simplicity.
+The model aims to minimize the operation count by individuating the most promising factorizations.
+The second set of constraints is to select all edges (i.e., all multiplications), exactly once.
+The first set of inequalities allows scheduling multiplications: once a vertex $s$ is selected (i.e., once a symbol is decided to be factorized), all multiplications involving $s$ can be grouped.
 \end{enumerate}

### 2016-02-17 TOMS_Paper/assembly.tex
Now use local optimality

@@
 We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms.
-Optimality and quasi-optimality are defined in terms of floating point operations given a memory bound.
-We provide an approach for exploring the space of legal transformations, which systematically reaches quasi-optimality.
+Global and local optimality are defined in terms of floating point operations given a memory bound.
+We provide an approach for exploring the space of legal transformations, which systematically reaches a local optimum.
 The conditions under which optimality is achieved are also discussed.
@@
 However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``quasi-optimality'', which relates operation count with inner loops).
+In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``local optimality'', which relates operation count with inner loops).
 Summarizing, our contributions are as follows:
@@
 The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
-\item We demonstrate that our algorithm systematically leads to quasi-optimality. We also explain under what conditions of the input problem optimality is achieved.
+\item We demonstrate that our algorithm systematically leads to a local optimum. We also explain under what conditions of the input problem global optimality is achieved.
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 \label{sec:optimal-impl}
-In this section, we characterize optimality and quasi-optimality for finite element integration as well as the space of legal transformations that needs be explored to achieve them.
+In this section, we characterize global and local optimality for finite element integration as well as the space of legal transformations that needs be explored to achieve them.
 How the exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.
@@
 
-We will show that multilinear loop nests, which arise naturally when translating bilinear or linear forms into code, are important because they have a structure that we can take advantage of to synthesize quasi-optimal code.
+We will show that multilinear loop nests, which arise naturally when translating bilinear or linear forms into code, are important because they have a structure that we can take advantage of to reach a local optimum.
 
@@
 
-The concept of quasi-optimality, which relies on the particular class of \textit{flop-decreasing} transformations, is also introduced.
+The concept of local optimality, which relies on the particular class of \textit{flop-decreasing} transformations, is also introduced.
 
@@
 
-\begin{Def}[Quasi-optimality of a loop nest]
+\begin{Def}[Local optimality of a loop nest]
 \label{def:mln-quasi-optimality}
-Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if:
+Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a locally optimal synthesis of $\Lambda$ if:
 \begin{itemize}
@@
 On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
-In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least quasi-optimality in all cases; (ii) optimality whenever the monomials are ``sufficiently structured''.
+In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) optimality whenever the monomials are ``sufficiently structured''.
 To this purpose, we will construct:
@@
 If the objective were reaching optimality and the expressions lacked structure, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
-Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
+Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee local optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
 For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold.
-Setting quasi-optimality, instead of optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
+Setting local optimality, instead of optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
 
@@
 \label{sec:optimal-synthesis}
-In this section, we build a transformation algorithm that, given a memory bound, systematically produces quasi-optimal finite element integration loop nests.
+In this section, we build a transformation algorithm that, given a memory bound, systematically reaches a local optimum for finite element integration loop nests.
 
@@
 \label{sec:proof}
-We demonstrate that the orchestration of sharing elimination and pre-evaluation as performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}).
+We demonstrate that the orchestration of sharing elimination and pre-evaluation as performed by the transformation algorithm guarantees local optimality (Definition~\ref{def:mln-quasi-optimality}).
 The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.
@@
 Assume that the pre-evaluation of different monomials does not result in identical tables.
-Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality} and satisfies Constraint~\ref{const:TH}.
+Then, $\Lambda' = \Gamma(\Lambda)$ is a local optimum in the sense of Definition~\ref{def:mln-quasi-optimality} and satisfies Constraint~\ref{const:TH}.
 \end{Prop}
@@
 \end{enumerate}
-\item \textit{Sharing elimination produces quasi-optimal (or optimal) loop nests.}\\
+\item \textit{Sharing elimination leads to a (at least) local optimum.}\\
 \end{enumerate}

### 2016-02-17 TOMS_Paper/assembly.tex
Small fixes here and there...

@@
 We provide an approach for exploring the space of legal transformations, which systematically reaches a local optimum.
-The conditions under which optimality is achieved are also discussed.
+The conditions under which a global optimum is achieved are also discussed.
 A theoretical analysis and extensive experimentation, which shows significant performance improvements over several state-of-the-art code generation systems, validate the approach.
@@
 \begin{itemize}
-\item We formalize finite element integration loop nests and we build the space of legal transformations impacting their operation count.
+\item We formalize the class of finite element integration loop nests and we build the space of legal transformations impacting their operation count.
 \item We provide an algorithm to select points in the transformation space.
 The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
-\item We demonstrate that our algorithm systematically leads to a local optimum. We also explain under what conditions of the input problem global optimality is achieved.
+\item We demonstrate that our approach systematically leads to a local optimum. We also explain under what conditions of the input problem global optimality is achieved.
 \item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation.
@@
 On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
-In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) optimality whenever the monomials are ``sufficiently structured''.
+In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) global optimality whenever the monomials are ``sufficiently structured''.
 To this purpose, we will construct:
@@
 \item Build the \textit{sharing graph} $G= (S, E)$.
-Each $s \in S$ represents a multilinear symbol or a temporary produced at step (2).
+Each $s \in S$ represents a multilinear symbol or a temporary produced at step (1).
 An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.
@@
 \item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression.
-Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.\\
-\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping the elements in $S^i$, which impact the expansion cost.}
-
-\item Map $G$ to an Integer Linear Programming (ILP) problem for determining how to optimally apply Strategy~\ref{strategy:ii}.
+Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ will not cause an increase in operation count.\\
+\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping $[s^1, ..., s^m]$, which impact the expansion cost.}
+
+\item Map $G$ to an Integer Linear Programming (ILP) model for determining how to optimally apply Strategy~\ref{strategy:ii}.
 The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}.
@@
 \end{gather*}
+\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}
 
@@
 After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, the ILP model is instantiated.
-At this point, the optimality of the algorithm reduces to prove the correctness of the model, which is relatively straightforward because of its simplicity.
+At this point, proving optimality reduces to convince ourselves about the correctness of the model, which is relatively straightforward because of its simplicity.
 The model aims to minimize the operation count by individuating the most promising factorizations.
@@
 The axes represent $q$, $nf$, and code transformation system.
-We show one subplot for each problem instance $\langle form, nf, q\rangle$, with the code transformation system varying within each subplot.
+We show one subplot for each problem instance ${\langle} form, nf, q {\rangle}$, with the code transformation system varying within each subplot.
 The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
@@
 Our transformation strategy does not always guarantee minimum execution time.
-In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime.
-The most significant of such test cases is the elastic model with $[q=4, nf=0]$.
+In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime.
+The most significant of such test cases is the elastic model with $[q=4,\ nf=0]$.
 There are two reasons for this.
@@
 
-\paragraph{Mass}
+\paragraph{Mass matrix}
 We start with the simplest of the bilinear forms investigated, the mass matrix.
@@
 On the other hand, \texttt{cfO2} always makes the optimal decision about whether applying pre-evaluation or not.
+Surprisingly, despite the simplicity of the form, the performance of the various code generation systems can differ significantly.
 
 \paragraph{Helmholtz}
-As happened with the mass matrix problem, when $q=1$ matrix insertion still hides the cost of local assembly.
-For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors.
-In particular, with
-\begin{itemize}
-\item $nf=0$, the adoption of pre-evaluation by \texttt{cfO2} results in increasingly notable speed-ups over \texttt{cfO1}, as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice.
-\item $nf=1$, \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$.
-This can indirectly be inferred from the large gap between \texttt{cfO1/cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it avoids pre-evaluation.
-\item $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
-\end{itemize}
-
-The sporadic slow-downs or only marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.
-
-An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$.
+As in the case of Mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular, with
+\begin{description}
+\item[$nf=0$] pre-evaluation makes \texttt{cfO2} notably faster than \texttt{cfO1}, especially for high values of $q$; \texttt{auto} correctly selects \texttt{tens}, which is comparable to \texttt{cfO2}.
+\item[$nf=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$.
+This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it rightfully avoids pre-evaluation because of the excessive expansion cost.
+\item[$nf=2$ and $nf=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
+\end{description}
+
+The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} are imputable to the presence of sharing in the generated code.
+
+An interesting experiment we additionally performed was relaxing the memory threshold by setting $T_H = L3_{size}$.
 We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle nf=2, q=2\rangle$.
-The effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is shared by the cores.
+This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.
 
@@
 The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should.
-To clarify this, consider the test case $\langle nf=0, q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
-\texttt{cfO2} finds pre-evaluation profitable -- that is, actually capable of reducing the operation count -- although it does not apply it because otherwise $T_H$ would be exceeded.
-However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that of \texttt{tens}.
-Our explanation is that despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this problem) that pre-evaluation would anyway be the optimal choice.
-This suggests that our model could be refined to handle the cases in which there is a significant gap between potential cache misses and save in flops.
+To clarify this, consider the test case $\langle nf=0,\ q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
+\texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$.
+However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}.
+The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice.
+This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.
 
@@
 \begin{itemize}
-\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to systematic sharing elimination and the use of symbolic execution to avoid iteration over the zero-valued regions in the basis function tables
-\item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
-\item \texttt{ufls} generally shows better runtime behaviour than \texttt{quad} and \texttt{tens}.
-This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
+\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to the perfect sharing elimination and the zero-valued blocks avoidance technique presented in Section~\ref{sec:zeros}.
+\item when $nf=1$, \texttt{auto} prefers \texttt{tens} over \texttt{quad}, which leads to sub-optimal operation counts and execution times.
+\item \texttt{ufls} often results in better execution times than \texttt{quad} and \texttt{tens}.
+This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, and a more effective code motion strategy.
 \end{itemize}
@@
 Firstly, the baseline is highly inefficient.
-With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) cannot exploit the structure inherent in the expressions.
+With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) fails at exploiting the structure inherent in the expressions.
 Secondly, the strategy for removing spatial and temporal sharing has a tremendous impact.

### 2016-02-26 TOMS_Paper/assembly.tex
Improves explonation of sharing elimination algo

@@
 $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
-For all multilinear operands in $P \in \mathbb{P}$ such that $|P| < |S_P|$, apply Strategy~\ref{strategy:i}.
+For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}.
 \\
-\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied.
-This is a consequence of Proposition~\ref{prop:multi-struct}.}
-
-\item Build the \textit{sharing graph} $G= (S, E)$.
-Each $s \in S$ represents a multilinear symbol or a temporary produced at step (1).
-An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.
-
-\item Let $T \subset S$ be the subset of vertices with at most one incident edges, or ``terminals''.
-Partition $T$ into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that the terminals in $T^i$ are either adjacent to the same vertex or disconnected.
-Apply Strategy~\ref{strategy:ii} factorizing the terminals in each $T^i$, provided that this leads to a reduction in operation count.\\
-\textit{Note: the last check ensures the flop-decreasing nature of the transformation; the cases in which the expansion cost exceeds the code motion gain are discarded.}
+\textit{Note: as a consequence of Proposition~\ref{prop:multi-struct}, $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied}
+
+\item For each sub-expression $expr$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1).
+Partition them into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that $T^i$ includes all instances of a given symbol in $expr$.
+Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T^i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
+\textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
+\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined in a tree visit.}
+
+\item Build the \textit{sharing graph} $G = (S, E)$.
+Each $s \in S$ represents a multilinear symbol or a temporary produced by the previous steps.
+An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.\\
+\textit{Note: the following steps will only impact bilinear forms, since otherwise $E = \emptyset$.}
 
 \item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression.
-Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ will not cause an increase in operation count.\\
-\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping $[s^1, ..., s^m]$, which impact the expansion cost.}
+Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.
 
@@
 \end{gather*}
-\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}
 
@@
 \normalfont
-Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (3) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''.
+Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''.
 We will further elaborate this aspect in Section~\ref{sec:proof}.
@@
 In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
-We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$.
+We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$.
 Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.

### 2016-02-26 TOMS_Paper/assembly.tex
Incorporate Paul's initial feedback

@@
 The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators.
-By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation approximated solution, are computed.
+By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation's approximated solution, are computed.
 The code for assembly operators must be carefully optimized: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.
@@
 
-However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
+However, in spite of such a considerable research effort, there is still no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
 In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``local optimality'', which relates operation count with inner loops).
@@
 \begin{itemize}
-\item We introduce an engine based on symbolic execution that allows skipping irrelevant floating point operations (e.g., those involving zero-valued quantities).
+\item We introduce a transformation based on symbolic execution that allows skipping irrelevant floating point operations (e.g., those involving zero-valued quantities).
 \end{itemize}
@@
 Local assembly is the subject of this article.
-It consists of computing the contributions of a single element in the discretized domain to the equation approximated solution.
+It consists of computing the contributions of a single element in the discretized domain to the equation's approximated solution.
 In global assembly, such local contributions are ``coupled'' by suitably inserting them into $A$ and $b$.
@@
 \item $i$ appears only as an array index, and
-\item whenever an array $a$ is indexed by $i$ ($a[i]$), all expressions in which this appears are affine in $a$.
+\item whenever an array $a$ is indexed by $i$ ($a[i]$), all expressions in which this appears are affine in $a[i]$.
 \end{enumerate}

### 2016-03-17 TOMS_Paper/assembly.tex
Change abstract and title (thanks to David/Paul)

@@
 % Page heads
-\markboth{F. Luporini et al.}{On Optimality of Finite Element Integration}
+\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration loops}
 
 % Title portion
-\title{On Optimality of Finite Element Integration}
+
+\title{An algorithm for the optimization of finite element integration loops}
 \author{Fabio Luporini
@@
 \begin{abstract}
-We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms.
-Global and local optimality are defined in terms of floating point operations given a memory bound.
-We provide an approach for exploring the space of legal transformations, which systematically reaches a local optimum.
-The conditions under which a global optimum is achieved are also discussed.
-A theoretical analysis and extensive experimentation, which shows significant performance improvements over several state-of-the-art code generation systems, validate the approach.
+We present an algorithm for the  optimization of a finite element integration loop nest.
+This algorithm is proven to achieve a locally optimal operation count.
+In specified circumstances the optimum achieved is global.
+Extensive numerical experiments demonstrate significant performance improvements over the state of the art in finite element code generation in almost all cases.
+This validates the algorithm presented here, and illustrate its limitations.
 \end{abstract}
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2015. On Optimality of Finite Element Integration.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2015. An algorithm for the optimization of finite element integration loops.}
 

### 2016-03-18 TOMS_Paper/assembly.tex
Integrate feedback from Paul and David

@@
 
-\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax
+\documentclass[prodmode,acmtoms]{acmsmall} % Aptara syntax
 
@@
 \begin{abstract}
-We present an algorithm for the  optimization of a finite element integration loop nest.
+We present an algorithm for the  optimization of a class of finite element integration loop nests.
 This algorithm is proven to achieve a locally optimal operation count.
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2015. An algorithm for the optimization of finite element integration loops.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration loops.}
 
@@
 \end{itemize}
+
+After reviewing basic concepts in finite element integration, we introduce in Section~\ref{sec:lnopt} a set of definitions mapping mathematical properties to the level of loop nests.
+This step is essential for understanding the two algorithms -- sharing elimination (Section~\ref{sec:sharing-elimination}) and pre-evaluation (Section~\ref{sec:pre-evaluation}) -- through which we construct the space of legal transformations.
+The main ``transformation algorithm'' in Section~\ref{sec:optimal-synthesis} delivers the local optimality claim by using a cost model to ``coordinate'' the application of sharing elimination and pre-evaluation.
+We elaborate on the correctness of the methodology in Section~\ref{sec:proof}.
+The numerical experiments are showed in Section~\ref{sec:perf-results}.
+We conclude discussing the limitations of the algorithms presented and future work.
 
@@
 Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary quite dramatically.
-The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms.
+The presence of derivatives or coefficient functions in the input form increases the rank of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms.
 On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''.
@@
 
-%TODO is "not expanded yet == still contracted" ?
-
 \begin{Def}[Finite element integration loop nest]
@@
 \begin{Def}[Flop-decreasing transformation]
-A sequence of transformations reducing the operation count is called flop-decreasing.
+A transformation which reduces the operation count is called flop-decreasing.
 \end{Def}
@@
 \begin{itemize}
-\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
+\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:mem-const})
 \item an algorithm to select points in the transformation space (Section~\ref{sec:optimal-synthesis})
@@
 If the objective were reaching optimality and the expressions lacked structure, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
-Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee local optimality\footnote{This requires coordination with other transformations, as discussed in the following sections.
-For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold.
-Setting local optimality, instead of optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
+Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee, after coordination with other transformations (an aspect which we discuss in the following sections), local optimality, with optimality being achieved if stronger preconditions hold.
+Setting local optimality, rather than optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
 
@@
 The tensor operation will reduce to the scalar expression $(a v^0_i + b v^1_i) (a v^0_i + b v^1_i) + ...$, in which $v^0_i$ and $v^1_i$ represent components of $v$ that depend on $L_i$.
-To minimize the operation count for expressions of this kind, we have two options.
+To minimize the operation count for expressions of this kind, we have two options:
 \begin{Strategy}
@@
 \begin{proof}
-This descends directly from Definition~\ref{def:linear-loop} and Definition~\ref{def:multi-linear-loop}, which essentially restrict the number of occurrences of a symbol $s_{\Lambda}$ in a summand to at most 1.
+This follows directly from Definition~\ref{def:linear-loop} and Definition~\ref{def:multi-linear-loop}, which essentially restrict the number of occurrences of a symbol $s_{\Lambda}$ in a summand to at most 1.
 \end{proof}
@@
 Finally, we observe that the $\sigma_{ei}$ sub-expressions can sometimes be considered ``weakly structured''.
-This happens when a relaxed version of Definition~\ref{def:struct-expr}, in which the factorization of $s_{\Lambda}$ only ``minimizes'' (rather than ``eliminates'') spatial sharing, applies (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
+This happens when a relaxed version of Definition~\ref{def:struct-expr} applies, in which the factorization of $s_{\Lambda}$ only ``minimizes'' (rather than ``eliminates'') spatial sharing (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).
 Weak structure will be exploited by Algorithm~\ref{algo:sharing-elimination} in the attempt to achieve optimality.
@@
 
+\noindent\rule[0.01ex]{\linewidth}{0.7pt}
+
 \begin{Algo}[Sharing elimination]
@@
 \normalfont 
-~
+The input of the algorithm is a tree representation a finite element integration loop nest.
 \begin{enumerate}
-\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$.
-$\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
+\item Perform a depth-first visit of the loop tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P_1, ..., P_p\rbrace$.
+$\mathbb{P}$ is such that all multilinear operands in each $P \in \mathbb{P}$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
 For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}.
@@
 
-\item For each sub-expression $expr$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1).
-Partition them into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that $T^i$ includes all instances of a given symbol in $expr$.
-Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T^i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
+\item For each sub-expression $\operatorname{expr}$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1).
+Partition them into disjoint sets, $\mathbb{T} = \lbrace T_1, ..., T_t\rbrace$, such that $T_i$ includes all instances of a given symbol in $\operatorname{expr}$.
+Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T_i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
 \textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
@@
 Each $s \in S$ represents a multilinear symbol or a temporary produced by the previous steps.
-An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.\\
+An edge $(s_i$, $s_j)$ indicates that a product $s_i s_j$ would appear if the sub-expressions including $s_i$ and $s_j$ were expanded.\\
 \textit{Note: the following steps will only impact bilinear forms, since otherwise $E = \emptyset$.}
 
-\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression.
-Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.
+\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S_1, ..., S_n\rbrace$, such that $S_i$ includes all instances of a given symbol $s$ in the expression.
+Transform $G$ by merging $\lbrace s_1, ..., s_m \rbrace \subset S_i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s_1, ..., s_m]$ would not cause an increase in operation count.
 
@@
 x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
-y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
+y_{ij} \text{: an edge in } E \text{ (1 if } s_i \text{ is factorized in the product } s_i s_j \text{ , 0 otherwise)}\\
 n_i \text{: the number of edges incident to } x_i \\
@@
 
-\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted expression, apply the most profitable between Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}.
-\\
+\item Perform a depth-first visit of the loop tree and, for each yet unhandled or hoisted expression, apply the most profitable between Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. \\
 \textit{Note: this pass speculatively assumes that expressions are (weakly) structured along the reduction loop.
@@
 
-\begin{Rem*}
-\normalfont
+\noindent\rule[1.0ex]{\linewidth}{0.7pt}
+
 Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''.
 We will further elaborate this aspect in Section~\ref{sec:proof}.
-\end{Rem*}
 
@@
 Consider again Figure~\ref{code:multi_loopnest_a}.
-We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$.
-For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step (1) has no effect.
+We have $\mathbb{P} = \lbrace P_0, P_1, P_2\rbrace$, with $P_0 = \lbrace b_j\rbrace$, $P_1 = \lbrace c_i\rbrace$, and $P_2 = \lbrace d_i\rbrace$.
+For all $P_i$, we have $|P_i| = 1 = |S_{P_i}|$, although applying Strategy~\ref{strategy:i} in step (1) has no effect.
 The sharing graph is $G = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $T = \lbrace c_i, d_i \rbrace$.
@@
 
-In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions.
-We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
-In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
-We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$.
+In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a very simple realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions.
+We observe that $\mathbb{P} = \lbrace P_0, P_1\rbrace$, with $P_0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$.
+In addition, $|P_i| = |S_{P_i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)).
+We then have (step (3)) $G = (\lbrace t_0, t_1, t_2, t_3 \rbrace, \lbrace (t_0, t_2), (t_1, t_3) \rbrace)$.
 Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.
@@
 \end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D.
-The operation counts are $E(f(z^0, z^1, ...) + IJK \cdot 18)$ (left) and $E(f(z^0, z^1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z^0, z^1, ...)$ representing the operation count for evaluating $z^0, z^1, ...$, and common sub-expressions being counted once.
-Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}
+The operation counts are $E(f(z_0, z_1, ...) + IJK \cdot 18)$ (left) and $E(f(z_0, z_1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z_0, z_1, ...)$ representing the operation count for evaluating $z_0, z_1, ...$, and common sub-expressions being counted once.
+The synthesis in Figure~\ref{code:poisson_b} is globally optimal apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}
 
@@
 We know that multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces.
-A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors.
+A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensor.
 In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after particular transformations of the expression tree.
 This leads to the following algorithm.
+
+\noindent\rule[0.01ex]{\linewidth}{0.7pt}
 
@@
 
+\noindent\rule[1.0ex]{\linewidth}{0.7pt}
+
 The pre-evaluation of a monomial introduces some critical issues:
@@
 Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms).
-For certain values of ${<}t, n, S{>}$, pre-evaluation will shift the loop nest from a CPU-bound to a memory-bound regime, which may be counter-productive for actual execution time.
+For certain values of ${<}t, n, S{>}$, pre-evaluation may dramatically increase the working set, which may be counter-productive for actual execution time.
 \item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count).
@@
 
-We expand on point 1) in the next section, while we address points 2) and 3) in Section~\ref{sec:optimal-synthesis}.
+We expand on point (1) in the next section, while we address points (2) and (3) in Section~\ref{sec:optimal-synthesis}.
 
@@
 Constraint~\ref{const:TH} has both theoretical and practical implications, which will be carefully analyzed in the next sections.
-
-\subsection{Limitation in the optimization space considered}
-\label{sec:completeness}
-We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization.
-Factorization addresses \textit{spatial redundancy}.
-The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.
-Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.
-We remark that sharing elimination and pre-evaluation aim to minimize spatial and temporal redundancy relying on \textit{a specific set of operators}.
-Theoretically, lower operation counts may be found by exploiting other domain-specific properties, such as redundancies in basis functions.
-Constraint~\ref{const:Le} also limits the applicability of code motion, which further reduced the transformation space to be analyzed.
 
@@
 Having defined $\theta$, we can now describe the transformation algorithm (Algorithm~\ref{algo:gamma}).
+
+\noindent\rule[0.01ex]{\linewidth}{0.7pt}
 
@@
 
+\noindent\rule[1.0ex]{\linewidth}{0.7pt}
+
 The output of the transformation algorithm is provided in Figure~\ref{code:loopnest-opt}, assuming as input the loop nest in Figure~\ref{code:loopnest}.
@@
 The compilation flags used were \texttt{-O3, -xHost}.
+The compilation flag \texttt{xHost} tells the Intel compiler to generate efficient code for the underlying platform.
 
@@
 
-The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time).
-They include the cost of both local assembly and matrix insertion, with the latter minimized through the choice of a mesh (details below) small enough to fit the L3 cache of the CPU.
+The values that we report are the average of three runs with ``warm cache''; that is, with all kernels retrieved directly from the Firedrake's cache, so code generation and compilation times are not counted.
+The timing includes however the cost of both local assembly and matrix insertion, with the latter minimized through the choice of a mesh (details below) small enough to fit the L3 cache of the CPU.
 
@@
 We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length.
-This ensured autovectorization in the majority of code variants\footnote{The compilation flag \texttt{xHost} helped the Intel compiler generating efficient vector code.
-We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}.
+This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}.
 The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code.
@@
 
-The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} are imputable to the presence of sharing in the generated code.
+The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.
 
@@
 However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}.
-The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice.
+The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the saving in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice.
 This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.
@@
 \item \texttt{ufls} often results in better execution times than \texttt{quad} and \texttt{tens}.
-This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, and a more effective code motion strategy.
+This is due to multiple factors, including avoidance of indirection arrays, preservation of loop structure, and a more effective code motion strategy.
 \end{itemize}
@@
 
+\section{Limitations and future work}
+\label{sec:completeness}
+We have defined sharing elimination and pre-evaluation as high level transformations on top of a specific set of rewrite operators, such as code motion and factorization, and we have used them to construct the transformation space.
+There are three main limitations in this process.
+Firstly, we do not have a systematic strategy to optimize sub-expressions independent of linear loops; despite we have a mechanism to determine ``how much computation'' should be hoisted at the level of the integration (reduction) loop, it is not clear how to effectively improve the heuristics used at step (6) in Algorithm~\ref{algo:sharing-elimination}.
+Secondly, lower operation counts may be found by exploiting domain-specific properties, such as redundancies in basis functions; this aspect is completely neglected in this article.
+Thirdly, with Constraint~\ref{const:Le} we have limited the applicability of code motion, which was essential given the complexity of the problem tackled.
+
+Another issue raised by the experimentation concerns selecting a proper threshold for Constraint~\ref{const:TH}.
+To solve this problem we probably need a sophisticated cost model, which is clearly out of the scope of this research.
+
+We also identify two possible research directions: a classification of forms that can attain optimality; a generalization of the methodology to other classes of loop nests, for instance those arising in spectral methods.
+
 \section{Conclusions}
 \label{sec:conclusions}
-We have developed transformation algorithms and a theory of optimality for finite element integration.
-The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made publicly available.
-Performance results suggest the effectiveness of our methodology.
-There are still several open problems.
-Among them, a classification of forms for which optimality can be reached and a generalization of the methodology to other loop nests, for instance those arising in spectral methods.
+We have developed a theory for the optimization of finite element integration loop nests.
+The article clearly explains what domain properties are exploited by our approach (e.g., linearity) and how these translate at the level of loop nests.
+All of the algorithms shown in this paper have been implemented in COFFEE, a compiler publicly available fully integrated with the Firedrake framework.
+The correctness of the transformation algorithm was discussed.
+The performance results achieved suggest the effectiveness of our methodology.
 

### 2016-03-21 TOMS_Paper/assembly.tex
fix grant numbers

@@
 
-This research is partly funded by the MAPDES project, by the
-Department of Computing at Imperial College London, by EPSRC through
-grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants
-NE/K008951/1 and NE/K006789/1, by the U.S.  National Science
-Foundation through grants 0811457 and 0926687, by the U.S. Army
-through contract W911NF-10-1-000, and by a HiPEAC collaboration
+This work was supported by by the
+Department of Computing at Imperial College London, the Engineering and
+Physical Sciences Research Council [grant number EP/L000407/1], and the Natural
+Environment Research Council [grant numbers 
+NE/K008951/1 and NE/K006789/1], and by a HiPEAC collaboration
 grant. The authors would like to thank Mr. Andrew T.T. McRae,
@@
 Author's addresses: Fabio Luporini $\&$ Paul H. J. Kelly, Department of Computing,
-Imperial College London; David A. Ham, Department of Computing and
-Department of Mathematics, Imperial College London; 
+Imperial College London; David A. Ham, Department of Mathematics, Imperial College London; 
 \end{bottomstuff}

### 2016-03-21 TOMS_Paper/assembly.tex
more bloody generated files

@@
 The need for rapid implementation of high performance, robust, and portable finite element methods has led to approaches based on automated code generation.
-This has been proven successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-paper}) projects.
-In these frameworks, the weak variational form of a problem is expressed at high level by means of a domain-specific language.
-The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators.
-By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation's approximated solution, are computed.
+This has been proven successful in the context of the FEniCS \cite{Fenics} and Firedrake \cite{firedrake-paper} projects.
+In these frameworks, the weak variational form of a problem is expressed in a high level mathematical syntax by means of the domain-specific language UFL \cite{UFL}.
+This mathematical specification is used by a domain-specific compiler, known as a form compiler, to generate low-level C or C++ code for the integration over a single element of the computational mesh of the variational problem's left and right hand side operators.
 The code for assembly operators must be carefully optimized: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.
@@
 This is a result of the complexity inherent in the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
-General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}).
-Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops.
+General-purpose compilers, such as those by \emph{GNU} and \emph{Intel}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}).
+Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO \cite{PLUTO}, focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops.
 The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems.
-In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''.
-In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count.
+\citeN{quadrature1} show how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''.
+\citeN{FFC-TC} and~\citeN{Francis} employ mathematical reformulations of finite element integration with the aim of minimizing the operation count.
 In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed.
-It is also worth mentioning an on-going effort to produce a new form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.
+It is also worth mentioning two new new form compilers, UFLACS \cite{Uflacs} and \cite{TSFC}, which particularly target the compile time challenges of the more complex variational forms.
 The performance evaluation in Section~\ref{sec:perf-results} includes most of these optimization systems.
@@
 In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``local optimality'', which relates operation count with inner loops).
-Summarizing, our contributions are as follows:
+In summary, our contributions are as follows:
 
@@
 
-In addition, in order to place COFFEE on the same level as other code generation systems from the viewpoint of low level optimization (which is essential for a fair performance comparison)
+In addition, in order to place COFFEE on the same level as other code generation systems from the viewpoint of low level optimization (which is essential for a fair performance comparison):
 
 \begin{itemize}
-\item We introduce a transformation based on symbolic execution that allows skipping irrelevant floating point operations (e.g., those involving zero-valued quantities).
+\item We introduce a transformation based on symbolic execution that allows irrelevant floating point operations to be skipped (for example those involving zero-valued quantities).
 \end{itemize}
 
-After reviewing basic concepts in finite element integration, we introduce in Section~\ref{sec:lnopt} a set of definitions mapping mathematical properties to the level of loop nests.
-This step is essential for understanding the two algorithms -- sharing elimination (Section~\ref{sec:sharing-elimination}) and pre-evaluation (Section~\ref{sec:pre-evaluation}) -- through which we construct the space of legal transformations.
-The main ``transformation algorithm'' in Section~\ref{sec:optimal-synthesis} delivers the local optimality claim by using a cost model to ``coordinate'' the application of sharing elimination and pre-evaluation.
+After reviewing basic concepts in finite element integration, in Section~\ref{sec:lnopt} we introduce a set of definitions mapping mathematical properties to the level of loop nests.
+This step is an essential precursor to the definition of the two algorithms -- sharing elimination (Section~\ref{sec:sharing-elimination}) and pre-evaluation (Section~\ref{sec:pre-evaluation}) -- through which we construct the space of legal transformations.
+The main transformation algorithm in Section~\ref{sec:optimal-synthesis} delivers the local optimality claim by using a cost model to coordinate the application of sharing elimination and pre-evaluation.
 We elaborate on the correctness of the methodology in Section~\ref{sec:proof}.
@@
 
-We consider the weak formulation of a linear variational problem
+Consider the weak formulation of a linear variational problem:
 \begin{equation}
@@
 where $a$ and $L$ are, respectively, a bilinear and a linear form.
-The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces.
+The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are suitable discrete function spaces.
 For simplicity, we assume $U = V$.
@@
 \end{equation}
-The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.
+The matrix $A$ and the vector $b$ are assembled and subsequently used to solve the linear system through (typically) an iterative method.
 
@@
 It consists of computing the contributions of a single element in the discretized domain to the equation's approximated solution.
-In global assembly, such local contributions are ``coupled'' by suitably inserting them into $A$ and $b$.
-
-We illustrate local assembly in a concrete example, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
+During global assembly, these local contributions are coupled by suitably inserting them into $A$ and $b$.
+
+We illustrate local assembly in a concrete example, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Poisson equation:
 \begin{equation}
@@
 The domain $\Omega$ of the equation is partitioned into a set of cells (elements) $T$ such that $\bigcup T = \Omega$ and $\bigcap T = \emptyset$.
-By defining $\lbrace \phi_i^K \rbrace$ as the set of local basis functions spanning $U$ on the element $K$, we can express the local element matrix as
+By defining $\lbrace \phi_i^K \rbrace$ as the set of basis functions with support on the element $K$ (i.e.
+those which do not vanish on this element), we can express the local element matrix as
 \begin{equation}
@@
 \label{sec:monomials}
-It has been shown (e.g., in~\citeN{Kirby:TC}) that local element tensors can be expressed as a sum of integrals over $K$, each integral being the product of derivatives of functions from sets of discrete spaces and, possibly, functions of some spatially varying coefficients.
+It has been shown (for example in~\citeN{Kirby:TC}) that local element tensors can be expressed as a sum of integrals over $K$, each integral being the product of derivatives of functions from sets of discrete spaces and, possibly, functions of some spatially varying coefficients.
 An integral of this form is called \textit{monomial}.
@@
 For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced.
-This implies that a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated.
-The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
+This implies that a change of variables from reference coordinates $X$ to real coordinates $x = F_K (X)$ is necessary any time a new element is evaluated.
+The basis functions $\{\phi_i^K\}$ are then replaced with local basis functions $\{\Phi_i\}$ such that $\Phi_i(X) = \phi_i^k(F_K(X)) = \phi_i^k(x) $.
+The numerical integration of \eqref{stiffness} over an element $K$ can then be expressed as follows:
 \begin{equation}
 \label{eq:quadrature}
-A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
+A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \Phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \Phi_i(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \Phi_j(X^q)}{\partial X_{\alpha_2}}  \det (F_K')W^q
 \end{equation}
-where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.
+where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $\det (F_K')$ the determinant of the Jacobian of the aforementioned change of coordinates.
 
@@
 \label{sec:tc}
-Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
+By exploiting the linearity, associativity and distributivity of the
+relevant mathematical operators, we can rewrite \eqref{eq:quadrature} as
 \begin{equation}
 \label{eq:tensor}
-A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
+A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n \left(\det (F_K') w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \left(\sum_{q=1}^{N} \Phi_{\alpha_3} \frac{\partial \Phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \Phi_{i_2}}{\partial X_{\alpha_2}}W^q\right)\right).
 \end{equation}
-A generalization of this transformation has been proposed in~\cite{Kirby:TC}.
-Because of only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables.
+A generalization of this transformation was introduced in~\cite{Kirby:TC}.
+Since it only involves reference element terms, the quadrature sum can be pre-evaluated and reused for each element.
 The evaluation of the local tensor can then be abstracted as
@@
 \end{equation}
-in which the pre-evaluated ``reference tensor'' $A_{i_1 i_2 \alpha}$ and the cell-dependent ``geometry tensor'' $G_{K}^\alpha$ are exposed.
+in which the pre-evaluated \emph{reference tensor}, $A_{i_1 i_2 \alpha}$, and the cell-dependent \emph{geometry tensor}, $G_{K}^\alpha$, are exposed.
 
@@
 Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary quite dramatically.
-The presence of derivatives or coefficient functions in the input form increases the rank of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms.
-On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''.
-The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.
-
-These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-TC}).
+The presence of derivatives or coefficient functions in the input form increases the rank of the geometry tensor, making the traditional quadrature mode preferable for sufficiently complex forms.
+On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains sufficiently small.
+The discretization, particularly the polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.
+
+These two modes are implemented in the FEniCS Form Compiler \cite{FFC-TC}.
 In this compiler, a heuristic is used to choose the most suitable mode for a given form.
-It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}).
-We will later comment on the efficacy of this approach (Section~\ref{sec:perf-results}).
-For the moment, we just recall that one of the goals of this research is to produce a system that goes beyond the dichotomy between quadrature and tensor modes.
+It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically \cite{Fenics}.
+We will return to the efficacy of this approach in section \ref{sec:perf-results}.
+One of the objectives of this paper is to produce a system that goes beyond the dichotomy between quadrature and tensor modes.
 We will reason in terms of loop nests, code motion, and code pre-evaluation, searching the entire implementation space for an optimal synthesis.
@@
 In this section, we characterize global and local optimality for finite element integration as well as the space of legal transformations that needs be explored to achieve them.
-How the exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.
+The method by which exploration is performed is discussed in Section~\ref{sec:optimal-synthesis}.
 
@@
 In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$.
-Expressions (right hand side of a statement) result from the translation of a form in high level matrix notation into code.
-In particular, $m$ is the number of ``terms'' (a monomial can be implemented by one or more terms), $\alpha_{eij}$ ($\beta_{eik}$) represents the product of a coefficient function (e.g., the inverse Jacobian matrix for the change of coordinates) with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
-We do not pose any restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex.
-We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact (e.g., products have not been expanded yet, distinct monomials can still be identified, etc.).
-This bring us to formalize the class of loop nests that we aim to optimize.
+Expressions (the right hand side of a statement) result from the translation of a form in high level matrix notation into code.
+In particular, $m$ is the number of monomials (a form is a sum of monomials), $\alpha_{eij}$ ($\beta_{eik}$) represents the product of a coefficient function (e.g., the inverse Jacobian matrix for the change of coordinates) with test  or trial functions, and $\sigma_{ei}$ is a function of coefficients and geometry.
+We do not pose any restrictions on function spaces (e.g., scalar- or vector-valued), coefficient expressions (linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex.
+We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact: products have not yet been expanded, distinct monomials can still be identified, and so on.
+This brings us to formalize the class of loop nests that we aim to optimize.
 

### 2016-03-21 TOMS_Paper/assembly.tex
more rewrites

@@
 
-The restriction to flop-decreasing transformations aims to exclude those ``deceptive'' solutions that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing, in fact, a ``global'' increase in operation count.
+The restriction to flop-decreasing transformations aims to exclude those apparent optimizations that, to achieve flop-optimal innermost loops, would rearrange the computation at the level of the outer loops causing, in fact, a global increase in operation count.
 
 We also observe that Definitions~\ref{def:mln-optimality} and~\ref{def:mln-quasi-optimality} do not take into account memory requirements.
-If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense.
-Henceforth we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated.
+If the execution of loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then optimizing the number of flops would be fruitless.
+Henceforth we assume we operate in a CPU-bound regime, evaluating arithmetic-intensive expressions.
 In the context of finite elements, this is often true for more complex multilinear forms and/or higher order elements.
@@
 Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured.
-On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$.
+However, multilinearity results in a certain degree of regularity in $\alpha_{eij}$ and $\beta_{eik}$.
 In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) global optimality whenever the monomials are ``sufficiently structured''.
@@
 Factorization eliminates spatial sharing and creates temporal sharing (Figure~\ref{code:multi_loopnest_b}).
-Finally, generalized code motion \cite{Luporini}, which seeks sub-expressions that are redundantly executed by at least one loop in the nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only checks invariance with respect to the innermost loop.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).
+Finally, generalized code motion \cite{Luporini}, which hoists sub-expressions that are redundantly executed by at least one loop in the nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only checks invariance with respect to the innermost loop.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).
 
@@
 If the objective were reaching optimality and the expressions lacked structure, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations.
-Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee, after coordination with other transformations (an aspect which we discuss in the following sections), local optimality, with optimality being achieved if stronger preconditions hold.
+Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee, after coordination with other transformations (an aspect which we discuss in the following sections), local optimality.
+Global optimality is achieved if stronger preconditions hold.
 Setting local optimality, rather than optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.
@@
 Often, the optimal implementation strategy for these operations is to be determined out of two alternatives.
-For instance, consider $J^{-1} \nabla v \cdot J^{-1} \nabla v$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $v$ a generic two-dimensional vector.
+For instance, consider $J^{-T} \nabla v \cdot J^{-T} \nabla v$, with $J^{-T}$ being the transposed inverse Jacobian matrix for the change of (two-dimensional) coordinates, and $v$ a generic two-dimensional vector.
 The tensor operation will reduce to the scalar expression $(a v^0_i + b v^1_i) (a v^0_i + b v^1_i) + ...$, in which $v^0_i$ and $v^1_i$ represent components of $v$ that depend on $L_i$.
@@
 On the other hand, Strategy~\ref{strategy:i} would have no effect as $v$ only depends on a single loop, $L_i$.
-In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion.
+In general, the choice between the two strategies depends on multiple factors: the loop sizes, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain due to code motion.
 A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}.
@@
 \label{def:struct-expr}
-We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if in order to eliminate the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
+We say that an expression is ``structured along a loop nest $\Lambda$'' if and only if,  for every symbol $s_{\Lambda}$ depending on at least one loop in $\Lambda$, the spatial sharing of $s_{\Lambda}$ may be eliminated by factorizing all occurrences of $s_{\Lambda}$ in the expression.
 \end{Def}

### 2016-03-21 TOMS_Paper/assembly.tex
more changes

@@
 
-Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''.
+Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest globally.
 We will further elaborate this aspect in Section~\ref{sec:proof}.
@@
 
-\subsection{Monomial pre-evaluation}
+\subsection{Pre-evaluation of reductions}
 \label{sec:pre-evaluation}
@@
 Each sub-expression is then factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
-This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py}}, as a consequence of the results in~\citeN{Kirby:TC}.
+This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py} and in \texttt{coffee/optimizer.py} in \citeN{FIXME-PUT-THE-COFFEE-DOI-CITE-HERE}.}, as a consequence of the results in~\citeN{Kirby:TC}.
 These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries.
@@
 \begin{enumerate}
-\item Depending on the complexity of a monomial, a certain number $t$ of temporary variables is required if pre-evaluation is performed.
+\item Depending on the complexity of a monomial, a certain number, $t$, of temporary variables is required if pre-evaluation is performed.
 Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms).
@@
 \label{sec:mem-const}
-We have just observed that the code motion induced by monomial pre-evaluation may excessively increase the working set size.
-Interestingly, even more aggressive code motion strategies are theoretically conceivable.
+We have just observed that the code motion induced by monomial pre-evaluation may dramatically increase the working set size.
+Even more aggressive code motion strategies are theoretically conceivable.
 Imagine $\Lambda$ is enclosed in a time stepping loop.
@@
 \begin{enumerate}
-\item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
+\item \textit{Coordination of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count in comparison with that achieved by sharing elimination.
 \item \textit{Optimizing over composite operations.} Consider a form comprising two monomials $m_1$ and $m_2$.
-Assume that pre-evaluation is profitable for $m_1$ but not for $m_2$, and that $m_1$ and $m_2$ share at least one term (e.g.
-some basis functions).
+Assume that pre-evaluation is profitable for $m_1$ but not for $m_2$, and that $m_1$ and $m_2$ share at least one term (for example some basis functions).
 If pre-evaluation were applied to $m_1$, sharing between $m_1$ and $m_2$ would be lost.
-We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
+We then need a mechanism to understand which transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
 \end{enumerate}
@@
 We recall that $\theta(m) = \mathrm{\theta}^{se}(m) - \mathrm{\theta}^{pre}(m)$, with $\mathrm{\theta}^{se}$ and $\mathrm{\theta}^{pre}$ representing the operation counts after applying sharing elimination and pre-evaluation.
-Since $\theta$ is expected to be used by a compiler, requirements are simplicity and velocity.
+Since $\theta$ is deployed in a working compiler, simplicity and efficiency are essential characteristics.
 In the following, we explain how to derive these two values.
@@
 This number captures the increase in arithmetic complexity due to the transformations exposing pre-evaluation opportunities.
-To contextualize, consider the example in Figure~\ref{code:increase_factor}.
-One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix.
-The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$).
+For context, consider the example in Figure~\ref{code:increase_factor}.
+One can think of this as the (simplified) loop nest originating from the integration of the action of a mass matrix.
+The sub-expression \texttt{$f_0$*$B_{i0}$+$f_1$*$B_{i1}$+$f_2$*$B_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$).
 In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $[L_i, L_j, L_k]$-dependent quantities (see Algorithm~\ref{algo:pre-evaluation}).
@@
 \label{sec:proof}
-We demonstrate that the orchestration of sharing elimination and pre-evaluation as performed by the transformation algorithm guarantees local optimality (Definition~\ref{def:mln-quasi-optimality}).
+We demonstrate that the orchestration of sharing elimination and pre-evaluation performed by the transformation algorithm guarantees local optimality (Definition~\ref{def:mln-quasi-optimality}).
 The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.
@@
 Let $\Gamma$ be the transformation algorithm.
-Let $X$ be the set of monomials that, according to $\Gamma$, need be pre-evaluated, and let $Y = M \setminus X$.
+Let $X$ be the set of monomials that, according to $\Gamma$, need to be pre-evaluated, and let $Y = M \setminus X$.
 Assume that the pre-evaluation of different monomials does not result in identical tables.
@@
 If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum.
-This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions for determining an optimum preserving Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}.
+This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions on order to determine an optimum which satisfies Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}.
 Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
@@
 
-Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} reaches optimality if stronger preconditions on the input variational form are satisfied.
-We here discuss such preconditions, in increasing order of complexity.
+Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} achieves a globally optimal flop count if stronger preconditions on the input variational form are satisfied.
+We state here these preconditions, in increasing order of complexity.
 \begin{enumerate}
@@
 Optimality is achieved if and only if all sub-expressions depending on coefficients are structured (see Section~\ref{sec:se-rln}).
-This would avoid ambiguity in factorization, which in turn would guarantee the output of step (7) in Algorithm~\ref{algo:sharing-elimination} be optimality.
+This avoids ambiguity in factorization, which in turn guarantees that the output of step (7) in Algorithm~\ref{algo:sharing-elimination} is optimal.
 \item There are multiple monomials, but either at most one coefficient (e.g., the coordinates field) or multiple coefficients not inducing sharing across different monomials are present.
@@
 \subsection{Expressing transformations through the COFFEE language}
-COFFEE implements sharing elimination and pre-evaluation by composing ``building-block'' operators, which we refer to as ``rewrite operators''.
+COFFEE implements sharing elimination and pre-evaluation by composing building block transformation operators, which we refer to as \emph{rewrite operators}.
 This has several advantages.
-Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
-Secondly, generality: COFFEE can be seen as a lightweight, low level computer algebra system, not necessarily tied to finite element integration.
-Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
+First, extendibility: novel transformations -- for instance, sum factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
+Second, generality: COFFEE can be seen as a lightweight, low level computer algebra system, not necessarily tied to finite element integration.
+Third, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
 The rewrite operators, whose (Python) implementation is based on manipulation of abstract syntax trees (ASTs), compose the COFFEE language.
@@
 
+{\color{red}FIXME: We need to put the exact versions of software in
+  here. for Firedrake this should be Zenodo DOIs. This is especially true
+  for the patched version of firedrake you used below}.
+
 The values that we report are the average of three runs with ``warm cache''; that is, with all kernels retrieved directly from the Firedrake's cache, so code generation and compilation times are not counted.
@@
 
-For Constraint~\ref{const:TH}, discussed in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the size of the processor L2 cache (the last level of private cache).
-When the threshold had an impact on the transformation process, the experiments were repeated with $T_H = L3_{size}$.
+For Constraint~\ref{const:TH}, discussed in Section~\ref{sec:mem-const}, we set $T_H = \operatorname{size}(\mathrm{L2})$; that is, the size of the processor L2 cache (the last level of private cache).
+When the threshold had an impact on the transformation process, the experiments were repeated with $T_H = \operatorname{size}(\mathrm{L3})$.
 The results are documented later, individually for each problem.
@@
 \begin{itemize}
-\item the polynomial order of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
-\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
+\item the polynomial degree of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
+\item the number of coefficient functions $\mathrm{nf} \in \lbrace0, 1, 2, 3\rbrace$
 \end{itemize}
@@
 
+{\color{red}FIXME: Please replace $nf$ with $\mathrm{nf}$ in all figure captions.}
+
 \begin{figure}
@@
 We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
-The axes represent $q$, $nf$, and code transformation system.
-We show one subplot for each problem instance ${\langle} form, nf, q {\rangle}$, with the code transformation system varying within each subplot.
+The axes represent $q$, $\mathrm{nf}$, and code transformation system.
+We show one subplot for each problem instance ${\langle} \textrm{form}, \mathrm{nf}, q {\rangle}$, with the code transformation system varying within each subplot.
 The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
-We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.
+We note that if a bar or a subplot are missing, then the form compiler failed at generating code because it either exceeded the system memory limit or was unable to handle the form.
 
@@
 In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime.
-The most significant of such test cases is the elastic model with $[q=4,\ nf=0]$.
+The most significant of such test cases is the elastic model with $[q=4,\ \mathrm{nf}=0]$.
 There are two reasons for this.
-Firstly, low level optimization can have a significant impact on the actual performance.
+First, low level optimization can have a significant impact on the actual performance.
 For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization.
-Secondly, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.
+Second, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.
 
@@
 The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code.
-In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization (\cite{SLP-vect}) is often triggered.
+In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered.
 In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.
@@
 As $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups.
-It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} would tend to be preferable.
-On the other hand, \texttt{cfO2} always makes the optimal decision about whether applying pre-evaluation or not.
+It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable.
+On the other hand, \texttt{cfO2} always makes the optimal decision about whether to apply pre-evaluation or not.
 Surprisingly, despite the simplicity of the form, the performance of the various code generation systems can differ significantly.
@@
 \begin{description}
-\item[$nf=0$] pre-evaluation makes \texttt{cfO2} notably faster than \texttt{cfO1}, especially for high values of $q$; \texttt{auto} correctly selects \texttt{tens}, which is comparable to \texttt{cfO2}.
-\item[$nf=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$.
+\item[$\mathrm{nf}=0$] pre-evaluation makes \texttt{cfO2} notably faster than \texttt{cfO1}, especially for high values of $q$; \texttt{auto} correctly selects \texttt{tens}, which is comparable to \texttt{cfO2}.
+\item[$\mathrm{nf}=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$.
 This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it rightfully avoids pre-evaluation because of the excessive expansion cost.
-\item[$nf=2$ and $nf=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
+\item[$\mathrm{nf}=2$ and $\mathrm{nf}=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
 \end{description}
@@
 
-An interesting experiment we additionally performed was relaxing the memory threshold by setting $T_H = L3_{size}$.
-We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle nf=2, q=2\rangle$.
+An interesting experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$.
+We found that this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2, q=2\rangle$.
 This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.
@@
 The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should.
-To clarify this, consider the test case $\langle nf=0,\ q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
+To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
 \texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$.
-However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}.
+However, running the same experiments with $T_H = \operatorname{size}(\mathrm{L3})$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}.
 The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the saving in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice.
@@
 \item the differences between \texttt{cfO2} and \texttt{cfO1} are due to the perfect sharing elimination and the zero-valued blocks avoidance technique presented in Section~\ref{sec:zeros}.
-\item when $nf=1$, \texttt{auto} prefers \texttt{tens} over \texttt{quad}, which leads to sub-optimal operation counts and execution times.
+\item when $\mathrm{nf}=1$, \texttt{auto} prefers \texttt{tens} over \texttt{quad}, which leads to sub-optimal operation counts and execution times.
 \item \texttt{ufls} often results in better execution times than \texttt{quad} and \texttt{tens}.
@@
 The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points.
-Firstly, the baseline is highly inefficient.
-With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) fails at exploiting the structure inherent in the expressions.
-Secondly, the strategy for removing spatial and temporal sharing has a tremendous impact.
+First, the baseline is highly inefficient.
+With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (the Intel compiler on an Intel platform at maximum optimization level) fails to exploiting the structure inherent in the expressions.
+Second, the strategy for removing spatial and temporal sharing has a tremendous impact.
 Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.
@@
 There are three main limitations in this process.
-Firstly, we do not have a systematic strategy to optimize sub-expressions independent of linear loops; despite we have a mechanism to determine ``how much computation'' should be hoisted at the level of the integration (reduction) loop, it is not clear how to effectively improve the heuristics used at step (6) in Algorithm~\ref{algo:sharing-elimination}.
-Secondly, lower operation counts may be found by exploiting domain-specific properties, such as redundancies in basis functions; this aspect is completely neglected in this article.
-Thirdly, with Constraint~\ref{const:Le} we have limited the applicability of code motion, which was essential given the complexity of the problem tackled.
+First, we do not have a systematic strategy to optimize sub-expressions which are independent of linear loops.
+Although we have a mechanism to determine how much computation should be hoisted to the level of the integration (reduction) loop, it is not clear how to effectively improve the heuristics used at step (6) in Algorithm~\ref{algo:sharing-elimination}.
+Second, lower operation counts may be found by exploiting domain-specific properties, such as redundancies in basis functions; this aspect is completely neglected in this article.
+Third, with Constraint~\ref{const:Le} we have limited the applicability of code motion.
+This constraint was essential given the complexity of the problem tackled.
 
 Another issue raised by the experimentation concerns selecting a proper threshold for Constraint~\ref{const:TH}.
-To solve this problem we probably need a sophisticated cost model, which is clearly out of the scope of this research.
-
-We also identify two possible research directions: a classification of forms that can attain optimality; a generalization of the methodology to other classes of loop nests, for instance those arising in spectral methods.
+To solve this problem would require a more sophisticated cost model, which is an interesting question deserving further research.
+
+We also identify two additional possible research directions: a complete classification of forms for which a global optimum is achieved; and a generalization of the methodology to other classes of loop nests, for instance those arising in spectral element methods.
 
@@
 We have developed a theory for the optimization of finite element integration loop nests.
-The article clearly explains what domain properties are exploited by our approach (e.g., linearity) and how these translate at the level of loop nests.
+The article details the domain properties which are exploited by our approach (e.g., linearity) and how these translate to transformations at the level of loop nests.
 All of the algorithms shown in this paper have been implemented in COFFEE, a compiler publicly available fully integrated with the Firedrake framework.

### 2016-04-05 TOMS_Paper/assembly.tex
select rather than individuate
some more very small rewrites, and one more TODO

@@
 After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, the ILP model is instantiated.
-At this point, proving optimality reduces to convince ourselves about the correctness of the model, which is relatively straightforward because of its simplicity.
-The model aims to minimize the operation count by individuating the most promising factorizations.
+At this point, proving optimality reduces to establishing the correctness of the model, which is relatively straightforward because of its simplicity.
+The model aims to minimize the operation count by selecting the most promising factorizations.
 The second set of constraints is to select all edges (i.e., all multiplications), exactly once.
-The first set of inequalities allows scheduling multiplications: once a vertex $s$ is selected (i.e., once a symbol is decided to be factorized), all multiplications involving $s$ can be grouped.
+The first set of inequalities allows multiplications to be scheduled: once a vertex $s$ is selected (i.e., once a symbol is decided to be factorized), all multiplications involving $s$ can be grouped.
 \end{enumerate}
@@
 
-Throughout the paper we have reiterated the idea that Algorithm~\ref{algo:gamma} achieves a globally optimal flop count if stronger preconditions on the input variational form are satisfied.
+Throughout the paper we have reiterated the claim that Algorithm~\ref{algo:gamma} achieves a globally optimal flop count if stronger preconditions on the input variational form are satisfied.
 We state here these preconditions, in increasing order of complexity.
@@
 This has several advantages.
-First, extendibility: novel transformations -- for instance, sum factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available.
+The first is extensibility.
+New transformations, such as sum factorization in spectral methods, could be expressed by composing the existing operators, or with small effort building on what is already available.
 Second, generality: COFFEE can be seen as a lightweight, low level computer algebra system, not necessarily tied to finite element integration.
-Third, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.
-The rewrite operators, whose (Python) implementation is based on manipulation of abstract syntax trees (ASTs), compose the COFFEE language.
+Third, robustness: the same operators are exploited, and therefore tested, by different optimization pipelines.
+The rewrite operators, whose (Python) implementation is based on manipulation of abstract syntax trees (ASTs), comprise the COFFEE language.
 A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion.
@@
 It provides an interface to build generic ASTs and only expects expressions to be in normal form (or sufficiently close to it).
-In Firedrake, for example, a modified version of the FEniCS Form Compiler producing ASTs, instead of strings, is used.
-Thus, COFFEE decouples the mathematical manipulation of a form from code optimization; or, in other words, relieves form compiler developers of caring about performance of generated code.
+For example, Firedrake has transitioned from a version of the FEniCS Form Compiler \cite{FFC-TC} modified to produce ASTs rather than strings, to a newly written compiler\footnote{TSFC, the two-stage form compiler \url{https://github.com/firedrakeproject/tsfc}}, while continuing to emply COFFEE.
+Thus, COFFEE decouples the mathematical manipulation of a form from code optimization; or, in other words, relieves form compiler developers of the task of fine scale loop optimization of generated code.
 
@@
 \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed.
-This technique, however, promotes non-contiguous memory loads and stores, which nullify the potential benefits of vectorization.
-COFFEE, instead, handles block-sparse basis function tables by restructuring loops in a way that low level optimization (especially vectorization) is only marginally affected.
-This is based on symbolic execution of the code, which enables a series of checks on array indices and loop bounds for determining which zero-valued blocks can be skipped without affecting data alignment.
+This technique, however, produces non-contiguous memory loads and stores, which nullify the potential benefits of vectorization.
+COFFEE, instead, handles block-sparse basis function tables by restructuring loops in such a manner that low level optimization (especially vectorization) is only marginally affected.
+This is based on symbolic execution of the code, which enables a series of checks on array indices and loop bounds which determine the zero-valued blocks which can be skipped without affecting data alignment.
 
@@
 
-For a fair comparison, small patches (publicly available) were written for the compatibility of \texttt{quad}, \texttt{tens}, and \texttt{ufls} with Firedrake.
+For a fair comparison, small patches (publicly available) were written to the make \texttt{quad}, \texttt{tens}, and \texttt{ufls} compatible with Firedrake.
 By executing all simulations in Firedrake, we guarantee that both matrix insertion and mesh iteration have a fixed cost, independent of the transformation system employed.
@@
 The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
-We note that if a bar or a subplot are missing, then the form compiler failed at generating code because it either exceeded the system memory limit or was unable to handle the form.
+We note that if a bar or a subplot are missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.
 
@@
 \paragraph{Autovectorization}
+
+{\color{red}I don't think you mean ``discretized'' in the next sentence. We need to discuss what this really means.}
+
 We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length.
@@
 We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly.
-As $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups.
-It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable.
+For $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups.
+It is worth noting why \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable.
 On the other hand, \texttt{cfO2} always makes the optimal decision about whether to apply pre-evaluation or not.
@@
 \paragraph{Helmholtz}
-As in the case of Mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular, with
+As in the case of Mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular:
 \begin{description}
@@
 \item[$\mathrm{nf}=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$.
-This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it rightfully avoids pre-evaluation because of the excessive expansion cost.
+This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it correctly avoids pre-evaluation because of the excessive expansion cost.
 \item[$\mathrm{nf}=2$ and $\mathrm{nf}=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
@@
 In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper.
-This is a positive aspect, since it indicates that our transformation algorithm scales well with form complexity.
+This is a positive result, since it indicates that our transformation algorithm scales well with form complexity.
 The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points.

### 2016-04-07 TOMS_Paper/assembly.tex
addressing issues raised by David

@@
 % Page heads
-\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration loops}
+\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration}
 
@@
 
-\title{An algorithm for the optimization of finite element integration loops}
+\title{An algorithm for the optimization of finite element integration}
 \author{Fabio Luporini
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration loops.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration.}
 
@@
 \textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
-\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined in a tree visit.}
+\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined through tree visits.}
 
@@
 
+The Zenodo system was used to archive all packages used to perform the experiments: Firedrake \cite{lawrence_mitchell_2016_49284}, PETSc \cite{barry_smith_2016_49285}, petsc4py \cite{firedrake_2016_49283}, FIAT \cite{marie_e_rognes_2016_49280}, UFL \cite{martin_sandve_alnaes_2016_49282}, FFC \cite{anders_logg_2016_49276}, PyOP2 \cite{florian_rathgeber_2016_49281} and COFFEE \cite{fabio_luporini_2016_49279}.
+
 We analyze the execution time of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods.
@@
 
-{\color{red}FIXME: We need to put the exact versions of software in
-  here. for Firedrake this should be Zenodo DOIs. This is especially true
-  for the patched version of firedrake you used below}.
-
 The values that we report are the average of three runs with ``warm cache''; that is, with all kernels retrieved directly from the Firedrake's cache, so code generation and compilation times are not counted.
@@
 
-For a fair comparison, small patches (publicly available) were written to the make \texttt{quad}, \texttt{tens}, and \texttt{ufls} compatible with Firedrake.
+For a fair comparison, small patches were written to the make \texttt{quad}, \texttt{tens}, and \texttt{ufls} compatible with Firedrake.
 By executing all simulations in Firedrake, we guarantee that both matrix insertion and mesh iteration have a fixed cost, independent of the transformation system employed.
@@
 
-{\color{red}I don't think you mean ``discretized'' in the next sentence. We need to discuss what this really means.}
-
-We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length.
+We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length.
 This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}.

### 2016-04-07 TOMS_Paper/assembly.tex
Fix last standing issue with plots. Add plotting script
Integrate Paul's suggestions. Still wondering about the title

@@
 We present an algorithm for the optimization of a class of finite element integration loop nests.
-This algorithm is proven to achieve a locally optimal operation count.
+This algorithm, which exploits fundamental mathematical properties of finite element operators, is proven to achieve a locally optimal operation count.
 In specified circumstances the optimum achieved is global.
 Extensive numerical experiments demonstrate significant performance improvements over the state of the art in finite element code generation in almost all cases.
-This validates the algorithm presented here, and illustrate its limitations.
+This validates the effectiveness of the algorithm presented here, and illustrates its limitations.
 \end{abstract}
@@
 NE/K008951/1 and NE/K006789/1], and by a HiPEAC collaboration
-grant. The authors would like to thank Mr. Andrew T.T. McRae,
+grant. The authors would like to thank Dr. Andrew T.T. McRae,
 Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
@@
 This is a result of the complexity inherent in the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals.
-General-purpose compilers, such as those by \emph{GNU} and \emph{Intel}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}).
-Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO \cite{PLUTO}, focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops.
+General-purpose compilers, such as those by \emph{GNU} and \emph{Intel}, fail to exploit the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this in Section~\ref{sec:perf-results}).
+Research compilers, for instance those based on polyhedral analysis of loop nests, such as PLUTO \cite{PLUTO}, focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops.
 The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems.
@@
 In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed.
-It is also worth mentioning two new new form compilers, UFLACS \cite{Uflacs} and \cite{TSFC}, which particularly target the compile time challenges of the more complex variational forms.
-The performance evaluation in Section~\ref{sec:perf-results} includes most of these optimization systems.
+It is also worth mentioning two new new form compilers, UFLACS \cite{Uflacs} and TSFC \cite{TSFC}, which particularly target the compilation time challenges of the more complex variational forms.
+The performance evaluation in Section~\ref{sec:perf-results} includes most of these systems.
 
 However, in spite of such a considerable research effort, there is still no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed?
-In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases (we will define ``local optimality'', which relates operation count with inner loops).
+In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations in all other cases.
+In particular, we will define ``local optimality'', which relates operation count with inner loops.
 In summary, our contributions are as follows:
@@
 
-{\color{red}FIXME: Please replace $nf$ with $\mathrm{nf}$ in all figure captions.}
-
 \begin{figure}
@@
 
-The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.
+The slow-downs (or marginal improvements) seen in a small number of cases exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.
 
@@
 First, the baseline is highly inefficient.
-With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (the Intel compiler on an Intel platform at maximum optimization level) fails to exploiting the structure inherent in the expressions.
+With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (the Intel compiler on an Intel platform at maximum optimization level) fails to exploit the structure inherent in the expressions.
 Second, the strategy for removing spatial and temporal sharing has a tremendous impact.
 Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.
+
+\section{Conclusions}
+\label{sec:conclusions}
+We have developed a theory for the optimization of finite element integration loop nests.
+The article details the domain properties which are exploited by our approach (e.g., linearity) and how these translate to transformations at the level of loop nests.
+All of the algorithms shown in this paper have been implemented in COFFEE, a compiler publicly available fully integrated with the Firedrake framework.
+The correctness of the transformation algorithm was discussed.
+The performance results achieved suggest the effectiveness of our methodology.
 
@@
 
-\section{Conclusions}
-\label{sec:conclusions}
-We have developed a theory for the optimization of finite element integration loop nests.
-The article details the domain properties which are exploited by our approach (e.g., linearity) and how these translate to transformations at the level of loop nests.
-All of the algorithms shown in this paper have been implemented in COFFEE, a compiler publicly available fully integrated with the Firedrake framework.
-The correctness of the transformation algorithm was discussed.
-The performance results achieved suggest the effectiveness of our methodology.
-
 % Bibliography

### 2016-04-07 TOMS_Paper/assembly.tex
minor fixes

@@
 
+\usepackage[utf8]{inputenc}
+
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 % Page heads
-\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration}
+\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration operators}
 
@@
 
-\title{An algorithm for the optimization of finite element integration}
+\title{An algorithm for the optimization of finite element integration operators}
 \author{Fabio Luporini
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration operators.}
 
@@
 The Zenodo system was used to archive all packages used to perform the experiments: Firedrake \cite{lawrence_mitchell_2016_49284}, PETSc \cite{barry_smith_2016_49285}, petsc4py \cite{firedrake_2016_49283}, FIAT \cite{marie_e_rognes_2016_49280}, UFL \cite{martin_sandve_alnaes_2016_49282}, FFC \cite{anders_logg_2016_49276}, PyOP2 \cite{florian_rathgeber_2016_49281} and COFFEE \cite{fabio_luporini_2016_49279}.
+The experiments can be reproduced using a publicly available benchmarking suite~\cite{florian_rathgeber_2016_49290}.
 

### 2016-04-07 TOMS_Paper/assembly.tex
minor fixes

@@
 Each sub-expression is then factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms.
-This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py} and in \texttt{coffee/optimizer.py} in \citeN{FIXME-PUT-THE-COFFEE-DOI-CITE-HERE}.}, as a consequence of the results in~\citeN{Kirby:TC}.
+This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py} in \citeN{fabio_luporini_2016_49279}.}, as a consequence of the results in~\citeN{Kirby:TC}.
 These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries.

### 2016-04-08 TOMS_Paper/assembly.tex
minor fixes
change title

@@
 % Page heads
-\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration operators}
+\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration loops}
 
@@
 
-\title{An algorithm for the optimization of finite element integration operators}
+\title{An algorithm for the optimization of finite element integration loops}
 \author{Fabio Luporini
@@
 
-\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration operators.}
+\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration loops.}
 
@@
 The Zenodo system was used to archive all packages used to perform the experiments: Firedrake \cite{lawrence_mitchell_2016_49284}, PETSc \cite{barry_smith_2016_49285}, petsc4py \cite{firedrake_2016_49283}, FIAT \cite{marie_e_rognes_2016_49280}, UFL \cite{martin_sandve_alnaes_2016_49282}, FFC \cite{anders_logg_2016_49276}, PyOP2 \cite{florian_rathgeber_2016_49281} and COFFEE \cite{fabio_luporini_2016_49279}.
-The experiments can be reproduced using a publicly available benchmarking suite~\cite{florian_rathgeber_2016_49290}.
+The experiments can be reproduced using a publicly available benchmark suite~\cite{florian_rathgeber_2016_49290}.
 

### 2016-11-30 TOMS_Paper/assembly.tex
Slightly smaller plots
Add stuff about flops performed and data transfer

@@
 \item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
+\item double-precision arithmetic
 \end{itemize}
@@
 \begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/mass}}
 \caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
@@
 \begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/helmholtz}}
 \caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation.
@@
 \begin{figure}
-\includegraphics[scale=0.77]{perf-results/elasticity}
+\includegraphics[scale=0.70]{perf-results/elasticity}
 \caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model.
@@
 \begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/hyperelasticity}}
 \caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model.
@@
 \paragraph{Mass matrix}
-We start with the simplest of the bilinear forms investigated, the mass matrix.
-Results are in Figure~\ref{fig:mass}.
-We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly.
-For $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups.
-It is worth noting why \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable.
+We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}.
+
+We first notice the lack of improvements when $q=1$.
+The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreads operations), but nontheless no gain in execution time is obtained.
+This is due to the fact that local assembly is memory-bound.
+For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requries three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed.
+In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format.
+This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which collocates the kernels in a memory-bound regime.
+
+For $q \geq 2$, instead, the transformation systems impact the performance, and \texttt{cfO2} generally shows the highest speed-ups.
+It is worth noting that \texttt{auto} does not always select the fastest implementation, as it always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable.
 On the other hand, \texttt{cfO2} always makes the optimal decision about whether to apply pre-evaluation or not.
-Surprisingly, despite the simplicity of the form, the performance of the various code generation systems can differ significantly.
+Unexpectedly, in spite of the simplicity of the form, the performance of the various code generation systems can differ significantly.
 
 \paragraph{Helmholtz}
-As in the case of Mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular:
+As in the case of mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular:
+
 \begin{description}
@@
 
+In the case $\langle \mathrm{nf}=0,\ q=2 \rangle$, the 2$\times$ speed-ups achieved by \texttt{cfO2} and \texttt{tens} are due to the remarkable drop in operation count achieved by pre-evaluating reductions: from order $10^4$ (\texttt{ufls} $3 \cdot 10^4$, \texttt{cfO1} $2 \cdot 10^4$) to order $10^3$.
+By unrolling loops and eliminating operations over zeros, \texttt{tens} eventually executes fewer operations than \texttt{cfO2} (roughly, $10^3$ instead of $2 \cdot 10^3$), but the performance is basically identical because of matrix insertion, which plays again the limiting role.
+With $10^2$ entries to be updated in the global matrix, following the same procedure of the previous paragraph, it is straightforward to verify that $\frac{\text{operation count}}{\text{bytes moved}} \ll 1$, a typical symptom of memory-boundedness.
+
+Another relevant problem instance is $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{cfO2} performs 1.5$\times$ more operations than \texttt{tens}, but still exhibits a better execution time.
+This is again caused by the chosen code generation strategies: while \texttt{tens} prefers to fully unroll the computation to detect all avoidable operations, \texttt{cfO2} preserves the loop structure.
+This choice mitigates the memory transfer (especially between L1 cache and registers) by maximizing register locality.
+
 The slow-downs (or marginal improvements) seen in a small number of cases exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.
+Further, the factorization applied by \texttt{ufls} hides code motion opportunities that the underlying compiler would otherwise have been able to exploit.
 
@@
 The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should.
-To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
+To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
 \texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$.
@@
 Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.
+
+We also tried to relate the achieved performance to the operation counts obtained by each transformation system.
+We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive.
+For the untransformed, original code, at the source level $37 \cdot 10^9$ operations are counted.
+However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have aggressively been applied.
+The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}.
+\texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}.
+This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by more than 3$\times$.
+
+Unsurprisingly, the memory transfer impact is less relevant in this hyperelastic than in all previous forms.
+With $\langle \mathrm{nf}=0,\ q=1 \rangle$, \texttt{cfO2} is already 1.4$\times$ faster than the untransformed code, and the gain increases with both $nf$ and $q$.
 

### 2016-12-05 TOMS_Paper/assembly.tex
More edits about flops vs performance

@@
 \begin{enumerate}
-\item There is a single monomial and only a specific coefficient (e.g., the coordinates field).
+\item There is a single monomial and only one coefficient (e.g., the coordinates field).
 This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
@@
 
-The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained.
+The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained and the relationship with the operation count.
 
@@
 To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}.
-\texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$.
+Although pre-evaluation is determined profitable by \texttt{cfO2} in terms of operation count, it is eventually not applied to avoid exceeding $T_H$.
 However, running the same experiments with $T_H = \operatorname{size}(\mathrm{L3})$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}.
@@
 \end{itemize}
+
+We now consider the test case $\langle \mathrm{nf}=2,\ q=3 \rangle$ to discuss the relationship between achieved performance and operation count, although analogous considerations apply to the other scenarios.
+It is difficult to estimate the number of flops executed by the code generated by the FEniCS Form Compiler, since loop-invariant code motion is extensively applied by the underlying compiler.
+The calculation is instead simple in all other cases: while \texttt{cfO2} performs roughly $2 \cdot 10^5$ operations, \texttt{quad}, \texttt{ufls} and \texttt{cfO1} are much closer to $10^6$ operations.
+In particular, \texttt{ufls} requires 4.3$\times$ more operations than \texttt{cfO2}, whereas the execution time speed-up is only slightly bigger than 2$\times$.
+The distance between these two values may be attributed to the memory transfer cost -- especially inserting the $3.6 \cdot 10^3$ non-zero entries of the local element matrix -- which, for reasons already discussed, becomes a bottleneck.
+\texttt{quad}'s operation count is even lower than that of \texttt{ufls} ($84 \cdot 10^4$ instead of $97 \cdot 10^5$), but as explained earlier, the use of indirection arrays dramatically affects the low level efficiency.
+Despite requiring roughly $10^5$ fewer operations than \texttt{cfO2}, \texttt{tens} is again affected by the poor low level efficiency of the generated code.
 
@@
 We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive.
-For the untransformed, original code, at the source level $37 \cdot 10^9$ operations are counted.
-However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have aggressively been applied.
+For the original code, at the source level $37 \cdot 10^9$ operations are counted.
+However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have been applied in aggressive way.
 The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}.
 \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}.
-This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by more than 3$\times$.
+This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by slightly more than 3$\times$.
 

### 2016-12-05 TOMS_Paper/assembly.tex
Minor addition about parallel runs

@@
 
-\begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/mass}}
-\caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
-\end{figure}
-
-\begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/helmholtz}}
-\caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation.
-The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.70]{perf-results/elasticity}
-\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model.
-The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
-\end{figure}
-
-\begin{figure}
- \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/hyperelasticity}}
-\caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model.
-The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
-\end{figure}
-
 We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
@@
 
-\paragraph{High level view}
+\subsubsection{High level view}
 Our transformation strategy does not always guarantee minimum execution time.
@@
 
-\paragraph{Autovectorization}
+\subsubsection{Autovectorization}
 
@@
 
-\paragraph{Mass matrix}
+\subsubsection{Mass matrix}
+
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.685]{perf-results/mass}}
+\caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
+\end{figure}
+
 We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}.
@@
 
-\paragraph{Helmholtz}
+\subsubsection{Helmholtz}
+
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.685]{perf-results/helmholtz}}
+\caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation.
+The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
+\end{figure}
+
 As in the case of mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular:
@@
 
-\paragraph{Elasticity}
+\subsubsection{Elasticity}
+
+\begin{figure}
+\includegraphics[scale=0.685]{perf-results/elasticity}
+\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model.
+The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
+\end{figure}
+
 The results for the elastic model are displayed in Figure~\ref{fig:elasticity}.
@@
 
-\paragraph{Hyperelasticity}
+\subsubsection{Hyperelasticity}
+
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.685]{perf-results/hyperelasticity}}
+\caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model.
+The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
+\end{figure}
+
 In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper.
@@
 Another issue raised by the experimentation concerns selecting a proper threshold for Constraint~\ref{const:TH}.
-To solve this problem would require a more sophisticated cost model, which is an interesting question deserving further research.
+Furthermore, in parallel runs, the L3 cache is no longer owned by a single process, so the effects of setting a proper threshold could become even more pronounced.
+To solve this problem would require more sophisticated cost model and/or objective functions, which is an interesting question deserving further research.
 

### 2016-12-05 TOMS_Paper/assembly.tex
Change section 7 headings
minor fixes

@@
 
-\subsection{Performance results}
-\label{sec:perf-results-forms}
-
 We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots.
@@
 
-\subsubsection{High level view}
+\subsection{High level view}
 Our transformation strategy does not always guarantee minimum execution time.
@@
 
-\subsubsection{Autovectorization}
+\subsection{Autovectorization}
 
@@
 
-\subsubsection{Mass matrix}
+\subsection{Mass matrix}
 
@@
 
-\subsubsection{Helmholtz}
+\subsection{Helmholtz}
 
@@
 By unrolling loops and eliminating operations over zeros, \texttt{tens} eventually executes fewer operations than \texttt{cfO2} (roughly, $10^3$ instead of $2 \cdot 10^3$), but the performance is basically identical because of matrix insertion, which plays again the limiting role.
-With $10^2$ entries to be updated in the global matrix, following the same procedure of the previous paragraph, it is straightforward to verify that $\frac{\text{operation count}}{\text{bytes moved}} \ll 1$, a typical symptom of memory-boundedness.
-
-Another relevant problem instance is $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{cfO2} performs 1.5$\times$ more operations than \texttt{tens}, but still exhibits a better execution time.
-This is again caused by the chosen code generation strategies: while \texttt{tens} prefers to fully unroll the computation to detect all avoidable operations, \texttt{cfO2} preserves the loop structure.
-This choice mitigates the memory transfer (especially between L1 cache and registers) by maximizing register locality.
+With $10^2$ entries to be updated in the global matrix, following the same procedure of the previous section, it is straightforward to verify that $\frac{\text{operation count}}{\text{bytes moved}} \ll 1$, a typical symptom of memory-boundedness.
+
+Another interesting problem instance is $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{cfO2} performs 1.5$\times$ more operations than \texttt{tens}, but still exhibits a better execution time.
+This is again a consequence of the code generation strategy: while \texttt{tens} prefers to fully unroll the computation to detect all avoidable operations, \texttt{cfO2} preserves the loop structure.
+This choice mitigates the memory transfer, especially between L1 cache and registers, by maximizing data locality.
 
@@
 
-An interesting experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$.
+A relevant experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$.
 We found that this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2, q=2\rangle$.
@@
 
-\subsubsection{Elasticity}
+\subsection{Elasticity}
 
@@
 We now consider the test case $\langle \mathrm{nf}=2,\ q=3 \rangle$ to discuss the relationship between achieved performance and operation count, although analogous considerations apply to the other scenarios.
-It is difficult to estimate the number of flops executed by the code generated by the FEniCS Form Compiler, since loop-invariant code motion is extensively applied by the underlying compiler.
-The calculation is instead simple in all other cases: while \texttt{cfO2} performs roughly $2 \cdot 10^5$ operations, \texttt{quad}, \texttt{ufls} and \texttt{cfO1} are much closer to $10^6$ operations.
+It is difficult to estimate the number of flops executed by the code produced by the FEniCS Form Compiler, since loop-invariant code motion is extensively applied by the underlying compiler.
+The calculation is instead simple in all other cases: while \texttt{cfO2} performs $2.3 \cdot 10^5$ operations, \texttt{quad}, \texttt{ufls} and \texttt{cfO1} are close to $10^6$ operations.
 In particular, \texttt{ufls} requires 4.3$\times$ more operations than \texttt{cfO2}, whereas the execution time speed-up is only slightly bigger than 2$\times$.
 The distance between these two values may be attributed to the memory transfer cost -- especially inserting the $3.6 \cdot 10^3$ non-zero entries of the local element matrix -- which, for reasons already discussed, becomes a bottleneck.
-\texttt{quad}'s operation count is even lower than that of \texttt{ufls} ($84 \cdot 10^4$ instead of $97 \cdot 10^5$), but as explained earlier, the use of indirection arrays dramatically affects the low level efficiency.
+\texttt{quad}'s operation count is even lower than that of \texttt{ufls} ($8.4 \cdot 10^5$ instead of $9.7 \cdot 10^5$), but as explained earlier, the use of indirection arrays dramatically affects the low level efficiency.
 Despite requiring roughly $10^5$ fewer operations than \texttt{cfO2}, \texttt{tens} is again affected by the poor low level efficiency of the generated code.
 
-\subsubsection{Hyperelasticity}
+\subsection{Hyperelasticity}
 
@@
 We also tried to relate the achieved performance to the operation counts obtained by each transformation system.
-We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive.
-For the original code, at the source level $37 \cdot 10^9$ operations are counted.
-However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have been applied in aggressive way.
-The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}.
+We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups were generally massive.
+In the original code, $37 \cdot 10^9$ operations were counted at the source level.
+However, the number of operations actually executed is in practice much lower, since by inspection of the assembly code generated by the underlying compiler it is clear that common sub-expressions elimination and loop-invariant code motion are applied in aggressive way.
+The actual operation count is, however, definitely larger than the $10^8$ operations performed by \texttt{ufls}.
 \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}.
@@
 
-Unsurprisingly, the memory transfer impact is less relevant in this hyperelastic than in all previous forms.
+Unsurprisingly, the memory transfer impact in the hyperelastic model is less relevant than in all other forms studied.
 With $\langle \mathrm{nf}=0,\ q=1 \rangle$, \texttt{cfO2} is already 1.4$\times$ faster than the untransformed code, and the gain increases with both $nf$ and $q$.

### 2016-12-21 TOMS_Paper/perf-results/overview/counted_flops.py
WIP: working on the scatter plot

@@ -1,1 +1,1 @@
-flops = OrderedDict([(('mass', 'original', 0, 1), 30), (('mass', 'coffee', 0, 1), 15), (('mass', 'original', 0, 2), 60), (('mass', 'coffee', 0, 2), 30), (('mass', 'original', 1, 1), 25), (('mass', 'coffee', 1, 1), 20)])
+flops = OrderedDict([(('mass', 'original', 0, 1), '256'), (('mass', 'original', 0, 2), '5600'), (('mass', 'original', 0, 3), '38400'), (('mass', 'original', 0, 4), '612500'), (('mass', 'original', 1, 1), '440'), (('mass', 'original', 1, 2), '12480'), (('mass', 'original', 1, 3), '255000'), (('mass', 'original', 1, 4), '2124885'), (('mass', 'original', 2, 1), '1568'), (('mass', 'original', 2, 2), '80000'), (('mass', 'original', 2, 3), '850640'), (('mass', 'original', 2, 4), '5460210'), (('mass', 'original', 3, 1), '2040'), (('mass', 'original', 3, 2), '164160'), (('mass', 'original', 3, 3), '1495040'), (('mass', 'original', 3, 4), '11692835'), (('mass', 'coffee', 0, 1), '32'), (('mass', 'coffee', 0, 2), '200'), (('mass', 'coffee', 0, 3), '800'), (('mass', 'coffee', 0, 4), '2450'), (('mass', 'coffee', 1, 1), '132'), (('mass', 'coffee', 1, 2), '2010'), (('mass', 'coffee', 1, 3), '16020'), (('mass', 'coffee', 1, 4), '85785'), (('mass', 'coffee', 2, 1), '358'), (('mass', 'coffee', 2, 2), '10847'), (('mass', 'coffee', 2, 3), '850640'), (('mass', 'coffee', 2, 4), '5460210'), (('mass', 'coffee', 3, 1), '1590'), (('mass', 'coffee', 3, 2), '121392'), (('mass', 'coffee', 3, 3), '1086464'), (('mass', 'coffee', 3, 4), '8434547'), (('helmholtz', 'original', 0, 1), '2560'), (('helmholtz', 'original', 0, 2), '56000'), (('helmholtz', 'original', 0, 3), '384000'), (('helmholtz', 'original', 0, 4), '6125000'), (('helmholtz', 'original', 1, 1), '3320'), (('helmholtz', 'original', 1, 2), '98880'), (('helmholtz', 'original', 1, 3), '2055000'), (('helmh
[…]

### 2016-12-21 TOMS_Paper/perf-results/overview/scatterplot.py

@@ -20,6 +20,7 @@ if len(sys.argv) != 3:
 converter = {'original': 'plain', 'coffee': 'coffee-O4'}
 
 problems = ['mass', 'helmholtz', 'elasticity', 'hyperelasticity']
+nfs = [0, 1, 2, 3]
 
 # Bring 'flops' in scope
 filename = os.path.join(sys.argv[1], 'counted_flops.py')
@@ -41,37 +42,88 @@ for (prob, mode, nf, q), v in flops.items():
     runtime = results[key]['nf %d' % nf]
 
     # Append (flops, runtime) for a <prob, nf, q> test case
-    mapper.setdefault((prob, nf, q), {})[mode] = (v, runtime)
+    mapper.setdefault((prob, nf, q), {})[mode] = (int(v), runtime)
 
 
+# Set up chart global properties
 plt.style.use('ggplot')
-
 fig = plt.figure()
 ax = fig.add_subplot(1, 1, 1)
-ax.set_ylabel(r'Speedup', fontsize=11, color='black')
-ax.set_xlabel(r'Operation count ratio', fontsize=11, color='black')
-ax.set_ylim([0.0, 5.0])
-ax.set_xlim([1.0, 3.0])
-ax.set_title("Scatter plot")
+ax.set_title("Correlation between flop reductions and achieved speed-ups", fontsize=12, y=1.03)
+# x-axis
+ax.set_xlabel(r'$\frac{\mathtt{baseline}\ \mathregular{flops}}{\mathtt{cfO2}\ \mathregular{flops}}$',
+              fontsize=15, color='black', labelpad=6.5)
+ax.set_xlim([0.0, 1.0])
+ax.set_xscale('log')
+# y-axis
+ax.set_ylabel(r'Speedup relative to baseline', fontsize=12, color='black', labelpad=12.0)
+ax.set_ylim([0.0, 1.0])
 
-set2 = brewer2mpl.get_map('Set1', 'qualitative', 4).hex_colors
-markers = ['x', 'v', '+', 'o']
-legend_font = FontProperties(size='xx-small')
+cm = brewer2mpl.get_map('Spectral', 'diverging', 4)
+markers = ['v', '^', '<', '>']
+legend_font = FontProperties(size='small')
 
+# Add points
+legend_points = []
 for (prob, nf, q), flops_by_mode in mapper.items():
     index = problems.index(prob)
-
-    # Append (flops gain, speed up)
[…]

### 2016-12-22 TOMS_Paper/perf-results/overview/scatterplot.py
Scatter plot generator almost ready

@@ -54,14 +54,16 @@ ax.set_title("Correlation between flop reductions and achieved speed-ups", fonts
 ax.set_xlabel(r'$\frac{\mathtt{baseline}\ \mathregular{flops}}{\mathtt{cfO2}\ \mathregular{flops}}$',
               fontsize=15, color='black', labelpad=6.5)
 ax.set_xlim([0.0, 1.0])
-ax.set_xscale('log')
+ax.set_xscale('symlog')
 # y-axis
 ax.set_ylabel(r'Speedup relative to baseline', fontsize=12, color='black', labelpad=12.0)
 ax.set_ylim([0.0, 1.0])
 
 cm = brewer2mpl.get_map('Spectral', 'diverging', 4)
-markers = ['v', '^', '<', '>']
-legend_font = FontProperties(size='small')
+#markers = ['v', '^', '<', '>']
+#markers = ['s', 'o', '+', 'v']
+markers = ['^', 's', 'p', 'o']
+legend_font = FontProperties(size='x-small')
 
 # Add points
 legend_points = []
@@ -88,7 +90,7 @@ for (prob, nf, q), flops_by_mode in mapper.items():
         legend_points.append((handle, ''))
     if nf == 0 and q == 1:
         handle = ax.plot([], [], ms=0, mew=0, lw=0)[0]
-        legend_points.append((handle, prob[:5]))
+        legend_points.append((handle, '              '))
     if nf == 0 and prob == 'mass':
         handle = ax.plot([], [], ms=0, mew=0, lw=0)[0]
         legend_points.append((handle, '$\mathrm{q=%d}$' % q))
@@ -96,8 +98,11 @@ for (prob, nf, q), flops_by_mode in mapper.items():
         handle = ax.plot([], [],
                          marker=markers[index],
                          ms=3*q, mew=1, mfc='black', mec='black', lw=0)[0]
-        legend_points.append((handle, ''))
+        legend_points.append((handle, '              '))
 
+
+# Speed-up >= 1 line
+ax.axhline(y=1, color='black', linestyle='--', linewidth=0.75)
 
 # Colorbar
 discrete_cm = cm.get_mpl_colormap().from_list('Custom cmap', cm.mpl_colors, len(nfs))
@@ -118,12 +123,11 @@ swap(legend_points, 2, 4)
 swap(legend_points, 3, 6)
 swap(legend_points, 4, 8)
 swap(legend_points, 5, 8)
-
[…]

### 2016-12-23 TOMS_Paper/assembly.tex
Lots of fixes towards final version

@@
 
-\section{Transformation Space}
+\section{Transformation space}
 \label{sec:optimal-impl}
@@
 
-\section{Code Generation}
+\section{Code generation}
 \label{sec:codegen}
@@
 
-\section{Performance Evaluation}
+\section{Performance evaluation}
 \label{sec:perf-results}
@@
 The experiments can be reproduced using a publicly available benchmark suite~\cite{florian_rathgeber_2016_49290}.
+Further, the generated code is available at \cite{generated_code}.
 
@@
 \begin{itemize}
-\item the space of test, trial, and coefficient functions: Lagrange
+\item the test, trial, and coefficient functions belong to the space of (possibly vector-valued) continuous piecewise-defined polynomials of degree $q$
 \item the mesh: tetrahedral with a total of 4374 elements
@@
 
+The memory transfer has sometimes a non-negligible impact, especially when low order polynomials are used.
+This is more evident in the simpler bilinear forms (e.g., \texttt{Mass}), when the speed-ups over the baseline are limited.
+The effect of the memory transfer is quantified, to some extent, in the sections below.
+For a comprehensive analysis of its cost, the reader is invited to refer to~\cite{extrusion-paper}, a study based upon the same Firedrake technology used in this article.
+
 \subsection{Autovectorization}
-
 We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length.
@@
 We first notice the lack of improvements when $q=1$.
-The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreads operations), but nontheless no gain in execution time is obtained.
+The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreds operations), but nonetheless no gain in execution time is obtained.
 This is due to the fact that local assembly is memory-bound.
-For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requries three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed.
+For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requires three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed.
 In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format.
@@
 \begin{figure}
-\includegraphics[scale=0.685]{perf-results/elasticity}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.685]{perf-results/elasticity}}
 \caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model.
@@
 With $\langle \mathrm{nf}=0,\ q=1 \rangle$, \texttt{cfO2} is already 1.4$\times$ faster than the untransformed code, and the gain increases with both $nf$ and $q$.
+
+\subsection{On the correlation between operation count and performance}
+As reiterated throughout the paper, the execution time of a kernel is influenced not only by the operation count, but also by the chosen code generation strategy (e.g., loop unrolling rather than preserving the loop structure).
+Figure~\ref{fig:flops-correlation} aggregates the \texttt{cfO2} results from the previous sections and displays them as a function of the reduction in executed flops.
+Besides summarizing the performance improvements achieved in this article, this figure highlights a number of aspects.
+The speed-ups achieved by \texttt{cfO2} tend to increase with the complexity of the operators and $q$.
+Further, the pre-multiplying functions have a limited impact in complex operators, and they occasionally have a negative effect in simple and small problems.
+Finally, this plot shows that even a powerful general-purpose compiler like Intel's is inadequate to fully exploit the structure of non-trivial finite element operators (analogous behaviors were observed with different compilers).
+
+\begin{figure}
+ \makebox[\textwidth][c]{\includegraphics[scale=0.55]{perf-results/overview}}
+\caption{Relationship between operation counts and performance achieved by \texttt{cfO2}.}
+\label{fig:flops-correlation}
+\end{figure}
 
@@
 
+The transformations and the cost models described in this paper assume that a local assembly kernel is executed sequentially by a single process for each mesh element.
+In Firedrake, parallelism is always achieved at the mesh level, by distributing element partitions to different processes.
+However, this strategy could be revisited, for instance, in the case of high order methods (not treated in this article), when the working set to execute an element does not fit in L3 cache.
+This would require parallelism at the level of the basis functions loops.
+The cost models should also be extended, since certain rewrite operators (e.g., factorization and generalized code motion) could on one hand reduce the operation count, but at the same time destroy otherwise easily parallelizable computation.
+
 We also identify two additional possible research directions: a complete classification of forms for which a global optimum is achieved; and a generalization of the methodology to other classes of loop nests, for instance those arising in spectral element methods.

### 2016-12-23 TOMS_Paper/reviewers_answers.txt
Add sketch answers to referees

Referee: 1

Comments to the Author This paper is overall well written. Examples are very
illustrative and help understand the article. My only concern is about how
authors relates their theortical results with numerical performance examples.
The main focus of this paper is about transformations that reduces FLOPs.
Authors present elimination and reduction pattern of code and show that there is
an optimal loop transformation that reduces FLOPs. However, the numeric
performance results is about "speed-up" not about FLOPs. Also, the speed-up is
more significant when the polynomial degree is higher. This confuses me either
vectorization improves the performance or actual FLOPs decreases. My suggestions
are as follows: 1. It would be better to include actual transformed code and
original code in the appendix for all computing model (mass, helmholtz,
elasticity, hyper elasticity). At least I wonder what makes hugh speed up in the
hyper elastic problem compared to usual mass matrix.  2. Please include actual
FLOP measurements of input code and transformed code and show some correlation
with performance.

Referee: 2

Comments to the Author

Report on “An algorithm for the optimization of finite element integration
loops”

The paper is well written, interesting does clearly merit publication.  The
usage of advanced compiler technology as demonstrated here is clearly novel and
it brings accelerated performance.

There is one issue that is not covered in the analysis and only briefly
mentioned in the general discussion: the transfer from memory. I understand that
it complicates the analysis but the least it should be quantified in the
numerical experiments. As was shown by Alnæs and Mardal in 2010, TOMS, the
transfer is significant. This part of the story should also be told as I guess
the most common user, low-order elements with simple equation, will not notice
any benefit. This has been an untold story in several FEniCS papers. It is to a
certain extent discussed already in the paper but it should also be quantified.

Referee: 3

Comments to the Author The FEM is widely used in scientific computing and the
integration to compute the elements (local element matrices)  can often be time
consuming. This manuscript is about optimizing the finite element integration
loops and the number of flops therein. The authors propose an approach that is
shown to be locally optimal. An implementation (in COFFEE) is compared against
other methods and speedups over the default generated code is demonstrated in
several cases.

Overall this is a well written and interesting paper, even to non-experts.
However, as a non-expert, I have several concerns that I think should be
addressed:

1) The focus is on minimizing flops, but on modern architectures, memory
accesses are typically more expensive and flops are "almost free". Although
clearly of theoretical interest, it would be nice to justify that this
optimization also is important to reduce run time.

2) All experiments and results are run in serial on single core. In practice,
most computers today are multicore (or manycore) so some discussion about
extensions to multithreaded case would be appropriate. I recognize a parallel
implementation is beyond the scope of the current paper.

3) Results: The authors present several charts with speedups on different
problems, which gives detailed information. However, it would be nice to add a
figure with aggregate results, such as a scatter plot of all instances or a
performance profile.

4) The manuscript is already quite long so if possible, do not increase the
total length (i.e. try find things to tighten or cut).Referee: 1

Comments to the Author This paper is overall well written. Examples are very
illustrative and help understand the article. My only concern is about how
authors relates their theortical results with numerical performance examples.
The main focus of this paper is about transformations that reduces FLOPs.
Authors present elimination and reduction pattern of code and show that there is
an optimal loop transformation that reduces FLOPs. However, the numeric
performance results is about "speed-up" not about FLOPs. Also, the speed-up is
more significant when the polynomial degree is higher. This confuses me either
vectorization improves the performance or actual FLOPs decreases. My suggestions
are as follows: 1. It would be better to include actual transformed code and
original code in the appendix for all computing model (mass, helmholtz,
elasticity, hyper elasticity). At least I wonder what makes hugh speed up in the
hyper elastic problem compared to usual mass matrix.  2. Please include actual
FLOP measurements of input code and transformed code and show some correlation
with performance.

Referee: 2

Comments to the Author

Report on “An algorithm for the optimization of finite element integration
loops”

The paper is well written, interesting does clearly merit publication.  The
usage of advanced compiler technology as demonstrated here is clearly novel and
it brings accelerated performance.

There is one issue that is not covered in the analysis and only briefly
mentioned in the general discussion: the transfer from memory. I understand that
it complicates the analysis but the least it should be quantified in the
numerical experiments. As was shown by Alnæs and Mardal in 2010, TOMS, the
transfer is significant. This part of the story should also be told as I guess
the most common user, low-order elements with simple equation, will not notice
any benefit. This has been an untold story in several FEniCS papers. It is to a
certain extent discussed already in the paper but it should also be quantified.

Referee: 3

Comments to the Author The FEM is widely used in scientific computing and the
integration to compute the elements (local element matrices)  can often be time
consuming. This manuscript is about optimizing the finite element integration
loops and the number of flops therein. The authors propose an approach that is
shown to be locally optimal. An implementation (in COFFEE) is compared against
other methods and speedups over the default generated code is demonstrated in
several cases.

Overall this is a well written and interesting paper, even to non-experts.
However, as a non-expert, I have several concerns that I think should be
addressed:

1) The focus is on minimizing flops, but on modern architectures, memory
accesses are typically more expensive and flops are "almost free". Although
clearly of theoretical interest, it would be nice to justify that this
optimization also is important to reduce run time.

2) All experiments and results are run in serial on single core. In practice,
most computers today are multicore (or manycore) so some discussion about
extensions to multithreaded case would be appropriate. I recognize a parallel
implementation is beyond the scope of the current paper.

3) Results: The authors present several charts with speedups on different
problems, which gives detailed information. However, it would be nice to add a
figure with aggregate results, such as a scatter plot of all instances or a
performance profile.

4) The manuscript is already quite long so if possible, do not increase the
total length (i.e. try find things to tighten or cut).Referee: 1

Comments to the Author This paper is overall well written. Examples are very
illustrative and help understand the article. My only concern is about how
authors relates their theortical results with numerical performance examples.
The main focus of this paper is about transformations that reduces FLOPs.
Authors present elimination and reduction pattern of code and show that there is
an optimal loop transformation that reduces FLOPs. However, the numeric
performance results is about "speed-up" not about FLOPs. Also, the speed-up is
more significant when the polynomial degree is higher. This confuses me either
vectorization improves the performance or actual FLOPs decreases. My suggestions
are as follows: 1. It would be better to include actual transformed code and
original code in the appendix for all computing model (mass, helmholtz,
elasticity, hyper elasticity). At least I wonder what makes hugh speed up in the
hyper elastic problem compared to usual mass matrix.  2. Please include actual
FLOP measurements of input code and transformed code and show some correlation
with performance.

Referee: 2

Comments to the Author

Report on “An algorithm for the optimization of finite element integration
loops”

The paper is well written, interesting does clearly merit publication.  The
usage of advanced compiler technology as demonstrated here is clearly novel and
it brings accelerated performance.

There is one issue that is not covered in the analysis and only briefly
mentioned in the general discussion: the transfer from memory. I understand that
it complicates the analysis but the least it should be quantified in the
numerical experiments. As was shown by Alnæs and Mardal in 2010, TOMS, the
transfer is significant. This part of the story should also be told as I guess
the most common user, low-order elements with simple equation, will not notice
any benefit. This has been an untold story in several FEniCS papers. It is to a
certain extent discussed already in the paper but it should also be quantified.

Referee: 3

Comments to the Author The FEM is widely used in scientific computing and the
integration to compute the elements (local element matrices)  can often be time
consuming. This manuscript is about optimizing the finite element integration
loops and the number of flops therein. The authors propose an approach that is
shown to be locally optimal. An implementation (in COFFEE) is compared against
other methods and speedups over the default generated code is demonstrated in
several cases.

Overall this is a well written and interesting paper, even to non-experts.
However, as a non-expert, I have several concerns that I think should be
addressed:

1) The focus is on minimizing flops, but on modern architectures, memory
accesses are typically more expensive and flops are "almost free". Although
clearly of theoretical interest, it would be nice to justify that this
optimization also is important to reduce run time.

2) All experiments and results are run in serial on single core. In practice,
most computers today are multicore (or manycore) so some discussion about
extensions to multithreaded case would be appropriate. I recognize a parallel
implementation is beyond the scope of the current paper.

3) Results: The authors present several charts with speedups on different
problems, which gives detailed information. However, it would be nice to add a
figure with aggregate results, such as a scatter plot of all instances or a
performance profile.

4) The manuscript is already quite long so if possible, do not increase the
total length (i.e. try find things to tighten or cut).

### 2016-12-23 TOMS_Paper/reviewers_response.txt
Write up response to reviewers

First of all, we would like to thank the referees for their insightful comments,
which helped improving the paper. In this new version, we have attempted to
address all of the referees' concerns. We explain how the paper was changed
below, by considering each of the requests/observations/questions raised by the
three referees.

Referee: 1
==========

| This paper is overall well written. Examples are very illustrative and help
| understand the article. My only concern is about how authors relates their
| theortical results with numerical performance examples. [...]

| My suggestions are as follows:
| 1. It would be better to include actual transformed code and original code in
| the appendix for all computing model (mass, helmholtz, elasticity, hyper
| elasticity). At least I wonder what makes hugh speed up in the hyper elastic
| problem compared to usual mass matrix.

The generated code has been uploaded to GitHub and given a DOI through Zenodo.
This is properly documented in the paper (see Section 7.1 - Experimental setup).
For reasons of space, it simply was not possible to append code in an appendix.

| 2. Please include actual FLOP measurements of input code and transformed code
| and show some correlation with performance.

There is a whole new section treating this fundamental aspect (7.8 - On the
correlation between operation count and performance). Further, a new figure was
added (Fig. 11): this clearly shows the relationship between the achieved
speed-ups and FLOPs reductions. We believe that this figure is particularly
useful, because it shows how well the system presented in the paper scales well
with the complexity of operators. It also provides a nice summary of the
performance improvements achieved. In addition to all this, Sections 7.4 through
7.7 (Mass, Helmholtz, ...) have been extended and now discuss the impact of the
reduced operation counts in a set of relevant test cases.

Referee: 2
==========

| The paper is well written, interesting does clearly merit publication.
| The usage of advanced compiler technology as demonstrated here is clearly
| novel and it brings accelerated performance.

| There is one issue that is not covered in the analysis and only briefly
| mentioned in the general discussion: the transfer from memory. I understand
| that it complicates the analysis but the least it should be quantified in
| the numerical experiments [...] It is to a certain extent discussed already
| in the paper but it should also be quantified.

Sections 7.4 through 7.7 now discuss, to some extent, the memory transfer cost.
We did that only for a small set of (relevant) test cases. We believe that it is
very difficult to thoroughly study this problem in a paper like ours, based on a
high dimensional experimentation space. This would probably require its own
study in a separate paper, introducing specific metrics and showing different
angles of the performance achieved. For all these reasons, we believe that it is
just simpler to refer to a paper that, in fact, has performed an analysis
analogous to what the referee is suggesting: [Bercea et al.  2016].  This is
referenced in Section 7.2, along with a high level discussion of the memory
transfer impact.  Notably important, the Bercea paper is based upon the same
Firedrake technology of our article (in fact, we are amongst its co-authors).

Referee: 3
==========

| Overall this is a well written and interesting paper, even to non-experts.
| However, as a non-expert, I have several concerns that I think should be
| addressed:

| 1) The focus is on minimizing flops, but on modern architectures, memory
| accesses are typically more expensive and flops are "almost free". Although
| clearly of theoretical interest, it would be nice to justify that this
| optimization also is important to reduce run time.

Figures 7, 8, 9, 10 show execution time speed-ups. Hence, we are already showing
that our work contributes to actually reducing run-times.

| 2) All experiments and results are run in serial on single core. In practice,
| most computers today are multicore (or manycore) so some discussion about
| extensions to multithreaded case would be appropriate. I recognize a parallel
| implementation is beyond the scope of the current paper.

We have added one paragraph in Section 9 and added a few sentences throughout
the text about this matter. Basically, in Firedrake parallelism is achieved at
the mesh level (by distributing mesh partitions to processes), while the
execution of a local assembly kernel is sequential. Clearly this might need to
be revisited in the case of higher order methods (as we would end up with big
working sets even for executing just a single element). This is discussed in
Section 9.

| 3) Results: The authors present several charts with speedups on different
| problems, which gives detailed information. However, it would be nice to add a
| figure with aggregate results, such as a scatter plot of all instances or a
| performance profile.

We agree with the referee about the usefulness of such a chart. We have added a
scatter plot (Fig. 11) as well as a section explaining what information we can
extrapolate from it.

| 4) The manuscript is already quite long so if possible, do not increase the
| total length (i.e. try find things to tighten or cut).

We tried our best, but we feel that we cannot really stay within 27 pages
without losing essential information. We observe, however, that other TOMS
papers went even beyond the 27 pages.

### 2016-12-30 TOMS_Paper/assembly.tex
minor fixes

@@
 \begin{description}
-\item[quad] Optimized quadrature mode. Work presented in~\citeN{quadrature1}, implemented in  in the FEniCS Form Compiler. 
+\item[quad] Optimized quadrature mode. Work presented in~\citeN{quadrature1}, implemented in the FEniCS Form Compiler.
 \item[tens] Tensor contraction mode. Work presented in~\citeN{FFC-TC}, implemented in the FEniCS Form Compiler.
@@
 We show one subplot for each problem instance ${\langle} \textrm{form}, \mathrm{nf}, q {\rangle}$, with the code transformation system varying within each subplot.
-The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.
+The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over the code produced by the FEniCS Form Compiler with no transformations applied (also referred to as ``baseline``).
 We note that if a bar or a subplot are missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.
@@
 As reiterated throughout the paper, the execution time of a kernel is influenced not only by the operation count, but also by the chosen code generation strategy (e.g., loop unrolling rather than preserving the loop structure).
-Figure~\ref{fig:flops-correlation} aggregates the \texttt{cfO2} results from the previous sections and displays them as a function of the reduction in executed flops.
+Figure~\ref{fig:flops-correlation} aggregates the \texttt{cfO2} results from the previous sections and displays them as a function of the reduction in flops counted at the source level.
 Besides summarizing the performance improvements achieved in this article, this figure highlights a number of aspects.
 The speed-ups achieved by \texttt{cfO2} tend to increase with the complexity of the operators and $q$.
-Further, the pre-multiplying functions have a limited impact in complex operators, and they occasionally have a negative effect in simple and small problems.
+Further, the pre-multiplying functions have a limited impact in complex operators.
 Finally, this plot shows that even a powerful general-purpose compiler like Intel's is inadequate to fully exploit the structure of non-trivial finite element operators (analogous behaviors were observed with different compilers).

### 2016-12-30 TOMS_Paper/reviewers_response.txt
Update response to reviewers

@@
 which helped improving the paper. In this new version, we have attempted to
-address all of the referees' concerns. We explain how the paper was changed
+address all of the referees' concerns. We explain how the paper has changed
 below, by considering each of the requests/observations/questions raised by the
-three referees.
+three referees individually.
 
@@
 useful, because it shows how well the system presented in the paper scales well
-with the complexity of operators. It also provides a nice summary of the
-performance improvements achieved. In addition to all this, Sections 7.4 through
-7.7 (Mass, Helmholtz, ...) have been extended and now discuss the impact of the
-reduced operation counts in a set of relevant test cases.
+with the complexity of the operators. It also provides a nice summary of the
+performance improvements achieved. Moreover, Sections 7.4 through 7.7 (Mass,
+Helmholtz, ...) have been extended and now also discuss the impact of the reduced
+operation counts in a set of relevant test cases.
 
@@
 
-Sections 7.4 through 7.7 now discuss, to some extent, the memory transfer cost.
-We did that only for a small set of (relevant) test cases. We believe that it is
-very difficult to thoroughly study this problem in a paper like ours, based on a
-high dimensional experimentation space. This would probably require its own
-study in a separate paper, introducing specific metrics and showing different
-angles of the performance achieved. For all these reasons, we believe that it is
-just simpler to refer to a paper that, in fact, has performed an analysis
-analogous to what the referee is suggesting: [Bercea et al.  2016].  This is
+Sections 7.4 through 7.7 now discuss the memory transfer cost, but only in a
+limited set of test cases. We believe that it is very difficult to thoroughly
+study this problem in a paper like ours, being based on a high dimensional
+experimentation space. A complete treatment would probably require a new paper,
+introducing specific metrics and showing different angles of the performance
+achieved.  However, there exists a very recent paper including exactly the type
+of analysis advocated by the referee: [Bercea et al.  2016]. This is now
 referenced in Section 7.2, along with a high level discussion of the memory
-transfer impact.  Notably important, the Bercea paper is based upon the same
-Firedrake technology of our article (in fact, we are amongst its co-authors).
+transfer impact.  It is worth noting that the Bercea paper is based upon the same
+Firedrake technology of our article.
 

### 2017-01-13 TOMS_Paper/reviewers_response.txt
Minor rewording of the response to reviewers

@@
 
-The generated code has been uploaded to GitHub and given a DOI through Zenodo.
-This is properly documented in the paper (see Section 7.1 - Experimental setup).
-For reasons of space, it simply was not possible to append code in an appendix.
+The generated code has been uploaded to GitHub and permanently
+archived with a DOI through Zenodo.  This is properly documented in
+the paper (see Section 7.1 - Experimental setup).  For reasons of
+space, it simply was not possible to append code in an appendix.
 
@@
 
-Sections 7.4 through 7.7 now discuss the memory transfer cost, but only in a
-limited set of test cases. We believe that it is very difficult to thoroughly
-study this problem in a paper like ours, being based on a high dimensional
-experimentation space. A complete treatment would probably require a new paper,
-introducing specific metrics and showing different angles of the performance
-achieved.  However, there exists a very recent paper including exactly the type
-of analysis advocated by the referee: [Bercea et al.  2016]. This is now
-referenced in Section 7.2, along with a high level discussion of the memory
-transfer impact.  It is worth noting that the Bercea paper is based upon the same
-Firedrake technology of our article.
+Sections 7.4 through 7.7 now discuss the memory transfer cost, but
+only in a limited set of test cases. We believe that it is very
+difficult to thoroughly study this problem in a paper like ours, being
+based on a high dimensional experimentation space. A complete
+treatment would probably require a new paper, introducing specific
+metrics and showing different angles of the performance achieved.
+However, with other coauthors we have very recently published a paper
+including exactly the type of analysis advocated by the referee:
+[Bercea et al.  2016]. This is now referenced in Section 7.2, along
+with a high level discussion of the memory transfer impact.  It is
+worth noting that the Bercea paper is based upon the same Firedrake
+technology of our article.
 
@@
 
-We tried our best, but we feel that we cannot really stay within 27 pages
-without losing essential information. We observe, however, that other TOMS
-papers went even beyond the 27 pages.
+We tried our best, but we feel that we cannot really stay within 27
+pages without losing essential information. We observe, however, that
+this paper is within the typical range of page lengths in TOMS.

### 2017-01-16 TOMS_Paper/perf-results/overview/scatterplot.py
Tweak scatterplot (eg larger font)

@@ -52,7 +52,7 @@ ax = fig.add_subplot(1, 1, 1)
 ax.set_title("Correlation between flop reductions and achieved speed-ups", fontsize=12, y=1.03)
 # x-axis
 ax.set_xlabel(r'$\frac{\mathtt{baseline}\ \mathregular{flops}}{\mathtt{cfO2}\ \mathregular{flops}}$',
-              fontsize=15, color='black', labelpad=6.5)
+              fontsize=16, color='black', labelpad=6.5)
 ax.set_xlim([0.0, 1.0])
 ax.set_xscale('symlog')
 # y-axis
@@ -63,7 +63,6 @@ cm = brewer2mpl.get_map('Spectral', 'diverging', 4)
 #markers = ['v', '^', '<', '>']
 #markers = ['s', 'o', '+', 'v']
 markers = ['^', 's', 'p', 'o']
-legend_font = FontProperties(size='x-small')
 
 # Add points
 legend_points = []
@@ -90,7 +89,7 @@ for (prob, nf, q), flops_by_mode in mapper.items():
         legend_points.append((handle, ''))
     if nf == 0 and q == 1:
         handle = ax.plot([], [], ms=0, mew=0, lw=0)[0]
-        legend_points.append((handle, '              '))
+        legend_points.append((handle, ' '*2))
     if nf == 0 and prob == 'mass':
         handle = ax.plot([], [], ms=0, mew=0, lw=0)[0]
         legend_points.append((handle, '$\mathrm{q=%d}$' % q))
@@ -98,7 +97,7 @@ for (prob, nf, q), flops_by_mode in mapper.items():
         handle = ax.plot([], [],
                          marker=markers[index],
                          ms=3*q, mew=1, mfc='black', mec='black', lw=0)[0]
-        legend_points.append((handle, '              '))
+        legend_points.append((handle, ' '*9))
 
 
 # Speed-up >= 1 line
@@ -109,6 +108,7 @@ discrete_cm = cm.get_mpl_colormap().from_list('Custom cmap', cm.mpl_colors, len(
 sm = plt.cm.ScalarMappable(cmap=discrete_cm, norm=plt.Normalize(0, 3))
 sm._A = []
 cbar = fig.colorbar(sm, ax=ax)
+cbar.ax.tick_params(labelsize=11.5)
[…]

### 2017-01-16 TOMS_Paper/perf-results/overview/scatterplot.py
Larger font

@@ -49,15 +49,17 @@ for (prob, mode, nf, q), v in flops.items():
 plt.style.use('ggplot')
 fig = plt.figure()
 ax = fig.add_subplot(1, 1, 1)
-ax.set_title("Correlation between flop reductions and achieved speed-ups", fontsize=12, y=1.03)
+ax.set_title("Correlation between flop reductions and achieved speed-ups", fontsize=13, y=1.03)
 # x-axis
 ax.set_xlabel(r'$\frac{\mathtt{baseline}\ \mathregular{flops}}{\mathtt{cfO2}\ \mathregular{flops}}$',
-              fontsize=16, color='black', labelpad=6.5)
+              fontsize=18, color='black', labelpad=6.5)
 ax.set_xlim([0.0, 1.0])
 ax.set_xscale('symlog')
 # y-axis
-ax.set_ylabel(r'Speedup relative to baseline', fontsize=12, color='black', labelpad=12.0)
+ax.set_ylabel(r'Speedup relative to baseline', fontsize=13, color='black', labelpad=12.0)
 ax.set_ylim([0.0, 1.0])
+
+ax.tick_params(labelsize=13)
 
 cm = brewer2mpl.get_map('Spectral', 'diverging', 4)
 #markers = ['v', '^', '<', '>']
@@ -108,10 +110,10 @@ discrete_cm = cm.get_mpl_colormap().from_list('Custom cmap', cm.mpl_colors, len(
 sm = plt.cm.ScalarMappable(cmap=discrete_cm, norm=plt.Normalize(0, 3))
 sm._A = []
 cbar = fig.colorbar(sm, ax=ax)
-cbar.ax.tick_params(labelsize=11.5)
+cbar.ax.tick_params(labelsize=14)
 cbar.set_ticks(np.array([0, 1, 2, 3], dtype=np.int32))
-cbar.set_ticklabels(np.array([r'$\mathrm{nf = 0}$', r'$\mathrm{nf = 1}$',
-                              r'$\mathrm{nf = 2}$', r'$\mathrm{nf = 3}$']))
+cbar.set_ticklabels(np.array([r'$\mathrm{nf} = 0$', r'$\mathrm{nf} = 1$',
+                              r'$\mathrm{nf} = 2$', r'$\mathrm{nf} = 3$']))
 
 # Legend
 def swap(l, i, j):

### 2017-01-16 TOMS_Paper/assembly.tex

@@
 The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over the code produced by the FEniCS Form Compiler with no transformations applied (also referred to as ``baseline``).
-We note that if a bar or a subplot are missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.
-
-The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained and the relationship with the operation count.
+We note that if a bar or a subplot is missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.
+
+The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we discuss, for each form, the performance results obtained and the relationship with the operation count.
 
 \subsection{High level view}
-Our transformation strategy does not always guarantee minimum execution time.
-In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime.
+Our transformation strategy does not always guarantee minimum execution
+time. In particular, about 5$\%$ of the test cases (3 out of 56, excluding
+marginal differences) \texttt{cfO2} did not produce the code with the lowest
+runtime.
 The most significant of such test cases is the elastic model with $[q=4,\ \mathrm{nf}=0]$.
@@
 
-The memory transfer has sometimes a non-negligible impact, especially when low order polynomials are used.
-This is more evident in the simpler bilinear forms (e.g., \texttt{Mass}), when the speed-ups over the baseline are limited.
-The effect of the memory transfer is quantified, to some extent, in the sections below.
-For a comprehensive analysis of its cost, the reader is invited to refer to~\cite{extrusion-paper}, a study based upon the same Firedrake technology used in this article.
+The cost of transferring data from memory sometimes has a noticable impact
+on performance, especially when
+low order polynomials are used. This is more evident in the simpler bilinear
+forms (e.g., \texttt{Mass}), for which the speed-ups over the baseline are
+limited.
+The effect of memory transfer costs is quantified, to some extent, in the sections below.
+For a comprehensive analysis of its cost, the reader is referred to~\cite{extrusion-paper}, a study based upon the same Firedrake technology used in this article.
 
 \subsection{Autovectorization}
-We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length.
-This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}.
-The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code.
-In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered.
-In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.
+We chose the mesh dimension and the function spaces such that the inner loop
+sizes are always a multiple of the machine vector length. This ensures
+autovectorization in the majority of code variants\footnote{We verified the
+  vectorization of inner loops by looking at both compiler reports and
+  assembly code.}. The most prominent exception is \texttt{quad}, where the
+presence of indirection arrays in the generated code impedes
+vectorization. In \texttt{tens}, loop nests are fully unrolled, so the
+standard loop vectorization is not feasible; the compiler reports suggest,
+however, that block vectorization \cite{SLP-vect} is often triggered. In
+\texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have
+identical structure, and the compuler regularly applies loop vectorization.
 
@@
 
-We first notice the lack of improvements when $q=1$.
-The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreds operations), but nonetheless no gain in execution time is obtained.
+We first notice the lack of improvements when $q=1$. The operation count of
+\texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of
+the alternatives (tens instead of hundreds of operations), but nonetheless no gain in execution time is obtained.
 This is due to the fact that local assembly is memory-bound.
@@
 In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format.
-This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which collocates the kernels in a memory-bound regime.
+This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which locates the kernels in a memory-bound regime.
 
@@
 
-A relevant experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$.
-We found that this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2, q=2\rangle$.
-This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.
+A relevant experiment we additionally performed was relaxing the memory
+threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$. We found that
+this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a
+maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2,
+q=2\rangle$. This effect might be expected to become worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.
 
@@
 
-We also tried to relate the achieved performance to the operation counts obtained by each transformation system.
-We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups were generally massive.
+It is also possible to relate the achieved performance to the operation counts
+obtained by each transformation system. We here limit ourselves to consider
+the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups were
+generally very large.
 In the original code, $37 \cdot 10^9$ operations were counted at the source level.
